### 2019-07-07 README.Rmd
update README, use README.Rmd

---
output: github_document
---

# mlr3spatiotemporal

Spatiotemporal extensions for mlr3

[![Travis build status](https://travis-ci.org/mlr-org/mlr3spatiotemporal.svg?branch=master)](https://travis-ci.org/mlr-org/mlr3spatiotemporal)
[![CRAN status](https://www.r-pkg.org/badges/version/mlr3spatiotemporal)](https://cran.r-project.org/package=mlr3spatiotemporal)
[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Coverage status](https://codecov.io/gh/mlr-org/mlr3spatiotemporal/branch/master/graph/badge.svg)](https://codecov.io/github/mlr-org/mlr3spatiotemporal?branch=master)

This package extends the [mlr3](https://github.com/mlr-org/mlr3) package framework by spatiotemporal resampling and visualization methods.

## Resampling methods

Currently, the following ones are implemented:

- Spatial CV using k-means clustering after Brenning2012 (ResampleCVKmeans) (package [sperrorest](https://github.com/pat-s/sperrorest)))

The following ones are planned:

- "Spatial Blocking" (package [blockCV](https://github.com/rvalavi/blockCV))
- "Environmental Blocking" (package [blockCV](https://github.com/rvalavi/blockCV))
- "Buffering" (package [blockCV](https://github.com/rvalavi/blockCV))

## Spatial tasks

- Task "ecuador" -> `mlr_tasks$get("ecuador")`

## Visualization methods

# References

<div id="refs" class="references">

<div id="ref-Brenning2012">

———. 2012. “Spatial Cross-Validation and Bootstrap for the Assessment of
Prediction Rules in Remote Sensing: The R Package Sperrorest.” In *2012
IEEE International Geoscience and Remote Sensing Symposium*, 5372–5.
<https://doi.org/10.1109/IGARSS.2012.6352393>.

</div>

</div>

### 2019-08-28 README.Rmd
upd README

@@
 
-Spatiotemporal extensions for mlr3
+Spatio-temporal extensions for mlr3
 
+<!-- badges: start -->
 [![Travis build status](https://travis-ci.org/mlr-org/mlr3spatiotemporal.svg?branch=master)](https://travis-ci.org/mlr-org/mlr3spatiotemporal)
 [![CRAN status](https://www.r-pkg.org/badges/version/mlr3spatiotemporal)](https://cran.r-project.org/package=mlr3spatiotemporal)
-[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
+[![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
 [![Coverage status](https://codecov.io/gh/mlr-org/mlr3spatiotemporal/branch/master/graph/badge.svg)](https://codecov.io/github/mlr-org/mlr3spatiotemporal?branch=master)
+<!-- badges: end -->
 
@@
 
-- Spatial CV using k-means clustering after Brenning2012 (ResampleCVKmeans) (package [sperrorest](https://github.com/pat-s/sperrorest)))
-
-The following ones are planned:
-
-- "Spatial Blocking" (package [blockCV](https://github.com/rvalavi/blockCV))
-- "Environmental Blocking" (package [blockCV](https://github.com/rvalavi/blockCV))
-- "Buffering" (package [blockCV](https://github.com/rvalavi/blockCV))
+| Literature Name | R Package | Reference | mlr3spatiotemporal Name |
+| --------------- | --------- | --------- | ----------------------- |
+| Spatial CV  | [sperrorest](https://github.com/giscience-fsu/sperrorest) | Brenning 2012 | `ResampleSpCVCoords`  |
+| Spatial Blocking | [blockCV](https://github.com/rvalavi/blockCV) | Valavi 2019 | `ResampleSpCVBlock`   |
+| Environmental Blocking | [blockCV](https://github.com/rvalavi/blockCV) | Valavi 2019 | `ResampleSpCVEnv`   |
+| Spatial Buffering | [blockCV](https://github.com/rvalavi/blockCV) | Valavi 2019 | `ResampleSpCVBuffer` |
 
@@
 
+Generic S3 function `autoplot()` for all implemented spatial resampling methods.
+
 # References
@@
 
-———. 2012. “Spatial Cross-Validation and Bootstrap for the Assessment of
+**Brenning**, A.. 2012. “Spatial Cross-Validation and Bootstrap for the Assessment of
 Prediction Rules in Remote Sensing: The R Package Sperrorest.” In *2012
@@
 
+<div id="ref-Valavi2019">
+
+**Valavi** R, Elith J, Lahoz-Monfort JJ, Guillera-Arroita G.
+blockCV: An r package for generating spatially or environmentally separated folds for k-fold cross-validation of species distribution models.
+Methods Ecol Evol.
+2019; 10:225–232.
+doi.org/10.1111/2041-210X.13107
+
 </div>
+
+</div>

### 2019-09-03 README.Rmd
. -> ,
add vis example to README

@@
 
+The following example shows the resampling method `"spcv-coords"`.
+
+Visualization of all partitions
+
+```{r spcv-coords-all-partitions}
+library(mlr3)
+library(mlr3spatiotemporal)
+task <- tsk("ecuador")
+resampling = rsmp("spcv-coords")
+resampling$param_set$values = list(folds = 5)
+resampling$instantiate(task)
+
+autoplot(resampling, task)
+```
+
+Visualization of the first fold only
+
+```{r spcv-coords-fold}
+autoplot(resampling, task, fold_id = 1)
+```
+
 # References
@@
 
-**Brenning**, A.. 2012. “Spatial Cross-Validation and Bootstrap for the Assessment of
+**Brenning**, A., 2012. “Spatial Cross-Validation and Bootstrap for the Assessment of
 Prediction Rules in Remote Sensing: The R Package Sperrorest.” In *2012

### 2019-09-03 README.Rmd
add task diplodia

@@
 - Task "ecuador" -> `mlr_tasks$get("ecuador")`
+- Task "diplodia" -> `mlr_tasks$get("diplodia")`
 
@@
 
+```{r setup, include=FALSE}
+set.seed(123)
+```
+
 ```{r spcv-coords-all-partitions}
@@
 library(mlr3spatiotemporal)
+library(ggplot2)
+
 task <- tsk("ecuador")

### 2019-09-03 R/ResamplingSpCVBuffer.R
<- to =

@@ -19,11 +19,11 @@
 #' @examples
 #' \dontrun{
 #' library(mlr3)
-#' task <- tsk("ecuador")
+#' task = tsk("ecuador")
 #'
 #' # Instantiate Resampling
-#' rcv <- rsmp("spcv-buffer")
-#' rcv$param_set$values <- list(range = 1000)
+#' rcv = rsmp("spcv-buffer")
+#' rcv$param_set$values = list(range = 1000)
 #' rcv$instantiate(task)
 #'
 #' # Individual sets:
@@ -34,7 +34,7 @@
 #' # Internal storage:
 #' rcv$instance
 #' }
-ResamplingSpCVBuffer <- R6Class("ResamplingSpCVBuffer",
+ResamplingSpCVBuffer = R6Class("ResamplingSpCVBuffer",
   inherit = mlr3::Resampling,
   public = list(
     initialize = function(id = "spcv-buffer", param_vals = list(range = 100)) {
@@ -51,12 +51,12 @@ ResamplingSpCVBuffer <- R6Class("ResamplingSpCVBuffer",
 
       assert_task(task)
 
-      groups <- task$groups
-      stratify <- self$param_set$values$stratify
+      groups = task$groups
+      stratify = self$param_set$values$stratify
 
       if (length(stratify) == 0L || isFALSE(stratify)) {
         if (is.null(groups)) {
-          instance <- private$.sample(task$row_ids, task$coordinates(), task$crs)
+          instance = private$.sample(task$row_ids, task$coordinates(), task$crs)
         } else {
           stopf("Grouping is not supported for spatial resampling methods.", call. = FALSE)
         }
@@ -67,8 +67,8 @@ ResamplingSpCVBuffer <- R6Class("ResamplingSpCVBuffer",
         stopf("Stratification is not supported for spatial resampling methods.", call. = FALSE)
       }
 
-      self$instance <- instance
-      self$task_hash <- task$hash
+      self$instance = instance
+      self$task_hash = task$hash
       invisible(self)
     }
   ),
@@ -81,11 +81,11 @@ ResamplingSpCVBuffer <- R6Class("ResamplingSpCVBuffer",
 
   private = list(
     .sample = function(ids, coords, crs) {
-      points <- sf::st_as_sf(coords,
+      points = sf::st_as_sf(coords,
         coords = c("x", "y"),
         crs = crs)
 
-      inds <- blockCV::buffering(
+      inds = blockCV::buffering(
         speciesData = points,
[…]

### 2019-09-03 R/Task_classif_diplodia.R

@@ -20,11 +20,11 @@
 #'   Modelling, 406, 109–120. https://doi.org/10/gf34bd
 NULL
 
-load_task_diplodia <- function(id = "diplodia") {
-  b <- as_data_backend(readRDS(system.file("extdata", "diplodia.rda",
+load_task_diplodia = function(id = "diplodia") {
+  b = as_data_backend(readRDS(system.file("extdata", "diplodia.rda",
     package = "mlr3spatiotemporal")))
-  b$hash <- "_mlr3_tasks_diplodia_"
-  task <- TaskClassifST$new(id, b, target = "diplo01", positive = "1",
+  b$hash = "_mlr3_tasks_diplodia_"
+  task = TaskClassifST$new(id, b, target = "diplo01", positive = "1",
     coordinates = c("x", "y"), coords_as_features = FALSE,
     crs = "+proj=utm +zone=30 +south +datum=WGS84 +units=m +no_defs")
 }

### 2019-09-03 R/Task_classif_ecuador.R

@@ -22,11 +22,11 @@
 #'   Geomorphology, 139-140: 271-284.
 NULL
 
-load_task_ecuador <- function(id = "ecuador") {
-  b <- as_data_backend(readRDS(system.file("extdata", "ecuador.rda",
+load_task_ecuador = function(id = "ecuador") {
+  b = as_data_backend(readRDS(system.file("extdata", "ecuador.rda",
     package = "mlr3spatiotemporal")))
-  b$hash <- "_mlr3_tasks_ecuador_"
-  task <- TaskClassifST$new(id, b, target = "slides", positive = "TRUE",
+  b$hash = "_mlr3_tasks_ecuador_"
+  task = TaskClassifST$new(id, b, target = "slides", positive = "TRUE",
     coordinates = c("x", "y"), coords_as_features = FALSE,
     crs = "+proj=utm +zone=17 +south +datum=WGS84 +units=m +no_defs")
 }

### 2019-09-03 R/zzz.R

@@ -7,11 +7,11 @@
 #' @importFrom R6 R6Class
 "_PACKAGE"
 
-.onLoad <- function(libname, pkgname) {
+.onLoad = function(libname, pkgname) {
 
   # add 'coordinates' as col roles
-  tmp <- mlr_reflections$task_col_roles$regr
-  mlr_reflections$task_col_roles <- list(
+  tmp = mlr_reflections$task_col_roles$regr
+  mlr_reflections$task_col_roles = list(
     regr = c(tmp, "coordinates"),
     classif = c(tmp, "coordinates")
   )

### 2019-09-03 tests/testthat/test_mlr_spcv_generic.R

@@ -5,12 +5,12 @@ context("mlr_resampling_spcv-all")
 task = mlr_tasks$get("ecuador")
 
 ### create custom data for grouping variable
-b <- as.data.table(readRDS(system.file("extdata", "ecuador.rda",
+b = as.data.table(readRDS(system.file("extdata", "ecuador.rda",
   package = "mlr3spatiotemporal"
 )))
 # add grouping variable
-data <- insert_named(b[1:150, ], list(grp = rep_len(letters[1:10], 150)))
-task_grp <- TaskClassifST$new("ecuador-grp", as_data_backend(data), target = "slides",
+data = insert_named(b[1:150, ], list(grp = rep_len(letters[1:10], 150)))
+task_grp = TaskClassifST$new("ecuador-grp", as_data_backend(data), target = "slides",
   coordinates = c("x", "y"))
 task_grp$set_col_role("grp", "groups")
 
@@ -32,7 +32,7 @@ test_that("stratification throws errors", {
   spcv_rsp_no_buffer = spcv_rsp
   spcv_rsp_no_buffer$`spcv-buffer` = NULL
   for (i in spcv_rsp_no_buffer) {
-    i$param_set$values <- list(folds = 5, stratify = TRUE)
+    i$param_set$values = list(folds = 5, stratify = TRUE)
     expect_error(i$instantiate(task))
   }
 
@@ -57,7 +57,7 @@ test_that("grouping throws errors when 'groups' and 'stratify' is set", {
   spcv_rsp_no_buffer$`spcv-buffer` = NULL
 
   for (i in spcv_rsp_no_buffer) {
-    i$param_set$values <- list(folds = 5, stratify = TRUE)
+    i$param_set$values = list(folds = 5, stratify = TRUE)
     expect_error(i$instantiate(task_grp), "Grouping is not supported for spatial resampling methods")
   }
 })
@@ -71,7 +71,7 @@ test_that("train and test set getter functions are working", {
 
   for (i in spcv_rsp_no_buffer) {
     # FIXME: seting folds=5 is only needed because of #17
-    i$param_set$values <- list(folds = 4)
+    i$param_set$values = list(folds = 4)
     i$instantiate(task)
     expect_silent(i$train_set(1))
     expect_silent(i$test_set(1))
@@ -85,7 +85,7 @@ test_that("cloning works", {
   )
 
   for (i in spcv_rsp) {
-    clone <- i$clone(deep = TRUE)
[…]

### 2019-09-03 tests/testthat/test_spcv_buffer.R

@@ -5,12 +5,12 @@ context("mlr_resampling_spcv-buffer")
 task = mlr_tasks$get("ecuador")
 
 ### create custom data for grouping variable
-b <- as.data.table(readRDS(system.file("extdata", "ecuador.rda",
+b = as.data.table(readRDS(system.file("extdata", "ecuador.rda",
                                        package = "mlr3spatiotemporal"
 )))
 # add grouping variable
-data <- insert_named(b[1:150, ], list(grp = rep_len(letters[1:10], 150)))
-task_grp <- TaskClassifST$new("ecuador-grp", as_data_backend(data), target = "slides",
+data = insert_named(b[1:150, ], list(grp = rep_len(letters[1:10], 150)))
+task_grp = TaskClassifST$new("ecuador-grp", as_data_backend(data), target = "slides",
                               coordinates = c("x", "y"))
 task_grp$set_col_role("grp", "groups")
 
@@ -52,7 +52,7 @@ test_that("cloning works", {
   )
 
   for (i in spcv_rsp) {
-    clone <- i$clone(deep = TRUE)
+    clone = i$clone(deep = TRUE)
     expect_true(all.equal(i, clone))
   }
 })

### 2019-09-03 README.Rmd
<- to =
Add tests for resampling methods (#18)
* make spcv tests generic

* suggest stringr

* fix typo

* test grouping and stratification combinations

* add spcv-block specific test

* remove unused .combine function

* test setter and getter, test cloning

* fix test and train getter tests

* rename

* move buffer tests into own file

* optimize error message

* add task diplodia

* add spcv-env test file

* <- to =

* fix spcv-env tst

* need assignment operator once

* typo

* format roxygen

* upd README badges

* add test for TaskClassifST

* formot roxygen

* minor notes

* last cover

@@
 [![Travis build status](https://travis-ci.org/mlr-org/mlr3spatiotemporal.svg?branch=master)](https://travis-ci.org/mlr-org/mlr3spatiotemporal)
-[![CRAN status](https://www.r-pkg.org/badges/version/mlr3spatiotemporal)](https://cran.r-project.org/package=mlr3spatiotemporal)
+[![CRAN\_Status\_Badge](http://www.r-pkg.org/badges/version-ago/mlr3spatiotemporal)](https://cran.r-project.org/package=mlr3spatiotemporal)
+[![cran checks](https://cranchecks.info/badges/worst/mlr3spatiotemporal)](https://cran.r-project.org/web/checks/check_results_mlr3spatiotemporal.html)
+[![CRAN Downloads](https://cranlogs.r-pkg.org/badges/mlr3spatiotemporal)](https://cran.rstudio.com/web/packages/mlr3spatiotemporal/index.html)
 [![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
@@
 
-task <- tsk("ecuador")
+task = tsk("ecuador")
 resampling = rsmp("spcv-coords")

### 2019-09-03 R/ResamplingSpCVCoords.R

@@ -3,29 +3,26 @@
 #' @format [R6::R6Class] inheriting from [Resampling].
 #' @import mlr3
 #'
-#' @description
-#' Spatial Cross validation following the "k-means" approach after Brenning 2012.
+#' @description Spatial Cross validation following the "k-means" approach after
+#' Brenning 2012.
 #'
-#' @section Fields:
-#' See [Resampling].
+#' @section Fields: See [Resampling].
 #'
-#' @section Methods:
-#' See [Resampling].
+#' @section Methods: See [Resampling].
 #'
-#' @references
-#' Brenning, A. (2012). Spatial cross-validation and bootstrap for the
-#' assessment of prediction rules in remote sensing: The R package sperrorest.
-#' 2012 IEEE International Geoscience and Remote Sensing Symposium.
+#' @references Brenning, A. (2012). Spatial cross-validation and bootstrap for
+#' the assessment of prediction rules in remote sensing: The R package
+#' sperrorest. 2012 IEEE International Geoscience and Remote Sensing Symposium.
 #' https://doi.org/10.1109/igarss.2012.6352393
 #'
 #' @export
 #' @examples
 #' library(mlr3)
-#' task <- tsk("ecuador")
+#' task = tsk("ecuador")
 #'
 #' # Instantiate Resampling
-#' rcv <- rsmp("spcv-coords")
-#' rcv$param_set$values <- list(folds = 3)
+#' rcv = rsmp("spcv-coords")
+#' rcv$param_set$values = list(folds = 3)
 #' rcv$instantiate(task)
 #'
 #' # Individual sets:
@@ -35,7 +32,7 @@
 #'
 #' # Internal storage:
 #' rcv$instance # table
-ResamplingSpCVCoords <- R6Class("ResamplingSpCVCoords",
+ResamplingSpCVCoords = R6Class("ResamplingSpCVCoords",
   inherit = mlr3::Resampling,
   public = list(
     initialize = function(id = "spcv-coords", param_vals = list(folds = 10L)) {
@@ -51,14 +48,14 @@ ResamplingSpCVCoords <- R6Class("ResamplingSpCVCoords",
     instantiate = function(task) {
 
       assert_task(task)
-      groups <- task$groups
+      groups = task$groups
 
 
-      stratify <- self$param_set$values$stratify
+      stratify = self$param_set$values$stratify
 
       if (length(stratify) == 0L || isFALSE(stratify)) {
         if (is.null(groups)) {
[…]

### 2019-09-03 R/helper.R

@@ -1,3 +1,4 @@
+# FIXME: Remove when exported in mlr3
 translate_types = function(x) {
   r_types = mlr_reflections$task_feature_types
   p_types = names(mlr_reflections$task_feature_types)

### 2019-09-03 R/ResamplingSpCVEnv.R
Add tests for resampling methods (#18)
* make spcv tests generic

* suggest stringr

* fix typo

* test grouping and stratification combinations

* add spcv-block specific test

* remove unused .combine function

* test setter and getter, test cloning

* fix test and train getter tests

* rename

* move buffer tests into own file

* optimize error message

* add task diplodia

* add spcv-env test file

* <- to =

* fix spcv-env tst

* need assignment operator once

* typo

* format roxygen

* upd README badges

* add test for TaskClassifST

* formot roxygen

* minor notes

* last coverage %

@@ -3,26 +3,27 @@
 #' @format [R6::R6Class] inheriting from [Resampling].
 #' @import mlr3
 #'
-#' @description
-#' Environmental Block Cross Validation. This strategy uses k-means clustering to specify blocks of smilar environmental conditions.
-#' Only numeric features can be used. The `features` used for building blocks can be specified in the `param_set`. By default, all numeric features are used.
-#' @section Fields:
-#' See [Resampling].
+#' @description Environmental Block Cross Validation. This strategy uses k-means
+#' clustering to specify blocks of smilar environmental conditions. Only numeric
+#' features can be used. The `features` used for building blocks can be
+#' specified in the `param_set`. By default, all numeric features are used.
+#' @section Fields: See [Resampling].
 #'
-#' @section Methods:
-#' See [Resampling].
+#' @section Methods: See [Resampling].
 #'
-#' @references
-#' Valavi R, Elith J, Lahoz-Monfort JJ, Guillera-Arroita G. blockCV: An r package for generating spatially or environmentally separated folds for k-fold cross-validation of species distribution models. Methods Ecol Evol. 2019; 10:225–232. https://doi.org/10.1111/2041-210X.13107
+#' @references Valavi R, Elith J, Lahoz-Monfort JJ, Guillera-Arroita G. blockCV:
+#' An r package for generating spatially or environmentally separated folds for
+#' k-fold cross-validation of species distribution models. Methods Ecol Evol.
+#' 2019; 10:225–232. https://doi.org/10.1111/2041-210X.13107
 #'
 #' @export
 #' @examples
 #' library(mlr3)
-#' task <- tsk("ecuador")
+#' task = tsk("ecuador")
 #'
 #' # Instantiate Resampling
-#' rcv <- rsmp("spcv-env")
-#' rcv$param_set$values <- list(folds = 4)
+#' rcv = rsmp("spcv-env")
+#' rcv$param_set$values = list(folds = 4)
 #' rcv$instantiate(task)
 #'
 #' # Individual sets:
@@ -32,7 +33,7 @@
 #'
 #' # Internal storage:
 #' rcv$instance
-ResamplingSpCVEnv <- R6Class("ResamplingSpCVEnv",
+ResamplingSpCVEnv = R6Class("ResamplingSpCVEnv",
   inherit = mlr3::Resampling,
   public = list(
     initialize = function(id = "spcv-env", param_vals = list(folds = 10L)) {
[…]

### 2019-09-03 R/TaskClassifST.R

@@ -4,17 +4,17 @@
 #' @import mlr3
 #'
 #' @usage NULL
-#' @format [R6::R6Class] object inheriting from [Task]/[TaskSupervised]/[TaskClassif].
+#' @format [R6::R6Class] object inheriting from
+#'   [Task]/[TaskSupervised]/[TaskClassif].
 #'
-#' @description
-#' This task specializes [Task] and [TaskSupervised] for spatio-temporal classification problems.
-#' The target column is assumed to be a factor.
-#' The `task_type` is set to `"classif"` and `"spatiotemporal"`.
+#' @description This task specializes [Task] and [TaskSupervised] for
+#' spatio-temporal classification problems. The target column is assumed to be a
+#' factor. The `task_type` is set to `"classif"` and `"spatiotemporal"`.
 #'
-#' A spatial example task is available via `mlr_tasks$get("ecuador")`.
-#' During initialization, coordinates need to be passed.
-#' By default, coordinates are not used as features.
-#' This can be changed by setting `coords_as_features = TRUE`.
+#' A spatial example task is available via `mlr_tasks$get("ecuador")`. During
+#' initialization, coordinates need to be passed. By default, coordinates are
+#' not used as features. This can be changed by setting `coords_as_features =
+#' TRUE`.
 #'
 #' @section Construction:
 #' ```
@@ -25,9 +25,9 @@
 #' * `id` :: `character(1)`\cr
 #'   Identifier for the task.
 #'
-#' * `backend` :: [DataBackend]\cr
-#'   Either a [DataBackend], or any object which is convertible to a DataBackend with `as_data_backend()`.
-#'   E.g., a `data.frame()` will be converted to a [DataBackendDataTable].
+#' * `backend` :: [DataBackend]\cr Either a [DataBackend], or any object which
+#' is convertible to a DataBackend with `as_data_backend()`. E.g., a
+#' `data.frame()` will be converted to a [DataBackendDataTable].
 #'
 #' * `target` :: `character(1)`\cr
 #'   Name of the target column.
@@ -73,7 +73,7 @@
 #'
 #' # possible properties:
 #' mlr3::mlr_reflections$task_properties$classif
-TaskClassifST <- R6::R6Class("TaskClassifST",
+TaskClassifST = R6::R6Class("TaskClassifST",
   inherit = TaskClassif,
   public = list(
     crs = NULL,
[…]

### 2019-09-03 tests/testthat/test_autoplot.R

@@ -19,7 +19,7 @@ vdiffr::expect_doppelganger("SpCVCoords - Fold 1-4", coords3)
 # SpCVBlock --------------------------------------------------------------------
 
 resa_Block = ResamplingSpCVBlock$new()
-resa_Block$param_set$values <- list(folds = 4)
+resa_Block$param_set$values = list(folds = 4)
 resa_Block$instantiate(task)
 Block1 = autoplot(resa_Block, task)
 Block2 = autoplot(resa_Block, task, 1)

### 2019-09-03 tests/testthat/test_mlr_spcv_generic.R

@@ -1,0 +1,91 @@
+context("mlr_resampling_spcv-all")
+
+# init objects -----------------------------------------------------------------
+
+task = mlr_tasks$get("ecuador")
+
+### create custom data for grouping variable
+b = as.data.table(readRDS(system.file("extdata", "ecuador.rda",
+  package = "mlr3spatiotemporal"
+)))
+# add grouping variable
+data = insert_named(b[1:150, ], list(grp = rep_len(letters[1:10], 150)))
+task_grp = TaskClassifST$new("ecuador-grp", as_data_backend(data), target = "slides",
+  coordinates = c("x", "y"))
+task_grp$set_col_role("grp", "groups")
+
+# run tests --------------------------------------------------------------------
+
+test_that("spcv has no duplicated ids", {
+  spcv_rsp = mlr_resamplings$mget(
+    as.data.table(mlr_resamplings)[map_lgl(key, stringr::str_detect, pattern = "spcv"), key]
+  )
+  for (i in spcv_rsp) {
+    expect_identical(i$duplicated_ids, FALSE)
+  }
+})
+
+test_that("stratification throws errors", {
+  spcv_rsp = mlr_resamplings$mget(
+    as.data.table(mlr_resamplings)[map_lgl(key, stringr::str_detect, pattern = "spcv"), key]
+  )
+  spcv_rsp_no_buffer = spcv_rsp
+  spcv_rsp_no_buffer$`spcv-buffer` = NULL
+  for (i in spcv_rsp_no_buffer) {
+    i$param_set$values = list(folds = 5, stratify = TRUE)
+    expect_error(i$instantiate(task))
+  }
+
+})
+
+test_that("grouping throws errors when 'groups' is set", {
+  spcv_rsp = mlr_resamplings$mget(
+    as.data.table(mlr_resamplings)[map_lgl(key, stringr::str_detect, pattern = "spcv"), key]
+  )
+  spcv_rsp_no_buffer = spcv_rsp
+  spcv_rsp_no_buffer$`spcv-buffer` = NULL
+  for (i in spcv_rsp) {
+    expect_error(i$instantiate(task_grp), "Grouping is not supported for spatial resampling methods")
+  }
+})
+
+test_that("grouping throws errors when 'groups' and 'stratify' is set", {
+  spcv_rsp = mlr_resamplings$mget(
[…]

### 2019-09-03 tests/testthat/test_spcv_block.R

@@ -1,0 +1,10 @@
+context("spcv-block")
+
+rsp = mlr_resamplings$get("spcv-block")
+task = mlr_tasks$get("ecuador")
+
+test_that("cols and rows are ignored when range is given", {
+  rsp = mlr_resamplings$get("spcv-block")
+  rsp$param_set$values = list(cols = 2, rows = 3, range = 100, folds = 4)
+  expect_warning(rsp$instantiate(task), "Cols and rows are ignored. Range is used to generated blocks")
+})

### 2019-09-04 README.Rmd

@@
 
-**Brenning**, A., 2012. “Spatial Cross-Validation and Bootstrap for the Assessment of
-Prediction Rules in Remote Sensing: The R Package Sperrorest.” In *2012
+**Brenning**, A, 2012. “Spatial cross-validation and bootstrap for the assessment of
+prediction rules in remote sensing: The R Package sperrorest.” In *2012
 IEEE International Geoscience and Remote Sensing Symposium*, 5372–5.
@@
 **Valavi** R, Elith J, Lahoz-Monfort JJ, Guillera-Arroita G.
-blockCV: An r package for generating spatially or environmentally separated folds for k-fold cross-validation of species distribution models.
+blockCV: An R package for generating spatially or environmentally separated folds for k-fold cross-validation of species distribution models.
 Methods Ecol Evol.

### 2019-09-17 R/ResamplingSpCVBuffer.R
update README, remove param_vals

@@ -37,14 +37,15 @@
 ResamplingSpCVBuffer = R6Class("ResamplingSpCVBuffer",
   inherit = mlr3::Resampling,
   public = list(
-    initialize = function(id = "spcv-buffer", param_vals = list(range = 100)) {
-      super$initialize(
-        id = id,
-        param_set = ParamSet$new(params = list(
+    initialize = function(id = "spcv-buffer") {
+      ps = ParamSet$new(params = list(
           ParamUty$new("stratify", default = NULL),
           ParamInt$new("range", lower = 1L, tags = "required")
-        )),
-        param_vals = param_vals
+      ))
+      ps$values = list(range = 100)
+      super$initialize(
+        id = id,
+        param_set = ps
       )
       require_namespaces(c("blockCV", "sf"))
     },

### 2019-09-17 inst/spatial-plots.R

@@ -6,7 +6,7 @@ set.seed(123)
 # kmeans and block
 
 task = mlr_tasks$get("ecuador")
-resampling_cv = mlr_resamplings$get("spcv-block", param_vals = list(folds = 4))
+resampling_cv = rsmp("spcv-block", folds = 4)
 resampling_cv$instantiate(task)
 
 coords = task$coordinates()
@@ -25,7 +25,7 @@ ggplot() +
 # buffer
 
 task = mlr_tasks$get("ecuador")
-resampling_cv = mlr_resamplings$get("spcv-buffer", param_vals = list(range = 3000))
+resampling_cv = rsmp("spcv-buffer", range = 3000)
 resampling_cv$instantiate(task)
 
 coords = task$coordinates()

### 2019-09-17 README.Rmd
update README, remove param_vals

@@
 [![Travis build status](https://travis-ci.org/mlr-org/mlr3spatiotemporal.svg?branch=master)](https://travis-ci.org/mlr-org/mlr3spatiotemporal)
-[![CRAN\_Status\_Badge](http://www.r-pkg.org/badges/version-ago/mlr3spatiotemporal)](https://cran.r-project.org/package=mlr3spatiotemporal)
+[![CRAN Status Badge](https://www.r-pkg.org/badges/version-ago/mlr3spatiotemporal)](https://cran.r-project.org/package=mlr3spatiotemporal)
 [![cran checks](https://cranchecks.info/badges/worst/mlr3spatiotemporal)](https://cran.r-project.org/web/checks/check_results_mlr3spatiotemporal.html)
@@
 task = tsk("ecuador")
-resampling = rsmp("spcv-coords")
-resampling$param_set$values = list(folds = 5)
+resampling = rsmp("spcv-coords", folds = 5)
 resampling$instantiate(task)

### 2019-10-22 README.Rmd
pkgdown: use mlr3pkgdowntemplate
fix readme figs

@@
 
-Spatio-temporal extensions for mlr3
+Spatio-temporal extensions for mlr3.
 
@@
 [![Travis build status](https://travis-ci.org/mlr-org/mlr3spatiotemporal.svg?branch=master)](https://travis-ci.org/mlr-org/mlr3spatiotemporal)
-[![CRAN Status Badge](https://www.r-pkg.org/badges/version-ago/mlr3spatiotemporal)](https://cran.r-project.org/package=mlr3spatiotemporal)
 [![cran checks](https://cranchecks.info/badges/worst/mlr3spatiotemporal)](https://cran.r-project.org/web/checks/check_results_mlr3spatiotemporal.html)
-[![CRAN Downloads](https://cranlogs.r-pkg.org/badges/mlr3spatiotemporal)](https://cran.rstudio.com/web/packages/mlr3spatiotemporal/index.html)
 [![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
@@
 set.seed(123)
+knitr::opts_chunk$set(fig.path = "man/figures/README-")
 ```

### 2019-10-22 README.Rmd
use .bib file in README

@@
 output: github_document
+bibliography: inst/references.bib
+nocite: | 
+  @brenning2012, @valavi2018
 ---
@@
 # References
-
-<div id="refs" class="references">
-
-<div id="ref-Brenning2012">
-
-**Brenning**, A, 2012. “Spatial cross-validation and bootstrap for the assessment of
-prediction rules in remote sensing: The R Package sperrorest.” In *2012
-IEEE International Geoscience and Remote Sensing Symposium*, 5372–5.
-<https://doi.org/10.1109/IGARSS.2012.6352393>.
-
-</div>
-
-<div id="ref-Valavi2019">
-
-**Valavi** R, Elith J, Lahoz-Monfort JJ, Guillera-Arroita G.
-blockCV: An R package for generating spatially or environmentally separated folds for k-fold cross-validation of species distribution models.
-Methods Ecol Evol.
-2019; 10:225–232.
-doi.org/10.1111/2041-210X.13107
-
-</div>
-
-</div>

### 2019-11-15 README.Rmd
mlr3spatiotemporal -> mlr3spatiotempcv

@@
 
-# mlr3spatiotemporal
+# mlr3spatiotempcv
 
-Spatio-temporal extensions for mlr3.
+Spatio-temporal resampling methods for mlr3.
 
 <!-- badges: start -->
-[![Travis build status](https://travis-ci.org/mlr-org/mlr3spatiotemporal.svg?branch=master)](https://travis-ci.org/mlr-org/mlr3spatiotemporal)
-[![cran checks](https://cranchecks.info/badges/worst/mlr3spatiotemporal)](https://cran.r-project.org/web/checks/check_results_mlr3spatiotemporal.html)
+[![Travis build status](https://travis-ci.org/mlr-org/mlr3spatiotempcv.svg?branch=master)](https://travis-ci.org/mlr-org/mlr3spatiotempcv)
+[![cran checks](https://cranchecks.info/badges/worst/mlr3spatiotempcv)](https://cran.r-project.org/web/checks/check_results_mlr3spatiotempcv.html)
 [![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
-[![Coverage status](https://codecov.io/gh/mlr-org/mlr3spatiotemporal/branch/master/graph/badge.svg)](https://codecov.io/github/mlr-org/mlr3spatiotemporal?branch=master)
+[![Coverage status](https://codecov.io/gh/mlr-org/mlr3spatiotempcv/branch/master/graph/badge.svg)](https://codecov.io/github/mlr-org/mlr3spatiotempcv?branch=master)
 <!-- badges: end -->
@@
 
-| Literature Name | R Package | Reference | mlr3spatiotemporal Name |
+| Literature Name | R Package | Reference | mlr3spatiotempcv Name |
 | --------------- | --------- | --------- | ----------------------- |
@@
 library(mlr3)
-library(mlr3spatiotemporal)
+library(mlr3spatiotempcv)
 library(ggplot2)

### 2019-11-15 R/Task_classif_diplodia.R

@@ -22,7 +22,7 @@ NULL
 
 load_task_diplodia = function(id = "diplodia") {
   b = as_data_backend(readRDS(system.file("extdata", "diplodia.rda",
-    package = "mlr3spatiotemporal")))
+    package = "mlr3spatiotempcv")))
   b$hash = "_mlr3_tasks_diplodia_"
   task = TaskClassifST$new(id, b, target = "diplo01", positive = "1",
     coordinates = c("x", "y"), coords_as_features = FALSE,

### 2019-11-15 inst/mlr3spatiotemporal_test.R

@@ -1,5 +1,5 @@
 library(mlr3)
-library(mlr3spatiotemporal)
+library(mlr3spatiotempcv)
 library(ggplot2)
 
 library(blockCV)

### 2019-11-15 tests/testthat/test_TaskClassifST.R

@@ -3,7 +3,7 @@ context("TaskClassifST")
 test_that("coordinates can be used as features", {
 
   b = as_data_backend(readRDS(system.file("extdata", "ecuador.rda",
-                                          package = "mlr3spatiotemporal")))
+                                          package = "mlr3spatiotempcv")))
   b$hash = "_mlr3_tasks_ecuador_"
   task = TaskClassifST$new("test", b, target = "slides", positive = "TRUE",
                            coordinates = c("x", "y"), coords_as_features = TRUE,

### 2019-11-15 R/TaskClassifST.R
mlr3spatiotemporal -> mlr3spatiotempcv

@@ -58,7 +58,7 @@
 #' @export
 #' @examples
 #' data = data.table::as.data.table(readRDS(system.file("extdata", "ecuador.rda",
-#'   package = "mlr3spatiotemporal")))
+#'   package = "mlr3spatiotempcv")))
 #' task = TaskClassifST$new("ecuador", backend = data, target = "slides",
 #'   positive = "TRUE", coordinates = c("x", "y"))
 #'

### 2019-11-15 tests/testthat.R

@@ -1,5 +1,5 @@
 library(checkmate)
 library(testthat)
-library(mlr3spatiotemporal)
+library(mlr3spatiotempcv)
 
-test_check("mlr3spatiotemporal")
+test_check("mlr3spatiotempcv")

### 2019-11-15 tests/testthat/test_mlr_spcv_generic.R

@@ -6,7 +6,7 @@ task = mlr_tasks$get("ecuador")
 
 ### create custom data for grouping variable
 b = as.data.table(readRDS(system.file("extdata", "ecuador.rda",
-  package = "mlr3spatiotemporal"
+  package = "mlr3spatiotempcv"
 )))
 # add grouping variable
 data = insert_named(b[1:150, ], list(grp = rep_len(letters[1:10], 150)))

### 2019-11-15 tests/testthat/test_spcv_buffer.R

@@ -6,7 +6,7 @@ task = mlr_tasks$get("ecuador")
 
 ### create custom data for grouping variable
 b = as.data.table(readRDS(system.file("extdata", "ecuador.rda",
-                                       package = "mlr3spatiotemporal"
+                                       package = "mlr3spatiotempcv"
 )))
 # add grouping variable
 data = insert_named(b[1:150, ], list(grp = rep_len(letters[1:10], 150)))

### 2019-12-29 README.Rmd
update README and pkgdown

@@
 <!-- badges: start -->
-[![Travis build status](https://travis-ci.org/mlr-org/mlr3spatiotempcv.svg?branch=master)](https://travis-ci.org/mlr-org/mlr3spatiotempcv)
-[![cran checks](https://cranchecks.info/badges/worst/mlr3spatiotempcv)](https://cran.r-project.org/web/checks/check_results_mlr3spatiotempcv.html)
+[![Build Status](https://img.shields.io/travis/mlr-org/mlr3spatiotempcv/master?label=Linux&logo=travis&style=flat-square)](https://travis-ci.org/mlr-org/mlr3spatiotempcv)
+[![CRAN Status Badge](https://www.r-pkg.org/badges/version-ago/mlr3spatiotempcv)](https://cran.r-project.org/package=mlr3spatiotempcv)
 [![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)

### 2020-01-10 README.Rmd
README: add circle ci badge, upd .bib, upd images

--- 2019-11-15 README.Rmd
@@
 <!-- badges: start -->
-[![Travis build status](https://travis-ci.org/mlr-org/mlr3spatiotempcv.svg?branch=master)](https://travis-ci.org/mlr-org/mlr3spatiotempcv)
-[![cran checks](https://cranchecks.info/badges/worst/mlr3spatiotempcv)](https://cran.r-project.org/web/checks/check_results_mlr3spatiotempcv.html)
+[![Build Status](https://img.shields.io/travis/mlr-org/mlr3spatiotempcv/master?label=Linux&logo=travis&style=flat-square)](https://travis-ci.org/mlr-org/mlr3spatiotempcv)
+[![CircleCI build status](https://circleci.com/gh/mlr-org/mlr3spatiotempcv.svg?style=svg)](https://circleci.com/gh/mlr-org/mlr3spatiotempcv)
+[![CRAN Status Badge](https://www.r-pkg.org/badges/version-ago/mlr3spatiotempcv)](https://cran.r-project.org/package=mlr3spatiotempcv)
 [![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
@@
 
-| Literature Name | R Package | Reference | mlr3spatiotempcv Name |
-| --------------- | --------- | --------- | ----------------------- |
-| Spatial CV  | [sperrorest](https://github.com/giscience-fsu/sperrorest) | Brenning 2012 | `ResampleSpCVCoords`  |
-| Spatial Blocking | [blockCV](https://github.com/rvalavi/blockCV) | Valavi 2019 | `ResampleSpCVBlock`   |
-| Environmental Blocking | [blockCV](https://github.com/rvalavi/blockCV) | Valavi 2019 | `ResampleSpCVEnv`   |
-| Spatial Buffering | [blockCV](https://github.com/rvalavi/blockCV) | Valavi 2019 | `ResampleSpCVBuffer` |
+| Literature Name        | R Package                                                 | Reference     | mlr3spatiotempcv Class Name | mlr3 sugar            |
+| ---------------        | ---------                                                 | ---------     | ---------------------       | --------------------- |
+| Spatial Buffering      | [blockCV](https://github.com/rvalavi/blockCV)             | Valavi 2019   | `ResampleSpCVBuffer`        | `rsmp("spcv-buffer")` |
+| Spatial Blocking       | [blockCV](https://github.com/rvalavi/blockCV)             | Valavi 2019   | `ResampleSpCVBlock`         | `rsmp("spcv-block")`  |
+| Spatial CV             | [sperrorest](https://github.com/giscience-fsu/sperrorest) | Brenning 2012 | `ResampleSpCVCoords`        | `rsmp("spcv-coords")` |
+| Environmental Blocking | [blockCV](https://github.com/rvalavi/blockCV)             | Valavi 2019   | `ResampleSpCVEnv`           | `rsmp("spcv-env")`    |
 

### 2020-01-10 README.Rmd
update README

@@
 
-| Literature Name        | R Package                                                 | Reference     | mlr3spatiotempcv Class Name | mlr3 sugar            |
-| ---------------        | ---------                                                 | ---------     | ---------------------       | --------------------- |
+| Literature             | Package                                                   | Reference     | Class Name           | mlr3 sugar            |
+| ---------------        | ---------                                                 | ---------     | -------------------- | --------------------- |
 | Spatial Buffering      | [blockCV](https://github.com/rvalavi/blockCV)             | Valavi 2019   | `ResampleSpCVBuffer` | `rsmp("spcv-buffer")` |
@@
 
-- Task "ecuador" -> `mlr_tasks$get("ecuador")`
-- Task "diplodia" -> `mlr_tasks$get("diplodia")`
+| Name     | Code                        | Type    |
+| ---      | ---                         | ---     |
+| ecuador  | `mlr_tasks$get("ecuador")`  | Classif |
+| diplodia | `mlr_tasks$get("diplodia")` | Classif |
 
-## Visualization methods
+## Visualization
 
@@
 
-The following example shows the resampling method `"spcv-coords"`.
-
-Visualization of all partitions
+### Visualization of all partitions
 
@@
 
-```{r spcv-coords-all-partitions}
+```{r spcv-coords-all-partitions, warning=FALSE}
 library(mlr3)
@@
 
-Visualization of the first fold only
+### Visualization of the first fold only
 
@@
 
+## More resources
+
+For detailed information on how to use spatial resampling in {mlr3} please read the section about [spatial analysis in the mlr3 book](https://mlr3book.mlr-org.com/spatial.html).
+
 # References

### 2020-01-13 README.Rmd
start a vignette

@@
 nocite: | 
-  @brenning2012, @valavi2018
+  @brenning2012, @valavi2018, @schratz2019
 ---
@@
 
-| Literature             | Package                                                   | Reference     | Class Name           | mlr3 sugar            |
+| Literature             | Package                                                   | Reference     | Class Name                      | mlr3 Sugar                     |
 | ---------------        | ---------                                                 | ---------     | --------------------            | ---------------------          |
-| Spatial Buffering      | [blockCV](https://github.com/rvalavi/blockCV)             | Valavi 2019   | `ResampleSpCVBuffer` | `rsmp("spcv-buffer")` |
-| Spatial Blocking       | [blockCV](https://github.com/rvalavi/blockCV)             | Valavi 2019   | `ResampleSpCVBlock`  | `rsmp("spcv-block")`  |
-| Spatial CV             | [sperrorest](https://github.com/giscience-fsu/sperrorest) | Brenning 2012 | `ResampleSpCVCoords` | `rsmp("spcv-coords")` |
-| Environmental Blocking | [blockCV](https://github.com/rvalavi/blockCV)             | Valavi 2019   | `ResampleSpCVEnv`    | `rsmp("spcv-env")`    |
+| Spatial Buffering      | [blockCV](https://github.com/rvalavi/blockCV)             | Valavi 2019   | `ResamplingSpCVBuffer`          | `rsmp("spcv-buffer")`          |
+| Spatial Blocking       | [blockCV](https://github.com/rvalavi/blockCV)             | Valavi 2019   | `ResamplingSpCVBlock`           | `rsmp("spcv-block")`           |
+| Spatial CV             | [sperrorest](https://github.com/giscience-fsu/sperrorest) | Brenning 2012 | `ResamplingSpCVCoords`          | `rsmp("spcv-coords")`          |
+| Environmental Blocking | [blockCV](https://github.com/rvalavi/blockCV)             | Valavi 2019   | `ResamplingSpCVEnv`             | `rsmp("spcv-env")`             |
+| ---                    | ---                                                       | ---           | ---                             | ---                            |
+| Repeated Spatial CV    | [sperrorest](https://github.com/giscience-fsu/sperrorest) | Brenning 2012 | `RepeatedRResamplingSpCVCoords` | `rsmp("repeated-spcv-coords")` |
 

### 2020-01-15 R/ResamplingRepeatedSpCVCoords.R
Add class RepeatedSpCVCoords + vignette (#30)
* add class `RepeatedSpCVCoords`

* skip vdiffr tests on CI

* add a "Getting Started" vignette

@@ -1,0 +1,122 @@
+#' @title Repeated Spatial Cross Validation Resampling
+#'
+#' @import mlr3
+#'
+#' @description Spatial Cross validation following the "k-means" approach after
+#' Brenning 2012.
+#'
+#' @references
+#' \cite{mlr3spatiotempcv}{brenning2012}
+#'
+#' @export
+#' @examples
+#' library(mlr3)
+#' task = tsk("diplodia")
+#'
+#' # Instantiate Resampling
+#' rrcv = rsmp("repeated-spcv-coords")
+#' rrcv$param_set$values = list(folds = 3, repeats = 5)
+#' rrcv$instantiate(task)
+#'
+#' # Individual sets:
+#' rrcv$iters
+#' rrcv$folds(1:6)
+#' rrcv$repeats(1:6)
+#'
+#' # Individual sets:
+#' rrcv$train_set(1)
+#' rrcv$test_set(1)
+#' intersect(rrcv$train_set(1), rrcv$test_set(1))
+#'
+#' # Internal storage:
+#' rrcv$instance # table
+ResamplingRepeatedSpCVCoords = R6Class("ResamplingRepeatedSpCVCoords",
+  inherit = mlr3::Resampling,
+
+  public = list(
+    #' @description
+    #' Create an "coordinate-based" repeated resampling instance.
+    #' @param id `character(1)`\cr
+    #'   Identifier for the resampling strategy.
+    initialize = function(id = "repeated-spcv-coords") {
+      ps = ParamSet$new(params = list(
+        ParamInt$new("repeats", lower = 1),
+        ParamInt$new("folds", lower = 1L, tags = "required")
+      ))
+      ps$values = list(folds = 10L)
+      super$initialize(
+        id = id,
+        param_set = ps,
+        man = "mlr3spatiotempcv::mlr_resamplings_repeated_spcvcoords"
+      )
+
+    },
+
+    folds = function(iters) {
+      iters = assert_integerish(iters, any.missing = FALSE, coerce = TRUE)
+      ((iters - 1L) %% as.integer(self$param_set$values$repeats)) + 1L
+    },
+
+    repeats = function(iters) {
+      iters = assert_integerish(iters, any.missing = FALSE, coerce = TRUE)
+      ((iters - 1L) %/% as.integer(self$param_set$values$folds)) + 1L
+    },
+
+    #' @description
+    #'  Materializes fixed training and test splits for a given task.
+    #' @param task [Task]\cr
+    #'  A task to instantiate.
[…]

### 2020-01-15 R/autoplot.R

@@ -13,6 +13,8 @@
 #'   mlr3 task object.
 #' @param fold_id [numeric]\cr
 #'   Fold IDs to plot.
+#' @param repeats_id [numeric]\cr
+#'   Repetition ID to plot.
 #' @param grid (`logical(1)`)\cr Should a gridded plot using
 #'   `cowplot::plot_grid()` be created? If `FALSE` only a list with all
 #'   \CRANpkg{ggplot2} resamplings is returned. Default is `TRUE`. Only applies
@@ -164,10 +166,44 @@ autoplot.ResamplingSpCVCoords = function(
     grid = grid)
 }
 
+#' @title Plot for Repeated Spatial Resampling
+#'
+#' @rdname autoplot_spatial_resampling
+#' @export
+#' @examples
+#' #####
+#' # RepeatedSpCVCoords
+#' #####
+#' \donttest{
+#' task = tsk("diplodia")
+#' resampling = rsmp("repeated-spcv-coords", folds = 10, repeats = 2)
+#' resampling$instantiate(task)
+#' autoplot(resampling, task)
+#' autoplot(resampling, task, 1)
+#' autoplot(resampling, task, fold_id = 2, repeats_id = 2)
+#' autoplot(resampling, task, c(1, 2, 3, 4))
+#' }
+autoplot.ResamplingRepeatedSpCVCoords = function(
+  object,
+  task,
+  fold_id = NULL,
+  repeats_id = 1,
+  grid = TRUE,
+  train_color = "#0072B5",
+  test_color = "#E18727",
+  ...) {
+  autoplot_spatial(resampling = object,
+    task = task,
+    fold_id = fold_id,
+    repeats_id = repeats_id,
+    grid = grid)
+}
+
 autoplot_spatial = function(
   resampling = NULL,
   task = NULL,
   fold_id = NULL,
+  repeats_id = NULL,
   grid = NULL,
   train_color = NULL,
   test_color = NULL) {
@@ -176,7 +212,16 @@ autoplot_spatial = function(
   coords$row_id = task$row_ids
   require_namespaces(c("sf", "cowplot"))
 
+  # instantiante if not yet done
+  if (!resampling$is_instantiated) {
+    resampling = resampling$instantiate(task)
+  }
+
   coords_resamp = merge(coords, resampling$instance, by = "row_id")
+
+  if (grepl("Repeated", class(resampling)[1])) {
+    coords_resamp = coords_resamp[rep == repeats_id, ]
+  }
 
[…]

### 2020-01-15 README.Rmd
Add class RepeatedSpCVCoords + vignette (#30)
* add class `RepeatedSpCVCoords`

* skip vdiffr tests on CI

* add a "Getting Started" vignette

@@
 | ---                    | ---                                                       | ---           | ---                             | ---                            |
-| Repeated Spatial CV    | [sperrorest](https://github.com/giscience-fsu/sperrorest) | Brenning 2012 | `RepeatedRResamplingSpCVCoords` | `rsmp("repeated-spcv-coords")` |
+| Repeated Spatial CV    | [sperrorest](https://github.com/giscience-fsu/sperrorest) | Brenning 2012 | `RepeatedResamplingSpCVCoords` | `rsmp("repeated-spcv-coords")` |
 
@@
 
-For detailed information on how to use spatial resampling in {mlr3} please read the section about [spatial analysis in the mlr3 book](https://mlr3book.mlr-org.com/spatial.html).
+For detailed information on how to use spatial resampling in {mlr3} please read the section about [spatial analysis in the mlr3 book](https://mlr3book.mlr-org.com/spatial.html) and consult the [Getting Started](https://mlr3spatiotempcv.mlr-org.com/articles/mlr3spatiotempcv.html) vignette.
 

### 2020-01-15 tests/testthat/test_mlr_spcv_generic.R

@@ -6,6 +6,7 @@ test_that("spcv has no duplicated ids", {
   spcv_rsp = mlr_resamplings$mget(
     as.data.table(mlr_resamplings)[map_lgl(key, grepl, pattern = "spcv"), key]
   )
+  spcv_rsp$`repeated-spcv-coords` = NULL
   for (i in spcv_rsp) {
     expect_identical(i$duplicated_ids, FALSE)
   }
@@ -15,9 +16,9 @@ test_that("stratification throws errors", {
   spcv_rsp = mlr_resamplings$mget(
     as.data.table(mlr_resamplings)[map_lgl(key, grepl, pattern = "spcv"), key]
   )
-  spcv_rsp_no_buffer = spcv_rsp
-  spcv_rsp_no_buffer$`spcv-buffer` = NULL
-  for (i in spcv_rsp_no_buffer) {
+  spcv_rsp$`spcv-buffer` = NULL
+  spcv_rsp$`repeated-spcv-coords` = NULL
+  for (i in spcv_rsp) {
     i$param_set$values = list(folds = 5, stratify = TRUE)
     expect_error(i$instantiate(task))
   }
@@ -29,6 +30,7 @@ test_that("grouping throws errors when 'groups' is set", {
   )
   spcv_rsp_no_buffer = spcv_rsp
   spcv_rsp_no_buffer$`spcv-buffer` = NULL
+  spcv_rsp$`repeated-spcv-coords` = NULL
   for (i in spcv_rsp) {
     expect_error(i$instantiate(task_grp),
       "Grouping is not supported for spatial resampling methods")
@@ -39,10 +41,10 @@ test_that("grouping throws errors when 'groups' and 'stratify' is set", {
   spcv_rsp = mlr_resamplings$mget(
     as.data.table(mlr_resamplings)[map_lgl(key, grepl, pattern = "spcv"), key]
   )
-  spcv_rsp_no_buffer = spcv_rsp
-  spcv_rsp_no_buffer$`spcv-buffer` = NULL
+  spcv_rsp$`spcv-buffer` = NULL
+  spcv_rsp$`repeated-spcv-coords` = NULL
 
-  for (i in spcv_rsp_no_buffer) {
+  for (i in spcv_rsp) {
     i$param_set$values = list(folds = 5, stratify = TRUE)
     expect_error(i$instantiate(task_grp),
       "Grouping is not supported for spatial resampling methods")
[…]

### 2020-01-15 tests/testthat/test_spcv_generic.R

@@ -19,6 +19,7 @@ test_that("stratification throws errors", {
   # for some tests
   spcv_rsp_no_buffer = spcv_rsp
   spcv_rsp_no_buffer$`spcv-buffer` = NULL
+  spcv_rsp_no_buffer$`repeated-spcv-coords` = NULL
 
   for (i in spcv_rsp_no_buffer) {
     i$param_set$values = list(folds = 5, stratify = TRUE)
@@ -45,17 +46,19 @@ test_that("grouping throws errors when 'groups' and 'stratify' is set", {
   spcv_rsp = mlr_resamplings$mget(
     as.data.table(mlr_resamplings)[map_lgl(key, grepl, pattern = "spcv"), key]
   )
-  spcv_rsp_no_buffer = spcv_rsp
-  spcv_rsp_no_buffer$`spcv-buffer` = NULL
+  spcv_rsp = spcv_rsp
+  spcv_rsp$`spcv-buffer` = NULL
+  spcv_rsp$`repeated-spcv-coords` = NULL
 
-  for (i in spcv_rsp_no_buffer) {
+  for (i in spcv_rsp) {
     i$param_set$values = list(folds = 5, stratify = TRUE)
     expect_error(i$instantiate(task_grp),
       "Grouping is not supported for spatial resampling methods")
   }
 
-  spcv_rsp$`spcv-buffer`$param_set$values = list(stratify = TRUE, range = 100)
-  expect_error(spcv_rsp$`spcv-buffer`$instantiate(task_grp),
+  buffer = rsmp("spcv-buffer")
+  buffer$param_set$values = list(stratify = TRUE, range = 100)
+  expect_error(buffer$instantiate(task_grp),
     "Grouping is not supported for spatial resampling methods")
 })
 
@@ -64,18 +67,18 @@ test_that("train and test set getter functions are working", {
     as.data.table(mlr_resamplings)[map_lgl(key, grepl, pattern = "spcv"), key]
   )
 
-  spcv_rsp_no_buffer = spcv_rsp
-  spcv_rsp_no_buffer$`spcv-buffer` = NULL
+  spcv_rsp$`spcv-buffer` = NULL
+  spcv_rsp$`repeated-spcv-coords` = NULL
 
-  for (i in spcv_rsp_no_buffer) {
+  for (i in spcv_rsp) {
     i$instantiate(task)
     expect_silent(i$train_set(1))
     expect_silent(i$test_set(1))
[…]

### 2020-01-19 R/zzz.R
Add ResamplingRepeatedSpCVEnv (#32)

@@ -43,7 +43,9 @@ register_mlr3 = function() {
     mlr_resamplings$add("spcv-buffer", ResamplingSpCVBuffer)
     mlr_resamplings$add("spcv-coords", ResamplingSpCVCoords)
     mlr_resamplings$add("spcv-env", ResamplingSpCVEnv)
+
     mlr_resamplings$add("repeated-spcv-coords", ResamplingRepeatedSpCVCoords)
+    mlr_resamplings$add("repeated-spcv-env", ResamplingRepeatedSpCVEnv)
   }
 
 }

### 2020-01-19 tests/testthat/test_mlr_spcv_generic.R

@@ -18,6 +18,8 @@ test_that("stratification throws errors", {
   )
   spcv_rsp$`spcv-buffer` = NULL
   spcv_rsp$`repeated-spcv-coords` = NULL
+  spcv_rsp$`repeated-spcv-env` = NULL
+
   for (i in spcv_rsp) {
     i$param_set$values = list(folds = 5, stratify = TRUE)
     expect_error(i$instantiate(task))
@@ -28,9 +30,9 @@ test_that("grouping throws errors when 'groups' is set", {
   spcv_rsp = mlr_resamplings$mget(
     as.data.table(mlr_resamplings)[map_lgl(key, grepl, pattern = "spcv"), key]
   )
-  spcv_rsp_no_buffer = spcv_rsp
-  spcv_rsp_no_buffer$`spcv-buffer` = NULL
   spcv_rsp$`repeated-spcv-coords` = NULL
+  spcv_rsp$`repeated-spcv-env` = NULL
+
   for (i in spcv_rsp) {
     expect_error(i$instantiate(task_grp),
       "Grouping is not supported for spatial resampling methods")
@@ -43,6 +45,7 @@ test_that("grouping throws errors when 'groups' and 'stratify' is set", {
   )
   spcv_rsp$`spcv-buffer` = NULL
   spcv_rsp$`repeated-spcv-coords` = NULL
+  spcv_rsp$`repeated-spcv-env` = NULL
 
   for (i in spcv_rsp) {
     i$param_set$values = list(folds = 5, stratify = TRUE)

### 2020-01-19 README.Rmd
use `tsk` instead of `mlr_tasks$get()`
Add ResamplingRepeatedSpCVEnv (#32)

@@
 | Repeated Spatial CV    | [sperrorest](https://github.com/giscience-fsu/sperrorest) | Brenning 2012 | `RepeatedResamplingSpCVCoords`  | `rsmp("repeated-spcv-coords")` |
+| Repeated Env Blocking  | [blockCV](https://github.com/rvalavi/blockCV)             | Valavi 2019   | `RepeatedResamplingSpCVEnv`     | `rsmp("repeated-spcv-env")`    |
 
@@
 | ---      | ---               | ---     |
-| ecuador  | `mlr_tasks$get("ecuador")`  | Classif |
-| diplodia | `mlr_tasks$get("diplodia")` | Classif |
+| ecuador  | `tsk("ecuador")`  | Classif |
+| diplodia | `tsk("diplodia")` | Classif |
 

### 2020-01-19 R/autoplot.R

@@ -199,6 +199,39 @@ autoplot.ResamplingRepeatedSpCVCoords = function(
     grid = grid)
 }
 
+#' @title Plot for Repeated Spatial Resampling
+#'
+#' @rdname autoplot_spatial_resampling
+#' @export
+#' @examples
+#' #####
+#' # RepeatedSpCVEnv
+#' #####
+#' \donttest{
+#' task = tsk("ecuador")
+#' resampling = rsmp("repeated-spcv-env", folds = 10, repeats = 2)
+#' resampling$instantiate(task)
+#' autoplot(resampling, task)
+#' autoplot(resampling, task, 1)
+#' autoplot(resampling, task, fold_id = 2, repeats_id = 2)
+#' autoplot(resampling, task, c(1, 2, 3, 4))
+#' }
+autoplot.ResamplingRepeatedSpCVEnv = function(
+  object,
+  task,
+  fold_id = NULL,
+  repeats_id = 1,
+  grid = TRUE,
+  train_color = "#0072B5",
+  test_color = "#E18727",
+  ...) {
+  autoplot_spatial(resampling = object,
+    task = task,
+    fold_id = fold_id,
+    repeats_id = repeats_id,
+    grid = grid)
+}
+
 autoplot_spatial = function(
   resampling = NULL,
   task = NULL,

### 2020-01-20 README.Rmd
add repeated blockCV to README

@@
 bibliography: inst/references.bib
+biblio-style: apalike
 nocite: | 
@@
 | Literature                | Package     | Reference     | Class Name      | mlr3 Sugar        |
-| ---------------        | ---------                                                 | ---------     | --------------------            | ---------------------          |
+| ---------------           | ---------   | ---------     | --------------- | ------------      |
 | Spatial Buffering         | [blockCV](https://github.com/rvalavi/blockCV)             | Valavi 2019   | `ResamplingSpCVBuffer`          | `rsmp("spcv-buffer")`          |
@@
 | ---                       | ---                                                       | ---           | ---                             | ---                            |
+| Repeated Spatial Blocking | [blockCV](https://github.com/rvalavi/blockCV)             | Valavi 2019   | `RepeatedResamplingSpCVBlock`   | `rsmp("repeated-spcv-block")`  |
 | Repeated Spatial CV       | [sperrorest](https://github.com/giscience-fsu/sperrorest) | Brenning 2012 | `RepeatedResamplingSpCVCoords`  | `rsmp("repeated-spcv-coords")` |

### 2020-01-20 README.Rmd
try to shrink table

@@
 | ---------------           | ---------   | ---------     | --------------- | ------------      |
-| Spatial Buffering         | [blockCV](https://github.com/rvalavi/blockCV)             | Valavi 2019   | `ResamplingSpCVBuffer`          | `rsmp("spcv-buffer")`          |
-| Spatial Blocking          | [blockCV](https://github.com/rvalavi/blockCV)             | Valavi 2019   | `ResamplingSpCVBlock`           | `rsmp("spcv-block")`           |
-| Spatial CV                | [sperrorest](https://github.com/giscience-fsu/sperrorest) | Brenning 2012 | `ResamplingSpCVCoords`          | `rsmp("spcv-coords")`          |
-| Environmental Blocking    | [blockCV](https://github.com/rvalavi/blockCV)             | Valavi 2019   | `ResamplingSpCVEnv`             | `rsmp("spcv-env")`             |
+| Spatial Buffering         | blockCV     | Valavi 2019   | `ResamplingSpCVBuffer`          | `rsmp("spcv-buffer")`          |
+| Spatial Blocking          | blockCV     | Valavi 2019   | `ResamplingSpCVBlock`           | `rsmp("spcv-block")`           |
+| Spatial CV                | sperrorest  | Brenning 2012 | `ResamplingSpCVCoords`          | `rsmp("spcv-coords")`          |
+| Environmental Blocking    | blockCV     | Valavi 2019   | `ResamplingSpCVEnv`             | `rsmp("spcv-env")`             |
 | ---                       | ---         | ---           | ---                             | ---                            |
-| Repeated Spatial Blocking | [blockCV](https://github.com/rvalavi/blockCV)             | Valavi 2019   | `RepeatedResamplingSpCVBlock`   | `rsmp("repeated-spcv-block")`  |
-| Repeated Spatial CV       | [sperrorest](https://github.com/giscience-fsu/sperrorest) | Brenning 2012 | `RepeatedResamplingSpCVCoords`  | `rsmp("repeated-spcv-coords")` |
-| Repeated Env Blocking     | [blockCV](https://github.com/rvalavi/blockCV)             | Valavi 2019   | `RepeatedResamplingSpCVEnv`     | `rsmp("repeated-spcv-env")`    |
+| Repeated Spatial Blocking | blockCV     | Valavi 2019   | `RepeatedResamplingSpCVBlock`   | `rsmp("repeated-spcv-block")`  |
+| Repeated Spatial CV       | sperrorest  | Brenning 2012 | `RepeatedResamplingSpCVCoords`  | `rsmp("repeated-spcv-coords")` |
+| Repeated Env Blocking     | blockCV     | Valavi 2019   | `RepeatedResamplingSpCVEnv`     | `rsmp("repeated-spcv-env")`    |
 

### 2020-01-20 README.Rmd
condense readme table

@@
 
-| Literature                | Package     | Reference     | Class Name      | mlr3 Sugar        |
-| ---------------           | ---------   | ---------     | --------------- | ------------      |
-| Spatial Buffering         | blockCV     | Valavi 2019   | `ResamplingSpCVBuffer`          | `rsmp("spcv-buffer")`          |
-| Spatial Blocking          | blockCV     | Valavi 2019   | `ResamplingSpCVBlock`           | `rsmp("spcv-block")`           |
-| Spatial CV                | sperrorest  | Brenning 2012 | `ResamplingSpCVCoords`          | `rsmp("spcv-coords")`          |
-| Environmental Blocking    | blockCV     | Valavi 2019   | `ResamplingSpCVEnv`             | `rsmp("spcv-env")`             |
-| ---                       | ---         | ---           | ---                             | ---                            |
-| Repeated Spatial Blocking | blockCV     | Valavi 2019   | `RepeatedResamplingSpCVBlock`   | `rsmp("repeated-spcv-block")`  |
-| Repeated Spatial CV       | sperrorest  | Brenning 2012 | `RepeatedResamplingSpCVCoords`  | `rsmp("repeated-spcv-coords")` |
-| Repeated Env Blocking     | blockCV     | Valavi 2019   | `RepeatedResamplingSpCVEnv`     | `rsmp("repeated-spcv-env")`    |
+| Literature                | Package     | Reference     | mlr3 Sugar                     |
+| ---------------           | ---------   | ---------     | ---------------                |
+| Spatial Buffering         | blockCV     | Valavi 2019   | `rsmp("spcv-buffer")`          |
+| Spatial Blocking          | blockCV     | Valavi 2019   | `rsmp("spcv-block")`           |
+| Spatial CV                | sperrorest  | Brenning 2012 | `rsmp("spcv-coords")`          |
+| Environmental Blocking    | blockCV     | Valavi 2019   | `rsmp("spcv-env")`             |
+| ---                       | ---         | ---           | ---                            |
+| Repeated Spatial Blocking | blockCV     | Valavi 2019   | `rsmp("repeated-spcv-block")`  |
+| Repeated Spatial CV       | sperrorest  | Brenning 2012 | `rsmp("repeated-spcv-coords")` |
+| Repeated Env Blocking     | blockCV     | Valavi 2019   | `rsmp("repeated-spcv-env")`    |
 

### 2020-01-21 README.Rmd
build macOS on Travis, deploy pkgdown on Circle CI

@@
 <!-- badges: start -->
-[![Build Status](https://img.shields.io/travis/mlr-org/mlr3spatiotempcv/master?label=Linux&logo=travis&style=flat-square)](https://travis-ci.org/mlr-org/mlr3spatiotempcv)
+[![Build Status](https://img.shields.io/travis/mlr-org/mlr3spatiotempcv/master?label=macOS&logo=travis&style=flat-square)](https://travis-ci.org/mlr-org/mlr3spatiotempcv)
 [![CircleCI build status](https://circleci.com/gh/mlr-org/mlr3spatiotempcv.svg?style=svg)](https://circleci.com/gh/mlr-org/mlr3spatiotempcv)

### 2020-01-21 tests/testthat/test-autoplot.R
`autoplot()`: support CV and RepeatedCV

@@ -180,3 +180,52 @@ test_that("autoplot works for RepeatedSpCVEnv", {
   vdiffr::expect_doppelganger("RepSpCVEnv - Fold 1, Rep 1", renv4)
   vdiffr::expect_doppelganger("RepSpCVEnv - Fold 1-4", renv5)
 })
+
+
+test_that("autoplot works for SpCVEnv", {
+
+  # CV -------------------------------------------------------------------------
+
+  set.seed(42)
+
+  task = tsk("ecuador")
+  resa_cv = rsmp("cv")
+  resa_cv$instantiate(task)
+  cv1 = autoplot(resa_cv, task)
+  cv2 = autoplot(resa_cv, task, 1)
+  cv3 = autoplot(resa_cv, task, c(1, 2, 3, 4))
+
+  skip_on_ci()
+  vdiffr::expect_doppelganger("CV all test sets", cv1)
+  vdiffr::expect_doppelganger("CV - Fold 1", cv2)
+  vdiffr::expect_doppelganger("CV - Fold 1-4", cv3)
+})
+
+test_that("autoplot works for RepeatedSpCVEnv", {
+
+  # RepeatedCV -----------------------------------------------------------------
+
+  set.seed(42)
+
+  task = tsk("ecuador")
+  rresa_cv = rsmp("repeated_cv", folds = 10, repeats = 2)
+  rresa_cv$instantiate(task)
+  rcv1 = autoplot(rresa_cv, task)
+  rcv2 = autoplot(rresa_cv, task, repeats_id = 2)
+  rcv3 = autoplot(rresa_cv, task, fold_id = 1)
+  rcv4 = autoplot(rresa_cv, task, fold_id = 1, repeats_id = 2)
+  rcv5 = autoplot(rresa_cv, task, fold_id = c(1, 2, 3, 4))
+
+  expect_true(is.ggplot(rcv1))
+  expect_true(is.ggplot(rcv2))
+  expect_true(is.ggplot(rcv3))
+  expect_true(is.ggplot(rcv4))
+  expect_true(is.ggplot(rcv5))
+
+  skip_on_ci()
+  vdiffr::expect_doppelganger("RepCV all test sets", rcv1)
+  vdiffr::expect_doppelganger("RepCV all test sets, Rep 2", rcv2)
+  vdiffr::expect_doppelganger("RepCV - Fold 1, Rep 2", rcv3)
[…]

### 2020-03-15 README.Rmd
Fix CI for Github Actions (#44)

@@
 <!-- badges: start -->
-[![Build Status](https://img.shields.io/travis/mlr-org/mlr3spatiotempcv/master?label=macOS&logo=travis&style=flat-square)](https://travis-ci.org/mlr-org/mlr3spatiotempcv)
-[![CircleCI build status](https://circleci.com/gh/mlr-org/mlr3spatiotempcv.svg?style=svg)](https://circleci.com/gh/mlr-org/mlr3spatiotempcv)
+[![R CMD Check via {tic}](https://img.shields.io/github/workflow/status/mlr-org/mlr3spatiotempcv/R%20CMD%20Check%20via%20%7Btic%7D?logo=github&label=R%20CMD%20Check%20via%20{tic}&style=flat-square)](https://github.com/mlr-org/mlr3spatiotempcv/actions)
 [![CRAN Status Badge](https://www.r-pkg.org/badges/version-ago/mlr3spatiotempcv)](https://cran.r-project.org/package=mlr3spatiotempcv)

### 2020-03-25 README.Rmd
update GHA badge

@@
 <!-- badges: start -->
-[![R CMD Check via {tic}](https://img.shields.io/github/workflow/status/mlr-org/mlr3spatiotempcv/R%20CMD%20Check%20via%20%7Btic%7D?logo=github&label=R%20CMD%20Check%20via%20{tic}&style=flat-square)](https://github.com/mlr-org/mlr3spatiotempcv/actions)
+[![R CMD Check via {tic}](https://github.com/mlr-org/mlr3spatiotempcv/workflows/R%20CMD%20Check%20via%20{tic}/badge.svg?branch=master)](https://github.com/mlr-org/mlr3spatiotempcv/actions)
 [![CRAN Status Badge](https://www.r-pkg.org/badges/version-ago/mlr3spatiotempcv)](https://cran.r-project.org/package=mlr3spatiotempcv)
@@
 
-task = tsk("ecuador")
-resampling = rsmp("spcv-coords", folds = 5)
+task <- tsk("ecuador")
+resampling <- rsmp("spcv-coords", folds = 5)
 resampling$instantiate(task)

### 2020-05-26 README.Rmd
update README

@@
 [![CRAN Status Badge](https://www.r-pkg.org/badges/version-ago/mlr3spatiotempcv)](https://cran.r-project.org/package=mlr3spatiotempcv)
-[![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
 [![Coverage status](https://codecov.io/gh/mlr-org/mlr3spatiotempcv/branch/master/graph/badge.svg)](https://codecov.io/github/mlr-org/mlr3spatiotempcv?branch=master)
+[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
 <!-- badges: end -->
 
-This package extends the [mlr3](https://github.com/mlr-org/mlr3) package framework by spatiotemporal resampling and visualization methods.
+This package extends the [mlr3](https://github.com/mlr-org/mlr3) package framework with spatiotemporal resampling and visualization methods.
+
+⚠️⚠️⚠️  
+The package is in very early stages and breaking changes withour further notice are expected.
+If you want to use if for your research, you might need to refactor your analysis along the way.  
+⚠️⚠️⚠️
 
@@
 | diplodia | `tsk("diplodia")` | Classif |
+| cookfarm | `tsk("cookfarm")` | Regr    |
 
@@
 
-### Visualization of all partitions
+### Visualization of all folds
 
@@
 
-### Visualization of the first fold only
+### Visualization of a specific fold
 

### 2020-06-14 README.Rmd
GHA: update tic templates

@@
 <!-- badges: start -->
-[![R CMD Check via {tic}](https://github.com/mlr-org/mlr3spatiotempcv/workflows/R%20CMD%20Check%20via%20{tic}/badge.svg?branch=master)](https://github.com/mlr-org/mlr3spatiotempcv/actions)
+[![tic](https://github.com/mlr-org/mlr3spatiotempcv/workflows/tic/badge.svg?branch=master)](https://github.com/mlr-org/mlr3spatiotempcv/actions)
 [![CRAN Status Badge](https://www.r-pkg.org/badges/version-ago/mlr3spatiotempcv)](https://cran.r-project.org/package=mlr3spatiotempcv)

### 2020-07-30 R/ResamplingSptCVCluto.R
- Add `ResamplingSptCVCluto` (#53)

@@ -1,0 +1,188 @@
+#' @title Spatio-Temporal Cluster Resampling
+#'
+#' @import mlr3
+#'
+#' @description Spatio-temporal cluster partitioning via the `vcluster`
+#'   executable of the
+#'   [CLUTO](http://glaros.dtc.umn.edu/gkhome/cluto/cluto/overview) clustering
+#'   application.
+#'
+#' This partitioning method relies on the external CLUTO library.
+#' To use it, CLUTO's executables need to be downloaded and installed into
+#' this package.
+#'
+#' See \url{https://gist.github.com/pat-s/6430470cf817050e27d26c43c0e9be72} for an
+#' installation approach that should work on Windows and Linux.
+#' macOS is not supported by CLUTO.
+#'
+#' Before using this method, please check the restrictive
+#' [copyright](http://glaros.dtc.umn.edu/gkhome/cluto/cluto/download) shown
+#' below.
+#'
+#' @details
+#' By default, `-clmethod='direct'` is passed to the `vcluster` executable in
+#' contrast to the upstream default `-clmethod='rb'`.
+#' There is no evidence or research that this method is the best among the
+#' available ones ("rb", "rbr", "direct", "agglo", "graph", "bagglo"). Also,
+#' various other parameters can be set via argument `cluto_parameters` to
+#' achieve different clustering results.
+#'
+#' Parameter `-clusterfile` is handled by \CRANpkg{skmeans} and cannot be
+#' changed.
+#'
+#' @section Copyright:
+#'
+#' CLUTO's copyright is as follows:
+#'
+#' The CLUTO package is copyrighted by the Regents of the University of
+#' Minnesota.
+#' It can be freely used for educational and research purposes by non-profit
+#' institutions and US government agencies only.
+#' Other organizations are allowed to use CLUTO only for evaluation purposes,
+#' and any further uses will require prior approval.
+#' The software may not be sold or redistributed without prior approval.
+#' One may make copies of the software for their use provided that the copies,
+#' are not sold or distributed, are used under the same terms and conditions.
+#' As unestablished research software, this code is provided on an “as is” basis
+#' without warranty of any kind, either expressed or implied.
+#' The downloading, or executing any part of this software constitutes an
+#' implicit agreement to these terms. These terms and conditions are subject to
[…]

### 2020-07-30 R/Task_classif_ecuador.R

@@ -1,5 +1,7 @@
 #' @title Ecuador Classification Task
 #'
+#' @docType data
+#' @usage data(ecuador)
 #' @name mlr_tasks_ecuador
 #' @format [R6::R6Class] inheriting from [TaskClassif].
 #' @importFrom mlr3 as_data_backend
@@ -22,9 +24,10 @@
 #'   rates of landslides along a humidity gradient in the tropical Andes.
 #'   Geomorphology, 139-140: 271-284.
 #' @template seealso_task
+"ecuador"
+
 load_task_ecuador = function(id = "ecuador") {
-  b = mlr3::as_data_backend(readRDS(system.file("extdata", "ecuador.rda",
-    package = "mlr3spatiotempcv")))
+  b = mlr3::as_data_backend(ecuador)
   b$hash = "_mlr3_tasks_ecuador_"
   task = TaskClassifST$new(
     id = "ecuador", b, target = "slides", positive = "TRUE",

### 2020-07-30 R/Task_regr_cookfarm_profiles.R

@@ -1,5 +1,7 @@
 #' @title Cookfarm Profiles Regression Task
 #'
+#' @docType data
+#' @usage data(cookfarm)
 #' @name mlr_tasks_cookfarm
 #' @format [R6::R6Class] inheriting from [TaskRegr].
 #' @importFrom mlr3 as_data_backend
@@ -37,11 +39,10 @@
 #' modeling the spatial and temporal variation of soil moisture in a dryland
 #' agricultural field. Submitted to Water Resources Research.
 #' @template seealso_task
-NULL
+"cookfarm"
 
 load_task_cookfarm = function(id = "cookfarm") {
-  b = mlr3::as_data_backend(readRDS(system.file("extdata", "cookfarm.rda",
-    package = "mlr3spatiotempcv")))
+  b = mlr3::as_data_backend(cookfarm)
   b$hash = "_mlr3_tasks_cookfarm_"
   task = TaskRegrST$new(
     id = "cookfarm", b, target = "PHIHOX",

### 2020-07-30 R/helper.R

@@ -4,3 +4,40 @@ translate_types = function(x) {
   p_types = names(mlr_reflections$task_feature_types)
   factor(map_values(x, r_types, p_types), levels = p_types)
 }
+
+check_cluto_path = function() {
+  if (!Sys.getenv("CLUTO_PATH") == "") {
+    vcluster_loc = normalizePath(Sys.getenv("CLUTO_PATH"))
+  } else {
+    cli::cli_alert_info("CLUTO executable path not set via env var
+          {.var CLUTO_PATH}.
+          Trying to find it in {.pkg mlr3spatiotempcv}.", wrap = TRUE)
+
+    vcluster_loc = switch(Sys.info()[["sysname"]],
+      "Windows" = {
+        if (!file.exists(system.file("vcluster.exe",
+          package = "mlr3spatiotempcv"))) {
+          cli::cli_alert_danger("{.file vcluster.exe} not found. Please
+              install CLUTO first. See {.code ?ResamplingSptCVCluto} for
+              instructions.", wrap = TRUE)
+          stop("CLUTO executable not found.")
+        }
+        normalizePath(system.file("vcluster.exe",
+          package = "mlr3spatiotempcv"))
+      },
+      "Linux" = {
+        if (!file.exists(system.file("vcluster",
+          package = "mlr3spatiotempcv"))) {
+          cli::cli_alert_danger("{.file vcluster} not found. Please install
+              CLUTO first. See {.code ?ResamplingSptCVCluto} for
+              instructions.", wrap = TRUE)
+          stop("CLUTO executable not found.")
+        }
+        system.file("vcluster",
+          package = "mlr3spatiotempcv")
+      },
+      "Darwin" = stop("macOS is not supported by CLUTO.")
+    )
+  }
+  return(vcluster_loc)
+}

### 2020-07-30 inst/extdata/cookfarm.R

@@ -1,31 +1,0 @@
-library(dplyr)
-library(sf)
-library(GSIF)
-data(cookfarm)
-set.seed(42)
-
-cookfarm_profiles = cookfarm$profiles
-
-cookfarm_mlr3 = cookfarm_profiles %>%
-  left_join(cookfarm$readings %>%
-    select("Date", "SOURCEID"), by = "SOURCEID") %>%
-  st_as_sf(coords = c("Easting", "Northing"), crs = 26911) %>%
-  mutate(x = st_coordinates(.)[, "X"]) %>%
-  mutate(y = st_coordinates(.)[, "Y"]) %>%
-  na.omit() %>%
-  sample_n(500)
-
-saveRDS(cookfarm_mlr3, "inst/extdata/cookfarm.R")
-
-# mapview::mapview(cookfarm_mlr3)
-
-### unique timestamps
-# length(unique(cookfarm_mlr3$Date))
-# = 365 (out of 500)
-
-### unique coordinates
-# st_coordinates(cookfarm_mlr3) %>%
-#   as.data.frame() %>%
-#   group_by(X, Y) %>%
-#   unique()
-# = 41

### 2020-07-30 tests/testthat/test_mlr_spcv_generic.R

@@ -43,6 +43,8 @@ test_that("train and test set getter functions are working", {
   spcv_rsp$`repeated-spcv-coords` = NULL
   spcv_rsp$`repeated-spcv-env` = NULL
   spcv_rsp$`spcv-block` = NULL
+  spcv_rsp$`spcv-cluto` = NULL
+  spcv_rsp$`repeated-spcv-cluto` = NULL
   # nolint end
 
   for (i in spcv_rsp) {
@@ -66,6 +68,8 @@ test_that("cloning works", {
   spcv_rsp$`repeated-spcv-coords` = NULL
   spcv_rsp$`repeated-spcv-env` = NULL
   spcv_rsp$`repeated-spcv-block` = NULL
+  spcv_rsp$`spcv-cluto` = NULL
+  spcv_rsp$`repeated-spcv-cluto` = NULL
   # nolint end
 
   for (i in spcv_rsp) {

### 2020-07-30 README.Rmd
- Add `ResamplingSptCVCluto` (#53)

@@
 | Environmental Blocking    | blockCV     | Valavi 2019   | `rsmp("spcv-env")`             |
+|           -               | -           | -             | `rsmp("spcv-cluto")`           |
 | ---                       | ---         | ---           | ---                            |
@@
 | Repeated Env Blocking     | blockCV     | Valavi 2019   | `rsmp("repeated-spcv-env")`    |
+|           -               | -           | -             | `rsmp("repeated-spcv-cluto")`  |
 
@@
 | diplodia | `tsk("diplodia")` | Classif |
+
+## Spatiotemporal tasks
+
+| Name     | Code              | Type    |
+| ---      | ---               | ---     |
 | cookfarm | `tsk("cookfarm")` | Regr    |
@@
 
-task <- tsk("ecuador")
-resampling <- rsmp("spcv-coords", folds = 5)
+task = tsk("ecuador")
+resampling = rsmp("spcv-coords", folds = 5)
 resampling$instantiate(task)
@@
 
+### Spatiotemporal Visualization
+
+Three-dimensional visualization via {plotly}
+
+```{r spatiotemp, eval=FALSE}
+library(plotly)
+task_st = tsk("cookfarm")
+resampling = rsmp("spcv-cluto", folds = 5)
+resampling$instantiate(task_st, time_var = "Date")
+autoplot(resampling, task_st, fold_id = 1, point_size = 3)
+```
+
+![](man/figures/spt-viz.png)
+
 ## More resources

### 2020-07-31 R/ResamplingSptCVCstf.R
add ResamplingSptCVCstf

@@ -1,0 +1,173 @@
+#' @title Create spatio-temporal Folds Using Predefined Groups
+#'
+#' @template rox_sptcv_cstf
+#' @export
+#' @examples
+#' library(mlr3)
+#' task = tsk("ecuador")
+#'
+#' # Instantiate Resampling
+#' rcv = rsmp("spcv-cstf", folds = 5)
+#' rcv$instantiate(task, time_var = "Date", space_var = "SOURCEID")
+#'
+#' # Individual sets:
+#' rcv$train_set(1)
+#' rcv$test_set(1)
+#' # check that no obs are in both sets
+#' intersect(rcv$train_set(1), rcv$test_set(1)) # good!
+#'
+#' # Internal storage:
+#' rcv$instance # table
+ResamplingSptCVCstf = R6Class("ResamplingSptCVCstf",
+  inherit = mlr3::Resampling,
+
+  public = list(
+    #' @description
+    #' Create a "Spacetime Folds" resampling instance.
+    #' @param id `character(1)`\cr
+    #'   Identifier for the resampling strategy.
+    initialize = function(id = "spcv-cstf") {
+      ps = ParamSet$new(params = list(
+        ParamInt$new("folds", lower = 1L, default = 10L, tags = "required")
+      ))
+      ps$values = list(folds = 10L)
+      super$initialize(
+        id = id,
+        param_set = ps
+      )
+    },
+
+    #' @description
+    #'  Materializes fixed training and test splits for a given task.
+    #' @param task [Task]\cr
+    #'   A task to instantiate.
+    #' @param space_var `[character]`\cr
+    #'   Column name identifying the spatial units.
+    #' @param time_var `[character]`\cr
+    #'   Column name identifying the temporal units.
+    #' @param class `[character]`\cr
+    #'   Column name identifying a class unit (e.g. land cover).
+    instantiate = function(task, space_var = NULL, time_var = NULL,
+      class = NULL) {
+
+      assert_task(task)
+      groups = task$groups
+
+      if (!is.null(groups)) {
+        stopf("Grouping is not supported for spatial resampling methods")
+      }
+
+      private$.sample(task, space_var, time_var, class)
+
+      self$task_hash = task$hash
+      invisible(self)
+    }
+  ),
+
+  active = list(
+    #' @field iters `integer(1)`\cr
+    #'   Returns the number of resampling iterations, depending on the
+    #'   values stored in the `param_set`.
+    iters = function() {
+      self$param_set$values$folds
+    }
+  ),
+
+  private = list(
[…]

### 2020-07-31 man-roxygen/rox_sptcv_cstf.R

@@ -1,0 +1,6 @@
+#' @description Implementation of `CAST::CreateSpaceTimeFolds()`.
+#'
+#' @inherit CAST::CreateSpacetimeFolds details
+#'
+#' @references
+#' \cite{mlr3spatiotempcv}{meyer2018}

### 2020-07-31 R/zzz.R
add ResamplingSptCVCstf

@@ -44,6 +44,7 @@ register_mlr3 = function() {
     x = utils::getFromNamespace("mlr_resamplings", ns = "mlr3")
     mlr_resamplings$add("spcv-block", ResamplingSpCVBlock)
     mlr_resamplings$add("spcv-buffer", ResamplingSpCVBuffer)
+    mlr_resamplings$add("spcv-cstf", ResamplingSptCVCstf)
     mlr_resamplings$add("spcv-cluto", ResamplingSptCVCluto)
     mlr_resamplings$add("spcv-coords", ResamplingSpCVCoords)
     mlr_resamplings$add("spcv-env", ResamplingSpCVEnv)

### 2020-07-31 tests/testthat/test_ResamplingSptCVCstf.R

@@ -1,0 +1,71 @@
+context("ResamplingSptCVCstf")
+
+test_that("mlr3spatiotempcv implementation is equal to CAST: time + space", {
+  library(mlr3)
+  library(mlr3spatiotempcv)
+  set.seed(42)
+
+  task = tsk("cookfarm")
+  rsp = rsmp("spcv-cstf", folds = 10)
+  rsp$instantiate(task, time_var = "Date", space_var = "SOURCEID")
+  rsp$test_set(1)
+
+  set.seed(42)
+  indices = CAST::CreateSpacetimeFolds(task$data(), "SOURCEID", "Date")
+  indices$indexOut[[1]]
+
+  expect_equal(rsp$instance$test, indices$indexOut)
+})
+
+test_that("mlr3spatiotempcv implementation is equal to CAST: time ", {
+  library(mlr3)
+  library(mlr3spatiotempcv)
+  set.seed(42)
+
+  task = tsk("cookfarm")
+  rsp = rsmp("spcv-cstf", folds = 10)
+  rsp$instantiate(task, time_var = "Date")
+  rsp$test_set(1)
+
+  set.seed(42)
+  indices = CAST::CreateSpacetimeFolds(task$data(), timevar = "Date")
+  indices$indexOut[[1]]
+
+  expect_equal(rsp$instance$test, indices$indexOut)
+})
+
+test_that("mlr3spatiotempcv implementation is equal to CAST: space", {
+  library(mlr3)
+  library(mlr3spatiotempcv)
+  set.seed(42)
+
+  task = tsk("cookfarm")
+  rsp = rsmp("spcv-cstf", folds = 10)
+  rsp$instantiate(task, space_var = "SOURCEID")
+  rsp$test_set(1)
+
+  set.seed(42)
+  indices = CAST::CreateSpacetimeFolds(task$data(), spacevar = "SOURCEID")
+  indices$indexOut[[1]]
+
+  expect_equal(rsp$instance$test, indices$indexOut)
+})
+
+test_that("mlr3spatiotempcv implementation is equal to CAST: space", {
+  library(mlr3)
+  library(mlr3spatiotempcv)
+  set.seed(42)
+
+  task = tsk("cookfarm")
+  rsp = rsmp("spcv-cstf", folds = 10)
+  rsp$instantiate(task, space_var = "SOURCEID", class = "TAXSUSDA")
+  rsp$test_set(1)
+
+  set.seed(42)
[…]

### 2020-07-31 README.Rmd
upd README with link to spatiotemp viz vignette
add ResamplingSptCVCstf

@@
 | Literature                  | Package    | Reference     | mlr3 Sugar                     |
-| ---------------           | ---------   | ---------     | ---------------                |
+| --------------------------- | ---------- | ------------- | ------------------------------ |
 | Spatial Buffering           | blockCV    | Valavi 2019   | `rsmp("spcv-buffer")`          |
@@
 | Environmental Blocking      | blockCV    | Valavi 2019   | `rsmp("spcv-env")`             |
-|           -               | -           | -             | `rsmp("spcv-cluto")`           |
+| -                           | -          | -             | `rsmp("sptcv-cluto")`          |
+| Leave-Location-and-Time-Out | CAST       | Meyer 2018    | `rsmp("sptcv-cstf")`           |
 | ---                         | ---        | ---           | ---                            |
@@
 | Repeated Env Blocking       | blockCV    | Valavi 2019   | `rsmp("repeated-spcv-env")`    |
-|           -               | -           | -             | `rsmp("repeated-spcv-cluto")`  |
+| -                           | -          | -             | `rsmp("repeated-sptcv-cluto")` |
 
@@
 
+(See vignette ["Spatiotemporal Visualization"](https://mlr3spatiotempcv.mlr-org.com/articles/spatiotemp-viz.html) for an interactive 3D HTML variant.)
+
 ```{r spatiotemp, eval=FALSE}

### 2020-07-31 R/TaskRegrST.R

@@ -14,7 +14,7 @@
 #'
 #' @family Task
 #' @export
-TaskRegrST <- R6::R6Class("TaskRegrST",
+TaskRegrST = R6::R6Class("TaskRegrST",
   inherit = TaskRegr,
   public = list(
 

### 2020-08-15 README.Rmd
update README
add RepeatedSptCVCstf

@@
 <!-- badges: start -->
-[![tic](https://github.com/mlr-org/mlr3spatiotempcv/workflows/tic/badge.svg?branch=master)](https://github.com/mlr-org/mlr3spatiotempcv/actions)
-[![CRAN Status Badge](https://www.r-pkg.org/badges/version-ago/mlr3spatiotempcv)](https://cran.r-project.org/package=mlr3spatiotempcv)
-[![Coverage status](https://codecov.io/gh/mlr-org/mlr3spatiotempcv/branch/master/graph/badge.svg)](https://codecov.io/github/mlr-org/mlr3spatiotempcv?branch=master)
-[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
-<!-- badges: end -->
+
+[![tic](https://github.com/mlr-org/mlr3spatiotempcv/workflows/tic/badge.svg?branch=master)](https://github.com/mlr-org/mlr3spatiotempcv/actions) [![CRAN Status Badge](https://www.r-pkg.org/badges/version-ago/mlr3spatiotempcv)](https://cran.r-project.org/package=mlr3spatiotempcv) [![Coverage status](https://codecov.io/gh/mlr-org/mlr3spatiotempcv/branch/master/graph/badge.svg)](https://codecov.io/github/mlr-org/mlr3spatiotempcv?branch=master) [![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental) <!-- badges: end -->
 
@@
 
-⚠️⚠️⚠️  
-The package is in very early stages and breaking changes withour further notice are expected.
-If you want to use if for your research, you might need to refactor your analysis along the way.  
+⚠️⚠️⚠️\
+The package is in very early stages and breaking changes withour further notice are expected. If you want to use if for your research, you might need to refactor your analysis along the way.\
 ⚠️⚠️⚠️
@@
 | Literature                           | Package    | Reference     | mlr3 Sugar                     |
-| --------------------------- | ---------- | ------------- | ------------------------------ |
+|--------------------------------------|------------|---------------|--------------------------------|
 | Spatial Buffering                    | blockCV    | Valavi 2019   | `rsmp("spcv-buffer")`          |
@@
 | Environmental Blocking               | blockCV    | Valavi 2019   | `rsmp("spcv-env")`             |
-| -                           | -          | -             | `rsmp("sptcv-cluto")`          |
+| \-                                   | \-         | \-            | `rsmp("sptcv-cluto")`          |
 | Leave-Location-and-Time-Out          | CAST       | Meyer 2018    | `rsmp("sptcv-cstf")`           |
@@
 | Repeated Env Blocking                | blockCV    | Valavi 2019   | `rsmp("repeated-spcv-env")`    |
-| -                           | -          | -             | `rsmp("repeated-sptcv-cluto")` |
+| \-                                   | \-         | \-            | `rsmp("repeated-sptcv-cluto")` |
+| Repeated Leave-Location-and-Time-Out | CAST       | Meyer 2018    | `rsmp("repeated-sptcv-cstf")`  |
 
@@
 | Name     | Code              | Type    |
-| ---      | ---               | ---     |
+|----------|-------------------|---------|
 | ecuador  | `tsk("ecuador")`  | Classif |
@@
 | Name     | Code              | Type |
-| ---      | ---               | ---     |
+|----------|-------------------|------|
 | cookfarm | `tsk("cookfarm")` | Regr |
@@
 
-Generic S3 function `autoplot()` for all implemented spatial resampling methods.
+S3 `autoplot()` for all implemented spatial resampling methods.
 

### 2020-08-15 R/ResamplingRepeatedSptCVCstf.R

@@ -1,0 +1,160 @@
+#' @title Repeated Spatio-Temporal Cluster Resampling
+#'
+#' @template rox_sptcv_cluto
+#'
+#' @export
+#' @examples
+#' \dontrun{
+#' library(mlr3)
+#' library(mlr3spatiotempcv)
+#' task = tsk("cookfarm")
+#'
+#' # Instantiate Resampling
+#' rrcv = rsmp("repeated-sptcv-cstf", folds = 3, repeats = 5)
+#' rrcv$instantiate(task, time_var = "Date")
+#' }
+#' # Individual sets:
+#' rrcv$iters
+#' rrcv$folds(1:6)
+#' rrcv$repeats(1:6)
+#'
+#' # Individual sets:
+#' rrcv$train_set(1)
+#' rrcv$test_set(1)
+#' intersect(rrcv$train_set(1), rrcv$test_set(1))
+#'
+#' # Internal storage:
+#' rrcv$instance # table
+ResamplingRepeatedSptCVCstf = R6Class("ResamplingRepeatedSptCVCstf",
+  inherit = mlr3::Resampling,
+
+  public = list(
+    #' @description
+    #' Create an "coordinate-based" repeated resampling instance.
+    #' @param id `character(1)`\cr
+    #'   Identifier for the resampling strategy.
+    initialize = function(id = "repeated-sptcv-cstf") {
+      ps = ParamSet$new(params = list(
+        ParamInt$new("folds", lower = 1L, default = 10L, tags = "required"),
+        ParamInt$new("repeats", lower = 1, default = 1L, tags = "required")
+      ))
+      ps$values = list(folds = 10L, repeats = 1)
+      super$initialize(
+        id = id,
+        param_set = ps,
+        man = "mlr3spatiotempcv::mlr_resamplings_repeated_SptCVCstf"
+      )
+    },
+
+    #' @description Translates iteration numbers to fold number.
+    #' @param iters `integer()`\cr
+    #'   Iteration number.
+    folds = function(iters) {
+      iters = assert_integerish(iters, any.missing = FALSE, coerce = TRUE)
+      ((iters - 1L) %% as.integer(self$param_set$values$repeats)) + 1L
+    },
+
+    #' @description Translates iteration numbers to repetition number.
+    #' @param iters `integer()`\cr
+    #'   Iteration number.
+    repeats = function(iters) {
+      iters = assert_integerish(iters, any.missing = FALSE, coerce = TRUE)
[…]

### 2020-08-15 R/helper_sample.R

@@ -1,0 +1,65 @@
+sample_cstf <- function(self, task, space_var, time_var, class, k, data) {
+
+  # if classification is used, make sure that classes are equally
+  # distributed across folds
+
+  if (!is.null(class)) {
+    unit = unique(data[, c(..space_var, ..class)])
+    # unit needs to be a data.frame here
+    unit$cstf_fold = caret::createFolds(
+      as.data.frame(unit)[, which(names(unit) == class)],
+      k = k, list = FALSE)
+    unit$cstf_fold = as.factor(as.character(unit$cstf_fold))
+    data = merge(data, unit,
+      by.x = c(space_var, class),
+      by.y = c(space_var, class), all.x = TRUE, sort = FALSE)
+    space_var = "cstf_fold"
+  }
+
+  if (!is.null(space_var)) {
+    checkmate::assert_factor(data[[space_var]], null.ok = TRUE)
+    if (k > uniqueN(data[[space_var]])) {
+      k = uniqueN(data[[space_var]])
+      cli::cli_alert_warning("Number of folds is higher than number of
+            unique locations.
+            Setting folds to the maximum amount of unique levels of variable
+            {space_var} which is {k}.", wrap = TRUE)
+    }
+  }
+  if (!is.null(time_var)) {
+    checkmate::assert_factor(data[[time_var]], null.ok = TRUE)
+    if (k > uniqueN(data[[time_var]])) {
+      k = uniqueN(data[[time_var]])
+      cli::cli_alert_warning("Number of folds is higher than number of
+            unique points in time.
+            Setting folds to the maximum amount of unique levels of variable
+            {time_var} which is {k}.", wrap = TRUE)
+    }
+  }
+
+  seed = sample(1:1000, 1)
+
+  # split space into k folds
+  if (!is.null(space_var)) {
+    set.seed(seed)
+    spacefolds = lapply(caret::createFolds(
+      1:uniqueN(data[[space_var]]),
+      k), function(y) {
+      unique(data[[space_var]])[y]
+    })
+  } else {
+    spacefolds = NULL
+  }
+  # split time into k folds
+  if (!is.null(time_var)) {
+    set.seed(seed)
+    timefolds = lapply(caret::createFolds(
+      1:uniqueN(data[[time_var]]),
+      k), function(y) {
+      unique(data[[time_var]])[y]
+    })
+  } else {
+    timefolds = NULL
+  }
+  return(list(spacefolds = spacefolds, timefolds = timefolds))
+}

### 2020-08-16 README.Rmd
Add `ResamplingRepeatedSptCVCstf` (#60)

@@
 task_st = tsk("cookfarm")
-resampling = rsmp("spcv-cluto", folds = 5)
+resampling = rsmp("sptcv-cluto", folds = 5)
 resampling$instantiate(task_st, time_var = "Date")

### 2020-08-16 R/ResamplingSptCVCluto.R

@@ -11,7 +11,7 @@
 #' task = tsk("cookfarm")
 #'
 #' # Instantiate Resampling
-#' rcv = rsmp("spcv-cluto", folds = 5)
+#' rcv = rsmp("sptcv-cluto", folds = 5)
 #' rcv$instantiate(task, "Date")
 #'
 #' # Individual sets:
@@ -31,7 +31,7 @@ ResamplingSptCVCluto = R6Class("ResamplingSptCVCluto",
     #' FIXME.
     #' @param id `character(1)`\cr
     #'   Identifier for the resampling strategy.
-    initialize = function(id = "spcv-cluto") {
+    initialize = function(id = "sptcv-cluto") {
       ps = ParamSet$new(params = list(
         ParamInt$new("folds", lower = 1L, default = 10L, tags = "required")
       ))

### 2020-08-16 R/autoplot.R

@@ -291,7 +291,7 @@ autoplot.ResamplingSpCVCoords = function( # nolint
 #' library(mlr3)
 #' library(mlr3spatiotempcv)
 #' task_st = tsk("cookfarm")
-#' resampling = rsmp("spcv-cluto", folds = 5)
+#' resampling = rsmp("sptcv-cluto", folds = 5)
 #' resampling$instantiate(task_st, "Date")
 #'
 #' # plot
@@ -353,7 +353,7 @@ autoplot.ResamplingSptCVCluto = function( # nolint
 #' library(mlr3)
 #' library(mlr3spatiotempcv)
 #' task_st = tsk("cookfarm")
-#' resampling = rsmp("spcv-cstf", folds = 5)
+#' resampling = rsmp("sptcv-cstf", folds = 5)
 #' resampling$instantiate(task_st, "Date")
 #'
 #' # plot
@@ -772,7 +772,7 @@ autoplot.ResamplingRepeatedSpCVBlock = function( # nolint
 #' library(mlr3)
 #' library(mlr3spatiotempcv)
 #' task_st = tsk("cookfarm")
-#' resampling = rsmp("repeated-spcv-cluto", folds = 5, repeats = 2)
+#' resampling = rsmp("repeated-sptcv-cluto", folds = 5, repeats = 2)
 #' resampling$instantiate(task_st, "Date")
 #'
 #' # plot

### 2020-08-16 R/helper_sample.R

@@ -1,0 +1,65 @@
+sample_cstf <- function(self, task, space_var, time_var, class, k, data) {
+
+  # if classification is used, make sure that classes are equally
+  # distributed across folds
+
+  if (!is.null(class)) {
+    select_cols = c(space_var, class)
+    unit = unique(data[, select_cols, with = FALSE])
+    # unit needs to be a data.frame here
+    unit$cstf_fold = caret::createFolds(
+      as.data.frame(unit)[, which(names(unit) == class)],
+      k = k, list = FALSE)
+    data = merge(data, unit,
+      by.x = c(space_var, class),
+      by.y = c(space_var, class), all.x = TRUE, sort = FALSE)
+    space_var = "cstf_fold"
+  }
+
+  if (!is.null(space_var)) {
+    if (k > uniqueN(data[[space_var]])) {
+      k = uniqueN(data[[space_var]])
+      cli::cli_alert_warning("Number of folds is higher than number of
+            unique locations.
+            Setting folds to the maximum amount of unique levels of variable
+            {space_var} which is {k}.", wrap = TRUE)
+    }
+  }
+  if (!is.null(time_var)) {
+    if (k > uniqueN(data[[time_var]])) {
+      k = uniqueN(data[[time_var]])
+      cli::cli_alert_warning("Number of folds is higher than number of
+            unique points in time.
+            Setting folds to the maximum amount of unique levels of variable
+            {time_var} which is {k}.", wrap = TRUE)
+    }
+  }
+
+  seed = sample(1:1000, 1)
+
+  # split space into k folds
+  if (!is.null(space_var)) {
+    set.seed(seed)
+    spacefolds = lapply(caret::createFolds(
+      1:uniqueN(data[[space_var]]),
+      k), function(y) {
+      unique(data[[space_var]])[y]
+    })
+  } else {
+    spacefolds = NULL
+  }
+  # split time into k folds
+  if (!is.null(time_var)) {
+    set.seed(seed)
+    timefolds = lapply(caret::createFolds(
+      1:uniqueN(data[[time_var]]),
+      k), function(y) {
+      unique(data[[time_var]])[y]
+    })
+  } else {
+    timefolds = NULL
+  }
+  return(list(
+    spacefolds = spacefolds, timefolds = timefolds,
+    space_var = space_var, data = data))
+}

### 2020-08-16 R/zzz.R

@@ -48,15 +48,16 @@ register_mlr3 = function() {
     x = utils::getFromNamespace("mlr_resamplings", ns = "mlr3")
     mlr_resamplings$add("spcv-block", ResamplingSpCVBlock)
     mlr_resamplings$add("spcv-buffer", ResamplingSpCVBuffer)
-    mlr_resamplings$add("spcv-cstf", ResamplingSptCVCstf)
-    mlr_resamplings$add("spcv-cluto", ResamplingSptCVCluto)
+    mlr_resamplings$add("sptcv-cstf", ResamplingSptCVCstf)
+    mlr_resamplings$add("sptcv-cluto", ResamplingSptCVCluto)
     mlr_resamplings$add("spcv-coords", ResamplingSpCVCoords)
     mlr_resamplings$add("spcv-env", ResamplingSpCVEnv)
 
     mlr_resamplings$add("repeated-spcv-coords", ResamplingRepeatedSpCVCoords)
     mlr_resamplings$add("repeated-spcv-env", ResamplingRepeatedSpCVEnv)
     mlr_resamplings$add("repeated-spcv-block", ResamplingRepeatedSpCVBlock)
-    mlr_resamplings$add("repeated-spcv-cluto", ResamplingRepeatedSptCVCluto)
+    mlr_resamplings$add("repeated-sptcv-cluto", ResamplingRepeatedSptCVCluto)
+    mlr_resamplings$add("repeated-sptcv-cstf", ResamplingRepeatedSptCVCstf)
   }
 
   utils::globalVariables(c(

### 2020-08-16 R/ResamplingRepeatedSptCVCstf.R
Add `ResamplingRepeatedSptCVCstf` (#60)

@@ -4,15 +4,13 @@
 #'
 #' @export
 #' @examples
-#' \dontrun{
 #' library(mlr3)
 #' library(mlr3spatiotempcv)
 #' task = tsk("cookfarm")
 #'
 #' # Instantiate Resampling
-#' rrcv = rsmp("repeated-sptcv-cluto", folds = 3, repeats = 5)
+#' rrcv = rsmp("repeated-sptcv-cstf", folds = 3, repeats = 5)
 #' rrcv$instantiate(task, time_var = "Date")
-#'
 #' # Individual sets:
 #' rrcv$iters
 #' rrcv$folds(1:6)
@@ -25,8 +23,7 @@
 #'
 #' # Internal storage:
 #' rrcv$instance # table
-#' }
-ResamplingRepeatedSptCVCluto = R6Class("ResamplingRepeatedSptCVCluto",
+ResamplingRepeatedSptCVCstf = R6Class("ResamplingRepeatedSptCVCstf",
   inherit = mlr3::Resampling,
 
   public = list(
@@ -34,7 +31,7 @@ ResamplingRepeatedSptCVCluto = R6Class("ResamplingRepeatedSptCVCluto",
     #' Create an "coordinate-based" repeated resampling instance.
     #' @param id `character(1)`\cr
     #'   Identifier for the resampling strategy.
-    initialize = function(id = "repeated-sptcv-cluto") {
+    initialize = function(id = "repeated-sptcv-cstf") {
       ps = ParamSet$new(params = list(
         ParamInt$new("folds", lower = 1L, default = 10L, tags = "required"),
         ParamInt$new("repeats", lower = 1, default = 1L, tags = "required")
@@ -43,7 +40,7 @@ ResamplingRepeatedSptCVCluto = R6Class("ResamplingRepeatedSptCVCluto",
       super$initialize(
         id = id,
         param_set = ps,
-        man = "mlr3spatiotempcv::mlr_resamplings_repeated_SptCVCluto"
+        man = "mlr3spatiotempcv::mlr_resamplings_repeated_SptCVCstf"
       )
     },
 
@@ -66,24 +63,15 @@ ResamplingRepeatedSptCVCluto = R6Class("ResamplingRepeatedSptCVCluto",
     #' @description
     #'  Materializes fixed training and test splits for a given task.
     #' @param task [Task]\cr
-    #'  A task to instantiate.
-    #' @param time_var [character]\cr
-    #'  The name of the variable which represents the time dimension.
[…]

### 2020-08-16 tests/testthat/test_mlr_spcv_generic.R

@@ -43,9 +43,9 @@ test_that("train and test set getter functions are working", {
   spcv_rsp$`repeated-spcv-coords` = NULL
   spcv_rsp$`repeated-spcv-env` = NULL
   spcv_rsp$`spcv-block` = NULL
-  spcv_rsp$`spcv-cluto` = NULL
-  spcv_rsp$`spcv-cstf` = NULL
-  spcv_rsp$`repeated-spcv-cluto` = NULL
+  spcv_rsp$`sptcv-cluto` = NULL
+  spcv_rsp$`sptcv-cstf` = NULL
+  spcv_rsp$`repeated-sptcv-cluto` = NULL
   # nolint end
 
   for (i in spcv_rsp) {
@@ -69,8 +69,8 @@ test_that("cloning works", {
   spcv_rsp$`repeated-spcv-coords` = NULL
   spcv_rsp$`repeated-spcv-env` = NULL
   spcv_rsp$`repeated-spcv-block` = NULL
-  spcv_rsp$`spcv-cluto` = NULL
-  spcv_rsp$`repeated-spcv-cluto` = NULL
+  spcv_rsp$`sptcv-cluto` = NULL
+  spcv_rsp$`repeated-sptcv-cluto` = NULL
   # nolint end
 
   for (i in spcv_rsp) {

### 2020-08-26 README.Rmd
adapt to mlr3 "extra_args" approach during Task construction

--- 2020-08-15 README.Rmd
@@
 | Literature                  | Package    | Reference     | mlr3 Sugar                     |
-|--------------------------------------|------------|---------------|--------------------------------|
+|-----------------------------|------------|---------------|--------------------------------|
 | Spatial Buffering           | blockCV    | Valavi 2019   | `rsmp("spcv-buffer")`          |
@@
 | \-                          | \-         | \-            | `rsmp("repeated-sptcv-cluto")` |
-| Repeated Leave-Location-and-Time-Out | CAST       | Meyer 2018    | `rsmp("repeated-sptcv-cstf")`  |
 

### 2020-08-26 R/TaskClassifST.R

@@ -60,8 +60,8 @@ TaskClassifST = R6::R6Class("TaskClassifST",
         id = id, backend = backend, target = target,
         positive = positive, extra_args = extra_args)
 
-      self$extra_args$coordinate_names = coordinate_names
-      self$extra_args$crs = checkmate::assert_character(crs, null.ok = TRUE)
+      self$extra_args$coordinate_names = extra_args$coordinate_names
+      self$extra_args$crs = checkmate::assert_character(extra_args$crs, null.ok = TRUE)
 
       info = self$col_info[id == target]
       levels = info$levels[[1L]]
@@ -81,15 +81,15 @@ TaskClassifST = R6::R6Class("TaskClassifST",
       }
 
       # check coordinates
-      assert_names(self$backend$colnames, must.include = coordinate_names)
-      for (coord in coordinate_names) {
+      assert_names(self$backend$colnames, must.include = extra_args$coordinate_names)
+      for (coord in extra_args$coordinate_names) {
         assert_numeric(self$data(cols = coord)[[1L]], any.missing = FALSE)
       }
 
       # mark columns as coordinates and check if coordinates should be included
       # as features
       self$col_roles$coordinates = extra_args$coordinate_names
-      if (isFALSE(coords_as_features)) {
+      if (isFALSE(extra_args$coords_as_features)) {
         self$col_roles$feature = setdiff(
           self$col_roles$feature,
           extra_args$coordinate_names)

### 2020-08-26 R/autoplot.R

@@ -174,7 +174,7 @@ autoplot.ResamplingSpCVBuffer = function( # nolint
 
     table = rbind(coords_train, coords_test)
 
-    sf_df = sf::st_as_sf(table, coords = c("x", "y"), crs = task$crs)
+    sf_df = sf::st_as_sf(table, coords = c("x", "y"), crs = task$extra_args$crs)
     sf_df$indicator = as.factor(as.character(sf_df$indicator))
 
     # reorder factor levels so that "train" comes first
@@ -209,7 +209,7 @@ autoplot.ResamplingSpCVBuffer = function( # nolint
 
     table = rbind(coords_train, coords_test)
 
-    sf_df = sf::st_as_sf(table, coords = c("x", "y"), crs = task$crs)
+    sf_df = sf::st_as_sf(table, coords = c("x", "y"), crs = task$extra_args$crs)
     # 'fold' needs to be a factor, otherwise `show.legend = "points" has no
     # effect
     sf_df$indicator = as.factor(as.character(sf_df$indicator))
@@ -858,7 +858,7 @@ autoplot_spatial = function(
 
       table[, indicator := ifelse(fold == i, "Test", "Train")]
 
-      sf_df = sf::st_as_sf(table, coords = c("x", "y"), crs = task$crs)
+      sf_df = sf::st_as_sf(table, coords = c("x", "y"), crs = task$extra_args$crs)
       sf_df$indicator = as.factor(as.character(sf_df$indicator))
 
       # reorder factor levels so that "train" comes first
@@ -894,7 +894,7 @@ autoplot_spatial = function(
       # Extract legend standalone, we only want one legend in the grid
       coords_resamp[, indicator := ifelse(fold == 1, "Test", "Train")]
 
-      sf_df = sf::st_as_sf(coords_resamp, coords = c("x", "y"), crs = task$crs)
+      sf_df = sf::st_as_sf(coords_resamp, coords = c("x", "y"), crs = task$extra_args$crs)
       # 'fold' needs to be a factor, otherwise `show.legend = "points" has no
       # effect
       sf_df$indicator = as.factor(as.character(sf_df$indicator))
@@ -919,7 +919,7 @@ autoplot_spatial = function(
     }
   } else {
     # Create one plot with all (test)-folds
-    sf_df = sf::st_as_sf(coords_resamp, coords = c("x", "y"), crs = task$crs)
[…]

### 2020-09-01 README.Rmd
upd README

--- 2020-08-15 README.Rmd
@@
 
-S3 `autoplot()` for all implemented spatial resampling methods.
+S3 `plot()` method for all implemented spatial resampling methods.
 
@@
 
-autoplot(resampling, task)
+plot(resampling, task, crs = 4326)
 ```
@@
 ```{r spcv-coords-fold}
-autoplot(resampling, task, fold_id = 1)
+plot(resampling, task, fold_id = 1, crs = 4326)
 ```
@@
 task_st = tsk("cookfarm")
-resampling = rsmp("spcv-cluto", folds = 5)
+resampling = rsmp("sptcv-cluto", folds = 5)
 resampling$instantiate(task_st, time_var = "Date")
-autoplot(resampling, task_st, fold_id = 1, point_size = 3)
+plot(resampling, task_st, fold_id = 1, point_size = 3)
 ```

### 2020-09-01 tests/testthat/test-autoplot_buffer.R
update vignettes

@@ -12,7 +12,6 @@ test_that("plot() works for 'spcv-buffer'", {
   p2 = autoplot(plots$rsp, plots$task, 1, crs = 4326)
   p3 = autoplot(plots$rsp, plots$task, c(1, 2), crs = 4326)
 
-  expect_true(is.ggplot(p1))
   expect_true(is.ggplot(p2))
   expect_true(is.ggplot(p2))
 

### 2020-09-01 R/ResamplingRepeatedSpCVBlock.R
- use underscores instead of hyphens for resampling method naming

@@ -8,7 +8,7 @@
 #' task = tsk("diplodia")
 #'
 #' # Instantiate Resampling
-#' rrcv = rsmp("repeated-spcv-block",
+#' rrcv = rsmp("repeated_spcv_block",
 #'   folds = 3, repeats = 2,
 #'   range = c(5000, 10000))
 #' rrcv$instantiate(task)
@@ -33,7 +33,7 @@ ResamplingRepeatedSpCVBlock = R6Class("ResamplingRepeatedSpCVBlock",
     #' Create an "coordinate-based" repeated resampling instance.
     #' @param id `character(1)`\cr
     #'   Identifier for the resampling strategy.
-    initialize = function(id = "repeated-spcv-block") {
+    initialize = function(id = "repeated_spcv_block") {
       ps = ParamSet$new(params = list(
         ParamInt$new("folds", lower = 1L, default = 10L, tags = "required"),
         ParamInt$new("repeats", lower = 1, default = 1L, tags = "required"),

### 2020-09-01 R/ResamplingSptCVCstf.R

@@ -7,7 +7,7 @@
 #' task = tsk("cookfarm")
 #'
 #' # Instantiate Resampling
-#' rcv = rsmp("sptcv-cstf", folds = 5)
+#' rcv = rsmp("sptcv_cstf", folds = 5)
 #' rcv$instantiate(task, time_var = "Date", space_var = "SOURCEID")
 #'
 #' # Individual sets:
@@ -26,7 +26,7 @@ ResamplingSptCVCstf = R6Class("ResamplingSptCVCstf",
     #' Create a "Spacetime Folds" resampling instance.
     #' @param id `character(1)`\cr
     #'   Identifier for the resampling strategy.
-    initialize = function(id = "sptcv-cstf") {
+    initialize = function(id = "sptcv_cstf") {
       ps = ParamSet$new(params = list(
         ParamInt$new("folds", lower = 1L, default = 10L, tags = "required")
       ))

### 2020-09-01 R/autoplot.R

@@ -45,7 +45,7 @@
 #' \dontrun{
 #' library(mlr3)
 #' task = tsk("ecuador")
-#' resampling = rsmp("spcv-block", range = 1000)
+#' resampling = rsmp("spcv_block", range = 1000)
 #' resampling$instantiate(task)
 #'
 #' ## Visualize all partitions
@@ -139,7 +139,7 @@ plot.ResamplingRepeatedSpCVBlock = function(x, ...) {
 #' ##########
 #' \dontrun{
 #' library(mlr3)
-#' resampling = rsmp("spcv-env", folds = 4, features = "dem")
+#' resampling = rsmp("spcv_env", folds = 4, features = "dem")
 #' resampling$instantiate(task)
 #'
 #' autoplot(resampling, task)
@@ -219,7 +219,7 @@ plot.ResamplingRepeatedSpCVEnv = function(x, ...) {
 #' ##########
 #' \dontrun{
 #' library(mlr3)
-#' resampling = rsmp("spcv-coords")
+#' resampling = rsmp("spcv_coords")
 #' resampling$instantiate(task)
 #'
 #' autoplot(resampling, task)
@@ -313,7 +313,7 @@ plot.ResamplingRepeatedSpCVCoords = function(x, ...) {
 #' library(mlr3)
 #' library(mlr3spatiotempcv)
 #' task_st = tsk("cookfarm")
-#' resampling = rsmp("sptcv-cluto", folds = 5)
+#' resampling = rsmp("sptcv_cluto", folds = 5)
 #' resampling$instantiate(task_st, "Date")
 #'
 #' # plot

### 2020-09-01 R/autoplot_spcv_buffer.R

@@ -6,7 +6,7 @@
 #' ##########
 #' \dontrun{
 #' library(mlr3)
-#' resampling = rsmp("spcv-buffer", theRange = 1000)
+#' resampling = rsmp("spcv_buffer", theRange = 1000)
 #' resampling$instantiate(task)
 #'
 #' autoplot(resampling, task, 1)

### 2020-09-01 tests/testthat/test-ResamplingRepeatedSpCVBlock.R

@@ -2,7 +2,7 @@ context("ResamplingRepeatedSpCVBlock")
 
 test_that("folds can be printed", {
   task = test_make_twoclass()
-  rsp = rsmp("repeated-spcv-block", folds = 3, repeats = 2, range = c(2, 4))
+  rsp = rsmp("repeated_spcv_block", folds = 3, repeats = 2, range = c(2, 4))
   rsp$instantiate(task)
 
   expect_equal(rsp$folds(1:6), c(1, 2, 1, 2, 1, 2))
@@ -10,7 +10,7 @@ test_that("folds can be printed", {
 
 test_that("reps can be printed", {
   task = test_make_twoclass()
-  rsp = rsmp("repeated-spcv-block", folds = 3, repeats = 2, range = c(2, 4))
+  rsp = rsmp("repeated_spcv_block", folds = 3, repeats = 2, range = c(2, 4))
   rsp$instantiate(task)
 
   expect_equal(rsp$repeats(1:9), c(1, 1, 1, 2, 2, 2, 3, 3, 3))
@@ -18,7 +18,7 @@ test_that("reps can be printed", {
 
 test_that("resampling iterations equals folds * repeats", {
   task = test_make_twoclass()
-  rsp = rsmp("repeated-spcv-block", folds = 3, repeats = 2, range = c(2, 4))
+  rsp = rsmp("repeated_spcv_block", folds = 3, repeats = 2, range = c(2, 4))
   rsp$instantiate(task)
 
   expect_equal(rsp$iters, 6)
@@ -26,7 +26,7 @@ test_that("resampling iterations equals folds * repeats", {
 
 test_that("error when neither cols & rows | range is specified", {
   task = test_make_twoclass()
-  rsmp = rsmp("repeated-spcv-block", repeats = 2)
+  rsmp = rsmp("repeated_spcv_block", repeats = 2)
   expect_error(
     rsmp$instantiate(task),
     "Either 'range' or 'cols' & 'rows' need to be set.")
@@ -34,7 +34,7 @@ test_that("error when neither cols & rows | range is specified", {
 
 test_that("error when length(range) != length(repeats)", {
   task = test_make_twoclass()
-  rsp = rsmp("repeated-spcv-block", repeats = 2, range = 5)
+  rsp = rsmp("repeated_spcv_block", repeats = 2, range = 5)
   expect_error(
     rsp$instantiate(task),
     ".*to be the same length as 'range'."
[…]

### 2020-09-01 tests/testthat/test-ResamplingSpCVBuffer.R

@@ -2,7 +2,7 @@ context("ResamplingSpCVBuffer")
 
 test_that("resampling iterations equals number of observations (two-class response)", {
   task = test_make_twoclass()
-  rsp = rsmp("spcv-buffer", theRange = 1)
+  rsp = rsmp("spcv_buffer", theRange = 1)
   rsp$instantiate(task)
 
   expect_equal(rsp$iters, 36)
@@ -10,7 +10,7 @@ test_that("resampling iterations equals number of observations (two-class respon
 
 test_that("resampling iterations equals number of observations (multi-class response)", {
   task = test_make_multiclass()
-  rsp = rsmp("spcv-buffer", theRange = 1)
+  rsp = rsmp("spcv_buffer", theRange = 1)
   rsp$instantiate(task)
 
   expect_equal(rsp$iters, 36)
@@ -18,7 +18,7 @@ test_that("resampling iterations equals number of observations (multi-class resp
 
 test_that("resampling iterations equals number of observations (continuous response)", {
   task = test_make_regr()
-  rsp = rsmp("spcv-buffer", theRange = 1)
+  rsp = rsmp("spcv_buffer", theRange = 1)
   rsp$instantiate(task)
 
   expect_equal(rsp$iters, 36)
@@ -26,7 +26,7 @@ test_that("resampling iterations equals number of observations (continuous respo
 
 test_that("buffered observations are excludes in train set", {
   task = test_make_twoclass()
-  rsp = rsmp("spcv-buffer", theRange = 1, spDataType = "PA")
+  rsp = rsmp("spcv_buffer", theRange = 1, spDataType = "PA")
   rsp$instantiate(task)
 
   expect_true(all(!c(1, 2, 7) %in% rsp$train_set(1)))
@@ -35,7 +35,7 @@ test_that("buffered observations are excludes in train set", {
 
 test_that("test sets only include positive observations", {
   task = test_make_twoclass()
-  rsp = rsmp("spcv-buffer", theRange = 1, spDataType = "PB", addBG = FALSE)
+  rsp = rsmp("spcv_buffer", theRange = 1, spDataType = "PB", addBG = FALSE)
   rsp$instantiate(task)
 
   expect_equal(rsp$iters, 18)
@@ -44,7 +44,7 @@ test_that("test sets only include positive observations", {
 
 test_that("test sets include background observations", {
[…]

### 2020-09-01 tests/testthat/test-ResamplingSptCVCstf.R

@@ -2,7 +2,7 @@ context("ResamplingSptCVCstf")
 
 test_that("mlr3spatiotempcv implementation is equal to CAST: time + space", {
   task = tsk("cookfarm")
-  rsp = rsmp("sptcv-cstf", folds = 10)
+  rsp = rsmp("sptcv_cstf", folds = 10)
   set.seed(42)
   rsp$instantiate(task, time_var = "Date", space_var = "SOURCEID")
   rsp$test_set(1)
@@ -16,7 +16,7 @@ test_that("mlr3spatiotempcv implementation is equal to CAST: time + space", {
 
 test_that("mlr3spatiotempcv implementation is equal to CAST: time ", {
   task = tsk("cookfarm")
-  rsp = rsmp("sptcv-cstf", folds = 10)
+  rsp = rsmp("sptcv_cstf", folds = 10)
   set.seed(42)
   rsp$instantiate(task, time_var = "Date")
   rsp$test_set(1)
@@ -30,7 +30,7 @@ test_that("mlr3spatiotempcv implementation is equal to CAST: time ", {
 
 test_that("mlr3spatiotempcv implementation is equal to CAST: space", {
   task = tsk("cookfarm")
-  rsp = rsmp("sptcv-cstf", folds = 10)
+  rsp = rsmp("sptcv_cstf", folds = 10)
   set.seed(42)
   rsp$instantiate(task, space_var = "SOURCEID")
   rsp$test_set(1)
@@ -44,7 +44,7 @@ test_that("mlr3spatiotempcv implementation is equal to CAST: space", {
 
 test_that("mlr3spatiotempcv implementation is equal to CAST: space", {
   task = tsk("cookfarm")
-  rsp = rsmp("sptcv-cstf", folds = 10)
+  rsp = rsmp("sptcv_cstf", folds = 10)
   set.seed(42)
   rsp$instantiate(task, space_var = "SOURCEID", class = "TAXSUSDA")
   rsp$test_set(1)

### 2020-09-01 tests/testthat/test-autoplot_buffer.R

@@ -1,11 +1,11 @@
 context("plot")
 
-# spcv-buffer ------------------------------------------------------------------
+# spcv_buffer ------------------------------------------------------------------
 
-test_that("plot() works for 'spcv-buffer'", {
+test_that("plot() works for 'spcv_buffer'", {
   set.seed(42)
 
-  plots = prepare_autoplot("spcv-buffer", theRange = 2)
+  plots = prepare_autoplot("spcv_buffer", theRange = 2)
 
   # autoplot() is used instead of plot() to prevent side-effect plotting
   expect_error(autoplot(plots$rsp, plots$task, crs = 4326))

### 2020-09-01 tests/testthat/test-autoplot_cluto.R

@@ -1,13 +1,13 @@
 context("plot")
 
-# SptCV-Cluto ------------------------------------------------------------------
+# sptcv_Cluto ------------------------------------------------------------------
 
-test_that("plot() works for 'sptcv-cluto'", {
+test_that("plot() works for 'sptcv_cluto'", {
   skip_on_os("mac")
   set.seed(42)
 
   task = tsk("cookfarm")
-  rsp = rsmp("sptcv-cluto", folds = 4)
+  rsp = rsmp("sptcv_cluto", folds = 4)
   rsp$instantiate(task, time_var = "Date")
 
   # autoplot() is used instead of plot() to prevent side-effect plotting
@@ -25,13 +25,13 @@ test_that("plot() works for 'sptcv-cluto'", {
   vdiffr::expect_doppelganger("SptCVCluto - Fold 1-4", p3)
 })
 
-test_that("plot() works for 'repeated-sptcv-cluto'", {
+test_that("plot() works for 'repeated_sptcv_cluto'", {
   skip_on_os("mac")
 
   set.seed(42)
 
   task = tsk("cookfarm")
-  rsp = rsmp("repeated-sptcv-cluto", folds = 4)
+  rsp = rsmp("repeated_sptcv_cluto", folds = 4)
   rsp$instantiate(task, time_var = "Date")
 
   # autoplot() is used instead of plot() to prevent side-effect plotting

### 2020-09-01 R/ResamplingSpCVBlock.R
- use underscores instead of hyphens for resampling method naming

@@ -8,7 +8,7 @@
 #' task = tsk("ecuador")
 #'
 #' # Instantiate Resampling
-#' rcv = rsmp("spcv-block", range = 1000)
+#' rcv = rsmp("spcv_block", range = 1000)
 #' rcv$instantiate(task)
 #'
 #' # Individual sets:
@@ -26,7 +26,7 @@ ResamplingSpCVBlock = R6Class("ResamplingSpCVBlock",
     #' Create an "Environmental Block" resampling instance.
     #' @param id `character(1)`\cr
     #'   Identifier for the resampling strategy.
-    initialize = function(id = "spcv-block") {
+    initialize = function(id = "spcv_block") {
       ps = ParamSet$new(params = list(
         ParamInt$new("folds", lower = 1L, default = 10L, tags = "required"),
         ParamInt$new("rows", lower = 1L),

### 2020-09-02 README.Rmd

@@
 |--------------------------------------|------------|---------------|--------------------------------|
-| Spatial Buffering                    | blockCV    | Valavi 2019   | `rsmp("spcv-buffer")`          |
-| Spatial Blocking                     | blockCV    | Valavi 2019   | `rsmp("spcv-block")`           |
-| Spatial CV                           | sperrorest | Brenning 2012 | `rsmp("spcv-coords")`          |
-| Environmental Blocking               | blockCV    | Valavi 2019   | `rsmp("spcv-env")`             |
-| \-                                   | \-         | \-            | `rsmp("sptcv-cluto")`          |
-| Leave-Location-and-Time-Out          | CAST       | Meyer 2018    | `rsmp("sptcv-cstf")`           |
+| Spatial Buffering                    | blockCV    | Valavi 2019   | `rsmp("spcv_buffer")`          |
+| Spatial Blocking                     | blockCV    | Valavi 2019   | `rsmp("spcv_block")`           |
+| Spatial CV                           | sperrorest | Brenning 2012 | `rsmp("spcv_coords")`          |
+| Environmental Blocking               | blockCV    | Valavi 2019   | `rsmp("spcv_env")`             |
+| \-                                   | \-         | \-            | `rsmp("sptcv_cluto")`          |
+| Leave-Location-and-Time-Out          | CAST       | Meyer 2018    | `rsmp("sptcv_cstf")`           |
 | ---                                  | ---        | ---           | ---                            |
-| Repeated Spatial Blocking            | blockCV    | Valavi 2019   | `rsmp("repeated-spcv-block")`  |
-| Repeated Spatial CV                  | sperrorest | Brenning 2012 | `rsmp("repeated-spcv-coords")` |
-| Repeated Env Blocking                | blockCV    | Valavi 2019   | `rsmp("repeated-spcv-env")`    |
-| \-                                   | \-         | \-            | `rsmp("repeated-sptcv-cluto")` |
-| Repeated Leave-Location-and-Time-Out | CAST       | Meyer 2018    | `rsmp("repeated-sptcv-cstf")`  |
+| Repeated Spatial Blocking            | blockCV    | Valavi 2019   | `rsmp("repeated_spcv_block")`  |
+| Repeated Spatial CV                  | sperrorest | Brenning 2012 | `rsmp("repeated_spcv_coords")` |
+| Repeated Env Blocking                | blockCV    | Valavi 2019   | `rsmp("repeated_spcv_env")`    |
+| \-                                   | \-         | \-            | `rsmp("repeated_sptcv_cluto")` |
+| Repeated Leave-Location-and-Time-Out | CAST       | Meyer 2018    | `rsmp("repeated_sptcv_cstf")`  |
 
@@
 
-```{r spcv-coords-all-partitions, warning=FALSE}
+```{r spcv_coords-all-partitions, warning=FALSE}
 library(mlr3)
@@
 task = tsk("ecuador")
-resampling = rsmp("spcv-coords", folds = 5)
+resampling = rsmp("spcv_coords", folds = 5)
 resampling$instantiate(task)
@@
 
-```{r spcv-coords-fold}
+```{r spcv_coords-fold}
 plot(resampling, task, fold_id = 1, crs = 4326)
@@
 task_st = tsk("cookfarm")
-resampling = rsmp("sptcv-cluto", folds = 5)
+resampling = rsmp("sptcv_cluto", folds = 5)
 resampling$instantiate(task_st, time_var = "Date")

### 2020-10-08 R/autoplot_spcv_buffer.R
replace {cowplot} by {patchwork} and make visualization of multiple folds more flexibe and eye appealing

@@ -6,11 +6,12 @@
 #' ##########
 #' \dontrun{
 #' library(mlr3)
+#' task = tsk("ecuador")
 #' resampling = rsmp("spcv_buffer", theRange = 1000)
 #' resampling$instantiate(task)
 #'
 #' autoplot(resampling, task, 1)
-#' autoplot(resampling, task, c(1, 2, 3, 4))
+#' autoplot(resampling, task, c(1, 2))
 #' }
 autoplot.ResamplingSpCVBuffer = function( # nolint
   object,
@@ -22,7 +23,7 @@ autoplot.ResamplingSpCVBuffer = function( # nolint
   crs = NULL,
   ...) {
 
-  require_namespaces(c("sf", "cowplot"))
+  require_namespaces(c("sf", "patchwork"))
   resampling = object
 
   resampling = assert_autoplot(resampling, fold_id, task)
@@ -36,12 +37,10 @@ autoplot.ResamplingSpCVBuffer = function( # nolint
     stopf("Plotting all folds of a LOOCV instance is not supported")
   }
 
-  indicator = NULL
+  plot_list = mlr3misc::map(fold_id, function(.x) {
 
-  plot_list = list()
-  for (i in fold_id) {
-    coords_train = coords[row_id %in% resampling$train_set(i)]
-    coords_test = coords[row_id %in% resampling$test_set(i)]
+    coords_train = coords[row_id %in% resampling$train_set(.x)]
+    coords_test = coords[row_id %in% resampling$test_set(.x)]
 
     coords_train$indicator = "Train"
     coords_test$indicator = "Test"
@@ -64,67 +63,32 @@ autoplot.ResamplingSpCVBuffer = function( # nolint
     # reorder factor levels so that "train" comes first
     sf_df$indicator = ordered(sf_df$indicator, levels = c("Train", "Test"))
 
-    plot_list[[length(plot_list) + 1]] =
-      ggplot() +
+    ggplot() +
       geom_sf(data = sf_df, aes(color = .data[["indicator"]])) +
       scale_color_manual(values = c(
         "Train" = train_color,
         "Test" = test_color
       )) +
-      labs(color = "Set") +
-      theme(legend.position = "none")
-  }
+      labs(color = "Set", title = sprintf("Fold %s", .x)) +
+      theme(plot.title = ggtext::element_textbox(
+        size = 10,
+        color = "black", fill = "#ebebeb", box.color = "black",
[…]

### 2020-10-08 R/autoplot.R
replace {cowplot} by {patchwork} and make visualization of multiple folds more flexibe and eye appealing

@@ -56,11 +56,11 @@
 #' autoplot(resampling, task, fold_id = 1)
 #'
 #' ## Visualize train/test splits of multiple folds
-#' autoplot(resampling, task, fold_id = c(1, 2, 3, 4))
+#' autoplot(resampling, task, fold_id = c(1, 2))
 #'
 #' # list of ggplot2 resamplings
 #' plot_list = autoplot(resampling, task,
-#'   fold_id = c(1, 2, 3, 4), grid = FALSE)
+#'   fold_id = c(1, 2), grid = FALSE)
 #' }
 autoplot.ResamplingSpCVBlock = function( # nolint
   object,
@@ -145,7 +145,7 @@ plot.ResamplingRepeatedSpCVBlock = function(x, ...) {
 #'
 #' autoplot(resampling, task)
 #' autoplot(resampling, task, 1)
-#' autoplot(resampling, task, c(1, 2, 3, 4))
+#' autoplot(resampling, task, c(1, 2))
 #' }
 autoplot.ResamplingSpCVEnv = function( # nolint
   object,
@@ -220,12 +220,13 @@ plot.ResamplingRepeatedSpCVEnv = function(x, ...) {
 #' ##########
 #' \dontrun{
 #' library(mlr3)
+#' task = tsk("ecuador")
 #' resampling = rsmp("spcv_coords")
 #' resampling$instantiate(task)
 #'
 #' autoplot(resampling, task)
 #' autoplot(resampling, task, 1)
-#' autoplot(resampling, task, c(1, 2, 3, 4))
+#' autoplot(resampling, task, c(1, 2))
 #' }
 autoplot.ResamplingSpCVCoords = function( # nolint
   object,
@@ -290,7 +291,7 @@ plot.ResamplingRepeatedSpCVCoords = function(x, ...) {
 
 # SptCV Cluto ------------------------------------------------------------------
 
-#' @title Plot for Spatio-temporal clustering
+#' @title Plot for spatio-temporal clustering
 #'
 #' @param tickformat_date `[character]`\cr
 #'   Date format for z-axis.
@@ -318,10 +319,9 @@ plot.ResamplingRepeatedSpCVCoords = function(x, ...) {
 #' resampling$instantiate(task_st, "Date")
 #'
 #' # plot
-#' # resampling = readRDS("inst/spt-plotting-data.rda")
 #' autoplot(resampling, task_st)
[…]

### 2020-10-08 R/autoplot_spcv_cstf.R

@@ -27,7 +27,7 @@
 #' # plot
 #' autoplot(resampling, task_st)
 #' autoplot(resampling, task_st, 1)
-#' autoplot(resampling, task_st, c(1, 2, 3, 4))
+#' autoplot(resampling, task_st, c(1, 2))
 #' }
 autoplot.ResamplingSptCVCstf = function( # nolint
   object,
@@ -49,7 +49,7 @@ autoplot.ResamplingSptCVCstf = function( # nolint
   resampling = object
   coords = task$coordinates()
   coords$row_id = task$row_ids
-  require_namespaces(c("sf", "cowplot"))
+  require_namespaces(c("sf", "patchwork"))
 
   resampling = assert_autoplot(resampling, fold_id, task)
 
@@ -110,11 +110,11 @@ autoplot.ResamplingSptCVCstf = function( # nolint
     } else {
 
       ### Multiplot of multiple partitions with train and test set
-      plot = list()
-      for (i in fold_id) {
 
-        coords_train = coords[row_id %in% resampling_sub$instance$train[[i]]]
-        coords_test = coords[row_id %in% resampling_sub$instance$test[[i]]]
+      plot = mlr3misc::map(fold_id, function(.x) {
+
+        coords_train = coords[row_id %in% resampling_sub$instance$train[[.x]]]
+        coords_test = coords[row_id %in% resampling_sub$instance$test[[.x]]]
 
         coords_train$indicator = "Train"
         coords_test$indicator = "Test"
@@ -127,52 +127,51 @@ autoplot.ResamplingSptCVCstf = function( # nolint
         table_data$Date = as.Date(table_data$Date)
 
         # create plots for each fold
-        plot[[length(plot) + 1]] = {
-          pl = plotly::plot_ly(table_data,
-            x = ~x, y = ~y, z = ~Date,
-            color = ~indicator, colors = c(
-              "#0072B5", "#E18727"
+        pl = plotly::plot_ly(table_data,
+          x = ~x, y = ~y, z = ~Date,
+          color = ~indicator, colors = c(
+            "#0072B5", "#E18727"
+          ),
+          # this is needed for later when doing 3D subplots
+          scene = paste0("scene", .x),
+          showlegend = ifelse(.x == 1, TRUE, FALSE)
+        )
+        pl = plotly::add_markers(pl, marker = list(size = point_size))
+        layout_args = list(pl,
+          "title" = sprintf("Fold #%s", .x),
+          list(
+            xaxis = list(
[…]

### 2020-10-08 inst/mlr3spatiotemporal_test.R

@@ -153,7 +153,7 @@ rcv = ResamplingSpCVBlock$new()
 rcv$param_set$values = list(folds = 4)
 rcv$instantiate(task)
 
-autoplot(rcv, task, fold_id = c(1, 2, 3, 4))
+autoplot(rcv, task, fold_id = c(1, 2))
 
 # Vizualiation - Multiplot
 

### 2020-10-08 tests/testthat/test-autoplot.R

@@ -12,7 +12,7 @@ test_that("errors are thrown for non-valid argument settings", {
   expect_message(plot(rsp, task))
   expect_error(plot(rsp, task, 30))
   expect_error(plot(rsp, task, c(1, 30)))
-  expect_list(plot(rsp, task, c(1, 2, 3, 4), plot_as_grid = FALSE))
+  expect_list(plot(rsp, task, c(1, 2), plot_as_grid = FALSE))
 })
 
 # cv ---------------------------------------------------------------------------
@@ -25,7 +25,7 @@ test_that("plot() works for 'cv'", {
   # autoplot() is used instead of plot() to prevent side-effect plotting
   p1 = autoplot(plots$rsp, plots$task, crs = 4326)
   p2 = autoplot(plots$rsp, plots$task, 1, crs = 4326)
-  p3 = autoplot(plots$rsp, plots$task, c(1, 2, 3, 4), crs = 4326)
+  p3 = autoplot(plots$rsp, plots$task, c(1, 2), crs = 4326)
 
   expect_true(is.ggplot(p1))
   expect_true(is.ggplot(p2))
@@ -33,7 +33,7 @@ test_that("plot() works for 'cv'", {
 
   vdiffr::expect_doppelganger("CV all test sets", p1)
   vdiffr::expect_doppelganger("CV - Fold 1", p2)
-  vdiffr::expect_doppelganger("CV - Fold 1-4", p3)
+  vdiffr::expect_doppelganger("CV - Fold 1-2", p3)
 })
 
 test_that("plot() works for 'repeated-cv'", {
@@ -45,7 +45,7 @@ test_that("plot() works for 'repeated-cv'", {
   # autoplot() is used instead of plot() to prevent side-effect plotting
   p1 = autoplot(plots$rsp, plots$task, crs = 4326)
   p2 = autoplot(plots$rsp, plots$task, 1, crs = 4326)
-  p3 = autoplot(plots$rsp, plots$task, c(1, 2, 3, 4), crs = 4326)
+  p3 = autoplot(plots$rsp, plots$task, c(1, 2), crs = 4326)
 
   expect_true(is.ggplot(p1))
   expect_true(is.ggplot(p2))
@@ -59,8 +59,8 @@ test_that("plot() works for 'repeated-cv'", {
 
[…]

### 2020-10-08 README.Rmd

@@
 
-plot(resampling, task, crs = 4326)
+autoplot(resampling, task, crs = 4326) +
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) +
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
 ```
 
-### Visualization of a specific fold
+### Visualization of specific folds
 
 ```{r spcv_coords-fold}
-plot(resampling, task, fold_id = 1, crs = 4326)
+autoplot(resampling, task, fold_id = c(1, 2), crs = 4326) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
 ```

### 2020-10-30 doc/spatiotemp-viz.R
add doc/ and Meta/ and vignette index

@@ -1,0 +1,57 @@
+## ---- include = FALSE---------------------------------------------------------
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  comment = "#>"
+)
+
+# don't run on mac
+os = ifelse(!Sys.info()[["sysname"]] == "Darwin", TRUE, FALSE)
+
+# - don't run on CRAN
+is_cran = ifelse(Sys.getenv("NOT_CRAN") != "false", FALSE, TRUE)
+
+# run on ci
+pkgdown = Sys.getenv("IN_PKGDOWN") != ""
+
+## ---- fig.align='center', eval=os&&!is_cran&&pkgdown, fig.width=8, fig.height=9----
+#  library(mlr3)
+#  library(mlr3spatiotempcv)
+#  task_st = tsk("cookfarm")
+#  resampling = rsmp("sptcv_cluto", folds = 5, time_var = "Date")
+#  resampling$instantiate(task_st)
+#
+#  pl = plot(resampling, task_st, c(1, 2, 3, 4),
+#    point_size = 3, axis_label_fontsize = 10)
+#
+#  # Warnings can be ignored
+#  pl_subplot = plotly::subplot(pl)
+#
+#  plotly::layout(pl_subplot,
+#    title = "Individual Folds",
+#    scene = list(
+#      domain = list(x = c(0, 0.5), y = c(0.5, 1)),
+#      aspectmode = "cube",
+#      camera = list(eye = list(z = 2.5))
+#    ),
+#    scene2 = list(
+#      domain = list(x = c(0.5, 1), y = c(0.5, 1)),
+#      aspectmode = "cube",
+#      camera = list(eye = list(z = 2.5))
+#    ),
+#    scene3 = list(
+#      domain = list(x = c(0, 0.5), y = c(0, 0.5)),
+#      aspectmode = "cube",
+#      camera = list(eye = list(z = 2.5))
+#    ),
+#    scene4 = list(
+#      domain = list(x = c(0.5, 1), y = c(0, 0.5)),
+#      aspectmode = "cube",
+#      camera = list(eye = list(z = 2.5))
+#    )
+#  )
+
+## ---- eval=os&&!is_cran&&!pkgdown, echo=FALSE, results='asis'-----------------
+#  cat("(The actual graphic here only rendered on the pkgdown site: please visit https://mlr3spatiotempcv.mlr-org.com.)")
+
+## ---- eval=!os, results='asis', echo=FALSE------------------------------------
[…]

### 2020-10-30 doc/spatiotemp-viz.Rmd
add doc/ and Meta/ and vignette index

---
title: "Spatiotemporal Visualization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatiotemporal Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# don't run on mac
os = ifelse(!Sys.info()[["sysname"]] == "Darwin", TRUE, FALSE)

# - don't run on CRAN
is_cran = ifelse(Sys.getenv("NOT_CRAN") != "false", FALSE, TRUE)

# run on ci
pkgdown = Sys.getenv("IN_PKGDOWN") != ""
```

{mlr3spatiotempcv} makes use of {plotly} to create the 3D plots for visualizing spatiotemporal folds created via the CLUTO algorithm.
Arranging multiple 3D plots in {plotly} is done via [3D subplots](https://plotly.com/r/3d-subplots/).

Unfortunately, {plotly}'s subplot implementation is not dynamic.
This means that multiple "scene" objects need to be specified in `plotly::layout()` to determine the coordinates of the respective subplot.
Depending on the number of chosen folds by the user in `autoplot()`, a different number of scenes with different coordinates needs to be given to align the plots properly.

Hence, manual action is needed to create a properly aligned grid of 3D plots.

Below is an example how to create a 2x2 grid showing four folds as 3D subplots.
It makes use of the returned 3D plotly objects which are returned in a list by `autoplot()`:

```{r, fig.align='center', eval=os&&!is_cran&&pkgdown, fig.width=8, fig.height=9}
library(mlr3)
library(mlr3spatiotempcv)
task_st = tsk("cookfarm")
resampling = rsmp("sptcv_cluto", folds = 5, time_var = "Date")
resampling$instantiate(task_st)

pl = plot(resampling, task_st, c(1, 2, 3, 4),
  point_size = 3, axis_label_fontsize = 10)

# Warnings can be ignored
pl_subplot = plotly::subplot(pl)

plotly::layout(pl_subplot,
  title = "Individual Folds",
  scene = list(
    domain = list(x = c(0, 0.5), y = c(0.5, 1)),
    aspectmode = "cube",
    camera = list(eye = list(z = 2.5))
  ),
  scene2 = list(
    domain = list(x = c(0.5, 1), y = c(0.5, 1)),
    aspectmode = "cube",
    camera = list(eye = list(z = 2.5))
  ),
  scene3 = list(
    domain = list(x = c(0, 0.5), y = c(0, 0.5)),
    aspectmode = "cube",
    camera = list(eye = list(z = 2.5))
  ),
  scene4 = list(
    domain = list(x = c(0.5, 1), y = c(0, 0.5)),
    aspectmode = "cube",
    camera = list(eye = list(z = 2.5))
  )
)
```

```{r, eval=os&&!is_cran&&!pkgdown, echo=FALSE, results='asis'}
cat("(The actual graphic here only rendered on the pkgdown site: please visit https://mlr3spatiotempcv.mlr-org.com.)")
```

```{r, eval=!os, results='asis', echo=FALSE}
cat("(The Cluto method does not work on macOS: please visit https://mlr3spatiotempcv.mlr-org.com to see the rendered plot.)")
```

Subplot titles can unfortunately not created dynamically.
However, there is also a manual workaround via [annotations](https://plotly.com/r/reference/#layout-annotations) show in this [RPubs post](https://rpubs.com/bcd/subplot-titles).

### 2020-10-30 doc/mlr3spatiotempcv.R

@@ -1,0 +1,18 @@
+## ----setup, include=FALSE-----------------------------------------------------
+set.seed(42)
+library("mlr3")
+library("mlr3spatiotempcv")
+
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  comment = "#>"
+)
+
+## -----------------------------------------------------------------------------
+mlr_reflections$task_types
+
+## -----------------------------------------------------------------------------
+mlr_reflections$task_col_roles
+
+## ---- warning=FALSE-----------------------------------------------------------
+as.data.table(mlr_resamplings)

### 2020-11-01 joss/joss.Rmd
add joss paper

---
# Example from https://joss.readthedocs.io/en/latest/submitting.html
title: "mlr3spatiotempcv: Spatiotemporal Resampling Methods for 'mlr3'"
tags:
  - resampling
  - machine learning
  - autocorrelation
  - spatial
  - temporal
authors:
  - name: Patrick Schratz
    orcid: 0000-0003-0748-6624
    affiliation: "1, 2"
  - name: Marc Becker
    orcid: 0000-0002-8115-0400
    affiliation: 2
  - name: Michel Lang
    orcid: 0000-0001-9754-0393
    affiliation: "2, 3"
affiliations:
 - name: Friedrich-Schiller-University Jena
   index: 1
 - name: Ludwig-Maximilians-University Munich
   index: 2
 - name: TU Dortmund University
   index: 3
citation_author: Schratz et. al
date: 01 November 2020
year: 2020
bibliography: paper.bib
# documentclass: joss
output: rticles::joss_article
csl: apa.csl
# journal: JOSS
---

# Summary

The R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is an extension package of the machine learning framework [mlr3](https://mlr3.mlr-org.com) [@mlr3].
It adds support for spatiotemporal resampling methods and respective visualization methods of.
Spatiotemporal resampling methods are needed to account for the existence of different autocorrelation types (spatial autocorrelation (SAC), temporal autocorrelation (TAC) or spatio-temporal autocorrelation (STAC)) in datasets.
These autocorrelation types cause biased performance estimates of varying magnitude in cross-validation when ignored.

At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: (CAST [@cast], blockCV [@blockCV] or sperrorest [@brenning2012]).
There are some resampling methods used in scientific studies for which the source code is either not available or a different programming language has been used [@roberts2017, @pohjankukka2017]

The goal of [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is to gather most available spatiotemporal resampling methods in R and make them available to users through a simplified and generic interface.
The latter is made possible by integrating the package directly into the [mlr3](https://mlr3.mlr-org.com) machine learning framework.
This simplifies the step of integrating package specific syntax into an overarching machine learning pipeline.
The hope is that the availability of such a package encourages researchers to more often apply spatiotemporal resampling methods in scientific case studies.

### Spatiotemporal Autocorrelation in Machine Learning

Data which includes spatial or temporal information requires special treatment in machine learning; similar to cost-sensitive, functional, multilabel, ordinal or survival datasets [@mlr3book].
In contrast to non-spatial/non-temporal data, observations inherit a natural grouping, either in space or time or in both space and time [@legendre1993].
This grouping causes observations to be autocorrelated, either in space (spatial autocorrelation (SAC)), time (temporal autocorrelation (TAC)) or both space and time (spatiotemporal autocorrelation (STAC)).
For simplicity, the acronym STAC is used as a generic term in the following chapter for all the different characteristics introduced above.

*What effects does STAC have in statistical/machine learning?*

The overarching problem is that STAC violates the assumption that the observations in the train and test datasets are independent [@hastie2001].
If this assumption is violated, the reliability of the resulting performance estimates, for example retrieved via cross-validation, is decreased.
The magnitude of this decrease is linked to the magnitude of STAC in the dataset, which cannot be determined easily.

One approach to account for the existence of STAC is to use dedicated resampling methods.
[mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) provides access to the most frequently used spatiotemporal resampling methods.
The following example showcases how a spatial dataset can be used to retrieve a bias-reduced performance estimate of a learner.

The following examples use the [ecuador](https://mlr3spatiotempcv.mlr-org.com/reference/mlr_tasks_ecuador.html) dataset created by [Jannes Muenchow](https://scholar.google.com/citations?user=Slq94Y4AAAAJ&hl=de&authuser=1&oi=ao).
It contains information on the occurrence of landslides (binary) in the Andes of Southern Ecuador.
The landslides were mapped from aerial photos taken in 2000.
The dataset is well suited to serve as an example because it it relatively small and of course due to the spatial nature of the observations.
Please refer to @muenchow2012 for a detailed description of the dataset.

To account for the spatial autocorrelation probably present in the landslide data, we will make use one of the most used spatial partitioning methods, a cluster-based k-means grouping [@brenning2012], (`"spcv_coords"` in [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com)).
This method performs a clustering in 2D space which contrasts with the commonly used random partitioning for non-spatial data.
The grouping has the effect that train and test data are more separated in space as they would be by conducting a random partitioning, thereby reducing the effect of STAC.

By contrast, when using the classical random partitioning approach with spatial data, train and test observations would be located side-by-side across the full study area (a visual example is provided further below).
This leads to a high similarity between train and test sets, resulting in "better" but biased performance estimates in every fold of a CV compared to the spatial CV approach.
However, these low error rates are mainly caused due to the STAC in the observations and the lack of appropriate partitioning methods and not by the power of the fitted model.

## Comparing Spatial Resampling & Non-Spatial Resampling

In the following a spatial and a non-spatial CV will be conducted to showcase the mentioned performance differences.

The performance of a simple classification tree (`"classif.rpart"`) is evaluated on a random partitioning (`"repeated_cv"`) with four folds and two repetitions.
The chosen evaluation measure is "classification error" (`"classif.ce"`).
The only difference in the spatial setting is that `"repeated_spcv_coords"` is chosen instead of `"repeated_cv"`.

### Non-Spatial Cross-Validation

```{r rr-nspcv}
library("mlr3")
library("mlr3spatiotempcv")
set.seed(42)

# be less verbose
lgr::get_logger("bbotk")$set_threshold("warn")
lgr::get_logger("mlr3")$set_threshold("warn")

task = tsk("ecuador")

learner = lrn("classif.rpart", maxdepth = 3, predict_type = "prob")
resampling_nsp = rsmp("repeated_cv", folds = 4, repeats = 2)
rr_nsp = resample(
  task = task, learner = learner,
  resampling = resampling_nsp
)

rr_nsp$aggregate(measures = msr("classif.ce"))
```

### Spatial Cross-Validation

```{r rr-spcv}
task = tsk("ecuador")

learner = lrn("classif.rpart", maxdepth = 3, predict_type = "prob")
resampling_sp = rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
rr_sp = resample(
  task = task, learner = learner,
  resampling = resampling_sp
)

rr_sp$aggregate(measures = msr("classif.ce"))
```

Here, the classification tree learner is around 0.05 percentage points worse when using Spatial Cross-Validation (SpCV) compared to Non-Spatial Cross-Validation (NSpCV).
The magnitude of this difference is variable as it depends on the dataset, the magnitude of STAC and the learner itself.
For algorithms with a higher tendency of overfitting to the training set, the difference between the two methods will be larger.

## Visualization of Spatiotemporal Partitions

Every partitioning method in [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) comes with a generic `plot()` method to visualize the created groups.
In a 2D space this happens via [ggplot2](https://ggplot2.tidyverse.org) [@ggplot2] while for spatiotemporal methods 3D visualizations via [plotly](https://github.com/ropensci/plotly) [@plotly] are created.

```{r vis-spcv, fig.cap="Visualization of spatial partitioning", fig.pos="h!", fig.height=6, fig.width=8}
autoplot(resampling_sp, task, fold_id = c(1:4)) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

Unless specified by the user, the coordinate reference system (CRS) defaults to EPSG code 4326 (WGS84).
This is because a lat/lon based CRS is better suited for plotting purposes than a Mercator (UTM) one.
Note that setting the correct CRS for the given data *during construction* is very important.
Even though EPSG 4326 is a good fallback and often used for visualization purposes, spatial offsets of up to multiple meters may occur if the wrong CRS was passed initially.

This example used an already created task via the sugar function `tsk()`.
In practice however, one needs to create a spatiotemporal task via `TaskClassifST()`/`TaskRegrST()` and set the `crs` argument.

The spatial grouping of the k-means based approach above (Figure 1) contrasts visually very well compared to the NSpCV (random) partitioning (Figure 2):

```{r vis-nspcv, fig.cap="Visualization of random partitioning", fig.pos="h!", fig.height=6, fig.width=8}
autoplot(resampling_nsp, task, fold_id = c(1:4), crs = 4326) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

## Choosing a Resampling Method

While the example used the `"spcv_coords"` method, this does not mean that this method is the best or only method suitable for this task.
Even though this method is quite popular, it was mainly chosen because of the clear visual grouping differences compared to random partitioning.

In fact, most often multiple spatial partitioning methods can be used for a dataset.
It is recommended (required) that users familiarize themselves with each implemented method and decide which method to choose based on the specific characteristics of the dataset.
For almost all methods implemented in [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com), there is a scientific publication describing the strengths and weaknesses of the respective approach (either linked in the help file of [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) or its respective dependency packages).

In the example above, a cross-validation without hyperparameter tuning was shown.
If a nested CV is desired, it is recommended to use the same spatial partitioning method for the inner loop (= tuning level).
See @schratz2019 for more details and chapter 11 of [Geocomputation with R](https://geocompr.robinlovelace.net/spatial-cv.html) [@lovelace2019].

A list of all implemented methods in [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) can be found in the [Getting Started](https://mlr3spatiotempcv.mlr-org.com/articles/mlr3spatiotempcv.html) vignette of the package.

# References

### 2020-11-04 joss/joss.Rmd
wording

@@
 The R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is an extension package of the machine learning framework [mlr3](https://mlr3.mlr-org.com) [@mlr3].
-It adds support for spatiotemporal resampling methods and respective visualization methods of.
+It adds support for spatiotemporal resampling methods and respective visualization methods.
 Spatiotemporal resampling methods are needed to account for the existence of different autocorrelation types (spatial autocorrelation (SAC), temporal autocorrelation (TAC) or spatio-temporal autocorrelation (STAC)) in datasets.
@@
 The landslides were mapped from aerial photos taken in 2000.
-The dataset is well suited to serve as an example because it it relatively small and of course due to the spatial nature of the observations.
+The dataset is well suited to serve as an example because it is relatively small and of course due to the spatial nature of the observations.
 Please refer to @muenchow2012 for a detailed description of the dataset.
 
-To account for the spatial autocorrelation probably present in the landslide data, we will make use one of the most used spatial partitioning methods, a cluster-based k-means grouping [@brenning2012], (`"spcv_coords"` in [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com)).
+To account for the spatial autocorrelation probably present in the landslide data, we will make use one of the most used spatial partitioning methods, a cluster-based $k$-means grouping [@brenning2012], (`"spcv_coords"` in [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com)).
 This method performs a clustering in 2D space which contrasts with the commonly used random partitioning for non-spatial data.
@@
 
-The performance of a simple classification tree (`"classif.rpart"`) is evaluated on a random partitioning (`"repeated_cv"`) with four folds and two repetitions.
+The performance of a simple classification tree (`"classif.rpart"`) is evaluated on a random repeated cross-validation (`"repeated_cv"`) with four folds and two repetitions.
 The chosen evaluation measure is "classification error" (`"classif.ce"`).

### 2020-11-05 joss/joss.Rmd
use ranger learner, put more weight on the bias in non-spatial cv
cc @mllg

@@
 
+Rephrased in other words: the higher performance results in a non-spatial CV scenario should not be relied on or used for reporting.
+The positive bias in these results is substantial and the actual performance of the model is nowhere near this result.
+Results from a SptCV are not unbiased but more close in reflecting the actual performance of the learner on unknown datasets.
+
 ## Comparing Spatial Resampling & Non-Spatial Resampling
@@
 library("mlr3spatiotempcv")
+library("mlr3learners")
 set.seed(42)
@@
 
-learner = lrn("classif.rpart", maxdepth = 3, predict_type = "prob")
+learner = lrn("classif.ranger", predict_type = "prob")
 resampling_nsp = rsmp("repeated_cv", folds = 4, repeats = 2)
@@
 
-rr_nsp$aggregate(measures = msr("classif.ce"))
+rr_nsp$aggregate(measures = msr("classif.auc"))
 ```
@@
 ```{r rr-spcv}
-task = tsk("ecuador")
-
-learner = lrn("classif.rpart", maxdepth = 3, predict_type = "prob")
 resampling_sp = rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
@@
 
-rr_sp$aggregate(measures = msr("classif.ce"))
+rr_sp$aggregate(measures = msr("classif.auc"))
 ```
 
-Here, the classification tree learner is around 0.05 percentage points worse when using Spatial Cross-Validation (SpCV) compared to Non-Spatial Cross-Validation (NSpCV).
+Here, the the random forest learners is around 0.15 worse in AUC when using Spatial Cross-Validation (SpCV) compared to Non-Spatial Cross-Validation (NSpCV).
 The magnitude of this difference is variable as it depends on the dataset, the magnitude of STAC and the learner itself.
-For algorithms with a higher tendency of overfitting to the training set, the difference between the two methods will be larger.
+Algorithms with a higher tendency of overfitting to the training set will have a large spread in such scenarios.
 

### 2020-11-06 joss/joss.Rmd
inc michels suggestions

@@
 By contrast, when using the classical random partitioning approach with spatial data, train and test observations would be located side-by-side across the full study area (a visual example is provided further below).
-This leads to a high similarity between train and test sets, resulting in "better" but biased performance estimates in every fold of a CV compared to the spatial CV approach.
+This leads to a high similarity between train and test sets, resulting in "better" but over-optimistic performance estimates in every fold of a CV compared to the spatial CV approach.
 However, these low error rates are mainly caused due to the STAC in the observations and the lack of appropriate partitioning methods and not by the power of the fitted model.
@@
 
-Here, the the random forest learners is around 0.15 worse in AUC when using Spatial Cross-Validation (SpCV) compared to Non-Spatial Cross-Validation (NSpCV).
+Here, the estimated AUC using Non-Spatial Cross-Validation (NSpCV) is around 0.15 better compared to the Spatial Cross-Validation (SpCV).
+Remember that the better performance score results from over-optimism in the estimation with NSpCV.
+The actual predictive performance on new data is much more realistically reflected by the performance estimated via SpCV.
 The magnitude of this difference is variable as it depends on the dataset, the magnitude of STAC and the learner itself.

### 2020-11-06 joss/joss.Rmd
optimize figure placement, split code in parts

@@
 [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) provides access to the most frequently used spatiotemporal resampling methods.
-The following example showcases how a spatial dataset can be used to retrieve a bias-reduced performance estimate of a learner.
+The following example, which uses the [ecuador](https://mlr3spatiotempcv.mlr-org.com/reference/mlr_tasks_ecuador.html) dataset created by [Jannes Muenchow](https://scholar.google.com/citations?user=Slq94Y4AAAAJ&hl=de&authuser=1&oi=ao), showcases how a spatial dataset can be used to retrieve a bias-reduced performance estimate of a learner.
 
-The following examples use the [ecuador](https://mlr3spatiotempcv.mlr-org.com/reference/mlr_tasks_ecuador.html) dataset created by [Jannes Muenchow](https://scholar.google.com/citations?user=Slq94Y4AAAAJ&hl=de&authuser=1&oi=ao).
-It contains information on the occurrence of landslides (binary) in the Andes of Southern Ecuador.
+This dataset contains information on the occurrence of landslides (binary) in the Andes of Southern Ecuador.
 The landslides were mapped from aerial photos taken in 2000.
@@
 
-In the following a spatial and a non-spatial CV will be conducted to showcase the mentioned performance differences.
+In the following, a spatial and a non-spatial CV will be conducted to showcase the mentioned performance differences.
 
-The performance of a simple classification tree (`"classif.rpart"`) is evaluated on a random repeated cross-validation (`"repeated_cv"`) with four folds and two repetitions.
-The chosen evaluation measure is "classification error" (`"classif.ce"`).
+The performance of a simple classification tree (`"classif.ranger"`) is evaluated on a random repeated cross-validation (`"repeated_cv"`) with four folds and two repetitions.
+The chosen evaluation measure is the "Area Under the ROC Curve" (`"classif.auc"`).
 The only difference in the spatial setting is that `"repeated_spcv_coords"` is chosen instead of `"repeated_cv"`.
 
-### Non-Spatial Cross-Validation
+### Data Preparation
 
-```{r rr-nspcv}
+The following code chunk first loads all necessary R packages and sets a low verbosity for this example.
+A seed is set for deterministic execution.
+
+Next the mlr3 builtin task `"ecuador"` is loaded via `tsk()`, which is a convenience wrapper around the underlying R6 construction method for a spatiotemporal classification task (`mlr3spatiotempcv::TaskClassifST$new()`).
+
+Last, the random forest learner (`"classif.ranger")` is initialized with default hyperparameters and the prediction type set to `"probability"` due to the binary response variable.
+
+```{r setup, echo=FALSE}
+# important for correct figure placement
+# see https://stackoverflow.com/a/49395389/4185785
+knitr::knit_hooks$set(plot = function(x, options) {
+  knitr::hook_plot_tex(x, options)
+})
+```
+
+```{r prepare}
 library("mlr3")
@@
 learner = lrn("classif.ranger", predict_type = "prob")
+```
+
+### Non-Spatial Cross-Validation
+
+```{r rr-nspcv}
 resampling_nsp = rsmp("repeated_cv", folds = 4, repeats = 2)
@@
 
-```{r vis-spcv, fig.cap="Visualization of spatial partitioning", fig.pos="h!", fig.height=6, fig.width=8}
+```{r vis-spcv, fig.cap="Visualization of spatial partitioning", fig.pos="!b", fig.height=6, fig.width=8}
 autoplot(resampling_sp, task, fold_id = c(1:4)) *
@@
 ```
+\
 
@@
 
-```{r vis-nspcv, fig.cap="Visualization of random partitioning", fig.pos="h!", fig.height=6, fig.width=8}
+```{r vis-nspcv, fig.cap="Visualization of random partitioning", fig.pos="!b", fig.height=6, fig.width=8}
 autoplot(resampling_nsp, task, fold_id = c(1:4), crs = 4326) *

### 2020-11-15 joss/joss.Rmd
First edit of joss article

@@
 
-At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: (CAST [@cast], blockCV [@blockCV] or sperrorest [@brenning2012]).
+At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: CAST [@cast], blockCV [@blockCV] or sperrorest [@brenning2012].
 There are some resampling methods used in scientific studies for which the source code is either not available or a different programming language has been used [@roberts2017, @pohjankukka2017]
 
-The goal of [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is to gather most available spatiotemporal resampling methods in R and make them available to users through a simplified and generic interface.
+The goal of [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is to gather the available spatiotemporal resampling methods in R and make them available to users through a simplified and generic interface.
 The latter is made possible by integrating the package directly into the [mlr3](https://mlr3.mlr-org.com) machine learning framework.
@@
 This grouping causes observations to be autocorrelated, either in space (spatial autocorrelation (SAC)), time (temporal autocorrelation (TAC)) or both space and time (spatiotemporal autocorrelation (STAC)).
-For simplicity, the acronym STAC is used as a generic term in the following chapter for all the different characteristics introduced above.
+For simplicity, the acronym STAC is used as a generic term in the following discussion for all the different characteristics introduced above.
 
@@
 The landslides were mapped from aerial photos taken in 2000.
-The dataset is well suited to serve as an example because it is relatively small and of course due to the spatial nature of the observations.
+The dataset is well suited to serve as an example because it is relatively small and, of course, due to the spatial nature of the observations.
 Please refer to @muenchow2012 for a detailed description of the dataset.
 
-To account for the spatial autocorrelation probably present in the landslide data, we will make use one of the most used spatial partitioning methods, a cluster-based $k$-means grouping [@brenning2012], (`"spcv_coords"` in [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com)).
+To account for the spatial autocorrelation probably present in the landslide data, we will make use of one of the most used spatial partitioning methods, a cluster-based $k$-means grouping [@brenning2012], (`"spcv_coords"` in [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com)).
 This method performs a clustering in 2D space which contrasts with the commonly used random partitioning for non-spatial data.
-The grouping has the effect that train and test data are more separated in space as they would be by conducting a random partitioning, thereby reducing the effect of STAC.
+The grouping has the effect that train and test data are more separated in space than they would be by conducting a random partitioning, thereby reducing the effect of STAC.
 
 By contrast, when using the classical random partitioning approach with spatial data, train and test observations would be located side-by-side across the full study area (a visual example is provided further below).
-This leads to a high similarity between train and test sets, resulting in "better" but over-optimistic performance estimates in every fold of a CV compared to the spatial CV approach.
+This leads to a high similarity between train and test sets, resulting in "better" but over-optimistic performance estimates in every fold of a CV compared to the Spatial CV (SpCV) approach.
 However, these low error rates are mainly caused due to the STAC in the observations and the lack of appropriate partitioning methods and not by the power of the fitted model.
@@
 The positive bias in these results is substantial and the actual performance of the model is nowhere near this result.
-Results from a SptCV are not unbiased but more close in reflecting the actual performance of the learner on unknown datasets.
+Results from a SpCV are not unbiased but are closer in reflecting the actual performance of the learner on unknown datasets.
 
@@
 
-Here, the estimated AUC using Non-Spatial Cross-Validation (NSpCV) is around 0.15 better compared to the Spatial Cross-Validation (SpCV).
+Here, the estimated AUC using Non-Spatial Cross-Validation (NSpCV) is around 0.15 better compared to the Spatial Cross-Validation.
 Remember that the better performance score results from over-optimism in the estimation with NSpCV.
@@
 
-This example used an already created task via the sugar function `tsk()`.
+This example used an already created task via the convenience function `tsk()`.
 In practice however, one needs to create a spatiotemporal task via `TaskClassifST()`/`TaskRegrST()` and set the `crs` argument.

### 2020-11-19 joss/joss.Rmd
reference fix

@@
 At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: CAST [@cast], blockCV [@blockCV] or sperrorest [@brenning2012].
-There are some resampling methods used in scientific studies for which the source code is either not available or a different programming language has been used [@roberts2017, @pohjankukka2017]
+There are some resampling methods used in scientific studies for which the source code is either not available or a different programming language has been used [@roberts2017; @pohjankukka2017]
 

### 2021-03-24 R/ResamplingSpCVBlock.R
autoplot(): Various changes (#98)
- add tests
- `autoplot()`: allow adjusting point size via `...`
- optimize spatiotemp-viz vignette
- update autoplot examples and related documentation
- Remove duplicate resources in Tasks "see also" fields

@@ -27,6 +27,12 @@ ResamplingSpCVBlock = R6Class("ResamplingSpCVBlock",
   inherit = mlr3::Resampling,
 
   public = list(
+
+    #' @field blocks `sf | list of sf objects`\cr
+    #'  Polygons (`sf` objects) as returned by \pkg{blockCV} which grouped
+    #'  observations into partitions.
+    blocks = NULL,
+
     #' @description
     #' Create an "Environmental Block" resampling instance.
     #' @param id `character(1)`\cr
@@ -122,6 +128,7 @@ ResamplingSpCVBlock = R6Class("ResamplingSpCVBlock",
 
   private = list(
     .sample = function(ids, coords, crs) {
+
       points = sf::st_as_sf(coords,
         coords = colnames(coords),
         crs = crs)
@@ -140,11 +147,26 @@ ResamplingSpCVBlock = R6Class("ResamplingSpCVBlock",
           progress = FALSE,
           verbose = FALSE))))
 
+      # Warning: In st_point_on_surface.sfc(sf::st_zm(x)) :
+      # st_point_on_surface may not give correct results for
+      # longitude/latitude data
+      blocks_sf = suppressWarnings(sf::st_as_sf(inds$blocks, crs = crs))
+
+      self$blocks = blocks_sf
+
+      # to be able to plot the spatial block later in autoplot(),
+      # we need to return them here
       data.table(
         row_id = ids,
-        fold = inds$foldID # ,
-        # key = "fold"
+        fold = inds$foldID
       )
+      # list(
+      #   resampling = data.table(
+      #     row_id = ids,
+      #     fold = inds$foldID
+      #   ),
+      #   blocks = blocks_sf
+      # )
     },
     # private get funs for train and test which are used by
     # Resampling$.get_set()

### 2021-03-24 R/autoplot.R

@@ -2,7 +2,8 @@
 
 #' @title Visualization Functions for SpCV Block Methods.
 #'
-#' @description Generic S3 `plot()` and `autoplot()` (ggplot2) methods.
+#' @description Generic S3 `plot()` and `autoplot()` (ggplot2) methods to
+#'   visualize mlr3 spatiotemporal resampling objects.
 #'
 #' @import ggplot2
 #'
@@ -20,17 +21,20 @@
 #' @param repeats_id `[numeric]`\cr
 #'   Repetition ID to plot.
 #' @param plot_as_grid `[logical(1)]`\cr
-#' Should a gridded plot using via \CRANpkg{patchwork} be created? If `FALSE`
-#' only a list with all \CRANpkg{ggplot2} resamplings is returned. Only applies
-#' if a numeric vector is passed to argument `fold_id`.
+#'   Should a gridded plot using via \CRANpkg{patchwork} be created? If `FALSE`
+#'   a list with of \CRANpkg{ggplot2} objects is returned.
+#'   Only applies if a numeric vector is passed to argument `fold_id`.
 #' @param train_color `[character(1)]`\cr
 #'   The color to use for the training set observations.
 #' @param test_color `[character(1)]`\cr
 #'   The color to use for the test set observations.
 #' @param crs `[character]`\cr
 #'   EPSG code of the CRS for x and y axes.
-#' @param ... Not used.
-#'
+#' @param show_blocks `[logical(1)]`\cr
+#'   Whether to show an overlay of the spatial blocks polygons.
+#' @param show_labels `[logical(1)]`\cr
+#'   Whether to show an overlay of the spatial block IDs.
+#' @param ... Passed to `geom_sf()`. Helpful for adjusting point sizes and shapes.
 #' @details
 #' By default a plot is returned; if `fold_id` is set, a gridded plot is
 #' created. If `plot_as_grid = FALSE`, a list of plot objects is returned.
@@ -41,6 +45,14 @@
 #' If you want to change the colors, call `<plot> + <color-palette>()`.
 #' @return [ggplot()] or list of ggplot2 objects.
 #' @name autoplot.ResamplingSpCVBlock
+#' @seealso
[…]

### 2021-03-24 tests/testthat/test-1-autoplot.R

@@ -8,9 +8,9 @@ test_that("errors are thrown for non-valid argument settings", {
 
   # these error checks apply to all resampling methods.
   expect_message(plot(rsp, task))
-  expect_error(plot(rsp, task, 30))
-  expect_error(plot(rsp, task, c(1, 30)))
-  expect_list(plot(rsp, task, c(1, 2), plot_as_grid = FALSE))
+  expect_error(plot(rsp, task, 30, crs = 4326))
+  expect_error(plot(rsp, task, c(1, 30), crs = 4326))
+  expect_list(plot(rsp, task, c(1, 2), plot_as_grid = FALSE, crs = 4326))
 })
 
 # cv ---------------------------------------------------------------------------
@@ -166,6 +166,47 @@ test_that("plot() works for 'repeated_spcv_block'", {
   vdiffr::expect_doppelganger("RepSpCVBlock - Fold 1, Rep 2", p5)
 })
 
+test_that("autplot blockCV shows correct blocks", {
+  skip_if_not_installed("vdiffr")
+  task = test_make_blockCV_test_task()
+
+  set.seed(42)
+  rsmp <- rsmp("spcv_block",
+    folds = 5,
+    rows = 3,
+    cols = 4)
+  rsmp$instantiate(task)
+
+  p1 = autoplot(rsmp, task,
+    fold_id = 1, size = 0.7,
+    show_blocks = TRUE, crs = 4326)
+
+  vdiffr::expect_doppelganger(
+    "autoplot show_blocks = TRUE show_labels = TRUE",
+    p1)
+})
+
+test_that("autplot blockCV shows correct blocks for repeated_cv", {
+  skip_if_not_installed("vdiffr")
+  task = test_make_blockCV_test_task()
+
+  set.seed(42)
+  rsmp <- rsmp("repeated_spcv_block",
+    folds = 5,
+    repeats = 2,
+    rows = 3,
+    cols = 4)
+  rsmp$instantiate(task)
+
+  p1 = autoplot(rsmp, task,
+    fold_id = 1, size = 0.7, repeats_id = 2,
+    show_blocks = TRUE, crs = 4326, show_labels = TRUE)
+
+  vdiffr::expect_doppelganger(
+    "autoplot repeated show_blocks = TRUE show_labels = TRUE",
+    p1)
+})
+
 # spcv_env ---------------------------------------------------------------------
 
[…]

### 2021-03-27 paper/paper.Rmd
add jss draft

--- 2020-11-19 joss/joss.Rmd
@@
 ---
-# Example from https://joss.readthedocs.io/en/latest/submitting.html
-title: "mlr3spatiotempcv: Spatiotemporal Resampling Methods for 'mlr3'"
-tags:
-  - resampling
-  - machine learning
-  - autocorrelation
-  - spatial
-  - temporal
-authors:
+documentclass: jss
+author:
   - name: Patrick Schratz
-    orcid: 0000-0003-0748-6624
-    affiliation: "1, 2"
+    affiliation: 'Friedrich-Schiller-University Jena \AND'
+    # affiliaton2: 'Ludwig-Maximilians-University Munich'
+    # use this syntax to add text on several lines
+    address: |
+      | Faculty of Chemistry and Earth Sciences
+      | Department of Geography
+      | GIScience group
+    email: \email{patrick.schratz@uni-jena.de}
   - name: Marc Becker
-    orcid: 0000-0002-8115-0400
-    affiliation: 2
+    affiliation: 'Ludwig-Maximilians-University Munich'
+    address: |
+      Department of Statistics
+      Statistical Learning and Data Science group
+    # To add another line, use \AND at the end of the previous one as above
   - name: Michel Lang
-    orcid: 0000-0001-9754-0393
-    affiliation: "2, 3"
-affiliations:
- - name: Friedrich-Schiller-University Jena
-   index: 1
- - name: Ludwig-Maximilians-University Munich
-   index: 2
- - name: TU Dortmund University
-   index: 3
-citation_author: Schratz et. al
-date: 01 November 2020
-year: 2020
-bibliography: paper.bib
-# documentclass: joss
-output: rticles::joss_article
-csl: apa.csl
-# journal: JOSS
----
-
-# Summary
-
+    affiliation: TU Dortmund University
+    address: |
+      | Faculty of Statistics
+title:
+  formatted: "\\pkg{mlr3spatiotempcv}: Spatiotemporal Resampling Methods for \\pkg{mlr3}"
+  # If you use tex in the formatted title, also supply version without
+  plain:     "mlr3spatiotempcv: Spatiotemporal Resampling Methods for mlr3"
+  # For running headers, if needed
+  # short:     "\\pkg{foo}: A Capitalized Title"
+abstract: >
   The R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is an extension package of the machine learning framework [mlr3](https://mlr3.mlr-org.com) [@mlr3].
@@
   These autocorrelation types cause biased performance estimates of varying magnitude in cross-validation when ignored.
-
-At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: CAST [@cast], blockCV [@blockCV] or sperrorest [@brenning2012].
+  At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: CAST [@cast], blockCV [@blockCV] or sperrorest [@sperrorest].
   There are some resampling methods used in scientific studies for which the source code is either not available or a different programming language has been used [@roberts2017; @pohjankukka2017]
@@
   The hope is that the availability of such a package encourages researchers to more often apply spatiotemporal resampling methods in scientific case studies.
+keywords:
+  # at least one keyword must be supplied
+  formatted: [resampling, machine learning, autocorrelation, spatial, temporal, "\\proglang{R}"]
+  plain:     [resampling, machine learning, autocorrelation, spatial, temporal, R]
+preamble: >
+  \usepackage{amsmath}
+  \usepackage{booktabs}
+  \usepackage{longtable}
+tables: true
+bibliography: paper.bib
+output: rticles::jss_article
+---
 
-### Spatiotemporal Autocorrelation in Machine Learning
+# Introduction
 
@@
 
-*What effects does STAC have in statistical/machine learning?*
+## Spatiotemporal Autocorrelation in Statistical/Machine Learning
 
@@
 
-To account for the spatial autocorrelation probably present in the landslide data, we will make use of one of the most used spatial partitioning methods, a cluster-based $k$-means grouping [@brenning2012], (`"spcv_coords"` in [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com)).
+To account for the spatial autocorrelation probably present in the landslide data, we will make use of one of the most used spatial partitioning methods, a cluster-based $k$-means grouping [@sperrorest], (`"spcv_coords"` in [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com)).
 This method performs a clustering in 2D space which contrasts with the commonly used random partitioning for non-spatial data.
@@
 
-Rephrased in other words: the higher performance results in a non-spatial CV scenario should not be relied on or used for reporting.
+Rephrased: the higher performance results in a non-spatial CV scenario should not be relied on or used for reporting.
 The positive bias in these results is substantial and the actual performance of the model is nowhere near this result.
@@
 
-## Comparing Spatial Resampling & Non-Spatial Resampling
+## Motivation
 
-In the following, a spatial and a non-spatial CV will be conducted to showcase the mentioned performance differences.
+In the last years awareness for spatiotemporal autocorrelation in ML tasks has grown [@pohjankukka2017; @roberts2017].
+Subsequently, scientists started writing code to account for the issue in their studies.
+With R being one of the most used languages with respect to modeling/statistics and academia in general, R packages were developed aiming to simplify the data splitting with respect to methods able to account for spatial autocorrelation [@cast, @blockCV].
+FIXME: adds refs
+This is a typical process for the R community as writing R packages has been made relatively straightforward in recent years thanks to efforts from various people/organizations such as Bioconductor, CRAN, ropensci or the tidyverse [@cran, @ropensci, @tidyverse]
+Also the effort of bundling code into a defined package structure makes it easier for others to re-use it, which is a good effort towards reproducible science.
+
+The downside of this culture is that "island solutions" are being created, i.e. small packages which serve a particular (sometimes very niche) use case.
+Usually the "Do one thing and do it well" UNIX philosophy is widely known and used in practice [@raymond2003art].
+However, the issue with R packages is that not every package is necessarily easily compatiable with other ones due to differing input and returns objects, making it problematic to write a clean and tidy workflow.
+Also many R packages are written by single persons which not necessarly have a complete overview over existing implementations.
+These points may lead to the following possible issues in practice:
+
+- Different syntax.
+- Support for only some predictor types (e.g. numeric and factor features but no ordered factors).
+- Different return values.
+- Different namings for the same method.
+- Unclear longterm maintenance.
+
+These points add substantial overhead and/or reproducibility issues for anyone who wants to make use of the provided implementations within their studies.
+\pkg{mlr3spatiotempcv} aims to provide a solution for some of the points outlined above by wrapping all spatiotemporal resampling methods available in \proglang{R} packages.
+By embedding the package within the \pkg{mlr3} ecosystem and following its unified syntax and scalable approach, \pkg{mlr3spatiotempcv} aims to provide future users longterm support for spatiotemporal partitioning methods in R.
+
+## The mlr3 ecosystem
+
+The \pkg{mlr3} R package and its extensions packages form the \pkg{mlr3} machine learning ecosystem [@mlr3].
+\pkg{mlr3} aims to wrap many algorithms available in R to enforce a consistent interface for such.
+This adds the ability for simplified benchmarking, visualization and further model evaluation regardless of the chosen algorithm.
+
+\pkg{mlr3spatiotempcv} complements this philosophy by aiming to apply the same standards for spatiotemporal resampling methods.
+
+# Comparing Spatial and Non-Spatial Resampling
+
+To introduce the effect of spatial autocorrelation in modeling, a spatial and non-spatial CV using \pkg{mlr3spatiotempcv} will be shown in the following.
+For simplicity, one of the most used methods (spatial k-means clustering after @sperrorest) will be used.
+Individual methods are discussed in FIXME.
 
 The performance of a simple classification tree (`"classif.ranger"`) is evaluated on a random repeated cross-validation (`"repeated_cv"`) with four folds and two repetitions.
-The chosen evaluation measure is the "Area Under the ROC Curve" (`"classif.auc"`).
-The only difference in the spatial setting is that `"repeated_spcv_coords"` is chosen instead of `"repeated_cv"`.
+In reality, more fold (between five and ten) and a higher number of repetitions should be used.
+The chosen evaluation measure is the "Area Under the ROC Curve" (`"classif.auc"`) which is a common evaluation measure for binary responses.
+The CV will be executed twice with the only difference being the use of a spatial resampling object (initialized with `rsmp("repeated_spcv_coords"`) instead of a non-spatial one (`rsmp("repeated_cv"`).
 
-### Data Preparation
+## Data Preparation
 
@@
 
-Next the mlr3 builtin task `"ecuador"` is loaded via `tsk()`, which is a convenience wrapper around the underlying R6 construction method for a spatiotemporal classification task (`mlr3spatiotempcv::TaskClassifST$new()`).
+Next the mlr3 task `"ecuador"`, which is included in \pkg{mlr3spatiotempcv}, is loaded via `tsk()`.
+This function is a convenience wrapper around the underlying R6 construction method for a spatiotemporal classification task (`mlr3spatiotempcv::TaskClassifST$new()`).
 
-Last, the random forest learner (`"classif.ranger")` is initialized with default hyperparameters and the prediction type set to `"probability"` due to the binary response variable.
+Last, the random forest learner (`lrn("classif.ranger")`) is initialized with default hyperparameters and the prediction type set to `"probability"` due to the binary response variable.
+A set of commonly used learners is available in \pkg{mlr3learners} [@mlr3learners], including the random forest implementation after @ranger.
+
+By setting the logging threshold, which can be controlled via the \pkg{lgr} package [@lgr], to `"warn"`, verbosity is reduced to keep the example tidy.
 
@@
 })
+knitr::opts_chunk$set(cache = FALSE)
 ```
 
-```{r prepare}
+```{r prepare, cache=FALSE}
 library("mlr3")
@@
 
-task = tsk("ecuador")
+task <- tsk("ecuador")
 
-learner = lrn("classif.ranger", predict_type = "prob")
+learner <- lrn("classif.ranger", predict_type = "prob")
 ```
@@
 
-```{r rr-nspcv}
-resampling_nsp = rsmp("repeated_cv", folds = 4, repeats = 2)
-rr_nsp = resample(
+```{r rr-nspcv, cache=FALSE}
+resampling_nsp <- rsmp("repeated_cv", folds = 4, repeats = 2)
+rr_nsp <- resample(
   task = task, learner = learner,
@@
 
-### Spatial Cross-Validation
+## Spatial Cross-Validation
 
-```{r rr-spcv}
-resampling_sp = rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
-rr_sp = resample(
+```{r rr-spcv, cache=FALSE}
+resampling_sp <- rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
+rr_sp <- resample(
   task = task, learner = learner,
@@
 
-Here, the estimated AUC using Non-Spatial Cross-Validation (NSpCV) is around 0.15 better compared to the Spatial Cross-Validation.
-Remember that the better performance score results from over-optimism in the estimation with NSpCV.
+Here, the estimated AUC using Non-Spatial Cross-Validation (NSpCV) is around 0.15 better compared to the Spatial Cross-Validation (SpCV).
+The better performance score results from over-optimism in the estimation with NSpCV.
 The actual predictive performance on new data is much more realistically reflected by the performance estimated via SpCV.
@@
 
-## Visualization of Spatiotemporal Partitions
+# Spatiotemporal Partitioning Methods
+
+At the time of writing, the following spatiotemporal resampling methods (FIXME ref table) were available across different R packages.
+This list might increase over time and it is recommended to consult the homepage of \pkg{mlr3spatiotempcv} for an up-to-date list.
+
+A low-level introduction is given for every method included at this point of this article.
+For a discussion with respect to usage see section FIXME.
+
+## Spatial Buffering
+
+The "buffering" method from the \pkg{blockCV} package is conceptually similar to a leave-one-out (LOO) cross-validation approach [@rest2014, @blockCV].
+To reduce the similarity between train and test, a circular buffer around the observations representing the test set is drawn.
+Within this buffer zone, all observations are removed from the training set, thereby reducing the similarity between train and test set and subsequently the influence of SAC.
+
+By default, the method considers all points for train/test creation.
+In an ecological context, this equals to the assumption of having *presence/absence* data for binary response variables.
+This means all non-presence values in the response variable are assumed to be confirmed absence values.
+
+The implementation also comes with support for *presence/background* data for binary responses.
+In this scenario, only the presence values are considered validated.
+All other response observations are assumed to be absent without a proof for this assumption.
+When setting argument `spDataType = "PB"`, the method only considers the *presence* observations in a CV scenario.
+Background data is used for training the model by default unless argument `addBG = FALSE` is set for which only *presence* values not falling within the buffer zone of the respective fold are used for model training.
+
+Due to its similarity to a LOO CV, this method will result in as many train/test evaluations as observations are present.
+In the example below, a buffer zone of 1000 m is used and the first fold shown is plotted.
+
+```{r}
+resampling_buffer <- rsmp("spcv_buffer", theRange = 1000)
+
+autoplot(resampling_buffer, size = 0.7, task = task, fold_id = 1, crs = 4326) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```
+
+<!-- TODO: Discuss issues setting a sensible range value -->
+
+## Spatial Blocking
+
+"Spatial Blocking" aims to create homogeneous blocks across the study area dividing observations into different zones of which each represents a partition.
+The way how these folds are created can be manifold:
+
+1.
+Supplying a numeric value in meters via argument `theRange` will create a quadratic block pattern across the study area following either a random (`selection = "random"`), systematic (`selection = "systematic"`) or checkerboard structure (`selection = "checkerboard"`).
+
+```{r}
+resampling_block_random <- rsmp("spcv_block", range = 500, folds = 5)
+
+autoplot(resampling_block_random,
+  size = 0.8, fold_id = 1,
+  task = task, crs = 4326, show_blocks = TRUE) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```
+
+## Spatial CV
+
+## Environmental Blocking
+
+## Leave-Location-and-Time-Out (LLOT)
+
+## Spatiotemporal Clustering (Cluto)
+
+\begin{table}[h]
+	\centering
+	\caption[t]{Spatiotemporal resampling methods, ordered alphabetically by class name (Notation column).}
+	\begingroup
+	\begin{tabular}{llll}
+		\\
+		Literature                  & Package          & Reference          & Notation                    \\
+		\toprule
+    Spatial Buffering           & \pkg{blockCV}    & \citet{blockCV}    & \texttt{rsmp("spcv\_buffer")} \\
+    Spatial Blocking            & \pkg{blockCV}    & \citet{blockCV}    & \texttt{rsmp("spcv\_block")}  \\
+    Spatial CV                  & \pkg{sperrorest} & \citet{sperrorest} & \texttt{rsmp("spcv\_cv")}     \\
+    Environmental Blocking      & \pkg{blockCV}    & \citet{blockCV}    & \texttt{rsmp("spcv\_env")}    \\
+    \midrule
+    Leave-Location-and-Time-Out & \pkg{CAST}       & \citet{cast}       & \texttt{rsmp("spcv\_cstf")}   \\
+    Spatiotemporal Clustering   & \pkg{skmeans}    & \citet{cluto}      & \texttt{rsmp("spcv\_cluto")}
+	\end{tabular}
+	\endgroup\label{tab:sptcv-methods}
+\end{table}
+
+##
+
+# Visualization of Spatiotemporal Partitions
 
@@
 
-## Choosing a Resampling Method
+# Choosing a Resampling Method
 

### 2021-03-27 paper/paper.tex
add jss draft

\documentclass[
]{jss}

\usepackage[utf8]{inputenc}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\author{
Patrick Schratz\\Friedrich-Schiller-University Jena \AND Marc
Becker\\Ludwig-Maximilians-University Munich \And Michel Lang\\TU
Dortmund University
}
\title{\pkg{mlr3spatiotempcv}: Spatiotemporal Resampling Methods for
\pkg{mlr3}}

\Plainauthor{Patrick Schratz, Marc Becker, Michel Lang}
\Plaintitle{mlr3spatiotempcv: Spatiotemporal Resampling Methods for
mlr3}

\Abstract{
The R package
\href{https://mlr3spatiotempcv.mlr-org.com}{mlr3spatiotempcv} is an
extension package of the machine learning framework
\href{https://mlr3.mlr-org.com}{mlr3} \citep{mlr3}. It adds support for
spatiotemporal resampling methods and respective visualization methods.
Spatiotemporal resampling methods are needed to account for the
existence of different autocorrelation types (spatial autocorrelation
(SAC), temporal autocorrelation (TAC) or spatio-temporal autocorrelation
(STAC)) in datasets. These autocorrelation types cause biased
performance estimates of varying magnitude in cross-validation when
ignored. At the time of writing, various R packages implementing
different spatiotemporal partitioning strategies exist: CAST
\citep{cast}, blockCV \citep{blockCV} or sperrorest \citep{sperrorest}.
There are some resampling methods used in scientific studies for which
the source code is either not available or a different programming
language has been used \citep{roberts2017, pohjankukka2017} The goal of
\href{https://mlr3spatiotempcv.mlr-org.com}{mlr3spatiotempcv} is to
gather the available spatiotemporal resampling methods in R and make
them available to users through a simplified and generic interface. The
latter is made possible by integrating the package directly into the
\href{https://mlr3.mlr-org.com}{mlr3} machine learning framework. This
simplifies the step of integrating package specific syntax into an
overarching machine learning pipeline. The hope is that the availability
of such a package encourages researchers to more often apply
spatiotemporal resampling methods in scientific case studies.
}

\Keywords{resampling, machine
learning, autocorrelation, spatial, temporal, \proglang{R}}
\Plainkeywords{resampling, machine
learning, autocorrelation, spatial, temporal, R}

%% publication information
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{}
%% \Acceptdate{2012-06-04}

\Address{
    Patrick Schratz\\
    Friedrich-Schiller-University Jena\\
    Faculty of Chemistry and Earth Sciences\\
Department of Geography\\
GIScience group\\
  E-mail: \email{patrick.schratz@uni-jena.de}\\

      Marc Becker\\
    Ludwig-Maximilians-University Munich\\
    Department of Statistics Statistical Learning and Data Science
group\\

      Michel Lang\\
    TU Dortmund University\\
    Faculty of Statistics\\

  }

% Pandoc citation processing

% Pandoc header

\usepackage{amsmath} \usepackage{booktabs} \usepackage{longtable}

\begin{document}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Data which includes spatial or temporal information requires special
treatment in machine learning; similar to cost-sensitive, functional,
multilabel, ordinal or survival datasets \citep{mlr3book}. In contrast
to non-spatial/non-temporal data, observations inherit a natural
grouping, either in space or time or in both space and time
\citep{legendre1993}. This grouping causes observations to be
autocorrelated, either in space (spatial autocorrelation (SAC)), time
(temporal autocorrelation (TAC)) or both space and time (spatiotemporal
autocorrelation (STAC)). For simplicity, the acronym STAC is used as a
generic term in the following discussion for all the different
characteristics introduced above.

\hypertarget{spatiotemporal-autocorrelation-in-statisticalmachine-learning}{%
\subsection{Spatiotemporal Autocorrelation in Statistical/Machine
Learning}\label{spatiotemporal-autocorrelation-in-statisticalmachine-learning}}

The overarching problem is that STAC violates the assumption that the
observations in the train and test datasets are independent
\citep{hastie2001}. If this assumption is violated, the reliability of
the resulting performance estimates, for example retrieved via
cross-validation, is decreased. The magnitude of this decrease is linked
to the magnitude of STAC in the dataset, which cannot be determined
easily.

One approach to account for the existence of STAC is to use dedicated
resampling methods.
\href{https://mlr3spatiotempcv.mlr-org.com}{mlr3spatiotempcv} provides
access to the most frequently used spatiotemporal resampling methods.
The following example, which uses the
\href{https://mlr3spatiotempcv.mlr-org.com/reference/mlr_tasks_ecuador.html}{ecuador}
dataset created by
\href{https://scholar.google.com/citations?user=Slq94Y4AAAAJ\&hl=de\&authuser=1\&oi=ao}{Jannes
Muenchow}, showcases how a spatial dataset can be used to retrieve a
bias-reduced performance estimate of a learner.

This dataset contains information on the occurrence of landslides
(binary) in the Andes of Southern Ecuador. The landslides were mapped
from aerial photos taken in 2000. The dataset is well suited to serve as
an example because it is relatively small and, of course, due to the
spatial nature of the observations. Please refer to \citet{muenchow2012}
for a detailed description of the dataset.

To account for the spatial autocorrelation probably present in the
landslide data, we will make use of one of the most used spatial
partitioning methods, a cluster-based \(k\)-means grouping
\citep{sperrorest}, (\texttt{"spcv\_coords"} in
\href{https://mlr3spatiotempcv.mlr-org.com}{mlr3spatiotempcv}). This
method performs a clustering in 2D space which contrasts with the
commonly used random partitioning for non-spatial data. The grouping has
the effect that train and test data are more separated in space than
they would be by conducting a random partitioning, thereby reducing the
effect of STAC.

By contrast, when using the classical random partitioning approach with
spatial data, train and test observations would be located side-by-side
across the full study area (a visual example is provided further below).
This leads to a high similarity between train and test sets, resulting
in ``better'' but over-optimistic performance estimates in every fold of
a CV compared to the Spatial CV (SpCV) approach. However, these low
error rates are mainly caused due to the STAC in the observations and
the lack of appropriate partitioning methods and not by the power of the
fitted model.

Rephrased: the higher performance results in a non-spatial CV scenario
should not be relied on or used for reporting. The positive bias in
these results is substantial and the actual performance of the model is
nowhere near this result. Results from a SpCV are not unbiased but are
closer in reflecting the actual performance of the learner on unknown
datasets.

\hypertarget{motivation}{%
\subsection{Motivation}\label{motivation}}

In the last years awareness for spatiotemporal autocorrelation in ML
tasks has grown \citep{pohjankukka2017, roberts2017}. Subsequently,
scientists started writing code to account for the issue in their
studies. With R being one of the most used languages with respect to
modeling/statistics and academia in general, R packages were developed
aiming to simplify the data splitting with respect to methods able to
account for spatial autocorrelation \citep[\citet{blockCV}]{cast}.
FIXME: adds refs This is a typical process for the R community as
writing R packages has been made relatively straightforward in recent
years thanks to efforts from various people/organizations such as
Bioconductor, CRAN, ropensci or the tidyverse \citep[\citet{ropensci},
\citet{tidyverse}]{cran} Also the effort of bundling code into a defined
package structure makes it easier for others to re-use it, which is a
good effort towards reproducible science.

The downside of this culture is that ``island solutions'' are being
created, i.e.~small packages which serve a particular (sometimes very
niche) use case. Usually the ``Do one thing and do it well'' UNIX
philosophy is widely known and used in practice \citep{raymond2003art}.
However, the issue with R packages is that not every package is
necessarily easily compatiable with other ones due to differing input
and returns objects, making it problematic to write a clean and tidy
workflow. Also many R packages are written by single persons which not
necessarly have a complete overview over existing implementations. These
points may lead to the following possible issues in practice:

\begin{itemize}
\tightlist
\item
  Different syntax.
\item
  Support for only some predictor types (e.g.~numeric and factor
  features but no ordered factors).
\item
  Different return values.
\item
  Different namings for the same method.
\item
  Unclear longterm maintenance.
\end{itemize}

These points add substantial overhead and/or reproducibility issues for
anyone who wants to make use of the provided implementations within
their studies. \pkg{mlr3spatiotempcv} aims to provide a solution for
some of the points outlined above by wrapping all spatiotemporal
resampling methods available in \proglang{R} packages. By embedding the
package within the \pkg{mlr3} ecosystem and following its unified syntax
and scalable approach, \pkg{mlr3spatiotempcv} aims to provide future
users longterm support for spatiotemporal partitioning methods in R.

\hypertarget{the-mlr3-ecosystem}{%
\subsection{The mlr3 ecosystem}\label{the-mlr3-ecosystem}}

The \pkg{mlr3} R package and its extensions packages form the \pkg{mlr3}
machine learning ecosystem \citep{mlr3}. \pkg{mlr3} aims to wrap many
algorithms available in R to enforce a consistent interface for such.
This adds the ability for simplified benchmarking, visualization and
further model evaluation regardless of the chosen algorithm.

\pkg{mlr3spatiotempcv} complements this philosophy by aiming to apply
the same standards for spatiotemporal resampling methods.

\hypertarget{comparing-spatial-and-non-spatial-resampling}{%
\section{Comparing Spatial and Non-Spatial
Resampling}\label{comparing-spatial-and-non-spatial-resampling}}

To introduce the effect of spatial autocorrelation in modeling, a
spatial and non-spatial CV using \pkg{mlr3spatiotempcv} will be shown in
the following. For simplicity, one of the most used methods (spatial
k-means clustering after \citet{sperrorest}) will be used. Individual
methods are discussed in FIXME.

The performance of a simple classification tree
(\texttt{"classif.ranger"}) is evaluated on a random repeated
cross-validation (\texttt{"repeated\_cv"}) with four folds and two
repetitions. In reality, more fold (between five and ten) and a higher
number of repetitions should be used. The chosen evaluation measure is
the ``Area Under the ROC Curve'' (\texttt{"classif.auc"}) which is a
common evaluation measure for binary responses. The CV will be executed
twice with the only difference being the use of a spatial resampling
object (initialized with \texttt{rsmp("repeated\_spcv\_coords"}) instead
of a non-spatial one (\texttt{rsmp("repeated\_cv"}).

\hypertarget{data-preparation}{%
\subsection{Data Preparation}\label{data-preparation}}

The following code chunk first loads all necessary R packages and sets a
low verbosity for this example. A seed is set for deterministic
execution.

Next the mlr3 task \texttt{"ecuador"}, which is included in
\pkg{mlr3spatiotempcv}, is loaded via \texttt{tsk()}. This function is a
convenience wrapper around the underlying R6 construction method for a
spatiotemporal classification task
(\texttt{mlr3spatiotempcv::TaskClassifST\$new()}).

Last, the random forest learner (\texttt{lrn("classif.ranger")}) is
initialized with default hyperparameters and the prediction type set to
\texttt{"probability"} due to the binary response variable. A set of
commonly used learners is available in \pkg{mlr3learners}
\citep{mlr3learners}, including the random forest implementation after
\citet{ranger}.

By setting the logging threshold, which can be controlled via the
\pkg{lgr} package \citep{lgr}, to \texttt{"warn"}, verbosity is reduced
to keep the example tidy.

\begin{CodeChunk}
\begin{CodeInput}
R> library("mlr3")
R> library("mlr3spatiotempcv")
R> library("mlr3learners")
R> set.seed(42)
R> 
R> # be less verbose
R> lgr::get_logger("bbotk")$set_threshold("warn")
R> lgr::get_logger("mlr3")$set_threshold("warn")
R> 
R> task <- tsk("ecuador")
R> 
R> learner <- lrn("classif.ranger", predict_type = "prob")
\end{CodeInput}
\end{CodeChunk}

\hypertarget{non-spatial-cross-validation}{%
\subsubsection{Non-Spatial
Cross-Validation}\label{non-spatial-cross-validation}}

\begin{CodeChunk}
\begin{CodeInput}
R> resampling_nsp <- rsmp("repeated_cv", folds = 4, repeats = 2)
R> rr_nsp <- resample(
+   task = task, learner = learner,
+   resampling = resampling_nsp
+ )
R> 
R> rr_nsp$aggregate(measures = msr("classif.auc"))
\end{CodeInput}
\begin{CodeOutput}
classif.auc 
  0.7600664 
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{spatial-cross-validation}{%
\subsection{Spatial Cross-Validation}\label{spatial-cross-validation}}

\begin{CodeChunk}
\begin{CodeInput}
R> resampling_sp <- rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
R> rr_sp <- resample(
+   task = task, learner = learner,
+   resampling = resampling_sp
+ )
R> 
R> rr_sp$aggregate(measures = msr("classif.auc"))
\end{CodeInput}
\begin{CodeOutput}
classif.auc 
  0.6100402 
\end{CodeOutput}
\end{CodeChunk}

Here, the estimated AUC using Non-Spatial Cross-Validation (NSpCV) is
around 0.15 better compared to the Spatial Cross-Validation (SpCV). The
better performance score results from over-optimism in the estimation
with NSpCV. The actual predictive performance on new data is much more
realistically reflected by the performance estimated via SpCV. The
magnitude of this difference is variable as it depends on the dataset,
the magnitude of STAC and the learner itself. Algorithms with a higher
tendency of overfitting to the training set will have a large spread in
such scenarios.

\hypertarget{spatiotemporal-partitioning-methods}{%
\section{Spatiotemporal Partitioning
Methods}\label{spatiotemporal-partitioning-methods}}

At the time of writing, the following spatiotemporal resampling methods
(FIXME ref table) were available across different R packages. This list
might increase over time and it is recommended to consult the homepage
of \pkg{mlr3spatiotempcv} for an up-to-date list.

A low-level introduction is given for every method included at this
point of this article. For a discussion with respect to usage see
section FIXME.

\hypertarget{spatial-buffering}{%
\subsection{Spatial Buffering}\label{spatial-buffering}}

The ``buffering'' method from the \pkg{blockCV} package is conceptually
similar to a leave-one-out (LOO) cross-validation approach
\citep[\citet{blockCV}]{rest2014}. To reduce the similarity between
train and test, a circular buffer around the observations representing
the test set is drawn. Within this buffer zone, all observations are
removed from the training set, thereby reducing the similarity between
train and test set and subsequently the influence of SAC.

By default, the method considers all points for train/test creation. In
an ecological context, this equals to the assumption of having
\emph{presence/absence} data for binary response variables. This means
all non-presence values in the response variable are assumed to be
confirmed absence values.

The implementation also comes with support for
\emph{presence/background} data for binary responses. In this scenario,
only the presence values are considered validated. All other response
observations are assumed to be absent without a proof for this
assumption. When setting argument \texttt{spDataType\ =\ "PB"}, the
method only considers the \emph{presence} observations in a CV scenario.
Background data is used for training the model by default unless
argument \texttt{addBG\ =\ FALSE} is set for which only \emph{presence}
values not falling within the buffer zone of the respective fold are
used for model training.

Due to its similarity to a LOO CV, this method will result in as many
train/test evaluations as observations are present. In the example
below, a buffer zone of 1000 m is used and the first fold shown is
plotted.

\begin{CodeChunk}
\begin{CodeInput}
R> resampling_buffer <- rsmp("spcv_buffer", theRange = 1000)
R> 
R> autoplot(resampling_buffer, size = 0.7, task = task, fold_id = 1, crs = 4326) *
+   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
\end{CodeInput}

\begin{center}\includegraphics{paper_files/figure-latex/unnamed-chunk-1-1} \end{center}

\end{CodeChunk}

\hypertarget{spatial-blocking}{%
\subsection{Spatial Blocking}\label{spatial-blocking}}

``Spatial Blocking'' aims to create homogeneous blocks across the study
area dividing observations into different zones of which each represents
a partition. The way how these folds are created can be manifold:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Supplying a numeric value in meters via argument \texttt{theRange}
  will create a quadratic block pattern across the study area following
  either a random (\texttt{selection\ =\ "random"}), systematic
  (\texttt{selection\ =\ "systematic"}) or checkerboard structure
  (\texttt{selection\ =\ "checkerboard"}).
\end{enumerate}

\begin{CodeChunk}
\begin{CodeInput}
R> resampling_block_random <- rsmp("spcv_block", range = 500, folds = 5)
R> 
R> autoplot(resampling_block_random, size = 0.8, fold_id = 1,
+          task = task, crs = 4326, show_blocks = TRUE) *
+   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
\end{CodeInput}
\begin{CodeOutput}
Warning in st_point_on_surface.sfc(sf::st_zm(x)): st_point_on_surface may not
give correct results for longitude/latitude data
\end{CodeOutput}

\begin{center}\includegraphics{paper_files/figure-latex/unnamed-chunk-2-1} \end{center}

\end{CodeChunk}

\hypertarget{spatial-cv}{%
\subsection{Spatial CV}\label{spatial-cv}}

\hypertarget{environmental-blocking}{%
\subsection{Environmental Blocking}\label{environmental-blocking}}

\hypertarget{leave-location-and-time-out-llot}{%
\subsection{Leave-Location-and-Time-Out
(LLOT)}\label{leave-location-and-time-out-llot}}

\hypertarget{spatiotemporal-clustering-cluto}{%
\subsection{Spatiotemporal Clustering
(Cluto)}\label{spatiotemporal-clustering-cluto}}

\begin{table}[h]
    \centering
    \caption[t]{Spatiotemporal resampling methods, ordered alphabetically by class name (Notation column).}
    \begingroup
    \begin{tabular}{llll}
        \\
        Literature                  & Package          & Reference          & Notation                    \\
        \toprule
    Spatial Buffering           & \pkg{blockCV}    & \citet{blockCV}    & \texttt{rsmp("spcv\_buffer")} \\
    Spatial Blocking            & \pkg{blockCV}    & \citet{blockCV}    & \texttt{rsmp("spcv\_block")}  \\
    Spatial CV                  & \pkg{sperrorest} & \citet{sperrorest} & \texttt{rsmp("spcv\_cv")}     \\
    Environmental Blocking      & \pkg{blockCV}    & \citet{blockCV}    & \texttt{rsmp("spcv\_env")}    \\
    \midrule
    Leave-Location-and-Time-Out & \pkg{CAST}       & \citet{cast}       & \texttt{rsmp("spcv\_cstf")}   \\
    Spatiotemporal Clustering   & \pkg{skmeans}    & \citet{cluto}      & \texttt{rsmp("spcv\_cluto")}
    \end{tabular}
    \endgroup\label{tab:sptcv-methods}
\end{table}

\hypertarget{section}{%
\subsection{}\label{section}}

\hypertarget{visualization-of-spatiotemporal-partitions}{%
\section{Visualization of Spatiotemporal
Partitions}\label{visualization-of-spatiotemporal-partitions}}

Every partitioning method in
\href{https://mlr3spatiotempcv.mlr-org.com}{mlr3spatiotempcv} comes with
a generic \texttt{plot()} method to visualize the created groups. In a
2D space this happens via \href{https://ggplot2.tidyverse.org}{ggplot2}
\citep{ggplot2} while for spatiotemporal methods 3D visualizations via
\href{https://github.com/ropensci/plotly}{plotly} \citep{plotly} are
created.

\begin{CodeChunk}
\begin{CodeInput}
R> autoplot(resampling_sp, task, fold_id = c(1:4)) *
+   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
\end{CodeInput}
\begin{CodeOutput}
CRS not set, transforming to WGS84 (EPSG: 4326).
CRS not set, transforming to WGS84 (EPSG: 4326).
CRS not set, transforming to WGS84 (EPSG: 4326).
CRS not set, transforming to WGS84 (EPSG: 4326).
\end{CodeOutput}
\begin{figure}[!b]

{\centering \includegraphics{paper_files/figure-latex/vis-spcv-1}

}

\caption[Visualization of spatial partitioning]{Visualization of spatial partitioning}\label{fig:vis-spcv}
\end{figure}
\end{CodeChunk}

\hfill\break

Unless specified by the user, the coordinate reference system (CRS)
defaults to EPSG code 4326 (WGS84). This is because a lat/lon based CRS
is better suited for plotting purposes than a Mercator (UTM) one. Note
that setting the correct CRS for the given data \emph{during
construction} is very important. Even though EPSG 4326 is a good
fallback and often used for visualization purposes, spatial offsets of
up to multiple meters may occur if the wrong CRS was passed initially.

This example used an already created task via the convenience function
\texttt{tsk()}. In practice however, one needs to create a
spatiotemporal task via \texttt{TaskClassifST()}/\texttt{TaskRegrST()}
and set the \texttt{crs} argument.

The spatial grouping of the k-means based approach above (Figure 1)
contrasts visually very well compared to the NSpCV (random) partitioning
(Figure 2):

\begin{CodeChunk}
\begin{CodeInput}
R> autoplot(resampling_nsp, task, fold_id = c(1:4), crs = 4326) *
+   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
\end{CodeInput}
\begin{figure}[!b]

{\centering \includegraphics{paper_files/figure-latex/vis-nspcv-1}

}

\caption[Visualization of random partitioning]{Visualization of random partitioning}\label{fig:vis-nspcv}
\end{figure}
\end{CodeChunk}

\hypertarget{choosing-a-resampling-method}{%
\section{Choosing a Resampling
Method}\label{choosing-a-resampling-method}}

While the example used the \texttt{"spcv\_coords"} method, this does not
mean that this method is the best or only method suitable for this task.
Even though this method is quite popular, it was mainly chosen because
of the clear visual grouping differences compared to random
partitioning.

In fact, most often multiple spatial partitioning methods can be used
for a dataset. It is recommended (required) that users familiarize
themselves with each implemented method and decide which method to
choose based on the specific characteristics of the dataset. For almost
all methods implemented in
\href{https://mlr3spatiotempcv.mlr-org.com}{mlr3spatiotempcv}, there is
a scientific publication describing the strengths and weaknesses of the
respective approach (either linked in the help file of
\href{https://mlr3spatiotempcv.mlr-org.com}{mlr3spatiotempcv} or its
respective dependency packages).

In the example above, a cross-validation without hyperparameter tuning
was shown. If a nested CV is desired, it is recommended to use the same
spatial partitioning method for the inner loop (= tuning level). See
\citet{schratz2019} for more details and chapter 11 of
\href{https://geocompr.robinlovelace.net/spatial-cv.html}{Geocomputation
with R} \citep{lovelace2019}.

A list of all implemented methods in
\href{https://mlr3spatiotempcv.mlr-org.com}{mlr3spatiotempcv} can be
found in the
\href{https://mlr3spatiotempcv.mlr-org.com/articles/mlr3spatiotempcv.html}{Getting
Started} vignette of the package.

\renewcommand\refname{References}
\bibliography{paper.bib}

\end{document}

### 2021-03-28 paper/paper.Rmd
add more method content

@@
 ```{r rr-nspcv, cache=FALSE}
-resampling_nsp <- rsmp("repeated_cv", folds = 4, repeats = 2)
+resampling_nsp <- rsmp("repeated_cv", folds = 4, repeats = 2)$instantiate(task)
 rr_nsp <- resample(
@@
 ```{r rr-spcv, cache=FALSE}
-resampling_sp <- rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
+resampling_sp <- rsmp("repeated_spcv_coords", folds = 4, repeats = 2)$instantiate(task)
 rr_sp <- resample(
@@
 ```{r}
-resampling_buffer <- rsmp("spcv_buffer", theRange = 1000)
+resampling_buffer <- rsmp("spcv_buffer", theRange = 1000)$instantiate(task)
 
@@
 
-"Spatial Blocking" aims to create homogeneous blocks across the study area dividing observations into different zones of which each represents a partition.
-The way how these folds are created can be manifold:
+"Spatial Blocking" (\pkg{blockCV} package) creates homogeneous blocks across the study area splitting observations into different zones [@bahn2012a, @wenger2012a].
+Then, a number of zones is combined randomly following desired the number of folds.
+For example, in a five-fold scenario, the method tries to combines zones to end up with partitions containing around 20\% of all observations.
 
-1.
-Supplying a numeric value in meters via argument `theRange` will create a quadratic block pattern across the study area following either a random (`selection = "random"`), systematic (`selection = "systematic"`) or checkerboard structure (`selection = "checkerboard"`).
+The way how the blocks are created can be manifold:
+
+1. Supplying a numeric value in meters via argument `theRange` (named just `range` in {mlr3spatiotempcv}) will create a quadratic block pattern across the study area.
+  The aggregation of zones can either be random (default) (`selection = "random"`), systematic (`selection = "systematic"`) or checkerboard structure (`selection = "checkerboard"`).
+
+`selection = "random"`
 
 ```{r}
-resampling_block_random <- rsmp("spcv_block", range = 500, folds = 5)
+resampling_block_random <- rsmp("spcv_block", range = 1000, folds = 5)
 
 autoplot(resampling_block_random,
-  size = 0.8, fold_id = 1,
-  task = task, crs = 4326, show_blocks = TRUE) *
+  size = 0.8, fold_id = 1, task = task, crs = 4326,
+  show_blocks = TRUE, show_labels = TRUE) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```
+
+`selection = "systematic"`: Here, aggregation follows the linear count from $1:n_{folds} from top to bottom.
+
+```{r}
+resampling_block_systematic <- rsmp("spcv_block",
+  range = 1000, folds = 5,
+  selection = "systematic")
+
+autoplot(resampling_block_systematic,
+  size = 0.8, fold_id = 1, task = task, crs = 4326,
+  show_blocks = TRUE, show_labels = TRUE) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```
+
+`selection = "checkerboard"`: This option ignores paramter `range` and always creates two partitions following a "1,2,1,2" pattern.
+
+```{r}
+resampling_block_checkerboard <- rsmp("spcv_block",
+  range = 1000, folds = 5,
+  selection = "checkerboard")
+
+autoplot(resampling_block_checkerboard,
+  size = 0.8, fold_id = 1, task = task, crs = 4326,
+  show_blocks = TRUE, show_labels = TRUE) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```
+
+## Environmental Blocking
+
+The last method from the \pkg{blockCV} package, "environmental blocking", makes use of *k-means* clustering [@hartigan1979a] in a possibly multivariate space.
+The user can specify a variable via argument `feature` which levels will be used to create clusters.
+Hereby, k-means will use Euclidean distance for cluster creation.
+Hence, the selected input variables should be of type numeric.
+To avoid a potential bias introduced by features with high variance when selecting multiple features, all features are standardized by default.
+
+The following example clusters by feature "distance to forest".
+
+```{r}
+resampling_env <- rsmp("spcv_env",
+  features = "distdeforest", folds = 5)
+
+autoplot(resampling_env,
+  size = 0.8, fold_id = 1, task = task, crs = 4326,
+  show_blocks = TRUE, show_labels = TRUE) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```
+
+It is also possible to cluster by multiple features:
+
+```{r}
+resampling_env_multi <- rsmp("spcv_env",
+  features = c("distdeforest", "slope"), folds = 5)
+
+autoplot(resampling_env_multi,
+  size = 0.8, fold_id = 1, task = task, crs = 4326,
+  show_blocks = TRUE, show_labels = TRUE) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
@@
 
-## Environmental Blocking
+The "Spatial CV" method after [@sperrorest] and implemented in package \pkg{sperrorest} is similar to method "environmental blocking" in that it also uses *k-means* clustering for fold creation.
+However, it does not make use of arbitrary features but uses the coordinates of all observations to create clusters in the spatial domain.
+This is what lead to the naming of `spcv_coords` in \pkg{mlr3spatiotempcv}.
+
+```{r}
+resampling_coords <- rsmp("spcv_coords", folds = 5)$instantiate(task)
+
+autoplot(resampling_coords,
+  size = 0.8, fold_id = 1, task = task, crs = 4326) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```
 
 ## Leave-Location-and-Time-Out (LLOT)
+
+Leave-Location-and-Time-Out from packge \pkg{CAST} is a spatiotemporal resampling method.
+It requires both a spatial and a temporal variable which are used to to separate observations in a multi-dimensional space.
+
+In this example the `cookfarm` dataset is used because it contains a temporal variable ("Date") which is not present in dataset `ecuador`.
+
+The visualization is done in a dynamic way via \pkg{plotly} to visualize both temporal and spatial dimensions.
+Because a dynamic image cannot be included in this manuscript, a static version generated by setting `static_image = TRUE` was used.
+
+```{r}
+task_spt = tsk("cookfarm")
+resampling_cstf <- rsmp("sptcv_cstf",
+  folds = 5,
+  time_var = "Date", space_var = "SOURCEID")
+
+autoplot(resampling_cstf,
+  fold_id = 1, task = task_spt, crs = 4326, static_image = TRUE,
+  file = "plot.pdf") *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```
+
+```{r, echo=FALSE}
+knitr::include_graphics("plot.pdf", dpi = 100)
+```
 

### 2021-03-28 paper/paper.tex
add more method content

@@
 \begin{CodeInput}
-R> resampling_nsp <- rsmp("repeated_cv", folds = 4, repeats = 2)
+R> resampling_nsp <- rsmp("repeated_cv", folds = 4, repeats = 2)$instantiate(task)
 R> rr_nsp <- resample(
@@
 \begin{CodeInput}
-R> resampling_sp <- rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
+R> resampling_sp <- rsmp("repeated_spcv_coords", folds = 4, repeats = 2)$instantiate(task)
 R> rr_sp <- resample(
@@
 \begin{CodeInput}
-R> resampling_buffer <- rsmp("spcv_buffer", theRange = 1000)
+R> resampling_buffer <- rsmp("spcv_buffer", theRange = 1000)$instantiate(task)
 R> 
@@
 
-``Spatial Blocking'' aims to create homogeneous blocks across the study
-area dividing observations into different zones of which each represents
-a partition. The way how these folds are created can be manifold:
+``Spatial Blocking'' (\pkg{blockCV} package) creates homogeneous blocks
+across the study area splitting observations into different zones
+\citep[\citet{wenger2012a}]{bahn2012a}. Then, a number of zones is
+combined randomly following desired the number of folds. For example, in
+a five-fold scenario, the method tries to combines zones to end up with
+partitions containing around 20\% of all observations.
+
+The way how the blocks are created can be manifold:
 
@@
   Supplying a numeric value in meters via argument \texttt{theRange}
-  will create a quadratic block pattern across the study area following
-  either a random (\texttt{selection\ =\ "random"}), systematic
+  (named just \texttt{range} in \{mlr3spatiotempcv\}) will create a
+  quadratic block pattern across the study area. The aggregation of
+  zones can either be random (default)
+  (\texttt{selection\ =\ "random"}), systematic
   (\texttt{selection\ =\ "systematic"}) or checkerboard structure
@@
 
+\texttt{selection\ =\ "random"}
+
 \begin{CodeChunk}
 \begin{CodeInput}
-R> resampling_block_random <- rsmp("spcv_block", range = 500, folds = 5)
+R> resampling_block_random <- rsmp("spcv_block", range = 1000, folds = 5)
 R> 
-R> autoplot(resampling_block_random, size = 0.8, fold_id = 1,
-+          task = task, crs = 4326, show_blocks = TRUE) *
+R> autoplot(resampling_block_random,
++   size = 0.8, fold_id = 1, task = task, crs = 4326,
++   show_blocks = TRUE, show_labels = TRUE) *
 +   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
@@
 \end{CodeInput}
-\begin{CodeOutput}
-Warning in st_point_on_surface.sfc(sf::st_zm(x)): st_point_on_surface may not
-give correct results for longitude/latitude data
-\end{CodeOutput}
 
 \begin{center}\includegraphics{paper_files/figure-latex/unnamed-chunk-2-1} \end{center}
+
+\end{CodeChunk}
+
+\texttt{selection\ =\ "systematic"}: Here, aggregation follows the
+linear count from \$1:n\_\{folds\} from top to bottom.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> resampling_block_systematic <- rsmp("spcv_block",
++   range = 1000, folds = 5,
++   selection = "systematic")
+R> 
+R> autoplot(resampling_block_systematic,
++   size = 0.8, fold_id = 1, task = task, crs = 4326,
++   show_blocks = TRUE, show_labels = TRUE) *
++   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
++   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+\end{CodeInput}
+
+\begin{center}\includegraphics{paper_files/figure-latex/unnamed-chunk-3-1} \end{center}
+
+\end{CodeChunk}
+
+\texttt{selection\ =\ "checkerboard"}: This option ignores paramter
+\texttt{range} and always creates two partitions following a ``1,2,1,2''
+pattern.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> resampling_block_checkerboard <- rsmp("spcv_block",
++   range = 1000, folds = 5,
++   selection = "checkerboard")
+R> 
+R> autoplot(resampling_block_checkerboard,
++   size = 0.8, fold_id = 1, task = task, crs = 4326,
++   show_blocks = TRUE, show_labels = TRUE) *
++   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
++   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+\end{CodeInput}
+
+\begin{center}\includegraphics{paper_files/figure-latex/unnamed-chunk-4-1} \end{center}
+
+\end{CodeChunk}
+
+\hypertarget{environmental-blocking}{%
+\subsection{Environmental Blocking}\label{environmental-blocking}}
+
+The last method from the \pkg{blockCV} package, ``environmental
+blocking'', makes use of \emph{k-means} clustering \citep{hartigan1979a}
+in a possibly multivariate space. The user can specify a variable via
+argument \texttt{feature} which levels will be used to create clusters.
+Hereby, k-means will use Euclidean distance for cluster creation. Hence,
+the selected input variables should be of type numeric. To avoid a
+potential bias introduced by features with high variance when selecting
+multiple features, all features are standardized by default.
+
+The following example clusters by feature ``distance to forest''.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> resampling_env <- rsmp("spcv_env",
++   features = "distdeforest", folds = 5)
+R> 
+R> autoplot(resampling_env,
++   size = 0.8, fold_id = 1, task = task, crs = 4326,
++   show_blocks = TRUE, show_labels = TRUE) *
++   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
++   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+\end{CodeInput}
+
+\begin{center}\includegraphics{paper_files/figure-latex/unnamed-chunk-5-1} \end{center}
+
+\end{CodeChunk}
+
+It is also possible to cluster by multiple features:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> resampling_env_multi <- rsmp("spcv_env",
++   features = c("distdeforest", "slope"), folds = 5)
+R> 
+R> autoplot(resampling_env_multi,
++   size = 0.8, fold_id = 1, task = task, crs = 4326,
++   show_blocks = TRUE, show_labels = TRUE) *
++   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
++   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+\end{CodeInput}
+
+\begin{center}\includegraphics{paper_files/figure-latex/unnamed-chunk-6-1} \end{center}
 
@@
 
-\hypertarget{environmental-blocking}{%
-\subsection{Environmental Blocking}\label{environmental-blocking}}
+The ``Spatial CV'' method after \citep{sperrorest} and implemented in
+package \pkg{sperrorest} is similar to method ``environmental blocking''
+in that it also uses \emph{k-means} clustering for fold creation.
+However, it does not make use of arbitrary features but uses the
+coordinates of all observations to create clusters in the spatial
+domain. This is what lead to the naming of \texttt{spcv\_coords} in
+\pkg{mlr3spatiotempcv}.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> resampling_coords <- rsmp("spcv_coords", folds = 5)$instantiate(task)
+R> 
+R> autoplot(resampling_coords,
++   size = 0.8, fold_id = 1, task = task, crs = 4326) *
++   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
++   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+\end{CodeInput}
+
+\begin{center}\includegraphics{paper_files/figure-latex/unnamed-chunk-7-1} \end{center}
+
+\end{CodeChunk}
 
@@
 (LLOT)}\label{leave-location-and-time-out-llot}}
+
+Leave-Location-and-Time-Out from packge \pkg{CAST} is a spatiotemporal
+resampling method. It requires both a spatial and a temporal variable
+which are used to to separate observations in a multi-dimensional space.
+
+In this example the \texttt{cookfarm} dataset is used because it
+contains a temporal variable (``Date'') which is not present in dataset
+\texttt{ecuador}.
+
+The visualization is done in a dynamic way via \pkg{plotly} to visualize
+the both temporal and spatial dimensions.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> task_spt = tsk("cookfarm")
+R> resampling_cstf <- rsmp("sptcv_cstf",
++   folds = 5,
++   time_var = "Date", space_var = "SOURCEID")
+R> 
+R> autoplot(resampling_cstf,
++   fold_id = 2, task = task_spt, crs = 4326, static_image = TRUE,
++   file = "plot.pdf") *
++   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
++   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+\end{CodeInput}
+\begin{CodeOutput}
+NULL
+\end{CodeOutput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{plot} \end{center}
+
+\end{CodeChunk}
 

### 2021-04-02 paper/paper.Rmd
- use pdf_book reference mechanism - use cache - adjust figure placements - restructure chapters - write CSTF methods

@@
 bibliography: paper.bib
-output: rticles::jss_article
+# output: rticles::jss_article
+output:
+  bookdown::pdf_book:
+    base_format: rticles::jss_article
 ---
+
+```{r setup, echo=FALSE}
+# important for correct figure placement
+# see https://stackoverflow.com/a/49395389/4185785
+knitr::knit_hooks$set(plot = function(x, options) {
+  knitr::hook_plot_tex(x, options)
+})
+# knitr::opts_chunk$set(cache = FALSE)
+```
 
@@
 
-# Comparing Spatial and Non-Spatial Resampling
+# Visualization of Spatiotemporal Partitions
 
-To introduce the effect of spatial autocorrelation in modeling, a spatial and non-spatial CV using \pkg{mlr3spatiotempcv} will be shown in the following.
-For simplicity, one of the most used methods (spatial k-means clustering after @sperrorest) will be used.
-Individual methods are discussed in FIXME.
+Every partitioning method in [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) comes with a generic `plot()` method to visualize the created groups.
+In a 2D space this happens via [ggplot2](https://ggplot2.tidyverse.org) [@ggplot2] while for spatiotemporal methods 3D visualizations are created via [plotly](https://github.com/ropensci/plotly) [@plotly].
 
-The performance of a simple classification tree (`"classif.ranger"`) is evaluated on a random repeated cross-validation (`"repeated_cv"`) with four folds and two repetitions.
-In reality, more fold (between five and ten) and a higher number of repetitions should be used.
-The chosen evaluation measure is the "Area Under the ROC Curve" (`"classif.auc"`) which is a common evaluation measure for binary responses.
-The CV will be executed twice with the only difference being the use of a spatial resampling object (initialized with `rsmp("repeated_spcv_coords"`) instead of a non-spatial one (`rsmp("repeated_cv"`).
+Unless specified by the user, the coordinate reference system (CRS) defaults to EPSG code 4326 (WGS84).
+This is because a lat/lon based CRS is better suited for plotting purposes than a Mercator (UTM) one.
+Note that setting the correct CRS for the given data *during construction* is very important.
+Even though EPSG 4326 is a good fallback and often used for visualization purposes, spatial offsets of up to multiple meters may occur if the wrong CRS was passed initially.
 
-## Data Preparation
+Note: The following example uses a built-in task via function `tsk()`.
+In practice however, one needs to create a spatiotemporal task first via `TaskClassifST()`/`TaskRegrST()`..
 
-The following code chunk first loads all necessary R packages and sets a low verbosity for this example.
-A seed is set for deterministic execution.
+The spatial grouping of the k-means based approach (`"spcv_coords"`) (Fig.\@ref(fig:vis-spcv)) contrasts visually very well to the NSpCV (random) partitioning (Fig.\@ref(fig:vis-nspcv)).
 
-Next the mlr3 task `"ecuador"`, which is included in \pkg{mlr3spatiotempcv}, is loaded via `tsk()`.
-This function is a convenience wrapper around the underlying R6 construction method for a spatiotemporal classification task (`mlr3spatiotempcv::TaskClassifST$new()`).
-
-Last, the random forest learner (`lrn("classif.ranger")`) is initialized with default hyperparameters and the prediction type set to `"probability"` due to the binary response variable.
-A set of commonly used learners is available in \pkg{mlr3learners} [@mlr3learners], including the random forest implementation after @ranger.
-
-By setting the logging threshold, which can be controlled via the \pkg{lgr} package [@lgr], to `"warn"`, verbosity is reduced to keep the example tidy.
-
-```{r setup, echo=FALSE}
-# important for correct figure placement
-# see https://stackoverflow.com/a/49395389/4185785
-knitr::knit_hooks$set(plot = function(x, options) {
-  knitr::hook_plot_tex(x, options)
-})
-knitr::opts_chunk$set(cache = FALSE)
-```
-
-```{r prepare, cache=FALSE}
-library("mlr3")
-library("mlr3spatiotempcv")
-library("mlr3learners")
-set.seed(42)
+```{r vis-spcv, fig.cap="Visualization of spatial partitioning", fig.pos="t!", fig.height=6, fig.width=8, cache=TRUE}
+library(mlr3)
+library(mlr3spatiotempcv)
 
@@
 
-task <- tsk("ecuador")
+set.seed(42)
 
-learner <- lrn("classif.ranger", predict_type = "prob")
+task = tsk("ecuador")
+
+resampling_sp <- rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
+resampling_sp$instantiate(task)
+
+autoplot(resampling_sp, task, fold_id = c(1:2), crs = 4326, size = 0.8) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
 ```
 
-### Non-Spatial Cross-Validation
+```{r vis-nspcv, fig.cap="Visualization of random partitioning", fig.pos="t!", fig.height=6, fig.width=8, cache=T}
+resampling_nsp <- rsmp("repeated_cv", folds = 4, repeats = 2)
+resampling_nsp$instantiate(task)
 
-```{r rr-nspcv, cache=FALSE}
-resampling_nsp <- rsmp("repeated_cv", folds = 4, repeats = 2)$instantiate(task)
-rr_nsp <- resample(
-  task = task, learner = learner,
-  resampling = resampling_nsp
-)
-
-rr_nsp$aggregate(measures = msr("classif.auc"))
+autoplot(resampling_nsp, task, fold_id = c(1:2), crs = 4326, size = 0.8) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
 ```
-
-## Spatial Cross-Validation
-
-```{r rr-spcv, cache=FALSE}
-resampling_sp <- rsmp("repeated_spcv_coords", folds = 4, repeats = 2)$instantiate(task)
-rr_sp <- resample(
-  task = task, learner = learner,
-  resampling = resampling_sp
-)
-
-rr_sp$aggregate(measures = msr("classif.auc"))
-```
-
-Here, the estimated AUC using Non-Spatial Cross-Validation (NSpCV) is around 0.15 better compared to the Spatial Cross-Validation (SpCV).
-The better performance score results from over-optimism in the estimation with NSpCV.
-The actual predictive performance on new data is much more realistically reflected by the performance estimated via SpCV.
-The magnitude of this difference is variable as it depends on the dataset, the magnitude of STAC and the learner itself.
-Algorithms with a higher tendency of overfitting to the training set will have a large spread in such scenarios.
 
@@
 
-At the time of writing, the following spatiotemporal resampling methods (FIXME ref table) were available across different R packages.
+At the time of writing, the following spatiotemporal resampling methods (Table \@ref(tab:sptcv-methods)) were available across different R packages.
 This list might increase over time and it is recommended to consult the homepage of \pkg{mlr3spatiotempcv} for an up-to-date list.
@@
 Due to its similarity to a LOO CV, this method will result in as many train/test evaluations as observations are present.
-In the example below, a buffer zone of 1000 m is used and the first fold shown is plotted.
+In the example below, a buffer zone of 1000 m is used and the first fold shown is plotted (Fig.\@ref(fig:buffer))).
 
-```{r}
-resampling_buffer <- rsmp("spcv_buffer", theRange = 1000)$instantiate(task)
+```{r buffer, fig.cap="Visualization of 'Spatial buffering' method.", out.width="50%", cache=T}
+resampling_buffer <- rsmp("spcv_buffer", theRange = 1000)
+resampling_buffer$instantiate(task)
 
-autoplot(resampling_buffer, size = 0.7, task = task, fold_id = 1, crs = 4326) *
+autoplot(resampling_buffer,
+  size = 0.8, task = task, fold_id = 1, crs = 4326) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
@@
 
-`selection = "random"`
+### Option: selection = "random"
 
-```{r}
+```{r block-random, cache = TRUE, fig.cap="Spatial blocking with option `selection = 'random'`.", out.width="50%", fig.pos="h"}
 resampling_block_random <- rsmp("spcv_block", range = 1000, folds = 5)
@@
 
-`selection = "systematic"`: Here, aggregation follows the linear count from $1:n_{folds} from top to bottom.
+### Option: selection = "systematic"
 
-```{r}
+Here, aggregation follows the linear count from $1:n_{folds}$ from top to bottom.
+
+```{r block-systematic, cache=TRUE, fig.cap="Spatial blocking with option `selection = 'systematic'`.", out.width="50%", fig.pos="h"}
 resampling_block_systematic <- rsmp("spcv_block",
   range = 1000, folds = 5,
-  selection = "systematic")
+  selection = "systematic"
+)
 
@@
 
-`selection = "checkerboard"`: This option ignores paramter `range` and always creates two partitions following a "1,2,1,2" pattern.
+### Option: selection = "checkerboard"
 
-```{r}
+This option ignores parameter `range` and always creates two partitions following a "1,2,1,2" pattern.
+
+```{r block-checkerboard, cache=TRUE, fig.cap="Spatial blocking with option `selection = 'checkerboard'`.", out.width="50%", fig.pos="h"}
 resampling_block_checkerboard <- rsmp("spcv_block",
@@
 
-```{r}
+```{r env-1, fig.cap="Environmental blocking using a single predictor.", cache=TRUE, out.width="50%", fig.pos="h"}
 resampling_env <- rsmp("spcv_env",
@@
 
-```{r}
+```{r env-2, fig.cap="Environmental blocking using multiple predictors.", cache=TRUE, out.width="50%", fig.pos="h"}
 resampling_env_multi <- rsmp("spcv_env",
@@
 
-```{r}
+```{r coords, fig.cap="Method 'Spatial CV' showing a k-means clustering approach based on coordinates.", cache=TRUE, out.width="50%", fig.pos="h"}
 resampling_coords <- rsmp("spcv_coords", folds = 5)$instantiate(task)
@@
 
-## Leave-Location-and-Time-Out (LLOT)
+## CreateSpacetimeFolds (Cstf)
 
-Leave-Location-and-Time-Out from packge \pkg{CAST} is a spatiotemporal resampling method.
+Leave-Location-and-Time-Out from packge \pkg{CAST} is a spatiotemporal resampling method [@meyer2018a].
 It requires both a spatial and a temporal variable which are used to to separate observations in a multi-dimensional space.
+It comes with satellite methods named "Leave-time-out" (LTO) and "Leave-location-out" (LLO) which only operatre in a two-dimensional space.
 
@@
 
-The visualization is done in a dynamic way via \pkg{plotly} to visualize both temporal and spatial dimensions.
-Because a dynamic image cannot be included in this manuscript, a static version generated by setting `static_image = TRUE` was used.
+Two visualization types are supported: a two-dimensional and a three-dimensional method which can be selected via the logical argument `plot3d`.
+The 3D visualization has been implemented via \pkg{plotly}.
+Because a dynamic image cannot be included in this manuscript, a static version, which can be generated by setting `static_image = TRUE`, is shown.
 
-```{r}
-task_spt = tsk("cookfarm")
-resampling_cstf <- rsmp("sptcv_cstf",
-  folds = 5,
-  time_var = "Date", space_var = "SOURCEID")
+In \pkg{mlr3spatiotempcv} the method is named `"sptcv_cstf"` because the underlying upstream implementation in package \pkg{CAST} is named `CreateSpacetimeFolds()`.
 
-autoplot(resampling_cstf,
-  fold_id = 1, task = task_spt, crs = 4326, static_image = TRUE,
-  file = "plot.pdf") *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+### Leave-Time-Out (LTO)
+
+To better showcase the partitioning of the "Leave-Time-Out" (LTO) method, the temporal variable `Date` was adjusted in this example.
+
+The following code uses the `cookfarm` dataset as the base, modifies to the `Date` variable to contain five unique levels and then creates a spatiotemporal regression task in \pkg{mlr3spatiotempcv} (Fig.\@ref(fig:lto)).
+
+```{r, results='hide', cache=TRUE}
+data = cookfarm_sample
+data$Date = rep(c(
+  "2020-01-01", "2020-02-01", "2020-03-01", "2020-04-01",
+  "2020-05-01"), times = 1, each = 100)
+b = mlr3::as_data_backend(data)
+b$hash = "_mlr3_tasks_cookfarm_"
+task_spt = TaskRegrST$new(
+  id = "cookfarm", b, target = "PHIHOX",
+  extra_args = list(
+    coordinate_names = c("x", "y"), coords_as_features = FALSE,
+    crs = 26911)
+)
+
+resampling_cstf_time = rsmp("sptcv_cstf", folds = 5, time_var = "Date")
+resampling_cstf_time$instantiate(task_spt)
+
+autoplot(resampling_cstf_time,
+  fold_id = 5, task = task_spt, crs = 4326, plot3D = TRUE
+)
 ```
 
-```{r, echo=FALSE}
-knitr::include_graphics("plot.pdf", dpi = 100)
+```{r lto, echo=FALSE, out.width="90%", fig.cap="Three-dimensional visualization of method `sptcv\\_cstf` and sub-method Leave-Time-Out (LTO).
+To support showcasing the temporal grouping, five folds and five temporal levels in variable `Date` were used in this example.", fig.pos="h"}
+knitr::include_graphics("lto.png")
+```
+
+### Leave-Location-Out (LLO)
+
+Method "Leave-Location-Out" (LLO) works similar to "Leave-Time-Out" (LTO) in that it operates in a two-dimensional space.
+Here, specific locations are left out for forming the partitions while the temporal information is ignored (Fig.\@ref(fig:llo)).
+
+```{r cache=TRUE, results='hide'}
+resampling_cstf_loc = rsmp("sptcv_cstf", folds = 5, space_var = "SOURCEID")
+resampling_cstf_loc$instantiate(task_spt)
+
+autoplot(resampling_cstf_loc,
+  fold_id = 5, task = task_spt, crs = 4326, plot3D = TRUE)
+```
+
+```{r llo, echo=FALSE, out.width="90%", fig.cap="Three-dimensional visualization of method 'sptcv\\_cstf' and sub-method 'Leave-Location-Out' (LLO).
+To support showcasing the spatial grouping, the viewing angle has been adjusted to a view from top.
+Points are stacked by temporal levels and only points from the same location are chosen to form a partition.", fig.pos="h"}
+knitr::include_graphics("llo.png")
+```
+
+### Leave-Location-and-Time-Out (LLTO)
+
+The "Leave-Location-and-Time-Out" (LLTO) method combines both "Leave-Time-Out" (LTO) and "Leave-Location-Out" (LLO).
+It creates partitions that operate on spatiotemporal clusters to reduce both spatial and temporal autocorrelations.
+To achieve this, observations are omitted and only a subset is used.
+In the special scenario with five temporal levels and five folds, the train/test split happens only at one temporal level (Fig.\@ref(fig:llto)).
+
+```{r cache=TRUE, results='hide'}
+resampling_cstf_time_loc = rsmp("sptcv_cstf",
+  folds = 5,
+  space_var = "SOURCEID", time_var = "Date")
+resampling_cstf_time_loc$instantiate(task_spt)
+
+autoplot(resampling_cstf_time_loc,
+  fold_id = 4, task = task_spt, crs = 4326, plot3D = TRUE,
+  show_omitted = TRUE)
+```
+
+```{r llto, echo=FALSE, out.width="90%", fig.cap="Three-dimensional visualization of method 'sptcv\\_cstf' and sub-method 'Leave-Location-And-Time-Out' (LLTO).
+The grey points show the points omitted by this method.", cache=TRUE, fig.pos="h"}
+knitr::include_graphics("llto.png")
 ```
@@
 
-##
+# Comparing Spatial and Non-Spatial Resampling
 
-# Visualization of Spatiotemporal Partitions
+To show the effect of spatial autocorrelation in modeling, a spatial and non-spatial CV using \pkg{mlr3spatiotempcv} will be shown in the following.
+For simplicity, one of the most used methods (spatial k-means clustering after @sperrorest) will be used.
+Individual methods are discussed in FIXME.
 
-Every partitioning method in [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) comes with a generic `plot()` method to visualize the created groups.
-In a 2D space this happens via [ggplot2](https://ggplot2.tidyverse.org) [@ggplot2] while for spatiotemporal methods 3D visualizations via [plotly](https://github.com/ropensci/plotly) [@plotly] are created.
+The performance of a simple classification tree (`"classif.ranger"`) is evaluated on a random repeated cross-validation (`"repeated_cv"`) with four folds and two repetitions.
+In reality, more fold (between five and ten) and a higher number of repetitions should be used.
+The chosen evaluation measure is the "Area Under the ROC Curve" (`"classif.auc"`) which is a common evaluation measure for binary responses.
+The CV will be executed twice with the only difference being the use of a spatial resampling object (initialized with `rsmp("repeated_spcv_coords"`) instead of a non-spatial one (`rsmp("repeated_cv"`).
 
-```{r vis-spcv, fig.cap="Visualization of spatial partitioning", fig.pos="!b", fig.height=6, fig.width=8}
-autoplot(resampling_sp, task, fold_id = c(1:4)) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+## Data Preparation
+
+The following code chunk first loads all necessary R packages and sets a low verbosity for this example.
+A seed is set for deterministic execution.
+
+Next the mlr3 task `"ecuador"`, which is included in \pkg{mlr3spatiotempcv}, is loaded via `tsk()`.
+This function is a convenience wrapper around the underlying R6 construction method for a spatiotemporal classification task (`mlr3spatiotempcv::TaskClassifST$new()`).
+
+Next, the random forest learner (`lrn("classif.ranger")`) is initialized with default hyperparameters and the prediction type set to `"probability"` due to the binary response variable.
+A set of commonly used learners is available in \pkg{mlr3learners} [@mlr3learners], including the random forest implementation after @ranger.
+
+By setting the logging threshold, which can be controlled via the \pkg{lgr} package [@lgr], to `"warn"`, verbosity is reduced to keep the example tidy.
+
+```{r prepare, cache=TRUE}
+library("mlr3")
+library("mlr3spatiotempcv")
+library("mlr3learners")
+
+task <- tsk("ecuador")
+
+learner <- lrn("classif.ranger", predict_type = "prob")
 ```
-\
 
-Unless specified by the user, the coordinate reference system (CRS) defaults to EPSG code 4326 (WGS84).
-This is because a lat/lon based CRS is better suited for plotting purposes than a Mercator (UTM) one.
-Note that setting the correct CRS for the given data *during construction* is very important.
-Even though EPSG 4326 is a good fallback and often used for visualization purposes, spatial offsets of up to multiple meters may occur if the wrong CRS was passed initially.
+### Non-Spatial Cross-Validation
 
-This example used an already created task via the convenience function `tsk()`.
-In practice however, one needs to create a spatiotemporal task via `TaskClassifST()`/`TaskRegrST()` and set the `crs` argument.
+Here, a four-fold, two times repeated non-spatial resampling structure is constructed and instantianeted via method `$instantiate()`.
+Next, the created resampling object `resampling_nsp` is passed to function `resample()` together with the Task and Learner created earlier to execute the cross-validation.
 
-The spatial grouping of the k-means based approach above (Figure 1) contrasts visually very well compared to the NSpCV (random) partitioning (Figure 2):
+Last, the results from the CV are aggregated using the measure `"classif.auc"`, which is the technical notation for the Area Under the Receivers Operating Curve (AUROC).
 
-```{r vis-nspcv, fig.cap="Visualization of random partitioning", fig.pos="!b", fig.height=6, fig.width=8}
-autoplot(resampling_nsp, task, fold_id = c(1:4), crs = 4326) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```{r rr-nspcv, cache=TRUE, cache.lazy = FALSE}
+resampling_nsp <- rsmp("repeated_cv", folds = 4, repeats = 2)
+resampling_nsp$instantiate(task)
+rr_nsp <- resample(
+  task = task, learner = learner,
+  resampling = resampling_nsp
+)
+
+rr_nsp$aggregate(measures = msr("classif.auc"))
 ```
+
+## Spatial Cross-Validation
+
+The same approach as for non-spatial resampling is used for spatial resampling, expect the method being `"repeated_spcv_coords"` instead of `"repeated_cv"`.
+
+```{r rr-spcv, cache=TRUE, cache.lazy = FALSE}
+resampling_sp <- rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
+resampling_sp$instantiate(task)
+rr_sp <- resample(
+  task = task, learner = learner,
+  resampling = resampling_sp
+)
+
+rr_sp$aggregate(measures = msr("classif.auc"))
+```
+
+The resulting AUC using Non-Spatial Cross-Validation (NSpCV) is around 0.15 better compared to the Spatial Cross-Validation (SpCV).
+The better performance score results from over-optimism in the estimation with NSpCV.
+The actual predictive performance on new data is much more realistically reflected by the performance estimated via SpCV.
+The magnitude of this difference is variable as it depends on the dataset, the magnitude of STAC and the learner itself.
+Algorithms with a higher tendency of overfitting to the training set will have a large spread in such scenarios.
 

### 2021-04-02 paper/paper.Rmd
- add cluto method - optimize "choosing a method"

@@
 The "Leave-Location-and-Time-Out" (LLTO) method combines both "Leave-Time-Out" (LTO) and "Leave-Location-Out" (LLO).
-It creates partitions that operate on spatiotemporal clusters to reduce both spatial and temporal autocorrelations.
+It creates partitions that operate on spatiotemporal clusters to reduce both spatial and temporal autocorrelation.
 To achieve this, observations are omitted and only a subset is used.
@@
 ## Spatiotemporal Clustering (Cluto)
+
+The "CLUTO" clustering algorithm by George Karypis is said to be an application "for clustering of low- and high-dimensional datasets and for analyzing the characteristics of the various clusters" [@cluto].
+
+The method can generally be used in {mlr3spatiotempcv} if users cope with the copyright statement of the tool, which restricts usage to non-profit organizations and US government agencies only, and make the application binary available locally.
+
+These restrictions in combination with
+
+- the requirement of a local binary which needs to be downloaded externally and added to a user's system path
+- the absent of updates since the year 2007
+- a http-only website 
+- no response from the copyright holder after asking for the current status of the project
+
+led to the decision to not explain and showcase this method in greater detail in this manuscript.
+Users are encouraged to visit the help page of the package for this method (`?ResamplingSptCVCluto`) for more information.
+
+# List of spatiotemporal resampling methods
+
+Table \@ref(tab:sptcv-methods) lists all available spatiotemporal resampling methods available in \pkg{mlr3spatiotempcv}.
+Column `Package` denotes the upstream package from which the method was incorporated.
+The internal identifier for the respective method in \pkg{mlr3spatiotempcv} is shown in column `Notation`.
 
@@
 
-# Choosing a Resampling Method
+# Discussion
 
-While the example used the `"spcv_coords"` method, this does not mean that this method is the best or only method suitable for this task.
-Even though this method is quite popular, it was mainly chosen because of the clear visual grouping differences compared to random partitioning.
+## Choosing a Resampling Method
+
+While in the example the `"spcv_coords"` method was used, this should not imply that this method is the best or only method suitable for this example task.
+Even though this method is frequently used, it was mainly chosen because of it's clear visual grouping differences when comparing spatial to random partitioning.
 
 In fact, most often multiple spatial partitioning methods can be used for a dataset.
-It is recommended (required) that users familiarize themselves with each implemented method and decide which method to choose based on the specific characteristics of the dataset.
-For almost all methods implemented in [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com), there is a scientific publication describing the strengths and weaknesses of the respective approach (either linked in the help file of [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) or its respective dependency packages).
+It is recommended (required) that users familiarize themselves with multiple methods and decide which method to choose based on the specific characteristics of the dataset at hand.
+There are various variables influencing the partitioning method choice for an environmental dataset:
+
+- Do natural groupings exists in the data?
+- How dense are the observations scattered both in space and time?
+- Are the predictor variables known to be highly affected by spatiotemporal autocorrelation by nature?
+
+Based on these criteria users must choose a matching partitioning method that is either more restrictive (by discarding observations for fold creation) or more liberal (by using all observations and eventually ignoring natural grouping patterns).
+For almost all methods implemented in \pkg{mlr3spatiotempcv}, there is a scientific publication describing the strengths and weaknesses of the respective approach (either linked in the help file of the package or its respective upstream packages).
 
 In the example above, a cross-validation without hyperparameter tuning was shown.
-If a nested CV is desired, it is recommended to use the same spatial partitioning method for the inner loop (= tuning level).
+In practice most often a nested CV is desired.
+In this case it is recommended to use the same spatial partitioning method for the inner loop (= tuning level).
 See @schratz2019 for more details and chapter 11 of [Geocomputation with R](https://geocompr.robinlovelace.net/spatial-cv.html) [@lovelace2019].
 
-A list of all implemented methods in [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) can be found in the [Getting Started](https://mlr3spatiotempcv.mlr-org.com/articles/mlr3spatiotempcv.html) vignette of the package.
+## A critical evaluation of all methods
+
+Even though all methods have their pros and cons and no universal method exists, this chapter should provide a critial view on the limitations of each method.
+
+<!-- spcv coords: mention generic name -->
+
+<!-- sptcv LLTO: discards many observations -->
+<!-- sptcv LLO: requires additional variable to specify groupings -->
+<!-- sptcv LTO: ignores spatial autocor entirely -->
+
+<!-- spcv buffer: discards observations, LOO very expensive -->
+
+<!-- missing: non-LOO method with buffer support -->
+
+<!-- env block: hard to decide which variable to use for clustering -->
+<!-- env block: more than one variable may lead to unforeseeable results -->
 

### 2021-04-04 paper/paper.Rmd
first readable draft

@@
 title:
-  formatted: "\\pkg{mlr3spatiotempcv}: Spatiotemporal Resampling Methods for \\pkg{mlr3}"
+  formatted: "\\pkg{mlr3spatiotempcv}: Spatiotemporal resampling methods for \\pkg{mlr3}"
   # If you use tex in the formatted title, also supply version without
-  plain:     "mlr3spatiotempcv: Spatiotemporal Resampling Methods for mlr3"
+  plain:     "mlr3spatiotempcv: spatiotemporal resampling methods for mlr3"
   # For running headers, if needed
@@
 With R being one of the most used languages with respect to modeling/statistics and academia in general, R packages were developed aiming to simplify the data splitting with respect to methods able to account for spatial autocorrelation [@cast, @blockCV].
-FIXME: adds refs
+
 This is a typical process for the R community as writing R packages has been made relatively straightforward in recent years thanks to efforts from various people/organizations such as Bioconductor, CRAN, ropensci or the tidyverse [@cran, @ropensci, @tidyverse]
@@
 
-The downside of this culture is that "island solutions" are being created, i.e. small packages which serve a particular (sometimes very niche) use case.
-Usually the "Do one thing and do it well" UNIX philosophy is widely known and used in practice [@raymond2003art].
-However, the issue with R packages is that not every package is necessarily easily compatiable with other ones due to differing input and returns objects, making it problematic to write a clean and tidy workflow.
-Also many R packages are written by single persons which not necessarly have a complete overview over existing implementations.
+The downside which comes with this is that "island solutions" are being created, i.e. small packages which serve a particular (sometimes very niche) use case.
+While the UNIX philosophy "Do one thing and do it well" [@raymond2003art] is widely known and used in the R community, R packages from different developers do often not play well with each other.
+This is due to differing input and returns objects, requiring the need to pre-and postprocess objects around such R packages which provide tools for specific modeling tasks.
+Also often R packages implement only an opinionated subset of methods which require users to combine multiple packages for different steps.
+
 These points may lead to the following possible issues in practice:
 
-- Different syntax.
+- Differing syntax.
 - Support for only some predictor types (e.g. numeric and factor features but no ordered factors).
-- Different return values.
-- Different namings for the same method.
+- Differing return values.
+- Differing namings for the same method.
 - Unclear longterm maintenance.
@@
 \pkg{mlr3spatiotempcv} aims to provide a solution for some of the points outlined above by wrapping all spatiotemporal resampling methods available in \proglang{R} packages.
-By embedding the package within the \pkg{mlr3} ecosystem and following its unified syntax and scalable approach, \pkg{mlr3spatiotempcv} aims to provide future users longterm support for spatiotemporal partitioning methods in R.
+By embedding the package within the \pkg{mlr3} ecosystem, \pkg{mlr3spatiotempcv} aims to provide longterm support for spatiotemporal partitioning methods in R while making all additional benefits of the machine learning framework available to users.
 
@@
 
-# Visualization of Spatiotemporal Partitions
+# Visualization of spatiotemporal partitions
 
@@
 
-Unless specified by the user, the coordinate reference system (CRS) defaults to EPSG code 4326 (WGS84).
-This is because a lat/lon based CRS is better suited for plotting purposes than a Mercator (UTM) one.
-Note that setting the correct CRS for the given data *during construction* is very important.
-Even though EPSG 4326 is a good fallback and often used for visualization purposes, spatial offsets of up to multiple meters may occur if the wrong CRS was passed initially.
+Visualization of partitioning methods is a vital part of \pkg{mlr3spatiotempcv} as most upstream packages lack support for such.
+The semantics are presented first because plotting will be used extensively when presenting the integrated methods.
 
-Note: The following example uses a built-in task via function `tsk()`.
-In practice however, one needs to create a spatiotemporal task first via `TaskClassifST()`/`TaskRegrST()`..
+Unless specified by the user, the coordinate reference system (CRS) defaults to EPSG code 32717 (WGS84).
+This is because a lat/lon based CRS is better suited for plotting purposes than a Mercator (UTM) one due tho a more descriptive axes labeling.
+Setting an appropriate CRS for the given data *during task construction* is very important.
+Even though EPSG 32717 is a good fallback and often used for visualization purposes, spatial offsets of up to multiple meters may occur if the wrong CRS was passed initially.
 
-The spatial grouping of the k-means based approach (`"spcv_coords"`) (Fig.\@ref(fig:vis-spcv)) contrasts visually very well to the NSpCV (random) partitioning (Fig.\@ref(fig:vis-nspcv)).
+The following example creates a spatiotemporal task from scratch.
+The data used here is also available as a built-in task via `tsk("ecuador")`.
 
-```{r vis-spcv, fig.cap="Visualization of spatial partitioning", fig.pos="t!", fig.height=6, fig.width=8, cache=TRUE}
+Because the spatial grouping of the k-means based approach (`"spcv_coords"`) (Fig.
+\@ref(fig:vis-spcv)) contrasts visually very well to the NSpCV (random) partitioning (Fig.
+\@ref(fig:vis-nspcv)), this method was chosen to showcase the 2D plotting capabilities of \pkg{mlr3spatiotempcv}.
+
+The partitions shown in the following are also used in the case study in section \@ref(sec:case-study).
+
+Besides the required arguments of the \pkg{mlr3} superclass `Task`, `TaskClassifST` / `TaskRegrST` classes also require the following arguments passed via a named list `extra_args`:
+
+- `coordinate_names`: Names of the features which represent the spatial coordinates.
+  This is automatically inferred when a `sf` object is passed.
+- `coords_as_features`: Whether the coordinates should be used as features during modeling.
+  By default coordinates are not used as predictors.
+- `crs`: The coordinate reference system of the data. Valid inputs are PROJ4 strings or EPSG codes in the format `ESPG: <code>`.
+
+```{r create task, cache=TRUE}
 library(mlr3)
@@
 
-task = tsk("ecuador")
+# equal to `tsk("ecuador")`
+b = mlr3::as_data_backend(ecuador)
+b$hash = "_mlr3_tasks_ecuador_" # optional
+task = TaskClassifST$new(
+  id = "ecuador", b, target = "slides", positive = "TRUE",
+  extra_args = list(
+    coordinate_names = c("x", "y"), coords_as_features = FALSE,
+    crs = "+proj=utm +zone=17 +south +datum=WGS84 +units=m +no_defs") # EPSG:32717
+)
+```
 
-resampling_sp <- rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
+Due to space limitations only the first two folds are plotted.
+
+```{r vis-spcv, fig.cap="Visualization of spatial partitioning", fig.pos="ht", fig.height=6, fig.width=8, cache=TRUE}
+resampling_sp = rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
 resampling_sp$instantiate(task)
 
-autoplot(resampling_sp, task, fold_id = c(1:2), crs = 4326, size = 0.8) *
+autoplot(resampling_sp, task, fold_id = c(1:2), crs = 32717, size = 0.8) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
@@
 
-```{r vis-nspcv, fig.cap="Visualization of random partitioning", fig.pos="t!", fig.height=6, fig.width=8, cache=T}
-resampling_nsp <- rsmp("repeated_cv", folds = 4, repeats = 2)
+```{r vis-nspcv, fig.cap="Visualization of random partitioning", fig.pos="ht", fig.height=6, fig.width=8, cache=T}
+resampling_nsp = rsmp("repeated_cv", folds = 4, repeats = 2)
 resampling_nsp$instantiate(task)
 
-autoplot(resampling_nsp, task, fold_id = c(1:2), crs = 4326, size = 0.8) *
+autoplot(resampling_nsp, task, fold_id = c(1:2), crs = 32717, size = 0.8) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
@@
 
-# Spatiotemporal Partitioning Methods
+# Spatiotemporal partitioning methods
 
-At the time of writing, the following spatiotemporal resampling methods (Table \@ref(tab:sptcv-methods)) were available across different R packages.
+At the time of writing, the following spatiotemporal resampling methods (Tab. \@ref(tab:sptcv-methods)) were available across different R packages.
 This list might increase over time and it is recommended to consult the homepage of \pkg{mlr3spatiotempcv} for an up-to-date list.
+The internal identifier for the respective method in \pkg{mlr3spatiotempcv} is shown in column `Notation`.
 
-A low-level introduction is given for every method included at this point of this article.
-For a discussion with respect to usage see section FIXME.
+\begin{table}[h]
+	\centering
+	\caption[t]{Spatiotemporal resampling methods, ordered alphabetically by class name (Notation column).}
+	\begingroup
+	\begin{tabular}{llll}
+		\\
+		Literature                  & Package          & Reference          & Notation                    \\
+		\toprule
+    Spatial Buffering           & \pkg{blockCV}    & \citet{blockCV}    & \texttt{rsmp("spcv\_buffer")} \\
+    Spatial Blocking            & \pkg{blockCV}    & \citet{blockCV}    & \texttt{rsmp("spcv\_block")}  \\
+    Spatial CV                  & \pkg{sperrorest} & \citet{sperrorest} & \texttt{rsmp("spcv\_cv")}     \\
+    Environmental Blocking      & \pkg{blockCV}    & \citet{blockCV}    & \texttt{rsmp("spcv\_env")}    \\
+    \midrule
+    Leave-Location-and-Time-Out & \pkg{CAST}       & \citet{cast}       & \texttt{rsmp("spcv\_cstf")}   \\
+    Spatiotemporal Clustering   & \pkg{skmeans}    & \citet{cluto}      & \texttt{rsmp("spcv\_cluto")}
+	\end{tabular}
+	\endgroup\label{tab:sptcv-methods}
+\end{table}
 
-## Spatial Buffering
+A low-level introduction including visualization is given for every method in the following.
+See section \@ref(sec:disc) for a discussion of the methods.
+
+## Spatial buffering
 
@@
 ```{r buffer, fig.cap="Visualization of 'Spatial buffering' method.", out.width="50%", cache=T}
-resampling_buffer <- rsmp("spcv_buffer", theRange = 1000)
+resampling_buffer = rsmp("spcv_buffer", theRange = 1000)
 resampling_buffer$instantiate(task)
@@
 autoplot(resampling_buffer,
-  size = 0.8, task = task, fold_id = 1, crs = 4326) *
+  size = 0.8, task = task, fold_id = 1, crs = 32717) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
@@
 
-## Spatial Blocking
+## Spatial blocking
 
@@
 
-The way how the blocks are created can be manifold:
-
-1. Supplying a numeric value in meters via argument `theRange` (named just `range` in {mlr3spatiotempcv}) will create a quadratic block pattern across the study area.
+The way how the blocks are created can be manifold: Supplying a numeric value in meters via argument `theRange` (named just `range` in {mlr3spatiotempcv}) will create a quadratic block pattern across the study area.
 The aggregation of zones can either be random (default) (`selection = "random"`), systematic (`selection = "systematic"`) or checkerboard structure (`selection = "checkerboard"`).
@@
 
-```{r block-random, cache = TRUE, fig.cap="Spatial blocking with option `selection = 'random'`.", out.width="50%", fig.pos="h"}
-resampling_block_random <- rsmp("spcv_block", range = 1000, folds = 5)
+```{r block-random, cache = TRUE, fig.cap="Spatial blocking with option `selection = 'random'`.", out.width="50%", fig.pos="ht"}
+resampling_block_random = rsmp("spcv_block", range = 1000, folds = 5)
 
 autoplot(resampling_block_random,
-  size = 0.8, fold_id = 1, task = task, crs = 4326,
+  size = 0.8, fold_id = 1, task = task, crs = 32717,
   show_blocks = TRUE, show_labels = TRUE) *
@@
 
-```{r block-systematic, cache=TRUE, fig.cap="Spatial blocking with option `selection = 'systematic'`.", out.width="50%", fig.pos="h"}
-resampling_block_systematic <- rsmp("spcv_block",
+```{r block-systematic, cache=TRUE, fig.cap="Spatial blocking with option `selection = 'systematic'`.", out.width="50%", fig.pos="ht"}
+resampling_block_systematic = rsmp("spcv_block",
   range = 1000, folds = 5,
@@
 autoplot(resampling_block_systematic,
-  size = 0.8, fold_id = 1, task = task, crs = 4326,
+  size = 0.8, fold_id = 1, task = task, crs = 32717,
   show_blocks = TRUE, show_labels = TRUE) *
@@
 
-```{r block-checkerboard, cache=TRUE, fig.cap="Spatial blocking with option `selection = 'checkerboard'`.", out.width="50%", fig.pos="h"}
-resampling_block_checkerboard <- rsmp("spcv_block",
+```{r block-checkerboard, cache=TRUE, fig.cap="Spatial blocking with option `selection = 'checkerboard'`.", out.width="40%", fig.pos="h!"}
+resampling_block_checkerboard = rsmp("spcv_block",
   range = 1000, folds = 5,
@@
 autoplot(resampling_block_checkerboard,
-  size = 0.8, fold_id = 1, task = task, crs = 4326,
+  size = 0.8, fold_id = 1, task = task, crs = 32717,
   show_blocks = TRUE, show_labels = TRUE) *
@@
 ```
+
+\newpage
+
+### Rectangular blocks
+
+Alternatively, arguments `rows` and `cols` can be specified instead of `theRange`.
+This allows to create rectangular blocks which honor the supplied `rows` and `cols` settings.
+
+```{r block-rows-and-cols, cache=TRUE, fig.cap="Spatial blocking with arguments `rows` and `cols`.", out.width="40%", fig.pos="ht"}
+resampling_block_rows_cols = rsmp("spcv_block",
+  rows = 6, cols = 3, folds = 5
+)
+
+autoplot(resampling_block_rows_cols,
+  size = 0.8, fold_id = 1, task = task, crs = 32717,
+  show_blocks = TRUE, show_labels = TRUE) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```
+
+\newpage
 
@@
 
-```{r env-1, fig.cap="Environmental blocking using a single predictor.", cache=TRUE, out.width="50%", fig.pos="h"}
-resampling_env <- rsmp("spcv_env",
+```{r env-1, fig.cap="Environmental blocking using a single predictor.", cache=TRUE, out.width="50%", fig.pos="ht"}
+resampling_env = rsmp("spcv_env",
   features = "distdeforest", folds = 5)
@@
 autoplot(resampling_env,
-  size = 0.8, fold_id = 1, task = task, crs = 4326,
+  size = 0.8, fold_id = 1, task = task, crs = 32717,
   show_blocks = TRUE, show_labels = TRUE) *
@@
 
-```{r env-2, fig.cap="Environmental blocking using multiple predictors.", cache=TRUE, out.width="50%", fig.pos="h"}
-resampling_env_multi <- rsmp("spcv_env",
+```{r env-2, fig.cap="Environmental blocking using multiple predictors.", cache=TRUE, out.width="50%", fig.pos="ht"}
+resampling_env_multi = rsmp("spcv_env",
   features = c("distdeforest", "slope"), folds = 5)
@@
 autoplot(resampling_env_multi,
-  size = 0.8, fold_id = 1, task = task, crs = 4326,
+  size = 0.8, fold_id = 1, task = task, crs = 32717,
   show_blocks = TRUE, show_labels = TRUE) *
@@
 
-```{r coords, fig.cap="Method 'Spatial CV' showing a k-means clustering approach based on coordinates.", cache=TRUE, out.width="50%", fig.pos="h"}
-resampling_coords <- rsmp("spcv_coords", folds = 5)$instantiate(task)
+```{r coords, fig.cap="Method 'Spatial CV' showing a k-means clustering approach based on coordinates.", cache=TRUE, out.width="50%", fig.pos="ht"}
+resampling_coords = rsmp("spcv_coords", folds = 5)$instantiate(task)
 
 autoplot(resampling_coords,
-  size = 0.8, fold_id = 1, task = task, crs = 4326) *
+  size = 0.8, fold_id = 1, task = task, crs = 32717) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
@@
 autoplot(resampling_cstf_time,
-  fold_id = 5, task = task_spt, crs = 4326, plot3D = TRUE
+  fold_id = 5, task = task_spt, crs = 32717, plot3D = TRUE
 )
@@
 ```{r lto, echo=FALSE, out.width="90%", fig.cap="Three-dimensional visualization of method `sptcv\\_cstf` and sub-method Leave-Time-Out (LTO).
-To support showcasing the temporal grouping, five folds and five temporal levels in variable `Date` were used in this example.", fig.pos="h"}
+To support showcasing the temporal grouping, five folds and five temporal levels in variable `Date` were used in this example.", fig.pos="ht"}
 knitr::include_graphics("lto.png")
@@
 autoplot(resampling_cstf_loc,
-  fold_id = 5, task = task_spt, crs = 4326, plot3D = TRUE)
+  fold_id = 5, task = task_spt, crs = 32717, plot3D = TRUE)
 ```
@@
 To support showcasing the spatial grouping, the viewing angle has been adjusted to a view from top.
-Points are stacked by temporal levels and only points from the same location are chosen to form a partition.", fig.pos="h"}
+Points are stacked by temporal levels and only points from the same location are chosen to form a partition.", fig.pos="ht"}
 knitr::include_graphics("llo.png")
@@
 autoplot(resampling_cstf_time_loc,
-  fold_id = 4, task = task_spt, crs = 4326, plot3D = TRUE,
+  fold_id = 4, task = task_spt, crs = 32717, plot3D = TRUE,
   show_omitted = TRUE)
@@
 ```{r llto, echo=FALSE, out.width="90%", fig.cap="Three-dimensional visualization of method 'sptcv\\_cstf' and sub-method 'Leave-Location-And-Time-Out' (LLTO).
-The grey points show the points omitted by this method.", cache=TRUE, fig.pos="h"}
+The grey points show the points omitted by this method.", cache=TRUE, fig.pos="ht"}
 knitr::include_graphics("llto.png")
@@
 
-# List of spatiotemporal resampling methods
+# Case study: Comparing spatial and non-spatial resampling {#sec:case-study}
 
-Table \@ref(tab:sptcv-methods) lists all available spatiotemporal resampling methods available in \pkg{mlr3spatiotempcv}.
-Column `Package` denotes the upstream package from which the method was incorporated.
-The internal identifier for the respective method in \pkg{mlr3spatiotempcv} is shown in column `Notation`.
+To show the effect of spatial autocorrelation on performance estimates in cross-validation, a spatial and non-spatial CV using method `"spatial_coords"` will be shown in the following.
 
-\begin{table}[h]
-	\centering
-	\caption[t]{Spatiotemporal resampling methods, ordered alphabetically by class name (Notation column).}
-	\begingroup
-	\begin{tabular}{llll}
-		\\
-		Literature                  & Package          & Reference          & Notation                    \\
-		\toprule
-    Spatial Buffering           & \pkg{blockCV}    & \citet{blockCV}    & \texttt{rsmp("spcv\_buffer")} \\
-    Spatial Blocking            & \pkg{blockCV}    & \citet{blockCV}    & \texttt{rsmp("spcv\_block")}  \\
-    Spatial CV                  & \pkg{sperrorest} & \citet{sperrorest} & \texttt{rsmp("spcv\_cv")}     \\
-    Environmental Blocking      & \pkg{blockCV}    & \citet{blockCV}    & \texttt{rsmp("spcv\_env")}    \\
-    \midrule
-    Leave-Location-and-Time-Out & \pkg{CAST}       & \citet{cast}       & \texttt{rsmp("spcv\_cstf")}   \\
-    Spatiotemporal Clustering   & \pkg{skmeans}    & \citet{cluto}      & \texttt{rsmp("spcv\_cluto")}
-	\end{tabular}
-	\endgroup\label{tab:sptcv-methods}
-\end{table}
-
-# Comparing Spatial and Non-Spatial Resampling
-
-To show the effect of spatial autocorrelation in modeling, a spatial and non-spatial CV using \pkg{mlr3spatiotempcv} will be shown in the following.
-For simplicity, one of the most used methods (spatial k-means clustering after @sperrorest) will be used.
-Individual methods are discussed in FIXME.
-
-The performance of a simple classification tree (`"classif.ranger"`) is evaluated on a random repeated cross-validation (`"repeated_cv"`) with four folds and two repetitions.
+The performance of a simple classification tree (`"classif.ranger"`) is evaluated on a repeated cross-validation (`"repeated_cv"`) with four folds and two repetitions.
 In reality, more fold (between five and ten) and a higher number of repetitions should be used.
@@
 
-## Data Preparation
+## Data preparation
 
-The following code chunk first loads all necessary R packages and sets a low verbosity for this example.
-A seed is set for deterministic execution.
+At first all necessary R packages are loaded and a lower verbosity is set to keep the output tidy.
+Also a seed is set for deterministic execution.
 
-Next the mlr3 task `"ecuador"`, which is included in \pkg{mlr3spatiotempcv}, is loaded via `tsk()`.
-This function is a convenience wrapper around the underlying R6 construction method for a spatiotemporal classification task (`mlr3spatiotempcv::TaskClassifST$new()`).
+Next the \pkg{mlr3spatiotempcv} built-in task `"ecuador"` is loaded via `tsk()`.
 
@@
 A set of commonly used learners is available in \pkg{mlr3learners} [@mlr3learners], including the random forest implementation after @ranger.
-
-By setting the logging threshold, which can be controlled via the \pkg{lgr} package [@lgr], to `"warn"`, verbosity is reduced to keep the example tidy.
 
@@
 
-task <- tsk("ecuador")
+task = tsk("ecuador")
 
-learner <- lrn("classif.ranger", predict_type = "prob")
+learner = lrn("classif.ranger", predict_type = "prob")
 ```
 
-### Non-Spatial Cross-Validation
+## Non-spatial cross-validation
 
-Here, a four-fold, two times repeated non-spatial resampling structure is constructed and instantianeted via method `$instantiate()`.
+A four-fold, two times repeated non-spatial resampling object is constructed and instantiated via method `$instantiate()`.
 Next, the created resampling object `resampling_nsp` is passed to function `resample()` together with the Task and Learner created earlier to execute the cross-validation.
 
-Last, the results from the CV are aggregated using the measure `"classif.auc"`, which is the technical notation for the Area Under the Receivers Operating Curve (AUROC).
+The results from the CV are then aggregated using the measure `"classif.auc"`, which is the package notation for the Area Under the Receivers Operating Curve (AUROC) measure.
 
 ```{r rr-nspcv, cache=TRUE, cache.lazy = FALSE}
-resampling_nsp <- rsmp("repeated_cv", folds = 4, repeats = 2)
+resampling_nsp = rsmp("repeated_cv", folds = 4, repeats = 2)
 resampling_nsp$instantiate(task)
-rr_nsp <- resample(
+rr_nsp = resample(
   task = task, learner = learner,
@@
 
-## Spatial Cross-Validation
+## Spatial cross-validation
 
-The same approach as for non-spatial resampling is used for spatial resampling, expect the method being `"repeated_spcv_coords"` instead of `"repeated_cv"`.
+The same approach as for non-spatial resampling is used for spatial resampling, expect replacing method `"repeated_cv"` by `"repeated_spcv_coords"`.
 
 ```{r rr-spcv, cache=TRUE, cache.lazy = FALSE}
-resampling_sp <- rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
+resampling_sp = rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
 resampling_sp$instantiate(task)
-rr_sp <- resample(
+rr_sp = resample(
   task = task, learner = learner,
@@
 
-The resulting AUC using Non-Spatial Cross-Validation (NSpCV) is around 0.15 better compared to the Spatial Cross-Validation (SpCV).
+## Summary
+
+The resulting AUC valiue of 0.77 using Non-Spatial Cross-Validation (NSpCV) is around 0.15 better compared to the Spatial Cross-Validation (SpCV) value of 0.61.
 The better performance score results from over-optimism in the estimation with NSpCV.
@@
 
-# Discussion
+The respective folds of the resamplings used in this example are shown in Fig. \@ref(fig:vis-spcv) and Fig. \@ref(fig:vis-nspcv).
 
-## Choosing a Resampling Method
+# Discussion {#sec:disc}
+
+## Choosing a resampling method
 
@@
 
-## A critical evaluation of all methods
+## Comparing spatio-temporal resampling methods
 
-Even though all methods have their pros and cons and no universal method exists, this chapter should provide a critial view on the limitations of each method.
+<!-- high autocor on the border of test sets -->
+<!-- sptcv LLTO: discards many observations -->
+All methods which make use of all variables in the dataset suffer from the problem that on the border between train and test sets, both in the spatial and temporal domain, autocorrelation is quite high.
+Only methods that omit some observations (`"spcv_buffer"`, `"sptcv_cstf"` with both `space_var` and `time_var` set) are able to reduce this problem.
+However, these then suffer from the fact of potentially not being able to properly represent the dataset because only a subset of observations is used.
+
+<!-- sptcv LLO & LTO: requires additional variable to specify groupings -->
+Methods `"spcv_cstf"` (with only one of `space_var` or `time_var` set) require a variable in the dataset which should be used for grouping.
+The specification of such is not always easy because the selection of such a variable also implies that the selected groups inherit substantial autorcorrelation and differ substantially from other groups.
+Also, if the variable contains too many groups, the difference within train/test splits may become undesirably high and tend towards a LOO CV.
+
+<!-- only some methods are suited for repeated execution -->
+<!-- spcv_coords: limited variance due to limited k-means possibilties -->
+For repeated executions of method `spcv_coords`, the variance between repetitions could be quite small due to the limited possibilities of different `k-means` clusters on the same dataset.
+This applies even more to `"spcv_block"` with options `selection = "systematic"` and `selection = "checkerboard"` as the fold assignments are the same in each repetition.
+Only methods with a random component (e.g. `"spcv_coords"` or `"spcv_block"` with option `selction = "random"`) may be suited for using multiple repetitions.
+
+<!-- spcv buffer: discards observations, LOO very expensive -->
+Method `"spcv_buffer"` makes use of the removal of observations to reduce the influence of spatial autocorrelation between train and test set.
+This might be pretty effective in reducing autocorrelation but also requires to discard many variables and comes with the downsides of LOO-CV: many model fits which can be computationally very demanding and a high variance due to the high similarity of training datasets [@blockCV].
 
 <!-- spcv coords: mention generic name -->
-
-<!-- sptcv LLTO: discards many observations -->
-<!-- sptcv LLO: requires additional variable to specify groupings -->
-<!-- sptcv LTO: ignores spatial autocor entirely -->
-
-<!-- spcv buffer: discards observations, LOO very expensive -->
-
-<!-- missing: non-LOO method with buffer support -->
+In @sperrorest, which is commonly used as the main reference for this k-means based spatial clustering approach, the method was not assigned a unique name.
+This brings potential confusion for users and the community as the term "spatial CV" is used a generic one for the idea of accounting for spatial autocorrelation for spatial datasets.
+Hence, in \pkg{mlr3spatiotempcv} this method was named `"spcv_coords"`, making use of the fact that the spatial coordinates are used for a cluster-based partition creation.
 
@@
 <!-- env block: more than one variable may lead to unforeseeable results -->
+Method `"spcv_env"` is tricky in that it requires the user to limit the clustering to (a) specific variable(s).
+This implies that only this/these variable(s) suffer from autocorrelation and others can be ignored.
+However, the influence of a single variable is hard to estimate and the same applies for estimating the influence of the non-accounted variables which might add substantial bias to the CV.
+
+# Summary and outlook
+
+<!-- missing: Guidelines how to choose a resampling method -->
+While various spatio-temporal resampling methods exists nowadays to account for spatio-temporal autocorrelation in modeling, there is still lots of uncertainties which method should be used for which dataset.
+The answer to this questions is not trivial and needs more research.
+One option could be to propose a checklist of dataset characteristics which would make a dataset more suitable for specific methods that could help users making an appropriate and objective decision.
+
+<!-- missing: non-LOO method with buffer support -->
+Also no methods yet exists that makes use of a buffer zone with blocking or clustering, bridging the gap between buffered LOO-CV and fold-based non-buffer methods (`"spcv_coords"` and `"spcv_block"`).
+
+We believe that making resampling methods available to widely used modeling frameworks is more benefitial than implementing such methods standalone in an R package.
+The latter often require special structured objects as inputs and cannot directly be re-used in susequent modeling steps of user workflows.
+This is not an issue specific to spatio-temporal resampling methods but a general philosophy of the authors.
+
+Spatio-temporal cross-validation is still not yet a paradigm that is completely established in scientfic workflows even though it is discussed intensively for more than a decade now.
+Making most available methods easily accessible to users is an important step to foster the acceptance of such in the community and lead to less biased cross-validation results in environmental and ecological studies.
 

### 2021-04-04 paper/paper.Rmd
refinements

@@
 
-```{r setup, echo=FALSE}
+```{r setup, echo=FALSE, purl=FALSE}
 # important for correct figure placement
@@
 # knitr::opts_chunk$set(cache = FALSE)
+```
+
+```{r setup-2, echo=FALSE, purl=TRUE}
+# for knitr spint later
+knitr::opts_chunk$set(fig.path = "./paper/figs/")
 ```
@@
 
+In addition, \pkg{mlr3spatiotempcv} adds 2D and 3D visualization possibilities for all implemented methods.
+The authors believe that the possibility for creating graphical representations of partitions is important, especially if spatio-temporal clustering or other grouping types were used to create the groups.
+
 ## The mlr3 ecosystem
@@
 
-Every partitioning method in [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) comes with a generic `plot()` method to visualize the created groups.
-In a 2D space this happens via [ggplot2](https://ggplot2.tidyverse.org) [@ggplot2] while for spatiotemporal methods 3D visualizations are created via [plotly](https://github.com/ropensci/plotly) [@plotly].
+Every partitioning method in [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) comes with generic `plot()` and `autoplot()` (\pkg{ggplot2} equivalent) method to visualize the created groups.
+In a 2D space this happens via package [ggplot2](https://ggplot2.tidyverse.org) [@ggplot2] while for spatiotemporal methods 3D visualizations are created via package [plotly](https://github.com/ropensci/plotly) [@plotly].
 
 Visualization of partitioning methods is a vital part of \pkg{mlr3spatiotempcv} as most upstream packages lack support for such.
-The semantics are presented first because plotting will be used extensively when presenting the integrated methods.
+Hence, plotting semantics are presented first because plotting will be used extensively when presenting the integrated methods.
 
-Unless specified by the user, the coordinate reference system (CRS) defaults to EPSG code 32717 (WGS84).
+Unless specified by the user, the coordinate reference system (CRS) defaults to EPSG code 4326 (WGS84).
 This is because a lat/lon based CRS is better suited for plotting purposes than a Mercator (UTM) one due tho a more descriptive axes labeling.
 Setting an appropriate CRS for the given data *during task construction* is very important.
-Even though EPSG 32717 is a good fallback and often used for visualization purposes, spatial offsets of up to multiple meters may occur if the wrong CRS was passed initially.
+Even though EPSG 4326 is a good fallback and often used for visualization purposes, spatial offsets of up to multiple meters may occur if the wrong CRS was passed initially.
 
 The following example creates a spatiotemporal task from scratch.
-The data used here is also available as a built-in task via `tsk("ecuador")`.
+This task is also available as a built-in task via `tsk("ecuador")`.
 
 Because the spatial grouping of the k-means based approach (`"spcv_coords"`) (Fig.
-\@ref(fig:vis-spcv)) contrasts visually very well to the NSpCV (random) partitioning (Fig.
+\@ref(fig:vis-spcv)) contrasts visually very well to NSpCV (random) partitioning (Fig.
 \@ref(fig:vis-nspcv)), this method was chosen to showcase the 2D plotting capabilities of \pkg{mlr3spatiotempcv}.
 
-The partitions shown in the following are also used in the case study in section \@ref(sec:case-study).
+The partition plots shown in the following are also used in the case study in section \@ref(sec:case-study).
 
@@
 At the time of writing, the following spatiotemporal resampling methods (Tab. \@ref(tab:sptcv-methods)) were available across different R packages.
-This list might increase over time and it is recommended to consult the homepage of \pkg{mlr3spatiotempcv} for an up-to-date list.
-The internal identifier for the respective method in \pkg{mlr3spatiotempcv} is shown in column `Notation`.
+This list might increase over time and it is recommended to check the homepage of \pkg{mlr3spatiotempcv} for an up-to-date list.
+In Tab. \@ref(tab:sptcv-methods), the internal identifier for the respective method in \pkg{mlr3spatiotempcv} is shown in column "Notation".
+Some of the methods listed are able to produce different partitioning results by setting different options, hence the actual number of available methods is higher than the one shown in Tab.
+\@ref(tab:sptcv-methods).
 
@@
 
-autoplot(resampling_buffer,
+plot(resampling_buffer,
   size = 0.8, task = task, fold_id = 1, crs = 32717) *
@@
 
-autoplot(resampling_block_random,
+plot(resampling_block_random,
   size = 0.8, fold_id = 1, task = task, crs = 32717,
@@
 
-autoplot(resampling_block_systematic,
+plot(resampling_block_systematic,
   size = 0.8, fold_id = 1, task = task, crs = 32717,
@@
 
-autoplot(resampling_block_checkerboard,
+plot(resampling_block_checkerboard,
   size = 0.8, fold_id = 1, task = task, crs = 32717,
@@
 
-autoplot(resampling_block_rows_cols,
+plot(resampling_block_rows_cols,
   size = 0.8, fold_id = 1, task = task, crs = 32717,
@@
 
-autoplot(resampling_env,
+plot(resampling_env,
   size = 0.8, fold_id = 1, task = task, crs = 32717,
@@
 
-autoplot(resampling_env_multi,
+plot(resampling_env_multi,
   size = 0.8, fold_id = 1, task = task, crs = 32717,
@@
 
-autoplot(resampling_coords,
+plot(resampling_coords,
   size = 0.8, fold_id = 1, task = task, crs = 32717) *
@@
 
-autoplot(resampling_cstf_time,
+plot(resampling_cstf_time,
   fold_id = 5, task = task_spt, crs = 32717, plot3D = TRUE
@@
 
-autoplot(resampling_cstf_loc,
+plot(resampling_cstf_loc,
   fold_id = 5, task = task_spt, crs = 32717, plot3D = TRUE)
@@
 
-autoplot(resampling_cstf_time_loc,
+plot(resampling_cstf_time_loc,
   fold_id = 4, task = task_spt, crs = 32717, plot3D = TRUE,
@@
 <!-- env block: more than one variable may lead to unforeseeable results -->
-Method `"spcv_env"` is tricky in that it requires the user to limit the clustering to (a) specific variable(s).
+Method `"spcv_env"` is complicated in that it requires the user to limit the clustering to (a) specific variable(s).
 This implies that only this/these variable(s) suffer from autocorrelation and others can be ignored.
@@
 
+```{r echo=FALSE}
+sessionInfo()
+```
+
 # References

### 2021-04-10 paper/paper.Rmd
expand and rotate methods table

@@
   \usepackage{longtable}
+  \usepackage{adjustbox}
+  \usepackage{multirow}
+  \usepackage{tabularx}
+  \usepackage{placeins}
 tables: true
 bibliography: paper.bib
-# output: rticles::jss_article
 output:
@@
   \centering
-	\caption[t]{Spatiotemporal resampling methods, ordered alphabetically by class name (Notation column).}
+  \caption[t]{Spatiotemporal resampling methods, ordered alphabetically by class name (Notation column). }
   \begingroup
-	\begin{tabular}{llll}
+  \begin{adjustbox}{angle=270}
+    \begin{tabular}{lllll}
       \\
-		Literature                  & Package          & Reference          & Notation                    \\
+      Type                            & Name                        & R Package(s)     & mlr3 Notation  & Literature \\
       \toprule
-    Spatial Buffering           & \pkg{blockCV}    & \citet{blockCV}    & \texttt{rsmp("spcv\_buffer")} \\
-    Spatial Blocking            & \pkg{blockCV}    & \citet{blockCV}    & \texttt{rsmp("spcv\_block")}  \\
-    Spatial CV                  & \pkg{sperrorest} & \citet{sperrorest} & \texttt{rsmp("spcv\_cv")}     \\
-    Environmental Blocking      & \pkg{blockCV}    & \citet{blockCV}    & \texttt{rsmp("spcv\_env")}    \\
+      \multirow{3}{*}{Spatial}        & Spatial Buffering           & \pkg{blockCV}, \pkg{sperrorest}    & "spcv\_buffer", "spcv\_disc()", "spcv\_loo()" & Lit1       \\
+                                      & Spatial Blocking            & \pkg{blockCV}, \pkg{sperrorest}  & "spcv\_block",
+                                      "spcv\_tiles()"  & Lit1       \\
+                                      & Spatial Clustering          & \pkg{sperrorest} & "spcv\_coords" & Lit1       \\
+       & Leave-Location-Out & \pkg{CAST}, \pkg{sperrorest}       & "sptcv\_cstf('space\_var')", "spcv\_factor"  & Lit1       \\
       \midrule
-    Leave-Location-and-Time-Out & \pkg{CAST}       & \citet{cast}       & \texttt{rsmp("spcv\_cstf")}   \\
-    Spatiotemporal Clustering   & \pkg{skmeans}    & \citet{cluto}      & \texttt{rsmp("spcv\_cluto")}
+      \multirow{2}{*}{Spatiotemporal} & Leave-Location-and-Time-Out & \pkg{CAST}       & "sptcv\_cstf"  & Lit1       \\
+                                      & Spatiotemporal Clustering   & \pkg{skmeans}    & "sptcv\_cluto" & Lit1       \\
+      \midrule
+      Feature space                   & Environmental Blocking      & \pkg{blockCV}    & "spcv\_env"    & Lit1       \\
     \end{tabular}
+  \end{adjustbox}
   \endgroup\label{tab:sptcv-methods}

### 2021-04-19 R/ResamplingSpCVTiles.R
- Add `spcv_tiles` and `repeated_spcv_tiles` (#121)

@@ -1,0 +1,304 @@
+#' @title (sperrorest) Spatial "Tiles" resampling
+#'
+#' @template rox_spcv_tiles
+#'
+#' @references
+#' `r format_bib("brenning2012")`
+#'
+#' @export
+#' @examples
+#' if (mlr3misc::require_namespaces("sperrorest", quietly = TRUE)) {
+#'   library(mlr3)
+#'   task = tsk("ecuador")
+#'
+#'   # Instantiate Resampling
+#'   rcv = rsmp("spcv_tiles", nsplit = c(4L, 3L), reassign = FALSE)
+#'   rcv$instantiate(task)
+#'
+#'   # Individual sets:
+#'   rcv$train_set(1)
+#'   rcv$test_set(1)
+#'   # check that no obs are in both sets
+#'   intersect(rcv$train_set(1), rcv$test_set(1)) # good!
+#'
+#'   # Internal storage:
+#'   rcv$instance # table
+#' }
+ResamplingSpCVTiles = R6Class("ResamplingSpCVTiles",
+  inherit = mlr3::Resampling,
+  public = list(
+    #' @description
+    #' Create a "Spatial 'Tiles' resampling" resampling instance.
+    #' @param id `character(1)`\cr
+    #'   Identifier for the resampling strategy.
+    #' For a list of available arguments, please see
+    #' [sperrorest::partition_tiles].
+    initialize = function(id = "spcv_Tiles") {
+      ps = ParamSet$new(params = list(
+        ParamUty$new("dsplit", custom_check = function(x) checkmate::assert_integer(x, len = 2)),
+        ParamUty$new("nsplit", custom_check = function(x) checkmate::assert_integer(x, len = 2)),
+        ParamFct$new("rotation", levels = c("none", "random", "user"), default = "none"),
+        ParamUty$new("user_rotation", default = NULL),
+        ParamFct$new("offset", levels = c("none", "random", "user"), default = "none"),
+        ParamUty$new("user_offset", default = NULL),
+        ParamLgl$new("reassign", default = TRUE),
+        ParamDbl$new("min_frac", default = 0.025, lower = 0, upper = 1),
+        ParamInt$new("min_n", default = 5, lower = 0),
+        ParamInt$new("iterate", default = 1, lower = 1)
+      ))
+      super$initialize(
+        id = id,
+        param_set = ps
+      )
+    },
+
+    #' @description
+    #'  Materializes fixed training and test splits for a given task.
+    #' @param task [Task]\cr
+    #'  A task to instantiate.
[…]

### 2021-04-21 paper/ABnotes.Rmd
quick write-up of proposed new intro

---
title: "Alex's Notes and Text Fragments for the JOSS Manuscript"
author: "Alexander Brenning"
date: "21 Apr 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Proposed New Introduction

*This is just a quick write-up, to be revised with more careful attention to referencing.*

Spatial and spatio-temporal prediction tasks are common in applications ranging from environmental sciences to archaeology and epidemiology. While sophisticated mathematical frameworks have long been developed in spatial statistics to characterize predictive uncertainties under well-defined mathematical assumptions such as intrinsic stationarity [e.g., cressie], computational estimation procedures have only been proposed more recently to assess predictive performances of spatial and spatio-temporal prediction models [cite brenning.2005; brenning.2012 and others].

Although alternatives such as the bootstrap exist and have some advantages [ref maybe Efron and Gong, or David Hand], cross-validation is a particularly well-established, easy-to-implement algorithm for the estimation of predictive performances of supervised regression and classification models [ref Efron and Gong and next section]. In its basic form, it is based on resampling the data without paying attention to any possible dependence structure, which may arise from, e.g., cluster sampling [ref], grouped data [brenning.lausen.2008; pena.brenning.20xx], or underlying environmental processes inducing some sort of spatial coherence or mixing [ref]. In treating dependent observations as independent, or ignoring autocorrelation, cross-validation test samples may in fact be heavily correlated with, or even pseudo-replicates of, the data used for training the model, which introduces a potentially severe bias, especially for flexible machine-learning models.

This cross-validation bias is well-known in spatial as well as non-spatial prediction [brenning.lausen.2008; ref]. It is most easily understood from a predictive modelling perspective by focusing on the question where (and when) the model should be used for prediction. In crop classification from remotely-sensed data, for instance, learning samples routinely contain multiple grid cells from a sample of fields with known crop type, e.g. 2000 grid cells from 100 fields scattered across a large study region. The purpose of training a model on this sample is to make predictions on other, new *fields* -- not *within* the same field, which obviously presents only one crop type. It would therefore be foolish to train a model on a simple random subsample of grid cells, and test it on the remaining data, i.e. using other grid cells from the same fields, as if we wanted to predict within a field; the results from this performance assessment would be over-optimitic, and perhaps badly so. To mimic the predictive situation for which the model is trained, we'd rather have to resample at the level of fields, not grid cells [pena.brenning.20xx]. If the model was to be applied to adjacent agricultural districts, it would even seem necessary to resample at a higher level of spatial aggregation, i.e. at the level of agricultural districts within the learning sample, in order to realistically mimic the actual prediction task. The cross-validation resampling needed therefore depends as much on the prediction task as on the data structure or dependency at hand.

While it is not the purpose of this paper to recommend specific resampling schemes for specific use cases, this example may suffice to motivate the use of spatial and spatio-temporal cross-validation techniques, and the need for a unified framework and computational toolbox that accommodate a variety of prediction tasks. `mlr3spatiotemporal` is such a toolbox.

This toolbox, implemented as an open-source R package, builds upon and generalizes several existing toolboxes that have been developed in recent years for more specific settings. The earliest and most general of these implementations is the `sperrorest` package [brenning2012], which has, however, not been integrated into established machine-learning frameworks such as `mlr` or `caret`, and lacks support of temporal prediction tasks. Other more recent implementations include `blockCV` for (what for?) [ref Valavi; Rest et al., 2014?] and `CAST` for spatio-temporal prediction tasks in a `caret` framework (correct?) [ref Meyer]. And `skmeans`. Reference to Table 1.

Thus, `mlr3spatiotemporal` implements for the first time a comprehensive state-of-the-art compilation of spatial and spatio-temporal partitioning schemes that is well-integrated into a major machine-learning framework in R, the `mlr3` ecosystem (Lang et al., 2019). This package is furthermore equipped with a variety of two- and three-dimensional diagram types. Our hope is that implementation will facilitate reproducible geospatial modelling and code-sharing across a broad range of application domains.

## Proposed section 2: Spatial and Spatio-temporal Cross-validation

the formal stuff, next on my list...

## General Comments

- Avoid the concept of SAC, TAC and STAC because it lacks a clear definition (in this paper, and a definition is not at all straightforward), and the link between STAC and a specific type of CV is far from obvious. Instead, refer to spatial, temporal, and spatio-temporal prediction tasks.

- I.e. the question is not "what's the structure of my data", but "what's the structure of the prediction problem". Perhaps this recommendation can be included in the Discussion.

### 2021-04-22 paper/ABnotes.Rmd
added references, paragraph on LOO-CV

@@
 Spatial and spatio-temporal prediction tasks are common in applications ranging from environmental sciences to archaeology and epidemiology.
-While sophisticated mathematical frameworks have long been developed in spatial statistics to characterize predictive uncertainties under well-defined mathematical assumptions such as intrinsic stationarity [e.g., cressie], computational estimation procedures have only been proposed more recently to assess predictive performances of spatial and spatio-temporal prediction models [cite brenning.2005; brenning.2012 and others].
+While sophisticated mathematical frameworks have long been developed in spatial statistics to characterize predictive uncertainties under well-defined mathematical assumptions such as intrinsic stationarity [e.g., cressie], computational estimation procedures have only been proposed more recently to assess predictive performances of spatial and spatio-temporal prediction models [@brenning.2005.nhess; @sperrorest; @pohjankukka2017; @roberts2017].
 
-Although alternatives such as the bootstrap exist and have some advantages [ref maybe Efron and Gong, or David Hand], cross-validation is a particularly well-established, easy-to-implement algorithm for the estimation of predictive performances of supervised regression and classification models [ref Efron and Gong and next section].
-In its basic form, it is based on resampling the data without paying attention to any possible dependence structure, which may arise from, e.g., cluster sampling [ref], grouped data [brenning.lausen.2008; pena.brenning.20xx], or underlying environmental processes inducing some sort of spatial coherence or mixing [ref].
-In treating dependent observations as independent, or ignoring autocorrelation, cross-validation test samples may in fact be heavily correlated with, or even pseudo-replicates of, the data used for training the model, which introduces a potentially severe bias, especially for flexible machine-learning models.
+Although alternatives such as the bootstrap exist and have some advantages [ref maybe Efron and Gong, or David Hand], cross-validation is a particularly well-established, easy-to-implement algorithm for the estimation of predictive performances of supervised regression and classification models [@efron.gong.1983 and next section].
+In its basic form, it is based on resampling the data without paying attention to any possible dependence structure, which may arise from, e.g., cluster sampling [ref], grouped data [@brenning.lausen.2008; @pena.brenning.2015.maipo], or underlying environmental processes inducing some sort of spatial coherence at the landscape scale [@brenning.2005.nhess].
+In treating dependent observations as independent, or ignoring autocorrelation, cross-validation test samples may in fact be heavily correlated with, or even pseudo-replicates of, the data used for training the model, which introduces a potentially severe bias, especially for flexible machine-learning models [@brenning.2005.nhess].
 
-This cross-validation bias is well-known in spatial as well as non-spatial prediction [brenning.lausen.2008; ref].
+This cross-validation bias is well-known in spatial as well as non-spatial prediction [@brenning.2005.nhess; @brenning.lausen.2008].
 It is most easily understood from a predictive modelling perspective by focusing on the question where (and when) the model should be used for prediction.
@@
 2000 grid cells from 100 fields scattered across a large study region.
-The purpose of training a model on this sample is to make predictions on other, new *fields* -- not *within* the same field, which obviously presents only one crop type.
+The purpose of training a model on this sample is to make predictions on other, new *fields* within the same geographic domain [*intra-domain* prediction; @brenning.2005.nhess] -- not *within* the same field, which obviously presents only one crop type.
 It would therefore be foolish to train a model on a simple random subsample of grid cells, and test it on the remaining data, i.e.
 using other grid cells from the same fields, as if we wanted to predict within a field; the results from this performance assessment would be over-optimitic, and perhaps badly so.
-To mimic the predictive situation for which the model is trained, we'd rather have to resample at the level of fields, not grid cells [pena.brenning.20xx].
-If the model was to be applied to adjacent agricultural districts, it would even seem necessary to resample at a higher level of spatial aggregation, i.e.
+To mimic the predictive situation for which the model is trained, we'd rather have to resample at the level of fields, not grid cells [@pena.brenning.2015.maipo].
+If the model was to be applied to adjacent agricultural districts, i.e.
+outside the learning sample's spatial domain [*extra-domain* prediction; @brenning.2005.nhess], it would even seem necessary to resample at a higher level of spatial aggregation, i.e.
 at the level of agricultural districts within the learning sample, in order to realistically mimic the actual prediction task.
@@
 
-While it is not the purpose of this paper to recommend specific resampling schemes for specific use cases, this example may suffice to motivate the use of spatial and spatio-temporal cross-validation techniques, and the need for a unified framework and computational toolbox that accommodate a variety of prediction tasks.
+While it is not the purpose of this paper to recommend specific resampling schemes for specific use cases, this example may suffice to motivate the use of spatial and spatio-temporal cross-validation techniques, and the need for a unified framework and computational toolbox that accommodate a variety of prediction tasks that may be applicable to a broad range of prediction tasks.
 `mlr3spatiotemporal` is such a toolbox.
@@
 This toolbox, implemented as an open-source R package, builds upon and generalizes several existing toolboxes that have been developed in recent years for more specific settings.
-The earliest and most general of these implementations is the `sperrorest` package [brenning2012], which has, however, not been integrated into established machine-learning frameworks such as `mlr` or `caret`, and lacks support of temporal prediction tasks.
-Other more recent implementations include `blockCV` for (what for?) [ref Valavi; Rest et al., 2014?] and `CAST` for spatio-temporal prediction tasks in a `caret` framework (correct?) [ref Meyer].
-And `skmeans`.
-Reference to Table 1.
+The earliest of these implementations is the `sperrorest` package [@sperrorest], which has, however, not been integrated into established machine-learning frameworks such as `mlr` or `caret`, and lacks support of temporal prediction tasks.
+Other more recent implementations include `blockCV` for (what for?) [ref @blockCV; @rest2014] and `CAST` for spatio-temporal prediction tasks in a `caret` framework (correct?) [@cast; @meyer2018a].
+And `skmeans` [ref].
+Reference to Table 1 FIXME.
 
-Thus, `mlr3spatiotemporal` implements for the first time a comprehensive state-of-the-art compilation of spatial and spatio-temporal partitioning schemes that is well-integrated into a major machine-learning framework in R, the `mlr3` ecosystem (Lang et al., 2019).
+Thus, `mlr3spatiotemporal` implements for the first time a comprehensive state-of-the-art compilation of spatial and spatio-temporal partitioning schemes that is well-integrated into a major machine-learning framework in R, the `mlr3` ecosystem [@mlr3].
 This package is furthermore equipped with a variety of two- and three-dimensional diagram types.
@@
 
+etc
+
+A variant of cross-validation is leave-one-out (LOO) cross-validation , which has long been established in geostatistics [cressie], sometimes with a focus on the spatial distribution of LOO error [@willmott.matsuura.2006].
+Since this is just a special case of non-spatial CV with $k=n$, it is sometimes also referred to as spatial CV [@willmott.matsuura.2006].
+Nevertheless, spatial variants have been proposed that apply an exclusion buffer to the test locations to separate them from the training data [@brenning.2005.nhess].
+One plausible approach for this is to use the range of autocorrelation of model residuals to determine the buffer distance as this ensures independence conditional on the predictors [@brenning.2005.nhess].
+When data is grouped, e.g., due to multi-level sampling designs or the study of spatial objects, LOO at the site level has been proposed [@martin.et.al.2008; @kasurak.et.al.2011]
+
 ## General Comments

### 2021-04-22 paper/ABnotes.Rmd
added refs: survey, LOO

@@
 
-Although alternatives such as the bootstrap exist and have some advantages [ref maybe Efron and Gong, or David Hand], cross-validation is a particularly well-established, easy-to-implement algorithm for the estimation of predictive performances of supervised regression and classification models [@efron.gong.1983 and next section].
+Although alternatives such as the bootstrap exist and have some advantages [@efron.gong.1983, or David Hand], cross-validation is a particularly well-established, easy-to-implement algorithm for the estimation of predictive performances of supervised machine-learning models [@efron.gong.1983 and next section] and statistical model selection [@arlot.celisse.2010].
 In its basic form, it is based on resampling the data without paying attention to any possible dependence structure, which may arise from, e.g., cluster sampling [ref], grouped data [@brenning.lausen.2008; @pena.brenning.2015.maipo], or underlying environmental processes inducing some sort of spatial coherence at the landscape scale [@brenning.2005.nhess].
@@
 
-This cross-validation bias is well-known in spatial as well as non-spatial prediction [@brenning.2005.nhess; @brenning.lausen.2008].
+This cross-validation bias is well-known in spatial as well as non-spatial prediction [@brenning.2005.nhess; @brenning.lausen.2008; @arlot.celisse.2010].
 It is most easily understood from a predictive modelling perspective by focusing on the question where (and when) the model should be used for prediction.
@@
 One plausible approach for this is to use the range of autocorrelation of model residuals to determine the buffer distance as this ensures independence conditional on the predictors [@brenning.2005.nhess].
-When data is grouped, e.g., due to multi-level sampling designs or the study of spatial objects, LOO at the site level has been proposed [@martin.et.al.2008; @kasurak.et.al.2011]
+When data is grouped, e.g., due to multi-level sampling designs or the study of spatial objects, LOO at has been proposed at the site level [@martin.et.al.2008; @kasurak.et.al.2011] or, in animal movement studies, at the animal level [@anderson.et.al.2005].
+Similarly, in the temporal domain, it has been proposed to leave out temporal observational units, as in leave-one-year-out CV [@anderson.et.al.2005; @brenning.2005.nhess].
 

### 2021-04-22 paper/paper.Rmd
keywords

@@
   # at least one keyword must be supplied
-  formatted: [resampling, machine learning, autocorrelation, spatial, temporal, "\\proglang{R}"]
-  plain:     [resampling, machine learning, autocorrelation, spatial, temporal, R]
+  formatted: [cross-validation, predictive performance, machine learning, autocorrelation, spatial, temporal, "\\proglang{R}"]
+  plain:     [cross-validation, predictive performance, autocorrelation, spatial, temporal, R]
 preamble: >

### 2021-04-22 paper/ABnotes.Rmd
added willmott reference

@@
 output: html_document
+bibliography: paper.bib
 ---
@@
 ## Proposed New Introduction
-
-*This is just a quick write-up, to be revised with more careful attention to referencing.*
 
@@
 - I.e. the question is not "what's the structure of my data", but "what's the structure of the prediction problem". Perhaps this recommendation can be included in the Discussion.
+
+## References

### 2021-04-22 paper/ABnotes.Rmd
wrote formal intro to CV

@@
 Spatial and spatio-temporal prediction tasks are common in applications ranging from environmental sciences to archaeology and epidemiology.
-While sophisticated mathematical frameworks have long been developed in spatial statistics to characterize predictive uncertainties under well-defined mathematical assumptions such as intrinsic stationarity [e.g., cressie], computational estimation procedures have only been proposed more recently to assess predictive performances of spatial and spatio-temporal prediction models [@brenning.2005.nhess; @sperrorest; @pohjankukka2017; @roberts2017].
+While sophisticated mathematical frameworks have long been developed in spatial statistics to characterize predictive uncertainties under well-defined mathematical assumptions such as intrinsic stationarity [e.g., cressie.2015], computational estimation procedures have only been proposed more recently to assess predictive performances of spatial and spatio-temporal prediction models [@brenning.2005.nhess; @sperrorest; @pohjankukka2017; @roberts2017].
 
-Although alternatives such as the bootstrap exist and have some advantages [@efron.gong.1983, or David Hand], cross-validation is a particularly well-established, easy-to-implement algorithm for the estimation of predictive performances of supervised machine-learning models [@efron.gong.1983 and next section] and statistical model selection [@arlot.celisse.2010].
+Although alternatives such as the bootstrap exist and have some advantages [@efron.gong.1983, or David Hand], cross-validation is a particularly well-established, easy-to-implement algorithm for *model assessment* of supervised machine-learning models [@efron.gong.1983 and next section] and *model selection* [@arlot.celisse.2010].
 In its basic form, it is based on resampling the data without paying attention to any possible dependence structure, which may arise from, e.g., cluster sampling [ref], grouped data [@brenning.lausen.2008; @pena.brenning.2015.maipo], or underlying environmental processes inducing some sort of spatial coherence at the landscape scale [@brenning.2005.nhess].
@@
 While it is not the purpose of this paper to recommend specific resampling schemes for specific use cases, this example may suffice to motivate the use of spatial and spatio-temporal cross-validation techniques, and the need for a unified framework and computational toolbox that accommodate a variety of prediction tasks that may be applicable to a broad range of prediction tasks.
-`mlr3spatiotemporal` is such a toolbox.
+`mlr3spatiotemporalcv` is such a toolbox.
 
@@
 
-Thus, `mlr3spatiotemporal` implements for the first time a comprehensive state-of-the-art compilation of spatial and spatio-temporal partitioning schemes that is well-integrated into a major machine-learning framework in R, the `mlr3` ecosystem [@mlr3].
+Thus, `mlr3spatiotemporalcv` implements for the first time a comprehensive state-of-the-art compilation of spatial and spatio-temporal partitioning schemes that is well-integrated into a major machine-learning framework in R, the `mlr3` ecosystem [@mlr3].
 This package is furthermore equipped with a variety of two- and three-dimensional diagram types.
@@
 
+In cross-validation for predictive model assessment, we consider the following formal setting.
+We are interested in predicting a numerical or categorical response $y$ of an object or instance using a feature vector $\mathbf{x} = (x^{(1)}, \ldots, x^{(p)}\in\mathbb{R}$ and a model $\hat{f}_L$ that has been trained on a learning sample $L = \{(y_i, \mathbf{x}_i),\ i = 1, ..., n\}$.
+Our goal is to estimate the expected value of the performance of $\hat{f}_L$,
+\[
+\mathit{perf(\hat{f}_L)} := E(l(Y,\hat{f}_L(X))),
+\]
+where $l$ is a real-valued loss function, and the expected value is with respect the probability distribution of $X$, the features of an instance $(Y,X)$ drawn randomly from the underlying population.
+This is referred to as the *actual* or *conditional* performance measure as it is conditional on $L$ [@hand.1996].
+The loss function can take a variety of forms such as the misclassification error $I(Y\neq\hat{f}_L(X))$ in classification, or the squared error $(Y-\hat{f}_L(X))^2$ in regression, among many other possible measures.
+The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance metrics for regression and classification [see e.g., @hand.1996 for classification, and ref??
+for regression tasks].
+
+Nevertheless, since we only have a sample $T$ of test data drawn from the population, we can realistically only *estimate* the conditional performance of $\hat{f}_L$:
+\[
+\widehat{\mathit{perf}}_T(\hat{f}_L) = \frac{1}{|T|}\sum_{i=1}^{|T|}l(Y,\hat{f}_L(X)).
+\]
+This representation as a point estimator for $\mathit{perf(\hat{f}_L)}$ underlines the importance of using a random sample for performance estimation to avoid estimation bias.
+Other estimators than the simple mean may be required when $T$ is not a simple random sample, for instance a stratified random sample [e.g., @thompson.2012].
+As always, judgment sampling may lead to uncontrollable bias.
+
+Since the re-use of the learning sample $L$ for testing, $T=L$, would result in the over-optimistic *resubstitution* or *apparent* model performance, cross-validation partitions the sample $L$ into disjoint training and test sets.
+Specifically, $L$ is split into $k$ partitions,
+\[
+L = L_1 \cup \ldots \cup L_k,\qquad L_i\cap L_j = \emptyset\quad \textrm{for all}\ i\neq j,
+\]
+and a model $\hat{f}_{(i)}$ is fitted on $L_{(i)} := L\setminus L_i$, while $L_i$ is withheld for testing.
+This is repeated for $i=1,\ldots,k$ in order to effectively use the entire sample for testing, while keeping training and test sets disjoint at all times.
+The $k$-fold CV estimator can therefore be written as
+\[
+\widehat{\mathit{perf}}_{L, CV}(f) := \frac{1}{k}\sum_{i=1}^k\widehat{\mathit{perf}}_{L_i}(\hat{f}_{L_{(i)}}),
+\]
+where $f$ is a machine-learning algorithm, i.e.
+a mapping that trains a model $\hat{f}_S$ using any suitable training sample $S$.
+The use of $k=5$ or $k=10$ folds is most commonly seen in practice, and these preferences are also supported by theory [ref??
+where did I get this from?].
+The $k$-fold CV estimator of model performance is a nearly unbiased estimator of the conditional performance measure when the observations were drawn independently [@efron.gong.1983].
+Since $\widehat{\mathit{perf}}_{L, CV}(f)$ still depends on the particular partitioning chosen for $L$, it is sometimes recommended to repeat the estimation using different random partitionings ($r$-repeated $k$-fold cross-validation, e.g.
+$k=100$; ref??).
+
+In traditional cross-validation, the partitioning is based on uniform random sampling, which ignores spatial or temporal autocorrelation or any existing grouping structure as well as the structure of the prediction task, as outlined in the Introduction, and may result in over-optimistic performance estimates.
+Several approaches have therefore been proposed in the literature and implemented in software to accommodate a variety of predictive situations (ref Table 1 FIXME).
+
 etc
 
-A variant of cross-validation is leave-one-out (LOO) cross-validation , which has long been established in geostatistics [cressie], sometimes with a focus on the spatial distribution of LOO error [@willmott.matsuura.2006].
+A variant of cross-validation is leave-one-out (LOO) cross-validation , which has long been established in geostatistics [@cressie.2015], sometimes with a focus on the spatial distribution of LOO error [@willmott.matsuura.2006].
 Since this is just a special case of non-spatial CV with $k=n$, it is sometimes also referred to as spatial CV [@willmott.matsuura.2006].
@@
 Similarly, in the temporal domain, it has been proposed to leave out temporal observational units, as in leave-one-year-out CV [@anderson.et.al.2005; @brenning.2005.nhess].
+
+It should be noted that $k$-fold cross-validation with a large value of $k$, and LOO-CV in particular, are not only very time-consuming since the model has to be trained $k$ times; these models will also be nearly identical since only a tiny fraction of the data is withheld, and therefore estimation bias increases.
+'Pure' LOO-CV is therefore not recommended for machine-learning model assessment [ref??].
 

### 2021-04-22 paper/paper.Rmd
list of authors

@@
     address: |
-      | Faculty of Chemistry and Earth Sciences
       | Department of Geography
-      | GIScience group
+      | Geographic Information Science group
     email: \email{patrick.schratz@uni-jena.de}
@@
       | Faculty of Statistics
+  - name: Alexander Brenning
+    affiliation: 'Friedrich Schiller University Jena'
+    address: |
+      | Department of Geography
+      | Geographic Information Science group
+    email: \email{alexander.brenning@uni-jena.de}
 title:

### 2021-05-02 paper/paper.Rmd
restructure partitioning methods section

@@
 
-```{r create task, cache=TRUE}
+```{r create task, cache=F}
 library(mlr3)
@@
 
+## Spatial
+
+### Buffering
+
 At the time of writing, the following spatiotemporal resampling methods (Tab. \@ref(tab:sptcv-methods)) were available across different R packages.
@@
   \begingroup
-  \begin{adjustbox}{angle=270}
+  <!-- \begin{adjustbox}{angle=270} -->
     \begin{tabular}{lllll}
@@
     \end{tabular}
-  \end{adjustbox}
+  <!-- \end{adjustbox} -->
   \endgroup\label{tab:sptcv-methods}
@@
 
-## Spatial buffering
+#### Spatial buffering (blockCV)
 
@@
 
-## Spatial blocking
+#### Spatial disc (sperrorest)
+
+### Blocking
+
+#### Spatial blocking (blockCV)
 
@@
 
-### Option: selection = "random"
+##### Option: selection = "random"
 
@@
 
-### Option: selection = "systematic"
+##### Option: selection = "systematic"
 
@@
 
-### Option: selection = "checkerboard"
+##### Option: selection = "checkerboard"
 
@@
 
-### Rectangular blocks
+##### Rectangular blocks
 
@@
 
-## Environmental Blocking
+#### Spatial tiling (sperrorest)
 
-The last method from the \pkg{blockCV} package, "environmental blocking", makes use of *k-means* clustering [@hartigan1979a] in a possibly multivariate space.
-The user can specify a variable via argument `feature` which levels will be used to create clusters.
-Hereby, k-means will use Euclidean distance for cluster creation.
-Hence, the selected input variables should be of type numeric.
-To avoid a potential bias introduced by features with high variance when selecting multiple features, all features are standardized by default.
+### Spatial Clustering (sperrorest)
 
-The following example clusters by feature "distance to forest".
-
-```{r env-1, fig.cap="Environmental blocking using a single predictor.", cache=TRUE, out.width="50%", fig.pos="ht"}
-resampling_env = rsmp("spcv_env",
-  features = "distdeforest", folds = 5)
-
-plot(resampling_env,
-  size = 0.8, fold_id = 1, task = task, crs = 32717,
-  show_blocks = TRUE, show_labels = TRUE) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
-```
-
-It is also possible to cluster by multiple features:
-
-```{r env-2, fig.cap="Environmental blocking using multiple predictors.", cache=TRUE, out.width="50%", fig.pos="ht"}
-resampling_env_multi = rsmp("spcv_env",
-  features = c("distdeforest", "slope"), folds = 5)
-
-plot(resampling_env_multi,
-  size = 0.8, fold_id = 1, task = task, crs = 32717,
-  show_blocks = TRUE, show_labels = TRUE) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
-```
-
-## Spatial CV
+#### Coordinate-based clustering
 
@@
 
-## CreateSpacetimeFolds (Cstf)
+### Feature space clustering (blockCV)
 
-Leave-Location-and-Time-Out from packge \pkg{CAST} is a spatiotemporal resampling method [@meyer2018a].
+The last method from the \pkg{blockCV} package, "environmental blocking", makes use of *k-means* clustering [@hartigan1979a] in a possibly multivariate space.
+The user can specify a variable via argument `feature` which levels will be used to create clusters.
+Hereby, k-means will use Euclidean distance for cluster creation.
+Hence, the selected input variables should be of type numeric.
+To avoid a potential bias introduced by features with high variance when selecting multiple features, all features are standardized by default.
+
+The following example clusters by feature "distance to forest".
+
+```{r env-1, fig.cap="Environmental blocking using a single predictor.", cache=F, out.width="50%", fig.pos="ht"}
+resampling_env = rsmp("spcv_env",
+  features = "distdeforest", folds = 5)
+
+plot(resampling_env,
+  size = 0.8, fold_id = 1, task = task, crs = 32717,
+  show_blocks = TRUE, show_labels = TRUE) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```
+
+It is also possible to cluster by multiple features:
+
+```{r env-2, fig.cap="Environmental blocking using multiple predictors.", cache=T, out.width="50%", fig.pos="ht"}
+resampling_env_multi = rsmp("spcv_env",
+  features = c("distdeforest", "slope"), folds = 5)
+
+plot(resampling_env_multi,
+  size = 0.8, fold_id = 1, task = task, crs = 32717,
+  show_blocks = TRUE, show_labels = TRUE) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```
+
+## Spatiotemporal
+
+### Factor-based grouping
+
+#### Predefined grouping
+
+FIXME
+
+#### Leave-Location-And-Time-Out (CAST)
+
+Leave-Location-and-Time-Out (LLTO) from packge \pkg{CAST} is a spatiotemporal resampling method [@meyer2018a].
 It requires both a spatial and a temporal variable which are used to to separate observations in a multi-dimensional space.
-It comes with satellite methods named "Leave-time-out" (LTO) and "Leave-location-out" (LLO) which only operatre in a two-dimensional space.
+It comes with satellite methods named "Leave-time-out" (LTO) and "Leave-location-out" (LLO) which only operate in a two-dimensional space.
 
@@
 
-### Leave-Time-Out (LTO)
+##### Leave-Time-Out (LTO)
 
@@
 
-### Leave-Location-Out (LLO)
+##### Leave-Location-Out (LLO)
 
@@
 
-### Leave-Location-and-Time-Out (LLTO)
+##### Leave-Location-and-Time-Out (LLTO)
 
@@
 
-## Spatiotemporal Clustering (Cluto)
+### Spatiotemporal Clustering (skmeans)
 

### 2021-05-02 paper/paper.Rmd
adjust caching settings
reorder feature space section

add content to spatial tiling section

@@
   - name: Patrick Schratz
-    affiliation: 'Friedrich-Schiller-University Jena \AND'
+    affiliation: 'Friedrich-Schiller-University Jena'
     # affiliaton2: 'Ludwig-Maximilians-University Munich'
@@
   - name: Marc Becker
-    affiliation: 'Ludwig-Maximilians-University Munich'
+    affiliation: 'Ludwig-Maximilians-University Munich \AND'
     address: |
@@
 
-```{r setup-2, echo=FALSE, purl=TRUE}
+```{r setup-2, echo=FALSE, purl=T}
 # for knitr spint later
-knitr::opts_chunk$set(fig.path = "./paper/figs/")
+knitr::opts_chunk$set(fig.path = "./paper/figs/",
+  cache = F,
+  cache.lazy = TRUE)
 ```
@@
 
-```{r create task, cache=F}
+```{r create task}
 library(mlr3)
@@
 
-```{r vis-spcv, fig.cap="Visualization of spatial partitioning", fig.pos="ht", fig.height=6, fig.width=8, cache=TRUE}
+```{r vis-spcv, fig.cap="Visualization of spatial partitioning", fig.pos="ht", fig.height=6, fig.width=8}
 resampling_sp = rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
@@
 
-```{r vis-nspcv, fig.cap="Visualization of random partitioning", fig.pos="ht", fig.height=6, fig.width=8, cache=T}
+```{r vis-nspcv, fig.cap="Visualization of random partitioning", fig.pos="ht", fig.height=6, fig.width=8}
 resampling_nsp = rsmp("repeated_cv", folds = 4, repeats = 2)
@@
 
-```{r buffer, fig.cap="Visualization of 'Spatial buffering' method.", out.width="50%", cache=T}
+```{r buffer, fig.cap="Visualization of 'Spatial buffering' method.", out.width="50%"}
 resampling_buffer = rsmp("spcv_buffer", theRange = 1000)
@@
 
-plot(resampling_buffer,
+autoplot(resampling_buffer,
   size = 0.8, task = task, fold_id = 1, crs = 32717) *
@@
 
-```{r block-random, cache = TRUE, fig.cap="Spatial blocking with option `selection = 'random'`.", out.width="50%", fig.pos="ht"}
+```{r block-random, fig.cap="Spatial blocking with option `selection = 'random'`.", out.width="50%", fig.pos="ht"}
 resampling_block_random = rsmp("spcv_block", range = 1000, folds = 5)
 
-plot(resampling_block_random,
+autoplot(resampling_block_random,
   size = 0.8, fold_id = 1, task = task, crs = 32717,
@@
 
-```{r block-systematic, cache=TRUE, fig.cap="Spatial blocking with option `selection = 'systematic'`.", out.width="50%", fig.pos="ht"}
+```{r block-systematic, fig.cap="Spatial blocking with option `selection = 'systematic'`.", out.width="50%", fig.pos="ht"}
 resampling_block_systematic = rsmp("spcv_block",
@@
 
-plot(resampling_block_systematic,
+autoplot(resampling_block_systematic,
   size = 0.8, fold_id = 1, task = task, crs = 32717,
@@
 
-```{r block-checkerboard, cache=TRUE, fig.cap="Spatial blocking with option `selection = 'checkerboard'`.", out.width="40%", fig.pos="h!"}
+```{r block-checkerboard, fig.cap="Spatial blocking with option `selection = 'checkerboard'`.", out.width="40%", fig.pos="h!"}
 resampling_block_checkerboard = rsmp("spcv_block",
@@
 
-plot(resampling_block_checkerboard,
+autoplot(resampling_block_checkerboard,
   size = 0.8, fold_id = 1, task = task, crs = 32717,
@@
 
-```{r block-rows-and-cols, cache=TRUE, fig.cap="Spatial blocking with arguments `rows` and `cols`.", out.width="40%", fig.pos="ht"}
+```{r block-rows-and-cols, fig.cap="Spatial blocking with arguments `rows` and `cols`.", out.width="40%", fig.pos="ht"}
 resampling_block_rows_cols = rsmp("spcv_block",
@@
 
-plot(resampling_block_rows_cols,
+autoplot(resampling_block_rows_cols,
   size = 0.8, fold_id = 1, task = task, crs = 32717,
@@
 
+Method `partition_tiles()` from package \pkg{sperrorest} also uses rectangular blocks to partition the study area.
+In contrast to "Spatial Blocking" from \pkg{blockCV} these blocks can be rotated (via argument `rotation`) and a minimum number of observations per block can be set (argument `min_n`).
+A predefined row/column grid is required via either argument `dsplit` or `nsplit` and no square blocks can be used.
+Small tiles can be merged with adjacent blocks via argument `reassign` or optionally be ignored (`reassign = FALSE`).
+The number of tiles set by the user equals the number of folds and blocks cannot be grouped into folds as in "Spatial Blocking".
+Also despite setting 12 blocks via `nsplit = c(3L, 4L)` in the following example, the final partition count is eleven because the bottom-left part of the study area does not contain observations.
+
+```{r tile, fig.cap="Spatial tiling with arguments `rows` and `cols`.", out.width="40%", fig.pos="ht"}
+resampling_tiles = rsmp("spcv_tiles",
+  nsplit = c(3L, 4L)
+)
+
+autoplot(resampling_tiles,
+  size = 0.8, fold_id = 1, task = task, crs = 32717) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```
+
 ### Spatial Clustering (sperrorest)
@@
 
-```{r coords, fig.cap="Method 'Spatial CV' showing a k-means clustering approach based on coordinates.", cache=TRUE, out.width="50%", fig.pos="ht"}
+```{r coords, fig.cap="Method 'Spatial CV' showing a k-means clustering approach based on coordinates.", out.width="50%", fig.pos="ht"}
 resampling_coords = rsmp("spcv_coords", folds = 5)$instantiate(task)
 
-plot(resampling_coords,
+autoplot(resampling_coords,
   size = 0.8, fold_id = 1, task = task, crs = 32717) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
-```
-
-### Feature space clustering (blockCV)
-
-The last method from the \pkg{blockCV} package, "environmental blocking", makes use of *k-means* clustering [@hartigan1979a] in a possibly multivariate space.
-The user can specify a variable via argument `feature` which levels will be used to create clusters.
-Hereby, k-means will use Euclidean distance for cluster creation.
-Hence, the selected input variables should be of type numeric.
-To avoid a potential bias introduced by features with high variance when selecting multiple features, all features are standardized by default.
-
-The following example clusters by feature "distance to forest".
-
-```{r env-1, fig.cap="Environmental blocking using a single predictor.", cache=F, out.width="50%", fig.pos="ht"}
-resampling_env = rsmp("spcv_env",
-  features = "distdeforest", folds = 5)
-
-plot(resampling_env,
-  size = 0.8, fold_id = 1, task = task, crs = 32717,
-  show_blocks = TRUE, show_labels = TRUE) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
-```
-
-It is also possible to cluster by multiple features:
-
-```{r env-2, fig.cap="Environmental blocking using multiple predictors.", cache=T, out.width="50%", fig.pos="ht"}
-resampling_env_multi = rsmp("spcv_env",
-  features = c("distdeforest", "slope"), folds = 5)
-
-plot(resampling_env_multi,
-  size = 0.8, fold_id = 1, task = task, crs = 32717,
-  show_blocks = TRUE, show_labels = TRUE) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
@@
 
-```{r, results='hide', cache=TRUE}
+```{r, results='hide'}
 data = cookfarm_sample
@@
 
-plot(resampling_cstf_time,
+autoplot(resampling_cstf_time,
   fold_id = 5, task = task_spt, crs = 32717, plot3D = TRUE
@@
 
-```{r cache=TRUE, results='hide'}
+```{r results='hide'}
 resampling_cstf_loc = rsmp("sptcv_cstf", folds = 5, space_var = "SOURCEID")
@@
 
-plot(resampling_cstf_loc,
+autoplot(resampling_cstf_loc,
   fold_id = 5, task = task_spt, crs = 32717, plot3D = TRUE)
@@
 
-```{r cache=TRUE, results='hide'}
+```{r results='hide'}
 resampling_cstf_time_loc = rsmp("sptcv_cstf",
@@
 
-plot(resampling_cstf_time_loc,
+autoplot(resampling_cstf_time_loc,
   fold_id = 4, task = task_spt, crs = 32717, plot3D = TRUE,
@@
 ```{r llto, echo=FALSE, out.width="90%", fig.cap="Three-dimensional visualization of method 'sptcv\\_cstf' and sub-method 'Leave-Location-And-Time-Out' (LLTO).
-The grey points show the points omitted by this method.", cache=TRUE, fig.pos="ht"}
+The grey points show the points omitted by this method.", fig.pos="ht"}
 knitr::include_graphics("llto.png")
@@
 Users are encouraged to visit the help page of the package for this method (`?ResamplingSptCVCluto`) for more information.
+
+## Feature space clustering
+
+The last method from the \pkg{blockCV} package, "environmental blocking", makes use of *k-means* clustering [@hartigan1979a] in a possibly multivariate space.
+The user can specify a variable via argument `feature` which levels will be used to create clusters.
+Hereby, k-means will use Euclidean distance for cluster creation.
+Hence, the selected input variables should be of type numeric.
+To avoid a potential bias introduced by features with high variance when selecting multiple features, all features are standardized by default.
+
+The following example clusters by feature "distance to forest".
+
+```{r env-1, fig.cap="Environmental blocking using a single predictor.", out.width="50%", fig.pos="ht"}
+resampling_env = rsmp("spcv_env",
+  features = "distdeforest", folds = 5)
+
+autoplot(resampling_env,
+  size = 0.8, fold_id = 1, task = task, crs = 32717,
+  show_blocks = TRUE, show_labels = TRUE) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```
+
+It is also possible to cluster by multiple features:
+
+```{r env-2, fig.cap="Environmental blocking using multiple predictors.", out.width="50%", fig.pos="ht"}
+resampling_env_multi = rsmp("spcv_env",
+  features = c("distdeforest", "slope"), folds = 5)
+
+autoplot(resampling_env_multi,
+  size = 0.8, fold_id = 1, task = task, crs = 32717,
+  show_blocks = TRUE, show_labels = TRUE) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```
 
@@
 
-```{r prepare, cache=TRUE}
+```{r prepare}
 library("mlr3")
@@
 
-```{r rr-nspcv, cache=TRUE, cache.lazy = FALSE}
+```{r rr-nspcv, cache.lazy = FALSE}
 resampling_nsp = rsmp("repeated_cv", folds = 4, repeats = 2)
@@
 
-```{r rr-spcv, cache=TRUE, cache.lazy = FALSE}
+```{r rr-spcv, cache.lazy = FALSE}
 resampling_sp = rsmp("repeated_spcv_coords", folds = 4, repeats = 2)

### 2021-05-02 paper/paper.Rmd
add spatial disc description

@@
 #### Spatial disc (sperrorest)
+
+Another buffering oriented resampling method is "Spatial Disc" from package \pkg{sperrorest}.
+The name spins from "Leave-one-disc-out" which again originates from the generic "Leave-one-out" method.
+More specifically, the 'disc' which is being left our here refers to the radial area of test sets created from this method.
+
+Around the radial test set a buffer zone / neutral area can be defined in which observations are excluded from the training set, increasing the spatial distance between train and test sets.
+
+```{r disc, fig.cap="Visualization of 'Spatial Disc' method.", out.width="50%"}
+resampling_disc = rsmp("spcv_disc", folds = 5, radius = 200L, buffer = 300L)
+resampling_disc$instantiate(task)
+
+autoplot(resampling_disc,
+  size = 0.8, task = task, fold_id = 1, crs = 32717) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```
 

### 2021-05-02 paper/paper.Rmd
update methods table

@@
   \usepackage{placeins}
+  \newcommand{\specialcell}[2][l]{%
+    \begin{tabular}[#1]{@{}l@{}}#2\end{tabular}}
 tables: true
@@
 })
-# knitr::opts_chunk$set(cache = FALSE)
 ```
@@
 knitr::opts_chunk$set(fig.path = "./paper/figs/",
-  cache = F,
+  cache = T,
   cache.lazy = TRUE)
@@
   \centering
+  \footnotesize
   \caption[t]{Spatiotemporal resampling methods, ordered alphabetically by class name (Notation column). }
   \begingroup
-  <!-- \begin{adjustbox}{angle=270} -->
     \begin{tabular}{lllll}
@@
       \toprule
-      \multirow{3}{*}{Spatial}        & Spatial Buffering           & \pkg{blockCV}, \pkg{sperrorest}    & "spcv\_buffer", "spcv\_disc()", "spcv\_loo()" & Lit1       \\
-                                      & Spatial Blocking            & \pkg{blockCV}, \pkg{sperrorest}  & "spcv\_block",
-                                      "spcv\_tiles()"  & Lit1       \\
-                                      & Spatial Clustering          & \pkg{sperrorest} & "spcv\_coords" & Lit1       \\
-       & Leave-Location-Out & \pkg{CAST}, \pkg{sperrorest}       & "sptcv\_cstf('space\_var')", "spcv\_factor"  & Lit1       \\
+      \multirow{6}{*}{Spatial}        & Spatial Buffering           & \specialcell{\pkg{blockCV} \\ \pkg{sperrorest}}    & \specialcell{ \texttt{"spcv\_buffer"} \\  \texttt{"spcv\_disc"}}  & Lit1       \\
+      \cmidrule{2-5}
+                                      & Spatial Blocking            & \specialcell{\pkg{blockCV} \\ \pkg{sperrorest}}  & \specialcell{  \texttt{"spcv\_block"} \\  \texttt{"spcv\_tiles"}}
+                                        & Lit1       \\
+  \cmidrule{2-5}
+                                      & Spatial Clustering          & \pkg{sperrorest} & \texttt{"spcv\_coords"} & Lit1       \\
+                                      \cmidrule{2-5}
+       & Leave-Location-Out & \specialcell{\pkg{CAST} \\ \pkg{mlr3}}       & \specialcell{ \texttt{"sptcv\_cstf('space\_var')"} \\ \texttt{"FIXME"}},  & Lit1       \\
       \midrule
-      \multirow{2}{*}{Spatiotemporal} & Leave-Location-and-Time-Out & \pkg{CAST}       & "sptcv\_cstf"  & Lit1       \\
-                                      & Spatiotemporal Clustering   & \pkg{skmeans}    & "sptcv\_cluto" & Lit1       \\
+      \multirow{2}{*}{Spatiotemporal} & Leave-Location-and-Time-Out & \pkg{CAST}       &  \texttt{"sptcv\_cstf"}  & Lit1       \\
+                                      & Spatiotemporal Clustering   & \pkg{skmeans}    &  \texttt{"sptcv\_cluto"} & Lit1       \\
       \midrule
-      Feature space                   & Environmental Blocking      & \pkg{blockCV}    & "spcv\_env"    & Lit1       \\
+      Feature space                   & Environmental Blocking      & \pkg{blockCV}    &  \texttt{"spcv\_env"}    & Lit1       \\
     \end{tabular}
-  <!-- \end{adjustbox} -->
   \endgroup\label{tab:sptcv-methods}

### 2021-05-05 paper/ABnotes.Rmd
changed ref names in bib file

@@
 ---
-title: "Alex's Notes and Text Fragments for the JOSS Manuscript"
+title: "Alex's Notes and Text Fragments for the JSS Manuscript"
 author: "Alexander Brenning"
-date: "21 Apr 2021"
+date: "05 May 2021"
 output: html_document
@@
 Spatial and spatio-temporal prediction tasks are common in applications ranging from environmental sciences to archaeology and epidemiology.
-While sophisticated mathematical frameworks have long been developed in spatial statistics to characterize predictive uncertainties under well-defined mathematical assumptions such as intrinsic stationarity [e.g., cressie.2015], computational estimation procedures have only been proposed more recently to assess predictive performances of spatial and spatio-temporal prediction models [@brenning.2005.nhess; @sperrorest; @pohjankukka2017; @roberts2017].
+While sophisticated mathematical frameworks have long been developed in spatial statistics to characterize predictive uncertainties under well-defined mathematical assumptions such as intrinsic stationarity [e.g., cressie2015], computational estimation procedures have only been proposed more recently to assess predictive performances of spatial and spatio-temporal prediction models [@brenning2005; @sperrorest; @pohjankukka2017; @roberts2017].
 
-Although alternatives such as the bootstrap exist and have some advantages [@efron.gong.1983, or David Hand], cross-validation is a particularly well-established, easy-to-implement algorithm for *model assessment* of supervised machine-learning models [@efron.gong.1983 and next section] and *model selection* [@arlot.celisse.2010].
-In its basic form, it is based on resampling the data without paying attention to any possible dependence structure, which may arise from, e.g., cluster sampling [ref], grouped data [@brenning.lausen.2008; @pena.brenning.2015.maipo], or underlying environmental processes inducing some sort of spatial coherence at the landscape scale [@brenning.2005.nhess].
-In treating dependent observations as independent, or ignoring autocorrelation, cross-validation test samples may in fact be heavily correlated with, or even pseudo-replicates of, the data used for training the model, which introduces a potentially severe bias, especially for flexible machine-learning models [@brenning.2005.nhess].
+Although alternatives such as the bootstrap exist and have some advantages [@efron1983; @hand1996], cross-validation is a particularly well-established, easy-to-implement algorithm for *model assessment* of supervised machine-learning models [@efron1983 and next section] and *model selection* [@arlot2010].
+In its basic form, it is based on resampling the data without paying attention to any possible dependence structure, which may arise from, e.g., cluster sampling [ref], grouped data [@brenning2008; @pena2015], or underlying environmental processes inducing some sort of spatial coherence at the landscape scale [@brenning2005].
+In treating dependent observations as independent, or ignoring autocorrelation, cross-validation test samples may in fact be heavily correlated with, or even pseudo-replicates of, the data used for training the model, which introduces a potentially severe bias, especially for flexible machine-learning models [@brenning2005].
 
-This cross-validation bias is well-known in spatial as well as non-spatial prediction [@brenning.2005.nhess; @brenning.lausen.2008; @arlot.celisse.2010].
+This cross-validation bias is well-known in spatial as well as non-spatial prediction [@brenning2005; @brenning2008; @arlot2010].
 It is most easily understood from a predictive modelling perspective by focusing on the question where (and when) the model should be used for prediction.
@@
 2000 grid cells from 100 fields scattered across a large study region.
-The purpose of training a model on this sample is to make predictions on other, new *fields* within the same geographic domain [*intra-domain* prediction; @brenning.2005.nhess] -- not *within* the same field, which obviously presents only one crop type.
+The purpose of training a model on this sample is to make predictions on other, new *fields* within the same geographic domain [*intra-domain* prediction; @brenning2005] -- not *within* the same field, which obviously presents only one crop type.
 It would therefore be foolish to train a model on a simple random subsample of grid cells, and test it on the remaining data, i.e.
 using other grid cells from the same fields, as if we wanted to predict within a field; the results from this performance assessment would be over-optimitic, and perhaps badly so.
-To mimic the predictive situation for which the model is trained, we'd rather have to resample at the level of fields, not grid cells [@pena.brenning.2015.maipo].
+To mimic the predictive situation for which the model is trained, we'd rather have to resample at the level of fields, not grid cells [@pena2015].
 If the model was to be applied to adjacent agricultural districts, i.e.
-outside the learning sample's spatial domain [*extra-domain* prediction; @brenning.2005.nhess], it would even seem necessary to resample at a higher level of spatial aggregation, i.e.
+outside the learning sample's spatial domain [*extra-domain* prediction; @brenning2005], it would even seem necessary to resample at a higher level of spatial aggregation, i.e.
 at the level of agricultural districts within the learning sample, in order to realistically mimic the actual prediction task.
@@
 
-the formal stuff, next on my list...
-
 In cross-validation for predictive model assessment, we consider the following formal setting.
@@
 where $l$ is a real-valued loss function, and the expected value is with respect the probability distribution of $X$, the features of an instance $(Y,X)$ drawn randomly from the underlying population.
-This is referred to as the *actual* or *conditional* performance measure as it is conditional on $L$ [@hand.1996].
+This is referred to as the *actual* or *conditional* performance measure as it is conditional on $L$ [@hand1996].
 The loss function can take a variety of forms such as the misclassification error $I(Y\neq\hat{f}_L(X))$ in classification, or the squared error $(Y-\hat{f}_L(X))^2$ in regression, among many other possible measures.
-The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance metrics for regression and classification [see e.g., @hand.1996 for classification, and ref??
+The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance metrics for regression and classification [see e.g., @hand1996 for classification, and ref??
 for regression tasks].
@@
 This representation as a point estimator for $\mathit{perf(\hat{f}_L)}$ underlines the importance of using a random sample for performance estimation to avoid estimation bias.
-Other estimators than the simple mean may be required when $T$ is not a simple random sample, for instance a stratified random sample [e.g., @thompson.2012].
+Other estimators than the simple mean may be required when $T$ is not a simple random sample, for instance a stratified random sample [e.g., @thompson2012].
 As always, judgment sampling may lead to uncontrollable bias.
@@
 where did I get this from?].
-The $k$-fold CV estimator of model performance is a nearly unbiased estimator of the conditional performance measure when the observations were drawn independently [@efron.gong.1983].
+The $k$-fold CV estimator of model performance is a nearly unbiased estimator of the conditional performance measure when the observations were drawn independently [@efron1983].
 Since $\widehat{\mathit{perf}}_{L, CV}(f)$ still depends on the particular partitioning chosen for $L$, it is sometimes recommended to repeat the estimation using different random partitionings ($r$-repeated $k$-fold cross-validation, e.g.
@@
 
-A variant of cross-validation is leave-one-out (LOO) cross-validation , which has long been established in geostatistics [@cressie.2015], sometimes with a focus on the spatial distribution of LOO error [@willmott.matsuura.2006].
-Since this is just a special case of non-spatial CV with $k=n$, it is sometimes also referred to as spatial CV [@willmott.matsuura.2006].
-Nevertheless, spatial variants have been proposed that apply an exclusion buffer to the test locations to separate them from the training data [@brenning.2005.nhess].
-One plausible approach for this is to use the range of autocorrelation of model residuals to determine the buffer distance as this ensures independence conditional on the predictors [@brenning.2005.nhess].
-When data is grouped, e.g., due to multi-level sampling designs or the study of spatial objects, LOO at has been proposed at the site level [@martin.et.al.2008; @kasurak.et.al.2011] or, in animal movement studies, at the animal level [@anderson.et.al.2005].
-Similarly, in the temporal domain, it has been proposed to leave out temporal observational units, as in leave-one-year-out CV [@anderson.et.al.2005; @brenning.2005.nhess].
+A variant of cross-validation is leave-one-out (LOO) cross-validation , which has long been established in geostatistics [@cressie2015], sometimes with a focus on the spatial distribution of LOO error [@willmott2006].
+Since this is just a special case of non-spatial CV with $k=n$, it is sometimes also referred to as spatial CV [@willmott2006].
+Nevertheless, spatial variants have been proposed that apply an exclusion buffer to the test locations to separate them from the training data [@brenning2005].
+One plausible approach for this is to use the range of autocorrelation of model residuals to determine the buffer distance as this ensures independence conditional on the predictors [@brenning2005].
+When data is grouped, e.g., due to multi-level sampling designs or the study of spatial objects, LOO at has been proposed at the site level [@martin2008; @kasurak2011] or, in animal movement studies, at the animal level [@anderson2005].
+Similarly, in the temporal domain, it has been proposed to leave out temporal observational units, as in leave-one-year-out CV [@anderson2005; @brenning2005].
 
-It should be noted that $k$-fold cross-validation with a large value of $k$, and LOO-CV in particular, are not only very time-consuming since the model has to be trained $k$ times; these models will also be nearly identical since only a tiny fraction of the data is withheld, and therefore estimation bias increases.
+It should be noted that $k$-fold cross-validation with a large value of $k$, and LOO-CV in particular ($k=n$), are not only very time-consuming since the model has to be trained $k$ times; these models will also be nearly identical since only a tiny fraction of the data is withheld, and therefore estimation bias increases.
 'Pure' LOO-CV is therefore not recommended for machine-learning model assessment [ref??].

### 2021-05-05 paper/paper.Rmd
updated intro sections with ABnotes text

@@
   \usepackage{amsmath}
+  \usepackage{amssymb}
   \usepackage{booktabs}
@@
 
-Data which includes spatial or temporal information requires special treatment in machine learning; similar to cost-sensitive, functional, multilabel, ordinal or survival datasets [@mlr3book].
-In contrast to non-spatial/non-temporal data, observations inherit a natural grouping, either in space or time or in both space and time [@legendre1993].
-This grouping causes observations to be autocorrelated, either in space (spatial autocorrelation (SAC)), time (temporal autocorrelation (TAC)) or both space and time (spatiotemporal autocorrelation (STAC)).
-For simplicity, the acronym STAC is used as a generic term in the following discussion for all the different characteristics introduced above.
+Spatial and spatio-temporal prediction tasks are common in applications ranging from environmental sciences to archaeology and epidemiology. 
+While sophisticated mathematical frameworks have long been developed in spatial statistics to characterize predictive uncertainties under well-defined mathematical assumptions such as intrinsic stationarity [e.g., cressie2015], computational estimation procedures have only been proposed more recently to assess predictive performances of spatial and spatio-temporal prediction models [@brenning2005; @sperrorest; @pohjankukka2017; @roberts2017].
 
-## Spatiotemporal Autocorrelation in Statistical/Machine Learning
+Although alternatives such as the bootstrap exist and have some advantages [@efron1983; @hand1996], cross-validation is a particularly well-established, easy-to-implement algorithm for *model assessment* of supervised machine-learning models [@efron1983 and next section] and *model selection* [@arlot2010].
+In its basic form, it is based on resampling the data without paying attention to any possible dependence structure, which may arise from, e.g., cluster sampling [ref], grouped data [@brenning2008; @pena2015], or underlying environmental processes inducing some sort of spatial coherence at the landscape scale [@brenning2005].
+In treating dependent observations as independent, or ignoring autocorrelation, cross-validation test samples may in fact be heavily correlated with, or even pseudo-replicates of, the data used for training the model, which introduces a potentially severe bias, especially for flexible machine-learning models [@brenning2005].
 
-The overarching problem is that STAC violates the assumption that the observations in the train and test datasets are independent [@hastie2001].
-If this assumption is violated, the reliability of the resulting performance estimates, for example retrieved via cross-validation, is decreased.
-The magnitude of this decrease is linked to the magnitude of STAC in the dataset, which cannot be determined easily.
+This cross-validation bias is well-known in spatial as well as non-spatial prediction [@brenning2005; @brenning2008; @arlot2010]. 
+It is most easily understood from a predictive modelling perspective by focusing on the question where (and when) the model should be used for prediction. 
+In crop classification from remotely-sensed data, for instance, learning samples routinely contain multiple grid cells from a sample of fields with known crop type, e.g.
+2000 grid cells from 100 fields scattered across a large study region.
+The purpose of training a model on this sample is to make predictions on other, new *fields* within the same geographic domain [*intra-domain* prediction; @brenning2005] &emdash; not *within* the same field, which obviously presents only one crop type.
+It would therefore be foolish to train a model on a simple random subsample of grid cells, and test it on the remaining data, i.e.
+using other grid cells from the same fields, as if we wanted to predict within a field; the results from this performance assessment would be over-optimitic, and perhaps badly so.
+To mimic the predictive situation for which the model is trained, we'd rather have to resample at the level of fields, not grid cells [@pena2015]. 
+If the model was to be applied to adjacent agricultural districts, i.e.
+outside the learning sample's spatial domain [*extra-domain* prediction; @brenning2005], it would even seem necessary to resample at a higher level of spatial aggregation, i.e.
+at the level of agricultural districts within the learning sample, in order to realistically mimic the actual prediction task.
+The cross-validation resampling needed therefore depends as much on the prediction task as on the data structure or dependency at hand.
 
-One approach to account for the existence of STAC is to use dedicated resampling methods.
-[mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) provides access to the most frequently used spatiotemporal resampling methods.
-The following example, which uses the [ecuador](https://mlr3spatiotempcv.mlr-org.com/reference/mlr_tasks_ecuador.html) dataset created by [Jannes Muenchow](https://scholar.google.com/citations?user=Slq94Y4AAAAJ&hl=de&authuser=1&oi=ao), showcases how a spatial dataset can be used to retrieve a bias-reduced performance estimate of a learner.
+While it is not the purpose of this paper to recommend specific resampling schemes for specific use cases, this example may suffice to motivate the use of spatial and spatio-temporal cross-validation techniques, and the need for a unified framework and computational toolbox that accommodate a variety of prediction tasks that may be applicable to a broad range of prediction tasks.
+\pkg{mlr3spatiotempcv} is such a toolbox.
 
-This dataset contains information on the occurrence of landslides (binary) in the Andes of Southern Ecuador.
-The landslides were mapped from aerial photos taken in 2000.
-The dataset is well suited to serve as an example because it is relatively small and, of course, due to the spatial nature of the observations.
-Please refer to @muenchow2012 for a detailed description of the dataset.
+This toolbox, implemented as an open-source R package, builds upon and generalizes several existing toolboxes that have been developed in recent years for more specific settings (see Table \@ref(tab:sptcv-methods)).
+The earliest of these implementations is the \pkg{sperrorest} package [@sperrorest], which has, however, not been integrated into established machine-learning frameworks such as \pkg{mlr} or \pkg{caret}, and lacks support of temporal prediction tasks.
+Other more recent implementations include \pkg{blockCV} for (what for?) [ref @blockCV; @rest2014] and \pkg{CAST} for spatio-temporal prediction tasks in a \pkg{caret} framework [@cast; @meyer2018a].
 
-To account for the spatial autocorrelation probably present in the landslide data, we will make use of one of the most used spatial partitioning methods, a cluster-based $k$-means grouping [@sperrorest], (`"spcv_coords"` in [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com)).
-This method performs a clustering in 2D space which contrasts with the commonly used random partitioning for non-spatial data.
-The grouping has the effect that train and test data are more separated in space than they would be by conducting a random partitioning, thereby reducing the effect of STAC.
+Thus, \pkg{mlr3spatiotempcv} implements for the first time a comprehensive state-of-the-art compilation of spatial and spatio-temporal partitioning schemes that is well-integrated into a major machine-learning framework in R, the \pkg{mlr3} ecosystem [@mlr3].
+This package is furthermore equipped with a variety of two- and three-dimensional diagram types.
+Our hope is that implementation will facilitate reproducible geospatial modelling and code-sharing across a broad range of application domains.
 
-By contrast, when using the classical random partitioning approach with spatial data, train and test observations would be located side-by-side across the full study area (a visual example is provided further below).
-This leads to a high similarity between train and test sets, resulting in "better" but over-optimistic performance estimates in every fold of a CV compared to the Spatial CV (SpCV) approach.
-However, these low error rates are mainly caused due to the STAC in the observations and the lack of appropriate partitioning methods and not by the power of the fitted model.
+# Spatial and Spatio-temporal Cross-validation
 
-Rephrased: the higher performance results in a non-spatial CV scenario should not be relied on or used for reporting.
-The positive bias in these results is substantial and the actual performance of the model is nowhere near this result.
-Results from a SpCV are not unbiased but are closer in reflecting the actual performance of the learner on unknown datasets.
+In cross-validation for predictive model assessment, we consider the following formal setting. 
+We are interested in predicting a numerical or categorical response $y$ of an object or instance using a feature vector $\mathbf{x} = (x^{(1)}, \ldots, x^{(p)}\in\mathbb{R}$ and a model $\hat{f}_L$ that has been trained on a learning sample $L = \{(y_i, \mathbf{x}_i),\ i = 1, ..., n\}$.
+Our goal is to estimate the expected value of the performance of $\hat{f}_L$,
+\[
+\mathit{perf(\hat{f}_L)} := E(l(Y,\hat{f}_L(X))),
+\]
+where $l$ is a real-valued loss function, and the expected value is with respect the probability distribution of $X$, the features of an instance $(Y,X)$ drawn randomly from the underlying population.
+This is referred to as the *actual* or *conditional* performance measure as it is conditional on $L$ [@hand1996]. 
+The loss function can take a variety of forms such as the misclassification error $I(Y\neq\hat{f}_L(X))$ in classification, or the squared error $(Y-\hat{f}_L(X))^2$ in regression, among many other possible measures.
+The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance metrics for regression and classification [see e.g., @hand1996 for classification, and ref??
+for regression tasks].
 
-## Motivation
+Nevertheless, since we only have a sample $T$ of test data drawn from the population, we can realistically only *estimate* the conditional performance of $\hat{f}_L$:
+\[
+\widehat{\mathit{perf}}_T(\hat{f}_L) = \frac{1}{|T|}\sum_{i=1}^{|T|}l(Y,\hat{f}_L(X)).
+\]
+This representation as a point estimator for $\mathit{perf(\hat{f}_L)}$ underlines the importance of using a random sample for performance estimation to avoid estimation bias.
+Other estimators than the simple mean may be required when $T$ is not a simple random sample, for instance a stratified random sample [e.g., @thompson2012].
+As always, judgment sampling may lead to uncontrollable bias.
 
-In the last years awareness for spatiotemporal autocorrelation in ML tasks has grown [@pohjankukka2017; @roberts2017].
-Subsequently, scientists started writing code to account for the issue in their studies.
-With R being one of the most used languages with respect to modeling/statistics and academia in general, R packages were developed aiming to simplify the data splitting with respect to methods able to account for spatial autocorrelation [@cast, @blockCV].
+Since the re-use of the learning sample $L$ for testing, $T=L$, would result in the over-optimistic *resubstitution* or *apparent* model performance, cross-validation partitions the sample $L$ into disjoint training and test sets.
+Specifically, $L$ is split into $k$ partitions,
+\[
+L = L_1 \cup \ldots \cup L_k,\qquad L_i\cap L_j = \emptyset\quad \textrm{for all}\ i\neq j,
+\]
+and a model $\hat{f}_{(i)}$ is fitted on $L_{(i)} := L\setminus L_i$, while $L_i$ is withheld for testing. 
+This is repeated for $i=1,\ldots,k$ in order to effectively use the entire sample for testing, while keeping training and test sets disjoint at all times. 
+The $k$-fold CV estimator can therefore be written as
+\[
+\widehat{\mathit{perf}}_{L, CV}(f) := \frac{1}{k}\sum_{i=1}^k\widehat{\mathit{perf}}_{L_i}(\hat{f}_{L_{(i)}}),
+\]
+where $f$ is a machine-learning algorithm, i.e. a mapping that trains a model $\hat{f}_S$ using any suitable training sample $S$. 
+The use of $k=5$ or $k=10$ folds is most commonly seen in practice, and these preferences are also supported by theory [ref??]. 
+The $k$-fold CV estimator of model performance is a nearly unbiased estimator of the conditional performance measure when the observations were drawn independently [@efron1983]. 
+Since $\widehat{\mathit{perf}}_{L, CV}(f)$ still depends on the particular partitioning chosen for $L$, it is sometimes recommended to repeat the estimation using different random partitionings ($r$-repeated $k$-fold cross-validation, e.g.
+$k=100$; ref??).
 
-This is a typical process for the R community as writing R packages has been made relatively straightforward in recent years thanks to efforts from various people/organizations such as Bioconductor, CRAN, ropensci or the tidyverse [@cran, @ropensci, @tidyverse]
-Also the effort of bundling code into a defined package structure makes it easier for others to re-use it, which is a good effort towards reproducible science.
+In traditional cross-validation, the partitioning is based on uniform random sampling, which ignores spatial or temporal autocorrelation or any existing grouping structure as well as the structure of the prediction task, as outlined in the Introduction, and may result in over-optimistic performance estimates.
+Several approaches have therefore been proposed in the literature and implemented in software to accommodate a variety of predictive situations (Table \@ref(tab:sptcv-methods)).
 
-The downside which comes with this is that "island solutions" are being created, i.e. small packages which serve a particular (sometimes very niche) use case.
-While the UNIX philosophy "Do one thing and do it well" [@raymond2003art] is widely known and used in the R community, R packages from different developers do often not play well with each other.
-This is due to differing input and returns objects, requiring the need to pre-and postprocess objects around such R packages which provide tools for specific modeling tasks.
-Also often R packages implement only an opinionated subset of methods which require users to combine multiple packages for different steps.
+etc
 
-These points may lead to the following possible issues in practice:
+A variant of cross-validation is leave-one-out (LOO) cross-validation , which has long been established in geostatistics [@cressie2015], sometimes with a focus on the spatial distribution of LOO error [@willmott2006].
+Since this is just a special case of non-spatial CV with $k=n$, it is sometimes also referred to as spatial CV [@willmott2006]. 
+Nevertheless, spatial variants have been proposed that apply an exclusion buffer to the test locations to separate them from the training data [@brenning2005]. 
+One plausible approach for this is to use the range of autocorrelation of model residuals to determine the buffer distance as this ensures independence conditional on the predictors [@brenning2005]. 
+When data is grouped, e.g., due to multi-level sampling designs or the study of spatial objects, LOO at has been proposed at the site level [@martin2008; @kasurak2011] or, in animal movement studies, at the animal level [@anderson2005].
+Similarly, in the temporal domain, it has been proposed to leave out temporal observational units, as in leave-one-year-out CV [@anderson2005; @brenning2005].
 
-- Differing syntax.
-- Support for only some predictor types (e.g. numeric and factor features but no ordered factors).
-- Differing return values.
-- Differing namings for the same method.
-- Unclear longterm maintenance.
-
-These points add substantial overhead and/or reproducibility issues for anyone who wants to make use of the provided implementations within their studies.
-\pkg{mlr3spatiotempcv} aims to provide a solution for some of the points outlined above by wrapping all spatiotemporal resampling methods available in \proglang{R} packages.
-By embedding the package within the \pkg{mlr3} ecosystem, \pkg{mlr3spatiotempcv} aims to provide longterm support for spatiotemporal partitioning methods in R while making all additional benefits of the machine learning framework available to users.
-
-In addition, \pkg{mlr3spatiotempcv} adds 2D and 3D visualization possibilities for all implemented methods.
-The authors believe that the possibility for creating graphical representations of partitions is important, especially if spatio-temporal clustering or other grouping types were used to create the groups.
-
-## The mlr3 ecosystem
-
-The \pkg{mlr3} R package and its extensions packages form the \pkg{mlr3} machine learning ecosystem [@mlr3].
-\pkg{mlr3} aims to wrap many algorithms available in R to enforce a consistent interface for such.
-This adds the ability for simplified benchmarking, visualization and further model evaluation regardless of the chosen algorithm.
-
-\pkg{mlr3spatiotempcv} complements this philosophy by aiming to apply the same standards for spatiotemporal resampling methods.
+It should be noted that $k$-fold cross-validation with a large value of $k$, and LOO-CV in particular ($k=n$), are not only very time-consuming since the model has to be trained $k$ times; these models will also be nearly identical since only a tiny fraction of the data is withheld, and therefore estimation bias increases.
+'Pure' LOO-CV is therefore not recommended for machine-learning model assessment [ref??].
 

### 2021-05-05 paper/paper.Rmd
mlr3 ecosystem section

@@
 'Pure' LOO-CV is therefore not recommended for machine-learning model assessment [ref??].
+
+# mlr3spatiotempcv within the mlr3 ecosystem
+
+With the increased awareness of the importance of spatial and spatio-temporal resampling strategies and the growing popularity of R in environmental modelling and geocomputation, it is important to equip machine-learning frameworks such as \pkg{mlr} with suitable algorithms.
+In this context, \pkg{mlr3} stands out as a unified, object-oriented and extendable framework designed to accommodate numerous machine-learning tasks with a variety of learners, feature and model selection tools, and model assessment capabilities [@mlr3; mlr3book].
+All of these are supported by advanced visualization tools, which are particularly important in a spatial and spatio-temporal setting.
+
+With its integrative approach and its aim to provide long-term support, \pkg{mlr3} overcomes the challenges of combining multiple specialized packages with poorly standardized interfaces.
+Issues that practitioners often face include varying argument lists of learners, different return values of `predict` methods, and support for only specific feature types. 
+These challenges result in substantial overhead and possible reproducibility issues, which are exacerbated by asynchronous development timelines of different components of the machine-learning pipelines.
 

### 2021-05-06 paper/paper.Rmd
rough draft paragraph on object architecture/interfaces, issue #126

@@
 
-etc
+- add a paragraph with more information on spCV using spatial partitionings
 
@@
 
+- add a paragraph on spatiotemporal partitioning
+
 # mlr3spatiotempcv within the mlr3 ecosystem
@@
 These challenges result in substantial overhead and possible reproducibility issues, which are exacerbated by asynchronous development timelines of different components of the machine-learning pipelines.
+
+- *add one/two paragraphs outlining mlr3's object model and mlr3psatiotempcv's place within. someting along the lines of:*
+
+Within \pkg{mlr}, partitioning strategies are represented by resampling objects (correct?) of class xyz, which are generated within mlr3 itself (for regular random CV, correct??) or within specialized extensions such as mlr3spatiotempcv.
+In the ML pipeline, these objects are used for model assessment by ...
+objects/tasks/whatever and for model selection by ...
+objects/tasks/..., providing a consistent framework for both tasks that is necessary to optimize learners for a given prediction problem [@schratz2019].
+
+Resampling objects in mlr3spatiotmpcv inherit from and can be created from established object classes for geospatial data in R, including simple features (add a general reference, maybe Lovelace), which facilitates its integration into domain-specific workflows in the geospatial sciences.
+This includes support for geographic coordinate reference systems and their respective distance measures in projected planar space or on the unprojected ellipsoid (corrrect?).
+Of course, non-spatial data tables with coordinates as variables are also supported.
 

### 2021-05-06 paper/paper.Rmd
rephrased intro of section on partitioning methods

@@
 
-## Spatial
-
-### Buffering
-
-At the time of writing, the following spatiotemporal resampling methods (Tab. \@ref(tab:sptcv-methods)) were available across different R packages.
-This list might increase over time and it is recommended to check the homepage of \pkg{mlr3spatiotempcv} for an up-to-date list.
-In Tab. \@ref(tab:sptcv-methods), the internal identifier for the respective method in \pkg{mlr3spatiotempcv} is shown in column "Notation".
-Some of the methods listed are able to produce different partitioning results by setting different options, hence the actual number of available methods is higher than the one shown in Tab.
-\@ref(tab:sptcv-methods).
+At the time of writing, \pkg{mlr3spatiotempcv} implements the partitioning methods identified in Table \@ref(tab:sptcv-methods) with their \pkg{mlr3} class name. 
+Several of the implemented algorithms are themselves versatile toolboxes with multiple options.
+Comprehensive and up-to-date information can be found in the package's online documentation.
+The following sections give an overview of most existing partitioning strategies and their visualization.
+The available methods are futher discussed in section \@ref(sec:disc).
 
@@
 
-A low-level introduction including visualization is given for every method in the following.
-See section \@ref(sec:disc) for a discussion of the methods.
+## Spatial
+
+### Buffering
 

### 2021-05-06 paper/paper.Rmd
spatio-temporal -> spatiotemporal (as this post convinced me: https://english.stackexchange.com/questions/490525/spatio-or-spatial )

@@
   It adds support for spatiotemporal resampling methods and respective visualization methods.
-  Spatiotemporal resampling methods are needed to account for the existence of different autocorrelation types (spatial autocorrelation (SAC), temporal autocorrelation (TAC) or spatio-temporal autocorrelation (STAC)) in datasets.
+  Spatiotemporal resampling methods are needed to account for the existence of different autocorrelation types (spatial autocorrelation (SAC), temporal autocorrelation (TAC) or spatiotemporal autocorrelation (STAC)) in datasets.
   These autocorrelation types cause biased performance estimates of varying magnitude in cross-validation when ignored.
@@
 
-Spatial and spatio-temporal prediction tasks are common in applications ranging from environmental sciences to archaeology and epidemiology. 
-While sophisticated mathematical frameworks have long been developed in spatial statistics to characterize predictive uncertainties under well-defined mathematical assumptions such as intrinsic stationarity [e.g., cressie2015], computational estimation procedures have only been proposed more recently to assess predictive performances of spatial and spatio-temporal prediction models [@brenning2005; @sperrorest; @pohjankukka2017; @roberts2017].
+Spatial and spatiotemporal prediction tasks are common in applications ranging from environmental sciences to archaeology and epidemiology. 
+While sophisticated mathematical frameworks have long been developed in spatial statistics to characterize predictive uncertainties under well-defined mathematical assumptions such as intrinsic stationarity [e.g., cressie2015], computational estimation procedures have only been proposed more recently to assess predictive performances of spatial and spatiotemporal prediction models [@brenning2005; @sperrorest; @pohjankukka2017; @roberts2017].
 
@@
 
-While it is not the purpose of this paper to recommend specific resampling schemes for specific use cases, this example may suffice to motivate the use of spatial and spatio-temporal cross-validation techniques, and the need for a unified framework and computational toolbox that accommodate a variety of prediction tasks that may be applicable to a broad range of prediction tasks.
+While it is not the purpose of this paper to recommend specific resampling schemes for specific use cases, this example may suffice to motivate the use of spatial and spatiotemporal cross-validation techniques, and the need for a unified framework and computational toolbox that accommodate a variety of prediction tasks that may be applicable to a broad range of prediction tasks.
 \pkg{mlr3spatiotempcv} is such a toolbox.
@@
 The earliest of these implementations is the \pkg{sperrorest} package [@sperrorest], which has, however, not been integrated into established machine-learning frameworks such as \pkg{mlr} or \pkg{caret}, and lacks support of temporal prediction tasks.
-Other more recent implementations include \pkg{blockCV} for (what for?) [ref @blockCV; @rest2014] and \pkg{CAST} for spatio-temporal prediction tasks in a \pkg{caret} framework [@cast; @meyer2018a].
+Other more recent implementations include \pkg{blockCV} for (what for?) [ref @blockCV; @rest2014] and \pkg{CAST} for spatiotemporal prediction tasks in a \pkg{caret} framework [@cast; @meyer2018a].
 
-Thus, \pkg{mlr3spatiotempcv} implements for the first time a comprehensive state-of-the-art compilation of spatial and spatio-temporal partitioning schemes that is well-integrated into a major machine-learning framework in R, the \pkg{mlr3} ecosystem [@mlr3].
+Thus, \pkg{mlr3spatiotempcv} implements for the first time a comprehensive state-of-the-art compilation of spatial and spatiotemporal partitioning schemes that is well-integrated into a major machine-learning framework in R, the \pkg{mlr3} ecosystem [@mlr3].
 This package is furthermore equipped with a variety of two- and three-dimensional diagram types.
@@
 
-# Spatial and Spatio-temporal Cross-validation
+# Spatial and Spatiotemporal Cross-validation
 
@@
 
-With the increased awareness of the importance of spatial and spatio-temporal resampling strategies and the growing popularity of R in environmental modelling and geocomputation, it is important to equip machine-learning frameworks such as \pkg{mlr} with suitable algorithms.
+With the increased awareness of the importance of spatial and spatiotemporal resampling strategies and the growing popularity of R in environmental modelling and geocomputation, it is important to equip machine-learning frameworks such as \pkg{mlr} with suitable algorithms.
 In this context, \pkg{mlr3} stands out as a unified, object-oriented and extendable framework designed to accommodate numerous machine-learning tasks with a variety of learners, feature and model selection tools, and model assessment capabilities [@mlr3; mlr3book].
-All of these are supported by advanced visualization tools, which are particularly important in a spatial and spatio-temporal setting.
+All of these are supported by advanced visualization tools, which are particularly important in a spatial and spatiotemporal setting.
 
@@
 
-## Comparing spatio-temporal resampling methods
+## Comparing spatiotemporal resampling methods
 
@@
 <!-- missing: Guidelines how to choose a resampling method -->
-While various spatio-temporal resampling methods exists nowadays to account for spatio-temporal autocorrelation in modeling, there is still lots of uncertainties which method should be used for which dataset.
+While various spatiotemporal resampling methods exists nowadays to account for spatiotemporal autocorrelation in modeling, there is still lots of uncertainties which method should be used for which dataset.
 The answer to this questions is not trivial and needs more research.
@@
 The latter often require special structured objects as inputs and cannot directly be re-used in susequent modeling steps of user workflows.
-This is not an issue specific to spatio-temporal resampling methods but a general philosophy of the authors.
+This is not an issue specific to spatiotemporal resampling methods but a general philosophy of the authors.
 
-Spatio-temporal cross-validation is still not yet a paradigm that is completely established in scientfic workflows even though it is discussed intensively for more than a decade now.
+Spatiotemporal cross-validation is still not yet a paradigm that is completely established in scientfic workflows even though it is discussed intensively for more than a decade now.
 Making most available methods easily accessible to users is an important step to foster the acceptance of such in the community and lead to less biased cross-validation results in environmental and ecological studies.

### 2021-05-06 paper/paper.Rmd
contributed mainly to former visualization section and to case study section

@@
 
-- *add one/two paragraphs outlining mlr3's object model and mlr3psatiotempcv's place within. someting along the lines of:*
-
 Within \pkg{mlr}, partitioning strategies are represented by resampling objects (correct?) of class xyz, which are generated within mlr3 itself (for regular random CV, correct??) or within specialized extensions such as mlr3spatiotempcv.
@@
 
-Resampling objects in mlr3spatiotmpcv inherit from and can be created from established object classes for geospatial data in R, including simple features (add a general reference, maybe Lovelace), which facilitates its integration into domain-specific workflows in the geospatial sciences.
-This includes support for geographic coordinate reference systems and their respective distance measures in projected planar space or on the unprojected ellipsoid (corrrect?).
-Of course, non-spatial data tables with coordinates as variables are also supported.
+Resampling objects in \pkg{mlr3spatiotmpcv} inherit from and can be created from established object classes for geospatial data in R, including simple features (add a general reference, maybe Lovelace), which facilitates its integration into domain-specific workflows in the geospatial sciences.
+This includes support for geographic coordinate reference systems (CRS) and their respective distance measures in projected planar space or using great circle distances on the unprojected ellipsoid (corrrect???).
+Non-spatial data tables with coordinates as variables are also supported.
+If no projection is specified, World Geodetic System 1984 (WGS84, EPSG: 4326) will be assumed.
 
-# Visualization of spatiotemporal partitions
+Partitioning objects in \pkg{mlr3spatiotempcv} are equipped with generic `plot()` and `autoplot()` methods that visualize the created groups.
+`autoplot()` is \pkg{ggplot2}-based and uses [ggplot2](https://ggplot2.tidyverse.org) [@ggplot2] in two-dimensional geographic space and [plotly](https://github.com/ropensci/plotly) [@plotly] in three dimensions, i.e.
+geographic space plus time.
 
-Every partitioning method in [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) comes with generic `plot()` and `autoplot()` (\pkg{ggplot2} equivalent) method to visualize the created groups.
-In a 2D space this happens via package [ggplot2](https://ggplot2.tidyverse.org) [@ggplot2] while for spatiotemporal methods 3D visualizations are created via package [plotly](https://github.com/ropensci/plotly) [@plotly].
+# Application in a Case Study
 
-Visualization of partitioning methods is a vital part of \pkg{mlr3spatiotempcv} as most upstream packages lack support for such.
-Hence, plotting semantics are presented first because plotting will be used extensively when presenting the integrated methods.
+We will use a well-known case study to demonstrate the application of spatial resampling techniques. 
+This binary classification task on landslide susceptibility in Ecuador [@muenchow2012] is available as a built-in task via `tsk("ecuador")`.
 
-Unless specified by the user, the coordinate reference system (CRS) defaults to EPSG code 4326 (WGS84).
-This is because a lat/lon based CRS is better suited for plotting purposes than a Mercator (UTM) one due tho a more descriptive axes labeling.
-Setting an appropriate CRS for the given data *during task construction* is very important.
-Even though EPSG 4326 is a good fallback and often used for visualization purposes, spatial offsets of up to multiple meters may occur if the wrong CRS was passed initially.
-
-The following example creates a spatiotemporal task from scratch.
-This task is also available as a built-in task via `tsk("ecuador")`.
+...
 
@@
 Several of the implemented algorithms are themselves versatile toolboxes with multiple options.
-Comprehensive and up-to-date information can be found in the package's online documentation.
+Comprehensive and up-to-date information can be found in the package's online documentation (<https://mlr3spatiotempcv.mlr-org.com>).
 The following sections give an overview of most existing partitioning strategies and their visualization.

### 2021-05-06 paper/paper.Rmd
contribued to case study section

@@
 
-To show the effect of spatial autocorrelation on performance estimates in cross-validation, a spatial and non-spatial CV using method `"spatial_coords"` will be shown in the following.
-
-The performance of a simple classification tree (`"classif.ranger"`) is evaluated on a repeated cross-validation (`"repeated_cv"`) with four folds and two repetitions.
-In reality, more fold (between five and ten) and a higher number of repetitions should be used.
-The chosen evaluation measure is the "Area Under the ROC Curve" (`"classif.auc"`) which is a common evaluation measure for binary responses.
-The CV will be executed twice with the only difference being the use of a spatial resampling object (initialized with `rsmp("repeated_spcv_coords"`) instead of a non-spatial one (`rsmp("repeated_cv"`).
+To show the effect of spatial autocorrelation on performance estimates in cross-validation, we will apply spatial and non-spatial cross-validation to the prediction of landslides using the random forest technique (`"classif.ranger"`).
+Specifically, we use spatial partitioning based on $k$-means clustering of the coordinates (method `rsmp("repeated_spcv_coords")`) and compare its results to a non-spatial (random) cross-validation (`rsmp("repeated_cv")`), both with (only) four folds and two repetitions for demonstration.
+We choose the area Under the ROC curve (`"classif.auc"`) as the performance measure for this 'soft' classification task.
 

### 2021-05-08 paper/paper.Rmd
refine ' mlr3spatiotempcv within the mlr3 ecosystem' section' (#126)'

@@
 
-Within \pkg{mlr}, partitioning strategies are represented by resampling objects (correct?) of class xyz, which are generated within mlr3 itself (for regular random CV, correct??) or within specialized extensions such as mlr3spatiotempcv.
-In the ML pipeline, these objects are used for model assessment by ...
-objects/tasks/whatever and for model selection by ...
-objects/tasks/..., providing a consistent framework for both tasks that is necessary to optimize learners for a given prediction problem [@schratz2019].
+Within the \pkg{mlr3} ecosystem, partitioning strategies are represented by their own objects of class \texttt{Resampling}, of which most are available within \pkg{mlr3} itself (e.g.
+regular random CV) and other specialized ones (e.g.
+\texttt{TaskClassifST} or \texttt{TaskRegrST}) are defined in extension packages such as \pkg{mlr3spatiotempcv}.
+In the ML pipeline, these objects define the data splits used for model evaluation and selection (hyperparameter tuning) by ML algorithms.
+They play an important role for an unbiased application of ML algorithms [@schratz2019].
 
-Resampling objects in \pkg{mlr3spatiotmpcv} inherit from and can be created from established object classes for geospatial data in R, including simple features (add a general reference, maybe Lovelace), which facilitates its integration into domain-specific workflows in the geospatial sciences.
-This includes support for geographic coordinate reference systems (CRS) and their respective distance measures in projected planar space or using great circle distances on the unprojected ellipsoid (corrrect???).
+Resampling objects in \pkg{mlr3spatiotmpcv} inherit from class \texttt{mlr3::Resampling} and can be created from established object classes for geospatial data in R, including simple features [@pebesma2018], which facilitates its integration into domain-specific workflows in the geospatial sciences.
+This includes support for geographic coordinate reference systems (CRS) and their respective distance measures in projected planar space or using great circle distances on the unprojected ellipsoid.
 Non-spatial data tables with coordinates as variables are also supported.
@@
 
-```{r create task}
+```{r create-task}
 library(mlr3)

### 2021-05-08 paper/paper.Rmd
fix all references and upd paper.bib

@@
   These autocorrelation types cause biased performance estimates of varying magnitude in cross-validation when ignored.
-  At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: CAST [@cast], blockCV [@blockCV] or sperrorest [@sperrorest].
+  At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: CAST [@cast], blockCV [@blockCV] or sperrorest [@brenning2012].
   There are some resampling methods used in scientific studies for which the source code is either not available or a different programming language has been used [@roberts2017; @pohjankukka2017]
@@
 Spatial and spatiotemporal prediction tasks are common in applications ranging from environmental sciences to archaeology and epidemiology. 
-While sophisticated mathematical frameworks have long been developed in spatial statistics to characterize predictive uncertainties under well-defined mathematical assumptions such as intrinsic stationarity [e.g., cressie2015], computational estimation procedures have only been proposed more recently to assess predictive performances of spatial and spatiotemporal prediction models [@brenning2005; @sperrorest; @pohjankukka2017; @roberts2017].
+While sophisticated mathematical frameworks have long been developed in spatial statistics to characterize predictive uncertainties under well-defined mathematical assumptions such as intrinsic stationarity [e.g., [@cressie1993], computational estimation procedures have only been proposed more recently to assess predictive performances of spatial and spatiotemporal prediction models [@brenning2005, @brenning2012; @pohjankukka2017; @roberts2017].
 
@@
 This toolbox, implemented as an open-source R package, builds upon and generalizes several existing toolboxes that have been developed in recent years for more specific settings (see Table \@ref(tab:sptcv-methods)).
-The earliest of these implementations is the \pkg{sperrorest} package [@sperrorest], which has, however, not been integrated into established machine-learning frameworks such as \pkg{mlr} or \pkg{caret}, and lacks support of temporal prediction tasks.
+The earliest of these implementations is the \pkg{sperrorest} package [@brenning2012], which has, however, not been integrated into established machine-learning frameworks such as \pkg{mlr} or \pkg{caret}, and lacks support of temporal prediction tasks.
 Other more recent implementations include \pkg{blockCV} for (what for?) [ref @blockCV; @rest2014] and \pkg{CAST} for spatiotemporal prediction tasks in a \pkg{caret} framework [@cast; @meyer2018a].
@@
 where $l$ is a real-valued loss function, and the expected value is with respect the probability distribution of $X$, the features of an instance $(Y,X)$ drawn randomly from the underlying population.
-This is referred to as the *actual* or *conditional* performance measure as it is conditional on $L$ [@hand1996]. 
+This is referred to as the *actual* or *conditional* performance measure as it is conditional on $L$ [@milligan2000]. 
 The loss function can take a variety of forms such as the misclassification error $I(Y\neq\hat{f}_L(X))$ in classification, or the squared error $(Y-\hat{f}_L(X))^2$ in regression, among many other possible measures.
@@
 
-A variant of cross-validation is leave-one-out (LOO) cross-validation , which has long been established in geostatistics [@cressie2015], sometimes with a focus on the spatial distribution of LOO error [@willmott2006].
+A variant of cross-validation is leave-one-out (LOO) cross-validation , which has long been established in geostatistics [@cressie1993], sometimes with a focus on the spatial distribution of LOO error [@willmott2006].
 Since this is just a special case of non-spatial CV with $k=n$, it is sometimes also referred to as spatial CV [@willmott2006]. 
@@
 
-With the increased awareness of the importance of spatial and spatiotemporal resampling strategies and the growing popularity of R in environmental modelling and geocomputation, it is important to equip machine-learning frameworks such as \pkg{mlr} with suitable algorithms.
-In this context, \pkg{mlr3} stands out as a unified, object-oriented and extendable framework designed to accommodate numerous machine-learning tasks with a variety of learners, feature and model selection tools, and model assessment capabilities [@mlr3; mlr3book].
+With the increased awareness of the importance of spatial and spatiotemporal resampling strategies and the growing popularity of R in environmental modeling and geocomputation, it is important to equip machine-learning frameworks such as \pkg{mlr3} with suitable algorithms.
+In this context, the \pkg{mlr3} ecosystem stands out as a unified, object-oriented and extendable framework designed to accommodate numerous machine-learning tasks with a variety of learners, feature and model selection tools, and model assessment capabilities [@mlr3; @mlr3book].
 All of these are supported by advanced visualization tools, which are particularly important in a spatial and spatiotemporal setting.
@@
 
-The "Spatial CV" method after [@sperrorest] and implemented in package \pkg{sperrorest} is similar to method "environmental blocking" in that it also uses *k-means* clustering for fold creation.
+The "Spatial CV" method after [@brenning2012] and implemented in package \pkg{sperrorest} is similar to method "environmental blocking" in that it also uses *k-means* clustering for fold creation.
 However, it does not make use of arbitrary features but uses the coordinates of all observations to create clusters in the spatial domain.
@@
 
-Leave-Location-and-Time-Out (LLTO) from packge \pkg{CAST} is a spatiotemporal resampling method [@meyer2018a].
+Leave-Location-and-Time-Out (LLTO) from packge \pkg{CAST} is a spatiotemporal resampling method [@meyer2018].
 It requires both a spatial and a temporal variable which are used to to separate observations in a multi-dimensional space.
@@
 <!-- spcv coords: mention generic name -->
-In @sperrorest, which is commonly used as the main reference for this k-means based spatial clustering approach, the method was not assigned a unique name.
+In @brenning2012, which is commonly used as the main reference for this k-means based spatial clustering approach, the method was not assigned a unique name.
 This brings potential confusion for users and the community as the term "spatial CV" is used a generic one for the idea of accounting for spatial autocorrelation for spatial datasets.

### 2021-05-16 paper/paper.Rmd
update references

@@
 
-Although alternatives such as the bootstrap exist and have some advantages [@efron1983; @hand1996], cross-validation is a particularly well-established, easy-to-implement algorithm for *model assessment* of supervised machine-learning models [@efron1983 and next section] and *model selection* [@arlot2010].
+Although alternatives such as the bootstrap exist and have some advantages [@efron1983; @milligan2000], cross-validation is a particularly well-established, easy-to-implement algorithm for *model assessment* of supervised machine-learning models [@efron1983 and next section] and *model selection* [@arlot2010].
 In its basic form, it is based on resampling the data without paying attention to any possible dependence structure, which may arise from, e.g., cluster sampling [ref], grouped data [@brenning2008; @pena2015], or underlying environmental processes inducing some sort of spatial coherence at the landscape scale [@brenning2005].
@@
 The loss function can take a variety of forms such as the misclassification error $I(Y\neq\hat{f}_L(X))$ in classification, or the squared error $(Y-\hat{f}_L(X))^2$ in regression, among many other possible measures.
-The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance metrics for regression and classification [see e.g., @hand1996 for classification, and ref??
+The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance metrics for regression and classification [see e.g., @milligan2000 for classification, and ref??
 for regression tasks].
@@
 
-```{r block-checkerboard, fig.cap="Spatial blocking with option `selection = 'checkerboard'`.", out.width="40%", fig.pos="h!"}
+```{r block-checkerboard, fig.cap="Spatial blocking with option `selection = 'checkerboard'`.", out.width="40%", fig.pos="ht!"}
 resampling_block_checkerboard = rsmp("spcv_block",

### 2021-05-16 paper/paper.Rmd
add literature references to spCV methods table

@@
   \usepackage{placeins}
+  \usepackage{scrextend}
+  \usepackage{tablefootnote}
   \newcommand{\specialcell}[2][l]{%
@@
       \toprule
-      \multirow{6}{*}{Spatial}        & Spatial Buffering           & \specialcell{\pkg{blockCV} \\ \pkg{sperrorest}}    & \specialcell{ \texttt{"spcv\_buffer"} \\  \texttt{"spcv\_disc"}}  & Lit1       \\
+      \multirow{6}{*}{Spatial}        & Spatial Buffering           & \specialcell{\pkg{blockCV} \\ \pkg{sperrorest}}    & \specialcell{ \texttt{"spcv\_buffer"} \\  \texttt{"spcv\_disc"} \tablefootnote{\label{endicott2017}\cite{endicott2017}} }  & \tablefootnote{\label{ploton2020}\cite{ploton2020}}  \tablefootnote{\label{diesing2020}\cite{diesing2020}}      \\
       \cmidrule{2-5}
-                                      & Spatial Blocking            & \specialcell{\pkg{blockCV} \\ \pkg{sperrorest}}  & \specialcell{  \texttt{"spcv\_block"} \\  \texttt{"spcv\_tiles"}}
-                                        & Lit1       \\
+                                      & Spatial Blocking            & \specialcell{\pkg{blockCV} \\ \pkg{sperrorest}}  & \specialcell{  \texttt{"spcv\_block"} \\  \texttt{"spcv\_tiles"} \tablefootnote{\label{bebber2017}\cite{bebber2017}} \tablefootnote{\label{brenning2015}\cite{brenning2015}} \tablefootnote{\label{zurell2020}\cite{zurell2020}} }
+                                        & \tablefootnote{\label{jensen2021}\cite{jensen2021}} \tablefootnote{\label{escobar2021}\cite{escobar2021}} \tablefootnote{\label{stewart2021}\cite{stewart2021}} \tablefootnote{\label{cerasoli2021}\cite{cerasoli2021}}        \\
   \cmidrule{2-5}
-                                      & Spatial Clustering          & \pkg{sperrorest} & \texttt{"spcv\_coords"} & Lit1       \\
+                                      & Spatial Clustering          & \pkg{sperrorest} & \texttt{"spcv\_coords"} & \tablefootnote{\label{note1}\cite{brenning2012}} \tablefootnote{\label{morera2021}\cite{morera2021}} \tablefootnote{\label{geiss2017}\cite{geiss2017}} \tablefootnote{\label{wu2020}\cite{wu2020}}      \\
                                       \cmidrule{2-5}
-       & Leave-Location-Out & \specialcell{\pkg{CAST} \\ \pkg{mlr3}}       & \specialcell{ \texttt{"sptcv\_cstf('space\_var')"} \\ \texttt{"FIXME"}},  & Lit1       \\
+       & Leave-Location-Out & \specialcell{\pkg{CAST} \\ \pkg{mlr3}}       & \specialcell{ \texttt{"sptcv\_cstf('space\_var')"} \\ \texttt{"FIXME"}},  & \footref{note1}       \\
       \midrule
-      \multirow{2}{*}{Spatiotemporal} & Leave-Location-and-Time-Out & \pkg{CAST}       &  \texttt{"sptcv\_cstf"}  & Lit1       \\
-                                      & Spatiotemporal Clustering   & \pkg{skmeans}    &  \texttt{"sptcv\_cluto"} & Lit1       \\
+      \multirow{2}{*}{Spatiotemporal} & Leave-Location-and-Time-Out & \pkg{CAST}       &  \texttt{"sptcv\_cstf"}  & \\
+                                      & Spatiotemporal Clustering   & \pkg{skmeans}    &  \texttt{"sptcv\_cluto"} &        \\
       \midrule
-      Feature space                   & Environmental Blocking      & \pkg{blockCV}    &  \texttt{"spcv\_env"}    & Lit1       \\
+      Feature space                   & Environmental Blocking      & \pkg{blockCV}    &  \texttt{"spcv\_env"}    & \footref{morera2021}     \\
+
     \end{tabular}

### 2021-05-20 paper/paper.Rmd
update resampling methods table

--- 2021-05-16 paper/paper.Rmd
@@
   \footnotesize
-  \caption[t]{Spatiotemporal resampling methods, ordered alphabetically by class name (Notation column). }
+  \caption[t]{Spatiotemporal resampling methods, ordered alphabetically by class name (Notation column).
+Column count refers to the number of studies which were found to have used this resampling technique until 2021-05.}
   \begingroup
@@
       \\
-      Type                            & Name                        & R Package(s)     & mlr3 Notation  & Literature \\
+      Type                            & Name                        & R Package(s)     & mlr3 Notation  & Count\\
       \toprule
-      \multirow{6}{*}{Spatial}        & Spatial Buffering           & \specialcell{\pkg{blockCV} \\ \pkg{sperrorest}}    & \specialcell{ \texttt{"spcv\_buffer"} \\  \texttt{"spcv\_disc"}}  & Lit1       \\
+      \multirow{6}{*}{Spatial}        & Spatial Buffering           & \specialcell{\pkg{blockCV} \\ \pkg{sperrorest}}    & \specialcell{ \texttt{"spcv\_buffer"} \tablefootnote{\label{ploton2020}\cite{ploton2020}}  \tablefootnote{\label{diesing2020}\cite{diesing2020}} \\  \texttt{"spcv\_disc"} \tablefootnote{\label{endicott2017}\cite{endicott2017}} \tablefootnote{\label{karasiak2021}\cite{karasiak2021}} \tablefootnote{\label{moller2021}\cite{moller2021}}}  & \specialcell{2 \\ 3}   \\
       \cmidrule{2-5}
-                                      & Spatial Blocking            & \specialcell{\pkg{blockCV} \\ \pkg{sperrorest}}  & \specialcell{  \texttt{"spcv\_block"} \\  \texttt{"spcv\_tiles"}}
-                                        & Lit1       \\
+                                      & Spatial Blocking            & \specialcell{\pkg{blockCV}  \\ \pkg{sperrorest}}  & \specialcell{  \texttt{"spcv\_block"} \tablefootnote{\label{jensen2021}\cite{jensen2021}} \tablefootnote{\label{escobar2021}\cite{escobar2021}} \tablefootnote{\label{stewart2021}\cite{stewart2021}} \\  \texttt{"spcv\_tiles"} \tablefootnote{\label{bebber2017}\cite{bebber2017}} \tablefootnote{\label{brenning2015}\cite{brenning2015}} \tablefootnote{\label{zurell2020}\cite{zurell2020}} }
+                                        & \specialcell{28 \\ 4} \\
   \cmidrule{2-5}
-                                      & Spatial Clustering          & \pkg{sperrorest} & \texttt{"spcv\_coords"} & Lit1       \\
+                                      & Spatial Clustering          & \pkg{sperrorest} & \texttt{"spcv\_coords"} \tablefootnote{\label{morera2021}\cite{morera2021}} \tablefootnote{\label{geiss2017}\cite{geiss2017}} \tablefootnote{\label{wu2020}\cite{wu2020}} &  21 \\
                                       \cmidrule{2-5}
-       & Leave-Location-Out & \specialcell{\pkg{CAST} \\ \pkg{mlr3}}       & \specialcell{ \texttt{"sptcv\_cstf('space\_var')"} \\ \texttt{"FIXME"}},  & Lit1       \\
+       & Predefined grouping & \specialcell{\pkg{CAST} \\ \pkg{mlr3}}       & \specialcell{ \texttt{"sptcv\_cstf"} \tablefootnote{\label{blothe2021}\cite{blothe2021}} \tablefootnote{\label{dolan2021}\cite{dolan2021}} \tablefootnote{\label{dumke2019}\cite{dumke2019}}  \\ \texttt{"FIXME"}}  & \specialcell{12 \\ -}       \\
       \midrule
-      \multirow{2}{*}{Spatiotemporal} & Leave-Location-and-Time-Out & \pkg{CAST}       &  \texttt{"sptcv\_cstf"}  & Lit1       \\
-                                      & Spatiotemporal Clustering   & \pkg{skmeans}    &  \texttt{"sptcv\_cluto"} & Lit1       \\
+      \multirow{2}{*}{Spatiotemporal} & Leave-Location-and-Time-Out & \pkg{CAST}       &  \texttt{"sptcv\_cstf"} \tablefootnote{\label{egli2020}\cite{egli2020}} \tablefootnote{\label{gao2019}\cite{gao2019}} \tablefootnote{\label{reitz2021}\cite{reitz2021}} & 6 \\
+                                      & Spatiotemporal Clustering   & \pkg{skmeans}    &  \texttt{"sptcv\_cluto"} &  -  \\
       \midrule
-      Feature space                   & Environmental Blocking      & \pkg{blockCV}    &  \texttt{"spcv\_env"}    & Lit1       \\
+      Feature space                   & Environmental Blocking      & \pkg{blockCV}    &  \texttt{"spcv\_env"} \footref{morera2021}   & 1    \\
+
     \end{tabular}

### 2021-05-22 paper/paper.Rmd
refine introduction, add refs

@@
 Spatial and spatiotemporal prediction tasks are common in applications ranging from environmental sciences to archaeology and epidemiology.
-While sophisticated mathematical frameworks have long been developed in spatial statistics to characterize predictive uncertainties under well-defined mathematical assumptions such as intrinsic stationarity [e.g., [@cressie1993], computational estimation procedures have only been proposed more recently to assess predictive performances of spatial and spatiotemporal prediction models [@brenning2005, @brenning2012; @pohjankukka2017; @roberts2017].
+While sophisticated mathematical frameworks have long been developed in spatial statistics to characterize predictive uncertainties under well-defined mathematical assumptions such as intrinsic stationarity, e.g.
+[@cressie1993], computational estimation procedures have only been proposed more recently to assess predictive performances of spatial and spatiotemporal prediction models [@brenning2005, @brenning2012; @pohjankukka2017; @roberts2017].
 
 Although alternatives such as the bootstrap exist and have some advantages [@efron1983; @milligan2000], cross-validation is a particularly well-established, easy-to-implement algorithm for *model assessment* of supervised machine-learning models [@efron1983 and next section] and *model selection* [@arlot2010].
-In its basic form, it is based on resampling the data without paying attention to any possible dependence structure, which may arise from, e.g., cluster sampling [ref], grouped data [@brenning2008; @pena2015], or underlying environmental processes inducing some sort of spatial coherence at the landscape scale [@brenning2005].
+In its basic form, it is based on resampling the data without paying attention to any possible dependence structure, which may arise from, e.g., cluster sampling [@henderson1982], grouped data [@brenning2008; @pena2015], or underlying environmental processes inducing some sort of spatial coherence at the landscape scale [@brenning2005].
 In treating dependent observations as independent, or ignoring autocorrelation, cross-validation test samples may in fact be heavily correlated with, or even pseudo-replicates of, the data used for training the model, which introduces a potentially severe bias, especially for flexible machine-learning models [@brenning2005].
@@
 2000 grid cells from 100 fields scattered across a large study region.
-The purpose of training a model on this sample is to make predictions on other, new *fields* within the same geographic domain [*intra-domain* prediction; @brenning2005] &emdash; not *within* the same field, which obviously presents only one crop type.
+The purpose of training a model on this sample is to make predictions on other, new *fields* within the same geographic domain [*intra-domain* prediction; @brenning2005] \textemdash; not *within* the same field, which obviously presents only one crop type.
 It would therefore be foolish to train a model on a simple random subsample of grid cells, and test it on the remaining data, i.e.
-using other grid cells from the same fields, as if we wanted to predict within a field; the results from this performance assessment would be over-optimitic, and perhaps badly so.
+using other grid cells from the same fields, as if we wanted to predict within a field.
+The results from this performance assessment would be over-optimistic, and perhaps badly so.
 To mimic the predictive situation for which the model is trained, we'd rather have to resample at the level of fields, not grid cells [@pena2015].
@@
 The earliest of these implementations is the \pkg{sperrorest} package [@brenning2012], which has, however, not been integrated into established machine-learning frameworks such as \pkg{mlr} or \pkg{caret}, and lacks support of temporal prediction tasks.
-Other more recent implementations include \pkg{blockCV} for (what for?) [ref @blockCV; @rest2014] and \pkg{CAST} for spatiotemporal prediction tasks in a \pkg{caret} framework [@cast; @meyer2018a].
+Other more recent implementations include \pkg{blockCV} for spatial partitioning methods tailored towards species distribution modeling [@blockCV; @rest2014] and \pkg{CAST} for spatiotemporal prediction tasks in a \pkg{caret} framework [@cast; @meyer2018].
+However, both are also missing integration into the major machine learning frameworks in R.
 
-Thus, \pkg{mlr3spatiotempcv} implements for the first time a comprehensive state-of-the-art compilation of spatial and spatiotemporal partitioning schemes that is well-integrated into a major machine-learning framework in R, the \pkg{mlr3} ecosystem [@mlr3].
+Thus, \pkg{mlr3spatiotempcv} implements for the first time a comprehensive state-of-the-art compilation of spatial and spatiotemporal partitioning schemes that is well-integrated into a major machine learning framework in R, the \pkg{mlr3} ecosystem [@mlr3].
 This package is furthermore equipped with a variety of two- and three-dimensional diagram types.
@@
 \[
-\mathit{perf(\hat{f}_L)} := E(l(Y,\hat{f}_L(X))),
+\mathit{perf(\hat{f}_L)} := E(l(Y,\hat{f}_L(X)))
 \]
@@
 The loss function can take a variety of forms such as the misclassification error $I(Y\neq\hat{f}_L(X))$ in classification, or the squared error $(Y-\hat{f}_L(X))^2$ in regression, among many other possible measures.
-The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance metrics for regression and classification [see e.g., @milligan2000 for classification, and ref??
-for regression tasks].
+The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance metrics for regression and classification (see e.g.
+@milligan2000 for classification).
 
@@
 where $f$ is a machine-learning algorithm, i.e. a mapping that trains a model $\hat{f}_S$ using any suitable training sample $S$.
-The use of $k=5$ or $k=10$ folds is most commonly seen in practice, and these preferences are also supported by theory [ref??].
+The use of $k=5$ or $k=10$ folds is most commonly seen in practice, and these preferences are also supported by theory [@bengio2004, @cawley2010].
 The $k$-fold CV estimator of model performance is a nearly unbiased estimator of the conditional performance measure when the observations were drawn independently [@efron1983].
-Since $\widehat{\mathit{perf}}_{L, CV}(f)$ still depends on the particular partitioning chosen for $L$, it is sometimes recommended to repeat the estimation using different random partitionings ($r$-repeated $k$-fold cross-validation, e.g.
-$k=100$; ref??).
+Since $\widehat{\mathit{perf}}_{L, CV}(f)$ still depends on the particular partitioning chosen for $L$, it is sometimes recommended to repeat the estimation using different random partitionings ($r$-repeated $k$-fold cross-validation) to reduce the influence of randomness when creating partitions [@vanwinckelen2012].
 
@@
 
-- add a paragraph with more information on spCV using spatial partitionings
+- FIXME: add a paragraph with more information on spCV using spatial partitionings
 
-A variant of cross-validation is leave-one-out (LOO) cross-validation , which has long been established in geostatistics [@cressie1993], sometimes with a focus on the spatial distribution of LOO error [@willmott2006].
+A variant of cross-validation is leave-one-out (LOO) cross-validation, which has long been established in geostatistics [@cressie1993], sometimes with a focus on the spatial distribution of LOO error [@willmott2006].
 Since this is just a special case of non-spatial CV with $k=n$, it is sometimes also referred to as spatial CV [@willmott2006].
@@
 
-- add a paragraph on spatiotemporal partitioning
+- FIXME: add a paragraph on spatiotemporal partitioning
 
@@
                                       \cmidrule{2-5}
-       & Predefined grouping & \specialcell{\pkg{CAST} \\ \pkg{mlr3}}       & \specialcell{ \texttt{"sptcv\_cstf"} \tablefootnote{\label{blothe2021}\cite{blothe2021}} \tablefootnote{\label{dolan2021}\cite{dolan2021}} \tablefootnote{\label{dumke2019}\cite{dumke2019}}  \\ \texttt{"FIXME"}}  & \specialcell{12 \\ -}       \\
+       & Predefined grouping & \specialcell{\pkg{CAST} \\ \pkg{mlr3}}       & \specialcell{ \texttt{"sptcv\_cstf"} (LLO) \tablefootnote{\label{blothe2021}\cite{blothe2021}} \tablefootnote{\label{dolan2021}\cite{dolan2021}} \tablefootnote{\label{dumke2019}\cite{dumke2019}}  \\ \texttt{"FIXME"}}  & \specialcell{12 \\ -}       \\
       \midrule
@@
 
-The "buffering" method from the \pkg{blockCV} package is conceptually similar to a leave-one-out (LOO) cross-validation approach [@rest2014, @blockCV].
-To reduce the similarity between train and test, a circular buffer around the observations representing the test set is drawn.
+The "buffering" method from the \pkg{blockCV} package is conceptually similar to a leave-one-out (LOO) cross-validation approach [@stone1974].
+To reduce the similarity between train and test, a circular buffer around the observations representing the test set is drawn [@rest2014, @blockCV].
 Within this buffer zone, all observations are removed from the training set, thereby reducing the similarity between train and test set and subsequently the influence of SAC.
@@
 
-"Spatial Blocking" (\pkg{blockCV} package) creates homogeneous blocks across the study area splitting observations into different zones [@bahn2012a, @wenger2012a].
-Then, a number of zones is combined randomly following desired the number of folds.
-For example, in a five-fold scenario, the method tries to combines zones to end up with partitions containing around 20\% of all observations.
+"Spatial Blocking" (\pkg{blockCV} package [@blockCV]) creates homogeneous blocks across the study area splitting observations into different zones [@bahn2012, @wenger2012].
+Multiple "blocks" are then combined into a partition, following the number of folds specified by the user.
+For example, in a five-fold scenario, the method would combine one fifth of all blocks into a partition (with five partitions in total).
 

### 2021-05-25 paper/paper.Rmd
minor edits in intro

--- 2021-05-20 paper/paper.Rmd
@@
 Spatial and spatiotemporal prediction tasks are common in applications ranging from environmental sciences to archaeology and epidemiology.
-While sophisticated mathematical frameworks have long been developed in spatial statistics to characterize predictive uncertainties under well-defined mathematical assumptions such as intrinsic stationarity [e.g., [@cressie1993], computational estimation procedures have only been proposed more recently to assess predictive performances of spatial and spatiotemporal prediction models [@brenning2005, @brenning2012; @pohjankukka2017; @roberts2017].
+While sophisticated mathematical frameworks have long been developed in spatial statistics to characterize predictive uncertainties under well-defined mathematical assumptions such as intrinsic stationarity (e.g., @cressie1993), computational estimation procedures have only been proposed more recently to assess predictive performances of spatial and spatiotemporal prediction models [@brenning2005; @brenning2012; @pohjankukka2017; @roberts2017].
 
-Although alternatives such as the bootstrap exist and have some advantages [@efron1983; @milligan2000], cross-validation is a particularly well-established, easy-to-implement algorithm for *model assessment* of supervised machine-learning models [@efron1983 and next section] and *model selection* [@arlot2010].
-In its basic form, it is based on resampling the data without paying attention to any possible dependence structure, which may arise from, e.g., cluster sampling [ref], grouped data [@brenning2008; @pena2015], or underlying environmental processes inducing some sort of spatial coherence at the landscape scale [@brenning2005].
+Although alternatives such as the bootstrap exist and have some advantages [@efron1983; @milligan2000], cross-validation is a particularly well-established, easy-to-implement algorithm for *model assessment* of supervised machine-learning models (@efron1983 and next section) and *model selection* [@arlot2010].
+In its basic form, it is based on resampling the data without paying attention to any possible dependence structure, which may arise from, e.g., grouped or structured data [@brenning2008; @pena2015], or underlying environmental processes inducing some sort of spatial coherence at the landscape scale [@brenning2005].
 In treating dependent observations as independent, or ignoring autocorrelation, cross-validation test samples may in fact be heavily correlated with, or even pseudo-replicates of, the data used for training the model, which introduces a potentially severe bias, especially for flexible machine-learning models [@brenning2005].
 
-This cross-validation bias is well-known in spatial as well as non-spatial prediction [@brenning2005; @brenning2008; @arlot2010].
+This cross-validation bias is well-known in spatial as well as non-spatial prediction [@brenning2005; @brenning2008; @arlot2010; @roberts2017].
 It is most easily understood from a predictive modelling perspective by focusing on the question where (and when) the model should be used for prediction.
-In crop classification from remotely-sensed data, for instance, learning samples routinely contain multiple grid cells from a sample of fields with known crop type, e.g.
-2000 grid cells from 100 fields scattered across a large study region.
-The purpose of training a model on this sample is to make predictions on other, new *fields* within the same geographic domain [*intra-domain* prediction; @brenning2005] &emdash; not *within* the same field, which obviously presents only one crop type.
-It would therefore be foolish to train a model on a simple random subsample of grid cells, and test it on the remaining data, i.e.
-using other grid cells from the same fields, as if we wanted to predict within a field; the results from this performance assessment would be over-optimitic, and perhaps badly so.
+In crop classification from remotely-sensed data, for instance, learning samples routinely contain multiple grid cells from a sample of fields with known crop type, for instance 2000 grid cells from 100 fields scattered across a large study region.
+The purpose of training a model on this sample is to make predictions on other, new *fields* within the same geographic domain (*intra-domain* prediction; @brenning2005) --- not *within* the same field, which obviously presents only one crop type.
+In this stituation it would therefore seem rather foolish to train a model on a simple random subsample of grid cells, and test it on the remaining data, i.e.
+using other grid cells from the same fields, as if we wanted to predict within a field.
+The results from this performance assessment would be over-optimitic, and perhaps badly so.
 To mimic the predictive situation for which the model is trained, we'd rather have to resample at the level of fields, not grid cells [@pena2015].
 If the model was to be applied to adjacent agricultural districts, i.e.
-outside the learning sample's spatial domain [*extra-domain* prediction; @brenning2005], it would even seem necessary to resample at a higher level of spatial aggregation, i.e.
+outside the learning sample's spatial domain (*extra-domain* prediction; @brenning2005), it would even seem necessary to resample at a higher level of spatial aggregation, i.e.
 at the level of agricultural districts within the learning sample, in order to realistically mimic the actual prediction task.
@@
 
-While it is not the purpose of this paper to recommend specific resampling schemes for specific use cases, this example may suffice to motivate the use of spatial and spatiotemporal cross-validation techniques, and the need for a unified framework and computational toolbox that accommodate a variety of prediction tasks that may be applicable to a broad range of prediction tasks.
+While it is not the purpose of this paper to recommend specific resampling schemes for specific use cases, this example may suffice to motivate the use of spatial and spatiotemporal cross-validation techniques, and the need for a unified framework and computational toolbox that accommodate a variety of prediction tasks that may be applicable to a broad range of application scenarios.
 \pkg{mlr3spatiotempcv} is such a toolbox.
@@
 This toolbox, implemented as an open-source R package, builds upon and generalizes several existing toolboxes that have been developed in recent years for more specific settings (see Table \@ref(tab:sptcv-methods)).
-The earliest of these implementations is the \pkg{sperrorest} package [@brenning2012], which has, however, not been integrated into established machine-learning frameworks such as \pkg{mlr} or \pkg{caret}, and lacks support of temporal prediction tasks.
-Other more recent implementations include \pkg{blockCV} for (what for?) [ref @blockCV; @rest2014] and \pkg{CAST} for spatiotemporal prediction tasks in a \pkg{caret} framework [@cast; @meyer2018a].
+The earliest and most comprehensive of these implementations is the \pkg{sperrorest} package [@brenning2012], which provides an extendible framework and includes predefined resampling strategies based on, for example, blocking, grouping and buffering.
+In contrast, \pkg{blockCV} was developed for block and buffer resampling with a focus on species distribution modelling [@blockCV; @rest2014].
+Neither of these have been integrated into established machine-learning frameworks such as \pkg{mlr} or \pkg{caret}, and both lack support of temporal prediction tasks.
+The \pkg{CAST} package, in contrast, focuses on spatiotemporal prediction tasks in a \pkg{caret} framework [@cast; @meyer2018a].
+One limitation of these packages is also that they focus on model assessment, while the proposed implementation within the \pkg{mlr} framework offers a seamless integration into model selection.
 
-Thus, \pkg{mlr3spatiotempcv} implements for the first time a comprehensive state-of-the-art compilation of spatial and spatiotemporal partitioning schemes that is well-integrated into a major machine-learning framework in R, the \pkg{mlr3} ecosystem [@mlr3].
+Thus, \pkg{mlr3spatiotempcv} implements for the first time a comprehensive state-of-the-art compilation of spatial and spatiotemporal partitioning schemes that is well-integrated into a comprehensive machine-learning framework in R, the \pkg{mlr3} ecosystem [@mlr3].
 This package is furthermore equipped with a variety of two- and three-dimensional diagram types.
-Our hope is that implementation will facilitate reproducible geospatial modelling and code-sharing across a broad range of application domains.
+Our hope is that this implementation will facilitate reproducible geospatial modelling and code-sharing across a broad range of application domains.
 
@@
 In cross-validation for predictive model assessment, we consider the following formal setting.
-We are interested in predicting a numerical or categorical response $y$ of an object or instance using a feature vector $\mathbf{x} = (x^{(1)}, \ldots, x^{(p)}\in\mathbb{R}$ and a model $\hat{f}_L$ that has been trained on a learning sample $L = \{(y_i, \mathbf{x}_i),\ i = 1, ..., n\}$.
+We are interested in predicting a numerical or categorical response $y$ of an object or instance using a feature vector $\mathbf{x} = (x^{(1)}, \ldots, x^{(p)})\in\mathbb{R}^p$ and a model $\hat{f}_L$ that has been trained on a learning sample $L = \{(y_i, \mathbf{x}_i),\ i = 1, ..., n\}$.
 Our goal is to estimate the expected value of the performance of $\hat{f}_L$,
@@
 The loss function can take a variety of forms such as the misclassification error $I(Y\neq\hat{f}_L(X))$ in classification, or the squared error $(Y-\hat{f}_L(X))^2$ in regression, among many other possible measures.
-The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance metrics for regression and classification [see e.g., @milligan2000 for classification, and ref??
-for regression tasks].
+The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance metrics for regression and classification (see e.g., @hand1997 for classification, and ref??
+for regression tasks).
 

### 2021-05-25 paper/paper.Rmd
using caligraphic letters for sets

@@
 In cross-validation for predictive model assessment, we consider the following formal setting.
-We are interested in predicting a numerical or categorical response $y$ of an object or instance using a feature vector $\mathbf{x} = (x^{(1)}, \ldots, x^{(p)})\in\mathbb{R}^p$ and a model $\hat{f}_L$ that has been trained on a learning sample $L = \{(y_i, \mathbf{x}_i),\ i = 1, ..., n\}$.
-Our goal is to estimate the expected value of the performance of $\hat{f}_L$,
+We are interested in predicting a numerical or categorical response $y$ of an object or instance using a feature vector $\mathbf{x} = (x^{(1)}, \ldots, x^{(p)})\in\mathbb{R}^p$ and a model $\hat{f}_\mathcal{L}$ that has been trained on a learning sample $\mathcal{L} = \{(y_i, \mathbf{x}_i),\ i = 1, ..., n\}$.
+Our goal is to estimate the expected value of the performance of $\hat{f}_\mathcal{L}$,
 \[
-\mathit{perf(\hat{f}_L)} := E(l(Y,\hat{f}_L(X))),
+\mathit{perf(\hat{f}_\mathcal{L})} := E(l(Y,\hat{f}_\mathcal{L}(X))),
 \]
@@
 This is referred to as the *actual* or *conditional* performance measure as it is conditional on $L$ [@milligan2000].
-The loss function can take a variety of forms such as the misclassification error $I(Y\neq\hat{f}_L(X))$ in classification, or the squared error $(Y-\hat{f}_L(X))^2$ in regression, among many other possible measures.
+The loss function can take a variety of forms such as the misclassification error $I(Y\neq\hat{f}_\mathcal{L}(X))$ in classification, or the squared error $(Y-\hat{f}_\mathcal{L}(X))^2$ in regression, among many other possible measures.
 The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance metrics for regression and classification (see e.g., @hand1997 for classification, and ref??
@@
 
-Nevertheless, since we only have a sample $T$ of test data drawn from the population, we can realistically only *estimate* the conditional performance of $\hat{f}_L$:
+Nevertheless, since we only have a sample $\mathcal{T}$ of test data drawn from the population, we can realistically only *estimate* the conditional performance of $\hat{f}_\mathcal{L}$:
 \[
-\widehat{\mathit{perf}}_T(\hat{f}_L) = \frac{1}{|T|}\sum_{i=1}^{|T|}l(Y,\hat{f}_L(X)).
+\widehat{\mathit{perf}}_T(\hat{f}_\mathcal{L}) = \frac{1}{|\mathcal{T}|}\sum_{i=1}^{|\mathcal{T}|}l(Y,\hat{f}_\mathcal{L}(X)).
 \]
-This representation as a point estimator for $\mathit{perf(\hat{f}_L)}$ underlines the importance of using a random sample for performance estimation to avoid estimation bias.
-Other estimators than the simple mean may be required when $T$ is not a simple random sample, for instance a stratified random sample [e.g., @thompson2012].
+This representation as a point estimator for $\mathit{perf(\hat{f}_\mathcal{L})}$ underlines the importance of using a random sample for performance estimation to avoid estimation bias.
+Other estimators than the simple mean may be required when $\mathcal{T}$ is not a simple random sample, for instance a stratified random sample [e.g., @thompson2012].
 As always, judgment sampling may lead to uncontrollable bias.
 
-Since the re-use of the learning sample $L$ for testing, $T=L$, would result in the over-optimistic *resubstitution* or *apparent* model performance, cross-validation partitions the sample $L$ into disjoint training and test sets.
-Specifically, $L$ is split into $k$ partitions,
+Since the re-use of the learning sample $\mathcal{L}$ for testing, $\mathcal{T}=\mathcal{L}$, would result in the over-optimistic *resubstitution* or *apparent* model performance, cross-validation partitions the sample $\mathcal{L}$ into disjoint training and test sets.
+Specifically, $\mathcal{L}$ is split into $k$ partitions,
 \[
-L = L_1 \cup \ldots \cup L_k,\qquad L_i\cap L_j = \emptyset\quad \textrm{for all}\ i\neq j,
+\mathcal{L} = \mathcal{L}_1 \cup \ldots \cup \mathcal{L}_k,\qquad \mathcal{L}_i\cap \mathcal{L}_j = \emptyset\quad \textrm{for all}\ i\neq j,
 \]
-and a model $\hat{f}_{(i)}$ is fitted on $L_{(i)} := L\setminus L_i$, while $L_i$ is withheld for testing.
+and a model $\hat{f}_{(i)}$ is fitted on $\mathcal{L}_{(i)} := \mathcal{L}\setminus \mathcal{L}_i$, while $\mathcal{L}_i$ is withheld for testing.
 This is repeated for $i=1,\ldots,k$ in order to effectively use the entire sample for testing, while keeping training and test sets disjoint at all times.
@@
 \[
-\widehat{\mathit{perf}}_{L, CV}(f) := \frac{1}{k}\sum_{i=1}^k\widehat{\mathit{perf}}_{L_i}(\hat{f}_{L_{(i)}}),
+\widehat{\mathit{perf}}_{\mathcal{L}, CV}(f) := \frac{1}{k}\sum_{i=1}^k\widehat{\mathit{perf}}_{\mathcal{L}_i}(\hat{f}_{\mathcal{L}_{(i)}}),
 \]
-where $f$ is a machine-learning algorithm, i.e. a mapping that trains a model $\hat{f}_S$ using any suitable training sample $S$.
+where $f$ is a machine-learning algorithm, i.e. a mapping that trains a model $\hat{f}_\mathcal{S}$ using any suitable training sample $\mathcal{S}$.
 The use of $k=5$ or $k=10$ folds is most commonly seen in practice, and these preferences are also supported by theory [ref??].
 The $k$-fold CV estimator of model performance is a nearly unbiased estimator of the conditional performance measure when the observations were drawn independently [@efron1983].
-Since $\widehat{\mathit{perf}}_{L, CV}(f)$ still depends on the particular partitioning chosen for $L$, it is sometimes recommended to repeat the estimation using different random partitionings ($r$-repeated $k$-fold cross-validation, e.g.
+Since $\widehat{\mathit{perf}}_{\mathcal{L}, CV}(f)$ still depends on the particular partitioning chosen for $L$, it is sometimes recommended to repeat the estimation using different random partitionings ($r$-repeated $k$-fold cross-validation, e.g.
 $k=100$; ref??).

### 2021-05-25 paper/paper.Rmd
minor edits; added "temporary" references

@@
 
-This cross-validation bias is well-known in spatial as well as non-spatial prediction [@brenning2005; @brenning2008; @arlot2010; @roberts2017].
+This cross-validation bias is well-known in spatial as well as non-spatial prediction [@brenning2005; @brenning2008; @arlot2010; @roberts2017] and in forecasting (ref FIXME: <https://doi.org/10.1016/j.csda.2017.11.003>).
 It is most easily understood from a predictive modelling perspective by focusing on the question where (and when) the model should be used for prediction.
@@
 The loss function can take a variety of forms such as the misclassification error $I(Y\neq\hat{f}_\mathcal{L}(X))$ in classification, or the squared error $(Y-\hat{f}_\mathcal{L}(X))^2$ in regression, among many other possible measures.
-The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance metrics for regression and classification (see e.g., @hand1997 for classification, and ref??
-for regression tasks).
+The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance metrics for regression and classification (see e.g., @hand1997 for classification, and ref FIXME perhaps <https://doi.org/10.1016/j.ijforecast.2006.03.001> for regression and forecasting tasks).
 

### 2021-05-25 paper/paper.Rmd
minor edits (forgot to save before last commit)

@@
 
-Nevertheless, since we only have a sample $\mathcal{T}$ of test data drawn from the population, we can realistically only *estimate* the conditional performance of $\hat{f}_\mathcal{L}$:
+Since we only have a sample $\mathcal{T}$ of test data drawn from the population, we can only *estimate* the conditional performance of $\hat{f}_\mathcal{L}$:
 \[
@@
 \]
-This representation as a point estimator for $\mathit{perf(\hat{f}_\mathcal{L})}$ underlines the importance of using a random sample for performance estimation to avoid estimation bias.
+This representation as a point estimator of $\mathit{perf(\hat{f}_\mathcal{L})}$ underlines the importance of using a random sample for model assessment to avoid estimation bias.
 Other estimators than the simple mean may be required when $\mathcal{T}$ is not a simple random sample, for instance a stratified random sample [e.g., @thompson2012].
@@
 
-Since the re-use of the learning sample $\mathcal{L}$ for testing, $\mathcal{T}=\mathcal{L}$, would result in the over-optimistic *resubstitution* or *apparent* model performance, cross-validation partitions the sample $\mathcal{L}$ into disjoint training and test sets.
+Since re-using the learning sample $\mathcal{L}$ for testing, i.e.
+$\mathcal{T}=\mathcal{L}$, would result in the over-optimistic *resubstitution* or *apparent* performance, cross-validation partitions the sample $\mathcal{L}$ into disjoint training and test sets.
 Specifically, $\mathcal{L}$ is split into $k$ partitions,

### 2021-05-25 paper/paper.Rmd
added conceptual paragraph on spatial CV types

@@
 
-- add a paragraph with more information on spCV using spatial partitionings
+Approaches based on spatial blocking or grouping require either the construction of spatial zones, or the use of pre-existing spatial structures in the data.
+Let's refer to these spatial units or zones as $\mathcal{Z_i}$, $1\le i\le n_z$.
+These zones are often constructed to serve as the $k=n_z$ spatial partitions, for example by performing $k$-means clustering of the sample coordinates (ref??) or generating the desired number of rectangular blocks.
+However, when $n_z$ is much larger than the desired number of folds, $k$, then a partitioning can be applied to the zones, i.e.
+the indices $1, \ldots, n_z$ are grouped into $k$ equally sized subsets $\mathcal{I}_1, \ldots, \mathcal{I}_k$.
+This approach has been applied, for example, in spatial CV at the agricultural field level (Pena??) or at the transect level (Kasurak??).
+In the following we will conceptually distinguish between CV *at the group (or block) level*, referring to the latter scenario, from spatial *block (or group)* CV, where the blocks or groups themselves define the CV partitions.
 

### 2021-05-25 paper/paper.Rmd

@@
 
+Thus, \pkg{mlr3spatiotempcv} implements for the first time a comprehensive state-of-the-art compilation of spatial and spatiotemporal partitioning schemes that is well-integrated into a major machine learning framework in R, the \pkg{mlr3} ecosystem [@mlr3].
+This package is furthermore equipped with a variety of two- and three-dimensional diagram types.
+Our hope is that implementation will facilitate reproducible geospatial modelling and code-sharing across a broad range of application domains.
+
 # Spatial and Spatiotemporal Cross-validation
@@
 where $f$ is a machine-learning algorithm, i.e. a mapping that trains a model $\hat{f}_\mathcal{S}$ using any suitable training sample $\mathcal{S}$.
-The use of $k=5$ or $k=10$ folds is most commonly seen in practice, and these preferences are also supported by theory [ref??].
+The use of $k=5$ or $k=10$ folds is most commonly seen in practice, and these preferences are also supported by theory [@bengio2004, @cawley2010].
 The $k$-fold CV estimator of model performance is a nearly unbiased estimator of the conditional performance measure when the observations were drawn independently [@efron1983].
-Since $\widehat{\mathit{perf}}_{\mathcal{L}, CV}(f)$ still depends on the particular partitioning chosen for $L$, it is sometimes recommended to repeat the estimation using different random partitionings ($r$-repeated $k$-fold cross-validation, e.g.
-$k=100$; ref??).
+Since $\widehat{\mathit{perf}}_{\mathcal{L}, CV}(f)$ still depends on the particular partitioning chosen for $L$, it is sometimes recommended to repeat the estimation using different random partitionings ($r$-repeated $k$-fold cross-validation) to reduce the influence of randomness when creating partitions [@vanwinckelen2012].
 
@@
 
-A variant of cross-validation is leave-one-out (LOO) cross-validation , which has long been established in geostatistics [@cressie1993], sometimes with a focus on the spatial distribution of LOO error [@willmott2006].
+A variant of cross-validation is leave-one-out (LOO) cross-validation, which has long been established in geostatistics [@cressie1993], sometimes with a focus on the spatial distribution of LOO error [@willmott2006].
 Since this is just a special case of non-spatial CV with $k=n$, it is sometimes also referred to as spatial CV [@willmott2006].

### 2021-05-25 paper/paper.Rmd
completed section 2

--- 2021-05-25 paper/paper.Rmd
@@
 
-This cross-validation bias is well-known in spatial as well as non-spatial prediction [@brenning2005; @brenning2008; @arlot2010; @roberts2017] and in forecasting (ref FIXME: <https://doi.org/10.1016/j.csda.2017.11.003>).
+This cross-validation bias is well-known in spatial as well as non-spatial prediction [@brenning2005; @brenning2008; @arlot2010; @roberts2017] and in forecasting [@bergmeir2018].
 It is most easily understood from a predictive modelling perspective by focusing on the question where (and when) the model should be used for prediction.
@@
 The loss function can take a variety of forms such as the misclassification error $I(Y\neq\hat{f}_\mathcal{L}(X))$ in classification, or the squared error $(Y-\hat{f}_\mathcal{L}(X))^2$ in regression, among many other possible measures.
-The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance metrics for regression and classification (see e.g., @hand1997 for classification, and ref FIXME perhaps <https://doi.org/10.1016/j.ijforecast.2006.03.001> for regression and forecasting tasks).
+The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance metrics for regression and classification (see e.g., @hand1997 for classification, and @hyndman2006 for regression and forecasting tasks).
 
@@
 
-- add a paragraph with more information on spCV using spatial partitionings
+Approaches based on spatial blocking or grouping require either the construction of spatial zones, or the use of pre-existing spatial structures in the data.
+Let's refer to these spatial units or zones as $\mathcal{Z_i}$, $1\le i\le n_z$.
+These zones are often constructed to serve as the $k=n_z$ spatial partitions, for example by performing $k$-means clustering of the sample coordinates [@russ2010] or generating the desired number of rectangular blocks.
+Similarly, the data can be partitioned in feature space instead of geographic space, which has been referred to as environmental blocking [@roberts2017].
 
-A variant of cross-validation is leave-one-out (LOO) cross-validation , which has long been established in geostatistics [@cressie1993], sometimes with a focus on the spatial distribution of LOO error [@willmott2006].
-Since this is just a special case of non-spatial CV with $k=n$, it is sometimes also referred to as spatial CV [@willmott2006].
+When $n_z$ is much larger than the desired number of folds, $k$, then a partitioning can be applied to the zones themselves.
+In this case, the zone indices $1, \ldots, n_z$ are grouped into $k$ equally sized subsets $\mathcal{I}_1, \ldots, \mathcal{I}_k$.
+This approach has been applied, for example, in spatial CV at the agricultural field level [@pena2015].
+In the following we will conceptually distinguish between CV *at the group (or block) level*, referring to the latter scenario, from spatial *block (or grouped)* CV, where the blocks or groups themselves define the CV partitions.
+
+A variant of cross-validation is leave-one-out (LOO) cross-validation, which has long been established in geostatistics [@cressie1993], sometimes with a focus on the spatial distribution of LOO error [@willmott2006].
+Although this is just a special case of non-spatial CV with $k=n$, it is sometimes also referred to as spatial CV [@willmott2006].
 Nevertheless, spatial variants have been proposed that apply an exclusion buffer to the test locations to separate them from the training data [@brenning2005].
 One plausible approach for this is to use the range of autocorrelation of model residuals to determine the buffer distance as this ensures independence conditional on the predictors [@brenning2005].
-When data is grouped, e.g., due to multi-level sampling designs or the study of spatial objects, LOO at has been proposed at the site level [@martin2008; @kasurak2011] or, in animal movement studies, at the animal level [@anderson2005].
-Similarly, in the temporal domain, it has been proposed to leave out temporal observational units, as in leave-one-year-out CV [@anderson2005; @brenning2005].
+When data is grouped, e.g., due to multi-level sampling designs or the study of spatial objects, it has been proposed to apply LOO at the site level [@martin2008; @kasurak2011] or, in animal movement studies, at the animal level [@anderson2005]; we will refer to this as leave-location-out (LLO) CV.
 
-It should be noted that $k$-fold cross-validation with a large value of $k$, and LOO-CV in particular ($k=n$), are not only very time-consuming since the model has to be trained $k$ times; these models will also be nearly identical since only a tiny fraction of the data is withheld, and therefore estimation bias increases.
-'Pure' LOO-CV is therefore not recommended for machine-learning model assessment [ref??].
+It should be noted that $k$-fold cross-validation with a large value of $k$, and LOO-CV in particular ($k=n$), is not only very time-consuming since the model has to be trained $k$ times; these models will also be nearly identical since only a tiny fraction of the data is withheld, and therefore estimation bias increases.
+'Pure' LOO-CV is therefore not recommended for machine-learning model assessment.
 
-- add a paragraph on spatiotemporal partitioning
+In the purely temporal domain, a special case is to leave out temporal observational units (or time slices; leave-time-out or LTO CV), as in leave-one-year-out CV [@anderson2005; @brenning2005]. 
+Cross-validation and hold-out validation strategies for time series have been discussed more extensively in the forecasting literature, considering also the effects of serial autocorrelation [@bergmeir2018].
+
+Turning to prediction tasks with spatiotemporal data, various spatial, temporal, or spatiotemporal partitioning strategies are being used, depending on the specific study objectives. 
+While the former two ignore the spatial and temporal dimension of the data, respectively, it has also been proposed to leave out spatiotemporal blocks [@meyer2018] or clusters [@cluto].
 

### 2021-05-25 paper/paper.Rmd
line breaks
added comments to section 4, minor changes in section 3

@@
 
-Within the \pkg{mlr3} ecosystem, partitioning strategies are represented by their own objects of class \texttt{Resampling}, of which most are available within \pkg{mlr3} itself (e.g.
-regular random CV) and other specialized ones (e.g.
-\texttt{TaskClassifST} or \texttt{TaskRegrST}) are defined in extension packages such as \pkg{mlr3spatiotempcv}.
-In the ML pipeline, these objects define the data splits used for model evaluation and selection (hyperparameter tuning) by ML algorithms.
-They play an important role for an unbiased application of ML algorithms [@schratz2019].
+Within the \pkg{mlr3} ecosystem, partitioning strategies are represented by their own objects of class \texttt{Resampling}, most of which are available within \pkg{mlr3} itself (e.g., regular random CV); other specialized strategies (e.g., \texttt{TaskClassifST} or \texttt{TaskRegrST}) are defined in extension packages such as \pkg{mlr3spatiotempcv}.
+In the machine-learning pipeline, these objects define the data splits used for model evaluation and selection (hyperparameter tuning) by machine-learning algorithms.
+<!--- They play an important role for an unbiased application of ML algorithms [@schratz2019]. [FIXME: Point gas been made before, it is unrelated to this section's topic.] -->
 
-Resampling objects in \pkg{mlr3spatiotmpcv} inherit from class \texttt{mlr3::Resampling} and can be created from established object classes for geospatial data in R, including simple features [@pebesma2018], which facilitates its integration into domain-specific workflows in the geospatial sciences.
+Resampling objects in \pkg{mlr3spatiotmpcv} inherit from class \texttt{mlr3::Resampling} and can be created from established object classes for geospatial data in R, including simple features [@pebesma2018], which facilitates their integration into domain-specific workflows in the geospatial sciences.
 This includes support for geographic coordinate reference systems (CRS) and their respective distance measures in projected planar space or using great circle distances on the unprojected ellipsoid.
 Non-spatial data tables with coordinates as variables are also supported.
-If no projection is specified, World Geodetic System 1984 (WGS84, EPSG: 4326) will be assumed.
+If no projection is specified, World Geodetic System 1984 (WGS84, EPSG: 4326) is assumed.
+<!--- FIXME: Is this correct even for distance metrics?
+Wouldn't it be more natural to simply use Euclidean distance when no projection is specified?
+In many cases coordinates will be UTM, not lat/lon...
+Just wondering.
+-->
 
@@
 
-At the time of writing, \pkg{mlr3spatiotempcv} implements the partitioning methods identified in Table \@ref(tab:sptcv-methods) with their \pkg{mlr3} class name.
+\pkg{mlr3spatiotempcv} currently implements the partitioning methods identified in Table \@ref(tab:sptcv-methods) with their \pkg{mlr3} class name.
 Several of the implemented algorithms are themselves versatile toolboxes with multiple options.
@@
 The following sections give an overview of most existing partitioning strategies and their visualization.
-The available methods are futher discussed in section \@ref(sec:disc).
+The available methods are further discussed in section \@ref(sec:disc).
 
@@
 
+(Subsection needs to be rewritten, see invisible comments.
+If this particular implementation is designed for binary classification problems, then a generic LOO with buffer is also needed, which should be introduced before this application-specific special case.)
+
+<!--- FIXME: Why refer to the blockCV package?
+This is a general method, as described (or at least outlined) above.
+If code from blockCV was used, when this should be acknowledged, but otherwise I don't undertstand why this method is "from" the blockCV package.
+-->
+
 The "buffering" method from the \pkg{blockCV} package is conceptually similar to a leave-one-out (LOO) cross-validation approach [@stone1974].
-To reduce the similarity between train and test, a circular buffer around the observations representing the test set is drawn [@rest2014, @blockCV].
-Within this buffer zone, all observations are removed from the training set, thereby reducing the similarity between train and test set and subsequently the influence of SAC.
+To reduce the similarity between training and test sets, a circular buffer around the observations representing the test set is drawn [@rest2014, @blockCV].
+Within this buffer zone, all observations are removed from the training set, thereby reducing the similarity between training and test set.
 
-By default, the method considers all points for train/test creation.
+By default, the method considers all points for training/test set creation.
 In an ecological context, this equals to the assumption of having *presence/absence* data for binary response variables.
 This means all non-presence values in the response variable are assumed to be confirmed absence values.
+<!--- FIXME: Why is this worth mentioning?
+What else could it be doing?
+LOO CV with buffer is a generic resampling method that can be applied to classification and regression problems.
+If this particular implementation only supports binary classification, then this should be mentioned explicitly.
+But mlr3 is not limited to binary classification problems, so why???
+-->
 
 The implementation also comes with support for *presence/background* data for binary responses.
-In this scenario, only the presence values are considered validated.
+In this scenario, only the presence values are considered validated. <!--- What is validated, in this context? Again, why overfit the code to such a very specific setting? -->
 All other response observations are assumed to be absent without a proof for this assumption.
@@
 
+<!--- again, why mention package name in title? -->
+
 Another buffering oriented resampling method is "Spatial Disc" from package \pkg{sperrorest}.
@@
 
-Around the radial test set a buffer zone / neutral area can be defined in which observations are excluded from the training set, increasing the spatial distance between train and test sets.
+Around the radial test set a buffer zone / neutral area can be defined in which observations are excluded from the training set, increasing the spatial distance between training and test sets.
 

### 2021-05-25 paper/paper.Rmd
added mention of spacv in python; removed duplicate paragraph

@@
 One limitation of these packages is also that they focus on model assessment, while the proposed implementation within the \pkg{mlr} framework offers a seamless integration into model selection.
+It is worth noting that a spatial cross-validation library \pkg{spacv} has also recently been developed for Python 3, which can be used with the \pkg{scikit-learn} machine-learning framework [@pedregosa2011].
 
@@
 Our hope is that this implementation will facilitate reproducible geospatial modelling and code-sharing across a broad range of application domains.
-
-Thus, \pkg{mlr3spatiotempcv} implements for the first time a comprehensive state-of-the-art compilation of spatial and spatiotemporal partitioning schemes that is well-integrated into a major machine learning framework in R, the \pkg{mlr3} ecosystem [@mlr3].
-This package is furthermore equipped with a variety of two- and three-dimensional diagram types.
-Our hope is that implementation will facilitate reproducible geospatial modelling and code-sharing across a broad range of application domains.
 

### 2021-05-25 paper/paper.Rmd
mention ENMeval

@@
 The earliest and most comprehensive of these implementations is the \pkg{sperrorest} package [@brenning2012], which provides an extendible framework and includes predefined resampling strategies based on, for example, blocking, grouping and buffering.
-In contrast, \pkg{blockCV} was developed for block and buffer resampling with a focus on species distribution modelling [@blockCV; @rest2014]. 
-Neither of these have been integrated into established machine-learning frameworks such as \pkg{mlr} or \pkg{caret}, and both lack support of temporal prediction tasks. 
+In contrast, \pkg{blockCV} and \pkg{ENMeval} were developed for block and buffer resampling with a focus on species distribution modelling [@blockCV; @rest2014; @muscarella2014]. 
+Neither of these have been integrated into established machine-learning frameworks such as \pkg{mlr} or \pkg{caret}, and all of them lack support of temporal prediction tasks. 
 The \pkg{CAST} package, in contrast, focuses on spatiotemporal prediction tasks in a \pkg{caret} framework [@cast; @meyer2018a]. 

### 2021-05-25 paper/paper.Rmd
added overview paragraph to Intro

@@
 Our hope is that this implementation will facilitate reproducible geospatial modelling and code-sharing across a broad range of application domains.
+
+The purpose of this paper is to give an overview of the methods implemented in the R package \pkg{mlr3spatiotempcv}.
+After presenting the conceptual background in the following section, we will outline the overall structure of the \pkg{mlr3spatiotempcv} package.
+We will then compare and contrast the various spatial and spatiotemporal partitioning techniques, before we demonstrate their application in a machine-learning model assessment in the following section.
+We finally discuss implications and provide an outlook.
 

### 2021-05-28 paper/paper.Rmd
clarified projection issues, dependency on upstream packages

@@
 In the machine-learning pipeline, these objects define the data splits used for model evaluation and selection (hyperparameter tuning) by machine-learning algorithms. 
-<!--- They play an important role for an unbiased application of ML algorithms [@schratz2019]. [FIXME: Point gas been made before, it is unrelated to this section's topic.] -->
+Spatial and spatiotemporal partitioning techniques in \pkg{mlr3spatiotempcv} are currently mostly imported and interfaced from other packages, in particular \pkg{sperrorest}, \pkg{blockCV}, \pkg{CAST} [@blockCV; @brenning2012; @CAST], in order to expose them to \pkg{mlr3} functionality.
 
 Resampling objects in \pkg{mlr3spatiotmpcv} inherit from class \texttt{mlr3::Resampling} and can be created from established object classes for geospatial data in R, including simple features [@pebesma2018], which facilitates their integration into domain-specific workflows in the geospatial sciences.
-This includes support for geographic coordinate reference systems (CRS) and their respective distance measures in projected planar space or using great circle distances on the unprojected ellipsoid.
+Support for projected (planar) and unprojected (geographic) coordinate reference systems (CRS) currently varies depending on the partitioning techniques used, since these inherit their behaviour from the underlying R package.
+It is planned to provide a more consistent interface that supports all CRS and allows the use of distance measures in projected planar space or using great circle distances on the unprojected ellipsoid, in the case of unprojected CRS.
 Non-spatial data tables with coordinates as variables are also supported.
-If no projection is specified, World Geodetic System 1984 (WGS84, EPSG: 4326) is assumed.
-<!--- FIXME: Is this correct even for distance metrics?
-Wouldn't it be more natural to simply use Euclidean distance when no projection is specified?
-In many cases coordinates will be UTM, not lat/lon...
-Just wondering.
--->
 

### 2021-05-28 paper/paper.Rmd
rewrote CLUTO subsection, changed it to "additional ..."

@@
 
-### Spatiotemporal Clustering (skmeans)
-
-The "CLUTO" clustering algorithm by George Karypis is said to be an application "for clustering of low- and high-dimensional datasets and for analyzing the characteristics of the various clusters" [@cluto].
-
-The method can generally be used in {mlr3spatiotempcv} if users cope with the copyright statement of the tool, which restricts usage to non-profit organizations and US government agencies only, and make the application binary available locally.
-
-These restrictions in combination with
-
-- the requirement of a local binary which needs to be downloaded externally and added to a user's system path
-- the absent of updates since the year 2007
-- a http-only website
-- no response from the copyright holder after asking for the current status of the project
-
-led to the decision to not explain and showcase this method in greater detail in this manuscript.
-Users are encouraged to visit the help page of the package for this method (`?ResamplingSptCVCluto`) for more information.
-
 ## Feature space clustering
@@
 ```
+
+### Additional partitioning techniques
+
+Considering the wealth and continuing development of spatial and spatiotemporal partitioning techniques for specific applications, we will refrain from providing a complete account of these algorithms.
+At present, \pkg{mlr3spatiotempcv} also supports spatiotemporal partitioning using the versatile CLUTO clustering algorithm [@cluto]. 
+CLUTO is available in R through the \pkg{skmeans} package, which provides an interface to a downloadable compiled library with a restriction to non-commercial uses (see `?ResamplingSptCVCluto` in \pkg{mlr3spatiotempcv} for more information).
+
+<!--- FIXME: It sounds like CLUTO can potentially be applied to spatial coordinates, space-time coordinates, or a combinations of coordinates and features.
+Does mlr3spatiotempcv support this, or is it limited to coordinates?
+If yes, perhaps this would be worth mentioning: "CLUTO can be used to partition data based on spatial coordinates, space--time coordinates, features, or a combination of these." -->
 

### 2021-05-29 paper/paper.Rmd
update author order, fix citations and bibfile

@@
     email: \email{patrick.schratz@uni-jena.de}
+  - name: Alexander Brenning
+    affiliation: 'Friedrich Schiller University Jena'
+    address: |
+      | Department of Geography
+      | Geographic Information Science group
+    email: \email{alexander.brenning@uni-jena.de}
   - name: Marc Becker
@@
       | Faculty of Statistics
-  - name: Alexander Brenning
-    affiliation: 'Friedrich Schiller University Jena'
-    address: |
-      | Department of Geography
-      | Geographic Information Science group
-    email: \email{alexander.brenning@uni-jena.de}
 title:
@@
 
-Although alternatives such as the bootstrap exist and have some advantages [@efron1983; @milligan2000], cross-validation is a particularly well-established, easy-to-implement algorithm for *model assessment* of supervised machine-learning models (@efron1983 and next section) and *model selection* [@arlot2010].
+Although alternatives such as the bootstrap exist and have some advantages [@efron1983; @hand1997], cross-validation is a particularly well-established, easy-to-implement algorithm for *model assessment* of supervised machine-learning models (@efron1983 and next section) and *model selection* [@arlot2010].
 In its basic form, it is based on resampling the data without paying attention to any possible dependence structure, which may arise from, e.g., grouped or structured data [@brenning2008; @pena2015], or underlying environmental processes inducing some sort of spatial coherence at the landscape scale [@brenning2005].
@@
 Neither of these have been integrated into established machine-learning frameworks such as \pkg{mlr} or \pkg{caret}, and all of them lack support of temporal prediction tasks.
-The \pkg{CAST} package, in contrast, focuses on spatiotemporal prediction tasks in a \pkg{caret} framework [@cast; @meyer2018a]. 
+The \pkg{CAST} package, in contrast, focuses on spatiotemporal prediction tasks in a \pkg{caret} framework [@cast; @meyer2018].
 One limitation of these packages is also that they focus on model assessment, while the proposed implementation within the \pkg{mlr} framework offers a seamless integration into model selection.
@@
 where $l$ is a real-valued loss function, and the expected value is with respect the probability distribution of $X$, the features of an instance $(Y,X)$ drawn randomly from the underlying population.
-This is referred to as the *actual* or *conditional* performance measure as it is conditional on $L$ [@milligan2000].
+This is referred to as the *actual* or *conditional* performance measure as it is conditional on $L$ [@hand1997].
 The loss function can take a variety of forms such as the misclassification error $I(Y\neq\hat{f}_\mathcal{L}(X))$ in classification, or the squared error $(Y-\hat{f}_\mathcal{L}(X))^2$ in regression, among many other possible measures.
@@
 In the machine-learning pipeline, these objects define the data splits used for model evaluation and selection (hyperparameter tuning) by machine-learning algorithms.
-Spatial and spatiotemporal partitioning techniques in \pkg{mlr3spatiotempcv} are currently mostly imported and interfaced from other packages, in particular \pkg{sperrorest}, \pkg{blockCV}, \pkg{CAST} [@blockCV; @brenning2012; @CAST], in order to expose them to \pkg{mlr3} functionality.
+Spatial and spatiotemporal partitioning techniques in \pkg{mlr3spatiotempcv} are currently mostly imported and interfaced from other packages, in particular \pkg{sperrorest}, \pkg{blockCV}, \pkg{CAST} [@blockCV; @brenning2012; @cast], in order to expose them to \pkg{mlr3} functionality.
 

### 2021-05-29 paper/paper.Rmd
clarify CLUTO clustering capabilities

@@
 CLUTO is available in R through the \pkg{skmeans} package, which provides an interface to a downloadable compiled library with a restriction to non-commercial uses (see `?ResamplingSptCVCluto` in \pkg{mlr3spatiotempcv} for more information).
-
-<!--- FIXME: It sounds like CLUTO can potentially be applied to spatial coordinates, space-time coordinates, or a combinations of coordinates and features.
-Does mlr3spatiotempcv support this, or is it limited to coordinates?
-If yes, perhaps this would be worth mentioning: "CLUTO can be used to partition data based on spatial coordinates, space--time coordinates, features, or a combination of these." -->
+With respect to functionality, \pkg{mlr3spatiotempcv} supports two-dimensional clustering based on coordinates and three-dimensional clustering on coordinates and time.
+Feature clustering is not supported.
 

### 2021-05-29 paper/paper.Rmd
reply to Alex blockCV::buffering comments

@@
 
-(Subsection needs to be rewritten, see invisible comments.
-If this particular implementation is designed for binary classification problems, then a generic LOO with buffer is also needed, which should be introduced before this application-specific special case.)
+<!--- (Subsection needs to be rewritten, see invisible comments.
+If this particular implementation is designed for binary classification problems, then a generic LOO with buffer is also needed, which should be introduced before this application-specific special case.) -->
+
+<!--- P: By default the buffering implementation of blockCV behaves like a normal LOOCV.
+It has additional features specific for SDM (presence-background data) but these are not active by default.
+That is why I did not add `sperrorest::partition_loo()` because `blockCV::buffering()` does the same + has support for SDM. -->
 
@@
 -->
+
+<!--- P: This addresses specifically the "buffering" function in the blockCV package which falls into the "buffering" category, the same applies to "spatial disc" from {sperrorest}.
+Is this a potential misunderstanding with the overarching level 3 header "Buffering"? -->
 
@@
 
+<!--- P: See my comment above, it might be that this is a misunderstanding. If not (or anyhow), we might want to clarify this more than to prevent such misunderstandings. -->
+
 The implementation also comes with support for *presence/background* data for binary responses.
 In this scenario, only the presence values are considered validated. <!--- What is validated, in this context? Again, why overfit the code to such a very specific setting? -->
+
+<!--- P: Validated here means: the existence of the presence observations is validated, the background observations are artifical. This is a common procedure in SDM.
+The blockCV buffering integration comes with support for presence/background handling which is important for SDM.
+I think this should be explaind shortly then - which is what this section is about. -->
+
 All other response observations are assumed to be absent without a proof for this assumption.

### 2021-05-29 paper/paper.Rmd
another comment reply

@@
 <!--- again, why mention package name in title? -->
+
+<!--- P: Because this is one method within the "buffering" category and so far my approach was to label these as "<method name> (<pkg>)".
+Apparently this is suboptimal from your point of view - how can we do better?
+Also the visual appearance of level 4 headings in the PDF is suboptimal - maybe we should do this someway different. -->
 

### 2021-05-29 paper/paper.Rmd
treat level 4 headers as proper headers

@@
     base_format: rticles::jss_article
+    includes:
+      in_header: header.tex
 ---
@@
 Apparently this is suboptimal from your point of view - how can we do better?
-Also the visual appearance of level 4 headings in the PDF is suboptimal - maybe we should do this someway different. -->
+Also the visual appearance of level 4 headings in the PDF is suboptimal - maybe we should do this someway different.
+I've fixed the leavel 4 headers, they appaer as proper formatted headers.
+I also opened an issue on GH to add some introductory information to each method category. -->
 

### 2021-05-29 paper/paper.Rmd
add intro for buffering categorie

@@
 
-#### Spatial buffering (blockCV)
+"Buffering" refers to the idea of removing observations between test and training set by drawing a cirular buffer around one or more observations [@rest2014, @brenning2012, @blockCV].
+By introducing more distance between observations, the influence of STAC should be reduced.
+Buffering mostly happens in the coordinate space due to the scattered appearance of observations in this space.
+\pkg{mlr3spatiotempcv} supports two "buffering" implementations: (1) "Spatial buffering" from package \pkg{blockCV} and (2) "Leave-one-disc-out cross-validation" from \pkg{sperrorest}.
+
+#### Spatial buffering --- \pkg{blockCV}
 
@@
 
-#### Spatial disc (sperrorest)
+#### Leave-one-disc-out
 
@@
 
-Another buffering oriented resampling method is "Spatial Disc" from package \pkg{sperrorest}.
-The name spins from "Leave-one-disc-out" which again originates from the generic "Leave-one-out" method.
+Another buffering oriented resampling method is "Leave-one-disc-out" from package \pkg{sperrorest}.
+The name originates from the generic "Leave-one-out" method.
 More specifically, the 'disc' which is being left our here refers to the radial area of test sets created from this method.
@@
 
-```{r disc, fig.cap="Visualization of 'Spatial Disc' method.", out.width="50%"}
+```{r disc, fig.cap="Visualization of 'Leave-one-disc-out' method.", out.width="50%"}
 resampling_disc = rsmp("spcv_disc", folds = 5, radius = 200L, buffer = 300L)

### 2021-05-29 paper/paper.Rmd
fix blockCV name in header
fix author display in header

@@
   - name: Alexander Brenning
-    affiliation: 'Friedrich Schiller University Jena'
+    affiliation: 'Friedrich Schiller University Jena \AND'
     address: |
@@
   - name: Marc Becker
-    affiliation: 'Ludwig-Maximilians-University Munich \AND'
+    affiliation: 'Ludwig-Maximilians-University Munich'
     address: |
@@
 
-#### Spatial buffering --- \pkg{blockCV}
+#### Spatial buffering --- blockCV
 

### 2021-05-30 paper/paper.Rmd
optimize blockCV::buffering sec

@@
 
-By default, the method considers all points for training/test set creation.
-In an ecological context, this equals to the assumption of having *presence/absence* data for binary response variables.
-This means all non-presence values in the response variable are assumed to be confirmed absence values.
+The implementation in the \pkg{blockCV} package comes with optional support for presence/absence data handling, as often used in the field of species distribution modeling (SDM) [@hijmans2013].
+However, by default, the method considers all points for training/test set creation and does not apply any specific SDM adjustments to the data.
+
+In an ecological/species distribution modeling context, the default assumption of a binary response variable is to have *presence/absence* data.
+This means all presence and absence values in the response variable are confirmed in their respective status.
 <!--- FIXME: Why is this worth mentioning?
@@
 Due to its similarity to a LOO CV, this method will result in as many train/test evaluations as observations are present.
-In the example below, a buffer zone of 1000 m is used and the first fold shown is plotted (Fig.\@ref(fig:buffer))).
+In the visualization of the first fold in Fig. \@ref(fig:buffer) a buffer zone of 1000 m was used without any custom presence/absence treament of the data.
 
-```{r buffer, fig.cap="Visualization of 'Spatial buffering' method.", out.width="50%"}
+```{r buffer, fig.cap="Visualization of the 'Spatial buffering' method from package `blockCV`.", out.width="50%", fig.pos="ht"}
 resampling_buffer = rsmp("spcv_buffer", theRange = 1000)

### 2021-05-30 paper/paper.Rmd
optimize spatial disc section

@@
 
-#### Leave-one-disc-out
+#### Leave-one-disc-out --- sperrorest
 
@@
 The name originates from the generic "Leave-one-out" method.
-More specifically, the 'disc' which is being left our here refers to the radial area of test sets created from this method.
+The 'disc' part in the name here refers to the radial area/buffer zone around the test set which is (optionally) being left out within each fold.
 
-Around the radial test set a buffer zone / neutral area can be defined in which observations are excluded from the training set, increasing the spatial distance between training and test sets.
+In addition, one can specify to sample with replacement by setting `replace = TRUE`.
+
+The method starts from a randomly selected observation and adds all points within the buffer zone area set via variable `radius` by the user.
+
+Around the radial test set a buffer zone / neutral area can be defined via variable `buffer` in which observations are excluded from the training set, increasing the spatial distance between training and test sets.
 
 ```{r disc, fig.cap="Visualization of 'Leave-one-disc-out' method.", out.width="50%"}
-resampling_disc = rsmp("spcv_disc", folds = 5, radius = 200L, buffer = 300L)
+resampling_disc = rsmp("spcv_disc", folds = 5, radius = 300L, buffer = 400L)
 resampling_disc$instantiate(task)

### 2021-06-11 paper/paper.Rmd
redo blocking intro

@@
 
-#### Spatial blocking (blockCV)
+"Spatial Blocking" creates homogeneous blocks across the study area splitting observations into different zones [@bahn2012, @wenger2012, @blockCV].
+The creation of these zones can be diverse: one option is to supply rows and columns to subdivide the data into a grid.
+Others make use of landscape patterns such as crop zones [@russ2010] or species presence [@blockCV] to determine an initial partitioning.
 
-"Spatial Blocking" (\pkg{blockCV} package [@blockCV]) creates homogeneous blocks across the study area splitting observations into different zones [@bahn2012, @wenger2012].
-Multiple "blocks" are then combined into a partition, following the number of folds specified by the user.
-For example, in a five-fold scenario, the method would combine one fifth of all blocks into a partition (with five partitions in total).
+Implementations of blocking methods often provide additional options to fine tune these initial partitions, allowing for fine-grained partition creation.
 
-The way how the blocks are created can be manifold: Supplying a numeric value in meters via argument `theRange` (named just `range` in {mlr3spatiotempcv}) will create a quadratic block pattern across the study area.
-The aggregation of zones can either be random (default) (`selection = "random"`), systematic (`selection = "systematic"`) or checkerboard structure (`selection = "checkerboard"`).
+\pkg{mlr3spatiotempcv} supports two "blocking" implementations: (1) "Spatial blocking" from package \pkg{blockCV} and (2) "Spatial Tiles" from \pkg{sperrorest}.
+
+#### Spatial blocking --- blockCV
+
+The "Spatial blocking" from package \pkg{blockCV} comes with various options to customize the zone creation.
+Besides the choice of relying on an initial random  (default) (`selection = "random"`), systematic (`selection = "systematic"`) or checkerboard pattern (`selection = "checkerboard"`) for the final partition aggregation, support for SDM specific background data is built-in.
+Supplying a numeric value in meters via argument `theRange` (named just `range` in {mlr3spatiotempcv}) will create a quadratic block pattern across the study area.
+Ultimately, users can also supply a user-defined polygon which defines the zones.
+
+After the initial block creation, multiple "blocks" are combined to form a partition, honoring the number of folds requested by the user.
+For example, in a five-fold scenario with a grid of four columns and five rows, the method would combine one fifth of all blocks (= four blocks) into a single partition.
+
+The following subsections showcase some of the introduced options.
 
@@
 
-#### Spatial tiling (sperrorest)
+#### Spatial Tiles --- sperrorest
 
@@
 ```{r tile, fig.cap="Spatial tiling with arguments `rows` and `cols`.", out.width="40%", fig.pos="ht"}
-resampling_tiles = rsmp("spcv_tiles",
-  nsplit = c(3L, 4L)
-)
+resampling_tiles = rsmp("spcv_tiles", nsplit = c(3L, 4L))
 

### 2021-06-11 paper/paper.Rmd
restructure group level CV part

@@
 
-### Factor-based grouping
+### Group-level CV
 
-#### Predefined grouping
+Group-level CV follows the principle of an inherit grouping structure of the observations, both in space and/or time, to determine (groups of) partitions.
+In contrast to blocking or clustering not the the spatial or temporal location is used but additional metadata such as the membership to a specific crop type [@brenning2006].
+The idea to follow a such an inherited grouping structure of the data is also referred to as "target-oriented validation" [@cast].
+Even though flexibility is limited due to the inherited grouping structure within the data, some implementations allow to specify the resulting number of folds by omitting parts of the dataset.
 
-FIXME
+\pkg{mlr3spatiotempcv} supports two "grouping" implementations: (1) The "Leave-location-and-time-out" family from package \pkg{CAST} and (2) "Custom Resampling" from \pkg{mlr3}.
 
-#### Leave-Location-And-Time-Out (CAST)
+#### Leave-Location-And-Time-Out --- CAST
 
-Leave-Location-and-Time-Out (LLTO) from packge \pkg{CAST} is a spatiotemporal resampling method [@meyer2018].
+Leave-Location-and-Time-Out (LLTO) from package \pkg{CAST} is a spatiotemporal resampling method [@meyer2018].
 It requires both a spatial and a temporal variable which are used to to separate observations in a multi-dimensional space.
@@
 ```
+
+### Custom Resampling --- mlr3
 

### 2021-06-13 paper/paper.Rmd
include example for custom resampling

@@
 
-### Group-level CV
+### Group-Level CV
 
@@
 
-### Custom Resampling --- mlr3
+### Custom (fixed) Resampling --- mlr3
+
+Support for user-defined partitioning strategies is built into \pkg{mlr3} directly since version FIXME.
+Labelled as "Custom CV", users are able to supply an external factor vector with the same length as number of observations.
+Alternatively to supplying an external factor vector, the name of a feature within the task can be used.
+
+The following example is taken from `sperrorest::partition_factor()`.
+
+FIXME: eval=TRUE when available upstream im mlr3
+
+```{r custom-cv,  fig.cap="Environmental blocking using a single predictor.", out.width="50%", fig.pos="ht", eval=FALSE}
+breaks <- quantile(task$data()$dem, seq(0, 1, length = 6))
+zclass <- cut(task$data()$dem, breaks, include.lowest = TRUE)
+
+resampling_custom = rsmp("custom_cv")
+resampling_custom$instantiate(task, split = zclass)
+```
 

### 2021-06-13 paper/paper.Rmd
harmonize
mention mlr3spatial and mlr3temporal

@@
 In the machine-learning pipeline, these objects define the data splits used for model evaluation and selection (hyperparameter tuning) by machine-learning algorithms.
-Spatial and spatiotemporal partitioning techniques in \pkg{mlr3spatiotempcv} are currently mostly imported and interfaced from other packages, in particular \pkg{sperrorest}, \pkg{blockCV}, \pkg{CAST} [@blockCV; @brenning2012; @cast], in order to expose them to \pkg{mlr3} functionality.
+Spatial and spatiotemporal partitioning techniques in \pkg{mlr3spatiotempcv} are currently mostly imported and interfaced from other packages, in particular \pkg{sperrorest}, \pkg{blockCV} and \pkg{CAST} [@blockCV; @brenning2012; @cast], in order to expose them to \pkg{mlr3} functionality.
+To reduce first-level dependencies, some methods were reimplemented instead of imported from upstream packages.
 
 Resampling objects in \pkg{mlr3spatiotmpcv} inherit from class \texttt{mlr3::Resampling} and can be created from established object classes for geospatial data in R, including simple features [@pebesma2018], which facilitates their integration into domain-specific workflows in the geospatial sciences.
-Support for projected (planar) and unprojected (geographic) coordinate reference systems (CRS) currently varies depending on the partitioning techniques used, since these inherit their behaviour from the underlying R package.
+Support for projected (planar) and unprojected (geographic) coordinate reference systems (CRS) currently varies depending on the partitioning techniques used, since these inherit their behavior from the underlying upstream R package.
 It is planned to provide a more consistent interface that supports all CRS and allows the use of distance measures in projected planar space or using great circle distances on the unprojected ellipsoid, in the case of unprojected CRS.
-Non-spatial data tables with coordinates as variables are also supported.
 
@@
 geographic space plus time.
+
+While \pkg{mlr3spatiotempcv} solely focuses on spatiotemporal resampling methods and their visualization, other packages such as \pkg{mlr3spatial} or \pkg{mlr3temporal} are planned in the mlr3 ecosystem to provide dedicated spatiotemporal learner and prediction methods.
 
@@
 ...
+<!-- FIXME: @alex: do you want to see more into here? -->
 
@@
 At first all necessary R packages are loaded and a lower verbosity is set to keep the output tidy.
-Also a seed is set for deterministic execution.
-
-Next the \pkg{mlr3spatiotempcv} built-in task `"ecuador"` is loaded via `tsk()`.
-
+Also a seed is set for deterministic executiona and the \pkg{mlr3spatiotempcv} built-in task `"ecuador"` is loaded via `tsk()`.
 Next, the random forest learner (`lrn("classif.ranger")`) is initialized with default hyperparameters and the prediction type set to `"probability"` due to the binary response variable.

### 2021-06-13 paper/paper.Rmd
spellcheck

@@
   At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: CAST [@cast], blockCV [@blockCV] or sperrorest [@brenning2012].
-  There are some resampling methods used in scientific studies for which the source code is either not available or a different programming language has been used [@roberts2017; @pohjankukka2017]
+  There are some resampling methods used in scientific studies for which the source code is either not available or a different programming language has been used [@roberts2017; @pohjankukka2017].
 
@@
 keywords:
-  # at least one keyword must be supplied
   formatted: [cross-validation, predictive performance, machine learning, autocorrelation, spatial, temporal, "\\proglang{R}"]
@@
 ```{r setup-2, echo=FALSE, purl=T}
-# for knitr spint later
+# for knitr spin later
 knitr::opts_chunk$set(fig.path = "./paper/figs/",
@@
 The purpose of training a model on this sample is to make predictions on other, new *fields* within the same geographic domain (*intra-domain* prediction; @brenning2005) --- not *within* the same field, which obviously presents only one crop type.
-In this stituation it would therefore seem rather foolish to train a model on a simple random subsample of grid cells, and test it on the remaining data, i.e.
+In this situation it would therefore seem rather foolish to train a model on a simple random subsample of grid cells, and test it on the remaining data, i.e.
 using other grid cells from the same fields, as if we wanted to predict within a field.
-The results from this performance assessment would be over-optimitic, and perhaps badly so.
+The results from this performance assessment would be over-optimistic, and perhaps badly so.
 To mimic the predictive situation for which the model is trained, we'd rather have to resample at the level of fields, not grid cells [@pena2015].
@@
 This toolbox, implemented as an open-source R package, builds upon and generalizes several existing toolboxes that have been developed in recent years for more specific settings (see Table \@ref(tab:sptcv-methods)).
-The earliest and most comprehensive of these implementations is the \pkg{sperrorest} package [@brenning2012], which provides an extendible framework and includes predefined resampling strategies based on, for example, blocking, grouping and buffering.
+The earliest and most comprehensive of these implementations is the \pkg{sperrorest} package [@brenning2012], which provides an extensible framework and includes predefined resampling strategies based on, for example, blocking, grouping and buffering.
 In contrast, \pkg{blockCV} and \pkg{ENMeval} were developed for block and buffer resampling with a focus on species distribution modelling [@blockCV; @rest2014; @muscarella2014].
@@
 Spatial and spatiotemporal partitioning techniques in \pkg{mlr3spatiotempcv} are currently mostly imported and interfaced from other packages, in particular \pkg{sperrorest}, \pkg{blockCV} and \pkg{CAST} [@blockCV; @brenning2012; @cast], in order to expose them to \pkg{mlr3} functionality.
-To reduce first-level dependencies, some methods were reimplemented instead of imported from upstream packages.
+To reduce first-level dependencies, some methods were re-implemented instead of imported from upstream packages.
 
@@
 
-"Buffering" refers to the idea of removing observations between test and training set by drawing a cirular buffer around one or more observations [@rest2014, @brenning2012, @blockCV].
+"Buffering" refers to the idea of removing observations between test and training set by drawing a circular buffer around one or more observations [@rest2014, @brenning2012, @blockCV].
 By introducing more distance between observations, the influence of STAC should be reduced.
@@
 This is a general method, as described (or at least outlined) above.
-If code from blockCV was used, when this should be acknowledged, but otherwise I don't undertstand why this method is "from" the blockCV package.
+If code from blockCV was used, when this should be acknowledged, but otherwise I don't understand why this method is "from" the blockCV package.
 -->
@@
 
-<!--- P: Validated here means: the existence of the presence observations is validated, the background observations are artifical. This is a common procedure in SDM.
+<!--- P: Validated here means: the existence of the presence observations is validated, the background observations are artificial. This is a common procedure in SDM.
 The blockCV buffering integration comes with support for presence/background handling which is important for SDM.
-I think this should be explaind shortly then - which is what this section is about. -->
+I think this should be explained shortly then - which is what this section is about. -->
 
@@
 Due to its similarity to a LOO CV, this method will result in as many train/test evaluations as observations are present.
-In the visualization of the first fold in Fig. \@ref(fig:buffer) a buffer zone of 1000 m was used without any custom presence/absence treament of the data.
+In the visualization of the first fold in Fig. \@ref(fig:buffer) a buffer zone of 1000 m was used without any custom presence/absence treatment of the data.
 
@@
 <!--- P: Because this is one method within the "buffering" category and so far my approach was to label these as "<method name> (<pkg>)".
-Apparently this is suboptimal from your point of view - how can we do better?
-Also the visual appearance of level 4 headings in the PDF is suboptimal - maybe we should do this someway different.
-I've fixed the leavel 4 headers, they appaer as proper formatted headers.
+Apparently this is sub-optimal from your point of view - how can we do better?
+Also the visual appearance of level 4 headings in the PDF is sub-optimal - maybe we should do this someway different.
+I've fixed the level 4 headers, they appear as proper formatted headers.
 I also opened an issue on GH to add some introductory information to each method category. -->
@@
 
-FIXME: eval=TRUE when available upstream im mlr3
+<!-- FIXME: eval=TRUE when available upstream im mlr3 -->
 
@@
 At first all necessary R packages are loaded and a lower verbosity is set to keep the output tidy.
-Also a seed is set for deterministic executiona and the \pkg{mlr3spatiotempcv} built-in task `"ecuador"` is loaded via `tsk()`.
+Also a seed is set for deterministic execution and the \pkg{mlr3spatiotempcv} built-in task `"ecuador"` is loaded via `tsk()`.
 Next, the random forest learner (`lrn("classif.ranger")`) is initialized with default hyperparameters and the prediction type set to `"probability"` due to the binary response variable.
@@
 
-The resulting AUC valiue of 0.77 using Non-Spatial Cross-Validation (NSpCV) is around 0.15 better compared to the Spatial Cross-Validation (SpCV) value of 0.61.
+The resulting AUC value of 0.77 using Non-Spatial Cross-Validation (NSpCV) is around 0.15 better compared to the Spatial Cross-Validation (SpCV) value of 0.61.
 The better performance score results from over-optimism in the estimation with NSpCV.
@@
 Methods `"spcv_cstf"` (with only one of `space_var` or `time_var` set) require a variable in the dataset which should be used for grouping.
-The specification of such is not always easy because the selection of such a variable also implies that the selected groups inherit substantial autorcorrelation and differ substantially from other groups.
+The specification of such is not always easy because the selection of such a variable also implies that the selected groups inherit substantial autocorrelation and differ substantially from other groups.
 Also, if the variable contains too many groups, the difference within train/test splits may become undesirably high and tend towards a LOO CV.
@@
 
-Spatiotemporal cross-validation is still not yet a paradigm that is completely established in scientfic workflows even though it is discussed intensively for more than a decade now.
+Spatiotemporal cross-validation is still not yet a paradigm that is completely established in scientific workflows even though it is discussed intensively for more than a decade now.
 Making most available methods easily accessible to users is an important step to foster the acceptance of such in the community and lead to less biased cross-validation results in environmental and ecological studies.

### 2021-06-13 paper/paper.Rmd
3d = space + time

@@
 Two visualization types are supported: a two-dimensional and a three-dimensional method which can be selected via the logical argument `plot3d`.
-The 3D visualization has been implemented via \pkg{plotly}.
+The 3D visualization (space + time) has been implemented via \pkg{plotly}.
 Because a dynamic image cannot be included in this manuscript, a static version, which can be generated by setting `static_image = TRUE`, is shown.
@@
 <!-- only some methods are suited for repeated execution -->
-<!-- spcv_coords: limited variance due to limited k-means possibilties -->
+<!-- spcv_coords: limited variance due to limited k-means possibilities -->
 For repeated executions of method `spcv_coords`, the variance between repetitions could be quite small due to the limited possibilities of different `k-means` clusters on the same dataset.
@@
 
-We believe that making resampling methods available to widely used modeling frameworks is more benefitial than implementing such methods standalone in an R package.
-The latter often require special structured objects as inputs and cannot directly be re-used in susequent modeling steps of user workflows.
+We believe that making resampling methods available to widely used modeling frameworks is more beneficial than implementing such methods standalone in an R package.
+The latter often require special structured objects as inputs and cannot directly be re-used in subsequent modeling steps of user workflows.
 This is not an issue specific to spatiotemporal resampling methods but a general philosophy of the authors.

### 2021-06-13 paper/paper.Rmd
add group-level CV to table

@@
                                       \cmidrule{2-5}
-       & Predefined grouping & \specialcell{\pkg{CAST} \\ \pkg{mlr3}}       & \specialcell{ \texttt{"sptcv\_cstf"} (LLO) \tablefootnote{\label{blothe2021}\cite{blothe2021}} \tablefootnote{\label{dolan2021}\cite{dolan2021}} \tablefootnote{\label{dumke2019}\cite{dumke2019}}  \\ \texttt{"FIXME"}}  & \specialcell{12 \\ -}       \\
+       & Group-Level CV & \specialcell{\pkg{CAST} \\ \pkg{mlr3}}       & \specialcell{ \texttt{"sptcv\_cstf"} (LLO) \tablefootnote{\label{blothe2021}\cite{blothe2021}} \tablefootnote{\label{dolan2021}\cite{dolan2021}} \tablefootnote{\label{dumke2019}\cite{dumke2019}}  \\ \texttt{"custom\_cv"} \tablefootnote{\label{brenning2006}\cite{brenning2006}} \tablefootnote{\label{pena2015}\cite{pena2015}}} & \specialcell{12 \\ 2}       \\
       \midrule
@@
 
-### Custom (fixed) Resampling --- mlr3
+#### Custom (fixed) Resampling --- mlr3
 

### 2021-06-15 paper/paper.Rmd
small edits, esp. coord.-based clustering

@@
 By introducing more distance between observations, the influence of STAC should be reduced.
-Buffering mostly happens in the coordinate space due to the scattered appearance of observations in this space.
+Buffering is mostly applied in the coordinate space due to the scattered appearance of observations in this space.
 \pkg{mlr3spatiotempcv} supports two "buffering" implementations: (1) "Spatial buffering" from package \pkg{blockCV} and (2) "Leave-one-disc-out cross-validation" from \pkg{sperrorest}.
@@
 
-<!--- FIXME: Why refer to the blockCV package?
-This is a general method, as described (or at least outlined) above.
-If code from blockCV was used, when this should be acknowledged, but otherwise I don't understand why this method is "from" the blockCV package.
--->
-
-<!--- P: This addresses specifically the "buffering" function in the blockCV package which falls into the "buffering" category, the same applies to "spatial disc" from {sperrorest}.
-Is this a potential misunderstanding with the overarching level 3 header "Buffering"? -->
+<!-- A: This paragraph is a general description of the approach, not a description of unique characteristics of this implementation.
+General approach of buffering is (and should be) described in the introdoctory paragraph of the Buffering section.
 
@@
 Within this buffer zone, all observations are removed from the training set, thereby reducing the similarity between training and test set.
+-->
 
 The implementation in the \pkg{blockCV} package comes with optional support for presence/absence data handling, as often used in the field of species distribution modeling (SDM) [@hijmans2013].
-However, by default, the method considers all points for training/test set creation and does not apply any specific SDM adjustments to the data.
+However, by default, the method considers all points for sampling the training and test sets.
 
@@
 
-The "Spatial CV" method after [@brenning2012] and implemented in package \pkg{sperrorest} is similar to method "environmental blocking" in that it also uses *k-means* clustering for fold creation.
-However, it does not make use of arbitrary features but uses the coordinates of all observations to create clusters in the spatial domain.
-This is what lead to the naming of `spcv_coords` in \pkg{mlr3spatiotempcv}.
+The "Spatial CV" method after [@russ2010; @brenning2012] and implemented in package \pkg{sperrorest} uses the coordinates of all observations to create clusters in the spatial domain.
+$k$-means clustering has been proposed as a clustering method.
+To avoid confusion with other spatial partitioning methods, we refer to this as coordinate-based clustering, or as `spcv_coords` in \pkg{mlr3spatiotempcv}.
 
-```{r coords, fig.cap="Method 'Spatial CV' showing a k-means clustering approach based on coordinates.", out.width="50%", fig.pos="ht"}
+This approach is very versatile as it adapts to irregularly-shaped study areas and ensures that exactly $k$ partitions are created.
+Nevertheless, despite the random selection of initial cluster centers, repeated partitionings may in some cases be nearly identical.
+Also, $k$-means clustering may be less suitable for data sets with existing clusters of points and/or with distant isolated sample locations.
+
+```{r coords, fig.cap="Method 'Spatial CV' showing k-means clustering based on coordinates.", out.width="50%", fig.pos="ht"}
 resampling_coords = rsmp("spcv_coords", folds = 5)$instantiate(task)

### 2021-06-24 paper/paper.Rmd
move parts from blockCV buffering to general buffering intro

@@
 "Buffering" refers to the idea of removing observations between test and training set by drawing a circular buffer around one or more observations [@rest2014, @brenning2012, @blockCV].
-By introducing more distance between observations, the influence of STAC should be reduced.
+Within this buffer zone, all observations are removed from the training set, thereby reducing the similarity between training and test set.
+
 Buffering is mostly applied in the coordinate space due to the scattered appearance of observations in this space.
@@
 
-<!--- (Subsection needs to be rewritten, see invisible comments.
-If this particular implementation is designed for binary classification problems, then a generic LOO with buffer is also needed, which should be introduced before this application-specific special case.) -->
-
-<!--- P: By default the buffering implementation of blockCV behaves like a normal LOOCV.
-It has additional features specific for SDM (presence-background data) but these are not active by default.
-That is why I did not add `sperrorest::partition_loo()` because `blockCV::buffering()` does the same + has support for SDM. -->
-
-<!-- A: This paragraph is a general description of the approach, not a description of unique characteristics of this implementation.
-General approach of buffering is (and should be) described in the introdoctory paragraph of the Buffering section.
-
 The "buffering" method from the \pkg{blockCV} package is conceptually similar to a leave-one-out (LOO) cross-validation approach [@stone1974].
-To reduce the similarity between training and test sets, a circular buffer around the observations representing the test set is drawn [@rest2014, @blockCV].
-Within this buffer zone, all observations are removed from the training set, thereby reducing the similarity between training and test set.
--->
-
-The implementation in the \pkg{blockCV} package comes with optional support for presence/absence data handling, as often used in the field of species distribution modeling (SDM) [@hijmans2013].
+It comes with optional support for presence/absence data handling, as often used in the field of species distribution modeling (SDM) [@hijmans2013].
 However, by default, the method considers all points for sampling the training and test sets.
@@
 This means all presence and absence values in the response variable are confirmed in their respective status.
-<!--- FIXME: Why is this worth mentioning?
-What else could it be doing?
-LOO CV with buffer is a generic resampling method that can be applied to classification and regression problems.
-If this particular implementation only supports binary classification, then this should be mentioned explicitly.
-But mlr3 is not limited to binary classification problems, so why???
--->
-
-<!--- P: See my comment above, it might be that this is a misunderstanding. If not (or anyhow), we might want to clarify this more than to prevent such misunderstandings. -->
-
-The implementation also comes with support for *presence/background* data for binary responses.
-In this scenario, only the presence values are considered validated. <!--- What is validated, in this context? Again, why overfit the code to such a very specific setting? -->
-
-<!--- P: Validated here means: the existence of the presence observations is validated, the background observations are artificial. This is a common procedure in SDM.
-The blockCV buffering integration comes with support for presence/background handling which is important for SDM.
-I think this should be explained shortly then - which is what this section is about. -->
+The \pkg{blockCV} package implementation also comes with support for *presence/background* data for binary responses.
 
@@
 #### Leave-one-disc-out --- sperrorest
-
-<!--- again, why mention package name in title? -->
-
-<!--- P: Because this is one method within the "buffering" category and so far my approach was to label these as "<method name> (<pkg>)".
-Apparently this is sub-optimal from your point of view - how can we do better?
-Also the visual appearance of level 4 headings in the PDF is sub-optimal - maybe we should do this someway different.
-I've fixed the level 4 headers, they appear as proper formatted headers.
-I also opened an issue on GH to add some introductory information to each method category. -->
 
@@
 ```{r custom-cv,  fig.cap="Environmental blocking using a single predictor.", out.width="50%", fig.pos="ht", eval=FALSE}
-breaks <- quantile(task$data()$dem, seq(0, 1, length = 6))
-zclass <- cut(task$data()$dem, breaks, include.lowest = TRUE)
+breaks = quantile(task$data()$dem, seq(0, 1, length = 6))
+zclass = cut(task$data()$dem, breaks, include.lowest = TRUE)
 

### 2021-06-24 paper/paper.Rmd
add intro to "clustering" and "spatial" chapter, adjust heading levels, change titles

@@
 Comprehensive and up-to-date information can be found in the package's online documentation (<https://mlr3spatiotempcv.mlr-org.com>).
-The following sections give an overview of most existing partitioning strategies and their visualization.
+The following sections, which follow the naming scheme `<method name> --- <R package>`, give an overview of most existing partitioning strategies and their visualization.
 The available methods are further discussed in section \@ref(sec:disc).
@@
   \footnotesize
-  \caption[t]{Spatiotemporal resampling methods, ordered alphabetically by class name (Notation column).
+  \caption[t]{Spatiotemporal resampling methods, ordered alphabetically by class name (column 'mlr3 Notation').
 Column count refers to the number of studies which were found to have used this resampling technique until 2021-05.}
@@
 
+Spatial partitioning methods operate in a two-dimensional space and split observations into training and test set by their spatial location.
+Methods such as grid-based partitioning or clustering are used to create groups of observations which may then be aggregated further to form the final partition sets.
+
+Judging by their common characteristics and the available implementations, spatial partitioning methods can be coarsely categorized into (1) buffering, (2) blocking and (3) clustering methods.
+If a method had the potential to belong to multiple categories, a deliberate category assignment was made for one specific group which matched the main characteristic of the respective method best.
+
 ### Buffering
@@
 Buffering is mostly applied in the coordinate space due to the scattered appearance of observations in this space.
-\pkg{mlr3spatiotempcv} supports two "buffering" implementations: (1) "Spatial buffering" from package \pkg{blockCV} and (2) "Leave-one-disc-out cross-validation" from \pkg{sperrorest}.
+However, there is no practical limitation that would prevent this method to be applied in a 2D + time setting.
+Because \pkg{mlr3spatiotempcv} supports two spatial-only "buffering" implementations, (1) "Spatial buffering" from package \pkg{blockCV} and (2) "Leave-one-disc-out cross-validation" from \pkg{sperrorest}, buffering was attributed to the category of spatial partitioning methods.
 
@@
 
-### Spatial Clustering (sperrorest)
+### Clustering
 
-#### Coordinate-based clustering
+The idea of cluster analysis, which originated in the 1930s in the anthropology field [@driver1932, @zubin1938], is to group objects with similar characteristics.
+In the spatial context, the similarity of spatial coordinates is used to form such groups.
+Different cluster algorithms such as hierarchical, density-based, grid-based or centroid-based clustering exist [@gower1967].
 
-The "Spatial CV" method after [@russ2010; @brenning2012] and implemented in package \pkg{sperrorest} uses the coordinates of all observations to create clusters in the spatial domain.
-$k$-means clustering has been proposed as a clustering method.
-To avoid confusion with other spatial partitioning methods, we refer to this as coordinate-based clustering, or as `spcv_coords` in \pkg{mlr3spatiotempcv}.
+#### Coordinate-based clustering --- sperrorest
+
+The so-called "Spatial CV" method after [@russ2010; @brenning2012] and implemented in package \pkg{sperrorest} uses the coordinates of all observations to create clusters in the spatial domain.
+$k$-means clustering, which belongs to the centroid-based clustering family, has been proposed as a clustering method.
+To avoid confusion with other spatial partitioning methods, we refer to this as coordinate-based clustering, or as `spcv_coords` in \pkg{mlr3spatiotempcv}, instead of using the generic "Spatial CV" term.
 
@@
 
-## Spatiotemporal
+## Spatial or Spatiotemporal
+
+<!-- FIXME: explain that most methdso can also be spatial-only -->
 
@@
 
-#### Custom (fixed) Resampling --- mlr3
+### Custom (fixed) Resampling --- mlr3
 
@@
 
-### Additional partitioning techniques
+## Additional partitioning techniques
 
@@
 
+<!-- FIXME: mention missing spatiotemporal buffering method -->
+
 <!-- missing: non-LOO method with buffer support -->
+<!-- FIXME: this is wrong, leave-one-disc-out does this -->
 Also no methods yet exists that makes use of a buffer zone with blocking or clustering, bridging the gap between buffered LOO-CV and fold-based non-buffer methods (`"spcv_coords"` and `"spcv_block"`).

### 2021-06-25 paper/paper.Rmd
add spatiotemporal intro

@@
 
-## Spatial or Spatiotemporal
+## Spatial, temporal or spatiotemporal
 
-<!-- FIXME: explain that most methdso can also be spatial-only -->
+Some methods have the ability to operate in multiple spaces, i.e. spatial, temporal or spatiotemporal (2D + time).
+To avoid redundancy and putting too much weight on certain methods, only the spatiotemporal examples of these methods will be shown in the following sections.
+However, these can be adapted for the respective other cases (spatial only, temporal only).
 
@@
 In contrast to blocking or clustering not the the spatial or temporal location is used but additional metadata such as the membership to a specific crop type [@brenning2006].
-The idea to follow a such an inherited grouping structure of the data is also referred to as "target-oriented validation" [@cast].
-Even though flexibility is limited due to the inherited grouping structure within the data, some implementations allow to specify the resulting number of folds by omitting parts of the dataset.
+The idea to follow an inherited grouping structure of the data is also referred to as "target-oriented validation" [@cast].
+Even though flexibility with respect to fold creation is limited due to the inherited grouping structure within the data, some implementations allow to specify the resulting number of folds by omitting parts of the dataset.
 
-\pkg{mlr3spatiotempcv} supports two "grouping" implementations: (1) The "Leave-location-and-time-out" family from package \pkg{CAST} and (2) "Custom Resampling" from \pkg{mlr3}.
+\pkg{mlr3spatiotempcv} supports two "grouping" implementations: (1) The "Leave-location-and-time-out" family from package \pkg{CAST} and (2) "Custom Resampling" from the \pkg{mlr3} package.
 
@@
 
-In this example the `cookfarm` dataset is used because it contains a temporal variable ("Date") which is not present in dataset `ecuador`.
+In this example the `cookfarm` dataset is used because it contains the variable ("Date") which serves as the temporal component in the dataset.
 
-Two visualization types are supported: a two-dimensional and a three-dimensional method which can be selected via the logical argument `plot3d`.
-The 3D visualization (space + time) has been implemented via \pkg{plotly}.
+Two visualization types are supported by `mlr3spatiotempcv::autoplot()` for this method: a two-dimensional and a three-dimensional one which can be selected via the logical argument `plot3D`.
+The 3D visualization (i.e. 2D + time) is done via package \pkg{plotly}.
 Because a dynamic image cannot be included in this manuscript, a static version, which can be generated by setting `static_image = TRUE`, is shown.
 
-In \pkg{mlr3spatiotempcv} the method is named `"sptcv_cstf"` because the underlying upstream implementation in package \pkg{CAST} is named `CreateSpacetimeFolds()`.
+In \pkg{mlr3spatiotempcv} the method is named `"sptcv_cstf"` which is an acronym originating from the underlying upstream implementation in package \pkg{CAST} `CreateSpacetimeFolds()`.
 
@@
 
-The following code uses the `cookfarm` dataset as the base, modifies to the `Date` variable to contain five unique levels and then creates a spatiotemporal regression task in \pkg{mlr3spatiotempcv} (Fig.\@ref(fig:lto)).
+The following code uses the `cookfarm` dataset as the base, modifies to the `Date` variable to contain five unique levels and then creates a spatiotemporal regression task in \pkg{mlr3spatiotempcv} (Fig.
+\@ref(fig:lto)).
 
@@
 Method "Leave-Location-Out" (LLO) works similar to "Leave-Time-Out" (LTO) in that it operates in a two-dimensional space.
-Here, specific locations are left out for forming the partitions while the temporal information is ignored (Fig.\@ref(fig:llo)).
+Here, specific locations are left out for forming the partitions while the temporal information is ignored (Fig. \@ref(fig:llo)).
 
@@
 To achieve this, observations are omitted and only a subset is used.
-In the special scenario with five temporal levels and five folds, the train/test split happens only at one temporal level (Fig.\@ref(fig:llto)).
+In the special scenario with five temporal levels and five folds, the train/test split happens only at one temporal level (Fig. \@ref(fig:llto)).
 

### 2021-06-25 paper/paper.Rmd
fix affil

@@
   - name: Patrick Schratz
-    affiliation: 'Friedrich-Schiller-University Jena'
+    affiliation: 'Friedrich Schiller University Jena'
     # affiliaton2: 'Ludwig-Maximilians-University Munich'

### 2021-06-25 paper/paper.Rmd
rewrite standalone implementation vs ecosystem

@@
 
-<!-- FIXME: mention missing spatiotemporal buffering method -->
+<!--mention missing spatiotemporal buffering method -->
+As of writing this paper, there is no buffering methods which operates in the spatiotemporal domain.
 
-<!-- missing: non-LOO method with buffer support -->
-<!-- FIXME: this is wrong, leave-one-disc-out does this -->
-Also no methods yet exists that makes use of a buffer zone with blocking or clustering, bridging the gap between buffered LOO-CV and fold-based non-buffer methods (`"spcv_coords"` and `"spcv_block"`).
+<!-- mention pros and cons of implementing methods in a standalone pkg vs a framework -->
+There are pros and cons on whether a new method should be integrated into an existing ecosystem such as \pkg{mlr3} directly first or distributed via a standalone package.
+On the upside, developers can benefit from existing package structures and orient themselves how previous methods were implemented.
+Also, the new method will be directly usable by all people using this ecosystem following an existing and familiar syntax.
+Contributors do not have to commit maintaining the package in the future, possibly risking its removal in case of major issues.
+In addition, they most likely get a thorough review from other community members during the implementation, possibly enhancing the code side of the implementation.
+On the contrary, people are forced to use the ecosystem and cannot easily use the method outside of it, whether its a different ecosystem or some other way.
+They might also be forced to follow certain guidelines of the ecosystem and potentially feel constrained in their way of doing things.
 
-We believe that making resampling methods available to widely used modeling frameworks is more beneficial than implementing such methods standalone in an R package.
-The latter often require special structured objects as inputs and cannot directly be re-used in subsequent modeling steps of user workflows.
-This is not an issue specific to spatiotemporal resampling methods but a general philosophy of the authors.
-
+<!-- signoff -->
 Spatiotemporal cross-validation is still not yet a paradigm that is completely established in scientific workflows even though it is discussed intensively for more than a decade now.

### 2021-06-25 paper/paper.Rmd
rewrite "how to choose a resampling method"

@@
 While various spatiotemporal resampling methods exists nowadays to account for spatiotemporal autocorrelation in modeling, there is still lots of uncertainties which method should be used for which dataset.
-The answer to this questions is not trivial and needs more research.
-One option could be to propose a checklist of dataset characteristics which would make a dataset more suitable for specific methods that could help users making an appropriate and objective decision.
+The answer to this question is not trivial and while there will most likely never be an easy answer to this, this topic needs to be discussed in more detail so that users feel more certain about their actions.
+One option could be a checklist of dataset characteristics showing which characteristics make a certain dataset more suitable for specific methods.
+This could then help users to make an appropriate and objective decision which resampling method might be a good fit for their dataset.
 

### 2021-06-25 paper/paper.Rmd
update llto figure
upd spatiotemp intro

@@
 To avoid redundancy and putting too much weight on certain methods, only the spatiotemporal examples of these methods will be shown in the following sections.
-However, these can be adapted for the respective other cases (spatial only, temporal only).
+For the respective other spaces, usually only one of the space or time variable inputs needs to be omitted.
 
@@
 ```{r llto, echo=FALSE, out.width="90%", fig.cap="Three-dimensional visualization of method 'sptcv\\_cstf' and sub-method 'Leave-Location-And-Time-Out' (LLTO).
-The grey points show the points omitted by this method.", fig.pos="ht"}
+The grey points show the points omitted by this method, i.e.
+the points which are not used in the training or test set in this fold.", fig.pos="ht"}
 knitr::include_graphics("llto.png")

### 2021-06-25 paper/paper.Rmd
small improvements

@@
 Comprehensive and up-to-date information can be found in the package's online documentation (<https://mlr3spatiotempcv.mlr-org.com>).
-The following sections, which follow the naming scheme `<method name> --- <R package>`, give an overview of most existing partitioning strategies and their visualization.
+The following sections, which follow the naming scheme `<method name> --- <R package>`, give an overview of most existing partitioning strategies and their visualization options.
 The available methods are further discussed in section \@ref(sec:disc).
@@
 
-Support for user-defined partitioning strategies is built into \pkg{mlr3} directly since version FIXME.
+Support for user-defined partitioning strategies is built into \pkg{mlr3} directly since version 0.12.0 FIXME.
 Labelled as "Custom CV", users are able to supply an external factor vector with the same length as number of observations.
@@
 
-The following example is taken from `sperrorest::partition_factor()`.
+The following example was taken from `sperrorest::partition_factor()`.
 
 <!-- FIXME: eval=TRUE when available upstream im mlr3 -->
+<!-- FIXME: add visualization when method is available -->
 
@@
 The last method from the \pkg{blockCV} package, "environmental blocking", makes use of *k-means* clustering [@hartigan1979a] in a possibly multivariate space.
-The user can specify a variable via argument `feature` which levels will be used to create clusters.
+The user can select a variable via argument `feature` specifying which levels will be used to create clusters.
 Hereby, k-means will use Euclidean distance for cluster creation.
@@
 To show the effect of spatial autocorrelation on performance estimates in cross-validation, we will apply spatial and non-spatial cross-validation to the prediction of landslides using the random forest technique (`"classif.ranger"`).
-Specifically, we use spatial partitioning based on $k$-means clustering of the coordinates (method `rsmp("repeated_spcv_coords")`) and compare its results to a non-spatial (random) cross-validation (`rsmp("repeated_cv")`), both with (only) four folds and two repetitions for demonstration.
-We choose the area Under the ROC curve (`"classif.auc"`) as the performance measure for this 'soft' classification task.
+Specifically, we use spatial partitioning based on $k$-means clustering of the coordinates (method `"repeated_spcv_coords"`) and compare its results to a non-spatial (random) cross-validation (`"repeated_cv"`), both with (only) four folds and two repetitions for demonstration.
+We choose the area Under the ROC curve (`"classif.auc"`) as the performance measure for this classification task.
 
@@
 Next, the random forest learner (`lrn("classif.ranger")`) is initialized with default hyperparameters and the prediction type set to `"probability"` due to the binary response variable.
-A set of commonly used learners is available in \pkg{mlr3learners} [@mlr3learners], including the random forest implementation after @ranger.
+A set of commonly used learners is available in package \pkg{mlr3learners} [@mlr3learners], including the random forest implementation after @ranger.
 
@@
 
-The results from the CV are then aggregated using the measure `"classif.auc"`, which is the package notation for the Area Under the Receivers Operating Curve (AUROC) measure.
+The results from the CV are then aggregated using the measure `"classif.auc"`, which is the \pkg{mlr3} package notation for the Area Under the Receivers Operating Curve (AUROC) measure.
 
@@
 The magnitude of this difference is variable as it depends on the dataset, the magnitude of STAC and the learner itself.
-Algorithms with a higher tendency of overfitting to the training set will have a large spread in such scenarios.
+Algorithms with a higher tendency of overfitting to the training set will have a larger spread in such scenarios.
 
@@
 <!--mention missing spatiotemporal buffering method -->
-As of writing this paper, there is no buffering methods which operates in the spatiotemporal domain.
+At the time of writing this article, there is no buffering methods which operates in the spatiotemporal domain.
+Though we encourage users generally to check out the supported methods in the \pkg{mlr3spatiotempcv} package directly to potentially discover new methods which might have been added to the package in the meantime.
 

### 2021-06-26 paper/paper.Rmd
optimize discussion

@@
 
-While in the example the `"spcv_coords"` method was used, this should not imply that this method is the best or only method suitable for this example task.
+The question which resampling method should be chosen for a dataset at hand comes up regularly in practice.
+Even though there is no definitive and easy answer, we would like to give some guidance in this section to help finding the most appropriate method.
+
+Although the case study example in section \@ref(sec:case-study) used the `"spcv_coords"` method, this should not imply that this method is the best or only method suitable for this example task.
 Even though this method is frequently used, it was mainly chosen because of it's clear visual grouping differences when comparing spatial to random partitioning.
@@
 In fact, most often multiple spatial partitioning methods can be used for a dataset.
-It is recommended (required) that users familiarize themselves with multiple methods and decide which method to choose based on the specific characteristics of the dataset at hand.
+It is recommended (required) that one familiarizes themselves with multiple methods and decide which method to choose based on the specific characteristics of the dataset at hand.
 There are various variables influencing the partitioning method choice for an environmental dataset:
@@
 Based on these criteria users must choose a matching partitioning method that is either more restrictive (by discarding observations for fold creation) or more liberal (by using all observations and eventually ignoring natural grouping patterns).
-For almost all methods implemented in \pkg{mlr3spatiotempcv}, there is a scientific publication describing the strengths and weaknesses of the respective approach (either linked in the help file of the package or its respective upstream packages).
+For almost all methods implemented in \pkg{mlr3spatiotempcv}, there is a scientific publication describing the strengths and weaknesses of the respective approach (either linked in the help file of the package or its respective upstream packages), hence we are refraining from repeating them again in this study.
 
 In the example above, a cross-validation without hyperparameter tuning was shown.
-In practice most often a nested CV is desired.
-In this case it is recommended to use the same spatial partitioning method for the inner loop (= tuning level).
-See @schratz2019 for more details and chapter 11 of [Geocomputation with R](https://geocompr.robinlovelace.net/spatial-cv.html) [@lovelace2019].
+In ML practice most often a nested CV should be desired to optimize model hyperparameter within the CV.
+In such scenarios it is recommended to use the same spatial partitioning method for the inner loop (= tuning level) that was chosen for the outer one.
+See @schratz2019 for more details as well as chapter 11 of [Geocomputation with R](https://geocompr.robinlovelace.net/spatial-cv.html) [@lovelace2019].
 
-## Comparing spatiotemporal resampling methods
+## Comparing and discussing spatiotemporal resampling methods
 
@@
 <!-- sptcv LLTO: discards many observations -->
-All methods which make use of all variables in the dataset suffer from the problem that on the border between train and test sets, both in the spatial and temporal domain, autocorrelation is quite high.
-Only methods that omit some observations (`"spcv_buffer"`, `"sptcv_cstf"` with both `space_var` and `time_var` set) are able to reduce this problem.
+All methods which make use of all variables in the dataset suffer from the problem that on the border between train and test sets, both in the spatial and temporal domain, autocorrelation is relatively high (unless the dataset inherits natural gaps between train and test sets).
+Only methods that omit some observations (`"spcv_buffer"`, `"spcv_tiles"`, `"sptcv_cstf"` with both `space_var` and `time_var` set) are able to reduce the impact of this issue.
 However, these then suffer from the fact of potentially not being able to properly represent the dataset because only a subset of observations is used.
@@
 <!-- sptcv LLO & LTO: requires additional variable to specify groupings -->
-Methods `"spcv_cstf"` (with only one of `space_var` or `time_var` set) require a variable in the dataset which should be used for grouping.
-The specification of such is not always easy because the selection of such a variable also implies that the selected groups inherit substantial autocorrelation and differ substantially from other groups.
-Also, if the variable contains too many groups, the difference within train/test splits may become undesirably high and tend towards a LOO CV.
+The `"spcv_cstf"` methods (with only one of `space_var` or `time_var` set) require a variable in the dataset which should be used for grouping.
+The specification of such is not always easy because the grouping should ensure that the selected groups inherit substantial autocorrelation within themselves and simultaneously differ substantially from other groups.
+Also, if the variable selected contains too many groups, the difference within train/test splits may become undesirably high and tend towards a LOO CV [@meyer2018].
 
@@
 <!-- spcv_coords: limited variance due to limited k-means possibilities -->
-For repeated executions of method `spcv_coords`, the variance between repetitions could be quite small due to the limited possibilities of different `k-means` clusters on the same dataset.
+For repeated executions of method `spcv_coords`, the variance between repetitions can become quite small due to the limited possibilities of differing `k-means`-based clusters on the same dataset [@brenning2012].
 This applies even more to `"spcv_block"` with options `selection = "systematic"` and `selection = "checkerboard"` as the fold assignments are the same in each repetition.
-Only methods with a random component (e.g. `"spcv_coords"` or `"spcv_block"` with option `selction = "random"`) may be suited for using multiple repetitions.
+In general, only methods making use of a random component (e.g. `"spcv_coords"` or `"spcv_block"` with option `selction = "random"`) may be suited for using multiple repetitions.
 
@@
 Method `"spcv_buffer"` makes use of the removal of observations to reduce the influence of spatial autocorrelation between train and test set.
-This might be pretty effective in reducing autocorrelation but also requires to discard many variables and comes with the downsides of LOO-CV: many model fits which can be computationally very demanding and a high variance due to the high similarity of training datasets [@blockCV].
+This can be a pretty effective way to reduce autocorrelation but also requires to discard many observations and comes with the downsides of a LOO-CV: many model fits which can be computationally very demanding and the results inheriting a high variance due to the high differences in the test sets (also commonly known as the "bias-variance tradeoff" [@wahba1998; @blockCV].
 
 <!-- spcv coords: mention generic name -->
-In @brenning2012, which is commonly used as the main reference for this k-means based spatial clustering approach, the method was not assigned a unique name.
-This brings potential confusion for users and the community as the term "spatial CV" is used a generic one for the idea of accounting for spatial autocorrelation for spatial datasets.
-Hence, in \pkg{mlr3spatiotempcv} this method was named `"spcv_coords"`, making use of the fact that the spatial coordinates are used for a cluster-based partition creation.
+In @brenning2012, which is the reference commonly used for the k-means based spatial partitioning approach, the method was not introduced with a unique name.
+Over time, studies which have used this method mainly referenced it using the generic "spatial CV" terminology.
+However, using this notation potentially introduces confusion as the term "spatial CV" is a generic one referring to the idea of accounting for spatial autocorrelation in spatial datasets and does not necessarily refer to a specific partitioning method in the first place.
+Hence, in package \pkg{mlr3spatiotempcv} this method was named `"spcv_coords"`, making use of the fact that the spatial coordinates are used for a cluster-based partition creation.
 
@@
 <!-- env block: more than one variable may lead to unforeseeable results -->
-Method `"spcv_env"` is complicated in that it requires the user to limit the clustering to (a) specific variable(s).
-This implies that only this/these variable(s) suffer from autocorrelation and others can be ignored.
-However, the influence of a single variable is hard to estimate and the same applies for estimating the influence of the non-accounted variables which might add substantial bias to the CV.
+Method `"spcv_env"` is complicated to use in that it requires the user to limit the clustering to one or more variables [@blockCV].
+This implies that only this/these variable(s) suffer from autocorrelation and others are ignored.
+However, it is not possible to estimate how much a single variable accounts for the autocorrelation in the data and the same applies for estimating the influence of the non-accounted variables which might add substantial bias to the CV.
 
@@
 
-```{r echo=FALSE}
+```{r echo=FALSE, cache=FALSE}
 sessionInfo()

### 2021-06-26 paper/paper.Rmd
update figure sizes

@@
 
-```{r buffer, fig.cap="Visualization of the 'Spatial buffering' method from package `blockCV`.", out.width="50%", fig.pos="ht"}
+```{r buffer, fig.cap="Visualization of the 'Spatial buffering' method from package `blockCV`.", out.width="40%", fig.pos="ht"}
 resampling_buffer = rsmp("spcv_buffer", theRange = 1000)
@@
 
-```{r disc, fig.cap="Visualization of 'Leave-one-disc-out' method.", out.width="50%"}
+```{r disc, fig.cap="Visualization of 'Leave-one-disc-out' method.", out.width="40%"}
 resampling_disc = rsmp("spcv_disc", folds = 5, radius = 300L, buffer = 400L)
@@
 
-```{r block-random, fig.cap="Spatial blocking with option `selection = 'random'`.", out.width="50%", fig.pos="ht"}
+```{r block-random, fig.cap="Spatial blocking with option `selection = 'random'`.", out.width="40%", fig.pos="ht"}
 resampling_block_random = rsmp("spcv_block", range = 1000, folds = 5)
@@
 
-```{r block-systematic, fig.cap="Spatial blocking with option `selection = 'systematic'`.", out.width="50%", fig.pos="ht"}
+```{r block-systematic, fig.cap="Spatial blocking with option `selection = 'systematic'`.", out.width="40%", fig.pos="ht"}
 resampling_block_systematic = rsmp("spcv_block",
@@
 
-```{r coords, fig.cap="Method 'Spatial CV' showing k-means clustering based on coordinates.", out.width="50%", fig.pos="ht"}
+```{r coords, fig.cap="Method 'Spatial CV' showing k-means clustering based on coordinates.", out.width="40%", fig.pos="ht"}
 resampling_coords = rsmp("spcv_coords", folds = 5)$instantiate(task)
@@
 
-```{r lto, echo=FALSE, out.width="90%", fig.cap="Three-dimensional visualization of method `sptcv\\_cstf` and sub-method Leave-Time-Out (LTO).
+```{r lto, echo=FALSE, out.width="70%", fig.cap="Three-dimensional visualization of method `sptcv\\_cstf` and sub-method Leave-Time-Out (LTO).
 To support showcasing the temporal grouping, five folds and five temporal levels in variable `Date` were used in this example.", fig.pos="ht"}
@@
 
-```{r llo, echo=FALSE, out.width="90%", fig.cap="Three-dimensional visualization of method 'sptcv\\_cstf' and sub-method 'Leave-Location-Out' (LLO).
+```{r llo, echo=FALSE, out.width="70%", fig.cap="Three-dimensional visualization of method 'sptcv\\_cstf' and sub-method 'Leave-Location-Out' (LLO).
 To support showcasing the spatial grouping, the viewing angle has been adjusted to a view from top.
@@
 
-```{r llto, echo=FALSE, out.width="90%", fig.cap="Three-dimensional visualization of method 'sptcv\\_cstf' and sub-method 'Leave-Location-And-Time-Out' (LLTO).
+```{r llto, echo=FALSE, out.width="70%", fig.cap="Three-dimensional visualization of method 'sptcv\\_cstf' and sub-method 'Leave-Location-And-Time-Out' (LLTO).
 The grey points show the points omitted by this method, i.e.
@@
 
-```{r custom-cv,  fig.cap="Environmental blocking using a single predictor.", out.width="50%", fig.pos="ht", eval=FALSE}
+```{r custom-cv,  fig.cap="Environmental blocking using a single predictor.", out.width="40%", fig.pos="ht", eval=FALSE}
 breaks = quantile(task$data()$dem, seq(0, 1, length = 6))
@@
 
-```{r env-1, fig.cap="Environmental blocking using a single predictor.", out.width="50%", fig.pos="ht"}
+```{r env-1, fig.cap="Environmental blocking using a single predictor.", out.width="40%", fig.pos="ht"}
 resampling_env = rsmp("spcv_env",
@@
 
-```{r env-2, fig.cap="Environmental blocking using multiple predictors.", out.width="50%", fig.pos="ht"}
+```{r env-2, fig.cap="Environmental blocking using multiple predictors.", out.width="40%", fig.pos="ht"}
 resampling_env_multi = rsmp("spcv_env",

### 2021-06-26 paper/paper.Rmd
move case study vis after execution, add new section "should new methods be added to existing ML ecosystems?"

@@
 While \pkg{mlr3spatiotempcv} solely focuses on spatiotemporal resampling methods and their visualization, other packages such as \pkg{mlr3spatial} or \pkg{mlr3temporal} are planned in the mlr3 ecosystem to provide dedicated spatiotemporal learner and prediction methods.
-
-# Application in a Case Study
-
-We will use a well-known case study to demonstrate the application of spatial resampling techniques.
-This binary classification task on landslide susceptibility in Ecuador [@muenchow2012] is available as a built-in task via `tsk("ecuador")`.
-
-...
-<!-- FIXME: @alex: do you want to see more into here? -->
-
-Because the spatial grouping of the k-means based approach (`"spcv_coords"`) (Fig.
-\@ref(fig:vis-spcv)) contrasts visually very well to NSpCV (random) partitioning (Fig.
-\@ref(fig:vis-nspcv)), this method was chosen to showcase the 2D plotting capabilities of \pkg{mlr3spatiotempcv}.
-
-The partition plots shown in the following are also used in the case study in section \@ref(sec:case-study).
-
-Besides the required arguments of the \pkg{mlr3} superclass `Task`, `TaskClassifST` / `TaskRegrST` classes also require the following arguments passed via a named list `extra_args`:
-
-- `coordinate_names`: Names of the features which represent the spatial coordinates.
-  This is automatically inferred when a `sf` object is passed.
-- `coords_as_features`: Whether the coordinates should be used as features during modeling.
-  By default coordinates are not used as predictors.
-- `crs`: The coordinate reference system of the data. Valid inputs are PROJ4 strings or EPSG codes in the format `ESPG: <code>`.
-
-```{r create-task}
-library(mlr3)
-library(mlr3spatiotempcv)
-
-# be less verbose
-lgr::get_logger("bbotk")$set_threshold("warn")
-lgr::get_logger("mlr3")$set_threshold("warn")
-
-set.seed(42)
-
-# equal to `tsk("ecuador")`
-b = mlr3::as_data_backend(ecuador)
-b$hash = "_mlr3_tasks_ecuador_" # optional
-task = TaskClassifST$new(
-  id = "ecuador", b, target = "slides", positive = "TRUE",
-  extra_args = list(
-    coordinate_names = c("x", "y"), coords_as_features = FALSE,
-    crs = "+proj=utm +zone=17 +south +datum=WGS84 +units=m +no_defs") # EPSG:32717
-)
-```
-
-Due to space limitations only the first two folds are plotted.
-
-```{r vis-spcv, fig.cap="Visualization of spatial partitioning", fig.pos="ht", fig.height=6, fig.width=8}
-resampling_sp = rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
-resampling_sp$instantiate(task)
-
-autoplot(resampling_sp, task, fold_id = c(1:2), crs = 32717, size = 0.8) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
-```
-
-```{r vis-nspcv, fig.cap="Visualization of random partitioning", fig.pos="ht", fig.height=6, fig.width=8}
-resampling_nsp = rsmp("repeated_cv", folds = 4, repeats = 2)
-resampling_nsp$instantiate(task)
-
-autoplot(resampling_nsp, task, fold_id = c(1:2), crs = 32717, size = 0.8) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
-```
 
@@
 
+We will use a well-known case study to demonstrate the application of spatial resampling techniques.
+This binary classification task on landslide susceptibility in Ecuador [@muenchow2012] is available as a built-in task via `tsk("ecuador")`.
+
+Because the spatial grouping of the k-means based approach (`"spcv_coords"`) (Fig.
+\@ref(fig:vis-spcv)) contrasts visually very well to NSpCV (random) partitioning (Fig.
+\@ref(fig:vis-nspcv)), this method was chosen to showcase the 2D plotting capabilities of \pkg{mlr3spatiotempcv}.
+
+Besides the required arguments of the \pkg{mlr3} superclass `Task`, the `TaskClassifST` / `TaskRegrST` classes from package \pkg{mlr3spatiotempcv}  also require the following arguments during construction which must be passed via a named list `extra_args`:
+
+- `coordinate_names`: Names of the features which represent the spatial coordinates.
+  This is automatically inferred when a `sf` object is passed.
+- `coords_as_features`: Whether the coordinates should be used as features during modeling.
+  By default coordinates are not used as predictors.
+- `crs`: The coordinate reference system of the data. Valid inputs are PROJ4 strings or EPSG codes in the format `ESPG:<code>`.
+
+At first all necessary R packages are loaded and a lower verbosity is set to keep the output tidy.
+Also a seed is set for deterministic execution and the \pkg{mlr3spatiotempcv} built-in task `"ecuador"` is loaded via `tsk()`.
+
+```{r create-task}
+library(mlr3)
+library(mlr3spatiotempcv)
+
+# be less verbose
+lgr::get_logger("bbotk")$set_threshold("warn")
+lgr::get_logger("mlr3")$set_threshold("warn")
+
+set.seed(42)
+
+# equal to `tsk("ecuador")`
+b = mlr3::as_data_backend(ecuador)
+b$hash = "_mlr3_tasks_ecuador_" # optional
+task = TaskClassifST$new(
+  id = "ecuador", b, target = "slides", positive = "TRUE",
+  extra_args = list(
+    coordinate_names = c("x", "y"), coords_as_features = FALSE,
+    crs = "EPSG:32717")
+)
+```
+
 To show the effect of spatial autocorrelation on performance estimates in cross-validation, we will apply spatial and non-spatial cross-validation to the prediction of landslides using the random forest technique (`"classif.ranger"`).
@@
 
-At first all necessary R packages are loaded and a lower verbosity is set to keep the output tidy.
-Also a seed is set for deterministic execution and the \pkg{mlr3spatiotempcv} built-in task `"ecuador"` is loaded via `tsk()`.
 Next, the random forest learner (`lrn("classif.ranger")`) is initialized with default hyperparameters and the prediction type set to `"probability"` due to the binary response variable.
@@
 ```{r prepare}
-library("mlr3")
-library("mlr3spatiotempcv")
 library("mlr3learners")
-
-task = tsk("ecuador")
 
@@
 rr_sp$aggregate(measures = msr("classif.auc"))
+```
+
+## Fold visualization
+
+Finally, we visualize the folds that were used during performance estimation making use of the generic `autoplot()` function in package \pkg{mlr3spatiotempcv}.
+Due to space limitations only the first two folds are shown.
+
+```{r vis-spcv, fig.cap="Visualization of spatial partitioning", fig.pos="ht", fig.height=6, fig.width=8}
+resampling_sp = rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
+resampling_sp$instantiate(task)
+
+autoplot(resampling_sp, task, fold_id = c(1:2), crs = 32717, size = 0.8) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```
+
+```{r vis-nspcv, fig.cap="Visualization of random partitioning", fig.pos="ht", fig.height=6, fig.width=8}
+resampling_nsp = rsmp("repeated_cv", folds = 4, repeats = 2)
+resampling_nsp$instantiate(task)
+
+autoplot(resampling_nsp, task, fold_id = c(1:2), crs = 32717, size = 0.8) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
 ```
@@
 
-# Summary and outlook
-
-<!-- missing: Guidelines how to choose a resampling method -->
-While various spatiotemporal resampling methods exists nowadays to account for spatiotemporal autocorrelation in modeling, there is still lots of uncertainties which method should be used for which dataset.
-The answer to this question is not trivial and while there will most likely never be an easy answer to this, this topic needs to be discussed in more detail so that users feel more certain about their actions.
-One option could be a checklist of dataset characteristics showing which characteristics make a certain dataset more suitable for specific methods.
-This could then help users to make an appropriate and objective decision which resampling method might be a good fit for their dataset.
-
-<!--mention missing spatiotemporal buffering method -->
-At the time of writing this article, there is no buffering methods which operates in the spatiotemporal domain.
-Though we encourage users generally to check out the supported methods in the \pkg{mlr3spatiotempcv} package directly to potentially discover new methods which might have been added to the package in the meantime.
+## Should new methods be added to existing ML ecosystems or made available in a new package?
 
@@
 
+# Summary and outlook
+
+<!-- missing: Guidelines how to choose a resampling method -->
+The \pkg{mlr3spatiotempcv} package is the first package to bundle and categorize spatiotemporal resampling methods in R.
+While various spatiotemporal resampling methods exists nowadays to account for spatiotemporal autocorrelation in modeling, there is still lots of uncertainty which method should be used for which dataset.
+The answer to this question is not trivial and while there will most likely never be an easy answer to this, this topic needs to be discussed in more detail so that users feel more certain about their actions.
+One option could be the proposal of a dataset characteristic checklist showing which characteristics make a certain dataset more suitable for specific methods.
+This could then help users to make an appropriate and objective decision which resampling method might be a good fit for their dataset.
+
+<!--mention missing spatiotemporal buffering method -->
+At the time of writing this article, there is no buffering method which operates in the spatiotemporal domain.
+We encourage readers of this article to check out the supported methods in the latest released version of the \pkg{mlr3spatiotempcv} package directly to potentially discover new methods which might have been added to the package in the meantime.
+
 <!-- signoff -->

### 2021-06-26 paper/paper.Rmd
machine-learning -> machine learning

@@
 
-Although alternatives such as the bootstrap exist and have some advantages [@efron1983; @hand1997], cross-validation is a particularly well-established, easy-to-implement algorithm for *model assessment* of supervised machine-learning models (@efron1983 and next section) and *model selection* [@arlot2010].
+Although alternatives such as the bootstrap exist and have some advantages [@efron1983; @hand1997], cross-validation is a particularly well-established, easy-to-implement algorithm for *model assessment* of supervised machine learning models (@efron1983 and next section) and *model selection* [@arlot2010].
 In its basic form, it is based on resampling the data without paying attention to any possible dependence structure, which may arise from, e.g., grouped or structured data [@brenning2008; @pena2015], or underlying environmental processes inducing some sort of spatial coherence at the landscape scale [@brenning2005].
-In treating dependent observations as independent, or ignoring autocorrelation, cross-validation test samples may in fact be heavily correlated with, or even pseudo-replicates of, the data used for training the model, which introduces a potentially severe bias, especially for flexible machine-learning models [@brenning2005].
+In treating dependent observations as independent, or ignoring autocorrelation, cross-validation test samples may in fact be heavily correlated with, or even pseudo-replicates of, the data used for training the model, which introduces a potentially severe bias, especially for flexible machine learning models [@brenning2005].
 
@@
 
-While it is not the purpose of this paper to recommend specific resampling schemes for specific use cases, this example may suffice to motivate the use of spatial and spatiotemporal cross-validation techniques, and the need for a unified framework and computational toolbox that accommodate a variety of prediction tasks that may be applicable to a broad range of application scenarios.
+While it is not the purpose of this article to recommend specific resampling schemes for specific use cases, this example may suffice to motivate the use of spatial and spatiotemporal cross-validation techniques, and the need for a unified framework and computational toolbox that accommodate a variety of prediction tasks that may be applicable to a broad range of application scenarios.
 \pkg{mlr3spatiotempcv} is such a toolbox.
@@
 In contrast, \pkg{blockCV} and \pkg{ENMeval} were developed for block and buffer resampling with a focus on species distribution modelling [@blockCV; @rest2014; @muscarella2014].
-Neither of these have been integrated into established machine-learning frameworks such as \pkg{mlr} or \pkg{caret}, and all of them lack support of temporal prediction tasks.
+Neither of these have been integrated into established machine learning frameworks such as \pkg{mlr} or \pkg{caret}, and all of them lack support of temporal prediction tasks.
 The \pkg{CAST} package, in contrast, focuses on spatiotemporal prediction tasks in a \pkg{caret} framework [@cast; @meyer2018].
 One limitation of these packages is also that they focus on model assessment, while the proposed implementation within the \pkg{mlr} framework offers a seamless integration into model selection.
-It is worth noting that a spatial cross-validation library \pkg{spacv} has also recently been developed for Python 3, which can be used with the \pkg{scikit-learn} machine-learning framework [@pedregosa2011].
+It is worth noting that a spatial cross-validation library \pkg{spacv} has also recently been developed for Python 3, which can be used with the \pkg{scikit-learn} machine learning framework [@pedregosa2011].
 
-Thus, \pkg{mlr3spatiotempcv} implements for the first time a comprehensive state-of-the-art compilation of spatial and spatiotemporal partitioning schemes that is well-integrated into a comprehensive machine-learning framework in R, the \pkg{mlr3} ecosystem [@mlr3].
+Thus, \pkg{mlr3spatiotempcv} implements for the first time a comprehensive state-of-the-art compilation of spatial and spatiotemporal partitioning schemes that is well-integrated into a comprehensive machine learning framework in R, the \pkg{mlr3} ecosystem [@mlr3].
 This package is furthermore equipped with a variety of two- and three-dimensional diagram types.
@@
 After presenting the conceptual background in the following section, we will outline the overall structure of the \pkg{mlr3spatiotempcv} package.
-We will then compare and contrast the various spatial and spatiotemporal partitioning techniques, before we demonstrate their application in a machine-learning model assessment in the following section.
+We will then compare and contrast the various spatial and spatiotemporal partitioning techniques, before we demonstrate their application in a machine learning model assessment in the following section.
 We finally discuss implications and provide an outlook.
@@
 \]
-where $f$ is a machine-learning algorithm, i.e. a mapping that trains a model $\hat{f}_\mathcal{S}$ using any suitable training sample $\mathcal{S}$.
+where $f$ is a machine learning algorithm, i.e. a mapping that trains a model $\hat{f}_\mathcal{S}$ using any suitable training sample $\mathcal{S}$.
 The use of $k=5$ or $k=10$ folds is most commonly seen in practice, and these preferences are also supported by theory [@bengio2004, @cawley2010].
@@
 It should be noted that $k$-fold cross-validation with a large value of $k$, and LOO-CV in particular ($k=n$), is not only very time-consuming since the model has to be trained $k$ times; these models will also be nearly identical since only a tiny fraction of the data is withheld, and therefore estimation bias increases.
-'Pure' LOO-CV is therefore not recommended for machine-learning model assessment.
+'Pure' LOO-CV is therefore not recommended for machine learning model assessment.
 
@@
 
-With the increased awareness of the importance of spatial and spatiotemporal resampling strategies and the growing popularity of R in environmental modeling and geocomputation, it is important to equip machine-learning frameworks such as \pkg{mlr3} with suitable algorithms.
-In this context, the \pkg{mlr3} ecosystem stands out as a unified, object-oriented and extendable framework designed to accommodate numerous machine-learning tasks with a variety of learners, feature and model selection tools, and model assessment capabilities [@mlr3; @mlr3book].
+With the increased awareness of the importance of spatial and spatiotemporal resampling strategies and the growing popularity of R in environmental modeling and geocomputation, it is important to equip machine learning frameworks such as \pkg{mlr3} with suitable algorithms.
+In this context, the \pkg{mlr3} ecosystem stands out as a unified, object-oriented and extendable framework designed to accommodate numerous machine learning tasks with a variety of learners, feature and model selection tools, and model assessment capabilities [@mlr3; @mlr3book].
 All of these are supported by advanced visualization tools, which are particularly important in a spatial and spatiotemporal setting.
@@
 Issues that practitioners often face include varying argument lists of learners, different return values of `predict` methods, and support for only specific feature types.
-These challenges result in substantial overhead and possible reproducibility issues, which are exacerbated by asynchronous development timelines of different components of the machine-learning pipelines.
+These challenges result in substantial overhead and possible reproducibility issues, which are exacerbated by asynchronous development timelines of different components of the machine learning pipelines.
 
 Within the \pkg{mlr3} ecosystem, partitioning strategies are represented by their own objects of class \texttt{Resampling}, most of which are available within \pkg{mlr3} itself (e.g., regular random CV); other specialized strategies (e.g., \texttt{TaskClassifST} or \texttt{TaskRegrST}) are defined in extension packages such as \pkg{mlr3spatiotempcv}.
-In the machine-learning pipeline, these objects define the data splits used for model evaluation and selection (hyperparameter tuning) by machine-learning algorithms.
+In the machine learning pipeline, these objects define the data splits used for model evaluation and selection (hyperparameter tuning) by machine learning algorithms.
 Spatial and spatiotemporal partitioning techniques in \pkg{mlr3spatiotempcv} are currently mostly imported and interfaced from other packages, in particular \pkg{sperrorest}, \pkg{blockCV} and \pkg{CAST} [@blockCV; @brenning2012; @cast], in order to expose them to \pkg{mlr3} functionality.

### 2021-06-26 paper/paper.Rmd
optimize intro

@@
 In crop classification from remotely-sensed data, for instance, learning samples routinely contain multiple grid cells from a sample of fields with known crop type, for instance 2000 grid cells from 100 fields scattered across a large study region.
-The purpose of training a model on this sample is to make predictions on other, new *fields* within the same geographic domain (*intra-domain* prediction; @brenning2005) --- not *within* the same field, which obviously presents only one crop type.
-In this situation it would therefore seem rather foolish to train a model on a simple random subsample of grid cells, and test it on the remaining data, i.e.
+The purpose of training a model on this sample is to make predictions on other, new fields within the same geographic domain (*intra-domain* prediction; @brenning2005) --- not *within* the same field, which obviously presents only one crop type.
+In this situation it would therefore seem rather unwise to train a model on a simple random subsample of grid cells, and test it on the remaining data, i.e.
 using other grid cells from the same fields, as if we wanted to predict within a field.
 The results from this performance assessment would be over-optimistic, and perhaps badly so.
-To mimic the predictive situation for which the model is trained, we'd rather have to resample at the level of fields, not grid cells [@pena2015].
+To mimic the predictive situation for which the model is trained, one would rather have to resample at the level of fields, not grid cells [@pena2015].
 If the model was to be applied to adjacent agricultural districts, i.e.
@@
 This toolbox, implemented as an open-source R package, builds upon and generalizes several existing toolboxes that have been developed in recent years for more specific settings (see Table \@ref(tab:sptcv-methods)).
-The earliest and most comprehensive of these implementations is the \pkg{sperrorest} package [@brenning2012], which provides an extensible framework and includes predefined resampling strategies based on, for example, blocking, grouping and buffering.
+The earliest and most comprehensive of these implementations is the \pkg{sperrorest} R package [@brenning2012], which provides an extensible framework and includes predefined resampling strategies based on, for example, blocking, grouping and buffering.
 In contrast, \pkg{blockCV} and \pkg{ENMeval} were developed for block and buffer resampling with a focus on species distribution modelling [@blockCV; @rest2014; @muscarella2014].
-Neither of these have been integrated into established machine learning frameworks such as \pkg{mlr} or \pkg{caret}, and all of them lack support of temporal prediction tasks.
-The \pkg{CAST} package, in contrast, focuses on spatiotemporal prediction tasks in a \pkg{caret} framework [@cast; @meyer2018].
-One limitation of these packages is also that they focus on model assessment, while the proposed implementation within the \pkg{mlr} framework offers a seamless integration into model selection.
-It is worth noting that a spatial cross-validation library \pkg{spacv} has also recently been developed for Python 3, which can be used with the \pkg{scikit-learn} machine learning framework [@pedregosa2011].
+Neither of these however have been integrated into established machine learning frameworks such as \pkg{mlr}/\pkg{mlr3} or \pkg{caret}/\pkg{tidymodels}, and all of them lack support of temporal prediction tasks.
+The \pkg{CAST} package, in contrast, focuses on spatiotemporal prediction tasks and makes use of some functions of the \pkg{caret} framework [@cast; @meyer2018].
+One limitation of these packages is the sole focus on model assessment, while the proposed implementation within the \pkg{mlr3} framework offers a seamless integration into model selection.
+It is worth noting that a spatial cross-validation library named \pkg{spacv} has also recently been developed for Python3, which can be used with the \pkg{scikit-learn} machine learning framework [@pedregosa2011].
 
 Thus, \pkg{mlr3spatiotempcv} implements for the first time a comprehensive state-of-the-art compilation of spatial and spatiotemporal partitioning schemes that is well-integrated into a comprehensive machine learning framework in R, the \pkg{mlr3} ecosystem [@mlr3].
-This package is furthermore equipped with a variety of two- and three-dimensional diagram types.
-Our hope is that this implementation will facilitate reproducible geospatial modelling and code-sharing across a broad range of application domains.
+This package is furthermore equipped with a variety of two- and three-dimensional visualization capabilities.
+Our hope is that this implementation will facilitate reproducible geospatial modeling and code-sharing across a broad range of application domains.
 
 The purpose of this paper is to give an overview of the methods implemented in the R package \pkg{mlr3spatiotempcv}.
-After presenting the conceptual background in the following section, we will outline the overall structure of the \pkg{mlr3spatiotempcv} package.
-We will then compare and contrast the various spatial and spatiotemporal partitioning techniques, before we demonstrate their application in a machine learning model assessment in the following section.
-We finally discuss implications and provide an outlook.
+After presenting the conceptual background in the following section, the overall structure of the \pkg{mlr3spatiotempcv} package is outlined.
+Next, various spatial and spatiotemporal partitioning techniques are contrasted and compared, before their application is demonstrated in a machine learning model assessment in the following section.
+Finally, limitations of existing methods are discussed and an outlook is given.
 
-# Spatial and Spatiotemporal Cross-validation
+# Spatial and spatiotemporal cross-validation
 

### 2021-06-27 paper/paper.Rmd
minor
group level vs custom cv

@@
 
-### Group-Level CV
+### Group-level resampling
 
@@
 
-<!-- FIXME: eval=TRUE when available upstream im mlr3 -->
-<!-- FIXME: add visualization when method is available -->
+FIXME: eval=TRUE when available upstream im mlr3
+FIXME: add visualization when method is available
 
@@
 
+<!-- group level CV vs custom CV -->
+\pkg{mlr3spatiotempcv} supports two factor-based partitioning schemes: "group-level resampling" and "custom resampling".
+Both are similar in that they require external user input for partition creation, i.e. a factor variable which assigns all observations into a distinct group.
+While this the entire idea of "custom resampling" and each group becomes a test set in the CV, "group-level resampling" methods apply additional aggregation on top of these initial groups: for example, users can supply an integer value to an argument named `fold` in the LLO method (e.g.
+`5L`) which combines the initial groups (e.g.
+$n = 20$) into their final fold-level partitions, which would result in four groups per partition [@meyer2018].
+The important point is that observations from within one group will either be assigned to the train or test set and never be split up.
+"Group-level resampling" can be, in contrast to "custom resampling", repeated multiple times because multiple different aggregations are possible in theory.
+However, it also requires the amount of initial groups to be substantially higher than the amount of folds so that a sensible aggregation is possible in the first place.
+
 <!-- spcv coords: mention generic name -->

### 2021-06-27 paper/paper.Rmd
rewrite

@@
 Only methods that omit some observations (`"spcv_buffer"`, `"spcv_tiles"`, `"sptcv_cstf"` with both `space_var` and `time_var` set) are able to reduce the impact of this issue.
-However, these then suffer from the fact of potentially not being able to properly represent the dataset because only a subset of observations is used.
+The downside is the removal of observations from the dataset which possibly reduces the generalization power of the trained model and by this, depending on the specific observations that were removed, adds substantial bias to the CV results.
 

### 2021-06-27 paper/paper.Rmd
discuss setting `theRange` arg in blocking CV

@@
 
+<!-- spcv_block: how to set argument `theRange` -->
+Method `"spcv_block"` requires the user to supply a value for argument `theRange` which denotes the block size used to create the blocking scheme across the study area.
+This specification is non-trivial and therefore package \pkg{blockCV} provides some helper functions (`spatialAutoRange()` and `rangeExplorer()`) to conduct a data-driven estimation of the distance at which the spatial autocorrelation in the data levels off [@blockCV].
+According to the package authors, this estimate should then be used for argument `theRange` to have a sensible value for the block sizes created in method `"spcv_block"`.
+
 <!-- sptcv LLO & LTO: requires additional variable to specify groupings -->

### 2021-07-02 paper/paper.Rmd
prepare custom cv

@@
 ```{r buffer, fig.cap="Visualization of the 'Spatial buffering' method from package `blockCV`.", out.width="40%", fig.pos="ht"}
+task = tsk("ecuador")
 resampling_buffer = rsmp("spcv_buffer", theRange = 1000)
@@
 
-```{r custom-cv,  fig.cap="Environmental blocking using a single predictor.", out.width="40%", fig.pos="ht", eval=FALSE}
+```{r custom-cv,  fig.cap="Custom resampling using an external factor vector.", out.width="40%", fig.pos="ht", eval=FALSE}
 breaks = quantile(task$data()$dem, seq(0, 1, length = 6))
@@
 resampling_custom = rsmp("custom_cv")
-resampling_custom$instantiate(task, split = zclass)
+resampling_custom$instantiate(task, f = zclass)
+
+autoplot(resampling_custom,
+  size = 0.8, fold_id = 1, task = task) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
 ```

### 2021-07-02 paper/paper.Rmd
fix env partitioning, remove deprecated `crs` arg

@@
 autoplot(resampling_buffer,
-  size = 0.8, task = task, fold_id = 1, crs = 32717) *
+  size = 0.8, task = task, fold_id = 1) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
@@
 autoplot(resampling_disc,
-  size = 0.8, task = task, fold_id = 1, crs = 32717) *
+  size = 0.8, task = task, fold_id = 1) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
@@
 autoplot(resampling_block_random,
-  size = 0.8, fold_id = 1, task = task, crs = 32717,
+  size = 0.8, fold_id = 1, task = task,
   show_blocks = TRUE, show_labels = TRUE) *
@@
 autoplot(resampling_block_systematic,
-  size = 0.8, fold_id = 1, task = task, crs = 32717,
+  size = 0.8, fold_id = 1, task = task,
   show_blocks = TRUE, show_labels = TRUE) *
@@
 autoplot(resampling_block_checkerboard,
-  size = 0.8, fold_id = 1, task = task, crs = 32717,
+  size = 0.8, fold_id = 1, task = task,
   show_blocks = TRUE, show_labels = TRUE) *
@@
 autoplot(resampling_block_rows_cols,
-  size = 0.8, fold_id = 1, task = task, crs = 32717,
+  size = 0.8, fold_id = 1, task = task,
   show_blocks = TRUE, show_labels = TRUE) *
@@
 autoplot(resampling_tiles,
-  size = 0.8, fold_id = 1, task = task, crs = 32717) *
+  size = 0.8, fold_id = 1, task = task) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
@@
 autoplot(resampling_coords,
-  size = 0.8, fold_id = 1, task = task, crs = 32717) *
+  size = 0.8, fold_id = 1, task = task) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
@@
 autoplot(resampling_cstf_time,
-  fold_id = 5, task = task_spt, crs = 32717, plot3D = TRUE
+  fold_id = 5, task = task_spt, plot3D = TRUE
 )
@@
 autoplot(resampling_cstf_loc,
-  fold_id = 5, task = task_spt, crs = 32717, plot3D = TRUE)
+  fold_id = 5, task = task_spt, plot3D = TRUE)
 ```
@@
 autoplot(resampling_cstf_time_loc,
-  fold_id = 4, task = task_spt, crs = 32717, plot3D = TRUE,
+  fold_id = 4, task = task_spt, plot3D = TRUE,
   show_omitted = TRUE)
@@
 Labelled as "Custom CV", users are able to supply an external factor vector with the same length as number of observations.
-Alternatively to supplying an external factor vector, the name of a feature within the task can be used.
+Each level of the factor vector will result in one partition.
+Alternatively to supplying an external factor vector, the name of a feature (type `character`) present within the task can be supplied via argument `col`.
 
-The following example was taken from `sperrorest::partition_factor()`.
+The following example was taken from `sperrorest::partition_factor()` because the resulting partitions separate nicely from each other in a visual way.
 
-FIXME: eval=TRUE when available upstream im mlr3
-FIXME: add visualization when method is available
-
-```{r custom-cv,  fig.cap="Custom resampling using an external factor vector.", out.width="40%", fig.pos="ht", eval=FALSE}
+```{r custom-cv, fig.cap="Custom resampling using an external factor vector defining the partitions. Each coloring represents one partition.", out.width="40%", fig.pos="ht"}
 breaks = quantile(task$data()$dem, seq(0, 1, length = 6))
@@
 
-autoplot(resampling_custom,
-  size = 0.8, fold_id = 1, task = task) *
+autoplot(resampling_custom, size = 0.8, task = task) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
@@
 autoplot(resampling_env,
-  size = 0.8, fold_id = 1, task = task, crs = 32717,
-  show_blocks = TRUE, show_labels = TRUE) *
+  size = 0.8, fold_id = 1, task = task) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
@@
 autoplot(resampling_env_multi,
-  size = 0.8, fold_id = 1, task = task, crs = 32717,
-  show_blocks = TRUE, show_labels = TRUE) *
+  size = 0.8, fold_id = 1, task = task) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
@@
 
-autoplot(resampling_sp, task, fold_id = c(1:2), crs = 32717, size = 0.8) *
+autoplot(resampling_sp, task, fold_id = c(1:2), size = 0.8) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
@@
 
-autoplot(resampling_nsp, task, fold_id = c(1:2), crs = 32717, size = 0.8) *
+autoplot(resampling_nsp, task, fold_id = c(1:2), size = 0.8) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *

### 2021-07-05 paper/JMNotes.Rmd
put down first remarks

# remarks JM

- abstract: it is at least uncommon to cite other works in the abstract.
If not explicitly required by the journal, I would refrain from it since you most likely cite all these works in the paper itself.
- The last sentence of the abstract does not add much.
Delete it or rewrite in a more compelling way.

- I like the crop classification example, however, it is not entirely clear to me what agricultural districs are. 
And though the crop classification indroduces non-standard spatial resampling strategies, it does not make reference to the more common spatial resampling procedure (k-means clustering) where only the coordinates are considered.
However,

- Valid inputs are PROJ4 strings
Just for your information: PROJ strings are from now on discouraged if a CRS is available.
Instead one should use CRS codes or WKT2 (see most recent comments at https://github.com/r-spatial/discuss/issues/43).
In any case, please us the more generic PROJ notation without any version number instead of PROJ4. 
For example, on my laptop I now use PROJ8.

### 2021-07-05 paper/paper.Rmd
attach packages in chunk buffer to make pdf compilation work

@@
 ```{r buffer, fig.cap="Visualization of the 'Spatial buffering' method from package `blockCV`.", out.width="40%", fig.pos="ht"}
+library("mlr3")
+library("mlr3spatiotempcv")
 task = tsk("ecuador")

### 2021-07-05 paper/JMNotes.Rmd
mention s2 spherical geometries

@@
 
+>>It is planned to provide a more consistent interface that supports all CRS and allows the use of distance measures in projected planar space or using great circle distances on the unprojected ellipsoid, in the case of unprojected CRS.
+
+- In case, you didn't know: from 1.0.0 onwards, sf uses s2 by default, i.e. spherical geometries.
+Hence, I am not sure if there is really much more work to do (see https://r-spatial.github.io/sf/articles/sf7.html).
+
 - Valid inputs are PROJ4 strings

### 2021-07-06 paper/paper.Rmd
slightly rephrase
refer to section numbers to stay consistent with the rest of the paper

@@
 
-In traditional cross-validation, the partitioning is based on uniform random sampling, which ignores spatial or temporal autocorrelation or any existing grouping structure as well as the structure of the prediction task, as outlined in the Introduction, and may result in over-optimistic performance estimates.
+In traditional cross-validation, the partitioning is based on uniform random sampling, which ignores spatial or temporal autocorrelation or any existing grouping structure as well as the structure of the prediction task, as outlined in section \@ref(introduction), and may result in over-optimistic performance estimates.
 Several approaches have therefore been proposed in the literature and implemented in software to accommodate a variety of predictive situations (Table \@ref(tab:sptcv-methods)).
@@
 Spatial and spatiotemporal partitioning techniques in \pkg{mlr3spatiotempcv} are currently mostly imported and interfaced from other packages, in particular \pkg{sperrorest}, \pkg{blockCV} and \pkg{CAST} [@blockCV; @brenning2012; @cast], in order to expose them to \pkg{mlr3} functionality.
-To reduce first-level dependencies, some methods were re-implemented instead of imported from upstream packages.
+To reduce first-level dependencies, some methods were re-implemented instead of importing them from upstream packages.
 

### 2021-07-06 paper/JMNotes.Rmd
add table 1 observations

@@
 
-- I like the crop classification example, however, it is not entirely clear to me what agricultural districs are. 
-And though the crop classification indroduces non-standard spatial resampling strategies, it does not make reference to the more common spatial resampling procedure (k-means clustering) where only the coordinates are considered.
-However,
+- I like the crop classification example, however, it is not entirely clear to me what agricultural districts are. 
+And though the crop classification introduces non-standard spatial resampling strategies, it does not make reference to the more common spatial resampling procedure (k-means clustering) where clustered coordinates are considered, which might be a more gentle introduction to spatial CV for readers who are not already familiar with the concepts.
+But I might be wrong.
+Looking at Table 1, the spatial blocking/tiles approach seems to be even more popular than the kmean clustering approach.
+
+- Table 1: Wow, not too many papers have so far used spatial CV.
+Is this a comprehensive list?
+Have you somewhere listed the references by chosen strategy?
+If so, put it in the Appendix and reference the Appendix in the caption.
+Not entirely clear, what the footnotes refer to.
+These seem to be studies which have used the corresponding spatial CV approach but it is not all of them (e.g., spcv_block has been used by 28 papers but only 3 are referenced). 
+Please clarify.
 

### 2021-07-06 paper/paper.Rmd
slightly rephrase

@@
 
-Judging by their common characteristics and the available implementations, spatial partitioning methods can be coarsely categorized into (1) buffering, (2) blocking and (3) clustering methods.
-If a method had the potential to belong to multiple categories, a deliberate category assignment was made for one specific group which matched the main characteristic of the respective method best.
+Spatial partitioning methods can be roughly divided into three groups: (1) buffering, (2) blocking and (3) clustering methods (Table \@ref(tab:sptcv-methods)).
+If a method could have been placed into multiple categories, we deliberately assigned the group which matched best the main characteristic of the respective method.
 

### 2021-07-06 paper/paper.Rmd
ask for clarification of scattered appearance
request to clearly state that buffering could have also been put into the spatiotemporal category

@@
 
-Buffering is mostly applied in the coordinate space due to the scattered appearance of observations in this space.
-However, there is no practical limitation that would prevent this method to be applied in a 2D + time setting.
-Because \pkg{mlr3spatiotempcv} supports two spatial-only "buffering" implementations, (1) "Spatial buffering" from package \pkg{blockCV} and (2) "Leave-one-disc-out cross-validation" from \pkg{sperrorest}, buffering was attributed to the category of spatial partitioning methods.
+<!--JM: not entirely sure what scattered appearance refers to-->
+Buffering is most often applied in coordinate space due to the scattered appearance of observations in this space.
+However, this method could be easily extended to a 2D + time setting.
+<!-- JM: I guess you mean that you could have placed buffering also into spatiotemporal partition. I would advise to say so instead of letting the reader guess.-->
+Since \pkg{mlr3spatiotempcv} supports two spatial-only "buffering" implementations, (1) "Spatial buffering" from package \pkg{blockCV} and (2) "Leave-one-disc-out cross-validation" from \pkg{sperrorest}, buffering was attributed to the category of spatial partitioning methods.
 

### 2021-07-06 paper/paper.Rmd
reference table 1 again

@@
 <!-- JM: I guess you mean that you could have placed buffering also into spatiotemporal partition. I would advise to say so instead of letting the reader guess.-->
-Since \pkg{mlr3spatiotempcv} supports two spatial-only "buffering" implementations, (1) "Spatial buffering" from package \pkg{blockCV} and (2) "Leave-one-disc-out cross-validation" from \pkg{sperrorest}, buffering was attributed to the category of spatial partitioning methods.
+Since \pkg{mlr3spatiotempcv} supports two spatial-only "buffering" implementations, (1) "Spatial buffering" from package \pkg{blockCV} and (2) "Leave-one-disc-out cross-validation" from \pkg{sperrorest}, buffering was attributed to the category of spatial partitioning methods (see Table \@ref(tab:sptcv-methods)).
 

### 2021-07-06 paper/JMNotes.Rmd
praise the visualizations
musing about leave-one-disc-out

@@
 But I might be wrong.
-Looking at Table 1, the spatial blocking/tiles approach seems to be even more popular than the kmean clustering approach.
+Looking at Table 1, the spatial blocking/tiles approach seems to be even more popular than the kmeans clustering approach.
 
@@
 
->>It is planned to provide a more consistent interface that supports all CRS and allows the use of distance measures in projected planar space or using great circle distances on the unprojected ellipsoid, in the case of unprojected CRS.
+- I really do like the visualizations, great job! And they help the reader a lot to better grasp the differences between the resampling strategies.
 
@@
 Hence, I am not sure if there is really much more work to do (see https://r-spatial.github.io/sf/articles/sf7.html).
+I am referring to this sentence in the paper:
+
+>It is planned to provide a more consistent interface that supports all CRS and allows the use of distance measures in projected planar space or using great circle distances on the unprojected ellipsoid, in the case of unprojected CRS.
+
+- The leave-one-disc-out method seems to be quite promising since the direct neighbors between the training and test sets are removed thus removing spatial autocorrelation which might still be present when using the kmeans clustering approach.
+In the back of my mind, I knew that this was possible but I didn't know that this was already implemented in **sperrorest**.
+I wonder why AB has often preferred the kmeans clustering approach over the leave-one-disc-out approach.
+Maybe something which can be discussed.
 

### 2021-07-06 paper/paper.Rmd
rephrase spatial blocking intro

@@
 
-The "Spatial blocking" from package \pkg{blockCV} comes with various options to customize the zone creation.
-Besides the choice of relying on an initial random  (default) (`selection = "random"`), systematic (`selection = "systematic"`) or checkerboard pattern (`selection = "checkerboard"`) for the final partition aggregation, support for SDM specific background data is built-in.
-Supplying a numeric value in meters via argument `theRange` (named just `range` in {mlr3spatiotempcv}) will create a quadratic block pattern across the study area.
-Ultimately, users can also supply a user-defined polygon which defines the zones.
+The "Spatial blocking" from package \pkg{blockCV} comes with various blocking options such as random (default) (`selection = "random"`), systematic (`selection = "systematic"`) or checkerboard pattern (`selection = "checkerboard"`).
+However, instead of using these options, the user might also supply a numeric value in meters via argument `theRange` (named just `range` in {mlr3spatiotempcv}) which will create a quadratic block pattern across the study area.
+<!--JM: to stay consistent please add the argument which lets the user indicate a user-defined polygon-->
+This kind of blocking might be of special interest for SDM studies as well as the possibility to supply a user-defined polygon with predefined blocking zones.
 
-After the initial block creation, multiple "blocks" are combined to form a partition, honoring the number of folds requested by the user.
+After the initial block creation, multiple "blocks" are combined to form a partition in correspondence with the number of folds requested by the user.
 For example, in a five-fold scenario with a grid of four columns and five rows, the method would combine one fifth of all blocks (= four blocks) into a single partition.

### 2021-07-06 paper/paper.Rmd
change sentence so it becomes clear that range is an additional argument

@@
 The "Spatial blocking" from package \pkg{blockCV} comes with various blocking options such as random (default) (`selection = "random"`), systematic (`selection = "systematic"`) or checkerboard pattern (`selection = "checkerboard"`).
-However, instead of using these options, the user might also supply a numeric value in meters via argument `theRange` (named just `range` in {mlr3spatiotempcv}) which will create a quadratic block pattern across the study area.
+Additionally, the user might supply a numeric value in meters via argument `theRange` (named just `range` in {mlr3spatiotempcv}) which will ensure a quadratic block pattern across the study area.
 <!--JM: to stay consistent please add the argument which lets the user indicate a user-defined polygon-->

### 2021-07-06 paper/paper.Rmd
expand comment

@@
 Additionally, the user might supply a numeric value in meters via argument `theRange` (named just `range` in {mlr3spatiotempcv}) which will ensure a quadratic block pattern across the study area.
-<!--JM: to stay consistent please add the argument which lets the user indicate a user-defined polygon-->
+<!--JM: to stay consistent please add the argument which lets the user indicate a user-defined polygon or are you referring to the rectangular block design specified with rows and cols?
+If so, say so.
+If not, clarify and additionally mention the rectangular block design here-->
 This kind of blocking might be of special interest for SDM studies as well as the possibility to supply a user-defined polygon with predefined blocking zones.

### 2021-07-06 paper/paper.Rmd
change order of words

@@
 
-The idea of cluster analysis, which originated in the 1930s in the anthropology field [@driver1932, @zubin1938], is to group objects with similar characteristics.
+The idea of cluster analysis, which originated in the 1930s in the field of anthropology [@driver1932, @zubin1938], is to group objects with similar characteristics.
 In the spatial context, the similarity of spatial coordinates is used to form such groups.

### 2021-07-06 paper/paper.Rmd
slightly rephrase kmeans intro

@@
 
-The so-called "Spatial CV" method after [@russ2010; @brenning2012] and implemented in package \pkg{sperrorest} uses the coordinates of all observations to create clusters in the spatial domain.
-$k$-means clustering, which belongs to the centroid-based clustering family, has been proposed as a clustering method.
+The so-called "Spatial CV" method after [@russ2010; @brenning2012] and implemented in package \pkg{sperrorest} uses the coordinates of all observations to create clusters in the spatial domain with the help of the $k$-means clustering algorithm.^[$k$-means clustering belongs to the centroid-based clustering family.]
 To avoid confusion with other spatial partitioning methods, we refer to this as coordinate-based clustering, or as `spcv_coords` in \pkg{mlr3spatiotempcv}, instead of using the generic "Spatial CV" term.

### 2021-07-06 paper/paper.Rmd
delete superfluous the

@@
 Group-level CV follows the principle of an inherit grouping structure of the observations, both in space and/or time, to determine (groups of) partitions.
-In contrast to blocking or clustering not the the spatial or temporal location is used but additional metadata such as the membership to a specific crop type [@brenning2006].
+In contrast to blocking or clustering not the spatial or temporal location is used but additional metadata such as the membership to a specific crop type [@brenning2006].
 The idea to follow an inherited grouping structure of the data is also referred to as "target-oriented validation" [@cast].

### 2021-07-06 paper/paper.Rmd
delete superfluous to
add notes on CAST

@@
 Leave-Location-and-Time-Out (LLTO) from package \pkg{CAST} is a spatiotemporal resampling method [@meyer2018].
-It requires both a spatial and a temporal variable which are used to to separate observations in a multi-dimensional space.
+It requires both a spatial and a temporal variable which are used to separate observations in a multi-dimensional space.
 It comes with satellite methods named "Leave-time-out" (LTO) and "Leave-location-out" (LLO) which only operate in a two-dimensional space.
@@
 
-The following code uses the `cookfarm` dataset as the base, modifies to the `Date` variable to contain five unique levels and then creates a spatiotemporal regression task in \pkg{mlr3spatiotempcv} (Fig.
+The following code uses the `cookfarm` dataset as the base, modifies the `Date` variable in such a way that it contains five unique levels and then creates a spatiotemporal regression task in \pkg{mlr3spatiotempcv} (Fig.
 \@ref(fig:lto)).
@@
 The grey points show the points omitted by this method, i.e.
-the points which are not used in the training or test set in this fold.", fig.pos="ht"}
+the points which are neither used in the training nor test set in this fold.", fig.pos="ht"}
 knitr::include_graphics("llto.png")

### 2021-07-06 paper/JMNotes.Rmd
add note on Figure 7
add notes on CAST

@@
 
+- Figure 7: spatial tiling. 
+It would be great if one could see the used rectangular blocks.
+- Figure 9: in fact, one only sees 4 time levels, not 5 as stated in the caption
+- LLTO: maybe you could expand a little bit on why observations need to be omitted.
+For readers new to the LLTO concept, the reasoning for this might not be entirely clear.
+
 - Valid inputs are PROJ4 strings

### 2021-07-06 paper/paper.Rmd
recommend to delete FIXME
add notes on features

@@
 
+<!--JM: you might want to delete the FIXME-->
 Support for user-defined partitioning strategies is built into \pkg{mlr3} directly since version 0.12.0 FIXME.
@@
 The last method from the \pkg{blockCV} package, "environmental blocking", makes use of *k-means* clustering [@hartigan1979a] in a possibly multivariate space.
-The user can select a variable via argument `feature` specifying which levels will be used to create clusters.
+<!--JM: pls rewrite next sentence. Do not use levels when you later on say the feature has to be numeric-->
+The user can select a variable via argument `features` which specifies which levels will be used to create clusters.
 Hereby, k-means will use Euclidean distance for cluster creation.

### 2021-07-12 paper/paper.Rmd
ask why feature clustering is not supported
remove references from abstract

--- 2021-07-06 paper/paper.Rmd
@@
 abstract: >
-  The R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is an extension package of the machine learning framework [mlr3](https://mlr3.mlr-org.com) [@mlr3].
+  The R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is an extension package of the machine learning framework [mlr3](https://mlr3.mlr-org.com).
   It adds support for spatiotemporal resampling methods and respective visualization methods.
@@
   These autocorrelation types cause biased performance estimates of varying magnitude in cross-validation when ignored.
-  At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: CAST [@cast], blockCV [@blockCV] or sperrorest [@brenning2012].
-  There are some resampling methods used in scientific studies for which the source code is either not available or a different programming language has been used [@roberts2017; @pohjankukka2017].
+  At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: CAST, blockCV or sperrorest.
+  There are some resampling methods used in scientific studies for which the source code is either not available or a different programming language has been used.
 
@@
 
-In traditional cross-validation, the partitioning is based on uniform random sampling, which ignores spatial or temporal autocorrelation or any existing grouping structure as well as the structure of the prediction task, as outlined in section \@ref(introduction), and may result in over-optimistic performance estimates.
+In traditional cross-validation, the partitioning is based on uniform random sampling, which ignores spatial or temporal autocorrelation or any existing grouping structure as well as the structure of the prediction task, as outlined in the Introduction, and may result in over-optimistic performance estimates.
 Several approaches have therefore been proposed in the literature and implemented in software to accommodate a variety of predictive situations (Table \@ref(tab:sptcv-methods)).
@@
 Spatial and spatiotemporal partitioning techniques in \pkg{mlr3spatiotempcv} are currently mostly imported and interfaced from other packages, in particular \pkg{sperrorest}, \pkg{blockCV} and \pkg{CAST} [@blockCV; @brenning2012; @cast], in order to expose them to \pkg{mlr3} functionality.
-To reduce first-level dependencies, some methods were re-implemented instead of importing them from upstream packages.
+To reduce first-level dependencies, some methods were re-implemented instead of imported from upstream packages.
 
@@
 
-Spatial partitioning methods can be roughly divided into three groups: (1) buffering, (2) blocking and (3) clustering methods (Table \@ref(tab:sptcv-methods)).
-If a method could have been placed into multiple categories, we deliberately assigned the group which matched best the main characteristic of the respective method.
+Judging by their common characteristics and the available implementations, spatial partitioning methods can be coarsely categorized into (1) buffering, (2) blocking and (3) clustering methods.
+If a method had the potential to belong to multiple categories, a deliberate category assignment was made for one specific group which matched the main characteristic of the respective method best.
 
@@
 
-<!--JM: not entirely sure what scattered appearance refers to-->
-Buffering is most often applied in coordinate space due to the scattered appearance of observations in this space.
-However, this method could be easily extended to a 2D + time setting.
-<!-- JM: I guess you mean that you could have placed buffering also into spatiotemporal partition. I would advise to say so instead of letting the reader guess.-->
-Since \pkg{mlr3spatiotempcv} supports two spatial-only "buffering" implementations, (1) "Spatial buffering" from package \pkg{blockCV} and (2) "Leave-one-disc-out cross-validation" from \pkg{sperrorest}, buffering was attributed to the category of spatial partitioning methods (see Table \@ref(tab:sptcv-methods)).
+Buffering is mostly applied in the coordinate space due to the scattered appearance of observations in this space.
+However, there is no practical limitation that would prevent this method to be applied in a 2D + time setting.
+Because \pkg{mlr3spatiotempcv} supports two spatial-only "buffering" implementations, (1) "Spatial buffering" from package \pkg{blockCV} and (2) "Leave-one-disc-out cross-validation" from \pkg{sperrorest}, buffering was attributed to the category of spatial partitioning methods.
 
@@
 ```{r buffer, fig.cap="Visualization of the 'Spatial buffering' method from package `blockCV`.", out.width="40%", fig.pos="ht"}
-library("mlr3")
-library("mlr3spatiotempcv")
 task = tsk("ecuador")
@@
 
-The "Spatial blocking" from package \pkg{blockCV} comes with various blocking options such as random (default) (`selection = "random"`), systematic (`selection = "systematic"`) or checkerboard pattern (`selection = "checkerboard"`).
-Additionally, the user might supply a numeric value in meters via argument `theRange` (named just `range` in {mlr3spatiotempcv}) which will ensure a quadratic block pattern across the study area.
-<!--JM: to stay consistent please add the argument which lets the user indicate a user-defined polygon or are you referring to the rectangular block design specified with rows and cols?
-If so, say so.
-If not, clarify and additionally mention the rectangular block design here-->
-This kind of blocking might be of special interest for SDM studies as well as the possibility to supply a user-defined polygon with predefined blocking zones.
+The "Spatial blocking" from package \pkg{blockCV} comes with various options to customize the zone creation.
+Besides the choice of relying on an initial random  (default) (`selection = "random"`), systematic (`selection = "systematic"`) or checkerboard pattern (`selection = "checkerboard"`) for the final partition aggregation, support for SDM specific background data is built-in.
+Supplying a numeric value in meters via argument `theRange` (named just `range` in {mlr3spatiotempcv}) will create a quadratic block pattern across the study area.
+Ultimately, users can also supply a user-defined polygon which defines the zones.
 
-After the initial block creation, multiple "blocks" are combined to form a partition in correspondence with the number of folds requested by the user.
+After the initial block creation, multiple "blocks" are combined to form a partition, honoring the number of folds requested by the user.
 For example, in a five-fold scenario with a grid of four columns and five rows, the method would combine one fifth of all blocks (= four blocks) into a single partition.
@@
 
-The idea of cluster analysis, which originated in the 1930s in the field of anthropology [@driver1932, @zubin1938], is to group objects with similar characteristics.
+The idea of cluster analysis, which originated in the 1930s in the anthropology field [@driver1932, @zubin1938], is to group objects with similar characteristics.
 In the spatial context, the similarity of spatial coordinates is used to form such groups.
@@
 
-The so-called "Spatial CV" method after [@russ2010; @brenning2012] and implemented in package \pkg{sperrorest} uses the coordinates of all observations to create clusters in the spatial domain with the help of the $k$-means clustering algorithm.^[$k$-means clustering belongs to the centroid-based clustering family.]
+The so-called "Spatial CV" method after [@russ2010; @brenning2012] and implemented in package \pkg{sperrorest} uses the coordinates of all observations to create clusters in the spatial domain.
+$k$-means clustering, which belongs to the centroid-based clustering family, has been proposed as a clustering method.
 To avoid confusion with other spatial partitioning methods, we refer to this as coordinate-based clustering, or as `spcv_coords` in \pkg{mlr3spatiotempcv}, instead of using the generic "Spatial CV" term.
@@
 Group-level CV follows the principle of an inherit grouping structure of the observations, both in space and/or time, to determine (groups of) partitions.
-In contrast to blocking or clustering not the spatial or temporal location is used but additional metadata such as the membership to a specific crop type [@brenning2006].
+In contrast to blocking or clustering not the the spatial or temporal location is used but additional metadata such as the membership to a specific crop type [@brenning2006].
 The idea to follow an inherited grouping structure of the data is also referred to as "target-oriented validation" [@cast].

### 2021-07-12 paper/paper.Rmd
rewrite end of abstract

@@
 
-  The goal of [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is to gather the available spatiotemporal resampling methods in R and make them available to users through a simplified and generic interface.
-  The latter is made possible by integrating the package directly into the [mlr3](https://mlr3.mlr-org.com) machine learning framework.
-  This simplifies the step of integrating package specific syntax into an overarching machine learning pipeline.
-  The hope is that the availability of such a package encourages researchers to more often apply spatiotemporal resampling methods in scientific case studies.
+  The goal of [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is to gather the available spatiotemporal resampling methods in R and make them available to users through a simple and generic interface.
+  The latter is made possible by integrating the package directly into the [mlr3](https://mlr3.mlr-org.com) machine learning framework which already has support for generic non-spatiotemporal resampling methods such as random partitioning.
+  This simplifies the step of integrating package specific syntax into an overarching machine learning pipeline, making it easier for users to apply a variety of different spatiotemporal resampling methods.
 keywords:

### 2021-07-12 paper/paper.Rmd
fix figure caption
re-explain why obs need to be omitted in LLTO

@@
 
-```{r tile, fig.cap="Spatial tiling with arguments `rows` and `cols`.", out.width="40%", fig.pos="ht"}
+```{r tile, fig.cap="Spatial tiling", out.width="40%", fig.pos="ht"}
 resampling_tiles = rsmp("spcv_tiles", nsplit = c(3L, 4L))
@@
 It creates partitions that operate on spatiotemporal clusters to reduce both spatial and temporal autocorrelation.
-To achieve this, observations are omitted and only a subset is used.
-In the special scenario with five temporal levels and five folds, the train/test split happens only at one temporal level (Fig. \@ref(fig:llto)).
+To achieve this, observations need to be omitted as the split in test and train sets only provides one dimension to separate observations whereas two (space and time) are needed.
+In other words, accounting for one dimension as in Fig.
+\@ref(fig:lto) or Fig.
+\@ref(fig:llo), by only having one time step or location in either test or training set, would have no effect on the respective other.
+To achieve an effect for both, observations need to be omitted as shown in Fig. \@ref(fig:llto).
 
@@
   By default coordinates are not used as predictors.
-- `crs`: The coordinate reference system of the data. Valid inputs are PROJ4 strings or EPSG codes in the format `ESPG:<code>`.
+- `crs`: The coordinate reference system of the data. Valid inputs are PROJ strings or EPSG codes in the format `ESPG:<code>`.
 

### 2021-07-12 paper/paper.Rmd
optimize feature space clustering intro
clarify mlr3spatiotempcv capabilities

@@
 The last method from the \pkg{blockCV} package, "environmental blocking", makes use of *k-means* clustering [@hartigan1979a] in a possibly multivariate space.
-The user can select a variable via argument `feature` specifying which levels will be used to create clusters.
+The user can select a variable via argument `feature` which denotes the variable that will be used to create clusters.
 Hereby, k-means will use Euclidean distance for cluster creation.
@@
 CLUTO is available in R through the \pkg{skmeans} package, which provides an interface to a downloadable compiled library with a restriction to non-commercial uses (see `?ResamplingSptCVCluto` in \pkg{mlr3spatiotempcv} for more information).
-With respect to functionality, \pkg{mlr3spatiotempcv} supports two-dimensional clustering based on coordinates and three-dimensional clustering on coordinates and time.
-Feature clustering is not supported.
+With respect to functionality, \pkg{mlr3spatiotempcv} supports two-dimensional clustering based on coordinates, time or features and three-dimensional clustering on coordinates and time.
 

### 2021-07-16 paper/paper.Rmd
explain reference count in methods table

@@
   \caption[t]{Spatiotemporal resampling methods, ordered alphabetically by class name (column 'mlr3 Notation').
-Column count refers to the number of studies which were found to have used this resampling technique until 2021-05.}
+    Column count refers to the number of studies which were found to have used this resampling technique until 2021-05.
+    For each method, up to three randomly chosen references were added to the table as footnotes.}
   \begingroup

### 2021-07-16 paper/paper.Rmd
remove CRS handling promises
discuss irregularly shaped study areas

@@
 Support for projected (planar) and unprojected (geographic) coordinate reference systems (CRS) currently varies depending on the partitioning techniques used, since these inherit their behavior from the underlying upstream R package.
-It is planned to provide a more consistent interface that supports all CRS and allows the use of distance measures in projected planar space or using great circle distances on the unprojected ellipsoid, in the case of unprojected CRS.
 
@@
 According to the package authors, this estimate should then be used for argument `theRange` to have a sensible value for the block sizes created in method `"spcv_block"`.
+Rectangular partitioning can be problematic for irregularly shaped study areas as it might result in certain partitions containing substantially less observations than others.
 
@@
 In general, only methods making use of a random component (e.g. `"spcv_coords"` or `"spcv_block"` with option `selction = "random"`) may be suited for using multiple repetitions.
+On the upside, `"spcv_coords"` works well with irregularly shaped study areas due to the ability to create partitions which adjust well to the given study area shape.
 

### 2021-07-16 paper/paper.Rmd
expand on the downsides of buffer methods

@@
 
-<!-- spcv buffer: discards observations, LOO very expensive -->
-Method `"spcv_buffer"` makes use of the removal of observations to reduce the influence of spatial autocorrelation between train and test set.
-This can be a pretty effective way to reduce autocorrelation but also requires to discard many observations and comes with the downsides of a LOO-CV: many model fits which can be computationally very demanding and the results inheriting a high variance due to the high differences in the test sets (also commonly known as the "bias-variance tradeoff" [@wahba1998; @blockCV].
+<!-- spcv buffer and spcv disc: discards observations, LOO very expensive -->
+Methods `"spcv_buffer"` and `"spcv_disc"` (optionally) make use of the removal of observations to reduce the influence of spatial autocorrelation between train and test set.
+This can be a pretty effective way to reduce autocorrelation but also requires to discard many observations.
+In addition it comes with the downsides of a LOO-CV: many model fits which can be computationally very demanding and the results inheriting a high variance due to the high differences in the test sets (also commonly known as the "bias-variance tradeoff" [@wahba1998; @blockCV].
+If a large buffer size is chosen relatively to the overall study area size, a large portion of it will not be sampled.
+Choosing an appropriate buffer size is not easy and it should be taken care not to select a value based on subjective interpretation of the study area characteristics.
 

### 2021-07-16 paper/paper.Rmd
mention tilt angle in figure 9

@@
 ```{r lto, echo=FALSE, out.width="70%", fig.cap="Three-dimensional visualization of method `sptcv\\_cstf` and sub-method Leave-Time-Out (LTO).
-To support showcasing the temporal grouping, five folds and five temporal levels in variable `Date` were used in this example.", fig.pos="ht"}
+To support showcasing the temporal grouping, five folds and five temporal levels in variable `Date` were used in this example.
+Due to the tilt angle of the static 3D plot the last two levels are hard to distinguish visually but they are separated in the same way as the the first two levels on the top of the plot.", fig.pos="ht"}
 knitr::include_graphics("lto.png")

### 2021-07-16 paper/paper.Rmd
rephrase LLTO intro

@@
 It creates partitions that operate on spatiotemporal clusters to reduce both spatial and temporal autocorrelation.
-To achieve this, observations need to be omitted as the split in test and train sets only provides one dimension to separate observations whereas two (space and time) are needed.
+If there are multiple observations over time per location, and the modeling purpose is to reduce the bias in the predictions as potentially introduced by spatial as well as temporal autocorrelation, LLTO might be an appropriate choice.
+To reduce autocorrelation in both dimensions, the methods omits observations as the splitting in test and train sets only provides separation in one dimension whereas two (space and time) are needed to account for both types, spatial and temporal autocorrelation.
 In other words, accounting for one dimension as in Fig.
@@
 \@ref(fig:llo), by only having one time step or location in either test or training set, would have no effect on the respective other.
-To achieve an effect for both, observations need to be omitted as shown in Fig. \@ref(fig:llto).
+Thus, the removal of observation can be seen as a proxied buffering in multiple dimensions as shown in Fig. \@ref(fig:llto).
 

### 2021-07-16 paper/paper.Rmd
simplify buffering intro
explicitily mention `rasterLayer` arg

@@
 
-Buffering is mostly applied in the coordinate space due to the scattered appearance of observations in this space.
+Buffering is mostly applied in the coordinate space.
 However, there is no practical limitation that would prevent this method to be applied in a 2D + time setting.
@@
 Supplying a numeric value in meters via argument `theRange` (named just `range` in {mlr3spatiotempcv}) will create a quadratic block pattern across the study area.
-Ultimately, users can also supply a user-defined polygon which defines the zones.
+Ultimately, users can also supply a user-defined polygon via argument `rasterLayer` to define the zones.
 

### 2021-07-16 paper/paper.Rmd

@@
 However, there is no practical limitation that would prevent this method to be applied in a 2D + time setting.
-Because \pkg{mlr3spatiotempcv} supports two spatial-only "buffering" implementations, (1) "Spatial buffering" from package \pkg{blockCV} and (2) "Leave-one-disc-out cross-validation" from \pkg{sperrorest}, buffering was attributed to the category of spatial partitioning methods.
+Since \pkg{mlr3spatiotempcv} supports two spatial-only "buffering" implementations, (1) "Spatial buffering" from package \pkg{blockCV} and (2) "Leave-one-disc-out cross-validation" from \pkg{sperrorest}, buffering was attributed to the category of spatial partitioning methods (see Table \@ref(tab:sptcv-methods)).
 
@@
 
-The "Spatial blocking" from package \pkg{blockCV} comes with various options to customize the zone creation.
-Besides the choice of relying on an initial random  (default) (`selection = "random"`), systematic (`selection = "systematic"`) or checkerboard pattern (`selection = "checkerboard"`) for the final partition aggregation, support for SDM specific background data is built-in.
-Supplying a numeric value in meters via argument `theRange` (named just `range` in {mlr3spatiotempcv}) will create a quadratic block pattern across the study area.
-Ultimately, users can also supply a user-defined polygon via argument `rasterLayer` to define the zones.
+The "Spatial blocking" from package \pkg{blockCV} comes with various blocking options such as random (default) (`selection = "random"`), systematic (`selection = "systematic"`) or checkerboard pattern (`selection = "checkerboard"`).
+Additionally, the user might supply a numeric value in meters via argument `theRange` (named just `range` in {mlr3spatiotempcv}) which will ensure a quadratic block pattern across the study area.
+This kind of blocking might be of special interest for SDM studies as well as the possibility to supply a user-defined polygon via argument `rasterLayer` with predefined blocking zones.
 
-After the initial block creation, multiple "blocks" are combined to form a partition, honoring the number of folds requested by the user.
+After the initial block creation, multiple "blocks" are combined to form a partition in correspondence with the number of folds requested by the user.
 For example, in a five-fold scenario with a grid of four columns and five rows, the method would combine one fifth of all blocks (= four blocks) into a single partition.

### 2021-07-20 paper/paper.Rmd
emphasize the need to check for the prediction purpose of the model throughout the paper

@@
   These autocorrelation types cause biased performance estimates of varying magnitude in cross-validation when ignored.
-  At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: CAST, blockCV or sperrorest.
-  There are some resampling methods used in scientific studies for which the source code is either not available or a different programming language has been used.
+  At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: CAST, blockCV and sperrorest.
 
@@
   This simplifies the step of integrating package specific syntax into an overarching machine learning pipeline, making it easier for users to apply a variety of different spatiotemporal resampling methods.
+  The mlr3spatiotempcv package refrains from making suggestions which resampling methods users should apply as this decision mainly depends on the prediction purpose of the fitted model.
+  No best method exists; only valid methods for the modeling purpose at hand.
 keywords:
@@
 In crop classification from remotely-sensed data, for instance, learning samples routinely contain multiple grid cells from a sample of fields with known crop type, for instance 2000 grid cells from 100 fields scattered across a large study region.
-The purpose of training a model on this sample is to make predictions on other, new fields within the same geographic domain (*intra-domain* prediction; @brenning2005) --- not *within* the same field, which obviously presents only one crop type.
-In this situation it would therefore seem rather unwise to train a model on a simple random subsample of grid cells, and test it on the remaining data, i.e.
+The purpose of training a model on this particular sample is to make predictions on other, new fields within the same geographic domain (*intra-domain* prediction; @brenning2005) --- not *within* the same field, which obviously presents only one crop type.
+In this specific situation it would therefore seem rather unwise to train a model on a simple random subsample of grid cells, and test it on the remaining data, i.e.
 using other grid cells from the same fields, as if we wanted to predict within a field.
@@
 
-While it is not the purpose of this article to recommend specific resampling schemes for specific use cases, this example may suffice to motivate the use of spatial and spatiotemporal cross-validation techniques, and the need for a unified framework and computational toolbox that accommodate a variety of prediction tasks that may be applicable to a broad range of application scenarios.
+While it is not the purpose of this article to recommend specific resampling schemes for specific use cases, this example may suffice to motivate the use of appropriate spatial and spatiotemporal cross-validation techniques, and the need for a unified framework and computational toolbox that accommodate a variety of prediction tasks that may be applicable to a broad range of application scenarios.
 \pkg{mlr3spatiotempcv} is such a toolbox.
@@
 This approach has been applied, for example, in spatial CV at the agricultural field level [@pena2015].
-In the following we will conceptually distinguish between CV *at the group (or block) level*, referring to the latter scenario, from spatial *block (or grouped)* CV, where the blocks or groups themselves define the CV partitions.
+We would like to emphasize the conceptual distinction between CV *at the group (or block) level*, referring to the latter scenario, from *custom (or grouped)* CV, where the blocks or groups themselves define the CV partitions.
 
@@
 Even though this method is frequently used, it was mainly chosen because of it's clear visual grouping differences when comparing spatial to random partitioning.
+Another advantage is its ability to adapt to the irregularly shaped study area of this example.
+Last, there are no obvious spatial groupings to our knowledge which need to be accounted for, also with respect to prediction purposes of the fitted model.
 
-In fact, most often multiple spatial partitioning methods can be used for a dataset.
-It is recommended (required) that one familiarizes themselves with multiple methods and decide which method to choose based on the specific characteristics of the dataset at hand.
+Most often multiple spatial partitioning methods can be used for a dataset.
+It is recommended (required) that one familiarizes themselves with multiple methods and decide which method to choose based on the specific characteristics of the dataset at hand and the prediction purpose of the fitted model, i.e.
+for which areas/zones a model should make predictions for.
 There are various variables influencing the partitioning method choice for an environmental dataset:
@@
 - How dense are the observations scattered both in space and time?
-- Are the predictor variables known to be highly affected by spatiotemporal autocorrelation by nature?
+- Are existing predictor variables known to be highly affected by spatiotemporal autocorrelation by nature?
 
-Based on these criteria users must choose a matching partitioning method that is either more restrictive (by discarding observations for fold creation) or more liberal (by using all observations and eventually ignoring natural grouping patterns).
-For almost all methods implemented in \pkg{mlr3spatiotempcv}, there is a scientific publication describing the strengths and weaknesses of the respective approach (either linked in the help file of the package or its respective upstream packages), hence we are refraining from repeating them again in this study.
+Based on these criteria users should choose a matching partitioning method that is either more restrictive (by discarding observations for fold creation) or more liberal (by using all observations and eventually ignoring natural grouping patterns).
+For almost all methods implemented in \pkg{mlr3spatiotempcv}, there is a scientific publication describing the application areas of the respective approach (either linked in the help file of the package or its respective upstream packages), hence we are refraining from repeating them again in this study.
 
-In the example above, a cross-validation without hyperparameter tuning was shown.
+Last, a note on best practices for the example shown in \@ref(sec:case-study):
+In this example a cross-validation without hyperparameter tuning was applied.
 In ML practice most often a nested CV should be desired to optimize model hyperparameter within the CV.
-In such scenarios it is recommended to use the same spatial partitioning method for the inner loop (= tuning level) that was chosen for the outer one.
+In such scenarios it is recommended to use the same spatial partitioning method for the inner loop (= tuning level) that was chosen for the outer one for consistency reasons.
 See @schratz2019 for more details as well as chapter 11 of [Geocomputation with R](https://geocompr.robinlovelace.net/spatial-cv.html) [@lovelace2019].
@@
 
+Even though the prediction purpose should be the first decision criteria for choosing an appropriate spatiotemporal resampling method, we would like to discuss the individual methods in more detail in the following section.
+
 <!-- high autocor on the border of test sets -->
 <!-- sptcv LLTO: discards many observations -->
-All methods which make use of all variables in the dataset suffer from the problem that on the border between train and test sets, both in the spatial and temporal domain, autocorrelation is relatively high (unless the dataset inherits natural gaps between train and test sets).
-Only methods that omit some observations (`"spcv_buffer"`, `"spcv_tiles"`, `"sptcv_cstf"` with both `space_var` and `time_var` set) are able to reduce the impact of this issue.
+All methods which make use of all variables in the dataset suffer from the problem that on the border between train and test sets, both in the spatial and temporal domain, autocorrelation is relatively high (unless the dataset inherits natural groupings or gaps between train and test sets).
+Methods that omit some observations (`"spcv_buffer"`, `"spcv_tiles"`, `"sptcv_cstf"` with both `space_var` and `time_var` set) are able to reduce the impact of autocorrelation in scenarios without a clear grouping or sufficient distance between train and test set observations.
 The downside is the removal of observations from the dataset which possibly reduces the generalization power of the trained model and by this, depending on the specific observations that were removed, adds substantial bias to the CV results.
@@
 While various spatiotemporal resampling methods exists nowadays to account for spatiotemporal autocorrelation in modeling, there is still lots of uncertainty which method should be used for which dataset.
-The answer to this question is not trivial and while there will most likely never be an easy answer to this, this topic needs to be discussed in more detail so that users feel more certain about their actions.
+The answer to this question is not trivial and while there will most likely never be an easy answer to this, users are advised to first think about the prediction purpose of the model and then about reducing autocorrelation between train and test sets to reduce the impact of spatiotemporal autocorrelation.
 One option could be the proposal of a dataset characteristic checklist showing which characteristics make a certain dataset more suitable for specific methods.
-This could then help users to make an appropriate and objective decision which resampling method might be a good fit for their dataset.
+This could then help users to make an appropriate and objective decision which resampling method might be a good fit for their dataset and prospective prediction task.
 
@@
 Spatiotemporal cross-validation is still not yet a paradigm that is completely established in scientific workflows even though it is discussed intensively for more than a decade now.
-Making most available methods easily accessible to users is an important step to foster the acceptance of such in the community and lead to less biased cross-validation results in environmental and ecological studies.
+Making most available methods easily accessible to users is an important step to foster the acceptance of such in the community and to lead to less biased cross-validation results in environmental and ecological studies.
 

### 2021-07-24 paper/paper.Rmd
Rename section custom rsmp to "Custom and grouped resampling"
rephrase a bit

@@
 
-### Group-level resampling
+### Group-level resampling {#sec:group-level}
 
@@
 
-### Custom (fixed) Resampling --- mlr3
+### Custom and grouped resampling --- mlr3
 
@@
 Labelled as "Custom CV", users are able to supply an external factor vector with the same length as number of observations.
+This partitioning concept is also ften referred to as "grouped resampling" and should not be mistaken with "group-level resampling" (section \@ref(sec:group-level)), which inherits additional aggregation of groups at the partition level.
 Each level of the factor vector will result in one partition.
@@
 Most often multiple spatial partitioning methods can be used for a dataset.
-It is recommended (required) that one familiarizes themselves with multiple methods and decide which method to choose based on the specific characteristics of the dataset at hand and the prediction purpose of the fitted model, i.e.
-for which areas/zones a model should make predictions for.
+It is recommended (required) to make oneself familiar with the available methods and to decide on one based on the specific characteristics of the dataset at hand and the prediction purpose of the fitted model.
+The prediction purpose is the model's ability for which areas or zones it actually can make predictions.
+For example, ...
+<!--JM: maybe refer back to the crop example of the intro to make the point better comprehensible than just giving the mere theory-->
+
 There are various variables influencing the partitioning method choice for an environmental dataset:

### 2021-07-24 paper/paper.Rmd
ask for examples

@@
 
-- Do natural groupings exists in the data?
+- Do natural groupings exists in the data? <!--Example-->
 - How dense are the observations scattered both in space and time?
-- Are existing predictor variables known to be highly affected by spatiotemporal autocorrelation by nature?
+The closer observations are in space and/or time, the more similar they are, and hence the greater is the effect of spatio-temporal autocorrelation.
+- Are some of the predictor variables under consideration known to be highly affected by spatiotemporal autocorrelation such as ... <!-- pls add an example -->?
 

### 2021-07-24 paper/paper.Rmd
add some suggestions how to rephrase certain sentences

@@
 Based on these criteria users should choose a matching partitioning method that is either more restrictive (by discarding observations for fold creation) or more liberal (by using all observations and eventually ignoring natural grouping patterns).
+<!--JM: what do you mean by "using all observations"?-->
 For almost all methods implemented in \pkg{mlr3spatiotempcv}, there is a scientific publication describing the application areas of the respective approach (either linked in the help file of the package or its respective upstream packages), hence we are refraining from repeating them again in this study.
@@
 The specification of such is not always easy because the grouping should ensure that the selected groups inherit substantial autocorrelation within themselves and simultaneously differ substantially from other groups.
-Also, if the variable selected contains too many groups, the difference within train/test splits may become undesirably high and tend towards a LOO CV [@meyer2018].
+Also, if the selected variable contains too many groups, the difference within train/test splits may become undesirably high and tend towards a LOO CV [@meyer2018].
 
@@
 In general, only methods making use of a random component (e.g. `"spcv_coords"` or `"spcv_block"` with option `selction = "random"`) may be suited for using multiple repetitions.
-On the upside, `"spcv_coords"` works well with irregularly shaped study areas due to the ability to create partitions which adjust well to the given study area shape.
+On the upside, `"spcv_coords"` works well with irregularly shaped study areas due to its ability to create partitions which adjust well to the shape of the study area.
 
@@
 In addition it comes with the downsides of a LOO-CV: many model fits which can be computationally very demanding and the results inheriting a high variance due to the high differences in the test sets (also commonly known as the "bias-variance tradeoff" [@wahba1998; @blockCV].
-If a large buffer size is chosen relatively to the overall study area size, a large portion of it will not be sampled.
-Choosing an appropriate buffer size is not easy and it should be taken care not to select a value based on subjective interpretation of the study area characteristics.
+If a large buffer size is chosen relative to the size of the study area, a large portion of it will remain unsampled.
+Choosing an appropriate buffer size is a difficult task.
+In any case, one should refrain oneself from selecting a value based on subjective notions with regard to study area characteristics.
 
@@
 Both are similar in that they require external user input for partition creation, i.e. a factor variable which assigns all observations into a distinct group.
-While this the entire idea of "custom resampling" and each group becomes a test set in the CV, "group-level resampling" methods apply additional aggregation on top of these initial groups: for example, users can supply an integer value to an argument named `fold` in the LLO method (e.g.
+The idea of "custom resampling" is that each group becomes a test set in the CV.
+By contrast, "group-level resampling" methods apply additional aggregation on top of these initial groups: for example, users can supply an integer value to an argument named `fold` in the LLO method (e.g.
 `5L`) which combines the initial groups (e.g.
@@
 Over time, studies which have used this method mainly referenced it using the generic "spatial CV" terminology.
-However, using this notation potentially introduces confusion as the term "spatial CV" is a generic one referring to the idea of accounting for spatial autocorrelation in spatial datasets and does not necessarily refer to a specific partitioning method in the first place.
+However, using this notation potentially introduces confusion as the term "spatial CV" is of a generic nature which refers to the idea of accounting for spatial autocorrelation in spatial datasets and does not necessarily refer to a specific partitioning method in the first place.
 Hence, in package \pkg{mlr3spatiotempcv} this method was named `"spcv_coords"`, making use of the fact that the spatial coordinates are used for a cluster-based partition creation.
@@
 There are pros and cons on whether a new method should be integrated into an existing ecosystem such as \pkg{mlr3} directly first or distributed via a standalone package.
-On the upside, developers can benefit from existing package structures and orient themselves how previous methods were implemented.
+On the upside, developers can benefit from existing package structures and orient themselves on how previous methods were implemented.
 Also, the new method will be directly usable by all people using this ecosystem following an existing and familiar syntax.

### 2021-07-26 paper/paper.Rmd
link back to example in intro, add examples

@@
 
-# Introduction
+# Introduction {#sec:intro}
 
@@
 The prediction purpose is the model's ability for which areas or zones it actually can make predictions.
-For example, ...
-<!--JM: maybe refer back to the crop example of the intro to make the point better comprehensible than just giving the mere theory-->
+In the crop example given in section \@ref(sec:intro) this would be to distinguish and predict different crop types. 
+Hence, the CV partitioning is done on the crop level to receive a honest estimate of the model's performance.
 
@@
 
-- Do natural groupings exists in the data? <!--Example-->
+- Do natural groupings exists in the data? Examples include agricultural fields, forest stands, lithology types, etc.
 - How dense are the observations scattered both in space and time?
   The closer observations are in space and/or time, the more similar they are, and hence the greater is the effect of spatio-temporal autocorrelation.
-- Are some of the predictor variables under consideration known to be highly affected by spatiotemporal autocorrelation such as ... <!-- pls add an example -->?
+- Are some of the predictor variables under consideration known to be highly affected by spatiotemporal autocorrelation such as temperature or precipitation?
 

### 2021-07-26 paper/paper.Rmd
small adjustments

@@
 
-Based on these criteria users should choose a matching partitioning method that is either more restrictive (by discarding observations for fold creation) or more liberal (by using all observations and eventually ignoring natural grouping patterns).
-<!--JM: what do you mean by "using all observations"?-->
+Based on these criteria users should choose a matching partitioning method that is either more restrictive (by discarding observations for fold creation) or more liberal (by not removing observations and eventually ignoring natural grouping patterns).
 For almost all methods implemented in \pkg{mlr3spatiotempcv}, there is a scientific publication describing the application areas of the respective approach (either linked in the help file of the package or its respective upstream packages), hence we are refraining from repeating them again in this study.
@@
 This applies even more to `"spcv_block"` with options `selection = "systematic"` and `selection = "checkerboard"` as the fold assignments are the same in each repetition.
-In general, only methods making use of a random component (e.g. `"spcv_coords"` or `"spcv_block"` with option `selction = "random"`) may be suited for using multiple repetitions.
+In general, only methods making use of a random component (e.g. `"spcv_coords"` or `"spcv_block"` with option `selection = "random"`) may be suited for using multiple repetitions.
 On the upside, `"spcv_coords"` works well with irregularly shaped study areas due to its ability to create partitions which adjust well to the shape of the study area.
@@
 If a large buffer size is chosen relative to the size of the study area, a large portion of it will remain unsampled.
-Choosing an appropriate buffer size is a difficult task.
-In any case, one should refrain oneself from selecting a value based on subjective notions with regard to study area characteristics.
+One should try to refrain from selecting a value based on subjective notions with respect to study area characteristics.
 
@@
 There are pros and cons on whether a new method should be integrated into an existing ecosystem such as \pkg{mlr3} directly first or distributed via a standalone package.
-On the upside, developers can benefit from existing package structures and orient themselves on how previous methods were implemented.
+On the upside, developers can benefit from existing package structures and orient themselves towards how previous methods were implemented.
 Also, the new method will be directly usable by all people using this ecosystem following an existing and familiar syntax.

### 2021-07-26 paper/paper.Rmd

@@
 
-## Should new methods be added to existing ML ecosystems or made available in a new package?
+## Contributing new resampling methods
 
-<!-- mention pros and cons of implementing methods in a standalone pkg vs a framework -->
-There are pros and cons on whether a new method should be integrated into an existing ecosystem such as \pkg{mlr3} directly first or distributed via a standalone package.
-On the upside, developers can benefit from existing package structures and orient themselves towards how previous methods were implemented.
+<!-- discuss pros and cons of implementing methods in a standalone pkg vs a framework -->
+There are pros and cons on whether a new method should be integrated into an existing ecosystem such as \pkg{mlr3} directly or distributed in its own package.
+When contributing a method to an existing ecosystem, developers can benefit from existing structures and orient themselves towards how previous methods were implemented.
+They do not need to worry about complying to various R package related best practices (such as CRAN policies) and can solely focus on the method implementation.
 Also, the new method will be directly usable by all people using this ecosystem following an existing and familiar syntax.
-Contributors do not have to commit maintaining the package in the future, possibly risking its removal in case of major issues.
-In addition, they most likely get a thorough review from other community members during the implementation, possibly enhancing the code side of the implementation.
+Contributions to (ML) ecosystems such as \pkg{mlr3} enables users to easily make use of a large variety of additional functionality, from preprocessing over hyperparameter tuning and feature selection to result visualization [@mlr3], which can be used in conjunction with the new resampling method.
+In addition, contributions most likely undergo a thorough review from other community members during the implementation, most often enhancing the code quality of the implementation.
+A direct contribution also reduces the maintenance load for the ecosystem developers as they do not need to wrap a new method into their system and watch for possible future changes.
+
 On the contrary, people are forced to use the ecosystem and cannot easily use the method outside of it, whether its a different ecosystem or some other way.
-They might also be forced to follow certain guidelines of the ecosystem and potentially feel constrained in their way of doing things.
+They might also be forced to follow certain (coding) guidelines of the ecosystem and potentially feel constrained in their personal way of doing things.
+Often, such ecosystems store their results in specialized classes, which make it hard for users to reuse them outside of the ecosystem.
 

### 2021-08-02 paper/paper.Rmd
applied minor editorial changes

@@
 abstract: >
-  The R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is an extension package of the machine learning framework [mlr3](https://mlr3.mlr-org.com).
+  The R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is an extension package of the machine-learning framework [mlr3](https://mlr3.mlr-org.com).
   It adds support for spatiotemporal resampling methods and respective visualization methods.
-  Spatiotemporal resampling methods are needed to account for the existence of different autocorrelation types (spatial autocorrelation (SAC), temporal autocorrelation (TAC) or spatiotemporal autocorrelation (STAC)) in datasets.
+  Spatiotemporal resampling methods are needed to account for the existence of different autocorrelation types (spatial autocorrelation, temporal autocorrelation or spatiotemporal autocorrelation in datasets.
   These autocorrelation types cause biased performance estimates of varying magnitude in cross-validation when ignored.
@@
   The goal of [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is to gather the available spatiotemporal resampling methods in R and make them available to users through a simple and generic interface.
-  The latter is made possible by integrating the package directly into the [mlr3](https://mlr3.mlr-org.com) machine learning framework which already has support for generic non-spatiotemporal resampling methods such as random partitioning.
-  This simplifies the step of integrating package specific syntax into an overarching machine learning pipeline, making it easier for users to apply a variety of different spatiotemporal resampling methods.
-  The mlr3spatiotempcv package refrains from making suggestions which resampling methods users should apply as this decision mainly depends on the prediction purpose of the fitted model.
+  The latter is made possible by integrating the package directly into the [mlr3](https://mlr3.mlr-org.com) machine-learning framework which already has support for generic non-spatiotemporal resampling methods such as random partitioning.
+  This simplifies the step of integrating package specific syntax into an overarching machine-learning pipeline, making it easier for users to choose from a variety of different spatiotemporal resampling methods.
+  The [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) package refrains from making suggestions which resampling methods users should apply as this decision mainly depends on the prediction purpose of the fitted model.
   No best method exists; only valid methods for the modeling purpose at hand.
@@
 Spatial and spatiotemporal prediction tasks are common in applications ranging from environmental sciences to archaeology and epidemiology.
-While sophisticated mathematical frameworks have long been developed in spatial statistics to characterize predictive uncertainties under well-defined mathematical assumptions such as intrinsic stationarity (e.g., @cressie1993), computational estimation procedures have only been proposed more recently to assess predictive performances of spatial and spatiotemporal prediction models [@brenning2005; @brenning2012; @pohjankukka2017; @roberts2017].
+While sophisticated mathematical frameworks have long been developed in spatial statistics to characterize predictive uncertainties under well-defined mathematical assumptions such as intrinsic stationarity [e.g., @cressie1993], computational estimation procedures have only been proposed more recently to assess predictive performances of spatial and spatiotemporal prediction models [@brenning2005; @brenning2012; @pohjankukka2017; @roberts2017].
 
-Although alternatives such as the bootstrap exist and have some advantages [@efron1983; @hand1997], cross-validation is a particularly well-established, easy-to-implement algorithm for *model assessment* of supervised machine learning models (@efron1983 and next section) and *model selection* [@arlot2010].
-In its basic form, it is based on resampling the data without paying attention to any possible dependence structure, which may arise from, e.g., grouped or structured data [@brenning2008; @pena2015], or underlying environmental processes inducing some sort of spatial coherence at the landscape scale [@brenning2005].
-In treating dependent observations as independent, or ignoring autocorrelation, cross-validation test samples may in fact be heavily correlated with, or even pseudo-replicates of, the data used for training the model, which introduces a potentially severe bias, especially for flexible machine learning models [@brenning2005].
+Although alternatives such as the bootstrap exist and have some advantages [@efron1983; @hand1997], cross-validation is a particularly well-established, easy-to-implement algorithm for *model assessment* of supervised machine-learning models [@efron1983 and next section] and *model selection* [@arlot2010].
+In its basic form, it is based on resampling the data without paying attention to any possible dependence structure, which may arise from, e.g., grouped or structured data, or underlying environmental processes inducing some sort of spatial coherence at the landscape scale.
+In treating dependent observations as independent, or ignoring autocorrelation, cross-validation test samples may in fact be heavily correlated with, or even pseudo-replicates of, the data used for training the model, which introduces a potentially severe bias in assessing the transferability of flexible machine-learning models.
 
 This cross-validation bias is well-known in spatial as well as non-spatial prediction [@brenning2005; @brenning2008; @arlot2010; @roberts2017] and in forecasting [@bergmeir2018].
-It is most easily understood from a predictive modelling perspective by focusing on the question where (and when) the model should be used for prediction.
+It is most easily understood from a predictive modeling perspective by focusing on the question where (and when) the model should be used for prediction.
 In crop classification from remotely-sensed data, for instance, learning samples routinely contain multiple grid cells from a sample of fields with known crop type, for instance 2000 grid cells from 100 fields scattered across a large study region.
-The purpose of training a model on this particular sample is to make predictions on other, new fields within the same geographic domain (*intra-domain* prediction; @brenning2005) --- not *within* the same field, which obviously presents only one crop type.
-In this specific situation it would therefore seem rather unwise to train a model on a simple random subsample of grid cells, and test it on the remaining data, i.e.
-using other grid cells from the same fields, as if we wanted to predict within a field.
+The purpose of training a model on this particular sample is to make predictions on other, new fields within the same geographic domain [*intra-domain* prediction, @brenning2005] --- not *within* the same field, which obviously presents only one crop type.
+In this specific situation it would therefore seem rather unwise to train a model on a simple random subsample of grid cells, and to test it on the remaining data, using other grid cells from the same fields, as if we wanted to predict within a field.
 The results from this performance assessment would be over-optimistic, and perhaps badly so.
@@
 If the model was to be applied to adjacent agricultural districts, i.e.
-outside the learning sample's spatial domain (*extra-domain* prediction; @brenning2005), it would even seem necessary to resample at a higher level of spatial aggregation, i.e.
+outside the learning sample's spatial domain [*extra-domain* prediction, @brenning2005], it would even seem necessary to resample at a higher level of spatial aggregation, i.e.
 at the level of agricultural districts within the learning sample, in order to realistically mimic the actual prediction task.
-The cross-validation resampling needed therefore depends as much on the prediction task as on the data structure or dependency at hand.
+The cross-validation resampling needed therefore depends as much on the prediction task itself as on the data structure or dependency at hand.
 
@@
 
-This toolbox, implemented as an open-source R package, builds upon and generalizes several existing toolboxes that have been developed in recent years for more specific settings (see Table \@ref(tab:sptcv-methods)).
+This toolbox, implemented as an open-source R package, builds upon and generalizes several existing toolboxes that have been developed in recent years for more specific settings (Table&nbsp;\@ref(tab:sptcv-methods)).
 The earliest and most comprehensive of these implementations is the \pkg{sperrorest} R package [@brenning2012], which provides an extensible framework and includes predefined resampling strategies based on, for example, blocking, grouping and buffering.
-In contrast, \pkg{blockCV} and \pkg{ENMeval} were developed for block and buffer resampling with a focus on species distribution modelling [@blockCV; @rest2014; @muscarella2014].
-Neither of these however have been integrated into established machine learning frameworks such as \pkg{mlr}/\pkg{mlr3} or \pkg{caret}/\pkg{tidymodels}, and all of them lack support of temporal prediction tasks.
+In contrast, \pkg{blockCV} and \pkg{ENMeval} were developed for block and buffer resampling with a focus on species distribution modeling [@blockCV; @rest2014; @muscarella2014].
+Neither of these however have been integrated into established machine-learning frameworks such as \pkg{mlr}/\pkg{mlr3} or \pkg{caret}/\pkg{tidymodels}, and all of them lack support of temporal prediction tasks.
 The \pkg{CAST} package, in contrast, focuses on spatiotemporal prediction tasks and makes use of some functions of the \pkg{caret} framework [@cast; @meyer2018].
 One limitation of these packages is the sole focus on model assessment, while the proposed implementation within the \pkg{mlr3} framework offers a seamless integration into model selection.
-It is worth noting that a spatial cross-validation library named \pkg{spacv} has also recently been developed for Python3, which can be used with the \pkg{scikit-learn} machine learning framework [@pedregosa2011].
+It is worth noting that a spatial cross-validation library named \pkg{spacv} has recently been developed for Python3, which can be used with the \pkg{scikit-learn} machine-learning framework [@pedregosa2011].
 
-Thus, \pkg{mlr3spatiotempcv} implements for the first time a comprehensive state-of-the-art compilation of spatial and spatiotemporal partitioning schemes that is well-integrated into a comprehensive machine learning framework in R, the \pkg{mlr3} ecosystem [@mlr3].
+Thus, \pkg{mlr3spatiotempcv} implements for the first time a comprehensive state-of-the-art compilation of spatial and spatiotemporal partitioning schemes that is well-integrated into a comprehensive machine-learning framework in R, the \pkg{mlr3} ecosystem [@mlr3].
 This package is furthermore equipped with a variety of two- and three-dimensional visualization capabilities.
@@
 After presenting the conceptual background in the following section, the overall structure of the \pkg{mlr3spatiotempcv} package is outlined.
-Next, various spatial and spatiotemporal partitioning techniques are contrasted and compared, before their application is demonstrated in a machine learning model assessment in the following section.
+Next, various spatial and spatiotemporal partitioning techniques are contrasted and compared, before their application is demonstrated in a machine-learning model assessment in the following section.
 Finally, limitations of existing methods are discussed and an outlook is given.
@@
 In cross-validation for predictive model assessment, we consider the following formal setting.
-We are interested in predicting a numerical or categorical response $y$ of an object or instance using a feature vector $\mathbf{x} = (x^{(1)}, \ldots, x^{(p)})\in\mathbb{R}^p$ and a model $\hat{f}_\mathcal{L}$ that has been trained on a learning sample $\mathcal{L} = \{(y_i, \mathbf{x}_i),\ i = 1, ..., n\}$.
+We are interested in predicting a numerical or categorical response $y$ of an object or instance using a feature vector $\mathbf{x} = (x^{(1)}, \ldots, x^{(p)})^t\in\mathbb{R}^p$ and a model $\hat{f}_\mathcal{L}$ that has been trained on a learning sample $\mathcal{L} = \{(y_i, \mathbf{x}_i),\ i = 1, ..., n\}$.
 Our goal is to estimate the expected value of the performance of $\hat{f}_\mathcal{L}$,
@@
 \]
-where $l$ is a real-valued loss function, and the expected value is with respect the probability distribution of $X$, the features of an instance $(Y,X)$ drawn randomly from the underlying population.
-This is referred to as the *actual* or *conditional* performance measure as it is conditional on $L$ [@hand1997].
+where $l$ is a real-valued loss function, and the expected value is with respect to the probability distribution of $X$, the features of an instance $(Y,X)$ drawn randomly from the underlying population.
+This is referred to as the *actual* or *conditional* performance measure as it is conditional on $\mathcal{L}$ [@hand1997].
 The loss function can take a variety of forms such as the misclassification error $I(Y\neq\hat{f}_\mathcal{L}(X))$ in classification, or the squared error $(Y-\hat{f}_\mathcal{L}(X))^2$ in regression, among many other possible measures.
-The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance metrics for regression and classification (see e.g., @hand1997 for classification, and @hyndman2006 for regression and forecasting tasks).
+The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance metrics for regression and classification (see, e.g., @hand1997 for classification, and @hyndman2006 for regression and forecasting tasks).
 
@@
 \[
-\widehat{\mathit{perf}}_T(\hat{f}_\mathcal{L}) = \frac{1}{|\mathcal{T}|}\sum_{i=1}^{|\mathcal{T}|}l(Y,\hat{f}_\mathcal{L}(X)).
+\widehat{\mathit{perf}}_T(\hat{f}_\mathcal{L}) = \frac{1}{|\mathcal{T}|}\sum_{(Y,X)\in\mathcal{T}}l(Y,\hat{f}_\mathcal{L}(X)).
 \]
@@
 Since re-using the learning sample $\mathcal{L}$ for testing, i.e.
-$\mathcal{T}=\mathcal{L}$, would result in the over-optimistic *resubstitution* or *apparent* performance, cross-validation partitions the sample $\mathcal{L}$ into disjoint training and test sets.
+$\mathcal{T}:=\mathcal{L}$, would yield the over-optimistic *resubstitution* or *apparent* performance, cross-validation partitions the sample $\mathcal{L}$ into disjoint training and test sets.
 Specifically, $\mathcal{L}$ is split into $k$ partitions,
@@
 \]
-where $f$ is a machine learning algorithm, i.e. a mapping that trains a model $\hat{f}_\mathcal{S}$ using any suitable training sample $\mathcal{S}$.
-The use of $k=5$ or $k=10$ folds is most commonly seen in practice, and these preferences are also supported by theory [@bengio2004, @cawley2010].
+where $f$ is a machine-learning algorithm, i.e. a mapping that trains a model $\hat{f}_\mathcal{S}$ using any suitable training sample $\mathcal{S}$.
+The use of $k=5$ or $k=10$ folds is most commonly seen in practice, and these preferences are also supported by theory [@bengio2004; @cawley2010].
 The $k$-fold CV estimator of model performance is a nearly unbiased estimator of the conditional performance measure when the observations were drawn independently [@efron1983].
-Since $\widehat{\mathit{perf}}_{\mathcal{L}, CV}(f)$ still depends on the particular partitioning chosen for $L$, it is sometimes recommended to repeat the estimation using different random partitionings ($r$-repeated $k$-fold cross-validation) to reduce the influence of randomness when creating partitions [@vanwinckelen2012].
+Since $\widehat{\mathit{perf}}_{\mathcal{L}, CV}(f)$ still depends on the particular partitioning chosen for $\mathcal{L}$, it is sometimes recommended to repeat the estimation using different random partitionings ($r$-repeated $k$-fold cross-validation) to reduce the influence of randomness when creating partitions [@vanwinckelen2012].
 
-In traditional cross-validation, the partitioning is based on uniform random sampling, which ignores spatial or temporal autocorrelation or any existing grouping structure as well as the structure of the prediction task, as outlined in section \@ref(introduction), and may result in over-optimistic performance estimates.
-Several approaches have therefore been proposed in the literature and implemented in software to accommodate a variety of predictive situations (Table \@ref(tab:sptcv-methods)).
+In traditional cross-validation, the partitioning is based on uniform random resampling, which ignores spatial or temporal autocorrelation or any existing grouping structure as well as the structure of the prediction task, and may result in over-optimistic performance estimates.
+Several approaches have therefore been proposed in the literature and implemented in software to accommodate a variety of predictive situations (Table&nbsp;\@ref(tab:sptcv-methods)).
 
-Approaches based on spatial blocking or grouping require either the construction of spatial zones, or the use of pre-existing spatial structures in the data.
-Let's refer to these spatial units or zones as $\mathcal{Z_i}$, $1\le i\le n_z$.
+Approaches based on *spatial blocking* or *grouping* require either the construction of spatial zones, or the use of pre-existing spatial structures in the data.
+Let's refer to these spatial units or zones as $\mathcal{Z}_i$, $1\le i\le n_z$.
 These zones are often constructed to serve as the $k=n_z$ spatial partitions, for example by performing $k$-means clustering of the sample coordinates [@russ2010] or generating the desired number of rectangular blocks.
-Similarly, the data can be partitioned in feature space instead of geographic space, which has been referred to as environmental blocking [@roberts2017].
+We refer to CV based on this resampling scheme as *block CV*.
+Similarly, the data can be partitioned in feature space instead of geographic space, which has been referred to as *environmental blocking* [@roberts2017].
 
@@
 This approach has been applied, for example, in spatial CV at the agricultural field level [@pena2015].
-We would like to emphasize the conceptual distinction between CV *at the group (or block) level*, referring to the latter scenario, from *custom (or grouped)* CV, where the blocks or groups themselves define the CV partitions.
+We would like to emphasize the conceptual distinction between CV *at the group (or block) level*, referring to this scenario, and the various previously types of block CV, where the blocks or groups themselves define the CV partitions.
 
@@
 Nevertheless, spatial variants have been proposed that apply an exclusion buffer to the test locations to separate them from the training data [@brenning2005].
-One plausible approach for this is to use the range of autocorrelation of model residuals to determine the buffer distance as this ensures independence conditional on the predictors [@brenning2005].
+We refer to this resampling scheme as *spatial buffering*.
+One approach that has been proposed for defining a separation distance is to use the range of autocorrelation of model residuals to determine the buffer distance as this ensures independence conditional on the predictors [@brenning2005; @roberts2017].
 When data is grouped, e.g., due to multi-level sampling designs or the study of spatial objects, it has been proposed to apply LOO at the site level [@martin2008; @kasurak2011] or, in animal movement studies, at the animal level [@anderson2005]; we will refer to this as leave-location-out (LLO) CV.
@@
 It should be noted that $k$-fold cross-validation with a large value of $k$, and LOO-CV in particular ($k=n$), is not only very time-consuming since the model has to be trained $k$ times; these models will also be nearly identical since only a tiny fraction of the data is withheld, and therefore estimation bias increases.
-'Pure' LOO-CV is therefore not recommended for machine learning model assessment.
+'Pure' LOO-CV is therefore not recommended for machine-learning model assessment.
 
@@
 
-With the increased awareness of the importance of spatial and spatiotemporal resampling strategies and the growing popularity of R in environmental modeling and geocomputation, it is important to equip machine learning frameworks such as \pkg{mlr3} with suitable algorithms.
-In this context, the \pkg{mlr3} ecosystem stands out as a unified, object-oriented and extendable framework designed to accommodate numerous machine learning tasks with a variety of learners, feature and model selection tools, and model assessment capabilities [@mlr3; @mlr3book].
+With the increased awareness of the importance of spatial and spatiotemporal resampling strategies and the growing popularity of R in environmental modeling and geocomputation, it is important to equip machine-learning frameworks such as \pkg{mlr3} with suitable algorithms.
+In this context, the \pkg{mlr3} ecosystem stands out as a unified, object-oriented and extensible framework designed to accommodate numerous machine-learning tasks with a variety of learners, feature and model selection tools, and model assessment capabilities [@mlr3; @mlr3book].
 All of these are supported by advanced visualization tools, which are particularly important in a spatial and spatiotemporal setting.
@@
 Issues that practitioners often face include varying argument lists of learners, different return values of `predict` methods, and support for only specific feature types.
-These challenges result in substantial overhead and possible reproducibility issues, which are exacerbated by asynchronous development timelines of different components of the machine learning pipelines.
+These challenges result in substantial overhead and possible reproducibility issues, which are exacerbated by asynchronous development timelines of different components of the machine-learning pipelines.
 
-Within the \pkg{mlr3} ecosystem, partitioning strategies are represented by their own objects of class \texttt{Resampling}, most of which are available within \pkg{mlr3} itself (e.g., regular random CV); other specialized strategies (e.g., \texttt{TaskClassifST} or \texttt{TaskRegrST}) are defined in extension packages such as \pkg{mlr3spatiotempcv}.
-In the machine learning pipeline, these objects define the data splits used for model evaluation and selection (hyperparameter tuning) by machine learning algorithms.
-Spatial and spatiotemporal partitioning techniques in \pkg{mlr3spatiotempcv} are currently mostly imported and interfaced from other packages, in particular \pkg{sperrorest}, \pkg{blockCV} and \pkg{CAST} [@blockCV; @brenning2012; @cast], in order to expose them to \pkg{mlr3} functionality.
-To reduce first-level dependencies, some methods were re-implemented instead of importing them from upstream packages.
+Within the \pkg{mlr3} ecosystem, partitioning strategies are represented by their own objects of class \texttt{Resampling}, most of which are available within \pkg{mlr3} itself (e.g., random CV); other specialized strategies (e.g., \texttt{TaskClassifST} or \texttt{TaskRegrST}) are defined in extension packages such as \pkg{mlr3spatiotempcv}.
+In the machine-learning pipeline, these objects define the data splits used for model assessment and selection (hyperparameter tuning) by machine-learning algorithms.
+Spatial and spatiotemporal partitioning techniques in \pkg{mlr3spatiotempcv} are currently mostly imported and interfaced from other packages, in particular \pkg{sperrorest}, \pkg{blockCV} and \pkg{CAST} [@brenning2012; @blockCV; @cast], in order to expose them to \pkg{mlr3} functionality.
+To reduce dependencies, some methods were re-implemented instead of importing them from upstream packages.
 
@@
 
-Partitioning objects in \pkg{mlr3spatiotempcv} are equipped with generic `plot()` and `autoplot()` methods that visualize the created groups.
-`autoplot()` is \pkg{ggplot2}-based and uses [ggplot2](https://ggplot2.tidyverse.org) [@ggplot2] in two-dimensional geographic space and [plotly](https://github.com/ropensci/plotly) [@plotly] in three dimensions, i.e.
-geographic space plus time.
+Partitioning objects in \pkg{mlr3spatiotempcv} are equipped with generic `plot()` and `autoplot()` methods that visualize the groups created.
+`autoplot()` is \pkg{ggplot2}-based and uses [ggplot2](https://ggplot2.tidyverse.org) [@ggplot2] in two-dimensional geographic space and [plotly](https://github.com/ropensci/plotly) [@plotly] in three dimensions, i.e., geographic space plus time.
 
@@
 
-\pkg{mlr3spatiotempcv} currently implements the partitioning methods identified in Table \@ref(tab:sptcv-methods) with their \pkg{mlr3} class name.
+\pkg{mlr3spatiotempcv} currently implements the partitioning methods identified in Table&nbsp;\@ref(tab:sptcv-methods) with their \pkg{mlr3} class name.
 Several of the implemented algorithms are themselves versatile toolboxes with multiple options.
@@
   \footnotesize
-  \caption[t]{Spatiotemporal resampling methods, ordered alphabetically by class name (column 'mlr3 Notation').
-    Column count refers to the number of studies which were found to have used this resampling technique until 2021-05.
+  \caption[t]{Spatiotemporal resampling methods implemented in the mlr3spatiotempcv package, ordered alphabetically by class name (column 'mlr3 Notation').
+    Column count refers to the number of studies which were found to have used this resampling technique until May 2021.
     For each method, up to three randomly chosen references were added to the table as footnotes.}
@@
 
-Spatial partitioning methods can be roughly divided into three groups: (1) buffering, (2) blocking and (3) clustering methods (Table \@ref(tab:sptcv-methods)).
+Spatial partitioning methods can be roughly divided into three groups: (1) buffering, (2) blocking and (3) clustering methods (Table&nbsp;\@ref(tab:sptcv-methods)).
 If a method could have been placed into multiple categories, we deliberately assigned the group which matched best the main characteristic of the respective method.
@@
 
-"Buffering" refers to the idea of removing observations between test and training set by drawing a circular buffer around one or more observations [@rest2014, @brenning2012, @blockCV].
+"Buffering" refers to the idea of removing observations located between test and training sets by drawing a circular buffer around one or more observations [@brenning2012; @rest2014; @blockCV].
 Within this buffer zone, all observations are removed from the training set, thereby reducing the similarity between training and test set.
@@
 However, there is no practical limitation that would prevent this method to be applied in a 2D + time setting.
-Since \pkg{mlr3spatiotempcv} supports two spatial-only "buffering" implementations, (1) "Spatial buffering" from package \pkg{blockCV} and (2) "Leave-one-disc-out cross-validation" from \pkg{sperrorest}, buffering was attributed to the category of spatial partitioning methods (see Table \@ref(tab:sptcv-methods)).
+Since \pkg{mlr3spatiotempcv} supports two spatial-only "buffering" implementations, (1) "Spatial buffering" from package \pkg{blockCV} and (2) "Leave-one-disc-out cross-validation" from \pkg{sperrorest}, buffering was attributed to the category of spatial partitioning methods (Table&nbsp;\@ref(tab:sptcv-methods)).
 
@@
 
-The "buffering" method from the \pkg{blockCV} package is conceptually similar to a leave-one-out (LOO) cross-validation approach [@stone1974].
-It comes with optional support for presence/absence data handling, as often used in the field of species distribution modeling (SDM) [@hijmans2013].
+The "buffering" method from the \pkg{blockCV} package is conceptually related to spatial leave-one-out (LOO) cross-validation.
+It comes with optional support for presence/absence data handling, as often used in the field of species distribution modeling [@hijmans2013].
 However, by default, the method considers all points for sampling the training and test sets.
@@
 Due to its similarity to a LOO CV, this method will result in as many train/test evaluations as observations are present.
-In the visualization of the first fold in Fig. \@ref(fig:buffer) a buffer zone of 1000 m was used without any custom presence/absence treatment of the data.
+In the visualization of the first fold in Figure&nbsp;\@ref(fig:buffer) a buffer zone of 1000 m was used without any custom presence/absence treatment of the data.
 
@@
 
-Another buffering oriented resampling method is "Leave-one-disc-out" from package \pkg{sperrorest}.
-The name originates from the generic "Leave-one-out" method.
-The 'disc' part in the name here refers to the radial area/buffer zone around the test set which is (optionally) being left out within each fold.
-
-In addition, one can specify to sample with replacement by setting `replace = TRUE`.
-
-The method starts from a randomly selected observation and adds all points within the buffer zone area set via variable `radius` by the user.
+Another buffering-oriented resampling method is leave-one-disc-out from package \pkg{sperrorest}.
+Here, 'disc' refers to the buffer zone around the test set which is (optionally) being removed from the training data.
 
 Around the radial test set a buffer zone / neutral area can be defined via variable `buffer` in which observations are excluded from the training set, increasing the spatial distance between training and test sets.
+In addition, one can specify to sample with replacement by setting `replace = TRUE`.
 
@@
 
-"Spatial Blocking" creates homogeneous blocks across the study area splitting observations into different zones [@bahn2012, @wenger2012, @blockCV].
-The creation of these zones can be diverse: one option is to supply rows and columns to subdivide the data into a grid.
+"Spatial Blocking" creates homogeneous blocks across the study area splitting observations into different zones [russ2010; @bahn2012; @wenger2012; @blockCV].
+The creation of these zones can be diverse: one option is to supply rows and columns to subdivide the data based on a rectangular grid.
 Others make use of landscape patterns such as crop zones [@russ2010] or species presence [@blockCV] to determine an initial partitioning.
 
-Implementations of blocking methods often provide additional options to fine tune these initial partitions, allowing for fine-grained partition creation.
+Implementations of blocking methods often provide additional options to fine tune these initial partitioning algorithms.
 
@@
 Additionally, the user might supply a numeric value in meters via argument `theRange` (named just `range` in {mlr3spatiotempcv}) which will ensure a quadratic block pattern across the study area.
-This kind of blocking might be of special interest for SDM studies as well as the possibility to supply a user-defined polygon via argument `rasterLayer` with predefined blocking zones.
+This kind of blocking might be of special interest for species distribution modeling as well as the possibility to supply a user-defined polygon via argument `rasterLayer` with predefined blocking zones.
 
-After the initial block creation, multiple "blocks" are combined to form a partition in correspondence with the number of folds requested by the user.
-For example, in a five-fold scenario with a grid of four columns and five rows, the method would combine one fifth of all blocks (= four blocks) into a single partition.
-
-The following subsections showcase some of the introduced options.
+The following subsections demonstrate the use of some of the introduced options.
 
@@
 
-This option ignores parameter `range` and always creates two partitions following a "1,2,1,2" pattern.
+This option ignores parameter `range` and always creates two partitions following a checkerboard pattern. The arguments `range` and `folds` are ignored in this case.
 
@@
 In contrast to "Spatial Blocking" from \pkg{blockCV} these blocks can be rotated (via argument `rotation`) and a minimum number of observations per block can be set (argument `min_n`).
-A predefined row/column grid is required via either argument `dsplit` or `nsplit` and no square blocks can be used.
+A predefined row/column grid is specified via one of the arguments `dsplit` or `nsplit` and no square blocks are supported.
 Small tiles can be merged with adjacent blocks via argument `reassign` or optionally be ignored (`reassign = FALSE`).
@@
 
-The idea of cluster analysis, which originated in the 1930s in the field of anthropology [@driver1932, @zubin1938], is to group objects with similar characteristics.
+Cluster analysis groups objects with similar characteristics.
 In the spatial context, the similarity of spatial coordinates is used to form such groups.
@@
 
-The so-called "Spatial CV" method after [@russ2010; @brenning2012] and implemented in package \pkg{sperrorest} uses the coordinates of all observations to create clusters in the spatial domain with the help of the $k$-means clustering algorithm.^[$k$-means clustering belongs to the centroid-based clustering family.]
+Coordinate-based clustering for spatial CV [@russ2010; @brenning2012] as implemented in package \pkg{sperrorest} uses the coordinates of all observations to create clusters in the spatial domain with the help of the $k$-means clustering algorithm.
 To avoid confusion with other spatial partitioning methods, we refer to this as coordinate-based clustering, or as `spcv_coords` in \pkg{mlr3spatiotempcv}, instead of using the generic "Spatial CV" term.
@@
 
-```{r coords, fig.cap="Method 'Spatial CV' showing k-means clustering based on coordinates.", out.width="40%", fig.pos="ht"}
+```{r coords, fig.cap="Method 'Spatial CV' showing $k$-means clustering based on coordinates.", out.width="40%", fig.pos="ht"}
 resampling_coords = rsmp("spcv_coords", folds = 5)$instantiate(task)
@@
 To avoid redundancy and putting too much weight on certain methods, only the spatiotemporal examples of these methods will be shown in the following sections.
-For the respective other spaces, usually only one of the space or time variable inputs needs to be omitted.
+For the respective other spaces, usually only either the space or time coordinates need to be omitted from the user input.
 
@@
 
-Group-level CV follows the principle of an inherit grouping structure of the observations, both in space and/or time, to determine (groups of) partitions.
-In contrast to blocking or clustering not the spatial or temporal location is used but additional metadata such as the membership to a specific crop type [@brenning2006].
+Group-level CV is based on the idea that an inherent grouping structure is present the observations or the sampling design in space and/or in time, which can be used to determine partitions.
+In contrast to blocking or clustering, the spatial or temporal location is not used explicitly, but additional metadata is exploited such as the membership in a specific crop type [@brenning2006].
 The idea to follow an inherited grouping structure of the data is also referred to as "target-oriented validation" [@cast].
@@
 
-\pkg{mlr3spatiotempcv} supports two "grouping" implementations: (1) The "Leave-location-and-time-out" family from package \pkg{CAST} and (2) "Custom Resampling" from the \pkg{mlr3} package.
+\pkg{mlr3spatiotempcv} supports two "grouping" implementations: (1) The "leave-location-and-time-out" family from package \pkg{CAST} and (2) "custom resampling" from the \pkg{mlr3} package.
 
@@
 
-Leave-Location-and-Time-Out (LLTO) from package \pkg{CAST} is a spatiotemporal resampling method [@meyer2018].
+Leave-location-and-time-out (LLTO) from package \pkg{CAST} is an implementation of spatiotemporal resampling method [@meyer2018].
 It requires both a spatial and a temporal variable which are used to separate observations in a multi-dimensional space.
-It comes with satellite methods named "Leave-time-out" (LTO) and "Leave-location-out" (LLO) which only operate in a two-dimensional space.
+It comes with satellite methods named "leave-time-out" (LTO) and "leave-location-out" (LLO) which only operate in a two-dimensional space.
 
@@
 
-The following code uses the `cookfarm` dataset as the base, modifies the `Date` variable in such a way that it contains five unique levels and then creates a spatiotemporal regression task in \pkg{mlr3spatiotempcv} (Fig.
-\@ref(fig:lto)).
+The following code uses the `cookfarm` dataset as the base, modifies the `Date` variable in such a way that it contains five unique levels and then creates a spatiotemporal regression task in \pkg{mlr3spatiotempcv} (Figure&nbsp;\@ref(fig:lto)).
 
@@
 Method "Leave-Location-Out" (LLO) works similar to "Leave-Time-Out" (LTO) in that it operates in a two-dimensional space.
-Here, specific locations are left out for forming the partitions while the temporal information is ignored (Fig. \@ref(fig:llo)).
+Here, specific locations are left out individually for forming test partitions while the temporal information is ignored (Figure&nbsp;\@ref(fig:llo)).
 
@@
 To reduce autocorrelation in both dimensions, the methods omits observations as the splitting in test and train sets only provides separation in one dimension whereas two (space and time) are needed to account for both types, spatial and temporal autocorrelation.
-In other words, accounting for one dimension as in Fig.
-\@ref(fig:lto) or Fig.
-\@ref(fig:llo), by only having one time step or location in either test or training set, would have no effect on the respective other.
-Thus, the removal of observation can be seen as a proxied buffering in multiple dimensions as shown in Fig. \@ref(fig:llto).
+In other words, accounting for one dimension as in Figure&nbsp;\@ref(fig:lto) or Figure&nbsp;\@ref(fig:llo), by only having one time step or location in either test or training set, would have no effect on the respective other.
+Thus, the removal of observation can be seen as a proxied buffering in multiple dimensions as shown in Figure&nbsp;\@ref(fig:llto).
 
@@
 Labelled as "Custom CV", users are able to supply an external factor vector with the same length as number of observations.
-This partitioning concept is also ften referred to as "grouped resampling" and should not be mistaken with "group-level resampling" (section \@ref(sec:group-level)), which inherits additional aggregation of groups at the partition level.
+This partitioning concept is also often referred to as "grouped resampling" and should not be mistaken with "group-level resampling" (section \@ref(sec:group-level)), which inherits additional aggregation of groups at the partition level.
 Each level of the factor vector will result in one partition.
@@
 The user can select a variable via argument `feature` which denotes the variable that will be used to create clusters.
-Hereby, k-means will use Euclidean distance for cluster creation.
+Hereby, $k$-means will use Euclidean distance for cluster creation.
 Hence, the selected input variables should be of type numeric.
@@
 
-It is also possible to cluster by multiple features:
+It is also possible to cluster based on multiple features:
 
@@
 At present, \pkg{mlr3spatiotempcv} also supports spatiotemporal partitioning using the versatile CLUTO clustering algorithm [@cluto].
-CLUTO is available in R through the \pkg{skmeans} package, which provides an interface to a downloadable compiled library with a restriction to non-commercial uses (see `?ResamplingSptCVCluto` in \pkg{mlr3spatiotempcv} for more information).
+CLUTO is available in R through the \pkg{skmeans} package, which provides an interface to a downloadable compiled library with a restriction to non-commercial uses (see `help("ResamplingSptCVCluto", package = "mlr3spatiotempcv")` for more information).
 With respect to functionality, \pkg{mlr3spatiotempcv} supports two-dimensional clustering based on coordinates, time or features and three-dimensional clustering on coordinates and time.
@@
 
-Because the spatial grouping of the k-means based approach (`"spcv_coords"`) (Fig.
-\@ref(fig:vis-spcv)) contrasts visually very well to NSpCV (random) partitioning (Fig.
-\@ref(fig:vis-nspcv)), this method was chosen to showcase the 2D plotting capabilities of \pkg{mlr3spatiotempcv}.
+Because the spatial grouping of the $k$-means-based approach (`"spcv_coords"`) (Figure&nbsp;\@ref(fig:vis-spcv)) contrasts visually very well with non-spatial (random) partitioning (Figure&nbsp;\@ref(fig:vis-nspcv)), this method was chosen to showcase the 2D plotting capabilities of \pkg{mlr3spatiotempcv}.
 
@@
 
-- `coordinate_names`: Names of the features which represent the spatial coordinates.
+- `coordinate_names`: Names of the features that represent the spatial coordinates.
   This is automatically inferred when a `sf` object is passed.
@@
 ```{r create-task}
-library(mlr3)
-library(mlr3spatiotempcv)
+library("mlr3")
+library("mlr3spatiotempcv")
 
@@
 Specifically, we use spatial partitioning based on $k$-means clustering of the coordinates (method `"repeated_spcv_coords"`) and compare its results to a non-spatial (random) cross-validation (`"repeated_cv"`), both with (only) four folds and two repetitions for demonstration.
-We choose the area Under the ROC curve (`"classif.auc"`) as the performance measure for this classification task.
+We choose the area under the ROC curve (`"classif.auc"`) as the performance measure for this classification task.
 
-## Data preparation
+## Model preparation
 
-Next, the random forest learner (`lrn("classif.ranger")`) is initialized with default hyperparameters and the prediction type set to `"probability"` due to the binary response variable.
-A set of commonly used learners is available in package \pkg{mlr3learners} [@mlr3learners], including the random forest implementation after @ranger.
+Next, the random forest learner (`lrn("classif.ranger")`) is initialized with default hyperparameters and the prediction type is set to `"probability"` due to the binary response variable.
+A set of commonly used learners is available in package \pkg{mlr3learners} [@mlr3learners], including the random forest implementation of @ranger.
 
@@
 A four-fold, two times repeated non-spatial resampling object is constructed and instantiated via method `$instantiate()`.
-Next, the created resampling object `resampling_nsp` is passed to function `resample()` together with the Task and Learner created earlier to execute the cross-validation.
+Next, the created resampling object `resampling_nsp` is passed to function `resample()` together with the task and learner objects created earlier to execute the model assessment.
 
-The results from the CV are then aggregated using the measure `"classif.auc"`, which is the \pkg{mlr3} package notation for the Area Under the Receivers Operating Curve (AUROC) measure.
+The results from the CV are then aggregated using the measure `"classif.auc"`, which is the \pkg{mlr3} package notation for the AUROC measure.
 
@@
 
-## Spatial cross-validation
+## Spatial cross-validation via coordinate-based clustering
 
-The same approach as for non-spatial resampling is used for spatial resampling, expect replacing method `"repeated_cv"` by `"repeated_spcv_coords"`.
+The same approach as for non-spatial resampling is used for spatial resampling, expect replacing method `"repeated_cv"` with `"repeated_spcv_coords"`.
 
@@
 
-## Fold visualization
+## Visualization of CV partitions
 
-Finally, we visualize the folds that were used during performance estimation making use of the generic `autoplot()` function in package \pkg{mlr3spatiotempcv}.
-Due to space limitations only the first two folds are shown.
+Finally, we visualize the partitions that were used during performance estimation making use of the generic `autoplot()` function in package \pkg{mlr3spatiotempcv} (Figure&nbsp;\@ref(fig:vis-spcv)).
 
-```{r vis-spcv, fig.cap="Visualization of spatial partitioning", fig.pos="ht", fig.height=6, fig.width=8}
+```{r vis-spcv, fig.cap="Visualization of spatial partitioning using coordinate-based clustering.
+Due to space limitations only the first two folds are shown.", fig.pos="ht", fig.height=6, fig.width=8}
 resampling_sp = rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
@@
 
-The resulting AUC value of 0.77 using Non-Spatial Cross-Validation (NSpCV) is around 0.15 better compared to the Spatial Cross-Validation (SpCV) value of 0.61.
+The non-spatial CV estimate of AUC (0.77) is around 0.15 better compared to the spatial CV estimate of 0.61.
 The better performance score results from over-optimism in the estimation with NSpCV.
@@
 
-The respective folds of the resamplings used in this example are shown in Fig. \@ref(fig:vis-spcv) and Fig. \@ref(fig:vis-nspcv).
+The respective folds of the resamplings used in this example are shown in Figures&nbsp;\@ref(fig:vis-spcv) and&nbsp;\@ref(fig:vis-nspcv).
 
@@
 
-The question which resampling method should be chosen for a dataset at hand comes up regularly in practice.
-Even though there is no definitive and easy answer, we would like to give some guidance in this section to help finding the most appropriate method.
+The question which resampling method should be chosen for a prediction task and dataset at hand comes up regularly in practice.
+Even though there is no definitive and easy answer, we would like to give some guidance in this section to help find an appropriate method.
 
 Although the case study example in section \@ref(sec:case-study) used the `"spcv_coords"` method, this should not imply that this method is the best or only method suitable for this example task.
-Even though this method is frequently used, it was mainly chosen because of it's clear visual grouping differences when comparing spatial to random partitioning.
+Even though this method is frequently used, it was mainly chosen because of its clear visual grouping differences when comparing spatial to random partitioning.
 Another advantage is its ability to adapt to the irregularly shaped study area of this example.
@@
   The closer observations are in space and/or time, the more similar they are, and hence the greater is the effect of spatio-temporal autocorrelation.
-- Are some of the predictor variables under consideration known to be highly affected by spatiotemporal autocorrelation such as temperature or precipitation?
+- Are some of the predictor variables under consideration known to be strongly autocorrelated such as air temperature or precipitation?
 
@@
 
-Last, a note on best practices for the example shown in \@ref(sec:case-study):
+Last, a note on best practices for the example shown in section \@ref(sec:case-study):
 In this example a cross-validation without hyperparameter tuning was applied.
@@
 According to the package authors, this estimate should then be used for argument `theRange` to have a sensible value for the block sizes created in method `"spcv_block"`.
-Rectangular partitioning can be problematic for irregularly shaped study areas as it might result in certain partitions containing substantially less observations than others.
+Rectangular partitioning can be problematic in irregularly shaped study areas where some of the resulting partitions may contain substantially fewer observations than others.
 
@@
 <!-- spcv_coords: limited variance due to limited k-means possibilities -->
-For repeated executions of method `spcv_coords`, the variance between repetitions can become quite small due to the limited possibilities of differing `k-means`-based clusters on the same dataset [@brenning2012].
+For repeated executions of method `spcv_coords`, the variance between repetitions can become quite small due to the limited possibilities of differing $k$-means-based clusters on the same dataset [@brenning2012].
 This applies even more to `"spcv_block"` with options `selection = "systematic"` and `selection = "checkerboard"` as the fold assignments are the same in each repetition.
@@
 The idea of "custom resampling" is that each group becomes a test set in the CV.
-By contrast, "group-level resampling" methods apply additional aggregation on top of these initial groups: for example, users can supply an integer value to an argument named `fold` in the LLO method (e.g.
+By contrast, "group-level resampling" methods resample these initial groups: for example, users can supply an integer value to an argument named `fold` in the LLO method (e.g.
 `5L`) which combines the initial groups (e.g.
@@
 "Group-level resampling" can be, in contrast to "custom resampling", repeated multiple times because multiple different aggregations are possible in theory.
-However, it also requires the amount of initial groups to be substantially higher than the amount of folds so that a sensible aggregation is possible in the first place.
+However, it also requires the number of initial groups to be substantially higher than the number of folds so that a sensible aggregation is possible in the first place.
 
 <!-- spcv coords: mention generic name -->
-In @brenning2012, which is the reference commonly used for the k-means based spatial partitioning approach, the method was not introduced with a unique name.
+In @brenning2012, which is the reference commonly used for the $k$-means-based spatial partitioning approach, the method was not introduced with a unique name.
 Over time, studies which have used this method mainly referenced it using the generic "spatial CV" terminology.
@@
 When contributing a method to an existing ecosystem, developers can benefit from existing structures and orient themselves towards how previous methods were implemented.
-They do not need to worry about complying to various R package related best practices (such as CRAN policies) and can solely focus on the method implementation.
+They do not need to worry about complying with various R package related best practices (such as CRAN policies) and can solely focus on the method implementation.
 Also, the new method will be directly usable by all people using this ecosystem following an existing and familiar syntax.
@@
 <!-- missing: Guidelines how to choose a resampling method -->
-The \pkg{mlr3spatiotempcv} package is the first package to bundle and categorize spatiotemporal resampling methods in R.
+The \pkg{mlr3spatiotempcv} package is the first package to bundle and categorize spatiotemporal resampling methods implemented in multiple other packages in R.
 While various spatiotemporal resampling methods exists nowadays to account for spatiotemporal autocorrelation in modeling, there is still lots of uncertainty which method should be used for which dataset.
-The answer to this question is not trivial and while there will most likely never be an easy answer to this, users are advised to first think about the prediction purpose of the model and then about reducing autocorrelation between train and test sets to reduce the impact of spatiotemporal autocorrelation.
+The answer to this question is not trivial and while there will most likely never be an easy answer to this, users are advised to first think about the prediction purpose of the model and then about reducing autocorrelation between training and test sets to reduce the impact of spatiotemporal autocorrelation.
 One option could be the proposal of a dataset characteristic checklist showing which characteristics make a certain dataset more suitable for specific methods.

### 2021-08-02 paper/paper.Rmd
edited abstract: focus on prediction task, less on autocorrelation; some rewording

@@
   The R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is an extension package of the machine-learning framework [mlr3](https://mlr3.mlr-org.com).
-  It adds support for spatiotemporal resampling methods and respective visualization methods.
-  Spatiotemporal resampling methods are needed to account for the existence of different autocorrelation types (spatial autocorrelation, temporal autocorrelation or spatiotemporal autocorrelation in datasets.
-  These autocorrelation types cause biased performance estimates of varying magnitude in cross-validation when ignored.
-  At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: CAST, blockCV and sperrorest.
-
-  The goal of [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is to gather the available spatiotemporal resampling methods in R and make them available to users through a simple and generic interface.
-  The latter is made possible by integrating the package directly into the [mlr3](https://mlr3.mlr-org.com) machine-learning framework which already has support for generic non-spatiotemporal resampling methods such as random partitioning.
-  This simplifies the step of integrating package specific syntax into an overarching machine-learning pipeline, making it easier for users to choose from a variety of different spatiotemporal resampling methods.
-  The [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) package refrains from making suggestions which resampling methods users should apply as this decision mainly depends on the prediction purpose of the fitted model.
-  No best method exists; only valid methods for the modeling purpose at hand.
+  It adds support for spatiotemporal resampling methods and their visualization.
+  Spatiotemporal resampling methods are needed to assess model performance in a variety of predictive situations in which spatial, temporal, or spatiotemporal autocorrelation is present.
+  Cross-validation performance estimates may be biased when this is ignored.
+  Currently various R packages implementing different spatiotemporal partitioning strategies exist: sperrorest, blockCV, CAST.
+  The goal of [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is to gather the available spatiotemporal resampling methods in R and make them available to users through a simple and common interface.
+  This is made possible by integrating the package directly into the [mlr3](https://mlr3.mlr-org.com) machine-learning framework, which already has support for generic non-spatiotemporal resampling methods such as random partitioning.
+  This simplifies the step of integrating package-specific syntax into an overarching machine-learning pipeline, making it easier for users to choose from a variety of different spatiotemporal resampling methods.
+  The users' decision to use a specfic resampling method should depend on the predictive task at hand, the autocorrelation within the data, and the structure of the sampling design used to collect the data.
 keywords:

### 2021-08-02 paper/paper.Rmd
edited title: mention purpose (machine learning) and software (R)

@@
 title:
-  formatted: "\\pkg{mlr3spatiotempcv}: Spatiotemporal resampling methods for \\pkg{mlr3}"
+  formatted: "\\pkg{mlr3spatiotempcv}: Spatiotemporal resampling methods for machine learning in R with \\pkg{mlr3}"
   # If you use tex in the formatted title, also supply version without
-  plain:     "mlr3spatiotempcv: spatiotemporal resampling methods for mlr3"
+  plain:     "mlr3spatiotempcv: spatiotemporal resampling methods for machine learning in R with mlr3"
   # For running headers, if needed

### 2021-08-02 paper/paper.Rmd
fixed misunderstanding related to crop classification; minor changes sec. 2, 3; and moved 4.1.2.2 up

@@
 In crop classification from remotely-sensed data, for instance, learning samples routinely contain multiple grid cells from a sample of fields with known crop type, for instance 2000 grid cells from 100 fields scattered across a large study region.
-The purpose of training a model on this particular sample is to make predictions on other, new fields within the same geographic domain [*intra-domain* prediction, @brenning2005] --- not *within* the same field, which obviously presents only one crop type.
+The purpose of training a model on this particular sample is to make predictions on other, new fields within the same geographic domain [*intra-domain* prediction, @brenning2005] --- not *within* the same field, which obviously presents only a single crop type that is already known from the training sample.
 In this specific situation it would therefore seem rather unwise to train a model on a simple random subsample of grid cells, and to test it on the remaining data, using other grid cells from the same fields, as if we wanted to predict within a field.
@@
 To mimic the predictive situation for which the model is trained, one would rather have to resample at the level of fields, not grid cells [@pena2015].
-If the model was to be applied to adjacent agricultural districts, i.e.
-outside the learning sample's spatial domain [*extra-domain* prediction, @brenning2005], it would even seem necessary to resample at a higher level of spatial aggregation, i.e.
-at the level of agricultural districts within the learning sample, in order to realistically mimic the actual prediction task.
+If the model was to be applied to adjacent agricultural regions, i.e., outside the learning sample's spatial domain [*extra-domain* prediction, @brenning2005], it would even seem necessary to resample at a higher level of spatial aggregation, i.e.
+at the level of agricultural sub-regions within the learning sample, in order to realistically mimic the actual prediction task.
 The cross-validation resampling needed therefore depends as much on the prediction task itself as on the data structure or dependency at hand.
@@
 This toolbox, implemented as an open-source R package, builds upon and generalizes several existing toolboxes that have been developed in recent years for more specific settings (Table&nbsp;\@ref(tab:sptcv-methods)).
-The earliest and most comprehensive of these implementations is the \pkg{sperrorest} R package [@brenning2012], which provides an extensible framework and includes predefined resampling strategies based on, for example, blocking, grouping and buffering.
+The earliest and most comprehensive of these implementations is the \pkg{sperrorest} R package [@brenning2012], which provides an extensible framework and includes predefined resampling strategies based on, for example, blocking, clustering, and buffering.
 In contrast, \pkg{blockCV} and \pkg{ENMeval} were developed for block and buffer resampling with a focus on species distribution modeling [@blockCV; @rest2014; @muscarella2014].
@@
 Let's refer to these spatial units or zones as $\mathcal{Z}_i$, $1\le i\le n_z$.
-These zones are often constructed to serve as the $k=n_z$ spatial partitions, for example by performing $k$-means clustering of the sample coordinates [@russ2010] or generating the desired number of rectangular blocks.
-We refer to CV based on this resampling scheme as *block CV*.
+These zones are often constructed to serve as the $k=n_z$ spatial partitions, for example by performing $k$-means clustering of the sample coordinates [@russ2010], which we refer to as *coordinate-based clustering*, or generating the desired number of rectangular blocks.
+The zones may also be defined by a modeler based on an arbitrary (e.g., polygonal) partitioning of the study region, which we refer to as *custom resampling*.
 Similarly, the data can be partitioned in feature space instead of geographic space, which has been referred to as *environmental blocking* [@roberts2017].
@@
 Although this is just a special case of non-spatial CV with $k=n$, it is sometimes also referred to as spatial CV [@willmott2006].
-Nevertheless, spatial variants have been proposed that apply an exclusion buffer to the test locations to separate them from the training data [@brenning2005].
+
+Spatial variants have been proposed that apply an exclusion buffer to the test locations to separate them from the training data [@brenning2005].
 We refer to this resampling scheme as *spatial buffering*.
 One approach that has been proposed for defining a separation distance is to use the range of autocorrelation of model residuals to determine the buffer distance as this ensures independence conditional on the predictors [@brenning2005; @roberts2017].
-When data is grouped, e.g., due to multi-level sampling designs or the study of spatial objects, it has been proposed to apply LOO at the site level [@martin2008; @kasurak2011] or, in animal movement studies, at the animal level [@anderson2005]; we will refer to this as leave-location-out (LLO) CV.
+When data is grouped, e.g., due to multi-level sampling designs or the study of spatial objects, it has been proposed to apply LOO at the site level [@martin2008; @kasurak2011] or, in animal movement studies, at the animal level [@anderson2005].
+We will broadly refer to such groups of observations as 'blocks' in a generic sense, and therefore refer to this type of LOO-CV as *leave-one-block-out CV*.
 
@@
 Turning to prediction tasks with spatiotemporal data, various spatial, temporal, or spatiotemporal partitioning strategies are being used, depending on the specific study objectives.
-While the former two ignore the spatial and temporal dimension of the data, respectively, it has also been proposed to leave out spatiotemporal blocks [@meyer2018] or clusters [@cluto].
+While the former two ignore the spatial and temporal dimension of the data, respectively, it has also been proposed to leave out random subsets of locations and time points [@meyer2018] or spatiotemporal clusters [@cluto].
+<!-- AB: This short paragraph does certaintly not adequately describe Meyer's LLTO with its unique buffer in space and time dimensions. But I'm not sure if it's worth the effort. -->
 
@@
 
+#### Leave-one-disc-out --- sperrorest
+
+Another buffering-oriented resampling method is leave-one-disc-out from package \pkg{sperrorest}.
+Here, 'disc' refers to the buffer zone around the test set which is (optionally) being removed from the training data.
+
+Around the radial test set a buffer zone / neutral area can be defined via variable `buffer` in which observations are excluded from the training set, increasing the spatial distance between training and test sets.
+In addition, one can specify to sample with replacement by setting `replace = TRUE`.
+
+```{r disc, fig.cap="Visualization of 'Leave-one-disc-out' method.", out.width="40%"}
+resampling_disc = rsmp("spcv_disc", folds = 5, radius = 300L, buffer = 400L)
+resampling_disc$instantiate(task)
+
+autoplot(resampling_disc,
+  size = 0.8, task = task, fold_id = 1) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```
+
 #### Spatial buffering --- blockCV
@@
 <!-- TODO: Discuss issues setting a sensible range value -->
-
-#### Leave-one-disc-out --- sperrorest
-
-Another buffering-oriented resampling method is leave-one-disc-out from package \pkg{sperrorest}.
-Here, 'disc' refers to the buffer zone around the test set which is (optionally) being removed from the training data.
-
-Around the radial test set a buffer zone / neutral area can be defined via variable `buffer` in which observations are excluded from the training set, increasing the spatial distance between training and test sets.
-In addition, one can specify to sample with replacement by setting `replace = TRUE`.
-
-```{r disc, fig.cap="Visualization of 'Leave-one-disc-out' method.", out.width="40%"}
-resampling_disc = rsmp("spcv_disc", folds = 5, radius = 300L, buffer = 400L)
-resampling_disc$instantiate(task)
-
-autoplot(resampling_disc,
-  size = 0.8, task = task, fold_id = 1) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
-```
 
@@
 Group-level CV is based on the idea that an inherent grouping structure is present the observations or the sampling design in space and/or in time, which can be used to determine partitions.
-In contrast to blocking or clustering, the spatial or temporal location is not used explicitly, but additional metadata is exploited such as the membership in a specific crop type [@brenning2006].
-The idea to follow an inherited grouping structure of the data is also referred to as "target-oriented validation" [@cast].
+In contrast to blocking or clustering, the spatial or temporal location is not used explicitly, but additional metadata is exploited such as the membership in a specific agricultural management unit [@brenning2006], or in a site in hierarchical sampling [@kasurak2011].
+<!--
+This has also been referred to as "target-oriented validation" [@cast].
+FIXME:
+AB: I don't think I agree with this.
+My interpretation of the Meyer paper was that in the Meyer paper, locations or time points are the targets.
+Fields consist of many grid cells, at multiple location.
+-->
 Even though flexibility with respect to fold creation is limited due to the inherited grouping structure within the data, some implementations allow to specify the resulting number of folds by omitting parts of the dataset.
@@
 The prediction purpose is the model's ability for which areas or zones it actually can make predictions.
-In the crop example given in section \@ref(sec:intro) this would be to distinguish and predict different crop types. 
-Hence, the CV partitioning is done on the crop level to receive a honest estimate of the model's performance.
+In the crop example given in section \@ref(sec:intro) this would be to distinguish and predict the crop type of 'new', unseen fields within an agricultural region. 
+Hence, the CV partitioning is done at the field level to receive a honest estimate of the model's performance in a relevant predictive situation.
 

### 2021-08-03 paper/paper.Rmd
corrected argument name theRange vs range

@@
 The "Spatial blocking" from package \pkg{blockCV} comes with various blocking options such as random (default) (`selection = "random"`), systematic (`selection = "systematic"`) or checkerboard pattern (`selection = "checkerboard"`).
-Additionally, the user might supply a numeric value in meters via argument `theRange` (named just `range` in {mlr3spatiotempcv}) which will ensure a quadratic block pattern across the study area.
+Additionally, the user might supply a numeric value in meters via argument `range` which will ensure a quadratic block pattern across the study area.
 This kind of blocking might be of special interest for species distribution modeling as well as the possibility to supply a user-defined polygon via argument `rasterLayer` with predefined blocking zones.
@@
 
-Alternatively, arguments `rows` and `cols` can be specified instead of `theRange`.
+Alternatively, arguments `rows` and `cols` can be specified instead of `range`.
 This allows to create rectangular blocks which honor the supplied `rows` and `cols` settings.
@@
 <!-- spcv_block: how to set argument `theRange` -->
-Method `"spcv_block"` requires the user to supply a value for argument `theRange` which denotes the block size used to create the blocking scheme across the study area.
+<!-- AB: argument is actually called range, not theRange -->
+Method `"spcv_block"` requires the user to supply a value for argument `range` which denotes the block size used to create the blocking scheme across the study area.
 This specification is non-trivial and therefore package \pkg{blockCV} provides some helper functions (`spatialAutoRange()` and `rangeExplorer()`) to conduct a data-driven estimation of the distance at which the spatial autocorrelation in the data levels off [@blockCV].
-According to the package authors, this estimate should then be used for argument `theRange` to have a sensible value for the block sizes created in method `"spcv_block"`.
+According to the package authors, this estimate should then be used for argument `range` to have a sensible value for the block sizes created in method `"spcv_block"`.
 Rectangular partitioning can be problematic in irregularly shaped study areas where some of the resulting partitions may contain substantially fewer observations than others.

### 2021-08-03 paper/paper.Rmd
move library calls to earlier code chunk (because of changed subsubsubsection order)

@@
 knitr::opts_chunk$set(fig.path = "./paper/figs/",
-  cache = T,
+  cache = FALSE,
   cache.lazy = TRUE)
@@
 ```{r disc, fig.cap="Visualization of 'Leave-one-disc-out' method.", out.width="40%"}
+library("mlr3")
+library("mlr3spatiotempcv")
+task = tsk("ecuador")
 resampling_disc = rsmp("spcv_disc", folds = 5, radius = 300L, buffer = 400L)
@@
 ```{r buffer, fig.cap="Visualization of the 'Spatial buffering' method from package `blockCV`.", out.width="40%", fig.pos="ht"}
-library("mlr3")
-library("mlr3spatiotempcv")
-task = tsk("ecuador")
 resampling_buffer = rsmp("spcv_buffer", theRange = 1000)

### 2021-08-03 paper/paper.Rmd
revised section 5.5 Summary

@@
 The non-spatial CV estimate of AUC (0.77) is around 0.15 better compared to the spatial CV estimate of 0.61.
-The better performance score results from over-optimism in the estimation with NSpCV.
-The actual predictive performance on new data is much more realistically reflected by the performance estimated via SpCV.
-The magnitude of this difference is variable as it depends on the dataset, the magnitude of STAC and the learner itself.
-Algorithms with a higher tendency of overfitting to the training set will have a larger spread in such scenarios.
+The better performance results from over-optimism in the estimation with non-spatial CV.
+Spatial CV, in contrast, provides a better measure of a model's ability to generalize from the training sample --- in the Ecuador case study, from the hillslopes and historical landslides in the training sample.
+It is also expected that spatial CV results better reflect the performance of a model on new, spatially separate data.
+The magnitude of this difference may depend on the dataset, the strength on spatial or spatiotemporal autocorrelation, and the learner itself.
+Algorithms with a higher tendency to overfit to the training set will tend to have a larger spread in such scenarios.
 
-The respective folds of the resamplings used in this example are shown in Figures&nbsp;\@ref(fig:vis-spcv) and&nbsp;\@ref(fig:vis-nspcv).
+<!-- AB: TODO: Check if the following paragraphs overlaps with the Discussion section; perhaps merge into Discussion. -->
+Since spatial CV can also be used for hyperparameter tuning in the \pkg{mlr3} framework, the model can be optimized to show an improved performance in a specific predictive setting [@schratz2019].
+This may, for example, involve a reduced maximum tree depth in the Ecuador case study, since this leads to stronger generalization and reduced overfitting.
 

### 2021-08-03 paper/paper.Rmd
add explanation to section 5: case study

@@
 set.seed(42)
+```
 
-# equal to `tsk("ecuador")`
+To set up the task `"ecuador"` manually, we would have to proceed as follows:
+
+```{r create-task}
+# equivalent to 'tsk("ecuador")':
 b = mlr3::as_data_backend(ecuador)

### 2021-08-03 paper/paper.Rmd
change case study title to step-by-step example
revised Conclusion and Outlook section

@@
 
-# Case study: Comparing spatial and non-spatial resampling {#sec:case-study}
+# Step-by-step example: Comparing spatial and non-spatial CV {#sec:case-study}
 
@@
 
-```{r create-task}
+```{r load-pkgs}
 library("mlr3")
@@
 
-# Summary and outlook
+# Conclusion and outlook
 
 <!-- missing: Guidelines how to choose a resampling method -->
+<!-- AB: I don't think we can present such guidelines in this paper, I think we discussed this recently and agreed that we will not propose such guidelines, we should rather focus on the methods and their implementation -->
 The \pkg{mlr3spatiotempcv} package is the first package to bundle and categorize spatiotemporal resampling methods implemented in multiple other packages in R.
-While various spatiotemporal resampling methods exists nowadays to account for spatiotemporal autocorrelation in modeling, there is still lots of uncertainty which method should be used for which dataset.
-The answer to this question is not trivial and while there will most likely never be an easy answer to this, users are advised to first think about the prediction purpose of the model and then about reducing autocorrelation between training and test sets to reduce the impact of spatiotemporal autocorrelation.
-One option could be the proposal of a dataset characteristic checklist showing which characteristics make a certain dataset more suitable for specific methods.
-This could then help users to make an appropriate and objective decision which resampling method might be a good fit for their dataset and prospective prediction task.
+The available resampling techniques allow the users to vary the scale or granularity of the resampled spatial units as well as their shape and possible buffer distance between training and test samples.
+These settings may account for the specific characteristics of spatiotemporal prediction tasks, but modelers now have to make the important decision of choosing a method that is adequate for their situation.
+They are advised to focus on the spatial or spatiotemporal structure of the learning sample at hand and of the model's prediction task.
 
-<!--mention missing spatiotemporal buffering method -->
-At the time of writing this article, there is no buffering method which operates in the spatiotemporal domain.
-We encourage readers of this article to check out the supported methods in the latest released version of the \pkg{mlr3spatiotempcv} package directly to potentially discover new methods which might have been added to the package in the meantime.
+The compilation of resampling techniques in \pkg{mlr3spatiotempcv} is by no means complete.
+Additional methods or parameters may therefore be added in the future as they become available in upstream package or are contributed directly to this package.
 
 <!-- signoff -->
-Spatiotemporal cross-validation is still not yet a paradigm that is completely established in scientific workflows even though it is discussed intensively for more than a decade now.
-Making most available methods easily accessible to users is an important step to foster the acceptance of such in the community and to lead to less biased cross-validation results in environmental and ecological studies.
+Spatiotemporal cross-validation a paradigm that is not yet fully established in scientific workflows although it has been discussed intensively for more than a decade now.
+We anticipate that making the existing methods easily accessible to users is an important step to foster the acceptance of spatiotemporal cross-validation in the community and to allow modelers to produce less biased cross-validation results in environmental and ecological studies.
 

### 2021-08-03 paper/paper.Rmd
revised Conclusion and Outlook (small addition)

@@
 These settings may account for the specific characteristics of spatiotemporal prediction tasks, but modelers now have to make the important decision of choosing a method that is adequate for their situation.
-They are advised to focus on the spatial or spatiotemporal structure of the learning sample at hand and of the model's prediction task.
+They are advised to focus on the spatial or spatiotemporal structure of the model's prediction task, consider the structure of the learning sample at hand, and think about how the autocorrelation between training and test samples might affect their model assessment and selection.
 

### 2021-08-03 paper/paper.Rmd
revised all figure captions; added a few TODO comments

@@
 knitr::opts_chunk$set(fig.path = "./paper/figs/",
-  cache = FALSE,
+  cache = TRUE,
   cache.lazy = TRUE)
@@
 
-```{r disc, fig.cap="Visualization of 'Leave-one-disc-out' method.", out.width="40%"}
+```{r disc, fig.cap='Visualization of one training set / test set combination generated with the leave-one-disc-out method from package sperrorest (method "spcv\\_disc" in mlr3spatiotempcv).
+The disc has a radius of 300 m and is surrounded by a 400-m buffer.', out.width="40%"}
 library("mlr3")
@@
 
-```{r buffer, fig.cap="Visualization of the 'Spatial buffering' method from package `blockCV`.", out.width="40%", fig.pos="ht"}
+```{r buffer, fig.cap='Visualization of the spatial buffering method from package blockCV (method "spcv\\_buffer" in  mlr3spatiotempcv). The buffer distance is 1000 m.', out.width="40%", fig.pos="ht"}
 resampling_buffer = rsmp("spcv_buffer", theRange = 1000)
@@
 
-```{r block-random, fig.cap="Spatial blocking with option `selection = 'random'`.", out.width="40%", fig.pos="ht"}
+```{r block-random, fig.cap='Random resampling of square spatial blocks using the implementation in package blockCV (method "spcv\\_block" with option selection = "random" in mlr3spatiotempcv).
+The size of the squares is 1000 m, and four out of the 19 blocks were assigned to this test sample.', out.width="40%", fig.pos="ht"}
 resampling_block_random = rsmp("spcv_block", range = 1000, folds = 5)
@@
 
-```{r block-systematic, fig.cap="Spatial blocking with option `selection = 'systematic'`.", out.width="40%", fig.pos="ht"}
+```{r block-systematic, fig.cap='Sytematic resampling of square spatial blocks using the implementation in package blockCV (method "spcv\\_block" with option selection = "systematic" in mlr3spatiotempcv).
+The size of the squares is 1000 m, and four out of the 19 blocks were assigned to this test sample.', out.width="40%", fig.pos="ht"}
 resampling_block_systematic = rsmp("spcv_block",
@@
 
-```{r block-checkerboard, fig.cap="Spatial blocking with option `selection = 'checkerboard'`.", out.width="40%", fig.pos="ht!"}
+```{r block-checkerboard, fig.cap='Checkerboard resampling of square spatial blocks using the implementation in package blockCV (method "spcv\\_block" with option selection = "checkerboard" in mlr3spatiotempcv).
+The size of the squares is 1000 m, and eleven out of the 19 blocks were assigned to this test sample.', out.width="40%", fig.pos="ht!"}
 resampling_block_checkerboard = rsmp("spcv_block",
@@
 
-```{r block-rows-and-cols, fig.cap="Spatial blocking with arguments `rows` and `cols`.", out.width="40%", fig.pos="ht"}
+```{r block-rows-and-cols, fig.cap='Random resampling of a specifid number of rectangular blocks as implemented in package blockCV (method "spcv\\_block" in mlr3spatiotempcv with arguments rows = 6 and cols = 3).', out.width="40%", fig.pos="ht"}
 resampling_block_rows_cols = rsmp("spcv_block",
@@
 
-```{r tile, fig.cap="Spatial tiling", out.width="40%", fig.pos="ht"}
+```{r tile, fig.cap='Leave-one-block-out resampling from package sperrorest (method "spcv\\_tiles" in package mlr3spatiotempcv with argument nsplit = c(3,4) indicating the number of rows and columns).', out.width="40%", fig.pos="ht"}
 resampling_tiles = rsmp("spcv_tiles", nsplit = c(3L, 4L))
@@
 
-```{r coords, fig.cap="Method 'Spatial CV' showing $k$-means clustering based on coordinates.", out.width="40%", fig.pos="ht"}
+```{r coords, fig.cap='Leave-one-block-out CV based on $k$-means clustering of the coordinates as implemented in package sperrorest (method "spcv\\_coords" in mlr3spatiotempcv).', out.width="40%", fig.pos="ht"}
 resampling_coords = rsmp("spcv_coords", folds = 5)$instantiate(task)
@@
 
-```{r lto, echo=FALSE, out.width="70%", fig.cap="Three-dimensional visualization of method `sptcv\\_cstf` and sub-method Leave-Time-Out (LTO).
-To support showcasing the temporal grouping, five folds and five temporal levels in variable `Date` were used in this example.
-Due to the tilt angle of the static 3D plot the last two levels are hard to distinguish visually but they are separated in the same way as the the first two levels on the top of the plot.", fig.pos="ht"}
+```{r lto, echo=FALSE, out.width="70%", fig.cap='Perspective plot of leave-time-out CV from package CAST (method "sptcv\\_cstf" and argument time\\_var = "Date").
+Only five folds and five time points were used in this example.
+Note that the blue dots correspond to five discrete time levels, which appear as a point cloud due to the viewing angle.', fig.pos="ht"}
 knitr::include_graphics("lto.png")
@@
 
-```{r llo, echo=FALSE, out.width="70%", fig.cap="Three-dimensional visualization of method 'sptcv\\_cstf' and sub-method 'Leave-Location-Out' (LLO).
-To support showcasing the spatial grouping, the viewing angle has been adjusted to a view from top.
-Points are stacked by temporal levels and only points from the same location are chosen to form a partition.", fig.pos="ht"}
+```{r llo, echo=FALSE, out.width="70%", fig.cap='Perspective plot of leave-location-out CV from package CAST (method "sptcv\\_cstf" and argument space\\_var = "SOURCEID").
+The data is viewed from the future towards the past.', fig.pos="ht"}
 knitr::include_graphics("llo.png")
@@
 
-```{r llto, echo=FALSE, out.width="70%", fig.cap="Three-dimensional visualization of method 'sptcv\\_cstf' and sub-method 'Leave-Location-And-Time-Out' (LLTO).
-The grey points show the points omitted by this method, i.e.
-the points which are neither used in the training nor test set in this fold.", fig.pos="ht"}
+```{r llto, echo=FALSE, out.width="70%", fig.cap='Perspective plot of leave-location-and-time-out CV from package CAST (method "sptcv\\_cstf" and arguments time\\_var = "Date" and space\\_var = "SOURCEID").
+The grey points are excluded from both the training and the test set in this fold.', fig.pos="ht"}
 knitr::include_graphics("llto.png")
@@
 
-```{r custom-cv, fig.cap="Custom resampling using an external factor vector defining the partitions. Each coloring represents one partition.", out.width="40%", fig.pos="ht"}
+<!-- TODO: AB: This diagram is inconsistent with the other diagrams - should display training and test samples. -->
+```{r custom-cv, fig.cap='Leave-one-level-out (custom) resampling from package mlr3 (method "custom\\_cv").
+A factor variable is used to define the partitions.
+Each color represents one partition.', out.width="40%", fig.pos="ht"}
 breaks = quantile(task$data()$dem, seq(0, 1, length = 6))
@@
   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+
 ```
@@
 
-```{r env-1, fig.cap="Environmental blocking using a single predictor.", out.width="40%", fig.pos="ht"}
+```{r env-1, fig.cap='Environmental leave-one-block-out CV from package blockCV using a single predictor to define blocks in feature space (method "spcv\\_env" in mlr3spatiotempcv with argument features = "distdeforest").', out.width="40%", fig.pos="ht"}
 resampling_env = rsmp("spcv_env",
@@
 
-```{r env-2, fig.cap="Environmental blocking using multiple predictors.", out.width="40%", fig.pos="ht"}
+<!-- TODO: AB: should be combined with the previous figure into a two-panel figure. Differences are much easier to recognize if plotted next to each other. -->
+```{r env-2, fig.cap='Environmental leave-one-block-out CV from package blockCV using two predictors to define blocks in feature space (method "spcv\\_env" in mlr3spatiotempcv with argument features = c("distdeforest", "slope")).', out.width="40%", fig.pos="ht"}
 resampling_env_multi = rsmp("spcv_env",
@@
 ```
+<!-- AB: TODO: In general, the code would look much simpler if we could get rid of the scale_x/y_continuous() calls. Improve default settings of autoplot()? -->
 
@@
 
-```{r vis-spcv, fig.cap="Visualization of spatial partitioning using coordinate-based clustering.
-Due to space limitations only the first two folds are shown.", fig.pos="ht", fig.height=6, fig.width=8}
+```{r vis-spcv, fig.cap="Spatial leave-one-block-out partitioning using coordinate-based clustering to create roughly equally sized polygonal blocks.
+Due to space limitations only the first two folds of the first repetition are shown.", fig.pos="ht", fig.height=6, fig.width=8}
 resampling_sp = rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
@@
 
-```{r vis-nspcv, fig.cap="Visualization of random partitioning", fig.pos="ht", fig.height=6, fig.width=8}
+```{r vis-nspcv, fig.cap="Random (non-spatial) four-fold CV partitioning. Only the first two folds of the first repetition are shown.", fig.pos="ht", fig.height=6, fig.width=8}
 resampling_nsp = rsmp("repeated_cv", folds = 4, repeats = 2)

### 2021-08-04 paper/paper.Rmd
revised and reorganized step-by-step section, removed repeated information

@@
 
-We will use a well-known case study to demonstrate the application of spatial resampling techniques.
-This binary classification task on landslide susceptibility in Ecuador [@muenchow2012] is available as a built-in task via `tsk("ecuador")`.
+We use a well-known case study to demonstrate the application of spatial and non-spatial resampling techniques for model assessment in \pkg{mlr3spatiotempcv}.
+The objective of landslide susceptibility modeling is to predict how prone to landslide initiation a location is.
+Models are fitted to historical landslide occurrences, but they need to learn generalizable relationships between predisposing variables and the response as opposed to perfectly reproducing or memorizing the historical distribution.
+This binary classification task on landslides in Ecuador [@muenchow2012] is available as a built-in task via `tsk("ecuador")`, but we generate it from the learning sample.
+Random forest is used as a classifier, and the area under the ROC curve (AUROC) as the performance measure.
 
-Because the spatial grouping of the $k$-means-based approach (`"spcv_coords"`) (Figure&nbsp;\@ref(fig:vis-spcv)) contrasts visually very well with non-spatial (random) partitioning (Figure&nbsp;\@ref(fig:vis-nspcv)), this method was chosen to showcase the 2D plotting capabilities of \pkg{mlr3spatiotempcv}.
+Spatial CV is implemented in the form of leave-one-block-out CV using coordinate-based $k$-means clustering to generate  irregularly shaped blocks of roughly equal size. 
+This approach is better suited for the shape of the present study area than a rectangular partitioning.
+Figures&nbsp;\@ref(fig:vis-spcv)) and&nbsp;\@ref(fig:vis-nspcv)) show the contrasting distributions of training and test samples.
+For demonstration purposes we use only four CV folds and two repetitions.
 
-Besides the required arguments of the \pkg{mlr3} superclass `Task`, the `TaskClassifST` / `TaskRegrST` classes from package \pkg{mlr3spatiotempcv}  also require the following arguments during construction which must be passed via a named list `extra_args`:
+## Task preparation
+
+In \pkg{mlr3}, machine-learning tasks with their respective dataset and response variable are represented by objects of the class `Task`.
+\pkg{mlr3spatiotempcv}'s spatial and spatiotemporal machine-learning tasks are derived from this superclass.
+Specifically, the `TaskClassifST` and `TaskRegrST` classes for classification and regression tasks require several additional arguments that must be passed as a named list using the `extra_args` argument:
 
@@
   This is automatically inferred when a `sf` object is passed.
-- `coords_as_features`: Whether the coordinates should be used as features during modeling.
-  By default coordinates are not used as predictors.
-- `crs`: The coordinate reference system of the data. Valid inputs are PROJ strings or EPSG codes in the format `ESPG:<code>`.
+- `coords_as_features`: Whether the coordinates should be used as features; by default they are not.
+- `crs`: The coordinate reference system of the data as a PROJ string or EPSG code in the format `ESPG:<code>`.
 
 At first all necessary R packages are loaded and a lower verbosity is set to keep the output tidy.
-Also a seed is set for deterministic execution and the \pkg{mlr3spatiotempcv} built-in task `"ecuador"` is loaded via `tsk()`.
+A random-number seed is set for reproducibility.
 
@@
 
-To set up the task `"ecuador"` manually, we would have to proceed as follows:
+The task `"ecuador"` is available in \pkg{mlr3spatiotempcv} through `tsk("ecuador")`.
+To set it up manually, we proceed as follows:
 
+<!-- AB: TODO: The comment says that b$hash is optional, but the reader will not know what it is good for.
+A more verbose comment would be helpful, maybe 'optional task identifier' or whatever the hash name may be good for.
+-->
 ```{r create-task}
-# equivalent to 'tsk("ecuador")':
 b = mlr3::as_data_backend(ecuador)
@@
 
-To show the effect of spatial autocorrelation on performance estimates in cross-validation, we will apply spatial and non-spatial cross-validation to the prediction of landslides using the random forest technique (`"classif.ranger"`).
-Specifically, we use spatial partitioning based on $k$-means clustering of the coordinates (method `"repeated_spcv_coords"`) and compare its results to a non-spatial (random) cross-validation (`"repeated_cv"`), both with (only) four folds and two repetitions for demonstration.
-We choose the area under the ROC curve (`"classif.auc"`) as the performance measure for this classification task.
-
 ## Model preparation
 
-Next, the random forest learner (`lrn("classif.ranger")`) is initialized with default hyperparameters and the prediction type is set to `"probability"` due to the binary response variable.
+Next, the random forest learner (`"classif.ranger"`) is initialized with default hyperparameters and the prediction type is set to `"probability"` because the model is used for soft classification.
 A set of commonly used learners is available in package \pkg{mlr3learners} [@mlr3learners], including the random forest implementation of @ranger.
@@
 
-A four-fold, two times repeated non-spatial resampling object is constructed and instantiated via method `$instantiate()`.
+A resampling is generated in \pkg{mlr3} with the `rsmp()` function, which is used to generate a resampling object for four-fold non-spatial CV with two repetitions.
+The object is constructed and instantiated with its `$instantiate()` method.
+
 Next, the created resampling object `resampling_nsp` is passed to function `resample()` together with the task and learner objects created earlier to execute the model assessment.
+This is the actual, potentially time-consuming CV estimation. With the present settings, eight random forest classifiers are fitted and evaluated in this step.
 
-The results from the CV are then aggregated using the measure `"classif.auc"`, which is the \pkg{mlr3} package notation for the AUROC measure.
+Model performances are calculated from the CV predictions using the AUROC (`"classif.auc"` in \pkg{mlr3} notation).
 
@@
 
-The same approach as for non-spatial resampling is used for spatial resampling, expect replacing method `"repeated_cv"` with `"repeated_spcv_coords"`.
+The model assessment is now repeated using spatial CV resampling, for which we only need to replace `"repeated_cv"` with `"repeated_spcv_coords"`.
 
@@
 Due to space limitations only the first two folds of the first repetition are shown.", fig.pos="ht", fig.height=6, fig.width=8}
-resampling_sp = rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
-resampling_sp$instantiate(task)
-
 autoplot(resampling_sp, task, fold_id = c(1:2), size = 0.8) *
@@
 ```{r vis-nspcv, fig.cap="Random (non-spatial) four-fold CV partitioning. Only the first two folds of the first repetition are shown.", fig.pos="ht", fig.height=6, fig.width=8}
-resampling_nsp = rsmp("repeated_cv", folds = 4, repeats = 2)
-resampling_nsp$instantiate(task)
-
 autoplot(resampling_nsp, task, fold_id = c(1:2), size = 0.8) *
@@
 
-The non-spatial CV estimate of AUC (0.77) is around 0.15 better compared to the spatial CV estimate of 0.61.
+The non-spatial CV estimate of AUC (0.76) is around 0.13 better compared to the spatial CV estimate of 0.64.
 The better performance results from over-optimism in the estimation with non-spatial CV.
-Spatial CV, in contrast, provides a better measure of a model's ability to generalize from the training sample --- in the Ecuador case study, from the hillslopes and historical landslides in the training sample.
+Spatial CV, in contrast, provides a better measure of a model's ability to generalize from the training sample --- in this case study, from the specific hillslopes and historical landslides in the training sample.
 It is also expected that spatial CV results better reflect the performance of a model on new, spatially separate data.
-The magnitude of this difference may depend on the dataset, the strength on spatial or spatiotemporal autocorrelation, and the learner itself.
+The magnitude of the difference between spatial and non-spatial CV estimates may depend on the dataset, the strength on spatial or spatiotemporal autocorrelation, and the learner itself.
 Algorithms with a higher tendency to overfit to the training set will tend to have a larger spread in such scenarios.

### 2021-08-04 paper/paper.Rmd
rewrote LODO-CV subsection, changed folds to 100

@@
 
-Another buffering-oriented resampling method is leave-one-disc-out from package \pkg{sperrorest}.
-Here, 'disc' refers to the buffer zone around the test set which is (optionally) being removed from the training data.
+Leave-one-disc-out resampling from package \pkg{sperrorest} randomly selects a circular test set and optionally excludes a buffer zone from the remaining training data.
+It thus ensures that a minimum separation distance between training and test data is maintained.
+Discs are centered at sample locations, their number being specified by the `folds` argument.
+It is therefore possible to sample up to $n$ discs without replacement (default) or any number when sampling with replacement (`replace = TRUE`).
+The number of discs is specified via the `folds` argument.
 
-Around the radial test set a buffer zone / neutral area can be defined via variable `buffer` in which observations are excluded from the training set, increasing the spatial distance between training and test sets.
-In addition, one can specify to sample with replacement by setting `replace = TRUE`.
+This resampling method is, strictly speaking, not a partitioning method since the discs are not a disjoint.
+Nevertheless, it is a logical extension of spatial LOO CV.
+Depending on the size of the discs, potentially a large number of discs (`folds` argument) is needed to effectively use the entire dataset.
 
+Leave-one-disc-out resampling becomes spatial LOO CV for a radius of 0 m and when each observation is at a unique location.
+
+<!-- AB: TODO: folds = 5 makes very little sense in the case of LODO-CV, since this would only sample 5 discs, which cover only a small fraction of the study area.
+Let's use folds = 100, this will at least make readers think about their choice of #folds.
+-->
 ```{r disc, fig.cap='Visualization of one training set / test set combination generated with the leave-one-disc-out method from package sperrorest (method "spcv\\_disc" in mlr3spatiotempcv).
@@
 task = tsk("ecuador")
-resampling_disc = rsmp("spcv_disc", folds = 5, radius = 300L, buffer = 400L)
+resampling_disc = rsmp("spcv_disc", folds = 100, radius = 300L, buffer = 400L)
 resampling_disc$instantiate(task)

### 2021-08-04 paper/paper.Rmd
rewrote leave-one-tile-out subsection

@@
 
-#### Spatial Tiles --- sperrorest
+#### Leave-one-tile-out resampling --- sperrorest
 
-Method `partition_tiles()` from package \pkg{sperrorest} also uses rectangular blocks to partition the study area.
-In contrast to "Spatial Blocking" from \pkg{blockCV} these blocks can be rotated (via argument `rotation`) and a minimum number of observations per block can be set (argument `min_n`).
-A predefined row/column grid is specified via one of the arguments `dsplit` or `nsplit` and no square blocks are supported.
-Small tiles can be merged with adjacent blocks via argument `reassign` or optionally be ignored (`reassign = FALSE`).
-The number of tiles set by the user equals the number of folds and blocks cannot be grouped into folds as in "Spatial Blocking".
-Also despite setting 12 blocks via `nsplit = c(3L, 4L)` in the following example, the final partition count is eleven because the bottom-left part of the study area does not contain observations.
+Leave-one-tile-out resampling is implemented in the `"spcv_tiles"` method imported from package \pkg{sperrorest}. 
+It uses rectangular blocks that can be rotated (argument `rotation`), and a minimum number or fraction of observations per block can optionally be achieved by iteratively merging small blocks into adjacent blocks  (argument `reassign` in conjunction with `min_n` or `min_frac`).
+Block size or number is specified via the argument `dsplit` or `nsplit`, respectively, and square blocks can be obtained with a single (or two identical) `dsplit` value.
+
+Note that the actual number of folds obtained may be smaller than `nsplit[1]*nsplit[2]` (or smaller than what would be expected base don `dsplit`) since some blocks may be empty or (optionally) merged into adjacent folds.
+In the example, there are only eleven folds instead of twelve because the southwestern part of the study area's bounding box does not contain observations  (Figure&nbsp;\@ref(fig:tile)).
 

### 2021-08-04 paper/paper.Rmd
revised coordinate-based clustering section

@@
 
-### Clustering
+### Clustering-based resampling
 
-Cluster analysis groups objects with similar characteristics.
-In the spatial context, the similarity of spatial coordinates is used to form such groups.
-Different cluster algorithms such as hierarchical, density-based, grid-based or centroid-based clustering exist [@gower1967].
+Cluster analysis provides a flexible approach to creating irregularly shaped spatial blocks for spatial resampling.
+Numerous techniques are available that can potentially be applied to the spatial coordinates of observations, to the features, or to a combination of both.
+In spatial model assessment, the focus has been on coordinate-based clustering, and specifically on leave-one-block-out resampling with blocks created by $k$-means clustering of the coordinates [@russ2010].
 
-#### Coordinate-based clustering --- sperrorest
+#### Leave-one-block-out resampling using coordinate-based clustering --- sperrorest
 
 Coordinate-based clustering for spatial CV [@russ2010; @brenning2012] as implemented in package \pkg{sperrorest} uses the coordinates of all observations to create clusters in the spatial domain with the help of the $k$-means clustering algorithm.
-To avoid confusion with other spatial partitioning methods, we refer to this as coordinate-based clustering, or as `spcv_coords` in \pkg{mlr3spatiotempcv}, instead of using the generic "Spatial CV" term.
+This can be regarded as a leave-one-block-out resampling method, or as a $k$-fold CV in which each test set is a spatial cluster. 
+This method is referred to as `spcv_coords` in \pkg{mlr3spatiotempcv}.
 
-This approach is very versatile as it adapts to irregularly-shaped study areas and ensures that exactly $k$ partitions are created.
+Coordinate-based clustering approach is very versatile as it adapts to irregularly-shaped study areas and ensures that exactly $k$ partitions are created, which are usually or very similar size when the sample locations are spread out evenly.
 Nevertheless, despite the random selection of initial cluster centers, repeated partitionings may in some cases be nearly identical.
-Also, $k$-means clustering may be less suitable for data sets with existing clusters of points and/or with distant isolated sample locations.
+Also, $k$-means clustering may be less suitable for data sets with pre-existing clusters of points and/or with distant isolated sample locations.
+When distinct clusters of points are present, as in multi-level sampling, it may be better to define clusters using a factor variable (see method `"custom_cv"` in Section \@ref(sec:custom-cv)).
 
@@
 
-### Custom and grouped resampling --- mlr3
+### Custom and grouped resampling --- mlr3 {#sec:custom-cv}
 

### 2021-08-05 paper/paper.Rmd
one sentence per line

@@
 
-Note that the actual number of folds obtained may be smaller than `nsplit[1]*nsplit[2]` (or smaller than what would be expected base don `dsplit`) since some blocks may be empty or (optionally) merged into adjacent folds.
+Note that the actual number of folds obtained may be smaller than `nsplit[1]*nsplit[2]` (or smaller than what would be expected based on `dsplit`) since some blocks may be empty or (optionally) merged into adjacent folds.
 In the example, there are only eleven folds instead of twelve because the southwestern part of the study area's bounding box does not contain observations  (Figure&nbsp;\@ref(fig:tile)).

### 2021-08-06 paper/paper.Rmd
revised blocking section, reduced #figures

--- 2021-08-04 paper/paper.Rmd
@@
 
-### Blocking
+### Spatial blocking
 
-"Spatial Blocking" creates homogeneous blocks across the study area splitting observations into different zones [russ2010; @bahn2012; @wenger2012; @blockCV].
-The creation of these zones can be diverse: one option is to supply rows and columns to subdivide the data based on a rectangular grid.
-Others make use of landscape patterns such as crop zones [@russ2010] or species presence [@blockCV] to determine an initial partitioning.
+Spatial blocking divides the study area into different zones [russ2010; @bahn2012; @wenger2012; @blockCV].
+Several methods have been implemented to create rectangular as well as irregularly shaped blocks. 
+These blocks can be used for leave-one-block-out resampling, or they can themselves be grouped into $k$ partitions for a $k$-fold CV at the block level.
 
-Implementations of blocking methods often provide additional options to fine tune these initial partitioning algorithms.
+\pkg{mlr3spatiotempcv} supports multiple resampling methods that are based on blocks: 
+(1) "Spatial blocking" from package \pkg{blockCV} implements $k$-fold CV resampling at the level of square or rectangular blocks, while (2) "Spatial tiles" from \pkg{sperrorest} offers leave-one-block-out resampling with square or rectangular, possibly rotated blocks.
+In addition, irregularly shaped blocks are available (3) through coordinate-based clustering as a leave-one-block-out technique from \pkg{sperrorest} (method `"spcv_coords"`, Section&nbsp;\@ref(sec:cluster-cv)), and (4) as a CV at the block level where blocks are defined by a factor variable that defines zones or groups (method `"custom_cv"` in \pkg{mlr3}, Section&nbsp;\@ref(sec:custom-cv)).
 
-\pkg{mlr3spatiotempcv} supports two "blocking" implementations: (1) "Spatial blocking" from package \pkg{blockCV} and (2) "Spatial Tiles" from \pkg{sperrorest}.
-
-#### Spatial blocking --- blockCV
+#### Square and rectangular blocks --- blockCV
 
@@
 
-The following subsections demonstrate the use of some of the introduced options.
-
-##### Option: selection = "random"
+This procedure supports both random and systematic resampling of square blocks (argument `selection = "random"` and `"systematic"`, respectively).
+Block sizes in meters) are determined by the `range` argument.
+Alternatively, rectangular blocks can be created by specifying the number of desired rows and columns (arguments `rows` and `cols`).
 
@@
 
-##### Option: selection = "systematic"
-
-Here, aggregation follows the linear count from $1:n_{folds}$ from top to bottom.
+In systematic resampling, the blocks are numbered row by row, and blocks $i+j\cdot\texttt{folds}$ are assigned to fold $i$.
+This may create undesired patterns when the number of columns is equal to or a multiple of the number of folds.
 
@@
 
+<!--
+
 ##### Option: selection = "checkerboard"
 
-This option ignores parameter `range` and always creates two partitions following a checkerboard pattern. The arguments `range` and `folds` are ignored in this case.
+-->
+
+The *checkerboard* partitioning is a special case of a systematic block partitioning (`selection = "checkerboard"`).
+It inherently supports only two folds, making it less appealing than the more commonly used five- or ten-fold resampling, which achieve larger training set sizes.
+
+<!--
+AB:
+The text previously stated that the range argument is ignored. This can't be true, there must be some argument that controls the size of checkerboard fields.
+-->
+
+<!--
+AB: Checkerboard pattern is just a special case of a systematic pattern and likely not of great practical relevance because there are only two folds -> omit figure.
 
@@
 resampling_block_checkerboard = rsmp("spcv_block",
-  range = 1000, folds = 5,
+  range = 1000, folds = 2,
   selection = "checkerboard")
@@
 
-\newpage
+-->
 
-##### Rectangular blocks
-
-Alternatively, arguments `rows` and `cols` can be specified instead of `range`.
-This allows to create rectangular blocks which honor the supplied `rows` and `cols` settings.
+<!--
+AB: This figure and subsubsubsection takes up a lot of space for very little additional information.
 
@@
 \newpage
+
+-->
 
@@
 
-### Clustering-based resampling
+### Clustering-based resampling {#sec:cluster-cv}
 
@@
 
-autoplot(resampling_custom, size = 0.8, task = task) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+# autoplot(resampling_custom, size = 0.8, task = task) *
+#   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+#   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
 

### 2021-08-06 paper/paper.Rmd
remove b$hash, remove FIXME, remove temporary comment

@@
 b = mlr3::as_data_backend(data)
-b$hash = "_mlr3_tasks_cookfarm_"
 task_spt = TaskRegrST$new(
@@
 
-<!--JM: you might want to delete the FIXME-->
-Support for user-defined partitioning strategies is built into \pkg{mlr3} directly since version 0.12.0 FIXME.
-Labelled as "Custom CV", users are able to supply an external factor vector with the same length as number of observations.
+Support for user-defined partitioning strategies is built into \pkg{mlr3} directly.
+In this so-called "Custom CV", users are able to supply an external factor vector with the same length as number of observations.
 This partitioning concept is also often referred to as "grouped resampling" and should not be mistaken with "group-level resampling" (section \@ref(sec:group-level)), which inherits additional aggregation of groups at the partition level.
@@
 
-# autoplot(resampling_custom, size = 0.8, task = task) *
-#   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-#   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+autoplot(resampling_custom, size = 0.8, task = task) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
 

### 2021-08-06 paper/paper.Rmd
some more revisions to spatial blocking section

@@
 
-The "Spatial blocking" from package \pkg{blockCV} comes with various blocking options such as random (default) (`selection = "random"`), systematic (`selection = "systematic"`) or checkerboard pattern (`selection = "checkerboard"`).
-Additionally, the user might supply a numeric value in meters via argument `range` which will ensure a quadratic block pattern across the study area.
-This kind of blocking might be of special interest for species distribution modeling as well as the possibility to supply a user-defined polygon via argument `rasterLayer` with predefined blocking zones.
+Block-level resampling from package \pkg{blockCV} supports both random and systematic resampling of square blocks (`method = "spcv_block"` with argument `selection = "random"` and `"systematic"`, respectively; Figure&nbpsp;~\@ref(fig:block-random) and \@ref(fig:block-systematic)).
+The implementation in \pkg{blockCV} has specific options that are designed for modeling presence-only data, which is a typical situation in species distribution modeling.
+Users can also supply a user-defined polygon via argument `rasterLayer` with predefined blocking zones.
 
-This procedure supports both random and systematic resampling of square blocks (argument `selection = "random"` and `"systematic"`, respectively).
-Block sizes in meters) are determined by the `range` argument.
+Block sizes (in meters) are determined by the `range` argument.
 Alternatively, rectangular blocks can be created by specifying the number of desired rows and columns (arguments `rows` and `cols`).

### 2021-08-06 paper/paper.Rmd
some quick edits in Discussion section

@@
 
-There are various variables influencing the partitioning method choice for an environmental dataset:
+There are various factors that might influence the choice of a partitioning method:
 
-- Do natural groupings exists in the data? Examples include agricultural fields, forest stands, lithology types, etc.
+- Do natural groupings exists in the data?
+Examples include multi-level sampling designs, land management units, or other characteristics of the study area or sampling design that may induce random effects.
 - How dense are the observations scattered both in space and time?
-  The closer observations are in space and/or time, the more similar they are, and hence the greater is the effect of spatio-temporal autocorrelation.
-- Are some of the predictor variables under consideration known to be strongly autocorrelated such as air temperature or precipitation?
+  The closer observations are in space and/or time, the more similar they are, and hence the greater is the effect of spatio-temporal autocorrelation on performance estimates.
+- Are some of the predictor variables under consideration known to be strongly autocorrelated such as air temperature or precipitation at the regional level?
 
@@
 Both are similar in that they require external user input for partition creation, i.e. a factor variable which assigns all observations into a distinct group.
-The idea of "custom resampling" is that each group becomes a test set in the CV.
+The idea of "custom resampling" is that each group becomes a test set in the CV (leave-one-group-out CV).
 By contrast, "group-level resampling" methods resample these initial groups: for example, users can supply an integer value to an argument named `fold` in the LLO method (e.g.
@@
 Hence, in package \pkg{mlr3spatiotempcv} this method was named `"spcv_coords"`, making use of the fact that the spatial coordinates are used for a cluster-based partition creation.
-
-<!-- env block: hard to decide which variable to use for clustering -->
-<!-- env block: more than one variable may lead to unforeseeable results -->
-Method `"spcv_env"` is complicated to use in that it requires the user to limit the clustering to one or more variables [@blockCV].
-This implies that only this/these variable(s) suffer from autocorrelation and others are ignored.
-However, it is not possible to estimate how much a single variable accounts for the autocorrelation in the data and the same applies for estimating the influence of the non-accounted variables which might add substantial bias to the CV.
 

### 2021-09-17 paper/paper.Rmd
use new method categorization (in table and text)

@@
 Comprehensive and up-to-date information can be found in the package's online documentation (<https://mlr3spatiotempcv.mlr-org.com>).
-The following sections, which follow the naming scheme `<method name> --- <R package>`, give an overview of most existing partitioning strategies and their visualization options.
+The following sections give an overview of most existing partitioning strategies and their visualization options.
 The available methods are further discussed in section \@ref(sec:disc).
@@
   \footnotesize
-  \caption[t]{Spatiotemporal resampling methods implemented in the mlr3spatiotempcv package, ordered alphabetically by class name (column 'mlr3 Notation').
-    Column count refers to the number of studies which were found to have used this resampling technique until May 2021.
+  \caption[t]{Spatiotemporal resampling methods available in the mlr3 ecosystem.
+    Column name shows the mlr3 method name as found in the `mlr3::mlr\_resamplings` dictionary.
+    Column "Count" refers to the number of studies which were found to have used this resampling technique until May 2021.
+    Literature appearances were only collected for methods with a "spcv" or "sptcv" prefix as methods which rely on custom grouping ("custom\_cv" and "cv") do not have a dedicated name or identifier to make them detectable in research papers.
     For each method, up to three randomly chosen references were added to the table as footnotes.}
@@
       \\
-      Type                            & Name                        & R Package(s)     & mlr3 Notation  & Count\\
+      Type                            & Name                        & R Package     & Sub-Type  & Count\\
       \toprule
-      \multirow{6}{*}{Spatial}        & Spatial Buffering           & \specialcell{\pkg{blockCV} \\ \pkg{sperrorest}}    & \specialcell{ \texttt{"spcv\_buffer"} \tablefootnote{\label{ploton2020}\cite{ploton2020}}  \tablefootnote{\label{diesing2020}\cite{diesing2020}} \\  \texttt{"spcv\_disc"} \tablefootnote{\label{endicott2017}\cite{endicott2017}} \tablefootnote{\label{karasiak2021}\cite{karasiak2021}} \tablefootnote{\label{moller2021}\cite{moller2021}}}  & \specialcell{2 \\ 3}   \\
+      \multirow{2}{*}{Spatial Leave-one-out}        &  \texttt{"spcv\_buffer"} \tablefootnote{\label{ploton2020}\cite{ploton2020}} \tablefootnote{\label{diesing2020}\cite{diesing2020}} & \pkg{blockCV}  & buffer, LOO & 2   \\
       \cmidrule{2-5}
-                                      & Spatial Blocking            & \specialcell{\pkg{blockCV}  \\ \pkg{sperrorest}}  & \specialcell{  \texttt{"spcv\_block"} \tablefootnote{\label{jensen2021}\cite{jensen2021}} \tablefootnote{\label{escobar2021}\cite{escobar2021}} \tablefootnote{\label{stewart2021}\cite{stewart2021}} \\  \texttt{"spcv\_tiles"} \tablefootnote{\label{bebber2017}\cite{bebber2017}} \tablefootnote{\label{brenning2015}\cite{brenning2015}} \tablefootnote{\label{zurell2020}\cite{zurell2020}} }
-                                        & \specialcell{28 \\ 4} \\
+       & \texttt{"spcv\_disc"} \tablefootnote{\label{endicott2017}\cite{endicott2017}} \tablefootnote{\label{karasiak2021}\cite{karasiak2021}} \tablefootnote{\label{moller2021}\cite{moller2021}} & \pkg{sperrorest}      & buffer, multiple observations  & 3   \\
+      \midrule
+      \multirow{4}{*}{Leave-one-block-out CV} & \texttt{"spcv\_coords"}  \tablefootnote{\label{morera2021}\cite{morera2021}} \tablefootnote{\label{geiss2017}\cite{geiss2017}} \tablefootnote{\label{wu2020}\cite{wu2020}} & \pkg{sperrorest}       &  clustering & 6 \\
             \cmidrule{2-5}
-                                      & Spatial Clustering          & \pkg{sperrorest} & \texttt{"spcv\_coords"} \tablefootnote{\label{morera2021}\cite{morera2021}} \tablefootnote{\label{geiss2017}\cite{geiss2017}} \tablefootnote{\label{wu2020}\cite{wu2020}} &  21 \\
+                                      & \texttt{"spcv\_tiles"} \tablefootnote{\label{bebber2017}\cite{bebber2017}} \tablefootnote{\label{brenning2015}\cite{brenning2015}} \tablefootnote{\label{zurell2020}\cite{zurell2020}} & \pkg{sperrorest}    &  geometric &  4  \\
                                                   \cmidrule{2-5}
-       & Group-Level CV & \specialcell{\pkg{CAST} \\ \pkg{mlr3}}       & \specialcell{ \texttt{"sptcv\_cstf"} (LLO) \tablefootnote{\label{blothe2021}\cite{blothe2021}} \tablefootnote{\label{dolan2021}\cite{dolan2021}} \tablefootnote{\label{dumke2019}\cite{dumke2019}}  \\ \texttt{"custom\_cv"} \tablefootnote{\label{brenning2006}\cite{brenning2006}} \tablefootnote{\label{pena2015}\cite{pena2015}}} & \specialcell{12 \\ 2}       \\
+                                      &  \texttt{"custom\_cv"}           & \pkg{mlr3} & custom  & - \\
       \midrule
-      \multirow{2}{*}{Spatiotemporal} & Leave-Location-and-Time-Out & \pkg{CAST}       &  \texttt{"sptcv\_cstf"} \tablefootnote{\label{egli2020}\cite{egli2020}} \tablefootnote{\label{gao2019}\cite{gao2019}} \tablefootnote{\label{reitz2021}\cite{reitz2021}} & 6 \\
-                                      & Spatiotemporal Clustering   & \pkg{skmeans}    &  \texttt{"sptcv\_cluto"} &  -  \\
+      \multirow{4}{*}{CV at the block level} & \texttt{"spcv\_block"}  \tablefootnote{\label{jensen2021}\cite{jensen2021}} \tablefootnote{\label{escobar2021}\cite{escobar2021}} \tablefootnote{\label{stewart2021}\cite{stewart2021}} & \pkg{blockCV}       &  geometric & 28 \\
+  \cmidrule{2-5}
+                                      & \texttt{"cv"}  & \pkg{mlr3}    &  custom, honors grouping &  -  \\
+  \cmidrule{2-5}
+                                      & \texttt{"spcv\_env"} \footref{morera2021}  & \pkg{blockCV}    &  clustering, feature space &  1  \\
       \midrule
-      Feature space                   & Environmental Blocking      & \pkg{blockCV}    &  \texttt{"spcv\_env"} \footref{morera2021}   & 1    \\
-
+      \multirow{3}{*}{CV for spatiotemporal data} & \texttt{"sptcv\_cstf"}  \tablefootnote{\label{egli2020}\cite{egli2020}} \tablefootnote{\label{gao2019}\cite{gao2019}} \tablefootnote{\label{reitz2021}\cite{reitz2021}} & \pkg{CAST}       &  custom & 6 \\
+                                                  \cmidrule{2-5}
+                                      & \texttt{"sptcv\_cluto"}  & \pkg{skmeans}    &  clustering & - \\
+                                      \bottomrule
     \end{tabular}
@@
 
-## Spatial
+On the most upper level, resampling methods are categorized through the way the fold partitioning is done:
 
-Spatial partitioning methods operate in a two-dimensional space and split observations into training and test set by their spatial location.
-Methods such as grid-based partitioning or clustering are used to create groups of observations which may then be aggregated further to form the final partition sets.
+- Leaving out single pixels (Spatial Leave-one-out) per fold with an optional buffer
+- Leaving out single blocks (Leave-one-block-out CV) per fold with an optional buffer
+- Leaving out multiple blocks (CV at the block level) per fold
 
-Spatial partitioning methods can be roughly divided into three groups: (1) buffering, (2) blocking and (3) clustering methods (Table&nbsp;\@ref(tab:sptcv-methods)).
-If a method could have been placed into multiple categories, we deliberately assigned the group which matched best the main characteristic of the respective method.
+A finer distinction can then be made by looking at how these groups are formed by the respective individual method:
 
-### Buffering
+- Using a geometric-based approach (rectangular or circular)
+- Using an unsupervised clustering approach
+- Using a custom/external input, i.e. specifying the groups via an external grouping variable
 
+The terminology "block" used in this categorization and in the manuscript in general refers to any kind of regular and irregularly shaped grouping of observations and not necessarily only to a rectangular partitioning as used in some methods.
+
+<!-- explain and introduce buffering -->
+The use of a buffer is not a direct grouping characteristic of a partitioning method in our view but only an optional add-on which could potentially be used by any method.
 "Buffering" refers to the idea of removing observations located between test and training sets by drawing a circular buffer around one or more observations [@brenning2012; @rest2014; @blockCV].
 Within this buffer zone, all observations are removed from the training set, thereby reducing the similarity between training and test set.
-
 Buffering is mostly applied in the coordinate space.
 However, there is no practical limitation that would prevent this method to be applied in a 2D + time setting.
-Since \pkg{mlr3spatiotempcv} supports two spatial-only "buffering" implementations, (1) "Spatial buffering" from package \pkg{blockCV} and (2) "Leave-one-disc-out cross-validation" from \pkg{sperrorest}, buffering was attributed to the category of spatial partitioning methods (Table&nbsp;\@ref(tab:sptcv-methods)).
+Since \pkg{mlr3spatiotempcv} supports two spatial-only "buffering" implementations, (1) `"spcv\_buffer"` from package \pkg{blockCV} and (2) `"spcv\_disc"` from \pkg{sperrorest}, buffering was attributed to the category of spatial partitioning methods (Table&nbsp;\@ref(tab:sptcv-methods)).
 
-#### Leave-one-disc-out --- sperrorest
+<!-- FIXME: graphic here -->
 
-Leave-one-disc-out resampling from package \pkg{sperrorest} randomly selects a circular test set and optionally excludes a buffer zone from the remaining training data.
-It thus ensures that a minimum separation distance between training and test data is maintained.
-Discs are centered at sample locations, their number being specified by the `folds` argument.
-It is therefore possible to sample up to $n$ discs without replacement (default) or any number when sampling with replacement (`replace = TRUE`).
-The number of discs is specified via the `folds` argument.
+## Spatial Leave-one-out
 
-This resampling method is, strictly speaking, not a partitioning method since the discs are not a disjoint.
-Nevertheless, it is a logical extension of spatial LOO CV.
-Depending on the size of the discs, potentially a large number of discs (`folds` argument) is needed to effectively use the entire dataset.
+Spatial Leave-one-out methods start from a single observation in space and create radial (buffer) zones around which define the partitions.
 
-Leave-one-disc-out resampling becomes spatial LOO CV for a radius of 0 m and when each observation is at a unique location.
+### Spatial Leave-one-out with buffer --- `"spcv_buffer"`
 
-<!-- AB: TODO: folds = 5 makes very little sense in the case of LODO-CV, since this would only sample 5 discs, which cover only a small fraction of the study area.
-Let's use folds = 100, this will at least make readers think about their choice of #folds.
--->
-```{r disc, fig.cap='Visualization of one training set / test set combination generated with the leave-one-disc-out method from package sperrorest (method "spcv\\_disc" in mlr3spatiotempcv).
-The disc has a radius of 300 m and is surrounded by a 400-m buffer.', out.width="40%"}
-library("mlr3")
-library("mlr3spatiotempcv")
-task = tsk("ecuador")
-resampling_disc = rsmp("spcv_disc", folds = 100, radius = 300L, buffer = 400L)
-resampling_disc$instantiate(task)
-
-autoplot(resampling_disc,
-  size = 0.8, task = task, fold_id = 1) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
-```
-
-#### Spatial buffering --- blockCV
-
-The "buffering" method from the \pkg{blockCV} package is conceptually related to spatial leave-one-out (LOO) cross-validation.
+The "buffering" method from the \pkg{blockCV} package is conceptually related to spatial leave-one-out cross-validation.
 It comes with optional support for presence/absence data handling, as often used in the field of species distribution modeling [@hijmans2013].
@@
 
-<!-- TODO: Discuss issues setting a sensible range value -->
+### Leave-one-disc-out with optional buffer --- `"spcv_disc"`
 
-### Spatial blocking
+Leave-one-disc-out resampling from package \pkg{sperrorest} randomly selects a circular test set starting from a single observations and optionally excludes a buffer zone from the remaining training data.
+It thus ensures that a minimum separation distance between training and test data is maintained.
+Discs are centered at sample locations, their number being specified by the `folds` argument.
+It is therefore possible to sample up to $n$ discs without replacement (default) or any number when sampling with replacement (`replace = TRUE`).
+The number of discs is specified via the `folds` argument.
 
-Spatial blocking divides the study area into different zones [russ2010; @bahn2012; @wenger2012; @blockCV].
-Several methods have been implemented to create rectangular as well as irregularly shaped blocks. 
-These blocks can be used for leave-one-block-out resampling, or they can themselves be grouped into $k$ partitions for a $k$-fold CV at the block level.
+This resampling method is, strictly speaking, not a partitioning method since the discs are not resulting in disjoint partitions - partitions will potentially overlap in subsequent folds.
+Nevertheless, it is a logical extension of spatial LOO CV.
+Depending on the size of the discs, potentially a large number of discs (`folds` argument) is needed to effectively use the entire dataset.
 
-\pkg{mlr3spatiotempcv} supports multiple resampling methods that are based on blocks: 
-(1) "Spatial blocking" from package \pkg{blockCV} implements $k$-fold CV resampling at the level of square or rectangular blocks, while (2) "Spatial tiles" from \pkg{sperrorest} offers leave-one-block-out resampling with square or rectangular, possibly rotated blocks.
-In addition, irregularly shaped blocks are available (3) through coordinate-based clustering as a leave-one-block-out technique from \pkg{sperrorest} (method `"spcv_coords"`, Section&nbsp;\@ref(sec:cluster-cv)), and (4) as a CV at the block level where blocks are defined by a factor variable that defines zones or groups (method `"custom_cv"` in \pkg{mlr3}, Section&nbsp;\@ref(sec:custom-cv)).
+Even though the buffer creates a circular area which defines the test partition, it is not categorized as a "geometric" method as the origin of the circle is observation-based.
+In contrast, "geometric" methods do not take into account single observation locations but only focus on the study area extend and will include all observations which fall into their respective covered area.
 
-#### Square and rectangular blocks --- blockCV
+Leave-one-disc-out resampling becomes spatial LOO CV for a radius of 0 m and when each observation is at a unique location.
 
-Block-level resampling from package \pkg{blockCV} supports both random and systematic resampling of square blocks (`method = "spcv_block"` with argument `selection = "random"` and `"systematic"`, respectively; Figure&nbpsp;~\@ref(fig:block-random) and \@ref(fig:block-systematic)).
-The implementation in \pkg{blockCV} has specific options that are designed for modeling presence-only data, which is a typical situation in species distribution modeling.
+<!-- AB: TODO: folds = 5 makes very little sense in the case of LODO-CV, since this would only sample 5 discs, which cover only a small fraction of the study area.
+Let's use folds = 100, this will at least make readers think about their choice of #folds.
+-->
+```{r disc, fig.cap='Visualization of one training set / test set combination generated with the leave-one-disc-out method from package sperrorest (method "spcv\\_disc" in mlr3spatiotempcv).
+The disc has a radius of 300 m and is surrounded by a 400-m buffer.', out.width="40%"}
+library("mlr3")
+library("mlr3spatiotempcv")
+task = tsk("ecuador")
+resampling_disc = rsmp("spcv_disc", folds = 100, radius = 300L, buffer = 400L)
+resampling_disc$instantiate(task)
+
+autoplot(resampling_disc,
+  size = 0.8, task = task, fold_id = 1) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```
+
+## Leave-one-block-out Cross-Validation
+
+Leave-one-block-out CV methods partition the dataset in a way that each of the resulting partitions will be used as a fold in the CV.
+
+### Clustering-based: Using coordinate-based clustering - `"spcv_coords"`
+
+Cluster analysis provides a flexible approach to creating irregularly shaped spatial blocks for spatial resampling.
+Numerous techniques are available that can potentially be applied to the spatial coordinates of observations, to the features, or to a combination of both.
+In spatial model assessment, the focus has been on coordinate-based clustering, and specifically on leave-one-block-out resampling with blocks created by $k$-means clustering of the coordinates [@russ2010].
+
+Coordinate-based clustering for spatial CV [@russ2010; @brenning2012] as implemented in package \pkg{sperrorest} uses the coordinates of all observations to create clusters in the spatial domain with the help of the $k$-means clustering algorithm.
+This can be regarded as a leave-one-block-out resampling method, or as a $k$-fold CV in which each test set is a spatial cluster.
+This method is referred to as `"spcv_coords"` in \pkg{mlr3spatiotempcv}.
+
+Coordinate-based clustering approach is very versatile as it adapts to irregularly-shaped study areas and ensures that exactly $k$ partitions are created, which are usually or very similar size when the sample locations are spread out evenly.
+Nevertheless, despite the random selection of initial cluster centers, repeated partitionings may in some cases be nearly identical.
+Also, $k$-means clustering may be less suitable for data sets with pre-existing clusters of points and/or with distant isolated sample locations.
+When distinct clusters of points are present, as in multi-level sampling, it may be better to define clusters using a factor variable (see method `"custom_cv"` in section \@ref(sec:custom-cv)).
+
+```{r coords, fig.cap='Leave-one-block-out CV based on $k$-means clustering of the coordinates as implemented in package sperrorest (method "spcv\\_coords" in mlr3spatiotempcv).', out.width="40%", fig.pos="ht"}
+resampling_coords = rsmp("spcv_coords", folds = 5)$instantiate(task)
+
+autoplot(resampling_coords,
+  size = 0.8, fold_id = 1, task = task) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```
+
+### Geometric: using rectangular blocks --- `"spcv_tiles"`
+
+Leave-one-tile-out resampling is implemented in the `"spcv_tiles"` method imported from package \pkg{sperrorest}.
+It uses rectangular blocks that can be rotated (argument `rotation`), and a minimum number or fraction of observations per block can optionally be achieved by iteratively merging small blocks into adjacent blocks  (argument `reassign` in conjunction with `min_n` or `min_frac`).
+Block size or number is specified via the argument `dsplit` or `nsplit`, respectively, and square blocks can be obtained with a single (or two identical) `dsplit` value.
+
+Note that the actual number of folds obtained may be smaller than `nsplit[1]*nsplit[2]` (or smaller than what would be expected based on `dsplit`) since some blocks may be empty or (optionally) merged into adjacent folds.
+In the example, there are only eleven folds instead of twelve because the southwestern part of the study area's bounding box does not contain observations  (Figure&nbsp;\@ref(fig:tile)).
+
+```{r tile, fig.cap='Leave-one-block-out resampling from package sperrorest (method "spcv\\_tiles" in package mlr3spatiotempcv with argument nsplit = c(3,4) indicating the number of rows and columns).', out.width="40%", fig.pos="ht"}
+resampling_tiles = rsmp("spcv_tiles", nsplit = c(3L, 4L))
+
+autoplot(resampling_tiles,
+  size = 0.8, fold_id = 1, task = task) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```
+
+### Custom: `"custom_cv"` in mlr3
+
+Support for user-defined partitioning strategies is built into \pkg{mlr3} directly.
+In this so-called "Custom CV", users are able to supply an external factor vector with the same length as number of observations.
+This partitioning concept should not be mistaken with "group-level resampling" (section \@ref(sec:group-level)), which conducts additional combination of data splits and performs CV at the block level.
+Each level of the factor vector will result in one partition.
+Alternatively to supplying an external factor vector, the name of a feature (type `character`) present within the task can be supplied via argument `col`.
+
+The following example was taken from `sperrorest::partition_factor()` because the resulting partitions separate nicely from each other in a visual way.
+
+<!-- TODO: AB: This diagram is inconsistent with the other diagrams - should display training and test samples. -->
+```{r custom-cv, fig.cap='Leave-one-level-out (custom) resampling from package mlr3 (method "custom\\_cv").
+A factor variable is used to define the partitions.
+Each color represents one partition.', out.width="40%", fig.pos="ht"}
+breaks = quantile(task$data()$dem, seq(0, 1, length = 6))
+zclass = cut(task$data()$dem, breaks, include.lowest = TRUE)
+
+resampling_custom = rsmp("custom_cv")
+resampling_custom$instantiate(task, f = zclass)
+
+autoplot(resampling_custom, size = 0.8, task = task) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```
+
+## Cross-Validation at the block level
+
+Methods which operate on the block level create dataset splits which are then combined to form the final partitions for the CV folds.
+In the special case of an equal amount of dataset splits and requested CV folds, these methods would perform a Leave-one-block-out CV.
+
+### Geometric: using rectangular blocks --- `"spcv_block"`
+
+The `"spcv_block"` method from package \pkg{blockCV} supports both random and systematic resampling of square blocks (`rsmp("spcv_block")` with argument `selection = "random"` and `"systematic"`, respectively; Figure&nbpsp;~\@ref(fig:block-random) and \@ref(fig:block-systematic)).
+This implementation comes with specific options that are designed for modeling presence-only data, which is a typical situation in species distribution modeling.
 Users can also supply a user-defined polygon via argument `rasterLayer` with predefined blocking zones.
@@
 
-#### Leave-one-tile-out resampling --- sperrorest
-
-Leave-one-tile-out resampling is implemented in the `"spcv_tiles"` method imported from package \pkg{sperrorest}.
-It uses rectangular blocks that can be rotated (argument `rotation`), and a minimum number or fraction of observations per block can optionally be achieved by iteratively merging small blocks into adjacent blocks  (argument `reassign` in conjunction with `min_n` or `min_frac`).
-Block size or number is specified via the argument `dsplit` or `nsplit`, respectively, and square blocks can be obtained with a single (or two identical) `dsplit` value.
-
-Note that the actual number of folds obtained may be smaller than `nsplit[1]*nsplit[2]` (or smaller than what would be expected based on `dsplit`) since some blocks may be empty or (optionally) merged into adjacent folds.
-In the example, there are only eleven folds instead of twelve because the southwestern part of the study area's bounding box does not contain observations  (Figure&nbsp;\@ref(fig:tile)).
-
-```{r tile, fig.cap='Leave-one-block-out resampling from package sperrorest (method "spcv\\_tiles" in package mlr3spatiotempcv with argument nsplit = c(3,4) indicating the number of rows and columns).', out.width="40%", fig.pos="ht"}
-resampling_tiles = rsmp("spcv_tiles", nsplit = c(3L, 4L))
-
-autoplot(resampling_tiles,
-  size = 0.8, fold_id = 1, task = task) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
-```
-
-### Clustering-based resampling {#sec:cluster-cv}
-
-Cluster analysis provides a flexible approach to creating irregularly shaped spatial blocks for spatial resampling.
-Numerous techniques are available that can potentially be applied to the spatial coordinates of observations, to the features, or to a combination of both.
-In spatial model assessment, the focus has been on coordinate-based clustering, and specifically on leave-one-block-out resampling with blocks created by $k$-means clustering of the coordinates [@russ2010].
-
-#### Leave-one-block-out resampling using coordinate-based clustering --- sperrorest
-
-Coordinate-based clustering for spatial CV [@russ2010; @brenning2012] as implemented in package \pkg{sperrorest} uses the coordinates of all observations to create clusters in the spatial domain with the help of the $k$-means clustering algorithm.
-This can be regarded as a leave-one-block-out resampling method, or as a $k$-fold CV in which each test set is a spatial cluster.
-This method is referred to as `spcv_coords` in \pkg{mlr3spatiotempcv}.
-
-Coordinate-based clustering approach is very versatile as it adapts to irregularly-shaped study areas and ensures that exactly $k$ partitions are created, which are usually or very similar size when the sample locations are spread out evenly.
-Nevertheless, despite the random selection of initial cluster centers, repeated partitionings may in some cases be nearly identical.
-Also, $k$-means clustering may be less suitable for data sets with pre-existing clusters of points and/or with distant isolated sample locations.
-When distinct clusters of points are present, as in multi-level sampling, it may be better to define clusters using a factor variable (see method `"custom_cv"` in Section \@ref(sec:custom-cv)).
-
-```{r coords, fig.cap='Leave-one-block-out CV based on $k$-means clustering of the coordinates as implemented in package sperrorest (method "spcv\\_coords" in mlr3spatiotempcv).', out.width="40%", fig.pos="ht"}
-resampling_coords = rsmp("spcv_coords", folds = 5)$instantiate(task)
-
-autoplot(resampling_coords,
-  size = 0.8, fold_id = 1, task = task) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
-```
-
-## Spatial, temporal or spatiotemporal
-
-Some methods have the ability to operate in multiple spaces, i.e. spatial, temporal or spatiotemporal (2D + time).
-To avoid redundancy and putting too much weight on certain methods, only the spatiotemporal examples of these methods will be shown in the following sections.
-For the respective other spaces, usually only either the space or time coordinates need to be omitted from the user input.
-
-### Group-level resampling {#sec:group-level}
+### Custom: `"cv"` in mlr3 {#sec:group-level}
 
@@
 
-#### Leave-Location-And-Time-Out --- CAST
-
-Leave-location-and-time-out (LLTO) from package \pkg{CAST} is an implementation of spatiotemporal resampling method [@meyer2018].
-It requires both a spatial and a temporal variable which are used to separate observations in a multi-dimensional space.
-It comes with satellite methods named "leave-time-out" (LTO) and "leave-location-out" (LLO) which only operate in a two-dimensional space.
-
-In this example the `cookfarm` dataset is used because it contains the variable ("Date") which serves as the temporal component in the dataset.
-
-Two visualization types are supported by `mlr3spatiotempcv::autoplot()` for this method: a two-dimensional and a three-dimensional one which can be selected via the logical argument `plot3D`.
-The 3D visualization (i.e. 2D + time) is done via package \pkg{plotly}.
-Because a dynamic image cannot be included in this manuscript, a static version, which can be generated by setting `static_image = TRUE`, is shown.
-
-In \pkg{mlr3spatiotempcv} the method is named `"sptcv_cstf"` which is an acronym originating from the underlying upstream implementation in package \pkg{CAST} `CreateSpacetimeFolds()`.
-
-##### Leave-Time-Out (LTO)
-
-To better showcase the partitioning of the "Leave-Time-Out" (LTO) method, the temporal variable `Date` was adjusted in this example.
-
-The following code uses the `cookfarm` dataset as the base, modifies the `Date` variable in such a way that it contains five unique levels and then creates a spatiotemporal regression task in \pkg{mlr3spatiotempcv} (Figure&nbsp;\@ref(fig:lto)).
-
-```{r, results='hide'}
-data = cookfarm_sample
-data$Date = rep(c(
-  "2020-01-01", "2020-02-01", "2020-03-01", "2020-04-01",
-  "2020-05-01"), times = 1, each = 100)
-b = mlr3::as_data_backend(data)
-task_spt = TaskRegrST$new(
-  id = "cookfarm", b, target = "PHIHOX",
-  extra_args = list(
-    coordinate_names = c("x", "y"), coords_as_features = FALSE,
-    crs = 26911)
-)
-
-resampling_cstf_time = rsmp("sptcv_cstf", folds = 5, time_var = "Date")
-resampling_cstf_time$instantiate(task_spt)
-
-autoplot(resampling_cstf_time,
-  fold_id = 5, task = task_spt, plot3D = TRUE
-)
-```
-
-```{r lto, echo=FALSE, out.width="70%", fig.cap='Perspective plot of leave-time-out CV from package CAST (method "sptcv\\_cstf" and argument time\\_var = "Date").
-Only five folds and five time points were used in this example.
-Note that the blue dots correspond to five discrete time levels, which appear as a point cloud due to the viewing angle.', fig.pos="ht"}
-knitr::include_graphics("lto.png")
-```
-
-##### Leave-Location-Out (LLO)
-
-Method "Leave-Location-Out" (LLO) works similar to "Leave-Time-Out" (LTO) in that it operates in a two-dimensional space.
-Here, specific locations are left out individually for forming test partitions while the temporal information is ignored (Figure&nbsp;\@ref(fig:llo)).
-
-```{r results='hide'}
-resampling_cstf_loc = rsmp("sptcv_cstf", folds = 5, space_var = "SOURCEID")
-resampling_cstf_loc$instantiate(task_spt)
-
-autoplot(resampling_cstf_loc,
-  fold_id = 5, task = task_spt, plot3D = TRUE)
-```
-
-```{r llo, echo=FALSE, out.width="70%", fig.cap='Perspective plot of leave-location-out CV from package CAST (method "sptcv\\_cstf" and argument space\\_var = "SOURCEID").
-The data is viewed from the future towards the past.', fig.pos="ht"}
-knitr::include_graphics("llo.png")
-```
-
-##### Leave-Location-and-Time-Out (LLTO)
-
-The "Leave-Location-and-Time-Out" (LLTO) method combines both "Leave-Time-Out" (LTO) and "Leave-Location-Out" (LLO).
-It creates partitions that operate on spatiotemporal clusters to reduce both spatial and temporal autocorrelation.
-If there are multiple observations over time per location, and the modeling purpose is to reduce the bias in the predictions as potentially introduced by spatial as well as temporal autocorrelation, LLTO might be an appropriate choice.
-To reduce autocorrelation in both dimensions, the methods omits observations as the splitting in test and train sets only provides separation in one dimension whereas two (space and time) are needed to account for both types, spatial and temporal autocorrelation.
-In other words, accounting for one dimension as in Figure&nbsp;\@ref(fig:lto) or Figure&nbsp;\@ref(fig:llo), by only having one time step or location in either test or training set, would have no effect on the respective other.
-Thus, the removal of observation can be seen as a proxied buffering in multiple dimensions as shown in Figure&nbsp;\@ref(fig:llto).
-
-```{r results='hide'}
-resampling_cstf_time_loc = rsmp("sptcv_cstf",
-  folds = 5,
-  space_var = "SOURCEID", time_var = "Date")
-resampling_cstf_time_loc$instantiate(task_spt)
-
-autoplot(resampling_cstf_time_loc,
-  fold_id = 4, task = task_spt, plot3D = TRUE,
-  show_omitted = TRUE)
-```
-
-```{r llto, echo=FALSE, out.width="70%", fig.cap='Perspective plot of leave-location-and-time-out CV from package CAST (method "sptcv\\_cstf" and arguments time\\_var = "Date" and space\\_var = "SOURCEID").
-The grey points are excluded from both the training and the test set in this fold.', fig.pos="ht"}
-knitr::include_graphics("llto.png")
-```
-
-### Custom and grouped resampling --- mlr3 {#sec:custom-cv}
-
-Support for user-defined partitioning strategies is built into \pkg{mlr3} directly.
-In this so-called "Custom CV", users are able to supply an external factor vector with the same length as number of observations.
-This partitioning concept is also often referred to as "grouped resampling" and should not be mistaken with "group-level resampling" (section \@ref(sec:group-level)), which inherits additional aggregation of groups at the partition level.
-Each level of the factor vector will result in one partition.
-Alternatively to supplying an external factor vector, the name of a feature (type `character`) present within the task can be supplied via argument `col`.
-
-The following example was taken from `sperrorest::partition_factor()` because the resulting partitions separate nicely from each other in a visual way.
-
-<!-- TODO: AB: This diagram is inconsistent with the other diagrams - should display training and test samples. -->
-```{r custom-cv, fig.cap='Leave-one-level-out (custom) resampling from package mlr3 (method "custom\\_cv").
-A factor variable is used to define the partitions.
-Each color represents one partition.', out.width="40%", fig.pos="ht"}
-breaks = quantile(task$data()$dem, seq(0, 1, length = 6))
-zclass = cut(task$data()$dem, breaks, include.lowest = TRUE)
-
-resampling_custom = rsmp("custom_cv")
-resampling_custom$instantiate(task, f = zclass)
-
-autoplot(resampling_custom, size = 0.8, task = task) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
-
-```
-
-## Feature space clustering
+### Clustering: using feature-based clustering --- `"spcv_env"`
 
@@
 
-## Additional partitioning techniques
+## Cross-Validation for spatiotemporal data
+
+Some methods have the ability to operate in multiple spaces, i.e. spatial, temporal or spatiotemporal (2D + time).
+To avoid redundancy and putting too much weight on certain methods, only the spatiotemporal examples of these methods will be shown in the following sections.
+For the respective other spaces, usually only either the space or time coordinates need to be omitted from the user input.
+
+### Custom: "Leave-location-and-time-out" --- `"spcv_cstf"`
+
+Leave-location-and-time-out (LLTO) from package \pkg{CAST} is an implementation of spatiotemporal resampling method [@meyer2018].
+It requires both a spatial and a temporal variable which are used to separate observations in a multi-dimensional space.
+It comes with satellite methods named "leave-time-out" (LTO) and "leave-location-out" (LLO) which only operate in a two-dimensional space.
+
+In this example the `cookfarm` dataset is used because it contains the variable ("Date") which serves as the temporal component in the dataset.
+
+The sub-methods "Leave-Time-Out" (LTO) and "Leave-Location-Out" (LLO) are conceptually similar to the "cv" method with a custom grouping from \pkg{mlr3} as they perform a CV at the block level using a predefined grouping structure.
+
+Two visualization types are supported by `mlr3spatiotempcv::autoplot()` for this method: a two-dimensional and a three-dimensional one which can be selected via the logical argument `plot3D`.
+The 3D visualization (i.e. 2D + time) is done via package \pkg{plotly}.
+Because a dynamic image cannot be included in this manuscript, a static version, which can be generated by setting `static_image = TRUE`, is shown.
+
+In \pkg{mlr3spatiotempcv} the method is named `"sptcv_cstf"` which is an acronym originating from the underlying upstream implementation in package \pkg{CAST} `CreateSpacetimeFolds()`.
+
+#### Leave-Time-Out (LTO)
+
+To better showcase the partitioning of the "Leave-Time-Out" (LTO) method, the temporal variable `Date` was adjusted in this example.
+
+The following code uses the `cookfarm` dataset as the base, modifies the `Date` variable in such a way that it contains five unique levels and then creates a spatiotemporal regression task in \pkg{mlr3spatiotempcv} (Figure&nbsp;\@ref(fig:lto)).
+
+```{r, results='hide'}
+data = cookfarm_sample
+data$Date = rep(c(
+  "2020-01-01", "2020-02-01", "2020-03-01", "2020-04-01",
+  "2020-05-01"), times = 1, each = 100)
+b = mlr3::as_data_backend(data)
+task_spt = TaskRegrST$new(
+  id = "cookfarm", b, target = "PHIHOX",
+  extra_args = list(
+    coordinate_names = c("x", "y"), coords_as_features = FALSE,
+    crs = 26911)
+)
+
+resampling_cstf_time = rsmp("sptcv_cstf", folds = 5, time_var = "Date")
+resampling_cstf_time$instantiate(task_spt)
+
+autoplot(resampling_cstf_time,
+  fold_id = 5, task = task_spt, plot3D = TRUE
+)
+```
+
+```{r lto, echo=FALSE, out.width="70%", fig.cap='Perspective plot of leave-time-out CV from package CAST (method "sptcv\\_cstf" and argument time\\_var = "Date").
+Only five folds and five time points were used in this example.
+Note that the blue dots correspond to five discrete time levels, which appear as a point cloud due to the viewing angle.', fig.pos="ht"}
+knitr::include_graphics("lto.png")
+```
+
+#### Leave-Location-Out (LLO)
+
+Method "Leave-Location-Out" (LLO) works similar to "Leave-Time-Out" (LTO) in that it operates in a two-dimensional space.
+Here, specific locations are left out individually for forming test partitions while the temporal information is ignored (Figure&nbsp;\@ref(fig:llo)).
+
+```{r results='hide'}
+resampling_cstf_loc = rsmp("sptcv_cstf", folds = 5, space_var = "SOURCEID")
+resampling_cstf_loc$instantiate(task_spt)
+
+autoplot(resampling_cstf_loc,
+  fold_id = 5, task = task_spt, plot3D = TRUE)
+```
+
+```{r llo, echo=FALSE, out.width="70%", fig.cap='Perspective plot of leave-location-out CV from package CAST (method "sptcv\\_cstf" and argument space\\_var = "SOURCEID").
+The data is viewed from the future towards the past.', fig.pos="ht"}
+knitr::include_graphics("llo.png")
+```
+
+#### Leave-Location-and-Time-Out (LLTO)
+
+The "Leave-Location-and-Time-Out" (LLTO) method combines both "Leave-Time-Out" (LTO) and "Leave-Location-Out" (LLO).
+It creates partitions that operate on spatiotemporal clusters to reduce both spatial and temporal autocorrelation.
+If there are multiple observations over time per location, and the modeling purpose is to reduce the bias in the predictions as potentially introduced by spatial as well as temporal autocorrelation, LLTO might be an appropriate choice.
+To reduce autocorrelation in both dimensions, the methods omits observations as the splitting in test and train sets only provides separation in one dimension whereas two (space and time) are needed to account for both types, spatial and temporal autocorrelation.
+In other words, accounting for one dimension as in Figure&nbsp;\@ref(fig:lto) or Figure&nbsp;\@ref(fig:llo), by only having one time step or location in either test or training set, would have no effect on the respective other.
+Thus, the removal of observation can be seen as a proxied buffering in multiple dimensions as shown in Figure&nbsp;\@ref(fig:llto).
+
+```{r results='hide'}
+resampling_cstf_time_loc = rsmp("sptcv_cstf",
+  folds = 5,
+  space_var = "SOURCEID", time_var = "Date")
+resampling_cstf_time_loc$instantiate(task_spt)
+
+autoplot(resampling_cstf_time_loc,
+  fold_id = 4, task = task_spt, plot3D = TRUE,
+  show_omitted = TRUE)
+```
+
+```{r llto, echo=FALSE, out.width="70%", fig.cap='Perspective plot of leave-location-and-time-out CV from package CAST (method "sptcv\\_cstf" and arguments time\\_var = "Date" and space\\_var = "SOURCEID").
+The grey points are excluded from both the training and the test set in this fold.', fig.pos="ht"}
+knitr::include_graphics("llto.png")
+```
+
+### Clustering: using CLUTO --- `"sptcv_cluto"`
 
@@
 With respect to functionality, \pkg{mlr3spatiotempcv} supports two-dimensional clustering based on coordinates, time or features and three-dimensional clustering on coordinates and time.
+
+<!-- Spatial blocking divides the study area into different zones [russ2010; @bahn2012; @wenger2012; @blockCV]. -->
+<!-- Several methods have been implemented to create rectangular as well as irregularly shaped blocks.  -->
+<!-- These blocks can be used for leave-one-block-out resampling, or they can themselves be grouped into $k$ partitions for a $k$-fold CV at the block level. -->
+
+<!-- \pkg{mlr3spatiotempcv} supports multiple resampling methods that are based on blocks:  -->
+<!-- (1) "Spatial blocking" from package \pkg{blockCV} implements $k$-fold CV resampling at the level of square or rectangular blocks, while (2) "Spatial tiles" from \pkg{sperrorest} offers leave-one-block-out resampling with square or rectangular, possibly rotated blocks.
+-->
+<!-- In addition, irregularly shaped blocks are available (3) through coordinate-based clustering as a leave-one-block-out technique from \pkg{sperrorest} (method `"spcv_coords"`, Section&nbsp;\@ref(sec:cluster-cv)), and (4) as a CV at the block level where blocks are defined by a factor variable that defines zones or groups (method `"custom_cv"` in \pkg{mlr3}, Section&nbsp;\@ref(sec:custom-cv)).
+-->
+
+### Clustering-based resampling {#sec:cluster-cv}
 

### 2021-09-18 paper/paper.Rmd
remove `scale_*_continuous()` lines for autoplot() parts

@@
 
-```{r buffer, fig.cap='Visualization of the spatial buffering method from package blockCV (method "spcv\\_buffer" in  mlr3spatiotempcv). The buffer distance is 1000 m.', out.width="40%", fig.pos="ht"}
-resampling_buffer = rsmp("spcv_buffer", theRange = 1000)
-resampling_buffer$instantiate(task)
+```{r buffer-non-eval, eval=FALSE}
+library("mlr3")
+library("mlr3spatiotempcv")
+task = tsk("ecuador")
+rsmp_buffer = rsmp("spcv_buffer", theRange = 1000)
+rsmp_buffer$instantiate(task)
 
-autoplot(resampling_buffer,
+autoplot(rsmp_buffer, size = 0.8, task = task, fold_id = 1)
+```
+
+```{r buffer-eval, fig.cap='Visualization of the spatial buffering method from package blockCV (method "spcv\\_buffer" in  mlr3spatiotempcv).
+The buffer distance is 1000 m.', out.width="40%", fig.pos="ht",echo=FALSE}
+library("mlr3")
+library("mlr3spatiotempcv")
+task = tsk("ecuador")
+rsmp_buffer = rsmp("spcv_buffer", theRange = 1000)
+rsmp_buffer$instantiate(task)
+
+autoplot(rsmp_buffer,
   size = 0.8, task = task, fold_id = 1) *
@@
 -->
-```{r disc, fig.cap='Visualization of one training set / test set combination generated with the leave-one-disc-out method from package sperrorest (method "spcv\\_disc" in mlr3spatiotempcv).
-The disc has a radius of 300 m and is surrounded by a 400-m buffer.', out.width="40%"}
+
+```{r disc-non-eval, eval=FALSE}
+rsmp_disc = rsmp("spcv_disc", folds = 100, radius = 300L, buffer = 400L)
+rsmp_disc$instantiate(task)
+
+autoplot(rsmp_disc,
+  size = 0.8, task = task, fold_id = 1)
+```
+
+```{r disc-eval, fig.cap='Visualization of one training set / test set combination generated with the leave-one-disc-out method from package sperrorest (method "spcv\\_disc" in mlr3spatiotempcv).
+The disc has a radius of 300 m and is surrounded by a 400-m buffer.', out.width="40%", echo=FALSE,fig.pos="hb"}
 library("mlr3")
@@
 task = tsk("ecuador")
-resampling_disc = rsmp("spcv_disc", folds = 100, radius = 300L, buffer = 400L)
-resampling_disc$instantiate(task)
+rsmp_disc = rsmp("spcv_disc", folds = 100, radius = 300L, buffer = 400L)
+rsmp_disc$instantiate(task)
 
-autoplot(resampling_disc,
+autoplot(rsmp_disc,
   size = 0.8, task = task, fold_id = 1) *
@@
 
-```{r coords, fig.cap='Leave-one-block-out CV based on $k$-means clustering of the coordinates as implemented in package sperrorest (method "spcv\\_coords" in mlr3spatiotempcv).', out.width="40%", fig.pos="ht"}
-resampling_coords = rsmp("spcv_coords", folds = 5)$instantiate(task)
+```{r coords-non-eval, eval=FALSE}
+rsmp_coords = rsmp("spcv_coords", folds = 5)$instantiate(task)
 
-autoplot(resampling_coords,
+autoplot(rsmp_coords, size = 0.8, fold_id = 1, task = task)
+```
+
+```{r coords-eval, fig.cap='Leave-one-block-out CV based on $k$-means clustering of the coordinates as implemented in package sperrorest (method "spcv\\_coords" in mlr3spatiotempcv).', out.width="40%", fig.pos="ht", echo=FALSE}
+rsmp_coords = rsmp("spcv_coords", folds = 5)$instantiate(task)
+
+autoplot(rsmp_coords,
   size = 0.8, fold_id = 1, task = task) *
@@
 
-```{r tile, fig.cap='Leave-one-block-out resampling from package sperrorest (method "spcv\\_tiles" in package mlr3spatiotempcv with argument nsplit = c(3,4) indicating the number of rows and columns).', out.width="40%", fig.pos="ht"}
-resampling_tiles = rsmp("spcv_tiles", nsplit = c(3L, 4L))
+```{r tile-non-eval, eval=FALSE}
+rsmp_tiles = rsmp("spcv_tiles", nsplit = c(3L, 4L))
 
-autoplot(resampling_tiles,
+autoplot(rsmp_tiles, size = 0.8, fold_id = 1, task = task)
+```
+
+```{r tile, fig.cap='Leave-one-block-out resampling from package sperrorest (method "spcv\\_tiles" in package mlr3spatiotempcv with argument nsplit = c(3,4) indicating the number of rows and columns).', out.width="40%", fig.pos="ht", echo=FALSE}
+rsmp_tiles = rsmp("spcv_tiles", nsplit = c(3L, 4L))
+
+autoplot(rsmp_tiles,
   size = 0.8, fold_id = 1, task = task) *
@@
 <!-- TODO: AB: This diagram is inconsistent with the other diagrams - should display training and test samples. -->
-```{r custom-cv, fig.cap='Leave-one-level-out (custom) resampling from package mlr3 (method "custom\\_cv").
-A factor variable is used to define the partitions.
-Each color represents one partition.', out.width="40%", fig.pos="ht"}
+```{r custom-cv-non-eval, eval=FALSE}
 breaks = quantile(task$data()$dem, seq(0, 1, length = 6))
@@
 
-resampling_custom = rsmp("custom_cv")
-resampling_custom$instantiate(task, f = zclass)
+rsmp_custom = rsmp("custom_cv")
+rsmp_custom$instantiate(task, f = zclass)
 
-autoplot(resampling_custom, size = 0.8, task = task) *
+autoplot(rsmp_custom, size = 0.8, task = task, fold_id = 1)
+```
+
+```{r custom-cv, fig.cap='Leave-one-level-out (custom) resampling from package mlr3 (method "custom\\_cv").
+A factor variable is used to define the partitions.
+Each color represents one partition.', out.width="40%", fig.pos="ht", echo=FALSE}
+breaks = quantile(task$data()$dem, seq(0, 1, length = 6))
+zclass = cut(task$data()$dem, breaks, include.lowest = TRUE)
+
+rsmp_custom = rsmp("custom_cv")
+rsmp_custom$instantiate(task, f = zclass)
+
+autoplot(rsmp_custom, size = 0.8, task = task, fold_id = 1) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
@@
 
+```{r block-random-non-eval, eval=FALSE}
+rsmp_block_random = rsmp("spcv_block", range = 1000, folds = 5)
+
+autoplot(rsmp_block_random, size = 0.8, fold_id = 1, task = task,
+  show_blocks = TRUE, show_labels = TRUE)
+```
+
 ```{r block-random, fig.cap='Random resampling of square spatial blocks using the implementation in package blockCV (method "spcv\\_block" with option selection = "random" in mlr3spatiotempcv).
-The size of the squares is 1000 m, and four out of the 19 blocks were assigned to this test sample.', out.width="40%", fig.pos="ht"}
-resampling_block_random = rsmp("spcv_block", range = 1000, folds = 5)
+The size of the squares is 1000 m, and four out of the 19 blocks were assigned to this test sample.', out.width="40%", fig.pos="ht", echo=FALSE}
+rsmp_block_random = rsmp("spcv_block", range = 1000, folds = 5)
 
-autoplot(resampling_block_random,
+autoplot(rsmp_block_random,
   size = 0.8, fold_id = 1, task = task,
@@
 
-```{r block-systematic, fig.cap='Sytematic resampling of square spatial blocks using the implementation in package blockCV (method "spcv\\_block" with option selection = "systematic" in mlr3spatiotempcv).
-The size of the squares is 1000 m, and four out of the 19 blocks were assigned to this test sample.', out.width="40%", fig.pos="ht"}
-resampling_block_systematic = rsmp("spcv_block",
+```{r block-systematic-non-eval, eval=FALSE}
+rsmp_block_systematic = rsmp("spcv_block",
   range = 1000, folds = 5,
@@
 
-autoplot(resampling_block_systematic,
+autoplot(rsmp_block_systematic, size = 0.8, fold_id = 1, task = task,
+  show_blocks = TRUE, show_labels = TRUE)
+```
+
+```{r block-systematic, fig.cap='Sytematic resampling of square spatial blocks using the implementation in package blockCV (method "spcv\\_block" with option selection = "systematic" in mlr3spatiotempcv).
+The size of the squares is 1000 m, and four out of the 19 blocks were assigned to this test sample.', out.width="40%", fig.pos="ht", echo=FALSE}
+rsmp_block_systematic = rsmp("spcv_block",
+  range = 1000, folds = 5,
+  selection = "systematic"
+)
+
+autoplot(rsmp_block_systematic,
   size = 0.8, fold_id = 1, task = task,
@@
 The size of the squares is 1000 m, and eleven out of the 19 blocks were assigned to this test sample.', out.width="40%", fig.pos="ht!"}
-resampling_block_checkerboard = rsmp("spcv_block",
+rsmp_block_checkerboard = rsmp("spcv_block",
   range = 1000, folds = 2,
@@
 
-autoplot(resampling_block_checkerboard,
+autoplot(rsmp_block_checkerboard,
   size = 0.8, fold_id = 1, task = task,
@@
 ```{r block-rows-and-cols, fig.cap='Random resampling of a specifid number of rectangular blocks as implemented in package blockCV (method "spcv\\_block" in mlr3spatiotempcv with arguments rows = 6 and cols = 3).', out.width="40%", fig.pos="ht"}
-resampling_block_rows_cols = rsmp("spcv_block",
+rsmp_block_rows_cols = rsmp("spcv_block",
   rows = 6, cols = 3, folds = 5
@@
 
-autoplot(resampling_block_rows_cols,
+autoplot(rsmp_block_rows_cols,
   size = 0.8, fold_id = 1, task = task,
@@
 
-```{r env-1, fig.cap='Environmental leave-one-block-out CV from package blockCV using a single predictor to define blocks in feature space (method "spcv\\_env" in mlr3spatiotempcv with argument features = "distdeforest").', out.width="40%", fig.pos="ht"}
-resampling_env = rsmp("spcv_env",
-  features = "distdeforest", folds = 5)
+```{r env-1-non-eval, eval=FALSE}
+rsmp_env = rsmp("spcv_env", features = "distdeforest", folds = 5)
 
-autoplot(resampling_env,
-  size = 0.8, fold_id = 1, task = task) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+rsmp_env_multi = rsmp("spcv_env", features = c("distdeforest", "slope"), folds = 5)
+
+plot_env_single = autoplot(rsmp_env, size = 0.8, fold_id = 1, task = task)
+
+plot_env_multi = autoplot(rsmp_env_multi, size = 0.8, fold_id = 1, task = task)
+
+library("patchwork")
+plot_env_single + plot_env_multi
 ```
@@
 <!-- TODO: AB: should be combined with the previous figure into a two-panel figure. Differences are much easier to recognize if plotted next to each other. -->
-```{r env-2, fig.cap='Environmental leave-one-block-out CV from package blockCV using two predictors to define blocks in feature space (method "spcv\\_env" in mlr3spatiotempcv with argument features = c("distdeforest", "slope")).', out.width="40%", fig.pos="ht"}
-resampling_env_multi = rsmp("spcv_env",
+```{r env-1-eval, fig.cap='Environmental leave-one-block-out CV from package blockCV using one (left, "distdeforest") and two (right, "distdeforest" and "slope") predictors to define blocks in the feature space.', out.width="100%", fig.pos="ht", echo=FALSE}
+rsmp_env = rsmp("spcv_env",
+  features = "distdeforest", folds = 5)
+
+rsmp_env_multi = rsmp("spcv_env",
   features = c("distdeforest", "slope"), folds = 5)
 
-autoplot(resampling_env_multi,
-  size = 0.8, fold_id = 1, task = task) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+plot_env_single = autoplot(rsmp_env,
+  size = 0.6, fold_id = 1, task = task) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02))
+
+plot_env_multi = autoplot(rsmp_env_multi,
+  size = 0.6, fold_id = 1, task = task) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02))
+
+library("patchwork")
+plot_env_single + plot_env_multi + plot_layout(guides = 'collect')
 ```
@@
 
-resampling_cstf_time = rsmp("sptcv_cstf", folds = 5, time_var = "Date")
-resampling_cstf_time$instantiate(task_spt)
+rsmp_cstf_time = rsmp("sptcv_cstf", folds = 5, time_var = "Date")
+rsmp_cstf_time$instantiate(task_spt)
 
-autoplot(resampling_cstf_time,
+autoplot(rsmp_cstf_time,
   fold_id = 5, task = task_spt, plot3D = TRUE
@@
 ```{r results='hide'}
-resampling_cstf_loc = rsmp("sptcv_cstf", folds = 5, space_var = "SOURCEID")
-resampling_cstf_loc$instantiate(task_spt)
+rsmp_cstf_loc = rsmp("sptcv_cstf", folds = 5, space_var = "SOURCEID")
+rsmp_cstf_loc$instantiate(task_spt)
 
-autoplot(resampling_cstf_loc,
+autoplot(rsmp_cstf_loc,
   fold_id = 5, task = task_spt, plot3D = TRUE)
@@
 ```{r results='hide'}
-resampling_cstf_time_loc = rsmp("sptcv_cstf",
+rsmp_cstf_time_loc = rsmp("sptcv_cstf",
   folds = 5,
   space_var = "SOURCEID", time_var = "Date")
-resampling_cstf_time_loc$instantiate(task_spt)
+rsmp_cstf_time_loc$instantiate(task_spt)
 
-autoplot(resampling_cstf_time_loc,
+autoplot(rsmp_cstf_time_loc,
   fold_id = 4, task = task_spt, plot3D = TRUE,
@@
 
-Next, the created resampling object `resampling_nsp` is passed to function `resample()` together with the task and learner objects created earlier to execute the model assessment.
+Next, the created resampling object `rsmp_nsp` is passed to function `resample()` together with the task and learner objects created earlier to execute the model assessment.
 This is the actual, potentially time-consuming CV estimation. With the present settings, eight random forest classifiers are fitted and evaluated in this step.
@@
 
-```{r rr-nspcv, cache.lazy = FALSE}
-resampling_nsp = rsmp("repeated_cv", folds = 4, repeats = 2)
-resampling_nsp$instantiate(task)
+```{r rr-nspcv-1, cache.lazy = FALSE, results='hide'}
+rsmp_nsp = rsmp("repeated_cv", folds = 4, repeats = 2)
+rsmp_nsp$instantiate(task)
 rr_nsp = resample(
   task = task, learner = learner,
-  resampling = resampling_nsp
+  resampling = rsmp_nsp
 )
+```
 
+```{r rr-nspcv-2, cache.lazy = FALSE}
 rr_nsp$aggregate(measures = msr("classif.auc"))
@@
 
-```{r rr-spcv, cache.lazy = FALSE}
-resampling_sp = rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
-resampling_sp$instantiate(task)
+```{r rr-spcv-1, cache.lazy = FALSE, results='hide'}
+rsmp_sp = rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
+rsmp_sp$instantiate(task)
 rr_sp = resample(
   task = task, learner = learner,
-  resampling = resampling_sp
+  resampling = rsmp_sp
 )
+```
 
+```{r rr-spcv-2, cache.lazy = FALSE}
 rr_sp$aggregate(measures = msr("classif.auc"))
@@
 
-```{r vis-spcv, fig.cap="Spatial leave-one-block-out partitioning using coordinate-based clustering to create roughly equally sized polygonal blocks.
-Due to space limitations only the first two folds of the first repetition are shown.", fig.pos="ht", fig.height=6, fig.width=8}
-autoplot(resampling_sp, task, fold_id = c(1:2), size = 0.8) *
+```{r vis-spcv-non-eval, eval=FALSE}
+autoplot(rsmp_sp, task, fold_id = c(1:2), size = 0.8)
+```
+
+```{r vis-spcv-eval, fig.cap="Spatial leave-one-block-out partitioning using coordinate-based clustering to create roughly equally sized polygonal blocks.
+Due to space limitations only the first two folds of the first repetition are shown.", fig.pos="ht", fig.height=6, fig.width=8, echo=FALSE}
+autoplot(rsmp_sp, task, fold_id = c(1:2), size = 0.8) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
@@
 
-```{r vis-nspcv, fig.cap="Random (non-spatial) four-fold CV partitioning. Only the first two folds of the first repetition are shown.", fig.pos="ht", fig.height=6, fig.width=8}
-autoplot(resampling_nsp, task, fold_id = c(1:2), size = 0.8) *
+```{r vis-nspcv-non-eval, eval=FALSE}
+autoplot(rsmp_nsp, task, fold_id = c(1:2), size = 0.8)
+```
+
+```{r vis-nspcv-eval, fig.cap="Random (non-spatial) four-fold CV partitioning. Only the first two folds of the first repetition are shown.", fig.pos="ht", fig.height=6, fig.width=8, echo=FALSE}
+autoplot(rsmp_nsp, task, fold_id = c(1:2), size = 0.8) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *

### 2021-09-18 paper/paper.Rmd
omit checkerboard plot, remove resolvec comments

@@
 
-<!--
-
-##### Option: selection = "checkerboard"
-
--->
-
 The *checkerboard* partitioning is a special case of a systematic block partitioning (`selection = "checkerboard"`).
 It inherently supports only two folds, making it less appealing than the more commonly used five- or ten-fold resampling, which achieve larger training set sizes.
-
-<!--
-AB:
-The text previously stated that the range argument is ignored. This can't be true, there must be some argument that controls the size of checkerboard fields.
--->
-
-<!--
-AB: Checkerboard pattern is just a special case of a systematic pattern and likely not of great practical relevance because there are only two folds -> omit figure.
-
-```{r block-checkerboard, fig.cap='Checkerboard resampling of square spatial blocks using the implementation in package blockCV (method "spcv\\_block" with option selection = "checkerboard" in mlr3spatiotempcv).
-The size of the squares is 1000 m, and eleven out of the 19 blocks were assigned to this test sample.', out.width="40%", fig.pos="ht!"}
-rsmp_block_checkerboard = rsmp("spcv_block",
-  range = 1000, folds = 2,
-  selection = "checkerboard")
-
-autoplot(rsmp_block_checkerboard,
-  size = 0.8, fold_id = 1, task = task,
-  show_blocks = TRUE, show_labels = TRUE) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
-```
-
--->
-
-<!--
-AB: This figure and subsubsubsection takes up a lot of space for very little additional information.
-
-```{r block-rows-and-cols, fig.cap='Random resampling of a specifid number of rectangular blocks as implemented in package blockCV (method "spcv\\_block" in mlr3spatiotempcv with arguments rows = 6 and cols = 3).', out.width="40%", fig.pos="ht"}
-rsmp_block_rows_cols = rsmp("spcv_block",
-  rows = 6, cols = 3, folds = 5
-)
-
-autoplot(rsmp_block_rows_cols,
-  size = 0.8, fold_id = 1, task = task,
-  show_blocks = TRUE, show_labels = TRUE) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
-```
-
-\newpage
-
--->
 
@@
 
-<!-- TODO: AB: should be combined with the previous figure into a two-panel figure. Differences are much easier to recognize if plotted next to each other. -->
 ```{r env-1-eval, fig.cap='Environmental leave-one-block-out CV from package blockCV using one (left, "distdeforest") and two (right, "distdeforest" and "slope") predictors to define blocks in the feature space.', out.width="100%", fig.pos="ht", echo=FALSE}
@@
 ```
-<!-- AB: TODO: In general, the code would look much simpler if we could get rid of the scale_x/y_continuous() calls. Improve default settings of autoplot()? -->
 
@@
 
-<!-- spcv_block: how to set argument `theRange` -->
-<!-- AB: argument is actually called range, not theRange -->
+<!-- spcv_block: how to set argument `range`` -->
 Method `"spcv_block"` requires the user to supply a value for argument `range` which denotes the block size used to create the blocking scheme across the study area.

### 2021-09-18 paper/paper.Rmd
clarify the term blocking again

@@
 
-The terminology "block" used in this categorization and in the manuscript in general refers to any kind of regular and irregularly shaped grouping of observations and not necessarily only to a rectangular partitioning as used in some methods.
+The terminology "block" used in this categorization and in the manuscript in general refers to any kind of regular and irregularly shaped grouping of observations which divides the study area into different zones [russ2010; @bahn2012; @wenger2012].
+This is different to the meaning of rectangular-only partitioning within some specific resampling methods [@blockCV].
 

### 2021-09-18 paper/paper.Rmd
resolve some comments

@@
 
-<!-- AB: TODO: folds = 5 makes very little sense in the case of LODO-CV, since this would only sample 5 discs, which cover only a small fraction of the study area.
-Let's use folds = 100, this will at least make readers think about their choice of #folds.
--->
-
 ```{r disc-non-eval, eval=FALSE}
@@
 
-<!-- TODO: AB: This diagram is inconsistent with the other diagrams - should display training and test samples. -->
 ```{r custom-cv-non-eval, eval=FALSE}
@@
 
-<!-- AB: TODO: The comment says that b$hash is optional, but the reader will not know what it is good for.
-A more verbose comment would be helpful, maybe 'optional task identifier' or whatever the hash name may be good for.
--->
 ```{r create-task}
 b = mlr3::as_data_backend(ecuador)
-b$hash = "_mlr3_tasks_ecuador_" # optional
 task = TaskClassifST$new(

### 2021-09-19 paper/paper.Rmd
add rsmp schema, upd cross-refs

@@
 
-<!-- FIXME: graphic here -->
+```{r rsmp-schema, echo=FALSE, fig.cap="Conceptual overview of various spatial partitioning schemas.
+Starting from unpartioned spatial observations (top left) either a spatial block partitioning or a 'spatial leave-one-out resampling' is applied in the first step.
+A spatial block partitioning can further be turned into a 'leave-one-block-out resampling' or a 'k-fold CV resampling at the block level'.
+The use of a buffer is theoretically possible in any scenario but in practice only offered by specific method implementations.", out.width="100%",fig.pos="ht"}
+knitr::include_graphics("pdf/rsmp-schema.pdf")
+```
 
@@
 Due to its similarity to a LOO CV, this method will result in as many train/test evaluations as observations are present.
-In the visualization of the first fold in Figure&nbsp;\@ref(fig:buffer) a buffer zone of 1000 m was used without any custom presence/absence treatment of the data.
+In the visualization of the first fold in Figure&nbsp;\@ref(fig:buffer-eval) a buffer zone of 1000 m was used without any custom presence/absence treatment of the data.
 
@@
 
-### Custom: `"custom_cv"` in mlr3
+### Custom: `"custom_cv"` in mlr3 {#sec:custom-cv}
 
@@
 Note that the blue dots correspond to five discrete time levels, which appear as a point cloud due to the viewing angle.', fig.pos="ht"}
-knitr::include_graphics("lto.png")
+knitr::include_graphics("png/lto.png")
 ```
@@
 The data is viewed from the future towards the past.', fig.pos="ht"}
-knitr::include_graphics("llo.png")
+knitr::include_graphics("png/llo.png")
 ```
@@
 The grey points are excluded from both the training and the test set in this fold.', fig.pos="ht"}
-knitr::include_graphics("llto.png")
+knitr::include_graphics("png/llto.png")
 ```
@@
 
-### Clustering-based resampling {#sec:cluster-cv}
-
 # Step-by-step example: Comparing spatial and non-spatial CV {#sec:case-study}
@@
 This approach is better suited for the shape of the present study area than a rectangular partitioning.
-Figures&nbsp;\@ref(fig:vis-spcv)) and&nbsp;\@ref(fig:vis-nspcv)) show the contrasting distributions of training and test samples.
+Figures&nbsp;\@ref(fig:vis-spcv-eval)) and&nbsp;\@ref(fig:vis-nspcv-eval)) show the contrasting distributions of training and test samples.
 For demonstration purposes we use only four CV folds and two repetitions.
@@
 
-Finally, we visualize the partitions that were used during performance estimation making use of the generic `autoplot()` function in package \pkg{mlr3spatiotempcv} (Figure&nbsp;\@ref(fig:vis-spcv)).
+Finally, we visualize the partitions that were used during performance estimation making use of the generic `autoplot()` function in package \pkg{mlr3spatiotempcv} (Figure&nbsp;\@ref(fig:vis-spcv-eval)).
 

### 2021-09-19 paper/paper.Rmd
add example and plot for `"cv"` and update description

@@
 rsmp_buffer = rsmp("spcv_buffer", theRange = 1000)
-rsmp_buffer$instantiate(task)
 
@@
 rsmp_buffer = rsmp("spcv_buffer", theRange = 1000)
-rsmp_buffer$instantiate(task)
+rsmp_buffer
 
@@
 rsmp_disc = rsmp("spcv_disc", folds = 100, radius = 300L, buffer = 400L)
-rsmp_disc$instantiate(task)
+rsmp_disc
 
-autoplot(rsmp_disc,
-  size = 0.8, task = task, fold_id = 1)
+autoplot(rsmp_disc, size = 0.8, task = task, fold_id = 1)
 ```
@@
 ```{r disc-eval, fig.cap='Visualization of one training set / test set combination generated with the leave-one-disc-out method from package sperrorest (method "spcv\\_disc" in mlr3spatiotempcv).
-The disc has a radius of 300 m and is surrounded by a 400-m buffer.', out.width="40%", echo=FALSE,fig.pos="hb"}
+The disc has a radius of 300 m and is surrounded by a 400-m buffer.', out.width="40%", echo=FALSE,fig.pos="ht"}
 library("mlr3")
@@
 rsmp_disc = rsmp("spcv_disc", folds = 100, radius = 300L, buffer = 400L)
-rsmp_disc$instantiate(task)
+rsmp_disc
 
@@
 ```{r coords-non-eval, eval=FALSE}
-rsmp_coords = rsmp("spcv_coords", folds = 5)$instantiate(task)
+rsmp_coords = rsmp("spcv_coords", folds = 5)
 
@@
 ```{r coords-eval, fig.cap='Leave-one-block-out CV based on $k$-means clustering of the coordinates as implemented in package sperrorest (method "spcv\\_coords" in mlr3spatiotempcv).', out.width="40%", fig.pos="ht", echo=FALSE}
-rsmp_coords = rsmp("spcv_coords", folds = 5)$instantiate(task)
+rsmp_coords = rsmp("spcv_coords", folds = 5)
 
@@
 ```{r custom-cv, fig.cap='Leave-one-level-out (custom) resampling from package mlr3 (method "custom\\_cv").
-A factor variable is used to define the partitions.
-Each color represents one partition.', out.width="40%", fig.pos="ht", echo=FALSE}
+A factor variable is used to define all partitions.', out.width="40%", fig.pos="ht", echo=FALSE}
 breaks = quantile(task$data()$dem, seq(0, 1, length = 6))
@@
 rsmp_block_systematic = rsmp("spcv_block",
-  range = 1000, folds = 5,
-  selection = "systematic"
+  range = 1000, folds = 5, selection = "systematic"
 )
@@
 
-Group-level CV is based on the idea that an inherent grouping structure is present the observations or the sampling design in space and/or in time, which can be used to determine partitions.
-In contrast to blocking or clustering, the spatial or temporal location is not used explicitly, but additional metadata is exploited such as the membership in a specific agricultural management unit [@brenning2006], or in a site in hierarchical sampling [@kasurak2011].
-<!--
-This has also been referred to as "target-oriented validation" [@cast].
-FIXME:
-AB: I don't think I agree with this.
-My interpretation of the Meyer paper was that in the Meyer paper, locations or time points are the targets.
-Fields consist of many grid cells, at multiple location.
--->
-Even though flexibility with respect to fold creation is limited due to the inherited grouping structure within the data, some implementations allow to specify the resulting number of folds by omitting parts of the dataset.
+The `"cv"` resampling strategy performs random, non-spatial partitioning.
+To use it in a spatiotemporal context, one can set the "group" column role in a mlr3 Task object for specific variable(s).
+This tells the resampling strategy to honor these grouping(s) during partitioning.
+In the final folds, these groups of observations will either belong to the test or training set and by this create possibly spatiotemporal resampling folds.
 
-\pkg{mlr3spatiotempcv} supports two "grouping" implementations: (1) The "leave-location-and-time-out" family from package \pkg{CAST} and (2) "custom resampling" from the \pkg{mlr3} package.
+The idea is to make use of inherent grouping structure in the observations or the sampling design in space and/or in time to determine partitions.
+In contrast to geometric blocking or clustering, the spatial or temporal location is not used explicitly, but additional metadata is exploited such as the membership in a specific agricultural management unit [@brenning2006], or in a site in hierarchical sampling [@kasurak2011].
+In \pkg{mlr} this concept was called "blocking".
+
+In the following example we create a custom grouping of eight classes using a kmeans clustering.
+Note that in practice usually the group creation would not be done in an unsupervised way but by making use of an inherent grouping information from the dataset.
+To showcase the honoring of the groups a uneven amount of folds (with respect to the number of groups) is requested.
+The test set in the first and second fold is composed out of three groups while the third fold is only made up from two groups.
+
+```{r cv-non-eval-1, eval=FALSE}
+task_cv = tsk("ecuador")
+group = as.factor(kmeans(task$coordinates(), 8)$cluster)
+task_cv$cbind(data.frame("group" = group))
+task_cv$set_col_roles("group", roles = "group")
+
+rsmp_cv_group = rsmp("cv", folds = 3)$instantiate(task_cv)
+
+print(rsmp_cv_group$instance)
+```
+
+```{r cv-non-eval-2, fig.cap='Cross-Validation at the block level including predefined groups from package mlr3 (method "cv").
+A factor variable is used to define the grouping.
+Each class is either assigned to the test or training set.', out.width="40%", fig.pos="ht", eval=FALSE}
+autoplot(rsmp_cv_group, size = 0.8, task = task_cv, fold_id = 1)
+```
+
+```{r cv-eval-1, echo=FALSE}
+task_cv = tsk("ecuador")
+group = as.factor(kmeans(task$coordinates(), 8)$cluster)
+task_cv$cbind(data.frame("group" = group))
+task_cv$set_col_roles("group", roles = "group")
+
+rsmp_cv_group = rsmp("cv", folds = 3)$instantiate(task_cv)
+
+print(rsmp_cv_group$instance)
+```
+
+```{r cv-eval-2, fig.cap='Cross-Validation at the block level including predefined groups from package mlr3 (method "cv").
+A factor variable is used to define the grouping.
+Each class is either assigned to the test or training set.', out.width="40%", fig.pos="ht", echo=FALSE}
+autoplot(rsmp_cv_group, size = 0.8, task = task_cv, fold_id = 1) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```
 
@@
 
-plot_env_single = autoplot(rsmp_env, size = 0.8, fold_id = 1, task = task)
+plot_env_single = autoplot(rsmp_env, size = 0.5, fold_id = 1, task = task)
 
-plot_env_multi = autoplot(rsmp_env_multi, size = 0.8, fold_id = 1, task = task)
+plot_env_multi = autoplot(rsmp_env_multi, size = 0.5, fold_id = 1, task = task)
 
@@
 
-```{r env-1-eval, fig.cap='Environmental leave-one-block-out CV from package blockCV using one (left, "distdeforest") and two (right, "distdeforest" and "slope") predictors to define blocks in the feature space.', out.width="100%", fig.pos="ht", echo=FALSE}
+```{r env-1-eval, fig.cap='Environmental leave-one-block-out CV from package blockCV using one (left, "distdeforest") and two (right, "distdeforest" and "slope") predictors to define blocks in the feature space.
+Due to feature space clustering observations are not (necessarily) grouped in the spatial domain.', out.width="100%", fig.pos="ht", echo=FALSE}
 rsmp_env = rsmp("spcv_env",
@@
 plot_env_single = autoplot(rsmp_env,
-  size = 0.6, fold_id = 1, task = task) *
+  size = 0.5, fold_id = 1, task = task) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
@@
 plot_env_multi = autoplot(rsmp_env_multi,
-  size = 0.6, fold_id = 1, task = task) *
+  size = 0.5, fold_id = 1, task = task) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
@@
 library("patchwork")
-plot_env_single + plot_env_multi + plot_layout(guides = 'collect')
+plot_env_single + plot_env_multi + plot_layout(guides = "collect")
 ```
@@
 ```{r llto, echo=FALSE, out.width="70%", fig.cap='Perspective plot of leave-location-and-time-out CV from package CAST (method "sptcv\\_cstf" and arguments time\\_var = "Date" and space\\_var = "SOURCEID").
-The grey points are excluded from both the training and the test set in this fold.', fig.pos="ht"}
+The grey points are excluded from both the training and the test set in this fold.', fig.pos="h"}
 knitr::include_graphics("png/llto.png")
@@
 ```{r create-task}
-b = mlr3::as_data_backend(ecuador)
+backend = mlr3::as_data_backend(ecuador)
 task = TaskClassifST$new(
-  id = "ecuador", b, target = "slides", positive = "TRUE",
+  id = "ecuador", backend = backend, target = "slides", positive = "TRUE",
   extra_args = list(
@@
 rsmp_nsp = rsmp("repeated_cv", folds = 4, repeats = 2)
-rsmp_nsp$instantiate(task)
+rsmp_nsp
 rr_nsp = resample(
@@
 rsmp_sp = rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
-rsmp_sp$instantiate(task)
+rsmp_sp
 rr_sp = resample(

### 2021-09-20 paper/paper.Rmd
rewrite case study summer section

@@
 
-## Summary
-
-The non-spatial CV estimate of AUC (0.76) is around 0.13 better compared to the spatial CV estimate of 0.64.
-The better performance results from over-optimism in the estimation with non-spatial CV.
-Spatial CV, in contrast, provides a better measure of a model's ability to generalize from the training sample --- in this case study, from the specific hillslopes and historical landslides in the training sample.
-It is also expected that spatial CV results better reflect the performance of a model on new, spatially separate data.
+## Interpretation
+
+The non-spatial CV estimate of AUC (0.76) is around 0.13 higher compared to the spatial CV estimate of 0.64.
+The higher non-spatial CV value results from over-optimism in the estimation with non-spatial CV and does imply better performance.
+Spatial CV, in contrast, provides a better/more accurate measure of a model's ability to generalize from the training sample --- in this case study, from the specific hillslopes and historical landslides in the training sample.
+It is also expected that spatial CV results reflect a more honest performance of a model on new, spatially separate data.
 The magnitude of the difference between spatial and non-spatial CV estimates may depend on the dataset, the strength on spatial or spatiotemporal autocorrelation, and the learner itself.
@@
 Since spatial CV can also be used for hyperparameter tuning in the \pkg{mlr3} framework, the model can be optimized to show an improved performance in a specific predictive setting [@schratz2019].
-This may, for example, involve a reduced maximum tree depth in the Ecuador case study, since this leads to stronger generalization and reduced overfitting.
+This may, for example, involve a reduced maximum tree depth in the Ecuador case study, since this parameter leads to stronger generalization and reduced overfitting.
 

### 2021-09-20 paper/paper.Rmd
adjust cv vs custom cv section to new grouping schema

@@
 The question which resampling method should be chosen for a prediction task and dataset at hand comes up regularly in practice.
-Even though there is no definitive and easy answer, we would like to give some guidance in this section to help find an appropriate method.
-
-Although the case study example in section \@ref(sec:case-study) used the `"spcv_coords"` method, this should not imply that this method is the best or only method suitable for this example task.
+Even though there is and will be no definitive answer to this, we would like to give some guidance in this section to help finding an appropriate method.
+
+Although the case study example in section \@ref(sec:case-study) used the `"spcv_coords"` method, this should not give the impression that this method is the best or only method suitable for this example task.
 Even though this method is frequently used, it was mainly chosen because of its clear visual grouping differences when comparing spatial to random partitioning.
@@
 Last, there are no obvious spatial groupings to our knowledge which need to be accounted for, also with respect to prediction purposes of the fitted model.
+Hence, this method is a valid choice for the given dataset.
 
@@
 It is recommended (required) to make oneself familiar with the available methods and to decide on one based on the specific characteristics of the dataset at hand and the prediction purpose of the fitted model.
-The prediction purpose is the model's ability for which areas or zones it actually can make predictions.
+The prediction purpose is the model's ability for which areas or zones it actually can make predictions for.
 In the crop example given in section \@ref(sec:intro) this would be to distinguish and predict the crop type of 'new', unseen fields within an agricultural region.
@@
 
-Even though the prediction purpose should be the first decision criteria for choosing an appropriate spatiotemporal resampling method, we would like to discuss the individual methods in more detail in the following section.
+Even though the prediction purpose should be the main decision criteria for choosing an appropriate spatiotemporal resampling method, we would like to discuss the individual methods in more detail in the following section.
 
@@
 Methods that omit some observations (`"spcv_buffer"`, `"spcv_tiles"`, `"sptcv_cstf"` with both `space_var` and `time_var` set) are able to reduce the impact of autocorrelation in scenarios without a clear grouping or sufficient distance between train and test set observations.
-The downside is the removal of observations from the dataset which possibly reduces the generalization power of the trained model and by this, depending on the specific observations that were removed, adds substantial bias to the CV results.
+The downside is the removal of observations from the dataset which possibly reduces the power of the trained model and by this, depending on the specific observations that were removed, adds substantial bias to the CV results.
 
@@
 <!-- spcv_coords: limited variance due to limited k-means possibilities -->
-For repeated executions of method `spcv_coords`, the variance between repetitions can become quite small due to the limited possibilities of differing $k$-means-based clusters on the same dataset [@brenning2012].
+For repeated executions of method `"spcv_coords"`, the variance between repetitions can become quite small due to the limited possibilities of differing $k$-means-based clusters on the same dataset [@brenning2012].
 This applies even more to `"spcv_block"` with options `selection = "systematic"` and `selection = "checkerboard"` as the fold assignments are the same in each repetition.
@@
 
-<!-- group level CV vs custom CV -->
-\pkg{mlr3spatiotempcv} supports two factor-based partitioning schemes: "group-level resampling" and "custom resampling".
-Both are similar in that they require external user input for partition creation, i.e. a factor variable which assigns all observations into a distinct group.
-The idea of "custom resampling" is that each group becomes a test set in the CV (leave-one-group-out CV).
-By contrast, "group-level resampling" methods resample these initial groups: for example, users can supply an integer value to an argument named `fold` in the LLO method (e.g.
-`5L`) which combines the initial groups (e.g.
-$n = 20$) into their final fold-level partitions, which would result in four groups per partition [@meyer2018].
-The important point is that observations from within one group will either be assigned to the train or test set and never be split up.
-"Group-level resampling" can be, in contrast to "custom resampling", repeated multiple times because multiple different aggregations are possible in theory.
-However, it also requires the number of initial groups to be substantially higher than the number of folds so that a sensible aggregation is possible in the first place.
+<!-- cv vs custom_cv -->
+\pkg{mlr3spatiotempcv} supports two factor-based partitioning schemes: `"cv"` (with column role `"groups"` set) and `"custom_cv"`.
+Both are similar in that they require external user input for partition creation, i.e. a factor variable which assigns all observations into distinct groups.
+The idea of `"custom_cv"` is that each group becomes a test set in the CV (leave-one-group-out CV).
+By contrast, method `"cv"` (with column role `"groups"` set) resamples the initial groups and performs CV at the block level.
+The important point for method "cv" is that observations from a specific group will either be assigned to the train or test set and never be split up.
+In contrast to method `"custom_cv"`, this approach can be repeated multiple times because multiple different fold combinations are possible.
+However, it also requires the number of initial groups to be (substantially) higher than the number of folds so that a sensible combination is possible in the first place.
 

### 2021-09-20 paper/paper.Rmd
clean comment

@@
 
-<!-- missing: Guidelines how to choose a resampling method -->
-<!-- AB: I don't think we can present such guidelines in this paper, I think we discussed this recently and agreed that we will not propose such guidelines, we should rather focus on the methods and their implementation -->
 The \pkg{mlr3spatiotempcv} package is the first package to bundle and categorize spatiotemporal resampling methods implemented in multiple other packages in R.
-The available resampling techniques allow the users to vary the scale or granularity of the resampled spatial units as well as their shape and possible buffer distance between training and test samples.
+The available resampling techniques allow the users to vary the scale or granularity of the resampled spatiotemporal units as well as their shape and possible buffer distance between training and test samples.
 These settings may account for the specific characteristics of spatiotemporal prediction tasks, but modelers now have to make the important decision of choosing a method that is adequate for their situation.

### 2021-09-22 paper/paper.Rmd
edited Table 1

@@
   \caption[t]{Spatiotemporal resampling methods available in the mlr3 ecosystem.
-    Column name shows the mlr3 method name as found in the `mlr3::mlr\_resamplings` dictionary.
-    Column "Count" refers to the number of studies which were found to have used this resampling technique until May 2021.
-    Literature appearances were only collected for methods with a "spcv" or "sptcv" prefix as methods which rely on custom grouping ("custom\_cv" and "cv") do not have a dedicated name or identifier to make them detectable in research papers.
+    The Name column shows the mlr3 method name as found in the `mlr3::mlr\_resamplings` dictionary.
+    \# pub. indicates the number of studies that were found to have used this resampling technique until May 2021.
+    <!--- Literature appearances were only collected for methods with a "spcv" or "sptcv" prefix as methods which rely on custom grouping ("custom\_cv" and "cv") do not have a dedicated name or identifier to make them detectable in research papers.
+--->
     For each method, up to three randomly chosen references were added to the table as footnotes.}
@@
       \\
-      Type                            & Name                        & R Package     & Sub-Type  & Count\\
+      Type                            & Sub-type & Name                        & R Package     & \# pub. \\
       \toprule
-      \multirow{2}{*}{Spatial Leave-one-out}        &  \texttt{"spcv\_buffer"} \tablefootnote{\label{ploton2020}\cite{ploton2020}} \tablefootnote{\label{diesing2020}\cite{diesing2020}} & \pkg{blockCV}  & buffer, LOO & 2   \\
+      \multirow{2}{*}{Spatial leave-one-out}        & single point, with buffer &  \texttt{"spcv\_buffer"} \tablefootnote{\label{ploton2020}\cite{ploton2020}} \tablefootnote{\label{diesing2020}\cite{diesing2020}} & \pkg{blockCV}  & 2   \\
       \cmidrule{2-5}
-       & \texttt{"spcv\_disc"} \tablefootnote{\label{endicott2017}\cite{endicott2017}} \tablefootnote{\label{karasiak2021}\cite{karasiak2021}} \tablefootnote{\label{moller2021}\cite{moller2021}} & \pkg{sperrorest}      & buffer, multiple observations  & 3   \\
+       & disc, with buffer  & \texttt{"spcv\_disc"} \tablefootnote{\label{endicott2017}\cite{endicott2017}} \tablefootnote{\label{karasiak2021}\cite{karasiak2021}} \tablefootnote{\label{moller2021}\cite{moller2021}} & \pkg{sperrorest}      & 3   \\
       \midrule
-      \multirow{4}{*}{Leave-one-block-out CV} & \texttt{"spcv\_coords"}  \tablefootnote{\label{morera2021}\cite{morera2021}} \tablefootnote{\label{geiss2017}\cite{geiss2017}} \tablefootnote{\label{wu2020}\cite{wu2020}} & \pkg{sperrorest}       &  clustering & 6 \\
+      \multirow{4}{*}{Leave-one-block-out CV} &  clustering of coordinates & \texttt{"spcv\_coords"}  \tablefootnote{\label{morera2021}\cite{morera2021}} \tablefootnote{\label{geiss2017}\cite{geiss2017}} \tablefootnote{\label{wu2020}\cite{wu2020}} & \pkg{sperrorest}       & 6 \\
             \cmidrule{2-5}
-                                      & \texttt{"spcv\_tiles"} \tablefootnote{\label{bebber2017}\cite{bebber2017}} \tablefootnote{\label{brenning2015}\cite{brenning2015}} \tablefootnote{\label{zurell2020}\cite{zurell2020}} & \pkg{sperrorest}    &  geometric &  4  \\
+                                      &  geometric: rectangular & \texttt{"spcv\_tiles"} \tablefootnote{\label{bebber2017}\cite{bebber2017}} \tablefootnote{\label{brenning2015}\cite{brenning2015}} \tablefootnote{\label{zurell2020}\cite{zurell2020}} & \pkg{sperrorest}    &  4  \\
                                                   \cmidrule{2-5}
-                                      &  \texttt{"custom\_cv"}           & \pkg{mlr3} & custom  & - \\
+                                      & custom  &  \texttt{"custom\_cv"}           & \pkg{mlr3} & - \\
       \midrule
-      \multirow{4}{*}{CV at the block level} & \texttt{"spcv\_block"}  \tablefootnote{\label{jensen2021}\cite{jensen2021}} \tablefootnote{\label{escobar2021}\cite{escobar2021}} \tablefootnote{\label{stewart2021}\cite{stewart2021}} & \pkg{blockCV}       &  geometric & 28 \\
+      \multirow{4}{*}{CV at the block level} &  geometric: rectangular & \texttt{"spcv\_block"}  \tablefootnote{\label{jensen2021}\cite{jensen2021}} \tablefootnote{\label{escobar2021}\cite{escobar2021}} \tablefootnote{\label{stewart2021}\cite{stewart2021}} & \pkg{blockCV}       & 28 \\
   \cmidrule{2-5}
-                                      & \texttt{"cv"}  & \pkg{mlr3}    &  custom, honors grouping &  -  \\
+                                      &  custom & \texttt{"cv"} with grouping  & \pkg{mlr3}    &  -  \\
   \cmidrule{2-5}
-                                      & \texttt{"spcv\_env"} \footref{morera2021}  & \pkg{blockCV}    &  clustering, feature space &  1  \\
+                                      &  clustering in feature space & \texttt{"spcv\_env"} \footref{morera2021}  & \pkg{blockCV}    &  1  \\
       \midrule
-      \multirow{3}{*}{CV for spatiotemporal data} & \texttt{"sptcv\_cstf"}  \tablefootnote{\label{egli2020}\cite{egli2020}} \tablefootnote{\label{gao2019}\cite{gao2019}} \tablefootnote{\label{reitz2021}\cite{reitz2021}} & \pkg{CAST}       &  custom & 6 \\
+      \multirow{3}{*}{CV for spatiotemporal data} &  custom & \texttt{"sptcv\_cstf"}  \tablefootnote{\label{egli2020}\cite{egli2020}} \tablefootnote{\label{gao2019}\cite{gao2019}} \tablefootnote{\label{reitz2021}\cite{reitz2021}} & \pkg{CAST}       & 6 \\
                                                   \cmidrule{2-5}
-                                      & \texttt{"sptcv\_cluto"}  & \pkg{skmeans}    &  clustering & - \\
+                                      &  clustering: space--time coord. & \texttt{"sptcv\_cluto"}  & \pkg{skmeans}    & - \\
                                       \bottomrule

### 2021-09-22 paper/paper.Rmd
re-organized, rephrased intro part of section 4

@@
 
-# Spatiotemporal partitioning methods
+# Spatiotemporal partitioning methods and their implementation
+
+At the most general level, resampling methods are categorized according to the level at which the data is partitioned and resampled (see Figure&nbsp;\@ref(fig:rsmp-schema)):
+
+- **Spatial leave-one-out resampling**: Each individual observation forms a test set;
+- **Leave-one-block-out CV**: Individual blocks are left out as test data, i.e. the number of folds equals the number of blocks;
+- **CV at the block level**: Blocks are grouped into $k$ partitions, each of which is used as a test fold.
+
+In this context, a block can refer to an arbitrarily shaped spatial (or spatiotemporal) group of observations, not necessarily a rectangular region.
+A finer distinction can then be made by looking at how the blocks are derived:
+
+- Using a geometry-based approach (rectangular or circular);
+- Using an unsupervised clustering approach;
+- Using a custom input, i.e. specifying the blocks with an external grouping variable.
+
+In some resampling schemes, separation buffers or guard zones can be imposed to separate the training and test data.
+
+```{r rsmp-schema, echo=FALSE, fig.cap="Conceptual overview of various spatial partitioning schemas.
+Starting from unpartioned spatial observations (top left) either a spatial block partitioning or a 'spatial leave-one-out resampling' is applied in the first step.
+A spatial block partitioning can further be turned into a 'leave-one-block-out resampling' or a 'k-fold CV resampling at the block level'.
+The use of a buffer is theoretically possible in any scenario but in practice only offered by specific method implementations.", out.width="100%",fig.pos="ht"}
+knitr::include_graphics("pdf/rsmp-schema.pdf")
+```
 
@@
 Comprehensive and up-to-date information can be found in the package's online documentation (<https://mlr3spatiotempcv.mlr-org.com>).
-The following sections give an overview of most existing partitioning strategies and their visualization options.
+The following sections give an overview of most implemented partitioning strategies and their visualization options.
 The available methods are further discussed in section \@ref(sec:disc).
@@
 
-On the most upper level, resampling methods are categorized through the way the fold partitioning is done:
+## Spatial leave-one-out
 
-- Leaving out single pixels (Spatial Leave-one-out) per fold with an optional buffer
-- Leaving out single blocks (Leave-one-block-out CV) per fold with an optional buffer
-- Leaving out multiple blocks (CV at the block level) per fold
+Spatial leave-one-out methods start from a single observation in space and create radial (buffer) zones around which define the partitions.
 
-A finer distinction can then be made by looking at how these groups are formed by the respective individual method:
-
-- Using a geometric-based approach (rectangular or circular)
-- Using an unsupervised clustering approach
-- Using a custom/external input, i.e. specifying the groups via an external grouping variable
-
-The terminology "block" used in this categorization and in the manuscript in general refers to any kind of regular and irregularly shaped grouping of observations which divides the study area into different zones [russ2010; @bahn2012; @wenger2012].
-This is different to the meaning of rectangular-only partitioning within some specific resampling methods [@blockCV].
-
-<!-- explain and introduce buffering -->
-The use of a buffer is not a direct grouping characteristic of a partitioning method in our view but only an optional add-on which could potentially be used by any method.
-"Buffering" refers to the idea of removing observations located between test and training sets by drawing a circular buffer around one or more observations [@brenning2012; @rest2014; @blockCV].
-Within this buffer zone, all observations are removed from the training set, thereby reducing the similarity between training and test set.
-Buffering is mostly applied in the coordinate space.
-However, there is no practical limitation that would prevent this method to be applied in a 2D + time setting.
-Since \pkg{mlr3spatiotempcv} supports two spatial-only "buffering" implementations, (1) `"spcv\_buffer"` from package \pkg{blockCV} and (2) `"spcv\_disc"` from \pkg{sperrorest}, buffering was attributed to the category of spatial partitioning methods (Table&nbsp;\@ref(tab:sptcv-methods)).
-
-```{r rsmp-schema, echo=FALSE, fig.cap="Conceptual overview of various spatial partitioning schemas.
-Starting from unpartioned spatial observations (top left) either a spatial block partitioning or a 'spatial leave-one-out resampling' is applied in the first step.
-A spatial block partitioning can further be turned into a 'leave-one-block-out resampling' or a 'k-fold CV resampling at the block level'.
-The use of a buffer is theoretically possible in any scenario but in practice only offered by specific method implementations.", out.width="100%",fig.pos="ht"}
-knitr::include_graphics("pdf/rsmp-schema.pdf")
-```
-
-## Spatial Leave-one-out
-
-Spatial Leave-one-out methods start from a single observation in space and create radial (buffer) zones around which define the partitions.
-
-### Spatial Leave-one-out with buffer --- `"spcv_buffer"`
+### Spatial leave-one-out with buffer --- `"spcv_buffer"`
 

### 2021-09-23 paper/paper.Rmd
rephrased/rewrote spcv_buffer text

@@
 
-Spatial leave-one-out methods start from a single observation in space and create radial (buffer) zones around which define the partitions.
+Spatial leave-one-out methods use individual observations in space as test partitions and apply circular buffer or guard zones around around these test points to enforce a minimum prediction distance.
+Leave-one-disc-out resampling modifies this approach to leave out circular regions centered at observation points. 
+Note that this straightforward extension of the LOO method does technically not establish a partitioning as used for CV resampling in the traditional sense.
 
@@
 
-The "buffering" method from the \pkg{blockCV} package is conceptually related to spatial leave-one-out cross-validation.
-It comes with optional support for presence/absence data handling, as often used in the field of species distribution modeling [@hijmans2013].
-However, by default, the method considers all points for sampling the training and test sets.
-
-In an ecological/species distribution modeling context, the default assumption of a binary response variable is to have *presence/absence* data.
-This means all presence and absence values in the response variable are confirmed in their respective status.
-The \pkg{blockCV} package implementation also comes with support for *presence/background* data for binary responses.
-
-All other response observations are assumed to be absent without a proof for this assumption.
-When setting argument `spDataType = "PB"`, the method only considers the *presence* observations in a CV scenario.
-Background data is used for training the model by default unless argument `addBG = FALSE` is set for which only *presence* values not falling within the buffer zone of the respective fold are used for model training.
-
-Due to its similarity to a LOO CV, this method will result in as many train/test evaluations as observations are present.
-In the visualization of the first fold in Figure&nbsp;\@ref(fig:buffer-eval) a buffer zone of 1000 m was used without any custom presence/absence treatment of the data.
+Leave-one-out CV with buffer and several adaptations for species distribution modeling [@hijmans2013] are implemented in the \pkg{blockCV} package as the so-called "buffering" method and integrated into \pkg{mlr3spatiotempcv} under the label `"spcv_buffer"`.
+In species distribution modeling, the response variable can either be recorded as presence/absence data or as presence/background information; both options are supported by this implementation. 
+By default, the dataset contains confirmed presence and confirmed absence observations, i.e.
+locations where a species was observed and not observed, respectively, and therefore spatial LOO CV in its usual sense can be carried out.
+Figure&nbsp;\@ref(fig:buffer-eval) shows the first test fold generated with this method for presence/absence data with a buffer distance of 1000 m.
 
@@
 ```
+
+In the presence/background (or presence-only) situation, in contrast, only presence observations are recorded, and all other locations within the study area are referred to as the background and considered as pseudo-absences.
+Presence/background modeling can be enabled with the argument `spDataType = "PB"`.
+In this situation, the method constructs test folds that are centered at the recorded presence locations, offering two different modes of operation.
+With `addBG = TRUE` (the default), all background points with a distance of `theRange` around a test (presence) point are included in the test fold as absence data.
+The `addBG = FALSE` setting, in contrast, no background data is added to the test fold, which then contains only one (presence) observation, and only the data at a distance of `theRange` or greater are included in the training sample, including background data from these areas.
+
+The application of LOO methods can be computationally expensive since the method cycles through the entire dataset and fits one model for each test fold.
 

### 2021-09-23 paper/paper.Rmd
spcv_buffer: clarified for addBG=TRUE: there's no separation buffer!

@@
 In this situation, the method constructs test folds that are centered at the recorded presence locations, offering two different modes of operation.
-With `addBG = TRUE` (the default), all background points with a distance of `theRange` around a test (presence) point are included in the test fold as absence data.
+With `addBG = TRUE` (the default), all background points with a distance of `theRange` around a test (presence) point are included in the test fold as absence data; note that in this case, there is no separation buffer between training and test samples.
 The `addBG = FALSE` setting, in contrast, no background data is added to the test fold, which then contains only one (presence) observation, and only the data at a distance of `theRange` or greater are included in the training sample, including background data from these areas.

### 2021-09-23 paper/paper.Rmd
edited, shortened leave-one-dist-out section

@@
 Leave-one-disc-out resampling modifies this approach to leave out circular regions centered at observation points.
-Note that this straightforward extension of the LOO method does technically not establish a partitioning as used for CV resampling in the traditional sense.
 
@@
 
-Leave-one-disc-out resampling from package \pkg{sperrorest} randomly selects a circular test set starting from a single observations and optionally excludes a buffer zone from the remaining training data.
+Leave-one-disc-out resampling from package \pkg{sperrorest} defines circular test sets that are centered at sample locations, and optionally excludes a buffer zone from the remaining training data.
 It thus ensures that a minimum separation distance between training and test data is maintained.
-Discs are centered at sample locations, their number being specified by the `folds` argument.
-It is therefore possible to sample up to $n$ discs without replacement (default) or any number when sampling with replacement (`replace = TRUE`).
-The number of discs is specified via the `folds` argument.
+The number of discs is specified by the `folds` argument, which defaults to the sample size $n$.
+Sample locations are selected randomly when `folds` is smaller than $n$; it is optionally possible to sample with replacement (`replace = TRUE`).
+Leave-one-disc-out resampling becomes spatial LOO CV for a radius of 0 m and when each observation is at a unique location.
 
-This resampling method is, strictly speaking, not a partitioning method since the discs are not resulting in disjoint partitions - partitions will potentially overlap in subsequent folds.
-Nevertheless, it is a logical extension of spatial LOO CV.
-Depending on the size of the discs, potentially a large number of discs (`folds` argument) is needed to effectively use the entire dataset.
+Note that the resampled discs will potentially overlap.
+Strictly speaking, this straightforward extension of spatial LOO does therefore not establish a disjoint partitioning as used for CV resampling in the traditional sense.
 
+<!-- AB: I don't think the following explanation is needed: -->
+<!--
 Even though the buffer creates a circular area which defines the test partition, it is not categorized as a "geometric" method as the origin of the circle is observation-based.
 In contrast, "geometric" methods do not take into account single observation locations but only focus on the study area extend and will include all observations which fall into their respective covered area.
-
-Leave-one-disc-out resampling becomes spatial LOO CV for a radius of 0 m and when each observation is at a unique location.
+-->
 

### 2021-09-23 paper/paper.Rmd
small edits in clustering-based LOBO section

@@
 
-## Leave-one-block-out Cross-Validation
+## Leave-one-block-out cross-validation
 
-Leave-one-block-out CV methods partition the dataset in a way that each of the resulting partitions will be used as a fold in the CV.
+Leave-one-block-out resampling methods partition the dataset spatially in order to use each of the resulting partitions as a CV test fold.
 
-### Clustering-based: Using coordinate-based clustering - `"spcv_coords"`
+### Clustering-based: using coordinates - `"spcv_coords"`
 
@@
 
-Coordinate-based clustering approach is very versatile as it adapts to irregularly-shaped study areas and ensures that exactly $k$ partitions are created, which are usually or very similar size when the sample locations are spread out evenly.
+The coordinate-based clustering approach is very versatile as it adapts to irregularly-shaped study areas and ensures that exactly $k$ partitions are created, which are usually of very similar size when the sample locations are spread out evenly.
 Nevertheless, despite the random selection of initial cluster centers, repeated partitionings may in some cases be nearly identical.
-Also, $k$-means clustering may be less suitable for data sets with pre-existing clusters of points and/or with distant isolated sample locations.
+Also, $k$-means clustering may be less suitable for data sets with pre-existing clusters of points and/or with isolated, distant sample locations.
 When distinct clusters of points are present, as in multi-level sampling, it may be better to define clusters using a factor variable (see method `"custom_cv"` in section \@ref(sec:custom-cv)).

### 2021-09-23 paper/paper.Rmd
revised custom_cv section

@@
 Support for user-defined partitioning strategies is built into \pkg{mlr3} directly.
-In this so-called "Custom CV", users are able to supply an external factor vector with the same length as number of observations.
+In this so-called "Custom CV", users supply a factor variable, each level of which defines a partition.
+<!-- AB: I think we can omit this sentence, otherwise we would have to repeat it for the other LOBO methods as well. Here we're in the leave-ONE-something-out section. -->
+<!--
 This partitioning concept should not be mistaken with "group-level resampling" (section \@ref(sec:group-level)), which conducts additional combination of data splits and performs CV at the block level.
-Each level of the factor vector will result in one partition.
-Alternatively to supplying an external factor vector, the name of a feature (type `character`) present within the task can be supplied via argument `col`.
-
-The following example was taken from `sperrorest::partition_factor()` because the resulting partitions separate nicely from each other in a visual way.
+-->
+The factor variable can either be specified through a factor vector of the same length as number of observations, or by passing the name of a feature of type `character` within the task (argument `col`).
+The following simple example from `sperrorest::partition_factor()` creates altitudinal zones that define the spatial partitions.
 

### 2021-09-23 paper/paper.Rmd
edited 'cv' and 'spcv_env' sections

@@
 
-## Leave-one-block-out cross-validation
+## Leave-one-block-out cross-validation {#sec:lobo-cv}
 
@@
 
-### Clustering-based: using coordinates - `"spcv_coords"`
+### Clustering-based: using coordinates --- `"spcv_coords"`
 
@@
 
-## Cross-Validation at the block level
+## Cross-validation at the block level {#sec:block-cv}
 
-Methods which operate on the block level create dataset splits which are then combined to form the final partitions for the CV folds.
-In the special case of an equal amount of dataset splits and requested CV folds, these methods would perform a Leave-one-block-out CV.
+Methods which operate at the block level first group the observations into blocks and then combine these blocks into CV partitions.
+In $k$-fold CV resampling at the block level, there are therefore $k$ partitions, each consisting of $1/k$-th of the blocks.
+The special case in which $k$ equals the number of blocks, CV at the block level simply becomes leave-one-block-out CV, for which dedicated implementations exist (Section&nbsp;\@ref(sec:lobo-cv)).
 
@@
 The `"spcv_block"` method from package \pkg{blockCV} supports both random and systematic resampling of square blocks (`rsmp("spcv_block")` with argument `selection = "random"` and `"systematic"`, respectively; Figure&nbpsp;~\@ref(fig:block-random) and \@ref(fig:block-systematic)).
-This implementation comes with specific options that are designed for modeling presence-only data, which is a typical situation in species distribution modeling.
-Users can also supply a user-defined polygon via argument `rasterLayer` with predefined blocking zones.
+There are additional options for modeling presence-only data, which is a typical situation in species distribution modeling.
+Users can also supply a user-defined polygon via an argument `rasterLayer` with predefined blocking zones.
 
@@
 
-The *checkerboard* partitioning is a special case of a systematic block partitioning (`selection = "checkerboard"`).
+*Checkerboard* partitioning is a special case of a systematic block partitioning (`selection = "checkerboard"`).
 It inherently supports only two folds, making it less appealing than the more commonly used five- or ten-fold resampling, which achieve larger training set sizes.
 
-### Custom: `"cv"` in mlr3 {#sec:group-level}
+### Custom: `"cv"` with grouping in mlr3 {#sec:group-level}
 
-The `"cv"` resampling strategy performs random, non-spatial partitioning.
-To use it in a spatiotemporal context, one can set the "group" column role in a mlr3 Task object for specific variable(s).
-This tells the resampling strategy to honor these grouping(s) during partitioning.
-In the final folds, these groups of observations will either belong to the test or training set and by this create possibly spatiotemporal resampling folds.
+Although the `"cv"` resampling strategy in \pkg{mlr3} performs random, non-spatial partitioning by default, it can also be used for CV at the block level.
+This is achieved by specifying the "group" column role in a mlr3 Task object, which uses the factor levels as blocks.
+A completely group or block of observations is therefore assigned to a specific partition, which consequently honor the grouping structure.
+In \pkg{mlr} this concept was referred to as "blocking".
 
-The idea is to make use of inherent grouping structure in the observations or the sampling design in space and/or in time to determine partitions.
-In contrast to geometric blocking or clustering, the spatial or temporal location is not used explicitly, but additional metadata is exploited such as the membership in a specific agricultural management unit [@brenning2006], or in a site in hierarchical sampling [@kasurak2011].
-In \pkg{mlr} this concept was called "blocking".
+In contrast to geometric or clustering-based blocks, the spatial or temporal location is not used explicitly, but rather implicitly through the spatial or spatiotemporal footprint of each user-defined block.
+<!-- The following sentence is really a repetition from the general background section, and similar statements could be made in the custom_cv section. -> omit this sentence  -->
+<!--
+additional metadata is exploited such as the membership in a specific agricultural management unit [@brenning2006], or in a site in hierarchical sampling [@kasurak2011].
+-->
 
-In the following example we create a custom grouping of eight classes using a kmeans clustering.
+The following example uses $k$-means clustering to generate classes that are used as blocks.
+<!--
 Note that in practice usually the group creation would not be done in an unsupervised way but by making use of an inherent grouping information from the dataset.
-To showcase the honoring of the groups a uneven amount of folds (with respect to the number of groups) is requested.
-The test set in the first and second fold is composed out of three groups while the third fold is only made up from two groups.
+-->
+To underline the honoring of the groups, we chose a number of groups (eight) that is not a multiple of the number of folds (three).
+The test sets in the first and second folds are therefore composed of three groups while the third one holds two groups.
 
@@
 
-The last method from the \pkg{blockCV} package, "environmental blocking", makes use of *k-means* clustering [@hartigan1979a] in a possibly multivariate space.
-The user can select a variable via argument `feature` which denotes the variable that will be used to create clusters.
-Hereby, $k$-means will use Euclidean distance for cluster creation.
-Hence, the selected input variables should be of type numeric.
+The last method from the \pkg{blockCV} package, referred to as "environmental blocking" [@roberts2017], makes use of *k-means* clustering [@hartigan1979a] in a possibly multivariate space to define blocks for resampling at the block level.
+The user can select one or multiple numeric features via argument `feature` from which the clusters are created.
+Hereby, $k$-means will use Euclidean distance.
 To avoid a potential bias introduced by features with high variance when selecting multiple features, all features are standardized by default.
 
-The following example clusters by feature "distance to forest".
+In the following example, the observations are clustered based on the "distance to forest" feature, which results in a distance-based zonification:
 
@@
 
-It is also possible to cluster based on multiple features:
+It is also possible to cluster based on multiple features, in this example the distance to deforestation and the slope angle:
 

### 2021-09-24 paper/paper.Rmd
revised (partly rewrote) spatiotemp methods section

@@
 
-## Cross-Validation for spatiotemporal data
-
-Some methods have the ability to operate in multiple spaces, i.e. spatial, temporal or spatiotemporal (2D + time).
-To avoid redundancy and putting too much weight on certain methods, only the spatiotemporal examples of these methods will be shown in the following sections.
-For the respective other spaces, usually only either the space or time coordinates need to be omitted from the user input.
-
-### Custom: "Leave-location-and-time-out" --- `"spcv_cstf"`
-
-Leave-location-and-time-out (LLTO) from package \pkg{CAST} is an implementation of spatiotemporal resampling method [@meyer2018].
-It requires both a spatial and a temporal variable which are used to separate observations in a multi-dimensional space.
-It comes with satellite methods named "leave-time-out" (LTO) and "leave-location-out" (LLO) which only operate in a two-dimensional space.
-
-In this example the `cookfarm` dataset is used because it contains the variable ("Date") which serves as the temporal component in the dataset.
-
-The sub-methods "Leave-Time-Out" (LTO) and "Leave-Location-Out" (LLO) are conceptually similar to the "cv" method with a custom grouping from \pkg{mlr3} as they perform a CV at the block level using a predefined grouping structure.
+## Cross-validation for spatiotemporal data
+
+Some of the implemented resampling methods operate in multiple dimensions, i.e. in space, time, or space--time.
+In this section, only examples of these methods in the spatiotemporal domain will be shown.
+For their application in lower dimensions, usually only either the space or time coordinates need to be omitted from the user input.
+
+### Custom: "Leave-location-and-time-out" and related methods --- `"spcv_cstf"`
+
+@meyer2018 proposed a spatiotemporal resampling method in which a test set is selected and all observations that correspond to the same location or time point are omitted from the training sample.
+This method is referred to as "leave-location-and-time-out" (LLTO) in package \pkg{CAST}.
+<!-- AB: The following information is kinda obvious: -->
+<!-- It requires both space coordinates and time point identifiers that are used to locate observations. --> 
+Additional methods that resample in the temporal and spatial domain only are named "leave-time-out" (LTO) and "leave-location-out" (LLO), respectively, in \pkg{CAST}.
+Note that despite their names, LLTO, LTO and LLO are conceptually not leave-*one*-out methods as they place a certain fraction of observations in the test set, as in ordinary CV.
+Also, LTO and LLO are conceptually similar to \pkg{mlr3}'s "cv" method with a custom grouping as they perform a CV at the block level using a grouping structure defined by time points (LTO) and locations (i.e., time series; LLO).
+
+In this section the `cookfarm` dataset is used as an example because has a temporal dimension identified by the variable "Date".
 
@@
 
+<!-- AB: No need to justify internal names of methods:
 In \pkg{mlr3spatiotempcv} the method is named `"sptcv_cstf"` which is an acronym originating from the underlying upstream implementation in package \pkg{CAST} `CreateSpacetimeFolds()`.
-
-#### Leave-Time-Out (LTO)
-
-To better showcase the partitioning of the "Leave-Time-Out" (LTO) method, the temporal variable `Date` was adjusted in this example.
-
-The following code uses the `cookfarm` dataset as the base, modifies the `Date` variable in such a way that it contains five unique levels and then creates a spatiotemporal regression task in \pkg{mlr3spatiotempcv} (Figure&nbsp;\@ref(fig:lto)).
+-->
+
+#### CV at the time-point level: "leave-time-out" (LTO)
+
+In the LTO method, the time points are resampled into the desired number of folds.
+In the terminology used in this work, this can be referred to as resampling at the level of time points, which effectively define blocks.
+Thus, observations from the same time point are jointly sampled into the same test (or training) fold, with no constraints on the temporal distance between the sampled time points.
+This method does therefore not implement block CV in the sense of the time series literature.
+
+In the `cookfarm` example, we reduce the `Date` variable to five unique levels for better visualization, and then create a spatiotemporal regression task in \pkg{mlr3spatiotempcv} (Figure&nbsp;\@ref(fig:lto)).
 
@@
 
-```{r lto, echo=FALSE, out.width="70%", fig.cap='Perspective plot of leave-time-out CV from package CAST (method "sptcv\\_cstf" and argument time\\_var = "Date").
+```{r lto, echo=FALSE, out.width="70%", fig.cap='Perspective plot of "leave-time-out" CV from package CAST (method "sptcv\\_cstf" and argument time\\_var = "Date").
 Only five folds and five time points were used in this example.
@@
 
-#### Leave-Location-Out (LLO)
-
-Method "Leave-Location-Out" (LLO) works similar to "Leave-Time-Out" (LTO) in that it operates in a two-dimensional space.
-Here, specific locations are left out individually for forming test partitions while the temporal information is ignored (Figure&nbsp;\@ref(fig:llo)).
+#### CV at the location level: "leave-location-out" (LLO)
+
+In contrast to LTO, the LLO method randomly resamples locations that may, for example, correspond to time series.
+The sampled locations form the test partition while the temporal information is ignored (Figure&nbsp;\@ref(fig:llo)).
+Unlike spatial CV methods that are based on geometric regions or the clustering of coordinates, the sampled test locations bear no particular spatial relationship.
 
@@
 
-```{r llo, echo=FALSE, out.width="70%", fig.cap='Perspective plot of leave-location-out CV from package CAST (method "sptcv\\_cstf" and argument space\\_var = "SOURCEID").
+```{r llo, echo=FALSE, out.width="70%", fig.cap='Perspective plot of "leave-location-out" CV from package CAST (method "sptcv\\_cstf" and argument space\\_var = "SOURCEID").
 The data is viewed from the future towards the past.', fig.pos="ht"}
@@
 
-#### Leave-Location-and-Time-Out (LLTO)
-
-The "Leave-Location-and-Time-Out" (LLTO) method combines both "Leave-Time-Out" (LTO) and "Leave-Location-Out" (LLO).
-It creates partitions that operate on spatiotemporal clusters to reduce both spatial and temporal autocorrelation.
-If there are multiple observations over time per location, and the modeling purpose is to reduce the bias in the predictions as potentially introduced by spatial as well as temporal autocorrelation, LLTO might be an appropriate choice.
-To reduce autocorrelation in both dimensions, the methods omits observations as the splitting in test and train sets only provides separation in one dimension whereas two (space and time) are needed to account for both types, spatial and temporal autocorrelation.
-In other words, accounting for one dimension as in Figure&nbsp;\@ref(fig:lto) or Figure&nbsp;\@ref(fig:llo), by only having one time step or location in either test or training set, would have no effect on the respective other.
-Thus, the removal of observation can be seen as a proxied buffering in multiple dimensions as shown in Figure&nbsp;\@ref(fig:llto).
+#### "Leave-location-and-time-out" (LLTO)
+
+In LLTO, a test set is first randomly sampled from the data set, and then all observations that correspond to the same location or time point are omitted from the training sample (Figure&nbsp;\@ref(fig:llto)).
+LLTO resampling mimics the situation where a model is trained on time series data from a number of locations and time points, and used to predict the time series at other locations and time points that are not included in the training sample.
+
+Conceptually, LLTO applies zero-distance buffering in both space and time: The buffer zones consist of all observations whose distance to the test sample in either space or time equals zero.
+In the mathematical sense, however, this buffering is not based on a valid metric (or distance function) in three-dimensional space (2D + time) as neither the identity of discernables nor the triangle inequality are satisfied by the underlying combined 'distance' measure.
+Also note that LLTO does not 'combine' LTO with LLO, as neither of these applies a buffer zone.
 

### 2021-09-24 paper/paper.Rmd
corrected affiliations

@@
     # use this syntax to add text on several lines
+    # must use German name, see https://www.lmu.de/de/die-lmu/struktur/zentrale-universitaetsverwaltung/kommunikation-und-presse/lmu-brand-guide/content-style-guide/index.html
     address: |
@@
   - name: Marc Becker
-    affiliation: 'Ludwig-Maximilians-University Munich'
+    affiliation: 'Ludwig-Maximilians-Universität München'
     address: |
-      Department of Statistics
-      Statistical Learning and Data Science group
+      | Department of Statistics
+      | Statistical Learning and Data Science group
     # To add another line, use \AND at the end of the previous one as above

### 2021-09-24 paper/paper.Rmd
revised interpretation section

@@
 Models are fitted to historical landslide occurrences, but they need to learn generalizable relationships between predisposing variables and the response as opposed to perfectly reproducing or memorizing the historical distribution.
-This binary classification task on landslides in Ecuador [@muenchow2012] is available as a built-in task via `tsk("ecuador")`, but we generate it from the learning sample.
+This binary classification task on landslides in Ecuador [@muenchow2012] is available as a built-in task via `tsk("ecuador")`, but here we generate it from the learning sample.
 Random forest is used as a classifier, and the area under the ROC curve (AUROC) as the performance measure.
@@
 Spatial CV is implemented in the form of leave-one-block-out CV using coordinate-based $k$-means clustering to generate  irregularly shaped blocks of roughly equal size.
-This approach is better suited for the shape of the present study area than a rectangular partitioning.
+This approach is better suited for the irregular shape of the present study area than a rectangular partitioning.
 Figures&nbsp;\@ref(fig:vis-spcv-eval)) and&nbsp;\@ref(fig:vis-nspcv-eval)) show the contrasting distributions of training and test samples.
@@
 
-Next, the created resampling object `rsmp_nsp` is passed to function `resample()` together with the task and learner objects created earlier to execute the model assessment.
-This is the actual, potentially time-consuming CV estimation. With the present settings, eight random forest classifiers are fitted and evaluated in this step.
+Next, the created resampling object `rsmp_nsp` is passed to the `resample()` function together with the task and learner objects created earlier to execute the model assessment.
+This is the actual, potentially time-consuming CV estimation.
+With the present settings, eight random forest classifiers are fitted and evaluated in this step --- one model fitted on each CV training set.
 
@@
 
-Finally, we visualize the partitions that were used during performance estimation making use of the generic `autoplot()` function in package \pkg{mlr3spatiotempcv} (Figure&nbsp;\@ref(fig:vis-spcv-eval)).
+Finally, we visualize (two of) the partitions that were used during performance estimation making use of the generic `autoplot()` function in package \pkg{mlr3spatiotempcv} (Figure&nbsp;\@ref(fig:vis-spcv-eval)).
 
@@
 
-The non-spatial CV estimate of AUC (0.76) is around 0.13 higher compared to the spatial CV estimate of 0.64.
-The higher non-spatial CV value results from over-optimism in the estimation with non-spatial CV and does imply better performance.
+The non-spatial CV estimate of AUC (0.76) is substantially higher compared to the spatial CV estimate of 0.64.
+Since test points in non-spatial CV may be from the same slopes or even the same landslides as the training data, the non-spatial CV result should be considered as an over-optimistic estimate of the model's ability to predict the susceptibility to "new" landslides.
 Spatial CV, in contrast, provides a better/more accurate measure of a model's ability to generalize from the training sample --- in this case study, from the specific hillslopes and historical landslides in the training sample.
-It is also expected that spatial CV results reflect a more honest performance of a model on new, spatially separate data.
+It is also expected that spatial CV results better represent the model's transferability to geologically and topographically similar areas adjacent to the training area.
 The magnitude of the difference between spatial and non-spatial CV estimates may depend on the dataset, the strength on spatial or spatiotemporal autocorrelation, and the learner itself.

### 2021-09-24 paper/paper.Rmd
edited Discussion 6.1 (+new 6.2), commented on 6.3 (now 6.4)

@@
 
-<!-- AB: TODO: Check if the following paragraphs overlaps with the Discussion section; perhaps merge into Discussion. -->
-Since spatial CV can also be used for hyperparameter tuning in the \pkg{mlr3} framework, the model can be optimized to show an improved performance in a specific predictive setting [@schratz2019].
-This may, for example, involve a reduced maximum tree depth in the Ecuador case study, since this parameter leads to stronger generalization and reduced overfitting.
-
 # Discussion {#sec:disc}
 
-## Choosing a resampling method
+## Choosing a resampling method for model assessment
 
@@
 Even though there is and will be no definitive answer to this, we would like to give some guidance in this section to help finding an appropriate method.
-
-Although the case study example in section \@ref(sec:case-study) used the `"spcv_coords"` method, this should not give the impression that this method is the best or only method suitable for this example task.
-Even though this method is frequently used, it was mainly chosen because of its clear visual grouping differences when comparing spatial to random partitioning.
-Another advantage is its ability to adapt to the irregularly shaped study area of this example.
-Last, there are no obvious spatial groupings to our knowledge which need to be accounted for, also with respect to prediction purposes of the fitted model.
-Hence, this method is a valid choice for the given dataset.
-
-Most often multiple spatial partitioning methods can be used for a dataset.
-It is recommended (required) to make oneself familiar with the available methods and to decide on one based on the specific characteristics of the dataset at hand and the prediction purpose of the fitted model.
-The prediction purpose is the model's ability for which areas or zones it actually can make predictions for.
-In the crop example given in section \@ref(sec:intro) this would be to distinguish and predict the crop type of 'new', unseen fields within an agricultural region.
-Hence, the CV partitioning is done at the field level to receive a honest estimate of the model's performance in a relevant predictive situation.
-
-There are various factors that might influence the choice of a partitioning method:
-
-- Do natural groupings exists in the data?
-Examples include multi-level sampling designs, land management units, or other characteristics of the study area or sampling design that may induce random effects.
-- How dense are the observations scattered both in space and time?
-  The closer observations are in space and/or time, the more similar they are, and hence the greater is the effect of spatio-temporal autocorrelation on performance estimates.
-- Are some of the predictor variables under consideration known to be strongly autocorrelated such as air temperature or precipitation at the regional level?
-
-Based on these criteria users should choose a matching partitioning method that is either more restrictive (by discarding observations for fold creation) or more liberal (by not removing observations and eventually ignoring natural grouping patterns).
-For almost all methods implemented in \pkg{mlr3spatiotempcv}, there is a scientific publication describing the application areas of the respective approach (either linked in the help file of the package or its respective upstream packages), hence we are refraining from repeating them again in this study.
-
-Last, a note on best practices for the example shown in section \@ref(sec:case-study):
-In this example a cross-validation without hyperparameter tuning was applied.
-In ML practice most often a nested CV should be desired to optimize model hyperparameter within the CV.
-In such scenarios it is recommended to use the same spatial partitioning method for the inner loop (= tuning level) that was chosen for the outer one for consistency reasons.
+As a general rule, we recommend to use a resampling scheme that (1) mimics the predictive situation in which the model will be applied operationally, and (2) is consistent with the structure of the data.
+Both aspects are outlined in this section, starting with two concrete modeling scenarios.
+
+Although the case study example in section \@ref(sec:case-study) used the `"spcv_coords"` method for coordindate-based clustering, this should not give the impression that this method is the only method suitable for this example task.
+In this application setting, we want to assess how well the model generalized from the concrete set of historical landslide occurrences, which is why we ensured that training and test sets contain different, "new" hillslopes and landslides.
+Coordinate-based clustering is particularly appealing in this setting because of its ability to adapt to the irregularly shaped study area of this example.
+Resampling at the level of sub-catchments could have been a viable alternative approach that can be implemented using custom resampling (`"custom_cv"` method); however, this may result in less balanced sizes of test sets as catchment sizes may vary.
+When the timing of landslides is known (event-based inventories) or multiple inventories have been compiled for different time points, it can also be recommendable to additionally sample training and test data from different time points, as with the LLTO and LTO [@meyer2018] or similar methods [@brenning2005].
+
+In other application scenarios such as the crop classification example given in section&nbsp;\@ref(sec:intro) [@pena2015], the objective is to predict the fruit-tree crop type of 'new', unseen fields within an agricultural region.
+In this use case we are not at all interested in "predicting" the already known crop type of (other) grid cells within the same agricultural field.
+Also, a model will likely be much better able to predict the crop type within the same field from multitemporal remote-sensing data since all crops within that field are subject to identical management practices (e.g., use of pesticides, pruning of fruit trees, tree spacing), while 'new' fields may be managed differently.
+As a consequence, grid cells from the same field should be grouped into a block, and resampling should be done at the field level (`"cv"` method with grouping) to receive an honest estimate of the model's performance in a relevant predictive situation.
+If, in contrast, the objective is to apply the model to an adjacent agricultural region (e.g., adjacent county) where the same crop types are present, it may be advisable to use coordinate-based clustering (`"spcv_coords"` method) to obtain larger, contiguous test regions.
+
+In summary, there are various factors that may be considered in judging the suitability of a resampling method:
+
+- Will the model be applied to predict 'new' outcomes at near or more distant spatial locations? 
+- Will it be applied to predict into the future, or hindcast gaps between spatiotemporal observations in the past?
+- How densely are the observations distributed in space and time? Are they more densely distributed that the intended spatial or temporal prediction distance?
+- Is the data naturally grouped, e.g. because of the spatial extent of the studied objects, or as a consequence of multi-level (cluster) sampling?
+- With an eye on environmental blocking and extrapolation in feature space, is it intended to apply the model to predict 'new' outcomes for unobserved values of predictor variables?
+
+Based on these criteria users may choose a matching resampling method that is either more restrictive (by discarding nearby observations for fold creation) or more liberal (by not removing observations and eventually ignoring natural grouping patterns).
+The specific publications related to the methods integrated into \pkg{mlr3spatiotempcv} may give further advice and provide additional use cases for the application of each respective approach.
+Users should therefore also refer to publications that are referenced of linked in the help files of this package or its respective upstream packages.
+
+## Resampling for hyperparameter tuning
+
+CV is also widely used to assess model performance when tuning the hyperparameters of flexible machine-learning models, and this is also supported by the \pkg{mlr3} framework.
+Using the CV methods introduced here, \pkg{mlr3} can therefore be used to optimized models to show an improved performance in specific spatial or spatiotemporal predictive setting [@schratz2019].
+Such an optimization may, for example, result in a reduced maximum tree depth or increased minimum node size in the Ecuador case study, since these hyperparameter settings would result in a stronger generalization and reduced overfitting.
+
+We recommend using a nested CV for this purpose.
+In nested CV, an "inner" CV is performed on each CV training set since hyperparameter tuning is an integral part of model fitting that should not be able to use information from the CV test set.
+In such scenarios it is recommended to use the same spatial resampling method for the inner CV (hyperparameter tuning) as for the outer CV (model assessment) in order to use the appropriate objective function for optimization.
 See @schratz2019 for more details as well as chapter 11 of [Geocomputation with R](https://geocompr.robinlovelace.net/spatial-cv.html) [@lovelace2019].
@@
 
+<!-- AB: I do not understand the point of this discussion section.
+It is not clear to the reader if mlr3spatiotempcv supports both approaches or supports/recommends one of them.
+Also, an implementation 'most likely' being reviewed in an unspecified process does not exactly inspire much trust...
+-->
+
 <!-- discuss pros and cons of implementing methods in a standalone pkg vs a framework -->
-There are pros and cons on whether a new method should be integrated into an existing ecosystem such as \pkg{mlr3} directly or distributed in its own package.
-When contributing a method to an existing ecosystem, developers can benefit from existing structures and orient themselves towards how previous methods were implemented.
-They do not need to worry about complying with various R package related best practices (such as CRAN policies) and can solely focus on the method implementation.
-Also, the new method will be directly usable by all people using this ecosystem following an existing and familiar syntax.
-Contributions to (ML) ecosystems such as \pkg{mlr3} enables users to easily make use of a large variety of additional functionality, from preprocessing over hyperparameter tuning and feature selection to result visualization [@mlr3], which can be used in conjunction with the new resampling method.
-In addition, contributions most likely undergo a thorough review from other community members during the implementation, most often enhancing the code quality of the implementation.
+There are pros and cons to whether a new method should be integrated into an existing modeling framework such as \pkg{mlr3} directly or distributed in its own package.
+When contributing a method to an existing framework, developers can benefit from existing structures and orient themselves towards how previous methods were implemented.
+They do not need to worry about complying with various R package related best practices (such as CRAN policies) and can solely focus on the implementation of an additional method.
+Also, the new method will be directly usable by all people using this framework following an existing and familiar syntax.
+Contributions to (ML) ecosystems such as \pkg{mlr3} enable users to easily make use of a large variety of additional functionality, from preprocessing over hyperparameter tuning and feature selection to result visualization [@mlr3], which can be used in conjunction with the new resampling method.
+In addition, contributions **most likely undergo a thorough review** from other community members during the implementation, most often enhancing the code quality of the implementation.
+<!-- AB: I do not necessarily agree with this since there is no formalized review process.
+'Most likely' is also not helpful.
+Would you read a paper from a journal that 'most likely' reviews papers thoroughly?
+-->
 A direct contribution also reduces the maintenance load for the ecosystem developers as they do not need to wrap a new method into their system and watch for possible future changes.
 
-On the contrary, people are forced to use the ecosystem and cannot easily use the method outside of it, whether its a different ecosystem or some other way.
+On the contrary, people are forced to use the ecosystem and cannot easily use the method outside of it, whether it is a different ecosystem or some other way.
 They might also be forced to follow certain (coding) guidelines of the ecosystem and potentially feel constrained in their personal way of doing things.

### 2021-09-24 paper/paper.Rmd
minor edits in Conclusion section

@@
 <!-- signoff -->
-Spatiotemporal cross-validation a paradigm that is not yet fully established in scientific workflows although it has been discussed intensively for more than a decade now.
-We anticipate that making the existing methods easily accessible to users is an important step to foster the acceptance of spatiotemporal cross-validation in the community and to allow modelers to produce less biased cross-validation results in environmental and ecological studies.
+Spatiotemporal cross-validation as a paradigm is not yet fully established in scientific workflows although it has been discussed intensively for more than a decade now.
+We anticipate that making the existing methods easily accessible to users is an important step to foster the acceptance of spatiotemporal cross-validation in the community and to allow modelers to produce bias-reduced model assessments in environmental and ecological studies.
 

### 2021-09-24 paper/paper.Rmd
minor edits (mainly terminology) in Sections 1, 2

@@
 
-Although alternatives such as the bootstrap exist and have some advantages [@efron1983; @hand1997], cross-validation is a particularly well-established, easy-to-implement algorithm for *model assessment* of supervised machine-learning models [@efron1983 and next section] and *model selection* [@arlot2010].
+Although alternatives such as the bootstrap exist and have some advantages [@efron1983; @hand1997], cross-validation (CV) is a particularly well-established, easy-to-implement algorithm for *model assessment* of supervised machine-learning models [@efron1983 and next section] and *model selection* [@arlot2010].
 In its basic form, it is based on resampling the data without paying attention to any possible dependence structure, which may arise from, e.g., grouped or structured data, or underlying environmental processes inducing some sort of spatial coherence at the landscape scale.
-In treating dependent observations as independent, or ignoring autocorrelation, cross-validation test samples may in fact be heavily correlated with, or even pseudo-replicates of, the data used for training the model, which introduces a potentially severe bias in assessing the transferability of flexible machine-learning models.
-
-This cross-validation bias is well-known in spatial as well as non-spatial prediction [@brenning2005; @brenning2008; @arlot2010; @roberts2017] and in forecasting [@bergmeir2018].
+In treating dependent observations as independent, or ignoring autocorrelation, CV test samples may in fact be heavily correlated with, or even pseudo-replicates of, the data used for training the model, which introduces a potentially severe bias in assessing the transferability of flexible machine-learning models.
+
+This CV bias is well-known in spatial as well as non-spatial prediction [@brenning2005; @brenning2008; @arlot2010; @roberts2017] and in forecasting [@bergmeir2018].
 It is most easily understood from a predictive modeling perspective by focusing on the question where (and when) the model should be used for prediction.
@@
 If the model was to be applied to adjacent agricultural regions, i.e., outside the learning sample's spatial domain [*extra-domain* prediction, @brenning2005], it would even seem necessary to resample at a higher level of spatial aggregation, i.e.
-at the level of agricultural sub-regions within the learning sample, in order to realistically mimic the actual prediction task.
-The cross-validation resampling needed therefore depends as much on the prediction task itself as on the data structure or dependency at hand.
+at the level of sub-regions within the learning sample, in order to realistically mimic the actual prediction task.
+The CV resampling needed therefore depends as much on the prediction task itself as on the data structure or dependency at hand.
 
@@
 This toolbox, implemented as an open-source R package, builds upon and generalizes several existing toolboxes that have been developed in recent years for more specific settings (Table&nbsp;\@ref(tab:sptcv-methods)).
-The earliest and most comprehensive of these implementations is the \pkg{sperrorest} R package [@brenning2012], which provides an extensible framework and includes predefined resampling strategies based on, for example, blocking, clustering, and buffering.
+The earliest and most comprehensive of these implementations is the \pkg{sperrorest} R package [@brenning2012], which provides an extensible framework and includes predefined resampling strategies based on, for example, geometric blocking, clustering, and buffering.
 In contrast, \pkg{blockCV} and \pkg{ENMeval} were developed for block and buffer resampling with a focus on species distribution modeling [@blockCV; @rest2014; @muscarella2014].
-Neither of these however have been integrated into established machine-learning frameworks such as \pkg{mlr}/\pkg{mlr3} or \pkg{caret}/\pkg{tidymodels}, and all of them lack support of temporal prediction tasks.
+However, neither of these have been integrated into established machine-learning frameworks such as \pkg{mlr}/\pkg{mlr3} or \pkg{caret}/\pkg{tidymodels}, and all of them lack support for temporal prediction tasks.
 The \pkg{CAST} package, in contrast, focuses on spatiotemporal prediction tasks and makes use of some functions of the \pkg{caret} framework [@cast; @meyer2018].
@@
 Next, various spatial and spatiotemporal partitioning techniques are contrasted and compared, before their application is demonstrated in a machine-learning model assessment in the following section.
-Finally, limitations of existing methods are discussed and an outlook is given.
-
-# Spatial and spatiotemporal cross-validation
-
-In cross-validation for predictive model assessment, we consider the following formal setting.
+Finally, recommendations for the selection of suitable resampling techniques are given.
+
+# Spatial and spatiotemporal CV
+
+In CV for predictive model assessment, we consider the following formal setting.
 We are interested in predicting a numerical or categorical response $y$ of an object or instance using a feature vector $\mathbf{x} = (x^{(1)}, \ldots, x^{(p)})^t\in\mathbb{R}^p$ and a model $\hat{f}_\mathcal{L}$ that has been trained on a learning sample $\mathcal{L} = \{(y_i, \mathbf{x}_i),\ i = 1, ..., n\}$.
@@
 The loss function can take a variety of forms such as the misclassification error $I(Y\neq\hat{f}_\mathcal{L}(X))$ in classification, or the squared error $(Y-\hat{f}_\mathcal{L}(X))^2$ in regression, among many other possible measures.
-The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance metrics for regression and classification (see, e.g., @hand1997 for classification, and @hyndman2006 for regression and forecasting tasks).
+The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance measures for regression and classification (see, e.g., @hand1997 for classification, and @hyndman2006 for regression and forecasting tasks).
 
@@
 Since re-using the learning sample $\mathcal{L}$ for testing, i.e.
-$\mathcal{T}:=\mathcal{L}$, would yield the over-optimistic *resubstitution* or *apparent* performance, cross-validation partitions the sample $\mathcal{L}$ into disjoint training and test sets.
+$\mathcal{T}:=\mathcal{L}$, would yield the over-optimistic *resubstitution* or *apparent* performance, CV partitions the sample $\mathcal{L}$ into disjoint training and test sets.
 Specifically, $\mathcal{L}$ is split into $k$ partitions,
@@
 
-In traditional cross-validation, the partitioning is based on uniform random resampling, which ignores spatial or temporal autocorrelation or any existing grouping structure as well as the structure of the prediction task, and may result in over-optimistic performance estimates.
+In traditional CV, the partitioning is based on uniform random resampling, which ignores spatial or temporal autocorrelation or any existing grouping structure as well as the structure of the prediction task, and may result in over-optimistic performance estimates.
 Several approaches have therefore been proposed in the literature and implemented in software to accommodate a variety of predictive situations (Table&nbsp;\@ref(tab:sptcv-methods)).
 
-Approaches based on *spatial blocking* or *grouping* require either the construction of spatial zones, or the use of pre-existing spatial structures in the data.
-Let's refer to these spatial units or zones as $\mathcal{Z}_i$, $1\le i\le n_z$.
-These zones are often constructed to serve as the $k=n_z$ spatial partitions, for example by performing $k$-means clustering of the sample coordinates [@russ2010], which we refer to as *coordinate-based clustering*, or generating the desired number of rectangular blocks.
-The zones may also be defined by a modeler based on an arbitrary (e.g., polygonal) partitioning of the study region, which we refer to as *custom resampling*.
-Similarly, the data can be partitioned in feature space instead of geographic space, which has been referred to as *environmental blocking* [@roberts2017].
+Approaches based on *spatial blocking* (or *grouping*) require either the construction of spatial zones, or the use of pre-existing spatial structures in the data.
+Let's refer to these spatial units or blocks as $\mathcal{Z}_i$, $1\le i\le n_z$.
+These blocks are often constructed to serve as the $k=n_z$ spatial partitions, for example by performing $k$-means clustering of the sample coordinates [@russ2010], which we refer to as *coordinate-based clustering*, or generating the desired number of rectangular blocks as an example of *geometric partitioning*.
+The blocks may also be defined by a modeler based on an arbitrary partitioning of the study region based on an external data source, which we refer to as *custom resampling*.
+This often used when the data is grouped.
+For example, when using to multi-level sampling designs or studying spatial objects, it has been proposed to apply LOO at the site level [@martin2008; @kasurak2011] or, in animal movement studies, at the animal level [@anderson2005].
+We will broadly refer to such groups of observations as 'blocks' in a generic sense, regardless of the shape or origin of the groups.
+Also, the data can be partitioned in feature space instead of geographic space, which has been referred to as "environmental blocking" [@roberts2017].
 
@@
 This approach has been applied, for example, in spatial CV at the agricultural field level [@pena2015].
-We would like to emphasize the conceptual distinction between CV *at the group (or block) level*, referring to this scenario, and the various previously types of block CV, where the blocks or groups themselves define the CV partitions.
-
-A variant of cross-validation is leave-one-out (LOO) cross-validation, which has long been established in geostatistics [@cressie1993], sometimes with a focus on the spatial distribution of LOO error [@willmott2006].
+We would like to emphasize the conceptual distinction between *CV at the block level*, referring to this scenario, and *leave-one-block-out CV*, where the blocks themselves define the CV partitions.
+Figure&nbsp;\@ref(fig:rsmp-schema) gives an overview of the conceptual framework and terminology used in this work.
+
+A variant of CV is leave-one-out (LOO) CV, which has long been established in geostatistics [@cressie1993], sometimes with a focus on the spatial distribution of LOO error [@willmott2006].
 Although this is just a special case of non-spatial CV with $k=n$, it is sometimes also referred to as spatial CV [@willmott2006].
 
-Spatial variants have been proposed that apply an exclusion buffer to the test locations to separate them from the training data [@brenning2005].
-We refer to this resampling scheme as *spatial buffering*.
-One approach that has been proposed for defining a separation distance is to use the range of autocorrelation of model residuals to determine the buffer distance as this ensures independence conditional on the predictors [@brenning2005; @roberts2017].
-When data is grouped, e.g., due to multi-level sampling designs or the study of spatial objects, it has been proposed to apply LOO at the site level [@martin2008; @kasurak2011] or, in animal movement studies, at the animal level [@anderson2005].
-We will broadly refer to such groups of observations as 'blocks' in a generic sense, and therefore refer to this type of LOO-CV as *leave-one-block-out CV*.
-
-It should be noted that $k$-fold cross-validation with a large value of $k$, and LOO-CV in particular ($k=n$), is not only very time-consuming since the model has to be trained $k$ times; these models will also be nearly identical since only a tiny fraction of the data is withheld, and therefore estimation bias increases.
-'Pure' LOO-CV is therefore not recommended for machine-learning model assessment.
+Spatial variants of CV have been proposed that apply an exclusion *buffer* or guard zone to the test locations to separate them from the training data [@brenning2005; @roberts2017].
+One approach that has been proposed for defining a separation distance is to use the range of autocorrelation of model residuals to determine the buffer distance as this seeks to establish independence conditional on the predictors [@brenning2005; @roberts2017].
+
+It should be noted that $k$-fold CV with a large value of $k$, and LOO CV in particular ($k=n$), is not only very time-consuming since the model has to be trained $k$ times; these models will also be nearly identical since only a tiny fraction of the data is withheld, and therefore estimation bias increases.
+'Pure' LOO CV is therefore not recommended for machine-learning model assessment.
 
 In the purely temporal domain, a special case is to leave out temporal observational units (or time slices; leave-time-out or LTO CV), as in leave-one-year-out CV [@anderson2005; @brenning2005].
-Cross-validation and hold-out validation strategies for time series have been discussed more extensively in the forecasting literature, considering also the effects of serial autocorrelation [@bergmeir2018].
+CV and hold-out validation strategies for time series have been discussed more extensively in the forecasting literature, considering also the effects of serial autocorrelation [@bergmeir2018]; these methods are not the focus of the implementation presented in this work.
 
 Turning to prediction tasks with spatiotemporal data, various spatial, temporal, or spatiotemporal partitioning strategies are being used, depending on the specific study objectives.
-While the former two ignore the spatial and temporal dimension of the data, respectively, it has also been proposed to leave out random subsets of locations and time points [@meyer2018] or spatiotemporal clusters [@cluto].
-<!-- AB: This short paragraph does certaintly not adequately describe Meyer's LLTO with its unique buffer in space and time dimensions. But I'm not sure if it's worth the effort. -->
+While the former two ignore the temporal and spatial dimension of the data, respectively, it has also been proposed to leave out random subsets of locations and time points [@meyer2018] or spatiotemporal clusters [@cluto].
+Details of these and other implementations are outlined in the respective subsections of Section&nbsp;\@ref(sec:implementation).
 
@@
 
-# Spatiotemporal partitioning methods and their implementation
+# Spatiotemporal partitioning methods and their implementation {#sec:implementation}
 

### 2021-09-26 paper/paper.Rmd
remove sentence on custom_cv vs cv difference

@@
 In this so-called "Custom CV", users supply a factor variable, each level of which defines a partition.
-<!-- AB: I think we can omit this sentence, otherwise we would have to repeat it for the other LOBO methods as well. Here we're in the leave-ONE-something-out section. -->
-<!--
-This partitioning concept should not be mistaken with "group-level resampling" (section \@ref(sec:group-level)), which conducts additional combination of data splits and performs CV at the block level.
--->
 The factor variable can either be specified through a factor vector of the same length as number of observations, or by passing the name of a feature of type `character` within the task (argument `col`).

### 2021-09-26 paper/paper.Rmd
remove cstv acronym explanation
remove section "contributing new resampling methods"

@@
 
+Contributions of new or missing spatiotemporal resampling methods is encouraged and can be done directly in \pkg{mlr3spatiotempcv}.
+Contributors can orient themselves by looking at the already implemented methods with respect to class structure, active bindings and methods.
+
 ## Spatial leave-one-out
@@
 Because a dynamic image cannot be included in this manuscript, a static version, which can be generated by setting `static_image = TRUE`, is shown.
-
-<!-- AB: No need to justify internal names of methods:
-In \pkg{mlr3spatiotempcv} the method is named `"sptcv_cstf"` which is an acronym originating from the underlying upstream implementation in package \pkg{CAST} `CreateSpacetimeFolds()`.
--->
 
@@
 
-## Contributing new resampling methods
-
-<!-- AB: I do not understand the point of this discussion section.
-It is not clear to the reader if mlr3spatiotempcv supports both approaches or supports/recommends one of them.
-Also, an implementation 'most likely' being reviewed in an unspecified process does not exactly inspire much trust...
--->
-
-<!-- discuss pros and cons of implementing methods in a standalone pkg vs a framework -->
-There are pros and cons to whether a new method should be integrated into an existing modeling framework such as \pkg{mlr3} directly or distributed in its own package.
-When contributing a method to an existing framework, developers can benefit from existing structures and orient themselves towards how previous methods were implemented.
-They do not need to worry about complying with various R package related best practices (such as CRAN policies) and can solely focus on the implementation of an additional method.
-Also, the new method will be directly usable by all people using this framework following an existing and familiar syntax.
-Contributions to (ML) ecosystems such as \pkg{mlr3} enable users to easily make use of a large variety of additional functionality, from preprocessing over hyperparameter tuning and feature selection to result visualization [@mlr3], which can be used in conjunction with the new resampling method.
-In addition, contributions **most likely undergo a thorough review** from other community members during the implementation, most often enhancing the code quality of the implementation.
-<!-- AB: I do not necessarily agree with this since there is no formalized review process.
-'Most likely' is also not helpful.
-Would you read a paper from a journal that 'most likely' reviews papers thoroughly?
--->
-A direct contribution also reduces the maintenance load for the ecosystem developers as they do not need to wrap a new method into their system and watch for possible future changes.
-
-On the contrary, people are forced to use the ecosystem and cannot easily use the method outside of it, whether it is a different ecosystem or some other way.
-They might also be forced to follow certain (coding) guidelines of the ecosystem and potentially feel constrained in their personal way of doing things.
-Often, such ecosystems store their results in specialized classes, which make it hard for users to reuse them outside of the ecosystem.
-
 # Conclusion and outlook

### 2021-09-26 paper/paper.Rmd
remove blank lines
minor edits to Table 1 caption

@@
   \caption[t]{Spatiotemporal resampling methods available in the mlr3 ecosystem.
-    The Name column shows the mlr3 method name as found in the `mlr3::mlr\_resamplings` dictionary.
-    \# pub. indicates the number of studies that were found to have used this resampling technique until May 2021.
-    <!--- Literature appearances were only collected for methods with a "spcv" or "sptcv" prefix as methods which rely on custom grouping ("custom\_cv" and "cv") do not have a dedicated name or identifier to make them detectable in research papers.
---->
+    The "Name" column shows the mlr3 method name as found in the \texttt{mlr3::mlr\_resamplings} dictionary.
+    "\# pub." indicates the number of studies that were found to have used this resampling technique until May 2021.
     For each method, up to three randomly chosen references were added to the table as footnotes.}

### 2021-09-26 paper/paper.Rmd
move spatialAutoRange example to methods section

@@
 
-The `"spcv_block"` method from package \pkg{blockCV} supports both random and systematic resampling of square blocks (`rsmp("spcv_block")` with argument `selection = "random"` and `"systematic"`, respectively; Figure&nbpsp;~\@ref(fig:block-random) and \@ref(fig:block-systematic)).
+The `"spcv_block"` method from package \pkg{blockCV} supports both random and systematic resampling of square blocks (`rsmp("spcv_block")` with argument `selection = "random"` and `"systematic"`, respectively; Figures \@ref(fig:block-random) and \@ref(fig:block-systematic)).
 There are additional options for modeling presence-only data, which is a typical situation in species distribution modeling.
@@
 
-Block sizes (in meters) are determined by the `range` argument.
-Alternatively, rectangular blocks can be created by specifying the number of desired rows and columns (arguments `rows` and `cols`).
+Block sizes (square) (in meters) are determined by the `range` argument.
+Rectangular blocks can be created by specifying the number of desired rows and columns (arguments `rows` and `cols`).
+Due to the non-trivial specification of argument `range` package \pkg{blockCV} provides the helper functions `spatialAutoRange()` and `rangeExplorer()` to conduct a data-driven estimation of the distance at which the spatial autocorrelation within the data levels off [@blockCV].
+According to the package authors, this estimate should then be used for argument `range` to have a sensible value for the block sizes created in method `"spcv_block"`.
+
+It should be noted that rectangular partitioning can be problematic in irregularly shaped study areas as in Figure \@ref(fig:block-random) where some of the resulting partitions may contain substantially fewer observations than others.
 
@@
 
-<!-- spcv_block: how to set argument `range`` -->
-Method `"spcv_block"` requires the user to supply a value for argument `range` which denotes the block size used to create the blocking scheme across the study area.
-This specification is non-trivial and therefore package \pkg{blockCV} provides some helper functions (`spatialAutoRange()` and `rangeExplorer()`) to conduct a data-driven estimation of the distance at which the spatial autocorrelation in the data levels off [@blockCV].
-According to the package authors, this estimate should then be used for argument `range` to have a sensible value for the block sizes created in method `"spcv_block"`.
-Rectangular partitioning can be problematic in irregularly shaped study areas where some of the resulting partitions may contain substantially fewer observations than others.
-
 <!-- sptcv LLO & LTO: requires additional variable to specify groupings -->

### 2021-09-26 paper/paper.Rmd
move sptcv_cstf paragraph from discussion to methods section

@@
 
+The `"spcv_cstf"` methods LLO and LTO (with only one of `space_var` or `time_var` set) require a variable in the dataset which should be used for grouping.
+The specification of the variable(s) which should be used for a spatial, temporal or spatiotemporal grouping is not trivial because the final partitioning should, in the optimal case, ensure that the selected groups inherit substantial autocorrelation within themselves and simultaneously differ substantially from other partitions.
+Also, if the selected variable contains too many groups, the difference within train/test splits may become undesirably high and tend towards a LOO CV [@meyer2018].
+
 ```{r results='hide'}
@@
 
-<!-- sptcv LLO & LTO: requires additional variable to specify groupings -->
-The `"spcv_cstf"` methods (with only one of `space_var` or `time_var` set) require a variable in the dataset which should be used for grouping.
-The specification of such is not always easy because the grouping should ensure that the selected groups inherit substantial autocorrelation within themselves and simultaneously differ substantially from other groups.
-Also, if the selected variable contains too many groups, the difference within train/test splits may become undesirably high and tend towards a LOO CV [@meyer2018].
-
 <!-- only some methods are suited for repeated execution -->

### 2021-09-26 paper/paper.Rmd
Figure&nbsp; -> Figure

@@
 We would like to emphasize the conceptual distinction between *CV at the block level*, referring to this scenario, and *leave-one-block-out CV*, where the blocks themselves define the CV partitions.
-Figure&nbsp;\@ref(fig:rsmp-schema) gives an overview of the conceptual framework and terminology used in this work.
+Figure \@ref(fig:rsmp-schema) gives an overview of the conceptual framework and terminology used in this work.
 
@@
 
-At the most general level, resampling methods are categorized according to the level at which the data is partitioned and resampled (see Figure&nbsp;\@ref(fig:rsmp-schema)):
+At the most general level, resampling methods are categorized according to the level at which the data is partitioned and resampled (see Figure \@ref(fig:rsmp-schema)):
 
@@
 locations where a species was observed and not observed, respectively, and therefore spatial LOO CV in its usual sense can be carried out.
-Figure&nbsp;\@ref(fig:buffer-eval) shows the first test fold generated with this method for presence/absence data with a buffer distance of 1000 m.
+Figure \@ref(fig:buffer-eval) shows the first test fold generated with this method for presence/absence data with a buffer distance of 1000 m.
 
@@
 Note that the actual number of folds obtained may be smaller than `nsplit[1]*nsplit[2]` (or smaller than what would be expected based on `dsplit`) since some blocks may be empty or (optionally) merged into adjacent folds.
-In the example, there are only eleven folds instead of twelve because the southwestern part of the study area's bounding box does not contain observations  (Figure&nbsp;\@ref(fig:tile)).
+In the example, there are only eleven folds instead of twelve because the southwestern part of the study area's bounding box does not contain observations  (Figure \@ref(fig:tile)).
 
@@
 
-In the `cookfarm` example, we reduce the `Date` variable to five unique levels for better visualization, and then create a spatiotemporal regression task in \pkg{mlr3spatiotempcv} (Figure&nbsp;\@ref(fig:lto)).
+In the `cookfarm` example, we reduce the `Date` variable to five unique levels for better visualization, and then create a spatiotemporal regression task in \pkg{mlr3spatiotempcv} (Figure \@ref(fig:lto)).
 
@@
 In contrast to LTO, the LLO method randomly resamples locations that may, for example, correspond to time series.
-The sampled locations form the test partition while the temporal information is ignored (Figure&nbsp;\@ref(fig:llo)).
+The sampled locations form the test partition while the temporal information is ignored (Figure \@ref(fig:llo)).
 Unlike spatial CV methods that are based on geometric regions or the clustering of coordinates, the sampled test locations bear no particular spatial relationship.
@@
 
-In LLTO, a test set is first randomly sampled from the data set, and then all observations that correspond to the same location or time point are omitted from the training sample (Figure&nbsp;\@ref(fig:llto)).
+In LLTO, a test set is first randomly sampled from the data set, and then all observations that correspond to the same location or time point are omitted from the training sample (Figure \@ref(fig:llto)).
 LLTO resampling mimics the situation where a model is trained on time series data from a number of locations and time points, and used to predict the time series at other locations and time points that are not included in the training sample.
@@
 
-Finally, we visualize (two of) the partitions that were used during performance estimation making use of the generic `autoplot()` function in package \pkg{mlr3spatiotempcv} (Figure&nbsp;\@ref(fig:vis-spcv-eval)).
+Finally, we visualize (two of) the partitions that were used during performance estimation making use of the generic `autoplot()` function in package \pkg{mlr3spatiotempcv} (Figure \@ref(fig:vis-spcv-eval)).
 

### 2021-09-26 paper/paper.Rmd
remove cv vs custom_cv paragraph

@@
 
-<!-- cv vs custom_cv -->
-\pkg{mlr3spatiotempcv} supports two factor-based partitioning schemes: `"cv"` (with column role `"groups"` set) and `"custom_cv"`.
-Both are similar in that they require external user input for partition creation, i.e. a factor variable which assigns all observations into distinct groups.
-The idea of `"custom_cv"` is that each group becomes a test set in the CV (leave-one-group-out CV).
-By contrast, method `"cv"` (with column role `"groups"` set) resamples the initial groups and performs CV at the block level.
-The important point for method "cv" is that observations from a specific group will either be assigned to the train or test set and never be split up.
-In contrast to method `"custom_cv"`, this approach can be repeated multiple times because multiple different fold combinations are possible.
-However, it also requires the number of initial groups to be (substantially) higher than the number of folds so that a sensible combination is possible in the first place.
-
 <!-- spcv coords: mention generic name -->

### 2021-09-26 paper/paper.Rmd
Table&nbsp; -> Table

@@
 
-This toolbox, implemented as an open-source R package, builds upon and generalizes several existing toolboxes that have been developed in recent years for more specific settings (Table&nbsp;\@ref(tab:sptcv-methods)).
+This toolbox, implemented as an open-source R package, builds upon and generalizes several existing toolboxes that have been developed in recent years for more specific settings (Table \@ref(tab:sptcv-methods)).
 The earliest and most comprehensive of these implementations is the \pkg{sperrorest} R package [@brenning2012], which provides an extensible framework and includes predefined resampling strategies based on, for example, geometric blocking, clustering, and buffering.
@@
 In traditional CV, the partitioning is based on uniform random resampling, which ignores spatial or temporal autocorrelation or any existing grouping structure as well as the structure of the prediction task, and may result in over-optimistic performance estimates.
-Several approaches have therefore been proposed in the literature and implemented in software to accommodate a variety of predictive situations (Table&nbsp;\@ref(tab:sptcv-methods)).
+Several approaches have therefore been proposed in the literature and implemented in software to accommodate a variety of predictive situations (Table \@ref(tab:sptcv-methods)).
 
@@
 
-\pkg{mlr3spatiotempcv} currently implements the partitioning methods identified in Table&nbsp;\@ref(tab:sptcv-methods) with their \pkg{mlr3} class name.
+\pkg{mlr3spatiotempcv} currently implements the partitioning methods identified in Table \@ref(tab:sptcv-methods) with their \pkg{mlr3} class name.
 Several of the implemented algorithms are themselves versatile toolboxes with multiple options.
@@
 This can be a pretty effective way to reduce autocorrelation but also requires to discard many observations.
-In addition it comes with the downsides of a LOO-CV: many model fits which can be computationally very demanding and the results inheriting a high variance due to the high differences in the test sets (also commonly known as the "bias-variance tradeoff" [@wahba1998; @blockCV].
-If a large buffer size is chosen relative to the size of the study area, a large portion of it will remain unsampled.
+However, this approach also comes with the downsides of a LOO-CV: many model fits which can be computationally very demanding and the results may inherit a high variance due to the high differences in the test sets (also commonly known as the "bias-variance tradeoff" [@wahba1998; @blockCV].
+Also if a large buffer size is chosen relative to the size of the study area, a large portion of it will remain unsampled.
 One should try to refrain from selecting a value based on subjective notions with respect to study area characteristics.
@@
 
-<!-- signoff -->
+<!-- sign-off -->
 Spatiotemporal cross-validation as a paradigm is not yet fully established in scientific workflows although it has been discussed intensively for more than a decade now.

### 2021-09-26 paper/paper.Rmd
Section&nbsp; -> Section

@@
 While the former two ignore the temporal and spatial dimension of the data, respectively, it has also been proposed to leave out random subsets of locations and time points [@meyer2018] or spatiotemporal clusters [@cluto].
-Details of these and other implementations are outlined in the respective subsections of Section&nbsp;\@ref(sec:implementation).
+Details of these and other implementations are outlined in the respective subsections of Section  \@ref(sec:implementation).
 
@@
 In $k$-fold CV resampling at the block level, there are therefore $k$ partitions, each consisting of $1/k$-th of the blocks.
-The special case in which $k$ equals the number of blocks, CV at the block level simply becomes leave-one-block-out CV, for which dedicated implementations exist (Section&nbsp;\@ref(sec:lobo-cv)).
+The special case in which $k$ equals the number of blocks, CV at the block level simply becomes leave-one-block-out CV, for which dedicated implementations exist (Section  \@ref(sec:lobo-cv)).
 
@@
 -->
-<!-- In addition, irregularly shaped blocks are available (3) through coordinate-based clustering as a leave-one-block-out technique from \pkg{sperrorest} (method `"spcv_coords"`, Section&nbsp;\@ref(sec:cluster-cv)), and (4) as a CV at the block level where blocks are defined by a factor variable that defines zones or groups (method `"custom_cv"` in \pkg{mlr3}, Section&nbsp;\@ref(sec:custom-cv)).
+<!-- In addition, irregularly shaped blocks are available (3) through coordinate-based clustering as a leave-one-block-out technique from \pkg{sperrorest} (method `"spcv_coords"`, Section  \@ref(sec:cluster-cv)), and (4) as a CV at the block level where blocks are defined by a factor variable that defines zones or groups (method `"custom_cv"` in \pkg{mlr3}, Section  \@ref(sec:custom-cv)).
 -->
@@
 
-In other application scenarios such as the crop classification example given in section&nbsp;\@ref(sec:intro) [@pena2015], the objective is to predict the fruit-tree crop type of 'new', unseen fields within an agricultural region.
+In other application scenarios such as the crop classification example given in Section  \@ref(sec:intro) [@pena2015], the objective is to predict the fruit-tree crop type of 'new', unseen fields within an agricultural region.
 In this use case we are not at all interested in "predicting" the already known crop type of (other) grid cells within the same agricultural field.

### 2021-09-26 paper/paper.Rmd
discernables -> detectability

@@
 Conceptually, LLTO applies zero-distance buffering in both space and time: The buffer zones consist of all observations whose distance to the test sample in either space or time equals zero.
-In the mathematical sense, however, this buffering is not based on a valid metric (or distance function) in three-dimensional space (2D + time) as neither the identity of discernables nor the triangle inequality are satisfied by the underlying combined 'distance' measure.
+In the mathematical sense, however, this buffering is not based on a valid metric (or distance function) in three-dimensional space (2D + time) as neither the identity of detectability nor the triangle inequality are satisfied by the underlying combined 'distance' measure.
 Also note that LLTO does not 'combine' LTO with LLO, as neither of these applies a buffer zone.

### 2021-09-26 paper/paper.Rmd
more &nbsp replacements
spell check

@@
 This approach is better suited for the irregular shape of the present study area than a rectangular partitioning.
-Figures&nbsp;\@ref(fig:vis-spcv-eval)) and&nbsp;\@ref(fig:vis-nspcv-eval)) show the contrasting distributions of training and test samples.
+Figures \@ref(fig:vis-spcv-eval)) and \@ref(fig:vis-nspcv-eval)) show the contrasting distributions of training and test samples.
 For demonstration purposes we use only four CV folds and two repetitions.
@@
 
-Although the case study example in section \@ref(sec:case-study) used the `"spcv_coords"` method for coordindate-based clustering, this should not give the impression that this method is the only method suitable for this example task.
+Although the case study example in section \@ref(sec:case-study) used the `"spcv_coords"` method for coordinate-based clustering, this should not give the impression that this method is the only method suitable for this example task.
 In this application setting, we want to assess how well the model generalized from the concrete set of historical landslide occurrences, which is why we ensured that training and test sets contain different, "new" hillslopes and landslides.

### 2021-09-26 paper/paper.Rmd
remove outdated commented paragraph

@@
 
-<!-- Spatial blocking divides the study area into different zones [russ2010; @bahn2012; @wenger2012; @blockCV]. -->
-<!-- Several methods have been implemented to create rectangular as well as irregularly shaped blocks.  -->
-<!-- These blocks can be used for leave-one-block-out resampling, or they can themselves be grouped into $k$ partitions for a $k$-fold CV at the block level. -->
-
-<!-- \pkg{mlr3spatiotempcv} supports multiple resampling methods that are based on blocks:  -->
-<!-- (1) "Spatial blocking" from package \pkg{blockCV} implements $k$-fold CV resampling at the level of square or rectangular blocks, while (2) "Spatial tiles" from \pkg{sperrorest} offers leave-one-block-out resampling with square or rectangular, possibly rotated blocks.
--->
-<!-- In addition, irregularly shaped blocks are available (3) through coordinate-based clustering as a leave-one-block-out technique from \pkg{sperrorest} (method `"spcv_coords"`, Section  \@ref(sec:cluster-cv)), and (4) as a CV at the block level where blocks are defined by a factor variable that defines zones or groups (method `"custom_cv"` in \pkg{mlr3}, Section  \@ref(sec:custom-cv)).
--->
-
 # Step-by-step example: Comparing spatial and non-spatial CV {#sec:case-study}

### 2021-09-26 paper/paper.Rmd
remove note regarding k-means clustering for custom_cv example

@@
 In contrast to geometric or clustering-based blocks, the spatial or temporal location is not used explicitly, but rather implicitly through the spatial or spatiotemporal footprint of each user-defined block.
-<!-- The following sentence is really a repetition from the general background section, and similar statements could be made in the custom_cv section. -> omit this sentence  -->
-<!--
-additional metadata is exploited such as the membership in a specific agricultural management unit [@brenning2006], or in a site in hierarchical sampling [@kasurak2011].
--->
 
 The following example uses $k$-means clustering to generate classes that are used as blocks.
-<!--
-Note that in practice usually the group creation would not be done in an unsupervised way but by making use of an inherent grouping information from the dataset.
--->
 To underline the honoring of the groups, we chose a number of groups (eight) that is not a multiple of the number of folds (three).

### 2021-09-27 paper/paper.Rmd
rephrased statement on user-contributed methods

@@
 
-Contributions of new or missing spatiotemporal resampling methods is encouraged and can be done directly in \pkg{mlr3spatiotempcv}.
-Contributors can orient themselves by looking at the already implemented methods with respect to class structure, active bindings and methods.
+Users are encouraged to contribute new or missing spatiotemporal resampling methods directly to \pkg{mlr3spatiotempcv}.
+The already implemented methods can be inspected to get to know the class structure, active bindings and methods.
 

### 2021-09-27 paper/paper.Rmd
edited discussion, removed commented-out text

@@
 
-<!-- AB: I don't think the following explanation is needed: -->
-<!--
-Even though the buffer creates a circular area which defines the test partition, it is not categorized as a "geometric" method as the origin of the circle is observation-based.
-In contrast, "geometric" methods do not take into account single observation locations but only focus on the study area extend and will include all observations which fall into their respective covered area.
--->
-
 ```{r disc-non-eval, eval=FALSE}
@@
 This method is referred to as "leave-location-and-time-out" (LLTO) in package \pkg{CAST}.
-<!-- AB: The following information is kinda obvious: -->
-<!-- It requires both space coordinates and time point identifiers that are used to locate observations. --> 
 Additional methods that resample in the temporal and spatial domain only are named "leave-time-out" (LTO) and "leave-location-out" (LLO), respectively, in \pkg{CAST}.
@@
 - Will it be applied to predict into the future, or hindcast gaps between spatiotemporal observations in the past?
+- Is it necessary to impose a separation distance or prediction horizon as a spatial or temporal buffer between training and prediction locations?
 - How densely are the observations distributed in space and time? Are they more densely distributed that the intended spatial or temporal prediction distance?
@@
 
-## Comparing and discussing spatiotemporal resampling methods
+## Additional practical issues
 
-Even though the prediction purpose should be the main decision criteria for choosing an appropriate spatiotemporal resampling method, we would like to discuss the individual methods in more detail in the following section.
+Since \pkg{mlr3spatiotempcv} harvests already implemented resampling methods from existing R packages, the broader overview presented in this work has highlighted that there are still several gaps that may need to be closed in the future, if specific use cases require those features.
 
-<!-- high autocor on the border of test sets -->
-<!-- sptcv LLTO: discards many observations -->
-All methods which make use of all variables in the dataset suffer from the problem that on the border between train and test sets, both in the spatial and temporal domain, autocorrelation is relatively high (unless the dataset inherits natural groupings or gaps between train and test sets).
-Methods that omit some observations (`"spcv_buffer"`, `"spcv_tiles"`, `"sptcv_cstf"` with both `space_var` and `time_var` set) are able to reduce the impact of autocorrelation in scenarios without a clear grouping or sufficient distance between train and test set observations.
-The downside is the removal of observations from the dataset which possibly reduces the power of the trained model and by this, depending on the specific observations that were removed, adds substantial bias to the CV results.
+<!-- risks of using buffer zone -->
+For example, buffering, or the use of a spatial or temporal separation distance between training and test sets, is currently only implemented for some methods (`"spcv_buffer"`, `"spcv_disc"`, and `"sptcv_cstf"` with both `space_var` and `time_var`).
+Its use should, however, be limited to use cases involving a prediction distance, as a buffer zone reduces the size of the training sample and introduces the risk of geographically biased training data.
 
@@
 <!-- spcv_coords: limited variance due to limited k-means possibilities -->
-For repeated executions of method `"spcv_coords"`, the variance between repetitions can become quite small due to the limited possibilities of differing $k$-means-based clusters on the same dataset [@brenning2012].
-This applies even more to `"spcv_block"` with options `selection = "systematic"` and `selection = "checkerboard"` as the fold assignments are the same in each repetition.
-In general, only methods making use of a random component (e.g. `"spcv_coords"` or `"spcv_block"` with option `selection = "random"`) may be suited for using multiple repetitions.
-On the upside, `"spcv_coords"` works well with irregularly shaped study areas due to its ability to create partitions which adjust well to the shape of the study area.
-
-<!-- spcv buffer and spcv disc: discards observations, LOO very expensive -->
-Methods `"spcv_buffer"` and `"spcv_disc"` (optionally) make use of the removal of observations to reduce the influence of spatial autocorrelation between train and test set.
-This can be a pretty effective way to reduce autocorrelation but also requires to discard many observations.
-However, this approach also comes with the downsides of a LOO-CV: many model fits which can be computationally very demanding and the results may inherit a high variance due to the high differences in the test sets (also commonly known as the "bias-variance tradeoff" [@wahba1998; @blockCV].
-Also if a large buffer size is chosen relative to the size of the study area, a large portion of it will remain unsampled.
-One should try to refrain from selecting a value based on subjective notions with respect to study area characteristics.
-
-<!-- spcv coords: mention generic name -->
-In @brenning2012, which is the reference commonly used for the $k$-means-based spatial partitioning approach, the method was not introduced with a unique name.
-Over time, studies which have used this method mainly referenced it using the generic "spatial CV" terminology.
-However, using this notation potentially introduces confusion as the term "spatial CV" is of a generic nature which refers to the idea of accounting for spatial autocorrelation in spatial datasets and does not necessarily refer to a specific partitioning method in the first place.
-Hence, in package \pkg{mlr3spatiotempcv} this method was named `"spcv_coords"`, making use of the fact that the spatial coordinates are used for a cluster-based partition creation.
+CV is often executed repeatedly to reduce the possible influence of random variability on CV estimates.
+In general, only methods that involve a random mechanism for generating or resampling blocks are suited for this.
+In leave-one-block-out CV, coordinate-based and environmental clustering (`"spcv_coords"`, `"spcv_env"` and `"sptcv_cluto"`) achieve this as their clusters are generated based on random seeds.
+However, experience with `"spcv_coords"` shows that clusters from repeated executions may in some situations be nearly identical to each other, resulting in very little variability between CV repetitions.
+This issue is even more critical in CV at the block level with `"spcv_block"` with options `selection = "systematic"` and `selection = "checkerboard"` because identical folds are assigned in each repetition.
+In contrast, `"spcv_block"` with option `selection = "random"` avoids this problem.
 

### 2021-09-27 paper/paper.Rmd
edited abstract

@@
 abstract: >
-  The R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is an extension package of the machine-learning framework [mlr3](https://mlr3.mlr-org.com).
-  It adds support for spatiotemporal resampling methods and their visualization.
-  Spatiotemporal resampling methods are needed to assess model performance in a variety of predictive situations in which spatial, temporal, or spatiotemporal autocorrelation is present.
-  Cross-validation performance estimates may be biased when this is ignored.
+  Spatial and spatiotemporal machine-learning models require a suitable framework for their model assessment, model selection, and hyperparameter tuning, in order to avoid error estimation bias and over-fitting.
+  This contribution reviews the state-of-the-art in spatial and spatiotemporal CV, and introduces the R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) as an extension package of the machine-learning framework [mlr3](https://mlr3.mlr-org.com), demonstrating its use in a step-by-step example.
   Currently various R packages implementing different spatiotemporal partitioning strategies exist: sperrorest, blockCV, CAST.
@@
   This is made possible by integrating the package directly into the [mlr3](https://mlr3.mlr-org.com) machine-learning framework, which already has support for generic non-spatiotemporal resampling methods such as random partitioning.
-  This simplifies the step of integrating package-specific syntax into an overarching machine-learning pipeline, making it easier for users to choose from a variety of different spatiotemporal resampling methods.
-  The users' decision to use a specfic resampling method should depend on the predictive task at hand, the autocorrelation within the data, and the structure of the sampling design used to collect the data.
+  This simplifies the the integration of package-specific syntax into an overarching machine-learning pipeline, making it easier for users to choose from a variety of spatiotemporal resampling methods.
+    The users' decision to use a specfic resampling method should depend on the predictive task at hand, the autocorrelation within the data, and the spatial structure of the sampling design or geographic objects being studied.
 keywords:

### 2021-10-14 paper/paper.Rmd
add some more context to similarity of repeated executions for spcv_env and sptcv_cluto

@@
 However, experience with `"spcv_coords"` shows that clusters from repeated executions may in some situations be nearly identical to each other, resulting in very little variability between CV repetitions.
+While this effect also depends on the variable used for clustering, similar effects could potentially also apply to `"spcv_env"` and `"sptcv_cluto"` methods.
+However, such effects are more difficult to quantify because selected features of these methods are always different in contrast to `"spcv_coords"` which always uses coordinates for clustering.
 This issue is even more critical in CV at the block level with `"spcv_block"` with options `selection = "systematic"` and `selection = "checkerboard"` because identical folds are assigned in each repetition.

### 2021-10-15 paper/paper.Rmd
final corrections

@@
   Spatial and spatiotemporal machine-learning models require a suitable framework for their model assessment, model selection, and hyperparameter tuning, in order to avoid error estimation bias and over-fitting.
-  This contribution reviews the state-of-the-art in spatial and spatiotemporal CV, and introduces the R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) as an extension package of the machine-learning framework [mlr3](https://mlr3.mlr-org.com), demonstrating its use in a step-by-step example.
-  Currently various R packages implementing different spatiotemporal partitioning strategies exist: sperrorest, blockCV, CAST.
+  This contribution reviews the state-of-the-art in spatial and spatiotemporal CV, and introduces the R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) as an extension package of the machine-learning framework [mlr3](https://mlr3.mlr-org.com).
+  Currently various R packages implementing different spatiotemporal partitioning strategies exist: blockCV, CAST, skmeans and sperrorest.
   The goal of [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is to gather the available spatiotemporal resampling methods in R and make them available to users through a simple and common interface.
   This is made possible by integrating the package directly into the [mlr3](https://mlr3.mlr-org.com) machine-learning framework, which already has support for generic non-spatiotemporal resampling methods such as random partitioning.
-  This simplifies the the integration of package-specific syntax into an overarching machine-learning pipeline, making it easier for users to choose from a variety of spatiotemporal resampling methods.
-    The users' decision to use a specfic resampling method should depend on the predictive task at hand, the autocorrelation within the data, and the spatial structure of the sampling design or geographic objects being studied.
+  One advantage is the use of a consistent nomenclature in an overarching machine-learning toolkit instead of a varying package-specific syntax, making it easier for users to choose from a variety of spatiotemporal resampling methods.
+  This package avoids giving recommendations which method to use in practice as this decision depends on the predictive task at hand, the autocorrelation within the data, and the spatial structure of the sampling design or geographic objects being studied.
 keywords:
@@
 
-Although alternatives such as the bootstrap exist and have some advantages [@efron1983; @hand1997], cross-validation (CV) is a particularly well-established, easy-to-implement algorithm for *model assessment* of supervised machine-learning models [@efron1983 and next section] and *model selection* [@arlot2010].
-In its basic form, it is based on resampling the data without paying attention to any possible dependence structure, which may arise from, e.g., grouped or structured data, or underlying environmental processes inducing some sort of spatial coherence at the landscape scale.
-In treating dependent observations as independent, or ignoring autocorrelation, CV test samples may in fact be heavily correlated with, or even pseudo-replicates of, the data used for training the model, which introduces a potentially severe bias in assessing the transferability of flexible machine-learning models.
+Although alternatives such as the bootstrap exist since some decades [@efron1983; @hand1997], cross-validation (CV) is a particularly well-established, easy-to-implement algorithm for *model assessment* of supervised machine-learning models [@efron1983 and next section] and *model selection* [@arlot2010].
+In its basic form, CV is based on resampling the data without paying attention to any possible dependence structure, which may arise from, e.g., grouped or structured data, or underlying environmental processes inducing some sort of spatial coherence at the landscape scale.
+In treating dependent observations as independent, or ignoring autocorrelation, CV test samples may in fact be heavily correlated with, or even pseudo-replicates of, the data used for training the model, which introduces a potentially severe bias in assessing the transferability of flexible machine-learning (ML) models.
 
@@
 The purpose of training a model on this particular sample is to make predictions on other, new fields within the same geographic domain [*intra-domain* prediction, @brenning2005] --- not *within* the same field, which obviously presents only a single crop type that is already known from the training sample.
-In this specific situation it would therefore seem rather unwise to train a model on a simple random subsample of grid cells, and to test it on the remaining data, using other grid cells from the same fields, as if we wanted to predict within a field.
+In this specific situation it would therefore seem rather unwise to train a model on a simple random subsample of grid cells, and to test it on the remaining data, using other grid cells from the same fields, as if one wanted to predict within a field.
 The results from this performance assessment would be over-optimistic, and perhaps badly so.
@@
 
-While it is not the purpose of this article to recommend specific resampling schemes for specific use cases, this example may suffice to motivate the use of appropriate spatial and spatiotemporal cross-validation techniques, and the need for a unified framework and computational toolbox that accommodate a variety of prediction tasks that may be applicable to a broad range of application scenarios.
+While it is not the purpose of this article to recommend specific resampling schemes for specific use cases, the example from above may suffice to motivate the use of appropriate spatial and spatiotemporal cross-validation techniques, and the need for a unified framework and computational toolbox that accommodate a variety of prediction tasks that may be applicable to a broad range of application scenarios.
 \pkg{mlr3spatiotempcv} is such a toolbox.
@@
 This toolbox, implemented as an open-source R package, builds upon and generalizes several existing toolboxes that have been developed in recent years for more specific settings (Table \@ref(tab:sptcv-methods)).
-The earliest and most comprehensive of these implementations is the \pkg{sperrorest} R package [@brenning2012], which provides an extensible framework and includes predefined resampling strategies based on, for example, geometric blocking, clustering, and buffering.
-In contrast, \pkg{blockCV} and \pkg{ENMeval} were developed for block and buffer resampling with a focus on species distribution modeling [@blockCV; @rest2014; @muscarella2014].
+The earliest and most comprehensive of these implementations is the \pkg{sperrorest} R package [@brenning2012], which provides an extensible framework and includes predefined resampling strategies based on geometric blocking, clustering, and buffering.
+In contrast, packages \pkg{blockCV} and \pkg{ENMeval} were developed for block and buffer resampling with a focus on species distribution modeling [@blockCV; @rest2014; @muscarella2014].
 However, neither of these have been integrated into established machine-learning frameworks such as \pkg{mlr}/\pkg{mlr3} or \pkg{caret}/\pkg{tidymodels}, and all of them lack support for temporal prediction tasks.
 The \pkg{CAST} package, in contrast, focuses on spatiotemporal prediction tasks and makes use of some functions of the \pkg{caret} framework [@cast; @meyer2018].
-One limitation of these packages is the sole focus on model assessment, while the proposed implementation within the \pkg{mlr3} framework offers a seamless integration into model selection.
+One limitation of all these packages is the sole focus on model assessment, while the proposed implementation within the \pkg{mlr3} framework also offers seamless integration into model selection and provides parallel execution and enhanced logging abilities.
 It is worth noting that a spatial cross-validation library named \pkg{spacv} has recently been developed for Python3, which can be used with the \pkg{scikit-learn} machine-learning framework [@pedregosa2011].
@@
 This package is furthermore equipped with a variety of two- and three-dimensional visualization capabilities.
-Our hope is that this implementation will facilitate reproducible geospatial modeling and code-sharing across a broad range of application domains.
+The hope is that this implementation will simplify and facilitate reproducible geospatial modeling and code-sharing across a broad range of application domains.
 
-The purpose of this paper is to give an overview of the methods implemented in the R package \pkg{mlr3spatiotempcv}.
+The purpose of this article is to give an overview of the methods implemented in the R package \pkg{mlr3spatiotempcv}.
 After presenting the conceptual background in the following section, the overall structure of the \pkg{mlr3spatiotempcv} package is outlined.
@@
 
-In CV for predictive model assessment, we consider the following formal setting.
-We are interested in predicting a numerical or categorical response $y$ of an object or instance using a feature vector $\mathbf{x} = (x^{(1)}, \ldots, x^{(p)})^t\in\mathbb{R}^p$ and a model $\hat{f}_\mathcal{L}$ that has been trained on a learning sample $\mathcal{L} = \{(y_i, \mathbf{x}_i),\ i = 1, ..., n\}$.
-Our goal is to estimate the expected value of the performance of $\hat{f}_\mathcal{L}$,
+In CV for predictive model assessment, the following formal setting is considered.
+The interest is in predicting a numerical or categorical response $y$ of an object or instance using a feature vector $\mathbf{x} = (x^{(1)}, \ldots, x^{(p)})^t\in\mathbb{R}^p$ and a model $\hat{f}_\mathcal{L}$ that has been trained on a learning sample $\mathcal{L} = \{(y_i, \mathbf{x}_i),\ i = 1, ..., n\}$.
+The goal is to estimate the expected value of the performance of $\hat{f}_\mathcal{L}$,
 \[
@@
 
-Since we only have a sample $\mathcal{T}$ of test data drawn from the population, we can only *estimate* the conditional performance of $\hat{f}_\mathcal{L}$:
+Since there is only a sample $\mathcal{T}$ of test data drawn from the population, one can only *estimate* the conditional performance of $\hat{f}_\mathcal{L}$:
 \[
@@
 \]
-where $f$ is a machine-learning algorithm, i.e. a mapping that trains a model $\hat{f}_\mathcal{S}$ using any suitable training sample $\mathcal{S}$.
+where $f$ is a ML algorithm, i.e. a mapping that trains a model $\hat{f}_\mathcal{S}$ using any suitable training sample $\mathcal{S}$.
 The use of $k=5$ or $k=10$ folds is most commonly seen in practice, and these preferences are also supported by theory [@bengio2004; @cawley2010].
@@
 
-Approaches based on *spatial blocking* (or *grouping*) require either the construction of spatial zones, or the use of pre-existing spatial structures in the data.
+Approaches based on *spatial blocking* (or somtimes calld *grouping*) require either the construction of spatial zones, or the use of pre-existing spatial structures in the data.
 Let's refer to these spatial units or blocks as $\mathcal{Z}_i$, $1\le i\le n_z$.
-These blocks are often constructed to serve as the $k=n_z$ spatial partitions, for example by performing $k$-means clustering of the sample coordinates [@russ2010], which we refer to as *coordinate-based clustering*, or generating the desired number of rectangular blocks as an example of *geometric partitioning*.
+These blocks are often constructed to serve as the $k=n_z$ spatial partitions, for example by performing $k$-means clustering of the sample coordinates [@russ2010], which we refer to as *coordinate-based clustering*; or generating the desired number of rectangular blocks as an example of *geometric partitioning*.
 The blocks may also be defined by a modeler based on an arbitrary partitioning of the study region based on an external data source, which we refer to as *custom resampling*.
@@
 We will broadly refer to such groups of observations as 'blocks' in a generic sense, regardless of the shape or origin of the groups.
-Also, the data can be partitioned in feature space instead of geographic space, which has been referred to as "environmental blocking" [@roberts2017].
+Also, data can be partitioned in feature space instead of geographic space, which has been referred to as "environmental blocking" [@roberts2017].
 
@@
 
-A variant of CV is leave-one-out (LOO) CV, which has long been established in geostatistics [@cressie1993], sometimes with a focus on the spatial distribution of LOO error [@willmott2006].
+One variant of CV is leave-one-out (LOO) CV, which has long been established in geostatistics [@cressie1993], sometimes with a focus on the spatial distribution of LOO error [@willmott2006].
 Although this is just a special case of non-spatial CV with $k=n$, it is sometimes also referred to as spatial CV [@willmott2006].
@@
 
-With the increased awareness of the importance of spatial and spatiotemporal resampling strategies and the growing popularity of R in environmental modeling and geocomputation, it is important to equip machine-learning frameworks such as \pkg{mlr3} with suitable algorithms.
-In this context, the \pkg{mlr3} ecosystem stands out as a unified, object-oriented and extensible framework designed to accommodate numerous machine-learning tasks with a variety of learners, feature and model selection tools, and model assessment capabilities [@mlr3; @mlr3book].
+With the increased awareness of the importance of spatial and spatiotemporal resampling strategies and the growing popularity of R in environmental modeling and geocomputation, it is important to equip ML frameworks such as \pkg{mlr3} with suitable algorithms.
+In this context, the \pkg{mlr3} ecosystem stands out as a unified, object-oriented and extensible framework designed to accommodate numerous ML tasks with a variety of learners, feature and model selection tools, and model assessment capabilities [@mlr3; @mlr3book].
 All of these are supported by advanced visualization tools, which are particularly important in a spatial and spatiotemporal setting.
@@
 With its integrative approach and its aim to provide long-term support, \pkg{mlr3} overcomes the challenges of combining multiple specialized packages with poorly standardized interfaces.
-Issues that practitioners often face include varying argument lists of learners, different return values of `predict` methods, and support for only specific feature types.
-These challenges result in substantial overhead and possible reproducibility issues, which are exacerbated by asynchronous development timelines of different components of the machine-learning pipelines.
+Issues that practitioners often face include varying argument lists of learners, different return values of `predict()` methods, and support for only specific feature types.
+These challenges result in substantial overhead and possible reproducibility issues, which are exacerbated by asynchronous development timelines of different components of the used ML pipelines.
 
-Within the \pkg{mlr3} ecosystem, partitioning strategies are represented by their own objects of class \texttt{Resampling}, most of which are available within \pkg{mlr3} itself (e.g., random CV); other specialized strategies (e.g., \texttt{TaskClassifST} or \texttt{TaskRegrST}) are defined in extension packages such as \pkg{mlr3spatiotempcv}.
-In the machine-learning pipeline, these objects define the data splits used for model assessment and selection (hyperparameter tuning) by machine-learning algorithms.
+Within the \pkg{mlr3} ecosystem, partitioning strategies are represented by their own objects of class \texttt{Resampling}, most of which are available within \pkg{mlr3} itself (e.g., random CV); other specialized strategies are defined in extension packages such as \pkg{mlr3spatiotempcv}.
+In the ML pipeline, these objects define the data splits used for model assessment and selection (hyperparameter tuning) by ML algorithms.
 Spatial and spatiotemporal partitioning techniques in \pkg{mlr3spatiotempcv} are currently mostly imported and interfaced from other packages, in particular \pkg{sperrorest}, \pkg{blockCV} and \pkg{CAST} [@brenning2012; @blockCV; @cast], in order to expose them to \pkg{mlr3} functionality.
-To reduce dependencies, some methods were re-implemented instead of importing them from upstream packages.
+To reduce dependencies, some methods were re-implemented instead of importing them from the respective upstream packages.
 
 Resampling objects in \pkg{mlr3spatiotmpcv} inherit from class \texttt{mlr3::Resampling} and can be created from established object classes for geospatial data in R, including simple features [@pebesma2018], which facilitates their integration into domain-specific workflows in the geospatial sciences.
-Support for projected (planar) and unprojected (geographic) coordinate reference systems (CRS) currently varies depending on the partitioning techniques used, since these inherit their behavior from the underlying upstream R package.
+Support for projected (planar) and unprojected (geographic) coordinate reference systems (CRS) currently varies depending on the partitioning techniques used, since these inherit their behavior from the underlying upstream packages.
 
-Partitioning objects in \pkg{mlr3spatiotempcv} are equipped with generic `plot()` and `autoplot()` methods that visualize the groups created.
-`autoplot()` is \pkg{ggplot2}-based and uses [ggplot2](https://ggplot2.tidyverse.org) [@ggplot2] in two-dimensional geographic space and [plotly](https://github.com/ropensci/plotly) [@plotly] in three dimensions, i.e., geographic space plus time.
+Partitioning objects in \pkg{mlr3spatiotempcv} are equipped with generic `plot()` and `autoplot()` methods to visualize the created partitions.
+`autoplot()` is \pkg{ggplot2}-based and uses [ggplot2](https://ggplot2.tidyverse.org) [@ggplot2] in two-dimensional geographic space and [plotly](https://github.com/ropensci/plotly) [@plotly] in the three dimensions case, i.e., geographic space plus time.
 
-While \pkg{mlr3spatiotempcv} solely focuses on spatiotemporal resampling methods and their visualization, other packages such as \pkg{mlr3spatial} or \pkg{mlr3temporal} are planned in the mlr3 ecosystem to provide dedicated spatiotemporal learner and prediction methods.
+While \pkg{mlr3spatiotempcv} solely focuses on spatiotemporal resampling methods and their visualization, other packages such as \pkg{mlr3spatial} or \pkg{mlr3temporal} are planned in the \pkg{mlr3} ecosystem to provide dedicated spatiotemporal learner and prediction methods.
 
@@
 ```{r rsmp-schema, echo=FALSE, fig.cap="Conceptual overview of various spatial partitioning schemas.
-Starting from unpartioned spatial observations (top left) either a spatial block partitioning or a 'spatial leave-one-out resampling' is applied in the first step.
+Starting from unpartioned spatial observations (top left) either a 'spatial block partitioning' or a 'spatial leave-one-out resampling' is applied in the first step.
 A spatial block partitioning can further be turned into a 'leave-one-block-out resampling' or a 'k-fold CV resampling at the block level'.
@@
 
-\pkg{mlr3spatiotempcv} currently implements the partitioning methods identified in Table \@ref(tab:sptcv-methods) with their \pkg{mlr3} class name.
+\pkg{mlr3spatiotempcv} currently implements the partitioning methods identified in Table \@ref(tab:sptcv-methods).
 Several of the implemented algorithms are themselves versatile toolboxes with multiple options.
@@
     "\# pub." indicates the number of studies that were found to have used this resampling technique until May 2021.
-    For each method, up to three randomly chosen references were added to the table as footnotes.}
+    For each method, up to three randomly selected references were added to the table as footnotes.}
   \begingroup
@@
 
-In the presence/background (or presence-only) situation, in contrast, only presence observations are recorded, and all other locations within the study area are referred to as the background and considered as pseudo-absences.
+In the presence/background (or presence-only) situation, in contrast, only presence observations are recorded, and all other locations within the study area are referred to as background and considered as pseudo-absences.
 Presence/background modeling can be enabled with the argument `spDataType = "PB"`.
@@
 With `addBG = TRUE` (the default), all background points with a distance of `theRange` around a test (presence) point are included in the test fold as absence data; note that in this case, there is no separation buffer between training and test samples.
-The `addBG = FALSE` setting, in contrast, no background data is added to the test fold, which then contains only one (presence) observation, and only the data at a distance of `theRange` or greater are included in the training sample, including background data from these areas.
+The `addBG = FALSE` setting, in contrast, for which no background data is added to the test fold, then contains only one (presence) observation, and only the data at a distance of `theRange` or greater are included in the training sample, including background data from these areas.
 
@@
 
-Note that the resampled discs will potentially overlap.
+It should be noted that the resampled discs will potentially overlap.
 Strictly speaking, this straightforward extension of spatial LOO does therefore not establish a disjoint partitioning as used for CV resampling in the traditional sense.
@@
 It uses rectangular blocks that can be rotated (argument `rotation`), and a minimum number or fraction of observations per block can optionally be achieved by iteratively merging small blocks into adjacent blocks  (argument `reassign` in conjunction with `min_n` or `min_frac`).
-Block size or number is specified via the argument `dsplit` or `nsplit`, respectively, and square blocks can be obtained with a single (or two identical) `dsplit` value.
+Block size or number is specified via the argument `dsplit` or `nsplit`, respectively, and square blocks can be obtained with a single (or two identical) `dsplit` value(s).
 
@@
 In this so-called "Custom CV", users supply a factor variable, each level of which defines a partition.
-The factor variable can either be specified through a factor vector of the same length as number of observations, or by passing the name of a feature of type `character` within the task (argument `col`).
-The following simple example from `sperrorest::partition_factor()` creates altitudinal zones that define the spatial partitions.
+The factor variable can either be specified through a factor vector of the same length as number of observations, or by passing the name of a feature within the task (argument `col`).
+The following simple example (taken from `sperrorest::partition_factor()`) creates altitudinal zones that define the spatial partitions.
 
@@
 In $k$-fold CV resampling at the block level, there are therefore $k$ partitions, each consisting of $1/k$-th of the blocks.
-The special case in which $k$ equals the number of blocks, CV at the block level simply becomes leave-one-block-out CV, for which dedicated implementations exist (Section  \@ref(sec:lobo-cv)).
+The special case in which $k$ equals the number of blocks, CV at the block level simply becomes leave-one-block-out CV, for which dedicated implementations exist (see section \@ref(sec:lobo-cv)).
 
@@
 
-The `"spcv_block"` method from package \pkg{blockCV} supports both random and systematic resampling of square blocks (`rsmp("spcv_block")` with argument `selection = "random"` and `"systematic"`, respectively; Figures \@ref(fig:block-random) and \@ref(fig:block-systematic)).
-There are additional options for modeling presence-only data, which is a typical situation in species distribution modeling.
-Users can also supply a user-defined polygon via an argument `rasterLayer` with predefined blocking zones.
+The `"spcv_block"` method from package \pkg{blockCV} supports both random and systematic resampling of square blocks with argument `selection = "random"` and `"systematic"`, respectively; (see figures \@ref(fig:block-random) and \@ref(fig:block-systematic)).
+There are additional options for modeling presence-only data, which is a typical use case in species distribution modeling.
+Users can furthermore supply a user-defined polygon via argument `rasterLayer` with predefined blocking zones.
 
-Block sizes (square) (in meters) are determined by the `range` argument.
+The size of the square blocks (in meters) are determined by the `range` argument.
 Rectangular blocks can be created by specifying the number of desired rows and columns (arguments `rows` and `cols`).
@@
 
-It should be noted that rectangular partitioning can be problematic in irregularly shaped study areas as in Figure \@ref(fig:block-random) where some of the resulting partitions may contain substantially fewer observations than others.
+It should be noted that rectangular partitioning can be problematic in irregularly shaped study areas as shown in Figure \@ref(fig:block-random) where some of the resulting partitions may contain substantially fewer observations than others.
 
@@
 ```{r block-random, fig.cap='Random resampling of square spatial blocks using the implementation in package blockCV (method "spcv\\_block" with option selection = "random" in mlr3spatiotempcv).
-The size of the squares is 1000 m, and four out of the 19 blocks were assigned to this test sample.', out.width="40%", fig.pos="ht", echo=FALSE}
+The size of the squares is 1000 m, and four out of the 19 blocks were assigned to the test partition.', out.width="40%", fig.pos="ht", echo=FALSE}
 rsmp_block_random = rsmp("spcv_block", range = 1000, folds = 5)
@@
 
-In systematic resampling, the blocks are numbered row by row, and blocks $i+j\cdot\texttt{folds}$ are assigned to fold $i$.
+In systematic resampling, the blocks are numbered row by row, and blocks $i+j\cdot\texttt{folds}$ are assigned to fold $i$ (see Figure \@ref(fig:block-systematic)).
 This may create undesired patterns when the number of columns is equal to or a multiple of the number of folds.
@@
 
-*Checkerboard* partitioning is a special case of a systematic block partitioning (`selection = "checkerboard"`).
+*Checkerboard* partitioning is a special case of a systematic block partitioning (`selection = "checkerboard"`) which is why we omitted a practical example for this option.
 It inherently supports only two folds, making it less appealing than the more commonly used five- or ten-fold resampling, which achieve larger training set sizes.
@@
 This is achieved by specifying the "group" column role in a mlr3 Task object, which uses the factor levels as blocks.
-A completely group or block of observations is therefore assigned to a specific partition, which consequently honor the grouping structure.
-In \pkg{mlr} this concept was referred to as "blocking".
+A complete group or block of observations is therefore assigned to a specific partition, which consequently honors the grouping structure.
+In the deprecated \pkg{mlr} package this concept was referred to as "blocking".
 
@@
 The following example uses $k$-means clustering to generate classes that are used as blocks.
-To underline the honoring of the groups, we chose a number of groups (eight) that is not a multiple of the number of folds (three).
-The test sets in the first and second folds are therefore composed of three groups while the third one holds two groups.
+To underline the honoring of the groups, a number of groups (eight) that is not a multiple of the number of folds (three) was chosen.
+The test sets in the first and second folds are therefore composed out of three groups while the third one holds two groups.
 
@@
 ```{r cv-non-eval-2, fig.cap='Cross-Validation at the block level including predefined groups from package mlr3 (method "cv").
-A factor variable is used to define the grouping.
+A factor variable (named "group") was used to define the grouping.
 Each class is either assigned to the test or training set.', out.width="40%", fig.pos="ht", eval=FALSE}
@@
 
-In the following example, the observations are clustered based on the "distance to forest" feature, which results in a distance-based zonification:
+In the following example, the observations are clustered based on the feature "distance to forest" (left sub-figure of Figure \@ref(fig:env-1-eval)), which results in a distance-based zonification.
+This method also allows to use multiple features for clustering.
+The right sub-figure of Figure \@ref(fig:env-1-eval) shows the outcome when using "distance to deforestation" and "slope angle".
 
@@
 ```
-
-It is also possible to cluster based on multiple features, in this example the distance to deforestation and the slope angle:
 
@@
 This method is referred to as "leave-location-and-time-out" (LLTO) in package \pkg{CAST}.
-Additional methods that resample in the temporal and spatial domain only are named "leave-time-out" (LTO) and "leave-location-out" (LLO), respectively, in \pkg{CAST}.
+Additional methods that resample in the temporal and spatial domain only are named "leave-time-out" (LTO) and "leave-location-out" (LLO), respectively.
 Note that despite their names, LLTO, LTO and LLO are conceptually not leave-*one*-out methods as they place a certain fraction of observations in the test set, as in ordinary CV.
@@
 
-In this section the `cookfarm` dataset is used as an example because has a temporal dimension identified by the variable "Date".
+In this section the `cookfarm` dataset is used as an example because it comes with a temporal dimension via variable "Date".
 
-Two visualization types are supported by `mlr3spatiotempcv::autoplot()` for this method: a two-dimensional and a three-dimensional one which can be selected via the logical argument `plot3D`.
-The 3D visualization (i.e. 2D + time) is done via package \pkg{plotly}.
-Because a dynamic image cannot be included in this manuscript, a static version, which can be generated by setting `static_image = TRUE`, is shown.
+`mlr3spatiotempcv::autoplot()` supports two visualization types for spatiotemporal methods which can be selected via the logical argument `plot3D`.
+The heavy lifiting of the 3D visualization (i.e. 2D + time) option is done via package \pkg{plotly}.
+Because a dynamic image cannot be included in this manuscript, static versions, which can be generated by setting `static_image = TRUE`, are shown (see for example Figure \@ref(fig:lto)).
 
@@
 
-In the `cookfarm` example, we reduce the `Date` variable to five unique levels for better visualization, and then create a spatiotemporal regression task in \pkg{mlr3spatiotempcv} (Figure \@ref(fig:lto)).
+In the `cookfarm` example dataset, the `Date` variable was reduced to five unique levels for better visualization, and then used to create a spatiotemporal regression task in \pkg{mlr3spatiotempcv} (Figure \@ref(fig:lto)).
 
@@
 The sampled locations form the test partition while the temporal information is ignored (Figure \@ref(fig:llo)).
-Unlike spatial CV methods that are based on geometric regions or the clustering of coordinates, the sampled test locations bear no particular spatial relationship.
+Unlike spatial CV methods that are based on geometric regions or the clustering of coordinates, the sampled test locations include no particular spatial relationship.
 
@@
 
-```{r llo, echo=FALSE, out.width="70%", fig.cap='Perspective plot of "leave-location-out" CV from package CAST (method "sptcv\\_cstf" and argument space\\_var = "SOURCEID").
-The data is viewed from the future towards the past.', fig.pos="ht"}
+```{r llo, echo=FALSE, out.width="70%", fig.cap='Birds-eye view of "leave-location-out" CV from package CAST (method "sptcv\\_cstf" and argument space\\_var = "SOURCEID").', fig.pos="ht"}
 knitr::include_graphics("png/llo.png")
@@
 Conceptually, LLTO applies zero-distance buffering in both space and time: The buffer zones consist of all observations whose distance to the test sample in either space or time equals zero.
-In the mathematical sense, however, this buffering is not based on a valid metric (or distance function) in three-dimensional space (2D + time) as neither the identity of detectability nor the triangle inequality are satisfied by the underlying combined 'distance' measure.
+In a mathematical sense, however, this buffering is not based on a valid metric (or distance function) in three-dimensional space (2D + time) as neither the identity of detectability nor the triangle inequality are satisfied by the underlying combined 'distance' measure.
 Also note that LLTO does not 'combine' LTO with LLO, as neither of these applies a buffer zone.
@@
 
-```{r llto, echo=FALSE, out.width="70%", fig.cap='Perspective plot of leave-location-and-time-out CV from package CAST (method "sptcv\\_cstf" and arguments time\\_var = "Date" and space\\_var = "SOURCEID").
+```{r llto, echo=FALSE, out.width="70%", fig.cap='Perspective plot of "leave-location-and-time-out" CV from package CAST (method "sptcv\\_cstf" and arguments time\\_var = "Date" and space\\_var = "SOURCEID").
 The grey points are excluded from both the training and the test set in this fold.', fig.pos="h"}
@@
 
-Considering the wealth and continuing development of spatial and spatiotemporal partitioning techniques for specific applications, we will refrain from providing a complete account of these algorithms.
 At present, \pkg{mlr3spatiotempcv} also supports spatiotemporal partitioning using the versatile CLUTO clustering algorithm [@cluto].
 CLUTO is available in R through the \pkg{skmeans} package, which provides an interface to a downloadable compiled library with a restriction to non-commercial uses (see `help("ResamplingSptCVCluto", package = "mlr3spatiotempcv")` for more information).
-With respect to functionality, \pkg{mlr3spatiotempcv} supports two-dimensional clustering based on coordinates, time or features and three-dimensional clustering on coordinates and time.
+Due to this restriction and the age of the latest release (14 years at the time of writing) this method is not explained in greater detail.
 
@@
 
-We use a well-known case study to demonstrate the application of spatial and non-spatial resampling techniques for model assessment in \pkg{mlr3spatiotempcv}.
+A well-known case study is used to demonstrate the application of spatial and non-spatial resampling techniques for model assessment in \pkg{mlr3spatiotempcv}.
 The objective of landslide susceptibility modeling is to predict how prone to landslide initiation a location is.
 Models are fitted to historical landslide occurrences, but they need to learn generalizable relationships between predisposing variables and the response as opposed to perfectly reproducing or memorizing the historical distribution.
-This binary classification task on landslides in Ecuador [@muenchow2012] is available as a built-in task via `tsk("ecuador")`, but here we generate it from the learning sample.
+This binary classification task on landslides in Ecuador [@muenchow2012] is available as a built-in task via `tsk("ecuador")`, but is generated from the learning sample in this example.
 Random forest is used as a classifier, and the area under the ROC curve (AUROC) as the performance measure.
@@
 This approach is better suited for the irregular shape of the present study area than a rectangular partitioning.
-Figures \@ref(fig:vis-spcv-eval)) and \@ref(fig:vis-nspcv-eval)) show the contrasting distributions of training and test samples.
-For demonstration purposes we use only four CV folds and two repetitions.
+Figures \@ref(fig:vis-spcv-eval) and \@ref(fig:vis-nspcv-eval) show the contrasting distributions of training and test samples.
+For demonstration purposes only four CV folds and two repetitions are used.
 
@@
 
-In \pkg{mlr3}, machine-learning tasks with their respective dataset and response variable are represented by objects of the class `Task`.
-\pkg{mlr3spatiotempcv}'s spatial and spatiotemporal machine-learning tasks are derived from this superclass.
+In \pkg{mlr3}, machine-learning tasks with their respective dataset and response variable are represented by objects of class `Task`.
+\pkg{mlr3spatiotempcv}'s spatial and spatiotemporal machine-learning tasks are also derived from this superclass.
 Specifically, the `TaskClassifST` and `TaskRegrST` classes for classification and regression tasks require several additional arguments that must be passed as a named list using the `extra_args` argument:
@@
 
-The task `"ecuador"` is available in \pkg{mlr3spatiotempcv} through `tsk("ecuador")`.
-To set it up manually, we proceed as follows:
+The task `"ecuador"` is available as an example task in \pkg{mlr3spatiotempcv} through `tsk("ecuador")`.
+To create it manually from a `data.frame` named `ecuador`, one would do:
 
@@
 
-A resampling is generated in \pkg{mlr3} with the `rsmp()` function, which is used to generate a resampling object for four-fold non-spatial CV with two repetitions.
-The object is constructed and instantiated with its `$instantiate()` method.
+To define a resampling strategy, the `rsmp()` function is used to generate a resampling object using four folds and two repetitions following a random sampling logic ("cv").
 
@@
 
-The model assessment is now repeated using spatial CV resampling, for which we only need to replace `"repeated_cv"` with `"repeated_spcv_coords"`.
+The model assessment is now repeated again using spatial CV resampling, for which the only required change is to replace `"repeated_cv"` by `"repeated_spcv_coords"`.
 
@@
 
-Finally, we visualize (two of) the partitions that were used during performance estimation making use of the generic `autoplot()` function in package \pkg{mlr3spatiotempcv} (Figure \@ref(fig:vis-spcv-eval)).
+Finally, we visualize (two of) the partitions that were used during performance estimation by making use of the generic `autoplot()` function in package \pkg{mlr3spatiotempcv} (Figure \@ref(fig:vis-spcv-eval)).
 
@@
 
-The non-spatial CV estimate of AUC (0.76) is substantially higher compared to the spatial CV estimate of 0.64.
-Since test points in non-spatial CV may be from the same slopes or even the same landslides as the training data, the non-spatial CV result should be considered as an over-optimistic estimate of the model's ability to predict the susceptibility to "new" landslides.
+If one takes a closer look at the results, the non-spatial CV estimate of AUC (0.76) is substantially higher compared to the spatial CV estimate of 0.64.
+Since test points in non-spatial CV may be from the same slopes or even the same landslides as the training data, the non-spatial CV result should/can be considered as an over-optimistic estimate of the model's ability to predict the susceptibility to "new" landslides.
 Spatial CV, in contrast, provides a better/more accurate measure of a model's ability to generalize from the training sample --- in this case study, from the specific hillslopes and historical landslides in the training sample.
@@
 The question which resampling method should be chosen for a prediction task and dataset at hand comes up regularly in practice.
-Even though there is and will be no definitive answer to this, we would like to give some guidance in this section to help finding an appropriate method.
+Even though there is and most likely will be no definitive answer to this question, we would like to give some guidance in this section to help finding an appropriate method.
 As a general rule, we recommend to use a resampling scheme that (1) mimics the predictive situation in which the model will be applied operationally, and (2) is consistent with the structure of the data.
@@
 
-We recommend using a nested CV for this purpose.
+We recommend using nested CV for this purpose.
 In nested CV, an "inner" CV is performed on each CV training set since hyperparameter tuning is an integral part of model fitting that should not be able to use information from the CV test set.
@@
 The \pkg{mlr3spatiotempcv} package is the first package to bundle and categorize spatiotemporal resampling methods implemented in multiple other packages in R.
-The available resampling techniques allow the users to vary the scale or granularity of the resampled spatiotemporal units as well as their shape and possible buffer distance between training and test samples.
+The available resampling techniques allow users to vary the scale or granularity of the resampled spatiotemporal units as well as their shape and possible buffer distance between training and test samples.
 These settings may account for the specific characteristics of spatiotemporal prediction tasks, but modelers now have to make the important decision of choosing a method that is adequate for their situation.

### 2021-10-15 paper/paper.Rmd
move Alex to last place

@@
     email: \email{patrick.schratz@uni-jena.de}
-  - name: Alexander Brenning
-    affiliation: 'Friedrich Schiller University Jena \AND'
-    address: |
-      | Department of Geography
-      | Geographic Information Science group
-    email: \email{alexander.brenning@uni-jena.de}
   - name: Marc Becker
-    affiliation: 'Ludwig-Maximilians-Universität München'
+    affiliation: 'Ludwig-Maximilians-Universität München \AND'
     address: |

### 2021-10-15 paper/paper.Rmd
spell check

@@
 
-Approaches based on *spatial blocking* (or somtimes calld *grouping*) require either the construction of spatial zones, or the use of pre-existing spatial structures in the data.
+Approaches based on *spatial blocking* (or sometimes called *grouping*) require either the construction of spatial zones, or the use of pre-existing spatial structures in the data.
 Let's refer to these spatial units or blocks as $\mathcal{Z}_i$, $1\le i\le n_z$.
@@
 `mlr3spatiotempcv::autoplot()` supports two visualization types for spatiotemporal methods which can be selected via the logical argument `plot3D`.
-The heavy lifiting of the 3D visualization (i.e. 2D + time) option is done via package \pkg{plotly}.
+The heavy lifting of the 3D visualization (i.e. 2D + time) option is done via package \pkg{plotly}.
 Because a dynamic image cannot be included in this manuscript, static versions, which can be generated by setting `static_image = TRUE`, are shown (see for example Figure \@ref(fig:lto)).

### 2021-10-18 paper/paper.Rmd
minor improvements in word choice

@@
 Partitioning objects in \pkg{mlr3spatiotempcv} are equipped with generic `plot()` and `autoplot()` methods to visualize the created partitions.
-`autoplot()` is \pkg{ggplot2}-based and uses [ggplot2](https://ggplot2.tidyverse.org) [@ggplot2] in two-dimensional geographic space and [plotly](https://github.com/ropensci/plotly) [@plotly] in the three dimensions case, i.e., geographic space plus time.
+`autoplot()` is \pkg{ggplot2}-based and uses [ggplot2](https://ggplot2.tidyverse.org) [@ggplot2] in two-dimensional geographic space and [plotly](https://github.com/ropensci/plotly) [@plotly] in the three dimensional case, i.e., geographic space plus time.
 
@@
 To underline the honoring of the groups, a number of groups (eight) that is not a multiple of the number of folds (three) was chosen.
-The test sets in the first and second folds are therefore composed out of three groups while the third one holds two groups.
+The test sets in the first and second folds are therefore composed of three groups while the third one holds two groups.
 
@@
 
-In this section the `cookfarm` dataset is used as an example because it comes with a temporal dimension via variable "Date".
+In this section the `cookfarm` dataset is used as an example because it has a temporal dimension identified by the variable "Date".
 

### 2021-10-19 paper/paper.Rmd
adjust title, citations (added URLs and DOIs) and add emails

@@
       | Statistical Learning and Data Science group
+    email: marc.becker@stat.uni-muenchen.de
     # To add another line, use \AND at the end of the previous one as above
@@
       | Faculty of Statistics
+    email: lang@statistik.tu-dortmund.de
   - name: Alexander Brenning
@@
 title:
-  formatted: "\\pkg{mlr3spatiotempcv}: Spatiotemporal resampling methods for machine learning in R with \\pkg{mlr3}"
+  formatted: "\\pkg{mlr3spatiotempcv}: Spatiotemporal resampling methods for machine learning in R"
   # If you use tex in the formatted title, also supply version without
-  plain:     "mlr3spatiotempcv: spatiotemporal resampling methods for machine learning in R with mlr3"
+  plain:     "mlr3spatiotempcv: spatiotemporal resampling methods for machine learning in R"
   # For running headers, if needed
@@
   Spatial and spatiotemporal machine-learning models require a suitable framework for their model assessment, model selection, and hyperparameter tuning, in order to avoid error estimation bias and over-fitting.
-  This contribution reviews the state-of-the-art in spatial and spatiotemporal CV, and introduces the R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) as an extension package of the machine-learning framework [mlr3](https://mlr3.mlr-org.com).
-  Currently various R packages implementing different spatiotemporal partitioning strategies exist: blockCV, CAST, skmeans and sperrorest.
-  The goal of [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is to gather the available spatiotemporal resampling methods in R and make them available to users through a simple and common interface.
-  This is made possible by integrating the package directly into the [mlr3](https://mlr3.mlr-org.com) machine-learning framework, which already has support for generic non-spatiotemporal resampling methods such as random partitioning.
+  This contribution reviews the state-of-the-art in spatial and spatiotemporal CV, and introduces the R package mlr3spatiotempcv as an extension package of the machine-learning framework **mlr3**.
+  Currently various R packages implementing different spatiotemporal partitioning strategies exist: **blockCV**, **CAST**, **skmeans** and **sperrorest**.
+  The goal of **mlr3spatiotempcv** is to gather the available spatiotemporal resampling methods in R and make them available to users through a simple and common interface.
+  This is made possible by integrating the package directly into the **mlr3** machine-learning framework, which already has support for generic non-spatiotemporal resampling methods such as random partitioning.
   One advantage is the use of a consistent nomenclature in an overarching machine-learning toolkit instead of a varying package-specific syntax, making it easier for users to choose from a variety of spatiotemporal resampling methods.
@@
     base_format: rticles::jss_article
+    citation_package: "natbib"
     includes:
@@
 In contrast, packages \pkg{blockCV} and \pkg{ENMeval} were developed for block and buffer resampling with a focus on species distribution modeling [@blockCV; @rest2014; @muscarella2014].
-However, neither of these have been integrated into established machine-learning frameworks such as \pkg{mlr}/\pkg{mlr3} or \pkg{caret}/\pkg{tidymodels}, and all of them lack support for temporal prediction tasks.
+However, neither of these have been integrated into established machine-learning frameworks such as \pkg{mlr}/\pkg{mlr3} [@mlr3] or \pkg{caret}/\pkg{tidymodels} [@kuhn2020], and all of them lack support for temporal prediction tasks.
 The \pkg{CAST} package, in contrast, focuses on spatiotemporal prediction tasks and makes use of some functions of the \pkg{caret} framework [@cast; @meyer2018].
@@
 
-Thus, \pkg{mlr3spatiotempcv} implements for the first time a comprehensive state-of-the-art compilation of spatial and spatiotemporal partitioning schemes that is well-integrated into a comprehensive machine-learning framework in R, the \pkg{mlr3} ecosystem [@mlr3].
+Thus, \pkg{mlr3spatiotempcv} implements for the first time a comprehensive state-of-the-art compilation of spatial and spatiotemporal partitioning schemes that is well-integrated into a comprehensive machine-learning framework in R, the \pkg{mlr3} ecosystem.
 This package is furthermore equipped with a variety of two- and three-dimensional visualization capabilities.
@@
 
-Leave-one-out CV with buffer and several adaptations for species distribution modeling [@hijmans2013] are implemented in the \pkg{blockCV} package as the so-called "buffering" method and integrated into \pkg{mlr3spatiotempcv} under the label `"spcv_buffer"`.
+Leave-one-out CV with buffer and several adaptations for species distribution modeling [@hijmans2020] are implemented in the \pkg{blockCV} package as the so-called "buffering" method and integrated into \pkg{mlr3spatiotempcv} under the label `"spcv_buffer"`.
 In species distribution modeling, the response variable can either be recorded as presence/absence data or as presence/background information; both options are supported by this implementation. 

### 2021-10-20 paper/paper.Rmd
Add Michels changes

@@
   Spatial and spatiotemporal machine-learning models require a suitable framework for their model assessment, model selection, and hyperparameter tuning, in order to avoid error estimation bias and over-fitting.
-  This contribution reviews the state-of-the-art in spatial and spatiotemporal CV, and introduces the R package mlr3spatiotempcv as an extension package of the machine-learning framework **mlr3**.
-  Currently various R packages implementing different spatiotemporal partitioning strategies exist: **blockCV**, **CAST**, **skmeans** and **sperrorest**.
-  The goal of **mlr3spatiotempcv** is to gather the available spatiotemporal resampling methods in R and make them available to users through a simple and common interface.
-  This is made possible by integrating the package directly into the **mlr3** machine-learning framework, which already has support for generic non-spatiotemporal resampling methods such as random partitioning.
+  This contribution reviews the state-of-the-art in spatial and spatiotemporal CV, and introduces the \proglang{R} package mlr3spatiotempcv as an extension package of the machine-learning framework **mlr3**.
+  Currently various \proglang{R} packages implementing different spatiotemporal partitioning strategies exist: \pkg{blockCV}, \pkg{CAST}, \pkg{kmeans} and \pkg{sperrorest}.
+  The goal of \pkg{mlr3spatiotempcv} is to gather the available spatiotemporal resampling methods in \proglang{R} and make them available to users through a simple and common interface.
+  This is made possible by integrating the package directly into the \pkg{mlr3} machine-learning framework, which already has support for generic non-spatiotemporal resampling methods such as random partitioning.
   One advantage is the use of a consistent nomenclature in an overarching machine-learning toolkit instead of a varying package-specific syntax, making it easier for users to choose from a variety of spatiotemporal resampling methods.
@@
 This CV bias is well-known in spatial as well as non-spatial prediction [@brenning2005; @brenning2008; @arlot2010; @roberts2017] and in forecasting [@bergmeir2018].
-It is most easily understood from a predictive modeling perspective by focusing on the question where (and when) the model should be used for prediction.
+It is most easily understood from a predictive modeling perspective by focusing on the question of where (and when) the model should be used for prediction.
 In crop classification from remotely-sensed data, for instance, learning samples routinely contain multiple grid cells from a sample of fields with known crop type, for instance 2000 grid cells from 100 fields scattered across a large study region.
@@
 This toolbox, implemented as an open-source R package, builds upon and generalizes several existing toolboxes that have been developed in recent years for more specific settings (Table \@ref(tab:sptcv-methods)).
-The earliest and most comprehensive of these implementations is the \pkg{sperrorest} R package [@brenning2012], which provides an extensible framework and includes predefined resampling strategies based on geometric blocking, clustering, and buffering.
+The earliest and most comprehensive of these implementations is the \pkg{sperrorest} \proglang{R} package [@brenning2012], which provides an extensible framework and includes predefined resampling strategies based on geometric blocking, clustering, and buffering.
 In contrast, packages \pkg{blockCV} and \pkg{ENMeval} were developed for block and buffer resampling with a focus on species distribution modeling [@blockCV; @rest2014; @muscarella2014].
@@
 
-The purpose of this article is to give an overview of the methods implemented in the R package \pkg{mlr3spatiotempcv}.
+The purpose of this article is to give an overview of the methods implemented in the \proglang{R} package \pkg{mlr3spatiotempcv}.
 After presenting the conceptual background in the following section, the overall structure of the \pkg{mlr3spatiotempcv} package is outlined.
@@
 where $l$ is a real-valued loss function, and the expected value is with respect to the probability distribution of $X$, the features of an instance $(Y,X)$ drawn randomly from the underlying population.
-This is referred to as the *actual* or *conditional* performance measure as it is conditional on $\mathcal{L}$ [@hand1997].
+This is referred to as the *actual* or *conditional* performance measure, as it is conditional on $\mathcal{L}$ [@hand1997].
 The loss function can take a variety of forms such as the misclassification error $I(Y\neq\hat{f}_\mathcal{L}(X))$ in classification, or the squared error $(Y-\hat{f}_\mathcal{L}(X))^2$ in regression, among many other possible measures.
@@
 Spatial variants of CV have been proposed that apply an exclusion *buffer* or guard zone to the test locations to separate them from the training data [@brenning2005; @roberts2017].
-One approach that has been proposed for defining a separation distance is to use the range of autocorrelation of model residuals to determine the buffer distance as this seeks to establish independence conditional on the predictors [@brenning2005; @roberts2017].
+One approach that has been proposed for defining a separation distance is to use the range of autocorrelation of model residuals to determine the buffer distance, as this seeks to establish independence conditional on the predictors [@brenning2005; @roberts2017].
 
@@
 All of these are supported by advanced visualization tools, which are particularly important in a spatial and spatiotemporal setting.
+Additionally, \pkg{mlr3pipelines} [@mlr3pipelines] provides a plethora of preprocessing operators to conveniently build ML pipelines which can be resampled, tuned and benchmarked as regular learners.
 
@@
       \\
-      Type                            & Sub-type & Name                        & R Package     & \# pub. \\
+      Type                            & Sub-type & Name                        & \proglang{R} package     & \# pub. \\
       \toprule
@@
 
-The question which resampling method should be chosen for a prediction task and dataset at hand comes up regularly in practice.
-Even though there is and most likely will be no definitive answer to this question, we would like to give some guidance in this section to help finding an appropriate method.
+The question of which resampling method should be chosen for a prediction task and dataset at hand comes up regularly in practice.
+Even though there is and most likely will be no definitive answer to this question, we would like to give some guidance in this section to help find an appropriate method.
 As a general rule, we recommend to use a resampling scheme that (1) mimics the predictive situation in which the model will be applied operationally, and (2) is consistent with the structure of the data.
@@
 - Is it necessary to impose a separation distance or prediction horizon as a spatial or temporal buffer between training and prediction locations?
-- How densely are the observations distributed in space and time? Are they more densely distributed that the intended spatial or temporal prediction distance?
+- How densely are the observations distributed in space and time? Are they more densely distributed than the intended spatial or temporal prediction distance?
 - Is the data naturally grouped, e.g. because of the spatial extent of the studied objects, or as a consequence of multi-level (cluster) sampling?
@@
 We recommend using nested CV for this purpose.
-In nested CV, an "inner" CV is performed on each CV training set since hyperparameter tuning is an integral part of model fitting that should not be able to use information from the CV test set.
+In nested CV, an "inner" CV is performed on each CV training set, since hyperparameter tuning is an integral part of model fitting that should not be able to use information from the CV test set.
 In such scenarios it is recommended to use the same spatial resampling method for the inner CV (hyperparameter tuning) as for the outer CV (model assessment) in order to use the appropriate objective function for optimization.
@@
 While this effect also depends on the variable used for clustering, similar effects could potentially also apply to `"spcv_env"` and `"sptcv_cluto"` methods.
-However, such effects are more difficult to quantify because selected features of these methods are always different in contrast to `"spcv_coords"` which always uses coordinates for clustering.
+However, such effects are more difficult to quantify because selected features of these methods are always different, in contrast to `"spcv_coords"` which always uses coordinates for clustering.
 This issue is even more critical in CV at the block level with `"spcv_block"` with options `selection = "systematic"` and `selection = "checkerboard"` because identical folds are assigned in each repetition.
@@
 <!-- sign-off -->
-Spatiotemporal cross-validation as a paradigm is not yet fully established in scientific workflows although it has been discussed intensively for more than a decade now.
+Spatiotemporal cross-validation as a paradigm is not yet fully established in scientific workflows, although it has been discussed intensively for more than a decade now.
 We anticipate that making the existing methods easily accessible to users is an important step to foster the acceptance of spatiotemporal cross-validation in the community and to allow modelers to produce bias-reduced model assessments in environmental and ecological studies.

### 2021-10-21 paper/paper.Rmd
table caption at bottom
reduce label sizes of Figure 10

@@
   \footnotesize
-  \caption[t]{Spatiotemporal resampling methods available in the mlr3 ecosystem.
-    The "Name" column shows the mlr3 method name as found in the \texttt{mlr3::mlr\_resamplings} dictionary.
-    "\# pub." indicates the number of studies that were found to have used this resampling technique until May 2021.
-    For each method, up to three randomly selected references were added to the table as footnotes.}
   \begingroup
@@
   \endgroup\label{tab:sptcv-methods}
+  \caption[b]{Spatiotemporal resampling methods available in the mlr3 ecosystem.
+    The "Name" column shows the mlr3 method name as found in the \texttt{mlr3::mlr\_resamplings} dictionary.
+    "\# pub." indicates the number of studies that were found to have used this resampling technique until May 2021.
+    For each method, up to three randomly selected references were added to the table as footnotes.}
 \end{table}
@@
 
-plot_env_single = autoplot(rsmp_env, size = 0.5, fold_id = 1, task = task)
+plot_env_single = autoplot(rsmp_env, size = 0.5, fold_id = 1, task = task) +
 
@@
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02))
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
+  ggplot2::theme(plot.title = ggtext::element_textbox(size = 8),
+    axis.text = ggplot2::element_text(size = 8))
 

### 2021-10-23 paper/paper.Rmd
last touches before subm

@@
       | Statistical Learning and Data Science group
-    email: marc.becker@stat.uni-muenchen.de
+    email: \email{marc.becker@stat.uni-muenchen.de}
     # To add another line, use \AND at the end of the previous one as above
@@
       | Faculty of Statistics
-    email: lang@statistik.tu-dortmund.de
+    email: \email{lang@statistik.tu-dortmund.de}
   - name: Alexander Brenning
@@
 # for knitr spin later
-knitr::opts_chunk$set(fig.path = "./paper/figs/",
+knitr::opts_chunk$set(fig.path = "./figs/",
   cache = TRUE,

### 2021-10-23 paper/code-repr.R

@@ -1,5 +1,6 @@
-fs::dir_create("paper/figs")
+# fs::dir_create("paper/figs")
 knitr::purl(input = "paper/paper.Rmd", output = "paper/code.R", quiet = TRUE)
-knitr::spin("paper/code.R")
-fs::file_delete(c("code.md", "paper/code.R", "code.html"))
-fs::dir_delete("paper/figs")
+source("paper/code.R", chdir = TRUE)
+# knitr::spin("paper/code.R")
+# fs::file_move("code.md", "paper/code.md")
+# fs::file_delete(c("code.md", "paper/code.R", "code.html"))

### 2022-05-17 paper/paper.Rmd
require blockCV and sperrorest namespaces in respective code chunks

@@
 library("mlr3spatiotempcv")
+requireNamespace("blockCV")
 task = tsk("ecuador")
@@
 ```{r tile-non-eval, eval=FALSE}
+requireNamespace("sperrorest")
 rsmp_tiles = rsmp("spcv_tiles", nsplit = c(3L, 4L))

### 2022-05-21 paper/paper.Rmd
add orcid IDs and update R code print output

@@
   - name: Patrick Schratz
+    orcid: 0000-0003-0748-6624
     affiliation: 'Friedrich Schiller University Jena'
@@
   - name: Marc Becker
+    orcid: 0000-0002-8115-0400
     affiliation: 'Ludwig-Maximilians-Universität München \AND'
@@
   - name: Michel Lang
+    orcid: 0000-0001-9754-0393
     affiliation: TU Dortmund University
@@
   - name: Alexander Brenning
+    orcid: 0000-0001-6640-679
     affiliation: 'Friedrich Schiller University Jena'
@@
   This contribution reviews the state-of-the-art in spatial and spatiotemporal CV, and introduces the \proglang{R} package mlr3spatiotempcv as an extension package of the machine-learning framework **mlr3**.
-  Currently various \proglang{R} packages implementing different spatiotemporal partitioning strategies exist: \pkg{blockCV}, \pkg{CAST}, \pkg{kmeans} and \pkg{sperrorest}.
+  Currently various \proglang{R} packages implementing different spatiotemporal partitioning strategies exist: \pkg{blockCV}, \pkg{CAST}, \pkg{skmeans} and \pkg{sperrorest}.
   The goal of \pkg{mlr3spatiotempcv} is to gather the available spatiotemporal resampling methods in \proglang{R} and make them available to users through a simple and common interface.
@@
   \usepackage{tablefootnote}
+  \usepackage{hyperref}
   \newcommand{\specialcell}[2][l]{%
@@
     citation_package: "natbib"
-    includes:
-      in_header: header.tex
+    # includes:
+    #   in_header: header.tex
 ---
@@
 })
+options(prompt = "R> ", continue = "+ ")
 ```

### 2022-05-21 paper/paper.Rmd
fix usage of code syntax in headers

@@
   \usepackage{tablefootnote}
-  \usepackage{hyperref}
   \newcommand{\specialcell}[2][l]{%
@@
     citation_package: "natbib"
-    # includes:
-    #   in_header: header.tex
+    includes:
+      in_header: header.tex
 ---

### 2022-05-23 paper/article.Rmd
- Update S3 methods of `as_task_regr.TaskRegrST` and `as_task_classif.TaskClassifST` (#178)
* paper: various style fixes

* update `as_task_classif` and `as_task_regr` S3 methods

--- 2022-05-21 paper/paper.Rmd
@@
   Spatial and spatiotemporal machine-learning models require a suitable framework for their model assessment, model selection, and hyperparameter tuning, in order to avoid error estimation bias and over-fitting.
-  This contribution reviews the state-of-the-art in spatial and spatiotemporal CV, and introduces the \proglang{R} package mlr3spatiotempcv as an extension package of the machine-learning framework **mlr3**.
+  This contribution reviews the state-of-the-art in spatial and spatiotemporal CV, and introduces the \proglang{R} package \pkg{mlr3spatiotempcv} as an extension package of the machine-learning framework \pkg{mlr3}.
   Currently various \proglang{R} packages implementing different spatiotemporal partitioning strategies exist: \pkg{blockCV}, \pkg{CAST}, \pkg{skmeans} and \pkg{sperrorest}.
@@
     \begin{tabular}[#1]{@{}l@{}}#2\end{tabular}}
-tables: true
+# tables: true
 bibliography: paper.bib
-output:
-  bookdown::pdf_book:
-    base_format: rticles::jss_article
-    citation_package: "natbib"
-    includes:
-      in_header: header.tex
+output: rticles::jss_article
+  # bookdown::pdf_book:
+  #   base_format: rticles::jss_article
+  #   citation_package: "natbib"
+    # includes:
+    #   in_header: header.tex
 ---
@@
 To mimic the predictive situation for which the model is trained, one would rather have to resample at the level of fields, not grid cells [@pena2015].
-If the model was to be applied to adjacent agricultural regions, i.e., outside the learning sample's spatial domain [*extra-domain* prediction, @brenning2005], it would even seem necessary to resample at a higher level of spatial aggregation, i.e.
-at the level of sub-regions within the learning sample, in order to realistically mimic the actual prediction task.
+If the model was to be applied to adjacent agricultural regions, i.e., outside the learning sample's spatial domain [*extra-domain* prediction, @brenning2005], it would even seem necessary to resample at a higher level of spatial aggregation, i.e., at the level of sub-regions within the learning sample, in order to realistically mimic the actual prediction task.
 The CV resampling needed therefore depends as much on the prediction task itself as on the data structure or dependency at hand.
@@
 
-This toolbox, implemented as an open-source R package, builds upon and generalizes several existing toolboxes that have been developed in recent years for more specific settings (Table \@ref(tab:sptcv-methods)).
+This toolbox, implemented as an open-source R package, builds upon and generalizes several existing toolboxes that have been developed in recent years for more specific settings (Table \ref{tab:sptcv-methods}).
 The earliest and most comprehensive of these implementations is the \pkg{sperrorest} \proglang{R} package [@brenning2012], which provides an extensible framework and includes predefined resampling strategies based on geometric blocking, clustering, and buffering.
@@
 
-Since re-using the learning sample $\mathcal{L}$ for testing, i.e.
-$\mathcal{T}:=\mathcal{L}$, would yield the over-optimistic *resubstitution* or *apparent* performance, CV partitions the sample $\mathcal{L}$ into disjoint training and test sets.
+Since re-using the learning sample $\mathcal{L}$ for testing, i.e., $\mathcal{T}:=\mathcal{L}$, would yield the over-optimistic *resubstitution* or *apparent* performance, CV partitions the sample $\mathcal{L}$ into disjoint training and test sets.
 Specifically, $\mathcal{L}$ is split into $k$ partitions,
@@
 \]
-where $f$ is a ML algorithm, i.e. a mapping that trains a model $\hat{f}_\mathcal{S}$ using any suitable training sample $\mathcal{S}$.
+where $f$ is a ML algorithm, i.e., a mapping that trains a model $\hat{f}_\mathcal{S}$ using any suitable training sample $\mathcal{S}$.
 The use of $k=5$ or $k=10$ folds is most commonly seen in practice, and these preferences are also supported by theory [@bengio2004; @cawley2010].
@@
 In traditional CV, the partitioning is based on uniform random resampling, which ignores spatial or temporal autocorrelation or any existing grouping structure as well as the structure of the prediction task, and may result in over-optimistic performance estimates.
-Several approaches have therefore been proposed in the literature and implemented in software to accommodate a variety of predictive situations (Table \@ref(tab:sptcv-methods)).
+Several approaches have therefore been proposed in the literature and implemented in software to accommodate a variety of predictive situations (Table \ref{tab:sptcv-methods}).
 
@@
 We would like to emphasize the conceptual distinction between *CV at the block level*, referring to this scenario, and *leave-one-block-out CV*, where the blocks themselves define the CV partitions.
-Figure \@ref(fig:rsmp-schema) gives an overview of the conceptual framework and terminology used in this work.
+Figure \ref{fig:rsmp-schema} gives an overview of the conceptual framework and terminology used in this work.
 
@@
 While the former two ignore the temporal and spatial dimension of the data, respectively, it has also been proposed to leave out random subsets of locations and time points [@meyer2018] or spatiotemporal clusters [@cluto].
-Details of these and other implementations are outlined in the respective subsections of Section  \@ref(sec:implementation).
+Details of these and other implementations are outlined in the respective subsections of Section \ref{sec:implementation}.
 
-# mlr3spatiotempcv within the mlr3 ecosystem
+# \pkg{mlr3spatiotempcv} within the \pkg{mlr3} ecosystem {short-title="mlr3spatiotempcv within the mlr3 ecosystem" #r-code}
 
@@
 
-Resampling objects in \pkg{mlr3spatiotmpcv} inherit from class \texttt{mlr3::Resampling} and can be created from established object classes for geospatial data in R, including simple features [@pebesma2018], which facilitates their integration into domain-specific workflows in the geospatial sciences.
+Resampling objects in \pkg{mlr3spatiotmpcv} inherit from class \texttt{mlr3::Resampling} and can be created from established object classes for geospatial data in \proglang{R}, including simple features [@pebesma2018], which facilitates their integration into domain-specific workflows in the geospatial sciences.
 Support for projected (planar) and unprojected (geographic) coordinate reference systems (CRS) currently varies depending on the partitioning techniques used, since these inherit their behavior from the underlying upstream packages.
@@
 Partitioning objects in \pkg{mlr3spatiotempcv} are equipped with generic `plot()` and `autoplot()` methods to visualize the created partitions.
-`autoplot()` is \pkg{ggplot2}-based and uses [ggplot2](https://ggplot2.tidyverse.org) [@ggplot2] in two-dimensional geographic space and [plotly](https://github.com/ropensci/plotly) [@plotly] in the three dimensional case, i.e., geographic space plus time.
+`autoplot()` is \pkg{ggplot2}-based and uses [ggplot2](https://ggplot2.tidyverse.org) [@ggplot2] in two-dimensional geographic space and [\pkg{plotly}](https://github.com/ropensci/plotly) [@plotly] in the three dimensional case, i.e., geographic space plus time.
 
@@
 
-At the most general level, resampling methods are categorized according to the level at which the data is partitioned and resampled (see Figure \@ref(fig:rsmp-schema)):
+At the most general level, resampling methods are categorized according to the level at which the data is partitioned and resampled (see Figure \ref{fig:rsmp-schema}):
 
 - **Spatial leave-one-out resampling**: Each individual observation forms a test set;
-- **Leave-one-block-out CV**: Individual blocks are left out as test data, i.e. the number of folds equals the number of blocks;
+- **Leave-one-block-out CV**: Individual blocks are left out as test data, i.e., the number of folds equals the number of blocks;
 - **CV at the block level**: Blocks are grouped into $k$ partitions, each of which is used as a test fold.
@@
 - Using an unsupervised clustering approach;
-- Using a custom input, i.e. specifying the blocks with an external grouping variable.
+- Using a custom input, i.e., specifying the blocks with an external grouping variable.
 
@@
 
-\pkg{mlr3spatiotempcv} currently implements the partitioning methods identified in Table \@ref(tab:sptcv-methods).
+\pkg{mlr3spatiotempcv} currently implements the partitioning methods identified in Table \ref{tab:sptcv-methods}.
 Several of the implemented algorithms are themselves versatile toolboxes with multiple options.
@@
 The following sections give an overview of most implemented partitioning strategies and their visualization options.
-The available methods are further discussed in section \@ref(sec:disc).
+The available methods are further discussed in section \ref{sec:disc}.
 
-\begin{table}[h]
+\begin{table}[ht]
   \centering
@@
     \end{tabular}
-  \endgroup\label{tab:sptcv-methods}
-  \caption[b]{Spatiotemporal resampling methods available in the mlr3 ecosystem.
-    The "Name" column shows the mlr3 method name as found in the \texttt{mlr3::mlr\_resamplings} dictionary.
+  \endgroup
+  \caption[b]{Spatiotemporal resampling methods available in the \pkg{mlr3} ecosystem.
+    The "Name" column shows the \pkg{mlr3} method name as found in the \texttt{mlr3::mlr\_resamplings} dictionary.
     "\# pub." indicates the number of studies that were found to have used this resampling technique until May 2021.
     For each method, up to three randomly selected references were added to the table as footnotes.}
+    \label{tab:sptcv-methods}
 \end{table}
@@
 
-### Spatial leave-one-out with buffer --- `"spcv_buffer"`
+### Spatial leave-one-out with buffer
 
@@
 In species distribution modeling, the response variable can either be recorded as presence/absence data or as presence/background information; both options are supported by this implementation.
-By default, the dataset contains confirmed presence and confirmed absence observations, i.e.
-locations where a species was observed and not observed, respectively, and therefore spatial LOO CV in its usual sense can be carried out.
-Figure \@ref(fig:buffer-eval) shows the first test fold generated with this method for presence/absence data with a buffer distance of 1000 m.
+By default, the dataset contains confirmed presence and confirmed absence observations, i.e., locations where a species was observed and not observed, respectively, and therefore spatial LOO CV in its usual sense can be carried out.
+Figure \ref{fig:buffer-eval} shows the first test fold generated with this method for presence/absence data with a buffer distance of 1000 m.
 
@@
 
-```{r buffer-eval, fig.cap='Visualization of the spatial buffering method from package blockCV (method "spcv\\_buffer" in  mlr3spatiotempcv).
+```{r buffer-eval, fig.cap='Visualization of the spatial buffering method from package \\pkg{blockCV} (method "spcv\\_buffer" in \\pkg{mlr3spatiotempcv}).
 The buffer distance is 1000 m.', out.width="40%", fig.pos="ht",echo=FALSE}
@@
 
-### Leave-one-disc-out with optional buffer --- `"spcv_disc"`
+### Leave-one-disc-out with optional buffer
 
@@
 
-```{r disc-eval, fig.cap='Visualization of one training set / test set combination generated with the leave-one-disc-out method from package sperrorest (method "spcv\\_disc" in mlr3spatiotempcv).
+```{r disc-eval, fig.cap='Visualization of one training set / test set combination generated with the leave-one-disc-out method from package \\pkg{sperrorest} (method `"spcv\\_disc"` in \\pkg{mlr3spatiotempcv}).
 The disc has a radius of 300 m and is surrounded by a 400-m buffer.', out.width="40%", echo=FALSE,fig.pos="ht"}
@@
 
-### Clustering-based: using coordinates --- `"spcv_coords"`
+### Clustering-based: using coordinates
 
@@
 Also, $k$-means clustering may be less suitable for data sets with pre-existing clusters of points and/or with isolated, distant sample locations.
-When distinct clusters of points are present, as in multi-level sampling, it may be better to define clusters using a factor variable (see method `"custom_cv"` in section \@ref(sec:custom-cv)).
+When distinct clusters of points are present, as in multi-level sampling, it may be better to define clusters using a factor variable (see method `"custom_cv"` in section \ref{sec:custom-cv}).
 
@@
 
-```{r coords-eval, fig.cap='Leave-one-block-out CV based on $k$-means clustering of the coordinates as implemented in package sperrorest (method "spcv\\_coords" in mlr3spatiotempcv).', out.width="40%", fig.pos="ht", echo=FALSE}
+```{r coords-eval, fig.cap='Leave-one-block-out CV based on $k$-means clustering of the coordinates as implemented in package \\pkg{sperrorest} (method `"spcv\\_coords"` in \\pkg{mlr3spatiotempcv}).', out.width="40%", fig.pos="ht", echo=FALSE}
 rsmp_coords = rsmp("spcv_coords", folds = 5)
@@
 
-### Geometric: using rectangular blocks --- `"spcv_tiles"`
+### Geometric: using rectangular blocks
 
@@
 Note that the actual number of folds obtained may be smaller than `nsplit[1]*nsplit[2]` (or smaller than what would be expected based on `dsplit`) since some blocks may be empty or (optionally) merged into adjacent folds.
-In the example, there are only eleven folds instead of twelve because the southwestern part of the study area's bounding box does not contain observations  (Figure \@ref(fig:tile)).
+In the example, there are only eleven folds instead of twelve because the southwestern part of the study area's bounding box does not contain observations (Figure \ref{fig:tile}).
 
@@
 
-```{r tile, fig.cap='Leave-one-block-out resampling from package sperrorest (method "spcv\\_tiles" in package mlr3spatiotempcv with argument nsplit = c(3,4) indicating the number of rows and columns).', out.width="40%", fig.pos="ht", echo=FALSE}
+```{r tile, fig.cap='Leave-one-block-out resampling from package \\pkg{sperrorest} (method `"spcv\\_tiles"` in package \\pkg{mlr3spatiotempcv} with argument `nsplit = c(3,4)` indicating the number of rows and columns).', out.width="40%", fig.pos="ht", echo=FALSE}
 rsmp_tiles = rsmp("spcv_tiles", nsplit = c(3L, 4L))
@@
 
-### Custom: `"custom_cv"` in mlr3 {#sec:custom-cv}
+<!-- FIXME: -->
+### Custom: "customcv" in \pkg{mlr3} {short-title="Custom: customcv in mlr3" #sec:custom-cv}
+<!-- {#sec:custom-cv} -->
+<!-- \subsubsection[Custom: \code{"customcv"} in mlr3]{Custom: \code{"custom\_cv"} in \pkg{mlr3}} -->
 
@@
 
-```{r custom-cv, fig.cap='Leave-one-level-out (custom) resampling from package mlr3 (method "custom\\_cv").
+```{r custom-cv, fig.cap='Leave-one-level-out (custom) resampling from package \\pkg{mlr3} (method `"custom\\_cv"`).
 A factor variable is used to define all partitions.', out.width="40%", fig.pos="ht", echo=FALSE}
@@
 In $k$-fold CV resampling at the block level, there are therefore $k$ partitions, each consisting of $1/k$-th of the blocks.
-The special case in which $k$ equals the number of blocks, CV at the block level simply becomes leave-one-block-out CV, for which dedicated implementations exist (see section \@ref(sec:lobo-cv)).
+The special case in which $k$ equals the number of blocks, CV at the block level simply becomes leave-one-block-out CV, for which dedicated implementations exist (see section \ref{sec:lobo-cv}).
 
-### Geometric: using rectangular blocks --- `"spcv_block"`
+### Geometric: using rectangular blocks
 
-The `"spcv_block"` method from package \pkg{blockCV} supports both random and systematic resampling of square blocks with argument `selection = "random"` and `"systematic"`, respectively; (see figures \@ref(fig:block-random) and \@ref(fig:block-systematic)).
+The `"spcv_block"` method from package \pkg{blockCV} supports both random and systematic resampling of square blocks with argument `selection = "random"` and `"systematic"`, respectively; (see figures \ref{fig:block-random} and \ref{fig:block-systematic}).
 There are additional options for modeling presence-only data, which is a typical use case in species distribution modeling.
@@
 
-It should be noted that rectangular partitioning can be problematic in irregularly shaped study areas as shown in Figure \@ref(fig:block-random) where some of the resulting partitions may contain substantially fewer observations than others.
+It should be noted that rectangular partitioning can be problematic in irregularly shaped study areas as shown in Figure \ref{fig:block-random} where some of the resulting partitions may contain substantially fewer observations than others.
 
@@
 
-```{r block-random, fig.cap='Random resampling of square spatial blocks using the implementation in package blockCV (method "spcv\\_block" with option selection = "random" in mlr3spatiotempcv).
+```{r block-random, fig.cap='Random resampling of square spatial blocks using the implementation in package \\pkg{blockCV} (method `"spcv\\_block"` with option `selection = "random"` in \\pkg{mlr3spatiotempcv}).
 The size of the squares is 1000 m, and four out of the 19 blocks were assigned to the test partition.', out.width="40%", fig.pos="ht", echo=FALSE}
@@
 
-In systematic resampling, the blocks are numbered row by row, and blocks $i+j\cdot\texttt{folds}$ are assigned to fold $i$ (see Figure \@ref(fig:block-systematic)).
+In systematic resampling, the blocks are numbered row by row, and blocks $i+j\cdot\texttt{folds}$ are assigned to fold $i$ (see Figure \ref{fig:block-systematic}).
 This may create undesired patterns when the number of columns is equal to or a multiple of the number of folds.
@@
 
-```{r block-systematic, fig.cap='Sytematic resampling of square spatial blocks using the implementation in package blockCV (method "spcv\\_block" with option selection = "systematic" in mlr3spatiotempcv).
+```{r block-systematic, fig.cap='Sytematic resampling of square spatial blocks using the implementation in package \\pkg{blockCV} (method `"spcv\\_block"` with option `selection = "systematic"` in \\pkg{mlr3spatiotempcv}).
 The size of the squares is 1000 m, and four out of the 19 blocks were assigned to this test sample.', out.width="40%", fig.pos="ht", echo=FALSE}
@@
 
-### Custom: `"cv"` with grouping in mlr3 {#sec:group-level}
+<!-- FIXME: underscore missing -->
+###  Custom: "cv" with grouping in \pkg{mlr3} {short-title="Custom cv with grouping in mlr3" #sec:custom-cv-grouping}
 
 Although the `"cv"` resampling strategy in \pkg{mlr3} performs random, non-spatial partitioning by default, it can also be used for CV at the block level.
-This is achieved by specifying the "group" column role in a mlr3 Task object, which uses the factor levels as blocks.
+This is achieved by specifying the "group" column role in a \pkg{mlr3} \texttt{Task} object, which uses the factor levels as blocks.
 A complete group or block of observations is therefore assigned to a specific partition, which consequently honors the grouping structure.
@@
 
-```{r cv-non-eval-2, fig.cap='Cross-Validation at the block level including predefined groups from package mlr3 (method "cv").
+```{r cv-non-eval-2, fig.cap='Cross-Validation at the block level including predefined groups from package \\pkg{mlr3} (method `"cv"`).
 A factor variable (named "group") was used to define the grouping.
@@
 
-```{r cv-eval-2, fig.cap='Cross-Validation at the block level including predefined groups from package mlr3 (method "cv").
+```{r cv-eval-2, fig.cap='Cross-Validation at the block level including predefined groups from package \\pkg{mlr3} (method `"cv"`).
 A factor variable is used to define the grouping.
@@
 
-### Clustering: using feature-based clustering --- `"spcv_env"`
+### Clustering: using feature-based clustering
 
@@
 
-In the following example, the observations are clustered based on the feature "distance to forest" (left sub-figure of Figure \@ref(fig:env-1-eval)), which results in a distance-based zonification.
+In the following example, the observations are clustered based on the feature "distance to forest" (left sub-figure of Figure \ref{fig:env-1-eval}), which results in a distance-based zonification.
 This method also allows to use multiple features for clustering.
-The right sub-figure of Figure \@ref(fig:env-1-eval) shows the outcome when using "distance to deforestation" and "slope angle".
+The right sub-figure of Figure \ref{fig:env-1-eval} shows the outcome when using "distance to deforestation" and "slope angle".
 
@@
 
-```{r env-1-eval, fig.cap='Environmental leave-one-block-out CV from package blockCV using one (left, "distdeforest") and two (right, "distdeforest" and "slope") predictors to define blocks in the feature space.
+```{r env-1-eval, fig.cap='Environmental leave-one-block-out CV from package \\pkg{blockCV} using one (left, `"distdeforest"`) and two (right, `"distdeforest"` and `"slope"`) predictors to define blocks in the feature space.
 Due to feature space clustering observations are not (necessarily) grouped in the spatial domain.', out.width="100%", fig.pos="ht", echo=FALSE}
@@
 
-Some of the implemented resampling methods operate in multiple dimensions, i.e. in space, time, or space--time.
+Some of the implemented resampling methods operate in multiple dimensions, i.e., in space, time, or space--time.
 In this section, only examples of these methods in the spatiotemporal domain will be shown.
@@
 
-### Custom: "Leave-location-and-time-out" and related methods --- `"spcv_cstf"`
+### Custom: "Leave-location-and-time-out" and related methods
 
@@
 `mlr3spatiotempcv::autoplot()` supports two visualization types for spatiotemporal methods which can be selected via the logical argument `plot3D`.
-The heavy lifting of the 3D visualization (i.e. 2D + time) option is done via package \pkg{plotly}.
-Because a dynamic image cannot be included in this manuscript, static versions, which can be generated by setting `static_image = TRUE`, are shown (see for example Figure \@ref(fig:lto)).
+The heavy lifting of the 3D visualization (i.e., 2D + time) option is done via package \pkg{plotly}.
+Because a dynamic image cannot be included in this manuscript, static versions, which can be generated by setting `static_image = TRUE`, are shown (see for example Figure \ref{fig:lto}).
 
@@
 
-In the `cookfarm` example dataset, the `Date` variable was reduced to five unique levels for better visualization, and then used to create a spatiotemporal regression task in \pkg{mlr3spatiotempcv} (Figure \@ref(fig:lto)).
+In the `cookfarm` example dataset, the `Date` variable was reduced to five unique levels for better visualization, and then used to create a spatiotemporal regression task in \pkg{mlr3spatiotempcv} (Figure \ref{fig:lto}).
 
@@
 
-```{r lto, echo=FALSE, out.width="70%", fig.cap='Perspective plot of "leave-time-out" CV from package CAST (method "sptcv\\_cstf" and argument time\\_var = "Date").
+```{r lto, echo=FALSE, out.width="70%", fig.cap='Perspective plot of `"leave-time-out"` CV from package \\pkg{CAST} (method `"sptcv\\_cstf"` and argument `time\\_var = "Date"`).
 Only five folds and five time points were used in this example.
@@
 In contrast to LTO, the LLO method randomly resamples locations that may, for example, correspond to time series.
-The sampled locations form the test partition while the temporal information is ignored (Figure \@ref(fig:llo)).
+The sampled locations form the test partition while the temporal information is ignored (Figure \ref{fig:llo}).
 Unlike spatial CV methods that are based on geometric regions or the clustering of coordinates, the sampled test locations include no particular spatial relationship.
@@
 
-```{r llo, echo=FALSE, out.width="70%", fig.cap='Birds-eye view of "leave-location-out" CV from package CAST (method "sptcv\\_cstf" and argument space\\_var = "SOURCEID").', fig.pos="ht"}
+```{r llo, echo=FALSE, out.width="70%", fig.cap='Birds-eye view of `"leave-location-out"` CV from package \\pkg{CAST} (method `"sptcv\\_cstf"` and argument `space\\_var = "SOURCEID"`).', fig.pos="ht"}
 knitr::include_graphics("png/llo.png")
@@
 
-In LLTO, a test set is first randomly sampled from the data set, and then all observations that correspond to the same location or time point are omitted from the training sample (Figure \@ref(fig:llto)).
+In LLTO, a test set is first randomly sampled from the data set, and then all observations that correspond to the same location or time point are omitted from the training sample (Figure \ref{fig:llto}).
 LLTO resampling mimics the situation where a model is trained on time series data from a number of locations and time points, and used to predict the time series at other locations and time points that are not included in the training sample.
@@
 
-```{r llto, echo=FALSE, out.width="70%", fig.cap='Perspective plot of "leave-location-and-time-out" CV from package CAST (method "sptcv\\_cstf" and arguments time\\_var = "Date" and space\\_var = "SOURCEID").
+```{r llto, echo=FALSE, out.width="70%", fig.cap='Perspective plot of `"leave-location-and-time-out"` CV from package \\pkg{CAST} (method `"sptcv\\_cstf"` and arguments `time\\_var = "Date"` and `space\\_var = "SOURCEID"`).
 The grey points are excluded from both the training and the test set in this fold.', fig.pos="h"}
@@
 
-### Clustering: using CLUTO --- `"sptcv_cluto"`
+### Clustering: using CLUTO
 
 At present, \pkg{mlr3spatiotempcv} also supports spatiotemporal partitioning using the versatile CLUTO clustering algorithm [@cluto].
-CLUTO is available in R through the \pkg{skmeans} package, which provides an interface to a downloadable compiled library with a restriction to non-commercial uses (see `help("ResamplingSptCVCluto", package = "mlr3spatiotempcv")` for more information).
+CLUTO is available in \proglang{R} through the \pkg{skmeans} package [@skmeans], which provides an interface to a downloadable compiled library with a restriction to non-commercial uses (see `help("ResamplingSptCVCluto", package = "mlr3spatiotempcv")` for more information).
 Due to this restriction and the age of the latest release (14 years at the time of writing) this method is not explained in greater detail.
@@
 This approach is better suited for the irregular shape of the present study area than a rectangular partitioning.
-Figures \@ref(fig:vis-spcv-eval) and \@ref(fig:vis-nspcv-eval) show the contrasting distributions of training and test samples.
+Figures \ref{fig:vis-spcv-eval} and \ref{fig:vis-nspcv-eval} show the contrasting distributions of training and test samples.
 For demonstration purposes only four CV folds and two repetitions are used.
@@
 
-Finally, we visualize (two of) the partitions that were used during performance estimation by making use of the generic `autoplot()` function in package \pkg{mlr3spatiotempcv} (Figure \@ref(fig:vis-spcv-eval)).
+Finally, we visualize (two of) the partitions that were used during performance estimation by making use of the generic `autoplot()` function in package \pkg{mlr3spatiotempcv} (Figure \ref{fig:vis-spcv-eval}).
 
@@
 
-Although the case study example in section \@ref(sec:case-study) used the `"spcv_coords"` method for coordinate-based clustering, this should not give the impression that this method is the only method suitable for this example task.
+Although the case study example in section \ref{sec:case-study} used the `"spcv_coords"` method for coordinate-based clustering, this should not give the impression that this method is the only method suitable for this example task.
 In this application setting, we want to assess how well the model generalized from the concrete set of historical landslide occurrences, which is why we ensured that training and test sets contain different, "new" hillslopes and landslides.
@@
 
-In other application scenarios such as the crop classification example given in Section  \@ref(sec:intro) [@pena2015], the objective is to predict the fruit-tree crop type of 'new', unseen fields within an agricultural region.
+In other application scenarios such as the crop classification example given in Section  \ref{sec:intro} [@pena2015], the objective is to predict the fruit-tree crop type of 'new', unseen fields within an agricultural region.
 In this use case we are not at all interested in "predicting" the already known crop type of (other) grid cells within the same agricultural field.
@@
 - How densely are the observations distributed in space and time? Are they more densely distributed than the intended spatial or temporal prediction distance?
-- Is the data naturally grouped, e.g. because of the spatial extent of the studied objects, or as a consequence of multi-level (cluster) sampling?
+- Is the data naturally grouped, e.g., because of the spatial extent of the studied objects, or as a consequence of multi-level (cluster) sampling?
 - With an eye on environmental blocking and extrapolation in feature space, is it intended to apply the model to predict 'new' outcomes for unobserved values of predictor variables?
@@
 
-The \pkg{mlr3spatiotempcv} package is the first package to bundle and categorize spatiotemporal resampling methods implemented in multiple other packages in R.
+The \pkg{mlr3spatiotempcv} package is the first package to bundle and categorize spatiotemporal resampling methods implemented in multiple other packages in \proglang{R}.
 The available resampling techniques allow users to vary the scale or granularity of the resampled spatiotemporal units as well as their shape and possible buffer distance between training and test samples.

### 2022-05-23 R/TaskClassifST.R

@@ -13,6 +13,10 @@
 #' changed by setting `extra_args$coords_as_features = TRUE`.
 #'
 #' @family Task
+#' @section S3 Methods:
+#' * `as_task_classif.TaskClassifST(x, ...) `\cr
+#'    * `x` [Task]\cr
+#'      \pkg{mlr3} [Task] object
 #' @export
 #' @examples
 #' if (mlr3misc::require_namespaces(c("sf", "blockCV"), quietly = TRUE)) {
@@ -142,3 +146,8 @@ TaskClassifST = R6::R6Class("TaskClassifST",
     extra_args = NULL
   )
 )
+
+#' @export
+as_task_classif.TaskClassifST = function(x, ...) {
+  TaskClassif$new(id = x$id, backend = x$backend, target = x$target_names, positive = x$positive)
+}

### 2022-05-23 paper/code.R
- Update S3 methods of `as_task_regr.TaskRegrST` and `as_task_classif.TaskClassifST` (#178)
* paper: various style fixes

* update `as_task_classif` and `as_task_regr` S3 methods

@@ -1,200 +1,0 @@
-if (!requireNamespace(c("blockCV"), quietly = TRUE)) install.packages(c("blockCV", "sf", "patchwork", "sperrorest", "ggtext", "plotly"))
-
-
-library("mlr3")
-library("mlr3spatiotempcv")
-task = tsk("ecuador")
-rsmp_buffer = rsmp("spcv_buffer", theRange = 1000)
-rsmp_buffer
-
-autoplot(rsmp_buffer,
-  size = 0.8, task = task, fold_id = 1) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
-
-
-library("mlr3")
-library("mlr3spatiotempcv")
-requireNamespace("blockCV")
-task = tsk("ecuador")
-rsmp_disc = rsmp("spcv_disc", folds = 100, radius = 300L, buffer = 400L)
-rsmp_disc
-
-autoplot(rsmp_disc,
-  size = 0.8, task = task, fold_id = 1) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
-
-
-rsmp_coords = rsmp("spcv_coords", folds = 5)
-
-autoplot(rsmp_coords,
-  size = 0.8, fold_id = 1, task = task) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
-
-
-rsmp_tiles = rsmp("spcv_tiles", nsplit = c(3L, 4L))
-
-autoplot(rsmp_tiles,
-  size = 0.8, fold_id = 1, task = task) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
-
-
-breaks = quantile(task$data()$dem, seq(0, 1, length = 6))
-zclass = cut(task$data()$dem, breaks, include.lowest = TRUE)
-
-rsmp_custom = rsmp("custom_cv")
-rsmp_custom$instantiate(task, f = zclass)
-
-autoplot(rsmp_custom, size = 0.8, task = task, fold_id = 1) *
[…]

### 2022-05-24 paper/article.Rmd

@@
   \usepackage{scrextend}
+  \usepackage{underscore}
   \usepackage{tablefootnote}
@@
 
-### Spatial leave-one-out with buffer
+### Spatial leave-one-out with buffer --- \code{"spcv_buffer"} {short-title='Spatial leave-one-out with buffer --- "spcv_buffer"' }
 
@@
 
-### Leave-one-disc-out with optional buffer
+### Leave-one-disc-out with optional buffer --- \code{"spcv_disc"} {short-title='Leave-one-disc-out with optional buffer --- "spcv_disc"' }
 
@@
 
-### Clustering-based: using coordinates
+### Clustering-based: using coordinates --- \code{"spcv_coords"} {short-title='Clustering-based: using coordinates --- "spcv_coords"' }
 
@@
 
-### Geometric: using rectangular blocks
+### Geometric: using rectangular blocks --- \code{"spcv_tiles"} {short-title='Geometric: using rectangular blocks --- "spcv_tiles"' }
 
@@
 
-<!-- FIXME: -->
-### Custom: "customcv" in \pkg{mlr3} {short-title="Custom: customcv in mlr3" #sec:custom-cv}
-<!-- {#sec:custom-cv} -->
-<!-- \subsubsection[Custom: \code{"customcv"} in mlr3]{Custom: \code{"custom\_cv"} in \pkg{mlr3}} -->
+### Custom: `"custom_cv"` in \pkg{mlr3} {short-title='Custom: "custom_cv" in mlr3' #sec:custom-cv}
 
@@
 
-### Geometric: using rectangular blocks
+### Geometric: using rectangular blocks --- \code{"spcv_block"} {short-title='Geometric: using rectangular blocks --- "spcv_block"' }
 
@@
 
-<!-- FIXME: underscore missing -->
-###  Custom: "cv" with grouping in \pkg{mlr3} {short-title="Custom cv with grouping in mlr3" #sec:custom-cv-grouping}
+###  Custom: `"cv"` with grouping in \pkg{mlr3} {short-title='Custom: "cv" with grouping in mlr3' #sec:custom-cv-grouping}
 
@@
 
-### Clustering: using feature-based clustering
+### Clustering: using feature-based clustering --- \code{"spcv_env"} {short-title='Clustering: using feature-based clustering --- "spcv_env"' }
 
@@
 
-#### CV at the time-point level: "leave-time-out" (LTO)
+#### CV at the time-point level: "leave-time-out" (LTO) --- \code{"sptcv_cstf"} {short-title='CV at the time-point level: "leave-time-out" (LTO) --- "sptcv_cstf"' }
 
@@
 
-#### CV at the location level: "leave-location-out" (LLO)
+#### CV at the location level: "leave-location-out" (LLO) --- \code{"sptcv_cstf"} {short-title='CV at the location level: "leave-location-out" (LLO) --- "sptcv_cstf"' }
 
@@
 
-#### "Leave-location-and-time-out" (LLTO)
+#### "Leave-location-and-time-out" (LLTO) --- \code{"sptcv_cstf"} {short-title='"Leave-location-and-time-out" (LLTO) --- "sptcv_cstf"' }
 

### 2022-05-26 doc/spatiotemp-viz.Rmd
adjust CLUTO to new approach and update test snapshots

@@
 task_st = tsk("cookfarm")
-resampling = rsmp("sptcv_cluto", folds = 5, time_var = "Date")
+resampling = rsmp("sptcv_cluto", folds = 5)
 resampling$instantiate(task_st)

### 2022-05-26 R/autoplot.R

@@ -382,7 +382,7 @@ plot.ResamplingRepeatedSpCVCoords = function(x, ...) {
 #'   library(mlr3)
 #'   library(mlr3spatiotempcv)
 #'   task_st = tsk("cookfarm")
-#'   resampling = rsmp("sptcv_cluto", folds = 5, time_var = "Date")
+#'   resampling = rsmp("sptcv_cluto", folds = 5)
 #'   resampling$instantiate(task_st)
 #'
 #'   # plot

### 2022-06-03 doc/spatiotemp-viz.Rmd
- chore: rename task cookfarm to `cookfarm_mlr3`

@@
 library(mlr3spatiotempcv)
-task_st = tsk("cookfarm")
+task_st = tsk("cookfarm_mlr3")
 resampling = rsmp("sptcv_cluto", folds = 5)

### 2022-06-03 R/ResamplingRepeatedSptCVCluto.R

@@ -12,7 +12,7 @@
 #' if (mlr3misc::require_namespaces("skmeans", quietly = TRUE)) {
 #'   library(mlr3)
 #'   library(mlr3spatiotempcv)
-#'   task = tsk("cookfarm")
+#'   task = tsk("cookfarm_mlr3")
 #'   task$set_col_roles("Date", "time")
 #'
 #'   # Instantiate Resampling

### 2022-06-03 R/ResamplingRepeatedSptCVCstf.R

@@ -14,7 +14,7 @@
 #' @export
 #' @examples
 #' library(mlr3)
-#' task = tsk("cookfarm")
+#' task = tsk("cookfarm_mlr3")
 #' task$set_col_roles("SOURCEID", roles = "space")
 #' task$set_col_roles("Date", roles = "time")
 #'

### 2022-06-03 R/Task_regr_cookfarm_profiles.R

@@ -8,7 +8,7 @@
 #' @section Usage:
 #' ```
 #' mlr_tasks$get("cookfarm")
-#' tsk("cookfarm")
+#' tsk("cookfarm_mlr3")
 #' ```
 #'
 #' @description
@@ -49,7 +49,7 @@
 #' @template seealso_task
 "cookfarm_mlr3"
 
-load_task_cookfarm = function(id = "cookfarm") {
+load_task_cookfarm = function(id = "cookfarm_mlr3") {
   b = mlr3::as_data_backend(cookfarm_mlr3)
   b$hash = "_mlr3_tasks_cookfarm_"
   task = TaskRegrST$new(

### 2022-06-03 R/as_task_regr_st.R

@@ -36,7 +36,7 @@
 #'   as_task_regr_st(cookfarm_sf, target = "PHIHOX")
 #'
 #'   # TaskRegrST
-#'   task = tsk("cookfarm")
+#'   task = tsk("cookfarm_mlr3")
 #'   as_task_regr_st(task)
 #' }
 as_task_regr_st = function(x, ...) {

### 2022-06-03 doc/mlr3spatiotempcv.Rmd
- chore: rename task cookfarm to `cookfarm_mlr3`

---
title: "Getting Started"
output: rmarkdown::html_vignette
bibliography: ../inst/references.bib
link-citations: yes
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
set.seed(42)
library("mlr3")
library("mlr3spatiotempcv")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This package adds resampling methods for the {mlr3} package framework suited for spatial, temporal and spatiotemporal data.
These methods can help to reduce the influence of [autocorrelation](https://mlr3book.mlr-org.com/spatiotemporal.html#spatiotemporal-intro) on performance estimates when performing cross-validation.
While this article gives a rather technical introduction to the package, a more applied approach can be found in the [mlr3book section on "Spatiotemporal Analysis"](https://mlr3book.mlr-org.com/spatiotemporal.html#spatiotemporal).

After loading the package via `library("mlr3spatiotempcv")`, the spatiotemporal resampling methods and example tasks provided by {mlr3spatiotempcv} are available to the user.

In mlr3, dictionaries are used for overview purposes of available methods.
The following shows which dictionaries get appended with new entries.

# Task Type

Additional task types:

-   `TaskClassifST`

-   `TaskRegrST`

```{r}
mlr_reflections$task_types
```

# Task Column Roles

Additional column roles:

-   `coordinates`

```{r}
mlr_reflections$task_col_roles
```

# Resampling Methods

Additional resampling methods:

-   `spcv_block`

-   `spcv_buffer`

-   `spcv_coords`

-   `spcv_env`

-   `sptcv_cluto`

-   `sptcv_cstf`

and their respective repeated versions.

```{r, warning=FALSE}
as.data.table(mlr_resamplings)
```

# Examples Tasks

Additional example tasks:

- `tsk("ecuador")` (spatial, classif)
- `tsk("cookfarm_mlr3")` (spatiotemp, regr)

# Upstream Packages and Scientific References

The following table lists all methods implemented in {mlr3spatiotempcv}, their upstream R package and scientific references.

+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+
| Literature                                                               | Package                                                                  | Reference                                                                | mlr3 Sugar                                                               |
+==========================================================================+==========================================================================+==========================================================================+==========================================================================+
| Spatial Buffering                                                        | [blockCV](https://github.com/rvalavi/blockCV)                            | @valavi2018                                                              | `rsmp("spcv_buffer")`                                                    |
+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+
| Spatial Blocking                                                         | [blockCV](https://github.com/rvalavi/blockCV)                            | @valavi2018                                                              | `rsmp("spcv_block")`                                                     |
+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+
| Spatial CV                                                               | [sperrorest](https://github.com/giscience-fsu/sperrorest)                | @brenning2012                                                            | `rsmp("spcv_coords")`                                                    |
+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+
| Environmental Blocking                                                   | [blockCV](https://github.com/rvalavi/blockCV)                            | @valavi2018                                                              | `rsmp("spcv_env")`                                                       |
+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+
| \-                                                                       | \-                                                                       | \-                                                                       | `rsmp("sptcv_cluto")`                                                    |
+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+
| Leave-Location-and-Time-Out                                              | [CAST](https://github.com/HannaMeyer/CAST)                               | @meyer2018                                                               | `rsmp("sptcv_cstf")`                                                     |
+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+
| Spatiotemporal Clustering                                                | [skmeans](https://github.com/cran/skmeans)                               | @zhao2002                                                                | `rsmp("repeated_sptcv_cluto")`                                           |
+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+
| ------------------------------------------------------------------------ | ------------------------------------------------------------------------ | ------------------------------------------------------------------------ | ------------------------------------------------------------------------ |
+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+
| Repeated Spatial Blocking                                                | [blockCV](https://github.com/rvalavi/blockCV)                            | @valavi2018                                                              | `rsmp("repeated_spcv_block")`                                            |
+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+
| Repeated Spatial CV                                                      | [sperrorest](https://github.com/giscience-fsu/sperrorest)                | @brenning2012                                                            | `rsmp("repeated_spcv_coords")`                                           |
+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+
| Repeated Env Blocking                                                    | [blockCV](https://github.com/rvalavi/blockCV)                            | @valavi2018                                                              | `rsmp("repeated_spcv_env")`                                              |
+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+
| \-                                                                       | \-                                                                       | \-                                                                       | `rsmp("repeated_sptcv_cluto")`                                           |
+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+
| Repeated Leave-Location-and-Time-Out                                     | [CAST](https://github.com/HannaMeyer/CAST)                               | @meyer2018                                                              | `rsmp("repeated_sptcv_cstf")`                                            |
+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+
| Repeated Spatiotemporal Clustering                                       | [skmeans](https://github.com/cran/skmeans)                               | @zhao2002                                                                | `rsmp("repeated_sptcv_cluto")`                                           |
+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+--------------------------------------------------------------------------+

# References

### 2022-06-03 R/TaskRegrST.R

@@ -5,7 +5,7 @@
 #' classification problems.
 #'
 #' A spatial example task is available via `tsk("ecuador")`, a spatiotemporal
-#' one via `tsk("cookfarm")`.
+#' one via `tsk("cookfarm_mlr3")`.
 #'
 #' The coordinate reference system passed during initialization must match the
 #' one which was used during data creation, otherwise offsets of multiple meters

### 2022-06-03 tests/testthat/test-1-autoplot.R
use plot col role for cstf plotting

@@ -329,7 +329,7 @@ test_that("plot() works for 'repeated_spcv_env'", {
 
 # sptcv_cstf -------------------------------------------------------------------
 
-# 2D -------------------------------------------------------------------------
+# 2D ---------------------------------------------------------------------------
 test_that("plot() works for 'sptcv_cstf' 2D - time_var", {
   skip_if_not_installed("vdiffr")
 
@@ -337,24 +337,22 @@ test_that("plot() works for 'sptcv_cstf' 2D - time_var", {
   skip_on_os("linux")
   skip_on_os("windows")
 
-  set.seed(42)
-
-  task = tsk("cookfarm_mlr3")
-  task$set_col_roles("Date", roles = "time")
+  tsk_cookfarm_sub$set_col_roles("Date", roles = "time")
+  tsk_cookfarm_sub$col_roles$space = character()
   rsp = rsmp("sptcv_cstf", folds = 4)
-  rsp$instantiate(task)
+  rsp$instantiate(tsk_cookfarm_sub)
 
-  p1 = autoplot(rsp, task = task)
-  p2 = autoplot(rsp, task, 1, plot3D = FALSE)
+  p1 = autoplot(rsp, task = tsk_cookfarm_sub)
+  p2 = autoplot(rsp, tsk_cookfarm_sub, 1, plot3D = FALSE)
   # plot() would force image printing here
-  p3 = autoplot(rsp, task, c(1, 2))
+  p3 = suppressMessages(autoplot(rsp, tsk_cookfarm_sub, c(1, 2)))
 
   expect_true(is.ggplot(p1))
   expect_true(is.ggplot(p2))
   expect_list(p3)
 
-  p4 = autoplot(rsp, task, repeats_id = 2)
-  p5 = autoplot(rsp, task, fold_id = 1, repeats_id = 2)
+  p4 = autoplot(rsp, tsk_cookfarm_sub, repeats_id = 2)
+  p5 = autoplot(rsp, tsk_cookfarm_sub, fold_id = 1, repeats_id = 2)
 
   vdiffr::expect_doppelganger("SptCVCstf 2D time_var all test sets", p1)
   vdiffr::expect_doppelganger("SptCVCstf 2D time_var - Fold 1", p2)
@@ -370,25 +368,22 @@ test_that("plot() works for 'sptcv_cstf' 2D - space_var", {
[…]

### 2022-06-17 paper/article.Rmd
add all changes

@@
     #   in_header: header.tex
+# nocite: |
+#   @ploton2020
 ---
@@
 # for knitr spin later
-knitr::opts_chunk$set(fig.path = "./figs/",
+knitr::opts_chunk$set(
   cache = TRUE,
@@
 ```
+
+\nocite{ploton2020}
 
@@
 
+From a user perspective, this package structure results in the following workflow for model assessment with \pkg{mlr3spatiotempcv} within \pkg{mlr3}: After choosing a ML algorithm that is supported by \pkg{mlr3} and setting up a learner object, users need to select hyperparameters that should be tuned and specify these in a `paradox::ParamSet`.
+Next, a suitable resampling scheme available within \pkg{mlr3spatiotempcv} is selected that mimics the spatial and/or temporal structure of the prediction task, such as spatial extrapolation, or forecasting of spatial time series.
+This information is used to create a `Resampling` object which is used within a (nested) CV to estimate the model performance.
+When using nested CV, the resampling schemes in the inner (tuning, `mlr3tuning::AutoTuner`) and outer loop (performance estimation, `resample()`) should be identical [@schratz2019].
+To evaluate the (nested) resampling, an adequate performance measure with respect to the response variable, such as the misclassification rate (classification) or the root-mean-square error (regression), must be selected and specified within `mlr3tuning::AutoTuner` and `resample$score()`.
+These choices now allow the user to execute the model assessment via either `resample()` (single model) or `benchmark()` (multiple models), and the results can be summarized visually (via \pkg{mlr3viz}) or in tabular form by accessing the respective fields of the returned `ResampleResult` object.
+
+Additional examples and tutorial can be found in the [mlr3book](https://mlr3book.mlr-org.com) or the [mlr3gallery](https://mlr-org.com/gallery).
+
 # Spatiotemporal partitioning methods and their implementation {#sec:implementation}
@@
 The use of a buffer is theoretically possible in any scenario but in practice only offered by specific method implementations.", out.width="100%",fig.pos="ht"}
-knitr::include_graphics("pdf/rsmp-schema.pdf")
+knitr::include_graphics(here::here("paper/pdf/rsmp-schema.pdf"))
 ```
@@
 
+% this is required to not expand the authors in the table citations on first use
+\shortcites{ploton2020, karasiak2021, moller2021, endicott2017, wu2020, geiss2017, morera2021, zurell2020, brenning2015, bebber2017, jensen2021, escobar2021, stewart2021, reitz2021, egli2020, gao2019}
+
 \begin{table}[ht]
@@
       \\
-      Type                            & Sub-type & Name                        & \proglang{R} package     & \# pub. \\
+      Type                            & Sub-type & Name                        & \proglang{R} package   & References \\
       \toprule
-      \multirow{2}{*}{Spatial leave-one-out}        & single point, with buffer &  \texttt{"spcv\_buffer"} \tablefootnote{\label{ploton2020}\cite{ploton2020}} \tablefootnote{\label{diesing2020}\cite{diesing2020}} & \pkg{blockCV}  & 2   \\
+      \multirow{4}{*}{\specialcell{Spatial \\ leave-one-\\out}}        & single point, with buffer &  \texttt{"spcv\_buffer"} & \pkg{blockCV} (2) & \specialcell{\cite{ploton2020} \\ \cite{diesing2020}} \\
       \cmidrule{2-5}
-       & disc, with buffer  & \texttt{"spcv\_disc"} \tablefootnote{\label{endicott2017}\cite{endicott2017}} \tablefootnote{\label{karasiak2021}\cite{karasiak2021}} \tablefootnote{\label{moller2021}\cite{moller2021}} & \pkg{sperrorest}      & 3   \\
+       & disc, with buffer  & \texttt{"spcv\_disc"}  & \pkg{sperrorest} (3) & \specialcell{\cite{karasiak2021} \\ \cite{moller2021} \\ \cite{endicott2017}} \\
       \midrule
-      \multirow{4}{*}{Leave-one-block-out CV} &  clustering of coordinates & \texttt{"spcv\_coords"}  \tablefootnote{\label{morera2021}\cite{morera2021}} \tablefootnote{\label{geiss2017}\cite{geiss2017}} \tablefootnote{\label{wu2020}\cite{wu2020}} & \pkg{sperrorest}       & 6 \\
+      \multirow{4}{*}{\specialcell{Leave-one-\\block-out \\ CV}} &  clustering of coordinates & \texttt{"spcv\_coords"} & \pkg{sperrorest} (6)      & \specialcell{\cite{morera2021} \\ \cite{geiss2017} \\ \cite{wu2020}} \\
             \cmidrule{2-5}
-                                      &  geometric: rectangular & \texttt{"spcv\_tiles"} \tablefootnote{\label{bebber2017}\cite{bebber2017}} \tablefootnote{\label{brenning2015}\cite{brenning2015}} \tablefootnote{\label{zurell2020}\cite{zurell2020}} & \pkg{sperrorest}    &  4  \\
+                                      &  geometric: rectangular & \texttt{"spcv\_tiles"} & \pkg{sperrorest}    & \specialcell{\cite{bebber2017} \\ \cite{zurell2020} \\ \cite{brenning2015}}  \\
                                                   \cmidrule{2-5}
-                                      & custom  &  \texttt{"custom\_cv"}           & \pkg{mlr3} & - \\
+                                      & custom  &  \texttt{"custom\_cv"}           & \pkg{mlr3} (0) & - \\
       \midrule
-      \multirow{4}{*}{CV at the block level} &  geometric: rectangular & \texttt{"spcv\_block"}  \tablefootnote{\label{jensen2021}\cite{jensen2021}} \tablefootnote{\label{escobar2021}\cite{escobar2021}} \tablefootnote{\label{stewart2021}\cite{stewart2021}} & \pkg{blockCV}       & 28 \\
+      \multirow{4}{*}{\specialcell{CV at the \\ block level}} &  geometric: rectangular & \texttt{"spcv\_block"}  & \pkg{blockCV} (28)      & \specialcell{\cite{jensen2021} \\ \cite{escobar2021} \\ \cite{stewart2021}} \\
   \cmidrule{2-5}
-                                      &  custom & \texttt{"cv"} with grouping  & \pkg{mlr3}    &  -  \\
+                                      &  custom & \texttt{"cv"} with grouping  & \pkg{mlr3} (0)  &  -  \\
   \cmidrule{2-5}
-                                      &  clustering in feature space & \texttt{"spcv\_env"} \footref{morera2021}  & \pkg{blockCV}    &  1  \\
+                                      &  clustering in feature space & \texttt{"spcv\_env"} & \pkg{blockCV} (1)   &  \cite{morera2021}  \\
       \midrule
-      \multirow{3}{*}{CV for spatiotemporal data} &  custom & \texttt{"sptcv\_cstf"}  \tablefootnote{\label{egli2020}\cite{egli2020}} \tablefootnote{\label{gao2019}\cite{gao2019}} \tablefootnote{\label{reitz2021}\cite{reitz2021}} & \pkg{CAST}       & 6 \\
+      \multirow{3}{*}{\specialcell{Spatiotemp. \\ CV}} &  custom & \texttt{"sptcv\_cstf"} & \pkg{CAST} (6)      & \specialcell{\cite{gao2019} \\ \cite{reitz2021} \\ \cite{egli2020}} \\
                                                   \cmidrule{2-5}
-                                      &  clustering: space--time coord. & \texttt{"sptcv\_cluto"}  & \pkg{skmeans}    & - \\
+                                      &  clustering: custom & \texttt{"sptcv\_cluto"}  & \pkg{skmeans} (0)    & - \\
                                       \bottomrule
@@
     The "Name" column shows the \pkg{mlr3} method name as found in the \texttt{mlr3::mlr\_resamplings} dictionary.
-    "\# pub." indicates the number of studies that were found to have used this resampling technique until May 2021.
-    For each method, up to three randomly selected references were added to the table as footnotes.}
+    The count in brackets after the package name represents the number of studies that were found having used this resampling technique until May 2021.
+    For each method, up to three randomly selected references were added to the table.}
     \label{tab:sptcv-methods}
@@
 task_cv = tsk("ecuador")
-group = as.factor(kmeans(task$coordinates(), 8)$cluster)
+group = as.factor(kmeans(task_cv$coordinates(), 8)$cluster)
 task_cv$cbind(data.frame("group" = group))
@@
 
-plot_env_single = autoplot(rsmp_env, size = 0.5, fold_id = 1, task = task) +
+plot_env_single = autoplot(rsmp_env, size = 0.3, fold_id = 1, task = task) +
 
-  plot_env_multi = autoplot(rsmp_env_multi, size = 0.5, fold_id = 1, task = task)
+  plot_env_multi = autoplot(rsmp_env_multi, size = 0.3, fold_id = 1, task = task)
 
@@
 plot_env_single = autoplot(rsmp_env,
-  size = 0.5, fold_id = 1, task = task) *
+  size = 0.3, fold_id = 1, task = task) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
@@
 plot_env_multi = autoplot(rsmp_env_multi,
-  size = 0.5, fold_id = 1, task = task) *
+  size = 0.3, fold_id = 1, task = task) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
@@
 
-In the `cookfarm` example dataset, the `Date` variable was reduced to five unique levels for better visualization, and then used to create a spatiotemporal regression task in \pkg{mlr3spatiotempcv} (Figure \ref{fig:lto}).
+In the `cookfarm_mlr3` example dataset, the `Date` variable was reduced to five unique levels for better visualization, and then used to create a spatiotemporal regression task in \pkg{mlr3spatiotempcv} (Figure \ref{fig:lto}).
+In `autoplot()`, a stratified sample based on the partitions is taken to reduce the number of points plotted.
 
 ```{r, results='hide'}
-data = cookfarm_sample
-data$Date = rep(c(
+data = cookfarm_mlr3
+set.seed(42)
+data$Date = sample(rep(c(
   "2020-01-01", "2020-02-01", "2020-03-01", "2020-04-01",
-  "2020-05-01"), times = 1, each = 100)
-b = mlr3::as_data_backend(data)
-task_spt = TaskRegrST$new(
-  id = "cookfarm", b, target = "PHIHOX",
-  extra_args = list(
+  "2020-05-01"), times = 1, each = 35768))
+task_spt = as_task_regr_st(data,
+  id = "cookfarm", target = "PHIHOX",
   coordinate_names = c("x", "y"), coords_as_features = FALSE,
   crs = 26911)
-)
+task_spt$set_col_roles("Date", roles = "time")
 
-rsmp_cstf_time = rsmp("sptcv_cstf", folds = 5, time_var = "Date")
-rsmp_cstf_time$instantiate(task_spt)
+rsmp_cstf_time = rsmp("sptcv_cstf", folds = 5)
 
 autoplot(rsmp_cstf_time,
-  fold_id = 5, task = task_spt, plot3D = TRUE
+  fold_id = 5, task = task_spt, plot3D = TRUE,
+  sample_fold_n = 3000L
 )
@@
 
-```{r lto, echo=FALSE, out.width="70%", fig.cap='Perspective plot of `"leave-time-out"` CV from package \\pkg{CAST} (method `"sptcv\\_cstf"` and argument `time\\_var = "Date"`).
+```{r lto, echo=FALSE, out.width="80%", fig.cap='Perspective plot of `"leave-time-out"` CV from package \\pkg{CAST} (method `"sptcv\\_cstf"` and column role `"time" = "Date"`).
 Only five folds and five time points were used in this example.
 Note that the blue dots correspond to five discrete time levels, which appear as a point cloud due to the viewing angle.', fig.pos="ht"}
-knitr::include_graphics("png/lto.png")
+# workflow
+
+knitr::include_graphics("pdf/lto.pdf")
 ```
@@
 
+To tell the resampling method to use the 'space' column for partitioning, the 'time' column needs to be unset and the 'space' column defined.
+Because the temporal variable "Date" is not in use in this scenario, `autoplot()` needs to be instructed explicitily to use it for 3D plotting via argument `plot_time_var`.
+
 ```{r results='hide'}
-rsmp_cstf_loc = rsmp("sptcv_cstf", folds = 5, space_var = "SOURCEID")
-rsmp_cstf_loc$instantiate(task_spt)
+task_spt$col_roles$time = character()
+task_spt$set_col_roles("SOURCEID", roles = "space")
+
+rsmp_cstf_loc = rsmp("sptcv_cstf", folds = 5)
 
 autoplot(rsmp_cstf_loc,
-  fold_id = 5, task = task_spt, plot3D = TRUE)
+  fold_id = 5, task = task_spt,
+  plot3D = TRUE, plot_time_var = "Date",
+  sample_fold_n = 3000L)
 ```
 
-```{r llo, echo=FALSE, out.width="70%", fig.cap='Birds-eye view of `"leave-location-out"` CV from package \\pkg{CAST} (method `"sptcv\\_cstf"` and argument `space\\_var = "SOURCEID"`).', fig.pos="ht"}
-knitr::include_graphics("png/llo.png")
+```{r llo, echo=FALSE, out.width="70%", fig.cap='Birds-eye view of `"leave-location-out"` CV from package \\pkg{CAST} (method `"sptcv\\_cstf"` and column role `"space" = "SOURCEID"`).', fig.pos="ht"}
+knitr::include_graphics("pdf/llo.pdf")
 ```
@@
 ```{r results='hide'}
-rsmp_cstf_time_loc = rsmp("sptcv_cstf",
-  folds = 5,
-  space_var = "SOURCEID", time_var = "Date")
-rsmp_cstf_time_loc$instantiate(task_spt)
+task_spt$set_col_roles("SOURCEID", roles = "space")
+task_spt$set_col_roles("Date", roles = "time")
+
+rsmp_cstf_time_loc = rsmp("sptcv_cstf", folds = 5)
 
@@
   fold_id = 4, task = task_spt, plot3D = TRUE,
-  show_omitted = TRUE)
+  show_omitted = TRUE, sample_fold_n = 3000L)
 ```
 
-```{r llto, echo=FALSE, out.width="70%", fig.cap='Perspective plot of `"leave-location-and-time-out"` CV from package \\pkg{CAST} (method `"sptcv\\_cstf"` and arguments `time\\_var = "Date"` and `space\\_var = "SOURCEID"`).
-The grey points are excluded from both the training and the test set in this fold.', fig.pos="h"}
-knitr::include_graphics("png/llto.png")
+```{r llto, echo=FALSE, out.width="70%", fig.cap='Perspective plot of `"leave-location-and-time-out"` CV from package \\pkg{CAST} (method `"sptcv\\_cstf"` and column roles `"time" = "Date"` and `"space" = "SOURCEID"`).
+The grey points are excluded from both the training and the test set in this example.', fig.pos="h"}
+knitr::include_graphics("pdf/llto.pdf")
 ```
@@
 For demonstration purposes only four CV folds and two repetitions are used.
+
+Besides the practical example shown below, additional tutorials covering \pkg{mlr3} use cases can be found at [mlr3gallery](https://mlr-org.com/gallery).
 
@@
 ```{r create-task}
-backend = mlr3::as_data_backend(ecuador)
-task = TaskClassifST$new(
-  id = "ecuador", backend = backend, target = "slides", positive = "TRUE",
-  extra_args = list(
+data("ecuador", package = "mlr3spatiotempcv")
+task = as_task_classif_st(ecuador, target = "slides", positive = "TRUE",
   coordinate_names = c("x", "y"), coords_as_features = FALSE,
   crs = "EPSG:32717")
-)
 ```
@@
 
-# References
+# References {-}

### 2022-06-23 paper/article.Rmd
update table caption
submission2

@@
   - name: Alexander Brenning
-    orcid: 0000-0001-6640-679
+    orcid: 0000-0001-6640-679X
     affiliation: 'Friedrich Schiller University Jena'
@@
     \begin{tabular}[#1]{@{}l@{}}#2\end{tabular}}
-# tables: true
 bibliography: paper.bib
 output: rticles::jss_article
-  # bookdown::pdf_book:
-  #   base_format: rticles::jss_article
-  #   citation_package: "natbib"
-    # includes:
-    #   in_header: header.tex
-# nocite: |
-#   @ploton2020
 ---
@@
 knitr::opts_chunk$set(
-  cache = TRUE,
+  cache = FALSE,
   cache.lazy = TRUE)
@@
   \endgroup
-  \caption[b]{Spatiotemporal resampling methods available in the \pkg{mlr3} ecosystem.
+  \caption[b]{Available spatiotemporal resampling methods in the \pkg{mlr3} ecosystem.
     The "Name" column shows the \pkg{mlr3} method name as found in the \texttt{mlr3::mlr\_resamplings} dictionary.

### 2022-06-23 paper/code-repr.R

@@ -1,5 +1,5 @@
 # fs::dir_create("paper/figs")
-knitr::purl(input = "paper/paper.Rmd", output = "paper/code.R", quiet = TRUE, documentation = 0)
+knitr::purl(input = "paper/article.Rmd", output = "paper/code.R", quiet = TRUE, documentation = 0)
 
 # add call to install suggested packages
 content = readLines("paper/code.R")
@@ -10,9 +10,9 @@ content = content[6:length(content)]
 # remove all commented lines
 content_no_knitr = content[!grepl("knitr::", content)]
 content_no_comments = content_no_knitr[!grepl("#", content_no_knitr)]
-writeLines(content_no_comments, "paper/article.R")
+writeLines(content_no_comments, "paper/code.R")
 
-source("paper/article.R", chdir = TRUE)
-# knitr::spin("paper/code.R")
-# fs::file_move("code.md", "paper/code.md")
-# fs::file_delete(c("code.md", "paper/code.R", "code.html"))
+source("paper/code.R", chdir = TRUE)
+styler.mlr::style_file("paper/code.R")
+
+knitr::spin("paper/code.R")

### 2022-07-08 tests/testthat/test-TaskClassifST.R
fix sf handling in task constructor

@@ -15,7 +15,8 @@ test_that("printing works", {
 test_that("Supplying a non-spatio temporal task gives descriptive error message", {
   expect_error(
     rsmp("spcv_coords")$instantiate(tsk("boston_housing")),
-    "Must inherit from class 'TaskClassifST' or 'TaskRegrST'.") # nolint
+    "Must inherit from class 'TaskClassifST' or 'TaskRegrST'."
+  ) # nolint
 })
 
 test_that("coordinates can be used as features", {
@@ -68,7 +69,25 @@ test_that("extra_args contains default arguments", {
     coordinate_names = c("x", "y")
   )
 
-  expect_equal(task$extra_args, list(crs = NA_character_,
+  expect_equal(task$extra_args, list(
+    crs = NA_character_,
     coordinate_names = c("x", "y"), coords_as_features = FALSE
   ))
+})
+
+test_that("Task creation works via constructor and sf object", {
+  data_sf = sf::st_as_sf(ecuador, coords = c("x", "y"))
+
+  # create mlr3 task
+  task = TaskClassifST$new("ecuador_sf",
+    backend = data_sf, target = "slides", positive = "TRUE"
+  )
+
+  task_as = as_task_classif_st(data_sf,
+    id = "ecuador_sf",
+    target = "slides", positive = "TRUE"
+  )
+
+
+  expect_equal(task, task_as, ignore_attr = "hash")
 })

### 2022-08-08 paper/article.Rmd
paper: create and save pdfs within article

@@
 
-autoplot(rsmp_cstf_time,
+p_lto = autoplot(rsmp_cstf_time,
   fold_id = 5, task = task_spt, plot3D = TRUE,
+  point_size = 6, axis_label_fontsize = 15,
   sample_fold_n = 3000L
 )
+
+p_lto_print = plotly::layout(p_lto,
+  scene = list(camera = list(eye = list(z = 0.58))),
+  showlegend = FALSE, title = "",
+  margin = list(l = 0, b = 0, r = 0, t = 0))
+
+plotly::save_image(p_lto_print, "pdf/lto.pdf",
+  scale = 2, width = 1000, height = 800)
 ```
@@
 
-autoplot(rsmp_cstf_loc,
+p_llo = autoplot(rsmp_cstf_loc,
   fold_id = 5, task = task_spt,
+  point_size = 6, axis_label_fontsize = 15,
   plot3D = TRUE, plot_time_var = "Date",
   sample_fold_n = 3000L)
+
+p_llo_print =
+  plotly::layout(p_llo,
+    scene = list(camera = list(eye = list(z = 2.5, x = -0.1, y = -0.1))),
+    showlegend = FALSE, title = "", polar = TRUE,
+    margin = list(l = 0, b = 0, r = 0, t = 0))
+
+plotly::save_image(p_llo_print, "pdf/llo.pdf",
+  scale = 2, width = 1000, height = 800)
 ```
@@
 
-autoplot(rsmp_cstf_time_loc,
+p_lto = autoplot(rsmp_cstf_time_loc, point_size = 6,
+  axis_label_fontsize = 15,
   fold_id = 4, task = task_spt, plot3D = TRUE,
   show_omitted = TRUE, sample_fold_n = 3000L)
+
+p_lto_print = plotly::layout(p_lto,
+  scene = list(camera = list(eye = list(z = 0.58))),
+  showlegend = FALSE, title = "",
+  margin = list(l = 0, b = 0, r = 0, t = 0))
+
+plotly::save_image(p_lto_print, "pdf/llto.pdf",
+  scale = 2, width = 1000, height = 800)
 ```

### 2022-08-08 paper/code.R

@@ -1,5 +1,5 @@
 if (!requireNamespace(c("blockCV"), quietly = TRUE)) install.packages(c("blockCV", "sf", "patchwork", "sperrorest", "ggtext", "plotly"))
-
+dir.create("pdf")
 
 library("mlr3")
 library("mlr3spatiotempcv")
@@ -125,33 +125,64 @@ task_spt$set_col_roles("Date", roles = "time")
 
 rsmp_cstf_time = rsmp("sptcv_cstf", folds = 5)
 
-autoplot(rsmp_cstf_time,
+p_lto = autoplot(rsmp_cstf_time,
   fold_id = 5, task = task_spt, plot3D = TRUE,
+  point_size = 6, axis_label_fontsize = 15,
   sample_fold_n = 3000L
 )
 
+p_lto_print = plotly::layout(p_lto,
+  scene = list(camera = list(eye = list(z = 0.58))),
+  showlegend = FALSE, title = "",
+  margin = list(l = 0, b = 0, r = 0, t = 0))
 
+plotly::save_image(p_lto_print, "pdf/lto.pdf",
+  scale = 2, width = 1000, height = 800)
+
+
+knitr::include_graphics("pdf/lto.pdf")
 
 task_spt$col_roles$time = character()
 task_spt$set_col_roles("SOURCEID", roles = "space")
 
 rsmp_cstf_loc = rsmp("sptcv_cstf", folds = 5)
 
-autoplot(rsmp_cstf_loc,
+p_llo = autoplot(rsmp_cstf_loc,
   fold_id = 5, task = task_spt,
+  point_size = 6, axis_label_fontsize = 15,
   plot3D = TRUE, plot_time_var = "Date",
   sample_fold_n = 3000L)
 
+p_llo_print =
+  plotly::layout(p_llo,
+    scene = list(camera = list(eye = list(z = 2.5, x = -0.1, y = -0.1))),
+    showlegend = FALSE, title = "", polar = TRUE,
+    margin = list(l = 0, b = 0, r = 0, t = 0))
+
+plotly::save_image(p_llo_print, "pdf/llo.pdf",
+  scale = 2, width = 1000, height = 800)
+
+knitr::include_graphics("pdf/llo.pdf")
 
 task_spt$set_col_roles("SOURCEID", roles = "space")
 task_spt$set_col_roles("Date", roles = "time")
 
 rsmp_cstf_time_loc = rsmp("sptcv_cstf", folds = 5)
 
[…]

### 2022-08-08 paper/create-plotly-figures.R

@@ -13,40 +13,39 @@ rsmp_cstf_time = rsmp("sptcv_cstf", folds = 5)
 # LTO --------
 p_lto = autoplot(rsmp_cstf_time,
   fold_id = 5, task = task_spt, plot3D = TRUE,
+  point_size = 6, axis_label_fontsize = 15,
   sample_fold_n = 3000L
 )
 
-p_lto_print = layout(p_lto, scene = list(camera = list(eye = list(z = 0.58))), showlegend = F, title = "",
-  margin = list(l = -150, b = 20, r = 0, t = -350))
+p_lto_print = plotly::layout(p_lto,
+  scene = list(camera = list(eye = list(z = 0.58))),
+  showlegend = FALSE, title = "",
+  margin = list(l = 0, b = 0, r = 0, t = 0))
 
 tmp = tempfile(fileext = ".pdf")
-plotly::save_image(p_lto_print, tmp,
-  scale = 2,
-  width = 900, height = 750)
-fs::file_move(tmp, "~/git/mlr-org/mlr3spatiotempcv/paper/pdf/lto.pdf")
+plotly::save_image(p_lto_print, here::here("paper/pdf/lto.pdf"),
+  scale = 2, width = 1000, height = 800)
 
 # LLO ------------
 
 task_spt$col_roles$time = character()
 task_spt$set_col_roles("SOURCEID", roles = "space")
 rsmp_cstf_loc = rsmp("sptcv_cstf", folds = 5)
+
 p_llo = autoplot(rsmp_cstf_loc,
   fold_id = 5, task = task_spt,
+  point_size = 6, axis_label_fontsize = 15,
   plot3D = TRUE, plot_time_var = "Date",
   sample_fold_n = 3000L)
 
 p_llo_print =
-  layout(p_llo,
+  plotly::layout(p_llo,
     scene = list(camera = list(eye = list(z = 2.5, x = -0.1, y = -0.1))),
-    showlegend = F, title = "", polar = T,
-    margin = list(l = -150, b = 20, r = 0, t = -350))
-# p_llo_print
+    showlegend = FALSE, title = "", polar = TRUE,
+    margin = list(l = 0, b = 0, r = 0, t = 0))
 
-tmp = tempfile(fileext = ".pdf")
-plotly::save_image(p_llo_print, tmp,
-  scale = 2,
-  width = 900, height = 750)
[…]

### 2022-08-08 paper/article.Rmd
paper: no abbr in abstract

@@
   Spatial and spatiotemporal machine-learning models require a suitable framework for their model assessment, model selection, and hyperparameter tuning, in order to avoid error estimation bias and over-fitting.
-  This contribution reviews the state-of-the-art in spatial and spatiotemporal CV, and introduces the \proglang{R} package \pkg{mlr3spatiotempcv} as an extension package of the machine-learning framework \pkg{mlr3}.
+  This contribution reviews the state-of-the-art in spatial and spatiotemporal cross-validation, and introduces the \proglang{R} package \pkg{mlr3spatiotempcv} as an extension package of the machine-learning framework \pkg{mlr3}.
   Currently various \proglang{R} packages implementing different spatiotemporal partitioning strategies exist: \pkg{blockCV}, \pkg{CAST}, \pkg{skmeans} and \pkg{sperrorest}.

### 2022-08-08 paper/article.Rmd
paper: remove comments from code

@@
     affiliation: 'Friedrich Schiller University Jena'
-    # affiliaton2: 'Ludwig-Maximilians-University Munich'
-    # use this syntax to add text on several lines
-    # must use German name, see https://www.lmu.de/de/die-lmu/struktur/zentrale-universitaetsverwaltung/kommunikation-und-presse/lmu-brand-guide/content-style-guide/index.html
     address: |
@@
     email: \email{marc.becker@stat.uni-muenchen.de}
-    # To add another line, use \AND at the end of the previous one as above
   - name: Michel Lang
@@
   formatted: "\\pkg{mlr3spatiotempcv}: Spatiotemporal resampling methods for machine learning in R"
-  # If you use tex in the formatted title, also supply version without
   plain:     "mlr3spatiotempcv: spatiotemporal resampling methods for machine learning in R"
-  # For running headers, if needed
-  # short:     "\\pkg{foo}: A Capitalized Title"
 abstract: >
@@
 ```{r setup, echo=FALSE, purl=FALSE}
-# important for correct figure placement
-# see https://stackoverflow.com/a/49395389/4185785
 knitr::knit_hooks$set(plot = function(x, options) {
@@
 
-```{r setup-2, echo=FALSE, purl=T}
-# for knitr spin later
+```{r setup-2, echo=FALSE, purl=TRUE}
 knitr::opts_chunk$set(
@@
 Note that the blue dots correspond to five discrete time levels, which appear as a point cloud due to the viewing angle.', fig.pos="ht"}
-# workflow
-
 knitr::include_graphics("pdf/lto.pdf")
@@
 
-# be less verbose
 lgr::get_logger("bbotk")$set_threshold("warn")

### 2022-08-08 paper/article.Rmd
paper: fix capitalization of figure, section and table

@@
 The following sections give an overview of most implemented partitioning strategies and their visualization options.
-The available methods are further discussed in section \ref{sec:disc}.
+The available methods are further discussed in Section \ref{sec:disc}.
 
@@
 Also, $k$-means clustering may be less suitable for data sets with pre-existing clusters of points and/or with isolated, distant sample locations.
-When distinct clusters of points are present, as in multi-level sampling, it may be better to define clusters using a factor variable (see method `"custom_cv"` in section \ref{sec:custom-cv}).
+When distinct clusters of points are present, as in multi-level sampling, it may be better to define clusters using a factor variable (see method `"custom_cv"` in Section \ref{sec:custom-cv}).
 
@@
 In $k$-fold CV resampling at the block level, there are therefore $k$ partitions, each consisting of $1/k$-th of the blocks.
-The special case in which $k$ equals the number of blocks, CV at the block level simply becomes leave-one-block-out CV, for which dedicated implementations exist (see section \ref{sec:lobo-cv}).
+The special case in which $k$ equals the number of blocks, CV at the block level simply becomes leave-one-block-out CV, for which dedicated implementations exist (see Section \ref{sec:lobo-cv}).
 
@@
 
-The `"spcv_block"` method from package \pkg{blockCV} supports both random and systematic resampling of square blocks with argument `selection = "random"` and `"systematic"`, respectively; (see figures \ref{fig:block-random} and \ref{fig:block-systematic}).
+The `"spcv_block"` method from package \pkg{blockCV} supports both random and systematic resampling of square blocks with argument `selection = "random"` and `"systematic"`, respectively; (see Figure \ref{fig:block-random} and Figure \ref{fig:block-systematic}).
 There are additional options for modeling presence-only data, which is a typical use case in species distribution modeling.
@@
 This approach is better suited for the irregular shape of the present study area than a rectangular partitioning.
-Figures \ref{fig:vis-spcv-eval} and \ref{fig:vis-nspcv-eval} show the contrasting distributions of training and test samples.
+Figure \ref{fig:vis-spcv-eval} and Figure \ref{fig:vis-nspcv-eval} show the contrasting distributions of training and test samples.
 For demonstration purposes only four CV folds and two repetitions are used.
@@
 
-Although the case study example in section \ref{sec:case-study} used the `"spcv_coords"` method for coordinate-based clustering, this should not give the impression that this method is the only method suitable for this example task.
+Although the case study example in Section \ref{sec:case-study} used the `"spcv_coords"` method for coordinate-based clustering, this should not give the impression that this method is the only method suitable for this example task.
 In this application setting, we want to assess how well the model generalized from the concrete set of historical landslide occurrences, which is why we ensured that training and test sets contain different, "new" hillslopes and landslides.

### 2022-08-09 R/zzz.R
update URLs from mlr3book

@@ -8,7 +8,7 @@
 #' @importFrom utils globalVariables packageVersion
 #' @section Main resources:
 #' * Book on mlr3: \url{https://mlr3book.mlr-org.com}
-#' * mlr3book section about spatiotemporal data: \url{https://mlr3book.mlr-org.com/08-special-spatiotemp.html}
+#' * mlr3book section about spatiotemporal data: \url{https://mlr3book.mlr-org.com/special.html#spatiotemporal}
 #' * package vignettes: \url{https://mlr3spatiotempcv.mlr-org.com/dev/articles/}
 #'
 #' ## Miscellaneous \pkg{mlr3} content

### 2022-11-12 paper/article.Rmd
paper: don't exceed page margins with code
style

@@
 
-rsmp_env_multi = rsmp("spcv_env", features = c("distdeforest", "slope"), folds = 5)
+rsmp_env_multi = rsmp("spcv_env", features = c("distdeforest", "slope"),
+  folds = 5)
 
-plot_env_single = autoplot(rsmp_env, size = 0.3, fold_id = 1, task = task) +
+plot_env_single = autoplot(rsmp_env, size = 0.3,
+  fold_id = 1, task = task) +
 
-  plot_env_multi = autoplot(rsmp_env_multi, size = 0.3, fold_id = 1, task = task)
+  plot_env_multi = autoplot(rsmp_env_multi, size = 0.3,
+  fold_id = 1, task = task)
 
@@
 ```{r vis-spcv-non-eval, eval=FALSE}
-autoplot(rsmp_sp, task, fold_id = c(1:2), size = 0.8)
+autoplot(rsmp_sp, task, fold_id = 1:2, size = 0.8)
 ```
@@
 Due to space limitations only the first two folds of the first repetition are shown.", fig.pos="ht", fig.height=6, fig.width=8, echo=FALSE}
-autoplot(rsmp_sp, task, fold_id = c(1:2), size = 0.8) *
+autoplot(rsmp_sp, task, fold_id = 1:2, size = 0.8) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
@@
 ```{r vis-nspcv-non-eval, eval=FALSE}
-autoplot(rsmp_nsp, task, fold_id = c(1:2), size = 0.8)
+autoplot(rsmp_nsp, task, fold_id = 1:2, size = 0.8)
 ```
@@
 ```{r vis-nspcv-eval, fig.cap="Random (non-spatial) four-fold CV partitioning. Only the first two folds of the first repetition are shown.", fig.pos="ht", fig.height=6, fig.width=8, echo=FALSE}
-autoplot(rsmp_nsp, task, fold_id = c(1:2), size = 0.8) *
+autoplot(rsmp_nsp, task, fold_id = 1:2, size = 0.8) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *

### 2023-05-22 R/ResamplingSptCVCluto.R
remove cluto

@@ -1,173 +1,0 @@
-#' @title (skmeans) Spatiotemporal clustering resampling via CLUTO
-#'
-#' @template rox_sptcv_cluto
-#' @name mlr_resamplings_sptcv_cluto
-#'
-#' @references
-#' `r format_bib("zhao2002")`
-#'
-#' @export
-#' @examples
-#' \dontrun{
-#' if (mlr3misc::require_namespaces("skmeans", quietly = TRUE)) {
-#'   library(mlr3)
-#'   library(mlr3spatiotempcv)
-#'   task = tsk("cookfarm_mlr3")
-#'   task$set_col_roles("Date", "time")
-#'
-#'   # Instantiate Resampling
-#'   rcv = rsmp("sptcv_cluto", folds = 5)
-#'   rcv$instantiate(task)
-#'
-#'   # Individual sets:
-#'   rcv$train_set(1)
-#'   rcv$test_set(1)
-#'   # check that no obs are in both sets
-#'   intersect(rcv$train_set(1), rcv$test_set(1)) # good!
-#'
-#'   # Internal storage:
-#'   rcv$instance # table
-#' }
-#' }
-ResamplingSptCVCluto = R6Class("ResamplingSptCVCluto",
-  inherit = mlr3::Resampling,
-  public = list(
-
-    #' @field clmethod [character]\cr
-    #'   Name of the clustering method to use within `vcluster`.
-    #'   See Details for more information.
-    clmethod = NULL,
-
-    #' @field cluto_parameters [character]\cr
-    #'   Additional parameters to pass to `vcluster`.
-    #'   Must be given as a single character string, e.g.
-    #'   `"param1='value1'param2='value2'"`.
-    #'   See the CLUTO documentation for a full list of supported parameters.
-    cluto_parameters = NULL,
-
-    #' @field verbose [logical]\cr
-    #'   Whether to show `vcluster` progress and summary output.
-    verbose = NULL,
-
-    #' @description
-    #' Create an repeated resampling instance using the CLUTO algorithm.
-    #' @param id `character(1)`\cr
-    #'   Identifier for the resampling strategy.
-    #' @param time_var [character]\cr
-    #'  The name of the variable which represents the time dimension.
-    #'  Must be of type numeric.
-    #' @param clmethod [character]\cr
-    #'   Name of the clustering method to use within `vcluster`.
-    #'   See Details for more information.
-    #' @param cluto_parameters [character]\cr
[…]

### 2023-05-22 R/autoplot_spcv_cstf.R

@@ -68,7 +68,6 @@
 #'   - [autoplot.ResamplingSpCVCoords()]
 #'   - [autoplot.ResamplingSpCVEnv()]
 #'   - [autoplot.ResamplingCV()]
-#'   - [autoplot.ResamplingSptCVCluto()]
 #' @examples
 #' \donttest{
 #' if (mlr3misc::require_namespaces(c("sf", "plotly"), quietly = TRUE)) {

### 2023-05-22 doc/mlr3spatiotempcv.R

@@ -1,18 +1,0 @@
-## ----setup, include=FALSE-----------------------------------------------------
-set.seed(42)
-library("mlr3")
-library("mlr3spatiotempcv")
-
-knitr::opts_chunk$set(
-  collapse = TRUE,
-  comment = "#>"
-)
-
-## -----------------------------------------------------------------------------
-mlr_reflections$task_types
-
-## -----------------------------------------------------------------------------
-mlr_reflections$task_col_roles
-
-## ---- warning=FALSE-----------------------------------------------------------
-as.data.table(mlr_resamplings)

### 2023-05-22 tests/testthat/test-ResamplingSptCVCluto.R

@@ -1,65 +1,0 @@
-# spatiotemporal ---------------------------------------------------------------
-
-# nocov start
-
-test_that("resampling iterations equals folds", {
-  skip_on_cran()
-  skip_on_os("mac")
-  skip_on_ci()
-  skip_if_not_installed("skmeans")
-  skip_if_not_installed("mlr3pipelines")
-  skip_if(Sys.info()[["machine"]] != "x86_64")
-
-  rsp = rsmp("sptcv_cluto", folds = 2)
-  rsp$instantiate(task_cluto)
-
-  expect_equal(rsp$iters, 2)
-})
-
-test_that("reps can be printed", {
-  skip_on_cran()
-  skip_on_os("mac")
-  skip_on_ci()
-  skip_if_not_installed("skmeans")
-  skip_if(Sys.info()[["machine"]] != "x86_64")
-
-  rsp = rsmp("repeated_sptcv_cluto",
-    folds = 3, repeats = 5
-  )
-  rsp$instantiate(task_cluto)
-
-  expect_equal(rsp$repeats(4:8), c(2, 2, 2, 3, 3))
-})
-
-test_that("resampling iterations equals folds * repeats", {
-  skip_on_cran()
-  skip_on_os("mac")
-  skip_on_ci()
-  skip_if_not_installed("skmeans")
-  skip_if(Sys.info()[["machine"]] != "x86_64")
-
-  rsp = rsmp("repeated_sptcv_cluto",
-    folds = 3, repeats = 2
-  )
-  rsp$instantiate(task_cluto)
-
-  expect_equal(rsp$iters, 6)
-})
-
-
-# spatial only -----------------------------------------------------------------
-
-test_that("clustering on coords only works", {
-  skip_on_cran()
-  skip_on_os("mac")
-  skip_on_ci()
-  skip_if_not_installed("skmeans")
-  skip_if(Sys.info()[["machine"]] != "x86_64")
-
-  rsp = rsmp("sptcv_cluto", folds = 2)
-  rsp$instantiate(task_cluto)
-
-  expect_equal(rsp$iters, 2)
-})
-
-# nocov end

### 2023-05-22 tests/testthat/test-mlr3pipelines-graph-integration.R

@@ -19,22 +19,3 @@ test_that("spcv methods work in a Graph Learner", {
 
   expect_true(all(out))
 })
-
-test_that("sptcv methods work in a Graph Learner", {
-  skip_if_not_installed("mlr3pipelines")
-
-  skip_on_cran()
-  skip_on_os("mac")
-  skip_if_not_installed("skmeans")
-  rsmps = rsmps(c("sptcv_cstf"), folds = 2)
-
-  out = mlr3misc::map_lgl(rsmps, function(x) {
-    test_graph_learner(task_cluto, x,
-      time_var = "Date",
-      learner = "regr.featureless",
-      measure = "regr.mse"
-    )
-  })
-
-  expect_true(all(out))
-})

### 2023-05-23 man-roxygen/rox_sptcv_cluto.R
Remove cluto algorithm and method (#224)

@@ -1,43 +1,0 @@
-#' @description
-#'  Spatiotemporal cluster partitioning via the `vcluster` executable of the
-#'  CLUTO clustering application.
-#'
-#' This partitioning method relies on the external CLUTO library.
-#' To use it, CLUTO's executables need to be downloaded and installed into
-#' this package.
-#'
-#' See \url{https://gist.github.com/pat-s/6430470cf817050e27d26c43c0e9be72} for
-#' an installation approach that should work on Windows and Linux. macOS is not
-#' supported by CLUTO.
-#'
-#' Before using this method, please check the restrictive copyright shown below.
-#'
-#' @details
-#' By default, `-clmethod='direct'` is passed to the `vcluster` executable in
-#' contrast to the upstream default `-clmethod='rb'`.
-#' There is no evidence or research that this method is the best among the
-#' available ones ("rb", "rbr", "direct", "agglo", "graph", "bagglo").
-#' Also, various other parameters can be set via argument `cluto_parameters` to
-#' achieve different clustering results.
-#'
-#' Parameter `-clusterfile` is handled by \CRANpkg{skmeans} and cannot be
-#' changed.
-#'
-#' @section Copyright:
-#'
-#' CLUTO's copyright is as follows:
-#'
-#' The CLUTO package is copyrighted by the Regents of the University of
-#' Minnesota.
-#' It can be freely used for educational and research purposes by non-profit
-#' institutions and US government agencies only.
-#' Other organizations are allowed to use CLUTO only for evaluation purposes,
-#' and any further uses will require prior approval.
-#' The software may not be sold or redistributed without prior approval.
-#' One may make copies of the software for their use provided that the copies,
-#' are not sold or distributed, are used under the same terms and conditions.
-#' As unestablished research software, this code is provided on an “as is” basis
-#' without warranty of any kind, either expressed or implied.
-#' The downloading, or executing any part of this software constitutes an
-#' implicit agreement to these terms. These terms and conditions are subject to
-#' change at any time without prior notice.

### 2023-05-23 tests/testthat/test-mlr_sptcv_generic.R

@@ -66,23 +66,6 @@ test_that("train and test set getter functions are working", {
   }
 })
 
-test_that("train and test set getter functions are working for sptcv", {
-  skip_on_cran()
-  skip_on_os("mac")
-  skip_if_not_installed("skmeans")
-
-  sptcv_rsp = rsmps(
-    c("repeated_sptcv_cstf"),
-    folds = 2
-  )
-
-  for (i in sptcv_rsp) {
-    i$instantiate(task_cluto)
-    expect_silent(i$train_set(1))
-    expect_silent(i$test_set(1))
-  }
-})
-
 test_that("cloning works", {
   skip_if_not_installed("skmeans")
 

### 2023-05-23 tic.R

@@ -7,12 +7,3 @@ if (ci_on_ghactions() && ci_has_env("BUILD_PKGDOWN")) {
     add_step(step_install_github("mlr-org/mlr3pkgdowntemplate"))
   do_pkgdown()
 }
-
-# write .Renviron with CLUTO path
-if (Sys.info()[["sysname"]] == "Windows") {
-  get_stage("before_script") %>%
-    add_code_step(writeLines(sprintf("CLUTO_PATH=%s/cluto/vcluster.exe", getwd()), ".Renviron"))
-} else if (Sys.info()[["sysname"]] == "Linux") {
-  get_stage("before_script") %>%
-    add_code_step(writeLines(sprintf("CLUTO_PATH=%s/cluto/vcluster", getwd()), ".Renviron"))
-}

### 2023-08-11 paper/article.Rmd
changes

@@
 In this specific situation it would therefore seem rather unwise to train a model on a simple random subsample of grid cells, and to test it on the remaining data, using other grid cells from the same fields, as if one wanted to predict within a field.
-The results from this performance assessment would be over-optimistic, and perhaps badly so.
+The results from this performance assessment would be over-optimistic, and will not reflect the model's true ability to make predictions for new fields.
 To mimic the predictive situation for which the model is trained, one would rather have to resample at the level of fields, not grid cells [@pena2015].
@@
 One limitation of all these packages is the sole focus on model assessment, while the proposed implementation within the \pkg{mlr3} framework also offers seamless integration into model selection and provides parallel execution and enhanced logging abilities.
-It is worth noting that a spatial cross-validation library named \pkg{spacv} has recently been developed for Python3, which can be used with the \pkg{scikit-learn} machine-learning framework [@pedregosa2011].
+During the intitial development of \pkg{mlr3spatiotempcv}, it has been the first toolbox aiming to integrate standalone spatiotemporal validation strategies into an existing machine-learning framework.
+Meanwhile the \pkg{spatialsample} has undertaken similar efforts for the \pkg{tidymodels} framework [@mahoney2023].
+In addition parallel efforts to support spatial validation in the Python world have been made by the \pkg{spacv} and \pkg{spatial-kfold} packages, yet often with only a subset of available methods compared to what \pkg{mlr3spatiotempcv} offers [@ghariani2023].
 
@@
 
-% this is required to not expand the authors in the table citations on first use
 \shortcites{ploton2020, karasiak2021, moller2021, endicott2017, wu2020, geiss2017, morera2021, zurell2020, brenning2015, bebber2017, jensen2021, escobar2021, stewart2021, reitz2021, egli2020, gao2019}
@@
       \midrule
-      \multirow{4}{*}{\specialcell{Leave-one-\\block-out \\ CV}} &  clustering of coordinates & \texttt{"spcv\_coords"} & \pkg{sperrorest} (6)      & \specialcell{\cite{morera2021} \\ \cite{geiss2017} \\ \cite{wu2020}} \\
+      \multirow{4}{*}{\specialcell{Leave-one-\\block-out \\ CV}} &  clustering of coordinates & \texttt{"spcv\_coords"}* & \pkg{sperrorest} (6)      & \specialcell{\cite{morera2021} \\ \cite{geiss2017} \\ \cite{wu2020}} \\
             \cmidrule{2-5}
@@
   \cmidrule{2-5}
-                                      &  clustering in feature space & \texttt{"spcv\_env"} & \pkg{blockCV} (1)   &  \cite{morera2021}  \\
+                                      &  clustering in feature space & \texttt{"spcv\_env"}* & \pkg{blockCV} (1)   &  \cite{morera2021}  \\
       \midrule
-      \multirow{3}{*}{\specialcell{Spatiotemp. \\ CV}} &  custom & \texttt{"sptcv\_cstf"} & \pkg{CAST} (6)      & \specialcell{\cite{gao2019} \\ \cite{reitz2021} \\ \cite{egli2020}} \\
-                                                  \cmidrule{2-5}
-                                      &  clustering: custom & \texttt{"sptcv\_cluto"}  & \pkg{skmeans} (0)    & - \\
+      \multirow{2}{*}{\specialcell{Spatiotemp. \\ CV}} &  custom & \texttt{"sptcv\_cstf"}* & \pkg{CAST} (6)      & \specialcell{\cite{gao2019} \\ \cite{reitz2021} \\ \cite{egli2020}} \\
                                       \bottomrule
@@
     The count in brackets after the package name represents the number of studies that were found having used this resampling technique until May 2021.
-    For each method, up to three randomly selected references were added to the table.}
+    For each method, up to three randomly selected references were added to the table.
+    Methods suffixed with a * in the "Name" column have been reimplemented in \pkg{mlr3spatiotempcv} for efficiency or to reduce upstream dependencies.}
     \label{tab:sptcv-methods}
@@
 ```{r buffer-eval, fig.cap='Visualization of the spatial buffering method from package \\pkg{blockCV} (method "spcv\\_buffer" in \\pkg{mlr3spatiotempcv}).
-The buffer distance is 1000 m.', out.width="40%", fig.pos="ht",echo=FALSE}
+The buffer distance is 1000 m.', out.width="55%", fig.pos="ht",echo=FALSE}
 library("mlr3")
@@
   size = 0.8, task = task, fold_id = 1) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
+  ggplot2::theme(text = ggplot2::element_text(size = 16),
+    plot.title = ggtext::element_textbox(size = 13,
+      r = ggplot2::unit(7, "pt"),
+      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8)) +
+  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 1.5)))
 ```
@@
 ```{r disc-eval, fig.cap='Visualization of one training set / test set combination generated with the leave-one-disc-out method from package \\pkg{sperrorest} (method `"spcv\\_disc"` in \\pkg{mlr3spatiotempcv}).
-The disc has a radius of 300 m and is surrounded by a 400-m buffer.', out.width="40%", echo=FALSE,fig.pos="ht"}
+The disc has a radius of 300 m and is surrounded by a 400-m buffer.', out.width="55%", echo=FALSE,fig.pos="ht"}
 library("mlr3")
@@
 
-```{r coords-eval, fig.cap='Leave-one-block-out CV based on $k$-means clustering of the coordinates as implemented in package \\pkg{sperrorest} (method `"spcv\\_coords"` in \\pkg{mlr3spatiotempcv}).', out.width="40%", fig.pos="ht", echo=FALSE}
+```{r coords-eval, fig.cap='Leave-one-block-out CV based on $k$-means clustering of the coordinates as implemented in package \\pkg{sperrorest} (method `"spcv\\_coords"` in \\pkg{mlr3spatiotempcv}).', out.width="55%", fig.pos="ht", echo=FALSE}
 rsmp_coords = rsmp("spcv_coords", folds = 5)
@@
   size = 0.8, fold_id = 1, task = task) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
+  ggplot2::theme(text = ggplot2::element_text(size = 16),
+    plot.title = ggtext::element_textbox(size = 13,
+      r = ggplot2::unit(7, "pt"),
+      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8)) +
+  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 1.5)))
 ```
@@
 
-```{r tile, fig.cap='Leave-one-block-out resampling from package \\pkg{sperrorest} (method `"spcv\\_tiles"` in package \\pkg{mlr3spatiotempcv} with argument `nsplit = c(3,4)` indicating the number of rows and columns).', out.width="40%", fig.pos="ht", echo=FALSE}
+```{r tile, fig.cap='Leave-one-block-out resampling from package \\pkg{sperrorest} (method `"spcv\\_tiles"` in package \\pkg{mlr3spatiotempcv} with argument `nsplit = c(3,4)` indicating the number of rows and columns).', out.width="55%", fig.pos="ht", echo=FALSE}
 rsmp_tiles = rsmp("spcv_tiles", nsplit = c(3L, 4L))
@@
 ```{r custom-cv, fig.cap='Leave-one-level-out (custom) resampling from package \\pkg{mlr3} (method `"custom\\_cv"`).
-A factor variable is used to define all partitions.', out.width="40%", fig.pos="ht", echo=FALSE}
+A factor variable is used to define all partitions.', out.width="55%", fig.pos="ht", echo=FALSE}
 breaks = quantile(task$data()$dem, seq(0, 1, length = 6))
@@
 autoplot(rsmp_custom, size = 0.8, task = task, fold_id = 1) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
+  ggplot2::theme(text = ggplot2::element_text(size = 16),
+    plot.title = ggtext::element_textbox(size = 13,
+      r = ggplot2::unit(7, "pt"),
+      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8)) +
+  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 1.5)))
 ```
@@
 ```{r block-random, fig.cap='Random resampling of square spatial blocks using the implementation in package \\pkg{blockCV} (method `"spcv\\_block"` with option `selection = "random"` in \\pkg{mlr3spatiotempcv}).
-The size of the squares is 1000 m, and four out of the 19 blocks were assigned to the test partition.', out.width="40%", fig.pos="ht", echo=FALSE}
+The size of the squares is 1000 m, and four out of the 19 blocks were assigned to the test partition.', out.width="55%", fig.pos="ht", echo=FALSE}
 rsmp_block_random = rsmp("spcv_block", range = 1000, folds = 5)
@@
   show_blocks = TRUE, show_labels = TRUE) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
+  ggplot2::theme(text = ggplot2::element_text(size = 16),
+    plot.title = ggtext::element_textbox(size = 13,
+      r = ggplot2::unit(7, "pt"),
+      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8)) +
+  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 1.5)))
 ```
@@
 ```{r block-systematic, fig.cap='Sytematic resampling of square spatial blocks using the implementation in package \\pkg{blockCV} (method `"spcv\\_block"` with option `selection = "systematic"` in \\pkg{mlr3spatiotempcv}).
-The size of the squares is 1000 m, and four out of the 19 blocks were assigned to this test sample.', out.width="40%", fig.pos="ht", echo=FALSE}
+The size of the squares is 1000 m, and four out of the 19 blocks were assigned to this test sample.', out.width="55%", fig.pos="ht", echo=FALSE}
 rsmp_block_systematic = rsmp("spcv_block",
@@
 A complete group or block of observations is therefore assigned to a specific partition, which consequently honors the grouping structure.
-In the deprecated \pkg{mlr} package this concept was referred to as "blocking".
 
 In contrast to geometric or clustering-based blocks, the spatial or temporal location is not used explicitly, but rather implicitly through the spatial or spatiotemporal footprint of each user-defined block.
+Studies which have applied custom grouping approaches include @meyer2018, @anderson2005, @kasurak2011.
 
@@
 rsmp_cv_group = rsmp("cv", folds = 3)$instantiate(task_cv)
-
-print(rsmp_cv_group$instance)
 ```
@@
 A factor variable (named "group") was used to define the grouping.
-Each class is either assigned to the test or training set.', out.width="40%", fig.pos="ht", eval=FALSE}
+Each class is either assigned to the test or training set.', out.width="55%", fig.pos="ht", eval=FALSE}
 autoplot(rsmp_cv_group, size = 0.8, task = task_cv, fold_id = 1)
@@
 A factor variable is used to define the grouping.
-Each class is either assigned to the test or training set.', out.width="40%", fig.pos="ht", echo=FALSE}
+Each class is either assigned to the test or training set.', out.width="55%", fig.pos="ht", echo=FALSE}
 autoplot(rsmp_cv_group, size = 0.8, task = task_cv, fold_id = 1) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
+  ggplot2::theme(text = ggplot2::element_text(size = 16),
+    plot.title = ggtext::element_textbox(size = 13,
+      r = ggplot2::unit(7, "pt"),
+      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8)) +
+  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 1.5)))
 ```
@@
   ggplot2::theme(plot.title = ggtext::element_textbox(size = 8),
-    axis.text = ggplot2::element_text(size = 8))
+    axis.text = ggplot2::element_text(size = 8)) +
+  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 0.8)))
 
@@
 
-A well-known case study is used to demonstrate the application of spatial and non-spatial resampling techniques for model assessment in \pkg{mlr3spatiotempcv}.
+A case study is used to demonstrate the application of spatial and non-spatial resampling techniques for model assessment in \pkg{mlr3spatiotempcv}.
 The objective of landslide susceptibility modeling is to predict how prone to landslide initiation a location is.
@@
 It is also expected that spatial CV results better represent the model's transferability to geologically and topographically similar areas adjacent to the training area.
+Yet it must be kept in mind that using spatial CV might also lead to an overpessimistic estimate of the model's predictive performance, as large parts of the study area are being left out during training.
+This goes in hand with the limited ability to predict to other areas, as discussed in [@wadoux2021, meyer2021].
 The magnitude of the difference between spatial and non-spatial CV estimates may depend on the dataset, the strength on spatial or spatiotemporal autocorrelation, and the learner itself.
@@
 
-<!-- risks of using buffer zone -->
 For example, buffering, or the use of a spatial or temporal separation distance between training and test sets, is currently only implemented for some methods (`"spcv_buffer"`, `"spcv_disc"`, and `"sptcv_cstf"` with both `space_var` and `time_var`).
@@
 
-<!-- only some methods are suited for repeated execution -->
-<!-- spcv_coords: limited variance due to limited k-means possibilities -->
 CV is often executed repeatedly to reduce the possible influence of random variability on CV estimates.
@@
 
-<!-- sign-off -->
 Spatiotemporal cross-validation as a paradigm is not yet fully established in scientific workflows, although it has been discussed intensively for more than a decade now.

### 2023-08-11 paper/article.Rmd
extend llo figure caption

@@
 
-```{r llo, echo=FALSE, out.width="70%", fig.cap='Birds-eye view of `"leave-location-out"` CV from package \\pkg{CAST} (method `"sptcv\\_cstf"` and column role `"space" = "SOURCEID"`).', fig.pos="ht"}
+```{r llo, echo=FALSE, out.width="70%", fig.cap='Birds-eye view of `"leave-location-out"` CV from package \\pkg{CAST} (method `"sptcv\\_cstf"` and column role `"space" = "SOURCEID"`).
+The readers viewpoint is located on top of the figure, looking down.
+The transects are individual locations being left out across multiple time steps.', fig.pos="ht"}
 knitr::include_graphics("pdf/llo.pdf")

### 2023-08-11 paper/article.Rmd
cv -> repeated_cv

@@
 
-To define a resampling strategy, the `rsmp()` function is used to generate a resampling object using four folds and two repetitions following a random sampling logic ("cv").
+To define a resampling strategy, the `rsmp()` function is used to generate a resampling object using four folds and two repetitions following a random sampling logic ("repeated_cv").
 

### 2023-08-11 paper/article.Rmd
wording

@@
 
-The model assessment is now repeated again using spatial CV resampling, for which the only required change is to replace `"repeated_cv"` by `"repeated_spcv_coords"`.
+The model assessment is now repeated again using spatial CV resampling, for which the only required change is to replace `"repeated_cv"` with `"repeated_spcv_coords"`.
 

### 2023-08-11 paper/article.Rmd
switch order of plots
shorten paragraph as suggested by rev2

@@
 
+```{r vis-nspcv-non-eval, eval=FALSE}
+autoplot(rsmp_nsp, task, fold_id = 1:2, size = 0.8)
+```
+
+```{r vis-nspcv-eval, fig.cap="Random (non-spatial) four-fold CV partitioning. Only the first two folds of the first repetition are shown.", fig.pos="ht", fig.height=6, fig.width=8, echo=FALSE}
+autoplot(rsmp_nsp, task, fold_id = 1:2, size = 0.8) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```
+
 ```{r vis-spcv-non-eval, eval=FALSE}
@@
 autoplot(rsmp_sp, task, fold_id = 1:2, size = 0.8) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
-```
-
-```{r vis-nspcv-non-eval, eval=FALSE}
-autoplot(rsmp_nsp, task, fold_id = 1:2, size = 0.8)
-```
-
-```{r vis-nspcv-eval, fig.cap="Random (non-spatial) four-fold CV partitioning. Only the first two folds of the first repetition are shown.", fig.pos="ht", fig.height=6, fig.width=8, echo=FALSE}
-autoplot(rsmp_nsp, task, fold_id = 1:2, size = 0.8) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
@@
 
-Although the case study example in Section \ref{sec:case-study} used the `"spcv_coords"` method for coordinate-based clustering, this should not give the impression that this method is the only method suitable for this example task.
-In this application setting, we want to assess how well the model generalized from the concrete set of historical landslide occurrences, which is why we ensured that training and test sets contain different, "new" hillslopes and landslides.
+Although the case study example in Section \ref{sec:case-study} used the `"spcv_coords"` method for coordinate-based clustering, other methods would be suitable as well.
+In this particular application setting, we want to assess how well the model generalized from the concrete set of historical landslide occurrences, which is why we ensured that training and test sets contain different, "new" hillslopes and landslides.
 Coordinate-based clustering is particularly appealing in this setting because of its ability to adapt to the irregularly shaped study area of this example.
@@
 
-In other application scenarios such as the crop classification example given in Section \ref{sec:intro} [@pena2015], the objective is to predict the fruit-tree crop type of 'new', unseen fields within an agricultural region.
-In this use case we are not at all interested in "predicting" the already known crop type of (other) grid cells within the same agricultural field.
-Also, a model will likely be much better able to predict the crop type within the same field from multitemporal remote-sensing data since all crops within that field are subject to identical management practices (e.g., use of pesticides, pruning of fruit trees, tree spacing), while 'new' fields may be managed differently.
-As a consequence, grid cells from the same field should be grouped into a block, and resampling should be done at the field level (`"cv"` method with grouping) to receive an honest estimate of the model's performance in a relevant predictive situation.
+In other scenarios, such as when predicting crop types across monocultural fields \ref{sec:intro} [@pena2015], it makes sense to group observations into blocks based on previously known boundaries and resample at the polygon level (`"cv"` method with grouping).
 If, in contrast, the objective is to apply the model to an adjacent agricultural region (e.g., adjacent county) where the same crop types are present, it may be advisable to use coordinate-based clustering (`"spcv_coords"` method) to obtain larger, contiguous test regions.

### 2023-08-15 paper/article.Rmd
increase label size and show_omitted points

@@
 
-autoplot(rsmp_buffer, size = 0.8, task = task, fold_id = 1)
+autoplot(rsmp_buffer, size = 0.8, task = task, fold_id = 1, show_omitted = TRUE)
 ```
@@
 autoplot(rsmp_buffer,
-  size = 0.8, task = task, fold_id = 1) *
+  size = 0.8, task = task, fold_id = 1, show_omitted = TRUE) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
@@
 
-autoplot(rsmp_disc, size = 0.8, task = task, fold_id = 1)
+autoplot(rsmp_disc, size = 0.8, task = task, fold_id = 1, show_omitted = TRUE)
 ```
@@
 autoplot(rsmp_disc,
-  size = 0.8, task = task, fold_id = 1) *
+  size = 0.8, task = task, fold_id = 1, show_omitted = TRUE) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
@@
 autoplot(rsmp_block_random, size = 0.8, fold_id = 1, task = task,
-  show_blocks = TRUE, show_labels = TRUE)
+  show_blocks = TRUE, show_labels = TRUE, label_size = 4)
 ```
@@
   size = 0.8, fold_id = 1, task = task,
-  show_blocks = TRUE, show_labels = TRUE) *
+  show_blocks = TRUE, show_labels = TRUE, label_size = 4) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
@@
 autoplot(rsmp_block_systematic, size = 0.8, fold_id = 1, task = task,
-  show_blocks = TRUE, show_labels = TRUE)
+  show_blocks = TRUE, show_labels = TRUE, label_size = 4)
 ```

### 2023-08-23 paper/article.Rmd
include wadoux ref
add knndm code, address reviewer comments in intro

@@
 \pkg{mlr3spatiotempcv} is such a toolbox.
+Yet all the included methods should only be used if no probability sample can be drawn from the data - which is often the case when working with spatial data.
+A design-based approach, if possible, is the preferred way to assess map accuracies [@wadoux2021].
 
@@
 In contrast, packages \pkg{blockCV} and \pkg{ENMeval} were developed for block and buffer resampling with a focus on species distribution modeling [@blockCV; @rest2014; @muscarella2014].
-However, neither of these have been integrated into established machine-learning frameworks such as \pkg{mlr}/\pkg{mlr3} [@mlr3] or \pkg{caret}/\pkg{tidymodels} [@kuhn2020], and all of them lack support for temporal prediction tasks.
+However, neither of these have been integrated into established machine-learning frameworks such as \pkg{mlr3} [@mlr3] or \pkg{caret}/\pkg{tidymodels} [@kuhn2020], and all of them lack support for temporal prediction tasks.
 The \pkg{CAST} package, in contrast, focuses on spatiotemporal prediction tasks and makes use of some functions of the \pkg{caret} framework [@cast; @meyer2018].
-One limitation of all these packages is the sole focus on model assessment, while the proposed implementation within the \pkg{mlr3} framework also offers seamless integration into model selection and provides parallel execution and enhanced logging abilities.
+One limitation of all the packages which provide only a subset of the machine learning workflow, e.g., resampling methods, is the sole focus on model assessment.
+Through the integration into the \pkg{mlr3} framework, \pkg{mlr3spatiotempcv} indirectly offers model selection capabilities, i.e., applying and evaluating various algorithms across different preprocessing setups while being able to make use of parallel execution and enhanced logging abilities.
 During the intitial development of \pkg{mlr3spatiotempcv}, it has been the first toolbox aiming to integrate standalone spatiotemporal validation strategies into an existing machine-learning framework.
-Meanwhile the \pkg{spatialsample} has undertaken similar efforts for the \pkg{tidymodels} framework [@mahoney2023].
+Meanwhile the \pkg{spatialsample} has undertaken similar efforts and added the ability to use spatial resampling methods for the \pkg{tidymodels} framework [@mahoney2023].
 In addition parallel efforts to support spatial validation in the Python world have been made by the \pkg{spacv} and \pkg{spatial-kfold} packages, yet often with only a subset of available methods compared to what \pkg{mlr3spatiotempcv} offers [@ghariani2023].
@@
 Leave-one-disc-out resampling modifies this approach to leave out circular regions centered at observation points.
+This method group naturally discards observations when applying buffers, meaning that these are not used for training or testing.
+Depending on the relative amount of discarded observations with respect to the overall number of observations, this might result in a substantial reduction of the sample size and thus decrease the applicability of the results.
 
@@
 ```
+
+### Clustering-based: using k-fold nearest-neighbour distance matching --- \code{"spcv_knndm"} {short-title='Clustering-based: using kNNDM --- "spcv_knndm"' }
+
+The k-fold nearest-neighboar distance matching method after @linnenbrinck2023 is derived from the NNDM method after @mila.
+By using geographical nearest neighbor distances between test and training points, the method aims to create realistic predictive situations within the CV folds that closely mimic the ones of either existing training and test points, if know, or a provided model domain in which the model is to be applied.
+This is done by minimizing the values ofthe Empirical Cumulative Distribution Function (ECDF) for the respective scenarios outlined above [@linnenbrinck2023].
+
+```{r knndm-non-eval, eval=FALSE}
+rsmp_knndm = rsmp("spcv_knndm", folds = 5)
+
+autoplot(rsmp_knndm, size = 0.8, fold_id = 1, task = task)
+```
+
+```{r coords-eval, fig.cap='Leave-one-block-out CV based on $k$-means clustering of the coordinates as implemented in package \\pkg{sperrorest} (method `"spcv\\_coords"` in \\pkg{mlr3spatiotempcv}).', out.width="40%", fig.pos="ht", echo=FALSE}
+rsmp_coords = rsmp("spcv_coords", folds = 5)
+
+autoplot(rsmp_coords,
+  size = 0.8, fold_id = 1, task = task) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```
+
+FIXME:
 

### 2023-08-23 R/bibentries.R
Add knndm method from {CAST} (#229)

@@ -109,6 +109,17 @@ bibentries = c( # nolint start
     author = "Ying Zhao and George Karypis",
     title = "Evaluation of Hierarchical Clustering Algorithms for Document Datasets",
     journal = "11th Conference of Information and Knowledge Management (CIKM)"
+  ),
+
+  linnenbrink2023 = bibentry("article",
+    author = "Linnenbrink, J. and Mila, C. and Ludwig, M. and Meyer, H.",
+    title = "kNNDM: k-fold Nearest Neighbour Distance Matching Cross-Validation for map accuracy estimation",
+    journal = "EGUsphere",
+    volume = "2023",
+    year = "2023",
+    pages = "1--16",
+    url = "https://egusphere.copernicus.org/preprints/2023/egusphere-2023-1308/",
+    doi = "10.5194/egusphere-2023-1308"
   )
 ) # nolint end
 # nocov end

### 2023-08-23 man-roxygen/rox_sptcv_knndm.R

@@ -1,0 +1,8 @@
+#' @inherit CAST::knndm description details
+#'
+#' @section Parameters:
+#' * `folds` (`integer(1)`)\cr
+#'   Number of folds.
+#' * `stratify`\cr
+#'   If `TRUE`, stratify on the target column.
+#'

### 2023-08-23 tests/testthat/test-ResamplingSpCVKnndm.R

@@ -1,0 +1,48 @@
+test_that("mlr3spatiotempcv indices are the same as CAST knndm: modeldomain", {
+  skip_if_not_installed("CAST")
+
+  task = tsk("ecuador")
+  points = sf::st_as_sf(task$coordinates(), crs = task$crs, coords = c("x", "y"))
+  modeldomain = sf::st_as_sfc(sf::st_bbox(points))
+
+  set.seed(42)
+  rsmp = rsmp("spcv_knndm", modeldomain = modeldomain)
+  suppressMessages(suppressWarnings(rsmp$instantiate(task)))
+
+  set.seed(42)
+  testknndm = suppressMessages(suppressWarnings(CAST::knndm(
+    tpoints = points,
+    modeldomain = modeldomain
+  )))
+
+  expect_equal(rsmp$instance$train, testknndm$indx_train)
+  expect_equal(rsmp$instance$test, testknndm$indx_test)
+})
+
+test_that("mlr3spatiotempcv indices are the same as CAST knndm: ppoints", {
+  skip_if_not_installed("CAST")
+
+  simarea = list(matrix(c(0, 0, 0, 100, 100, 100, 100, 0, 0, 0),
+    ncol = 2, byrow = TRUE))
+  simarea = sf::st_polygon(simarea)
+  train_points = sf::st_sample(simarea, 1000, type = "random")
+  train_points = sf::st_as_sf(train_points)
+  train_points$target = as.factor(sample(c("TRUE", "FALSE"), 1000, replace = TRUE))
+  pred_points = sf::st_sample(simarea, 1000, type = "regular")
+
+  task = mlr3spatial::as_task_classif_st(sf::st_as_sf(train_points),
+    "target", positive = "TRUE")
+
+  set.seed(42)
+  rsmp = rsmp("spcv_knndm", ppoints = pred_points)
+  suppressMessages(suppressWarnings(rsmp$instantiate(task)))
+
+  set.seed(42)
+  testknndm = suppressMessages(suppressWarnings(CAST::knndm(
+    tpoints = train_points,
+    ppoints = pred_points
+  )))
+
+  expect_equal(rsmp$instance$train, testknndm$indx_train)
+  expect_equal(rsmp$instance$test, testknndm$indx_test)
+})

### 2023-10-06 R/zzz.R
update mlr3book links

@@ -8,7 +8,7 @@
 #' @importFrom utils globalVariables packageVersion
 #' @section Main resources:
 #' * Book on mlr3: \url{https://mlr3book.mlr-org.com}
-#' * mlr3book section about spatiotemporal data: \url{https://mlr3book.mlr-org.com/special.html#spatiotemporal}
+#' * mlr3book section about spatiotemporal data: \url{https://mlr3book.mlr-org.com/chapters/chapter13/beyond_regression_and_classification.html#spatiotemp-cv}
 #' * package vignettes: \url{https://mlr3spatiotempcv.mlr-org.com/dev/articles/}
 #'
 #' ## Miscellaneous \pkg{mlr3} content

### 2023-12-24 paper/article.Rmd
align figure sizes (points, axes, width)

@@
 
-```{r coords-eval, fig.cap='Leave-one-block-out CV based on $k$-means clustering of the coordinates as implemented in package \\pkg{sperrorest} (method `"spcv\\_coords"` in \\pkg{mlr3spatiotempcv}).', out.width="40%", fig.pos="ht", echo=FALSE}
+```{r knndm-eval, fig.cap='Leave-one-block-out CV based on $k$-means clustering of the coordinates as implemented in package \\pkg{sperrorest} (method `"spcv\\_coords"` in \\pkg{mlr3spatiotempcv}).', out.width="55%", fig.pos="ht", echo=FALSE}
 rsmp_coords = rsmp("spcv_coords", folds = 5)
@@
 ```{r vis-nspcv-non-eval, eval=FALSE}
-autoplot(rsmp_nsp, task, fold_id = 1:2, size = 0.8)
-```
-
-```{r vis-nspcv-eval, fig.cap="Random (non-spatial) four-fold CV partitioning. Only the first two folds of the first repetition are shown.", fig.pos="ht", fig.height=6, fig.width=8, echo=FALSE}
-autoplot(rsmp_nsp, task, fold_id = 1:2, size = 0.8) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+autoplot(rsmp_nsp, task, fold_id = 1:2, size = 0.3)
+```
+
+```{r vis-nspcv-eval, fig.cap="Random (non-spatial) four-fold CV partitioning. Only the first two folds of the first repetition are shown.", out.width="100%", fig.pos="ht", echo=FALSE}
+autoplot(rsmp_nsp, task, fold_id = 1:2, size = 0.3) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) *
+  ggplot2::theme(plot.title = ggtext::element_textbox(size = 8),
+    axis.text = ggplot2::element_text(size = 8)) *
+  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 0.8)))
 ```
@@
 ```{r vis-spcv-non-eval, eval=FALSE}
-autoplot(rsmp_sp, task, fold_id = 1:2, size = 0.8)
+autoplot(rsmp_sp, task, fold_id = 1:2, size = 0.3)
 ```
@@
 ```{r vis-spcv-eval, fig.cap="Spatial leave-one-block-out partitioning using coordinate-based clustering to create roughly equally sized polygonal blocks.
-Due to space limitations only the first two folds of the first repetition are shown.", fig.pos="ht", fig.height=6, fig.width=8, echo=FALSE}
-autoplot(rsmp_sp, task, fold_id = 1:2, size = 0.8) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+Only the first two folds of the first repetition are shown.", fig.pos="ht", out.width="100%", echo=FALSE}
+autoplot(rsmp_sp, task, fold_id = 1:2, size = 0.3) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) *
+  ggplot2::theme(plot.title = ggtext::element_textbox(size = 8),
+    axis.text = ggplot2::element_text(size = 8)) *
+  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 0.8)))
 ```

### 2023-12-25 paper/article.Rmd
harmonize figures

--- 2023-08-15 paper/article.Rmd
@@
 In this specific situation it would therefore seem rather unwise to train a model on a simple random subsample of grid cells, and to test it on the remaining data, using other grid cells from the same fields, as if one wanted to predict within a field.
-The results from this performance assessment would be over-optimistic, and will not reflect the model's true ability to make predictions for new fields.
+The results from this performance assessment would be over-optimistic, and perhaps badly so.
 To mimic the predictive situation for which the model is trained, one would rather have to resample at the level of fields, not grid cells [@pena2015].
@@
 One limitation of all these packages is the sole focus on model assessment, while the proposed implementation within the \pkg{mlr3} framework also offers seamless integration into model selection and provides parallel execution and enhanced logging abilities.
-During the intitial development of \pkg{mlr3spatiotempcv}, it has been the first toolbox aiming to integrate standalone spatiotemporal validation strategies into an existing machine-learning framework.
-Meanwhile the \pkg{spatialsample} has undertaken similar efforts for the \pkg{tidymodels} framework [@mahoney2023].
-In addition parallel efforts to support spatial validation in the Python world have been made by the \pkg{spacv} and \pkg{spatial-kfold} packages, yet often with only a subset of available methods compared to what \pkg{mlr3spatiotempcv} offers [@ghariani2023].
+It is worth noting that a spatial cross-validation library named \pkg{spacv} has recently been developed for Python3, which can be used with the \pkg{scikit-learn} machine-learning framework [@pedregosa2011].
 
@@
 
+% this is required to not expand the authors in the table citations on first use
 \shortcites{ploton2020, karasiak2021, moller2021, endicott2017, wu2020, geiss2017, morera2021, zurell2020, brenning2015, bebber2017, jensen2021, escobar2021, stewart2021, reitz2021, egli2020, gao2019}
@@
       \midrule
-      \multirow{4}{*}{\specialcell{Leave-one-\\block-out \\ CV}} &  clustering of coordinates & \texttt{"spcv\_coords"}* & \pkg{sperrorest} (6)      & \specialcell{\cite{morera2021} \\ \cite{geiss2017} \\ \cite{wu2020}} \\
+      \multirow{4}{*}{\specialcell{Leave-one-\\block-out \\ CV}} &  clustering of coordinates & \texttt{"spcv\_coords"} & \pkg{sperrorest} (6)      & \specialcell{\cite{morera2021} \\ \cite{geiss2017} \\ \cite{wu2020}} \\
             \cmidrule{2-5}
@@
   \cmidrule{2-5}
-                                      &  clustering in feature space & \texttt{"spcv\_env"}* & \pkg{blockCV} (1)   &  \cite{morera2021}  \\
+                                      &  clustering in feature space & \texttt{"spcv\_env"} & \pkg{blockCV} (1)   &  \cite{morera2021}  \\
       \midrule
-      \multirow{2}{*}{\specialcell{Spatiotemp. \\ CV}} &  custom & \texttt{"sptcv\_cstf"}* & \pkg{CAST} (6)      & \specialcell{\cite{gao2019} \\ \cite{reitz2021} \\ \cite{egli2020}} \\
+      \multirow{3}{*}{\specialcell{Spatiotemp. \\ CV}} &  custom & \texttt{"sptcv\_cstf"} & \pkg{CAST} (6)      & \specialcell{\cite{gao2019} \\ \cite{reitz2021} \\ \cite{egli2020}} \\
+                                                  \cmidrule{2-5}
+                                      &  clustering: custom & \texttt{"sptcv\_cluto"}  & \pkg{skmeans} (0)    & - \\
                                       \bottomrule
@@
     The count in brackets after the package name represents the number of studies that were found having used this resampling technique until May 2021.
-    For each method, up to three randomly selected references were added to the table.
-    Methods suffixed with a * in the "Name" column have been reimplemented in \pkg{mlr3spatiotempcv} for efficiency or to reduce upstream dependencies.}
+    For each method, up to three randomly selected references were added to the table.}
     \label{tab:sptcv-methods}
@@
 
-autoplot(rsmp_buffer, size = 0.8, task = task, fold_id = 1, show_omitted = TRUE)
+autoplot(rsmp_buffer, size = 0.8, task = task, fold_id = 1)
 ```
@@
 ```{r buffer-eval, fig.cap='Visualization of the spatial buffering method from package \\pkg{blockCV} (method "spcv\\_buffer" in \\pkg{mlr3spatiotempcv}).
-The buffer distance is 1000 m.', out.width="55%", fig.pos="ht",echo=FALSE}
+The buffer distance is 1000 m.', out.width="60%", fig.pos="ht",echo=FALSE}
 library("mlr3")
@@
 autoplot(rsmp_buffer,
-  size = 0.8, task = task, fold_id = 1, show_omitted = TRUE) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
-  ggplot2::theme(text = ggplot2::element_text(size = 16),
-    plot.title = ggtext::element_textbox(size = 13,
-      r = ggplot2::unit(7, "pt"),
-      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8)) +
-  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 1.5)))
+  size = 0.8, task = task, fold_id = 1) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
 ```
@@
 
-autoplot(rsmp_disc, size = 0.8, task = task, fold_id = 1, show_omitted = TRUE)
+autoplot(rsmp_disc, size = 0.8, task = task, fold_id = 1)
 ```
@@
 ```{r disc-eval, fig.cap='Visualization of one training set / test set combination generated with the leave-one-disc-out method from package \\pkg{sperrorest} (method `"spcv\\_disc"` in \\pkg{mlr3spatiotempcv}).
-The disc has a radius of 300 m and is surrounded by a 400-m buffer.', out.width="55%", echo=FALSE,fig.pos="ht"}
+The disc has a radius of 300 m and is surrounded by a 400-m buffer.', out.width="60%", echo=FALSE,fig.pos="ht"}
 library("mlr3")
@@
 autoplot(rsmp_disc,
-  size = 0.8, task = task, fold_id = 1, show_omitted = TRUE) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
-  ggplot2::theme(text = ggplot2::element_text(size = 16),
-    plot.title = ggtext::element_textbox(size = 13,
-      r = ggplot2::unit(7, "pt"),
-      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8)) +
-  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 1.5)))
+  size = 0.8, task = task, fold_id = 1) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
 ```
@@
 
-```{r coords-eval, fig.cap='Leave-one-block-out CV based on $k$-means clustering of the coordinates as implemented in package \\pkg{sperrorest} (method `"spcv\\_coords"` in \\pkg{mlr3spatiotempcv}).', out.width="55%", fig.pos="ht", echo=FALSE}
+```{r coords-eval, fig.cap='Leave-one-block-out CV based on $k$-means clustering of the coordinates as implemented in package \\pkg{sperrorest} (method `"spcv\\_coords"` in \\pkg{mlr3spatiotempcv}).', out.width="60%", fig.pos="ht", echo=FALSE}
 rsmp_coords = rsmp("spcv_coords", folds = 5)
@@
   size = 0.8, fold_id = 1, task = task) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
-  ggplot2::theme(text = ggplot2::element_text(size = 16),
-    plot.title = ggtext::element_textbox(size = 13,
-      r = ggplot2::unit(7, "pt"),
-      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8)) +
-  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 1.5)))
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
 ```
@@
 
-```{r tile, fig.cap='Leave-one-block-out resampling from package \\pkg{sperrorest} (method `"spcv\\_tiles"` in package \\pkg{mlr3spatiotempcv} with argument `nsplit = c(3,4)` indicating the number of rows and columns).', out.width="55%", fig.pos="ht", echo=FALSE}
+```{r tile, fig.cap='Leave-one-block-out resampling from package \\pkg{sperrorest} (method `"spcv\\_tiles"` in package \\pkg{mlr3spatiotempcv} with argument `nsplit = c(3,4)` indicating the number of rows and columns).', out.width="60%", fig.pos="ht", echo=FALSE}
 rsmp_tiles = rsmp("spcv_tiles", nsplit = c(3L, 4L))
@@
 ```{r custom-cv, fig.cap='Leave-one-level-out (custom) resampling from package \\pkg{mlr3} (method `"custom\\_cv"`).
-A factor variable is used to define all partitions.', out.width="55%", fig.pos="ht", echo=FALSE}
+A factor variable is used to define all partitions.', out.width="60%", fig.pos="ht", echo=FALSE}
 breaks = quantile(task$data()$dem, seq(0, 1, length = 6))
@@
 autoplot(rsmp_custom, size = 0.8, task = task, fold_id = 1) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
-  ggplot2::theme(text = ggplot2::element_text(size = 16),
-    plot.title = ggtext::element_textbox(size = 13,
-      r = ggplot2::unit(7, "pt"),
-      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8)) +
-  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 1.5)))
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
 ```
@@
 autoplot(rsmp_block_random, size = 0.8, fold_id = 1, task = task,
-  show_blocks = TRUE, show_labels = TRUE, label_size = 4)
+  show_blocks = TRUE, show_labels = TRUE)
 ```
@@
 ```{r block-random, fig.cap='Random resampling of square spatial blocks using the implementation in package \\pkg{blockCV} (method `"spcv\\_block"` with option `selection = "random"` in \\pkg{mlr3spatiotempcv}).
-The size of the squares is 1000 m, and four out of the 19 blocks were assigned to the test partition.', out.width="55%", fig.pos="ht", echo=FALSE}
+The size of the squares is 1000 m, and four out of the 19 blocks were assigned to the test partition.', out.width="60%", fig.pos="ht", echo=FALSE}
 rsmp_block_random = rsmp("spcv_block", range = 1000, folds = 5)
@@
   size = 0.8, fold_id = 1, task = task,
-  show_blocks = TRUE, show_labels = TRUE, label_size = 4) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
-  ggplot2::theme(text = ggplot2::element_text(size = 16),
-    plot.title = ggtext::element_textbox(size = 13,
-      r = ggplot2::unit(7, "pt"),
-      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8)) +
-  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 1.5)))
+  show_blocks = TRUE, show_labels = TRUE) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
 ```
@@
 autoplot(rsmp_block_systematic, size = 0.8, fold_id = 1, task = task,
-  show_blocks = TRUE, show_labels = TRUE, label_size = 4)
+  show_blocks = TRUE, show_labels = TRUE)
 ```
@@
 ```{r block-systematic, fig.cap='Sytematic resampling of square spatial blocks using the implementation in package \\pkg{blockCV} (method `"spcv\\_block"` with option `selection = "systematic"` in \\pkg{mlr3spatiotempcv}).
-The size of the squares is 1000 m, and four out of the 19 blocks were assigned to this test sample.', out.width="55%", fig.pos="ht", echo=FALSE}
+The size of the squares is 1000 m, and four out of the 19 blocks were assigned to this test sample.', out.width="60%", fig.pos="ht", echo=FALSE}
 rsmp_block_systematic = rsmp("spcv_block",
@@
 A complete group or block of observations is therefore assigned to a specific partition, which consequently honors the grouping structure.
+In the deprecated \pkg{mlr} package this concept was referred to as "blocking".
 
 In contrast to geometric or clustering-based blocks, the spatial or temporal location is not used explicitly, but rather implicitly through the spatial or spatiotemporal footprint of each user-defined block.
-Studies which have applied custom grouping approaches include @meyer2018, @anderson2005, @kasurak2011.
 
@@
 rsmp_cv_group = rsmp("cv", folds = 3)$instantiate(task_cv)
+
+print(rsmp_cv_group$instance)
 ```
@@
 A factor variable (named "group") was used to define the grouping.
-Each class is either assigned to the test or training set.', out.width="55%", fig.pos="ht", eval=FALSE}
+Each class is either assigned to the test or training set.', out.width="60%", fig.pos="ht", eval=FALSE}
 autoplot(rsmp_cv_group, size = 0.8, task = task_cv, fold_id = 1)
@@
 A factor variable is used to define the grouping.
-Each class is either assigned to the test or training set.', out.width="55%", fig.pos="ht", echo=FALSE}
+Each class is either assigned to the test or training set.', out.width="60%", fig.pos="ht", echo=FALSE}
 autoplot(rsmp_cv_group, size = 0.8, task = task_cv, fold_id = 1) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
-  ggplot2::theme(text = ggplot2::element_text(size = 16),
-    plot.title = ggtext::element_textbox(size = 13,
-      r = ggplot2::unit(7, "pt"),
-      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8)) +
-  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 1.5)))
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
 ```
@@
   ggplot2::theme(plot.title = ggtext::element_textbox(size = 8),
-    axis.text = ggplot2::element_text(size = 8)) +
-  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 0.8)))
+    axis.text = ggplot2::element_text(size = 7),
+    legend.title = ggtext::element_textbox(size = 8),
+    legend.text = ggtext::element_textbox(size = 6),
+    legend.key.size = ggplot2::unit(0.4, "cm"))
 
@@
 
-```{r llo, echo=FALSE, out.width="70%", fig.cap='Birds-eye view of `"leave-location-out"` CV from package \\pkg{CAST} (method `"sptcv\\_cstf"` and column role `"space" = "SOURCEID"`).
-The readers viewpoint is located on top of the figure, looking down.
-The transects are individual locations being left out across multiple time steps.', fig.pos="ht"}
+```{r llo, echo=FALSE, out.width="70%", fig.cap='Birds-eye view of `"leave-location-out"` CV from package \\pkg{CAST} (method `"sptcv\\_cstf"` and column role `"space" = "SOURCEID"`).', fig.pos="ht"}
 knitr::include_graphics("pdf/llo.pdf")
@@
 
-A case study is used to demonstrate the application of spatial and non-spatial resampling techniques for model assessment in \pkg{mlr3spatiotempcv}.
+A well-known case study is used to demonstrate the application of spatial and non-spatial resampling techniques for model assessment in \pkg{mlr3spatiotempcv}.
 The objective of landslide susceptibility modeling is to predict how prone to landslide initiation a location is.
@@
 
-To define a resampling strategy, the `rsmp()` function is used to generate a resampling object using four folds and two repetitions following a random sampling logic ("repeated_cv").
+To define a resampling strategy, the `rsmp()` function is used to generate a resampling object using four folds and two repetitions following a random sampling logic ("cv").
 
@@
 
-The model assessment is now repeated again using spatial CV resampling, for which the only required change is to replace `"repeated_cv"` with `"repeated_spcv_coords"`.
+The model assessment is now repeated again using spatial CV resampling, for which the only required change is to replace `"repeated_cv"` by `"repeated_spcv_coords"`.
 
@@
 
+```{r vis-spcv-non-eval, eval=FALSE}
+autoplot(rsmp_sp, task, fold_id = 1:2, size = 0.8)
+```
+
+```{r vis-spcv-eval, fig.cap="Spatial leave-one-block-out partitioning using coordinate-based clustering to create roughly equally sized polygonal blocks.
+Due to space limitations only the first two folds of the first repetition are shown.", fig.pos="ht", out.width="100%", echo=FALSE}
+autoplot(rsmp_sp, task, fold_id = 1:2, size = 0.3) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) *
+  ggplot2::theme(plot.title = ggtext::element_textbox(size = 8),
+    axis.text = ggplot2::element_text(size = 7),
+    legend.title = ggtext::element_textbox(size = 8),
+    legend.text = ggtext::element_textbox(size = 7),
+    legend.key.size = ggplot2::unit(0.4, "cm"))
+```
+
 ```{r vis-nspcv-non-eval, eval=FALSE}
@@
 
-```{r vis-nspcv-eval, fig.cap="Random (non-spatial) four-fold CV partitioning. Only the first two folds of the first repetition are shown.", fig.pos="ht", fig.height=6, fig.width=8, echo=FALSE}
-autoplot(rsmp_nsp, task, fold_id = 1:2, size = 0.8) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
-```
-
-```{r vis-spcv-non-eval, eval=FALSE}
-autoplot(rsmp_sp, task, fold_id = 1:2, size = 0.8)
-```
-
-```{r vis-spcv-eval, fig.cap="Spatial leave-one-block-out partitioning using coordinate-based clustering to create roughly equally sized polygonal blocks.
-Due to space limitations only the first two folds of the first repetition are shown.", fig.pos="ht", fig.height=6, fig.width=8, echo=FALSE}
-autoplot(rsmp_sp, task, fold_id = 1:2, size = 0.8) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
+```{r vis-nspcv-eval, fig.cap="Random (non-spatial) four-fold CV partitioning. Only the first two folds of the first repetition are shown.", fig.pos="ht",out.width="100%", echo=FALSE}
+autoplot(rsmp_nsp, task, fold_id = 1:2, size = 0.3) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) *
+  ggplot2::theme(plot.title = ggtext::element_textbox(size = 8),
+    axis.text = ggplot2::element_text(size = 7),
+    legend.title = ggtext::element_textbox(size = 8),
+    legend.text = ggtext::element_textbox(size = 7),
+    legend.key.size = ggplot2::unit(0.4, "cm"))
 ```
@@
 It is also expected that spatial CV results better represent the model's transferability to geologically and topographically similar areas adjacent to the training area.
-Yet it must be kept in mind that using spatial CV might also lead to an overpessimistic estimate of the model's predictive performance, as large parts of the study area are being left out during training.
-This goes in hand with the limited ability to predict to other areas, as discussed in [@wadoux2021, meyer2021].
 The magnitude of the difference between spatial and non-spatial CV estimates may depend on the dataset, the strength on spatial or spatiotemporal autocorrelation, and the learner itself.
@@
 
-Although the case study example in Section \ref{sec:case-study} used the `"spcv_coords"` method for coordinate-based clustering, other methods would be suitable as well.
-In this particular application setting, we want to assess how well the model generalized from the concrete set of historical landslide occurrences, which is why we ensured that training and test sets contain different, "new" hillslopes and landslides.
+Although the case study example in Section \ref{sec:case-study} used the `"spcv_coords"` method for coordinate-based clustering, this should not give the impression that this method is the only method suitable for this example task.
+In this application setting, we want to assess how well the model generalized from the concrete set of historical landslide occurrences, which is why we ensured that training and test sets contain different, "new" hillslopes and landslides.
 Coordinate-based clustering is particularly appealing in this setting because of its ability to adapt to the irregularly shaped study area of this example.
@@
 
-In other scenarios, such as when predicting crop types across monocultural fields \ref{sec:intro} [@pena2015], it makes sense to group observations into blocks based on previously known boundaries and resample at the polygon level (`"cv"` method with grouping).
+In other application scenarios such as the crop classification example given in Section \ref{sec:intro} [@pena2015], the objective is to predict the fruit-tree crop type of 'new', unseen fields within an agricultural region.
+In this use case we are not at all interested in "predicting" the already known crop type of (other) grid cells within the same agricultural field.
+Also, a model will likely be much better able to predict the crop type within the same field from multitemporal remote-sensing data since all crops within that field are subject to identical management practices (e.g., use of pesticides, pruning of fruit trees, tree spacing), while 'new' fields may be managed differently.
+As a consequence, grid cells from the same field should be grouped into a block, and resampling should be done at the field level (`"cv"` method with grouping) to receive an honest estimate of the model's performance in a relevant predictive situation.
 If, in contrast, the objective is to apply the model to an adjacent agricultural region (e.g., adjacent county) where the same crop types are present, it may be advisable to use coordinate-based clustering (`"spcv_coords"` method) to obtain larger, contiguous test regions.
@@
 
+<!-- risks of using buffer zone -->
 For example, buffering, or the use of a spatial or temporal separation distance between training and test sets, is currently only implemented for some methods (`"spcv_buffer"`, `"spcv_disc"`, and `"sptcv_cstf"` with both `space_var` and `time_var`).
@@
 
+<!-- only some methods are suited for repeated execution -->
+<!-- spcv_coords: limited variance due to limited k-means possibilities -->
 CV is often executed repeatedly to reduce the possible influence of random variability on CV estimates.
@@
 
+<!-- sign-off -->
 Spatiotemporal cross-validation as a paradigm is not yet fully established in scientific workflows, although it has been discussed intensively for more than a decade now.

### 2023-12-25 paper/article.Rmd

--- 2023-12-24 paper/article.Rmd
@@
 A factor variable is used to define the grouping.
-Each class is either assigned to the test or training set.', out.width="55%", fig.pos="ht", echo=FALSE}
+Each class is either assigned to the test or training set.', out.width="60%", fig.pos="ht", echo=FALSE}
 autoplot(rsmp_cv_group, size = 0.8, task = task_cv, fold_id = 1) *
@@
 
+```{r vis-spcv-non-eval, eval=FALSE}
+autoplot(rsmp_sp, task, fold_id = 1:2, size = 0.3)
+```
+
+```{r vis-spcv-eval, fig.cap="Spatial leave-one-block-out partitioning using coordinate-based clustering to create roughly equally sized polygonal blocks.
+Due to space limitations only the first two folds of the first repetition are shown.", fig.pos="ht", out.width="100%", echo=FALSE}
+autoplot(rsmp_sp, task, fold_id = 1:2, size = 0.3) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) *
+  ggplot2::theme(plot.title = ggtext::element_textbox(size = 8),
+    axis.text = ggplot2::element_text(size = 7),
+    legend.title = ggtext::element_textbox(size = 8),
+    legend.text = ggtext::element_textbox(size = 7),
+    legend.key.size = ggplot2::unit(0.4, "cm"))
+```
+
 ```{r vis-nspcv-non-eval, eval=FALSE}
@@
   ggplot2::theme(plot.title = ggtext::element_textbox(size = 8),
-    axis.text = ggplot2::element_text(size = 8)) *
-  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 0.8)))
-```
-
-```{r vis-spcv-non-eval, eval=FALSE}
-autoplot(rsmp_sp, task, fold_id = 1:2, size = 0.3)
-```
-
-```{r vis-spcv-eval, fig.cap="Spatial leave-one-block-out partitioning using coordinate-based clustering to create roughly equally sized polygonal blocks.
-Only the first two folds of the first repetition are shown.", fig.pos="ht", out.width="100%", echo=FALSE}
-autoplot(rsmp_sp, task, fold_id = 1:2, size = 0.3) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) *
-  ggplot2::theme(plot.title = ggtext::element_textbox(size = 8),
-    axis.text = ggplot2::element_text(size = 8)) *
-  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 0.8)))
+    axis.text = ggplot2::element_text(size = 7),
+    legend.title = ggtext::element_textbox(size = 8),
+    legend.text = ggtext::element_textbox(size = 7),
+    legend.key.size = ggplot2::unit(0.4, "cm"))
 ```

### 2023-12-25 paper/article.Rmd
align figures

@@
 
-```{r knndm-eval, fig.cap='Leave-one-block-out CV based on $k$-means clustering of the coordinates as implemented in package \\pkg{sperrorest} (method `"spcv\\_coords"` in \\pkg{mlr3spatiotempcv}).', out.width="55%", fig.pos="ht", echo=FALSE}
+```{r knndm-eval, fig.cap='Leave-one-block-out CV based on $k$-means clustering of the coordinates as implemented in package \\pkg{sperrorest} (method `"spcv\\_coords"` in \\pkg{mlr3spatiotempcv}).', out.width="60%", fig.pos="ht", echo=FALSE}
 rsmp_coords = rsmp("spcv_coords", folds = 5)
@@
   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
-  ggplot2::theme(text = ggplot2::element_text(size = 16),
-    plot.title = ggtext::element_textbox(size = 13,
+  ggplot2::theme(text = ggplot2::element_text(size = 15),
+    plot.title = ggtext::element_textbox(size = 12,
       r = ggplot2::unit(7, "pt"),
@@
   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
-  ggplot2::theme(plot.title = ggtext::element_textbox(size = 8),
-    axis.text = ggplot2::element_text(size = 7),
-    legend.title = ggtext::element_textbox(size = 8),
-    legend.text = ggtext::element_textbox(size = 6),
-    legend.key.size = ggplot2::unit(0.4, "cm"))
+  ggplot2::theme(plot.title = ggtext::element_textbox(size = 8,
+    r = ggplot2::unit(4, "pt"),
+    height = ggplot2::unit(0.30, "inch"), linewidth = 0.5),
+  axis.text = ggplot2::element_text(size = 8.5),
+  legend.title = ggtext::element_textbox(size = 10),
+  legend.text = ggtext::element_textbox(size = 8),
+  legend.key.size = ggplot2::unit(0.4, "cm")) +
+  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 0.8)))
 
@@
   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) *
-  ggplot2::theme(plot.title = ggtext::element_textbox(size = 8),
-    axis.text = ggplot2::element_text(size = 7),
-    legend.title = ggtext::element_textbox(size = 8),
-    legend.text = ggtext::element_textbox(size = 7),
-    legend.key.size = ggplot2::unit(0.4, "cm"))
+  ggplot2::theme(plot.title = ggtext::element_textbox(size = 8,
+    r = ggplot2::unit(4, "pt"),
+    height = ggplot2::unit(0.30, "inch"), linewidth = 0.5),
+  axis.text = ggplot2::element_text(size = 8.5),
+  legend.title = ggtext::element_textbox(size = 10),
+  legend.text = ggtext::element_textbox(size = 8),
+  legend.key.size = ggplot2::unit(0.4, "cm")) *
+  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 0.8)))
 ```

### 2024-01-05 paper/article.Rmd
add knndm section and table row

--- 2023-12-25 paper/article.Rmd
@@
   cache = FALSE,
-  cache.lazy = TRUE)
+  cache.lazy = TRUE
+)
 ```
@@
       \midrule
-      \multirow{4}{*}{\specialcell{CV at the \\ block level}} &  geometric: rectangular & \texttt{"spcv\_block"}  & \pkg{blockCV} (28)      & \specialcell{\cite{jensen2021} \\ \cite{escobar2021} \\ \cite{stewart2021}} \\
+      \multirow{5}{*}{\specialcell{CV at the \\ block level}} &  geometric: rectangular & \texttt{"spcv\_block"}  & \pkg{blockCV} (28)      & \specialcell{\cite{jensen2021} \\ \cite{escobar2021} \\ \cite{stewart2021}} \\
   \cmidrule{2-5}
@@
                                       &  clustering in feature space & \texttt{"spcv\_env"}* & \pkg{blockCV} (1)   &  \cite{morera2021}  \\
+  \cmidrule{2-5}
+                                      &  clustering of coordinates + aggregation & \texttt{"spcv\_knndm"}* & \pkg{CAST} (1)   &  \cite{ludwig2023}  \\
       \midrule
@@
 The buffer distance is 1000 m.', out.width="60%", fig.pos="ht",echo=FALSE}
+set.seed(42)
 library("mlr3")
@@
 autoplot(rsmp_buffer,
-  size = 0.8, task = task, fold_id = 1, show_omitted = TRUE) *
+  size = 0.8, task = task, fold_id = 1, show_omitted = TRUE
+) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
-  ggplot2::theme(text = ggplot2::element_text(size = 16),
-    plot.title = ggtext::element_textbox(size = 13,
+  ggplot2::theme(
+    text = ggplot2::element_text(size = 16),
+    plot.title = ggtext::element_textbox(
+      size = 13,
       r = ggplot2::unit(7, "pt"),
-      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8)) +
+      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8
+    )
+  ) +
   ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 1.5)))
@@
 autoplot(rsmp_disc,
-  size = 0.8, task = task, fold_id = 1, show_omitted = TRUE) *
+  size = 0.8, task = task, fold_id = 1, show_omitted = TRUE
+) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
-  ggplot2::theme(text = ggplot2::element_text(size = 16),
-    plot.title = ggtext::element_textbox(size = 13,
+  ggplot2::theme(
+    text = ggplot2::element_text(size = 16),
+    plot.title = ggtext::element_textbox(
+      size = 13,
       r = ggplot2::unit(7, "pt"),
-      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8)) +
+      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8
+    )
+  ) +
   ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 1.5)))
@@
 autoplot(rsmp_coords,
-  size = 0.8, fold_id = 1, task = task) *
+  size = 0.8, fold_id = 1, task = task
+) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
-  ggplot2::theme(text = ggplot2::element_text(size = 16),
-    plot.title = ggtext::element_textbox(size = 13,
+  ggplot2::theme(
+    text = ggplot2::element_text(size = 16),
+    plot.title = ggtext::element_textbox(
+      size = 13,
       r = ggplot2::unit(7, "pt"),
-      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8)) +
+      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8
+    )
+  ) +
   ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 1.5)))
 ```
-
-### Clustering-based: using k-fold nearest-neighbour distance matching --- \code{"spcv_knndm"} {short-title='Clustering-based: using kNNDM --- "spcv_knndm"' }
-
-The k-fold nearest-neighboar distance matching method after @linnenbrinck2023 is derived from the NNDM method after @mila.
-By using geographical nearest neighbor distances between test and training points, the method aims to create realistic predictive situations within the CV folds that closely mimic the ones of either existing training and test points, if know, or a provided model domain in which the model is to be applied.
-This is done by minimizing the values ofthe Empirical Cumulative Distribution Function (ECDF) for the respective scenarios outlined above [@linnenbrinck2023].
-
-```{r knndm-non-eval, eval=FALSE}
-rsmp_knndm = rsmp("spcv_knndm", folds = 5)
-
-autoplot(rsmp_knndm, size = 0.8, fold_id = 1, task = task)
-```
-
-```{r knndm-eval, fig.cap='Leave-one-block-out CV based on $k$-means clustering of the coordinates as implemented in package \\pkg{sperrorest} (method `"spcv\\_coords"` in \\pkg{mlr3spatiotempcv}).', out.width="55%", fig.pos="ht", echo=FALSE}
-rsmp_coords = rsmp("spcv_coords", folds = 5)
-
-autoplot(rsmp_coords,
-  size = 0.8, fold_id = 1, task = task) *
-  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
-  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
-  ggplot2::theme(text = ggplot2::element_text(size = 16),
-    plot.title = ggtext::element_textbox(size = 13,
-      r = ggplot2::unit(7, "pt"),
-      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8)) +
-  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 1.5)))
-```
-
-FIXME:
 
@@
 autoplot(rsmp_tiles,
-  size = 0.8, fold_id = 1, task = task) *
+  size = 0.8, fold_id = 1, task = task
+) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
-  ggplot2::theme(text = ggplot2::element_text(size = 16),
-    plot.title = ggtext::element_textbox(size = 13,
+  ggplot2::theme(
+    text = ggplot2::element_text(size = 16),
+    plot.title = ggtext::element_textbox(
+      size = 13,
       r = ggplot2::unit(7, "pt"),
-      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8)) +
+      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8
+    )
+  ) +
   ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 1.5)))
@@
   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
-  ggplot2::theme(text = ggplot2::element_text(size = 16),
-    plot.title = ggtext::element_textbox(size = 13,
+  ggplot2::theme(
+    text = ggplot2::element_text(size = 16),
+    plot.title = ggtext::element_textbox(
+      size = 13,
       r = ggplot2::unit(7, "pt"),
-      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8)) +
+      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8
+    )
+  ) +
   ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 1.5)))
@@
 
-autoplot(rsmp_block_random, size = 0.8, fold_id = 1, task = task,
-  show_blocks = TRUE, show_labels = TRUE, label_size = 4)
+autoplot(rsmp_block_random,
+  size = 0.8, fold_id = 1, task = task,
+  show_blocks = TRUE, show_labels = TRUE, label_size = 4
+)
 ```
@@
   size = 0.8, fold_id = 1, task = task,
-  show_blocks = TRUE, show_labels = TRUE, label_size = 4) *
+  show_blocks = TRUE, show_labels = TRUE, label_size = 4
+) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
-  ggplot2::theme(text = ggplot2::element_text(size = 16),
-    plot.title = ggtext::element_textbox(size = 13,
+  ggplot2::theme(
+    text = ggplot2::element_text(size = 16),
+    plot.title = ggtext::element_textbox(
+      size = 13,
       r = ggplot2::unit(7, "pt"),
-      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8)) +
+      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8
+    )
+  ) +
   ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 1.5)))
@@
 
-autoplot(rsmp_block_systematic, size = 0.8, fold_id = 1, task = task,
-  show_blocks = TRUE, show_labels = TRUE, label_size = 4)
+autoplot(rsmp_block_systematic,
+  size = 0.8, fold_id = 1, task = task,
+  show_blocks = TRUE, show_labels = TRUE, label_size = 4
+)
 ```
@@
   size = 0.8, fold_id = 1, task = task,
-  show_blocks = TRUE, show_labels = TRUE, label_size = 4) *
+  show_blocks = TRUE, show_labels = TRUE, label_size = 4
+) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
-  ggplot2::theme(text = ggplot2::element_text(size = 16),
-    plot.title = ggtext::element_textbox(size = 13,
+  ggplot2::theme(
+    text = ggplot2::element_text(size = 15),
+    plot.title = ggtext::element_textbox(
+      size = 12,
       r = ggplot2::unit(7, "pt"),
-      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8)) +
+      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8
+    )
+  ) +
   ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 1.5)))
@@
 
-rsmp_env_multi = rsmp("spcv_env", features = c("distdeforest", "slope"),
-  folds = 5)
-
-plot_env_single = autoplot(rsmp_env, size = 0.3,
-  fold_id = 1, task = task) +
-
-  plot_env_multi = autoplot(rsmp_env_multi, size = 0.3,
-  fold_id = 1, task = task)
+rsmp_env_multi = rsmp("spcv_env",
+  features = c("distdeforest", "slope"),
+  folds = 5
+)
+
+plot_env_single = autoplot(rsmp_env,
+  size = 0.3,
+  fold_id = 1, task = task
+) +
+
+  plot_env_multi = autoplot(rsmp_env_multi,
+  size = 0.3,
+  fold_id = 1, task = task
+)
 
@@
 rsmp_env = rsmp("spcv_env",
-  features = "distdeforest", folds = 5)
+  features = "distdeforest", folds = 5
+)
 
 rsmp_env_multi = rsmp("spcv_env",
-  features = c("distdeforest", "slope"), folds = 5)
+  features = c("distdeforest", "slope"), folds = 5
+)
 
 plot_env_single = autoplot(rsmp_env,
-  size = 0.3, fold_id = 1, task = task) *
+  size = 0.3, fold_id = 1, task = task
+) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
-  ggplot2::theme(plot.title = ggtext::element_textbox(size = 8),
-    axis.text = ggplot2::element_text(size = 7),
-    legend.title = ggtext::element_textbox(size = 8),
-    legend.text = ggtext::element_textbox(size = 6),
-    legend.key.size = ggplot2::unit(0.4, "cm"))
+  ggplot2::theme(
+    plot.title = ggtext::element_textbox(
+      size = 8,
+      r = ggplot2::unit(4, "pt"),
+      height = ggplot2::unit(0.30, "inch"), linewidth = 0.5
+    ),
+    axis.text = ggplot2::element_text(size = 8.5),
+    legend.title = ggtext::element_textbox(size = 10),
+    legend.text = ggtext::element_textbox(size = 8),
+    legend.key.size = ggplot2::unit(0.4, "cm")
+  ) +
+  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 0.8)))
 
 plot_env_multi = autoplot(rsmp_env_multi,
-  size = 0.3, fold_id = 1, task = task) *
+  size = 0.3, fold_id = 1, task = task
+) *
   ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
-  ggplot2::theme(plot.title = ggtext::element_textbox(size = 8),
-    axis.text = ggplot2::element_text(size = 7),
-    legend.title = ggtext::element_textbox(size = 8),
-    legend.text = ggtext::element_textbox(size = 6),
-    legend.key.size = ggplot2::unit(0.4, "cm"))
+  ggplot2::theme(
+    plot.title = ggtext::element_textbox(
+      size = 8,
+      r = ggplot2::unit(4, "pt"),
+      height = ggplot2::unit(0.30, "inch"), linewidth = 0.5
+    ),
+    axis.text = ggplot2::element_text(size = 8.5),
+    legend.title = ggtext::element_textbox(size = 10),
+    legend.text = ggtext::element_textbox(size = 8),
+    legend.key.size = ggplot2::unit(0.4, "cm")
+  ) +
+  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 0.8)))
 
@@
 plot_env_single + plot_env_multi + plot_layout(guides = "collect")
+```
+
+### Nearest Neighbour Distance Matching --- \code{"spcv_knndm"} {short-title='Nearest Neighbour Distance Matching --- "spcv_knndm"' }
+
+The `spcv_knndm` method from the \pkg{CAST} package uses a two-stage approach: first groups of observations are created using Nearest Neighbour clustering.
+These are then merged into k clusters with the aim to minimize the differences between the empirical Nearest Neighbour distribution function of training and prediction points and training and test data during CV using the Wasserstein W statistic [@linnenbrink2023].
+
+The main goal is to ensure similar conditions during CV compared to the actual prediction scenario of the trained model.
+The k-fold variant is an extension of the LOOCV variant proposed initially by @mila2022 and more suited for large datasets.
+
+The following example uses the default settings of the implementation in the \pkg{CAST} package.
+The method requires a prediction area as an `sf` polygon or point object.
+In this case the former was used and passed as an argument to the `"spcv_knndm"` method.
+
+```{r knndm-non-eval, eval=FALSE}
+points = sf::st_as_sf(task$coordinates(), crs = task$crs, coords = c("x", "y"))
+modeldomain = sf::st_as_sfc(sf::st_bbox(points))
+
+rsmp_knndm = rsmp("spcv_knndm", modeldomain = modeldomain, folds = 5)
+
+autoplot(rsmp_knndm, size = 0.8, fold_id = 1, task = task, label_size = 4)
+```
+
+```{r knndm, fig.cap='Random resampling of square spatial blocks using the implementation in package \\pkg{CAST} (method `"spcv\\_knndm"`).
+Hierarchical clustering has been used together with the "ward.D2" link function and a target of five folds.', out.width="60%", fig.pos="ht", echo=FALSE}
+points = sf::st_as_sf(task$coordinates(), crs = task$crs, coords = c("x", "y"))
+modeldomain = sf::st_as_sfc(sf::st_bbox(points))
+
+rsmp_knndm = rsmp("spcv_knndm", modeldomain = modeldomain, folds = 5)
+
+suppressMessages(autoplot(rsmp_knndm,
+  size = 0.8, fold_id = 1, task = task, label_size = 4
+) *
+  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.02)) *
+  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) +
+  ggplot2::theme(
+    text = ggplot2::element_text(size = 16),
+    plot.title = ggtext::element_textbox(
+      size = 13,
+      r = ggplot2::unit(7, "pt"),
+      height = ggplot2::unit(0.46, "inch"), linewidth = 0.8
+    )
+  ) +
+  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 1.5))))
 ```
@@
   "2020-01-01", "2020-02-01", "2020-03-01", "2020-04-01",
-  "2020-05-01"), times = 1, each = 35768))
+  "2020-05-01"
+), times = 1, each = 35768))
 task_spt = as_task_regr_st(data,
@@
   coordinate_names = c("x", "y"), coords_as_features = FALSE,
-  crs = 26911)
+  crs = 26911
+)
 task_spt$set_col_roles("Date", roles = "time")
@@
   showlegend = FALSE, title = "",
-  margin = list(l = 0, b = 0, r = 0, t = 0))
+  margin = list(l = 0, b = 0, r = 0, t = 0)
+)
 
 plotly::save_image(p_lto_print, "pdf/lto.pdf",
-  scale = 2, width = 1000, height = 800)
+  scale = 2, width = 1000, height = 800
+)
 ```
@@
   plot3D = TRUE, plot_time_var = "Date",
-  sample_fold_n = 3000L)
+  sample_fold_n = 3000L
+)
 
@@
     showlegend = FALSE, title = "", polar = TRUE,
-    margin = list(l = 0, b = 0, r = 0, t = 0))
+    margin = list(l = 0, b = 0, r = 0, t = 0)
+  )
 
 plotly::save_image(p_llo_print, "pdf/llo.pdf",
-  scale = 2, width = 1000, height = 800)
+  scale = 2, width = 1000, height = 800
+)
 ```
@@
 
-p_lto = autoplot(rsmp_cstf_time_loc, point_size = 6,
+p_lto = autoplot(rsmp_cstf_time_loc,
+  point_size = 6,
   axis_label_fontsize = 15,
   fold_id = 4, task = task_spt, plot3D = TRUE,
-  show_omitted = TRUE, sample_fold_n = 3000L)
+  show_omitted = TRUE, sample_fold_n = 3000L
+)
 
@@
   showlegend = FALSE, title = "",
-  margin = list(l = 0, b = 0, r = 0, t = 0))
+  margin = list(l = 0, b = 0, r = 0, t = 0)
+)
 
 plotly::save_image(p_lto_print, "pdf/llto.pdf",
-  scale = 2, width = 1000, height = 800)
+  scale = 2, width = 1000, height = 800
+)
 ```
@@
 data("ecuador", package = "mlr3spatiotempcv")
-task = as_task_classif_st(ecuador, target = "slides", positive = "TRUE",
+task = as_task_classif_st(ecuador,
+  target = "slides", positive = "TRUE",
   coordinate_names = c("x", "y"), coords_as_features = FALSE,
-  crs = "EPSG:32717")
+  crs = "EPSG:32717"
+)
 ```
@@
   ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.02)) *
-  ggplot2::theme(plot.title = ggtext::element_textbox(size = 8),
-    axis.text = ggplot2::element_text(size = 7),
-    legend.title = ggtext::element_textbox(size = 8),
-    legend.text = ggtext::element_textbox(size = 7),
-    legend.key.size = ggplot2::unit(0.4, "cm"))
+  ggplot2::theme(
+    plot.title = ggtext::element_textbox(
+      size = 8,
+      r = ggplot2::unit(4, "pt"),
+      height = ggplot2::unit(0.30, "inch"), linewidth = 0.5
+    ),
+    axis.text = ggplot2::element_text(size = 8.5),
+    legend.title = ggtext::element_textbox(size = 10),
+    legend.text = ggtext::element_textbox(size = 8),
+    legend.key.size = ggplot2::unit(0.4, "cm")
+  ) *
+  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 0.8)))
 ```
@@
 For example, buffering, or the use of a spatial or temporal separation distance between training and test sets, is currently only implemented for some methods (`"spcv_buffer"`, `"spcv_disc"`, and `"sptcv_cstf"` with both `space_var` and `time_var`).
-Its use should, however, be limited to use cases involving a prediction distance, as a buffer zone reduces the size of the training sample and introduces the risk of geographically biased training data.
+Its use should, however, be limited to use cases involving a prediction distance, as a buffer zone reduces the size of the training sample and introduces the risk of geographically biased training data [@meyer2021, @zhu2015].
 

### 2024-01-24 paper/article.Rmd
revisions to manuscript and references

@@
 \pkg{mlr3spatiotempcv} is such a toolbox.
-Yet all the included methods should only be used if no probability sample can be drawn from the data - which is often the case when working with spatial data.
-A design-based approach, if possible, is the preferred way to assess map accuracies [@wadoux2021].
+We would like to note that design-based approaches are also a viable approach in situations in which it is possible to obtain additional independently sampled test data as a probability sample [@wadoux2021].
 
@@
 
-Since re-using the learning sample $\mathcal{L}$ for testing, i.e., $\mathcal{T}:=\mathcal{L}$, would yield the over-optimistic *resubstitution* or *apparent* performance, CV partitions the sample $\mathcal{L}$ into disjoint training and test sets.
+Since re-using the learning sample $\mathcal{L}$ for testing, i.e., $\mathcal{T}:=\mathcal{L}$, would over-optimistically estimate the model's predictive performance on new instances (so-called *resubstitution* or *apparent* model performance; @hand1997), CV partitions the sample $\mathcal{L}$ into disjoint training and test sets.
 Specifically, $\mathcal{L}$ is split into $k$ partitions,
@@
 One approach that has been proposed for defining a separation distance is to use the range of autocorrelation of model residuals to determine the buffer distance, as this seeks to establish independence conditional on the predictors [@brenning2005; @roberts2017].
-
-It should be noted that $k$-fold CV with a large value of $k$, and LOO CV in particular ($k=n$), is not only very time-consuming since the model has to be trained $k$ times; these models will also be nearly identical since only a tiny fraction of the data is withheld, and therefore estimation bias increases.
-'Pure' LOO CV is therefore not recommended for machine-learning model assessment.
+Recently, spatial LOO CV has been extended to generate spatial *prediction error profiles*, which relate prediction error to prediction distance [@brenning2023].
+
+It should be noted that $k$-fold CV with a large $k$, and LOO CV in particular ($k=n$) is computationally expensive due to the large number of models being fitted.
+There is ongoing debate regarding the bias--variance trade-off of these methods in model selection and assessment [@kohavi1995; @arlot2010; @zhang2015].
 
@@
 Leave-one-disc-out resampling modifies this approach to leave out circular regions centered at observation points.
-This method group naturally discards observations when applying buffers, meaning that these are not used for training or testing.
-Depending on the relative amount of discarded observations with respect to the overall number of observations, this might result in a substantial reduction of the sample size and thus decrease the applicability of the results.
+This group of methods discards observations when applying buffers, meaning that these are not used for training or testing.
+Depending on the fraction of data being discarded, this might result in a substantial reduction of the sample size and consequently pessimistically biased performance estimates.
 
@@
 Sample locations are selected randomly when `folds` is smaller than $n$; it is optionally possible to sample with replacement (`replace = TRUE`).
-Leave-one-disc-out resampling becomes spatial LOO CV for a radius of 0 m and when each observation is at a unique location.
+Leave-one-disc-out resampling becomes LOO CV and when each observation is at a unique location.
 
@@
 Since test points in non-spatial CV may be from the same slopes or even the same landslides as the training data, the non-spatial CV result should/can be considered as an over-optimistic estimate of the model's ability to predict the susceptibility to "new" landslides.
-Spatial CV, in contrast, provides a better/more accurate measure of a model's ability to generalize from the training sample --- in this case study, from the specific hillslopes and historical landslides in the training sample.
+Spatial CV, in contrast, provides a bias-reduced measure of a model's ability to generalize from the training sample --- in this case study, from the specific hillslopes and historical landslides in the training sample.
 It is also expected that spatial CV results better represent the model's transferability to geologically and topographically similar areas adjacent to the training area.
-Yet it must be kept in mind that using spatial CV might also lead to an overpessimistic estimate of the model's predictive performance, as large parts of the study area are being left out during training.
+Yet, it must be kept in mind that using spatial CV might also lead to an pessimistic estimate of the model's predictive performance if large parts of the study area are being left out during training, depending on the number of CV folds being used.
 This goes in hand with the limited ability to predict to other areas, as discussed in [@wadoux2021, meyer2021].

### 2024-01-30 paper/article.Rmd
adaptions after alex review

@@
   Spatial and spatiotemporal machine-learning models require a suitable framework for their model assessment, model selection, and hyperparameter tuning, in order to avoid error estimation bias and over-fitting.
-  This contribution reviews the state-of-the-art in spatial and spatiotemporal cross-validation, and introduces the \proglang{R} package \pkg{mlr3spatiotempcv} as an extension package of the machine-learning framework \pkg{mlr3}.
+  This contribution provides an overview of the state-of-the-art in spatial and spatiotemporal cross-validation techniques and their implementations in R while introducing the \proglang{R} package \pkg{mlr3spatiotempcv} as an extension package of the machine-learning framework \pkg{mlr3}.
   Currently various \proglang{R} packages implementing different spatiotemporal partitioning strategies exist: \pkg{blockCV}, \pkg{CAST}, \pkg{skmeans} and \pkg{sperrorest}.
@@
 One limitation of all the packages which provide only a subset of the machine learning workflow, e.g., resampling methods, is the sole focus on model assessment.
-Through the integration into the \pkg{mlr3} framework, \pkg{mlr3spatiotempcv} indirectly offers model selection capabilities, i.e., applying and evaluating various algorithms across different preprocessing setups while being able to make use of parallel execution and enhanced logging abilities.
+Through the integration into the \pkg{mlr3} framework, \pkg{mlr3spatiotempcv} indirectly offers model selection capabilities, i.e., applying and evaluating various algorithms across different preprocessing setups (e.g., all the preprocessing ‘pipeops’ of \pkg{mlr3pipelines}) while being able to make use of parallel execution and enhanced logging abilities.
 During the intitial development of \pkg{mlr3spatiotempcv}, it has been the first toolbox aiming to integrate standalone spatiotemporal validation strategies into an existing machine-learning framework.
@@
 
+Compared to other existing machine-learning frameworks in R (e.g., \pkg{caret} [@kuhn2020] or \pkg{tidymodels} [@tidymodels]), \pkg{mlr3} is the only one that provides a dedicated object-oriented framework for spatial and spatiotemporal resampling methods.
+In addition \pkg{mlr3} uses efficient core dependencies from the \pkg{data.table} [@datatable] and \pkg{R6} [@r6] packages, which are particularly well suited for large datasets.
+Through this object-oriented approach, \pkg{mlr3} uses substantially less memory than other frameworks due to the use of pointer references and the avoidance of deep object copies whenever possible [@mlr3book].
+
 Within the \pkg{mlr3} ecosystem, partitioning strategies are represented by their own objects of class \texttt{Resampling}, most of which are available within \pkg{mlr3} itself (e.g., random CV); other specialized strategies are defined in extension packages such as \pkg{mlr3spatiotempcv}.
@@
 Spatial and spatiotemporal partitioning techniques in \pkg{mlr3spatiotempcv} are currently mostly imported and interfaced from other packages, in particular \pkg{sperrorest}, \pkg{blockCV} and \pkg{CAST} [@brenning2012; @blockCV; @cast], in order to expose them to \pkg{mlr3} functionality.
-To reduce dependencies, some methods were re-implemented instead of importing them from the respective upstream packages.
+By closely following previously proposed methods and existing implementations, we allow users of these established and tested approaches to transition into \pkg{mlr3} without having to adjust their resampling procedures.
+To reduce unnecessary upstream dependencies, some methods were re-implemented instead of importing them from the respective upstream packages.
 
@@
 
-While \pkg{mlr3spatiotempcv} solely focuses on spatiotemporal resampling methods and their visualization, other packages such as \pkg{mlr3spatial} or \pkg{mlr3temporal} are planned in the \pkg{mlr3} ecosystem to provide dedicated spatiotemporal learner and prediction methods.
+While \pkg{mlr3spatiotempcv} solely focuses on spatiotemporal resampling methods and their visualization, other packages such as \pkg{mlr3spatial} or \pkg{mlr3temporal} are planned in the \pkg{mlr3} ecosystem to provide dedicated spatiotemporal learner and prediction methods (see Figure \ref{fig:mlr3-ecosystem}).
+
+```{r mlr3-ecosystem, echo=FALSE, fig.cap="Overview of the mlr3 ecosystem.
+The packages with gray dashed lines are still in development, all others have a stable interface.
+Source: @mlr3book", out.width="100%",fig.pos="ht"}
+knitr::include_graphics(here::here("paper/png/mlr3-ecosystem.png"))
+```
 
@@
 A spatial block partitioning can further be turned into a 'leave-one-block-out resampling' or a 'k-fold CV resampling at the block level'.
-The use of a buffer is theoretically possible in any scenario but in practice only offered by specific method implementations.", out.width="100%",fig.pos="ht"}
+The use of a buffer is theoretically possible in any scenario but in practice only offered by specific method implementations.
+In \\pkg{mlr3}, statistics are then calculated per fold and aggregated using the mean by default.", out.width="100%",fig.pos="ht"}
 knitr::include_graphics(here::here("paper/pdf/rsmp-schema.pdf"))
@@
 Yet, it must be kept in mind that using spatial CV might also lead to an pessimistic estimate of the model's predictive performance if large parts of the study area are being left out during training, depending on the number of CV folds being used.
-This goes in hand with the limited ability to predict to other areas, as discussed in [@wadoux2021, meyer2021].
 The magnitude of the difference between spatial and non-spatial CV estimates may depend on the dataset, the strength on spatial or spatiotemporal autocorrelation, and the learner itself.
@@
 
-## Resampling for hyperparameter tuning
-
-CV is also widely used to assess model performance when tuning the hyperparameters of flexible machine-learning models, and this is also supported by the \pkg{mlr3} framework.
+## Resampling in model optimization
+
+CV is also widely used to assess model performance when tuning hyperparameters of flexible machine-learning models or performing feature selection [@cai2018, @bischl2021] .
+The \pkg{mlr3} framework supports the use of CV for both approaches.
 Using the CV methods introduced here, \pkg{mlr3} can therefore be used to optimized models to show an improved performance in specific spatial or spatiotemporal predictive setting [@schratz2019].
 Such an optimization may, for example, result in a reduced maximum tree depth or increased minimum node size in the Ecuador case study, since these hyperparameter settings would result in a stronger generalization and reduced overfitting.
-
-We recommend using nested CV for this purpose.
-In nested CV, an "inner" CV is performed on each CV training set, since hyperparameter tuning is an integral part of model fitting that should not be able to use information from the CV test set.
+Usually both hyperparameter and feature selection (wrapper feature selection or filters) are combined within a single, nested optimization process [@schratz2021].
+
+In nested CV specifically, an "inner" CV is performed on each CV training set, since hyperparameter tuning is an integral part of model fitting that should not be able to use information from the outer CV test set.
 In such scenarios it is recommended to use the same spatial resampling method for the inner CV (hyperparameter tuning) as for the outer CV (model assessment) in order to use the appropriate objective function for optimization.

### 2024-01-30 paper/article.Rmd
fix missing refs

@@
 
-Compared to other existing machine-learning frameworks in R (e.g., \pkg{caret} [@kuhn2020] or \pkg{tidymodels} [@tidymodels]), \pkg{mlr3} is the only one that provides a dedicated object-oriented framework for spatial and spatiotemporal resampling methods.
+Compared to other existing machine-learning frameworks in R (e.g., \pkg{caret} [@kuhn2020] or \pkg{tidymodels} [@kuhn2020]), \pkg{mlr3} is the only one that provides a dedicated object-oriented framework for spatial and spatiotemporal resampling methods.
 In addition \pkg{mlr3} uses efficient core dependencies from the \pkg{data.table} [@datatable] and \pkg{R6} [@r6] packages, which are particularly well suited for large datasets.

### 2024-01-30 paper/article.Rmd
fix ludwig et al reference

@@
 \nocite{ploton2020}
+\nocite{ludwig2023}
 
@@
       \midrule
-      \multirow{5}{*}{\specialcell{CV at the \\ block level}} &  geometric: rectangular & \texttt{"spcv\_block"}  & \pkg{blockCV} (28)      & \specialcell{\cite{jensen2021} \\ \cite{escobar2021} \\ \cite{stewart2021}} \\
+      \multirow{6}{*}{\specialcell{CV at the \\ block level}} &  geometric: rectangular & \texttt{"spcv\_block"}  & \pkg{blockCV} (28)      & \specialcell{\cite{jensen2021} \\ \cite{escobar2021} \\ \cite{stewart2021}} \\
   \cmidrule{2-5}
@@
   \cmidrule{2-5}
-                                      &  clustering of coordinates + aggregation & \texttt{"spcv\_knndm"}* & \pkg{CAST} (1)   &  \cite{ludwig2023}  \\
+                                      & {\specialcell{clustering of coordinates + \\ aggregation}} & \texttt{"spcv\_knndm"}* & \pkg{CAST} (1)   &  Ludwig \textit{et al.} (2023)  \\
       \midrule

### 2024-01-31 paper/article.Rmd
minor edits

@@
 One limitation of all the packages which provide only a subset of the machine learning workflow, e.g., resampling methods, is the sole focus on model assessment.
-Through the integration into the \pkg{mlr3} framework, \pkg{mlr3spatiotempcv} indirectly offers model selection capabilities, i.e., applying and evaluating various algorithms across different preprocessing setups (e.g., all the preprocessing ‘pipeops’ of \pkg{mlr3pipelines}) while being able to make use of parallel execution and enhanced logging abilities.
-During the intitial development of \pkg{mlr3spatiotempcv}, it has been the first toolbox aiming to integrate standalone spatiotemporal validation strategies into an existing machine-learning framework.
+Through the integration into the \pkg{mlr3} framework, \pkg{mlr3spatiotempcv} indirectly offers model selection capabilities, i.e., applying and evaluating various algorithms across different preprocessing setups (e.g., all the preprocessing ‘pipelining operators’ of \pkg{mlr3pipelines}) while being able to make use of parallel execution and enhanced logging abilities.
+During the initial development of \pkg{mlr3spatiotempcv}, it has been the first toolbox aiming to integrate standalone spatiotemporal validation strategies into an existing machine-learning framework.
 Meanwhile the \pkg{spatialsample} has undertaken similar efforts and added the ability to use spatial resampling methods for the \pkg{tidymodels} framework [@mahoney2023].
@@
 
-```{r mlr3-ecosystem, echo=FALSE, fig.cap="Overview of the mlr3 ecosystem.
+```{r mlr3-ecosystem, echo=FALSE, fig.cap="Overview of the \\pkg{mlr3} ecosystem.
 The packages with gray dashed lines are still in development, all others have a stable interface.
-Source: @mlr3book", out.width="100%",fig.pos="ht"}
+Source: \\citet{mlr3book}.", out.width="100%",fig.pos="ht"}
 knitr::include_graphics(here::here("paper/png/mlr3-ecosystem.png"))
@@
 
-CV is also widely used to assess model performance when tuning hyperparameters of flexible machine-learning models or performing feature selection [@cai2018, @bischl2021] .
+CV is also widely used to assess model performance when tuning hyperparameters of flexible machine-learning models or performing feature selection [@cai2018; @bischl2021].
 The \pkg{mlr3} framework supports the use of CV for both approaches.
@@
 
-In nested CV specifically, an "inner" CV is performed on each CV training set, since hyperparameter tuning is an integral part of model fitting that should not be able to use information from the outer CV test set.
+In nested CV specifically, an "inner" CV is performed on each CV training set, since hyperparameter tuning is an integral part of model fitting that should not be able to use information from the outer CV test set as this would result in information leakage.
 In such scenarios it is recommended to use the same spatial resampling method for the inner CV (hyperparameter tuning) as for the outer CV (model assessment) in order to use the appropriate objective function for optimization.

### 2024-02-28 paper/submission4/jstatsoft-sub4/article.Rmd

--- 2024-01-31 paper/article.Rmd
@@
 One limitation of all the packages which provide only a subset of the machine learning workflow, e.g., resampling methods, is the sole focus on model assessment.
-Through the integration into the \pkg{mlr3} framework, \pkg{mlr3spatiotempcv} indirectly offers model selection capabilities, i.e., applying and evaluating various algorithms across different preprocessing setups (e.g., all the preprocessing ‘pipelining operators’ of \pkg{mlr3pipelines}) while being able to make use of parallel execution and enhanced logging abilities.
+Through the integration into the \pkg{mlr3} framework, \pkg{mlr3spatiotempcv} offers model selection capabilities, i.e., the seamless evaluation and application of various algorithms across different preprocessing and optimization settings while being able to make use of parallel execution and enhanced logging abilities.
 During the initial development of \pkg{mlr3spatiotempcv}, it has been the first toolbox aiming to integrate standalone spatiotemporal validation strategies into an existing machine-learning framework.
@@
 
-CV is also widely used to assess model performance when tuning hyperparameters of flexible machine-learning models or performing feature selection [@cai2018; @bischl2021].
+CV is also widely used to assess model performance when tuning hyperparameters or performing feature selection [@cai2018; @bischl2021].
 The \pkg{mlr3} framework supports the use of CV for both approaches.
-Using the CV methods introduced here, \pkg{mlr3} can therefore be used to optimized models to show an improved performance in specific spatial or spatiotemporal predictive setting [@schratz2019].
+Using the CV methods introduced here, \pkg{mlr3} can therefore be used to optimize models to show an improved performance in specific spatial or spatiotemporal predictive settings [@schratz2019].
 Such an optimization may, for example, result in a reduced maximum tree depth or increased minimum node size in the Ecuador case study, since these hyperparameter settings would result in a stronger generalization and reduced overfitting.

### 2024-03-15 paper/article.Rmd
trim code to 76 chrs

@@
 While the former two ignore the temporal and spatial dimension of the data, respectively, it has also been proposed to leave out random subsets of locations and time points [@meyer2018] or spatiotemporal clusters [@cluto].
-Details of these and other implementations are outlined in the respective subsections of Section \ref{sec:implementation}.
+Details of these and other implementations are outlined in Section \ref{sec:implementation}.
 
@@
 
-autoplot(rsmp_buffer, size = 0.8, task = task, fold_id = 1, show_omitted = TRUE)
+autoplot(rsmp_buffer, size = 0.8, task = task,
+  fold_id = 1, show_omitted = TRUE)
 ```
@@
   ) +
-  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 1.5)))
+  ggplot2::guides(colour = ggplot2::guide_legend(
+    override.aes = list(size = 1.5)))
 ```
@@
 
-autoplot(rsmp_disc, size = 0.8, task = task, fold_id = 1, show_omitted = TRUE)
+autoplot(rsmp_disc, size = 0.8, task = task,
+  fold_id = 1, show_omitted = TRUE)
 ```
@@
   ) +
-  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 0.8)))
+  ggplot2::guides(colour = ggplot2::guide_legend(
+    override.aes = list(size = 0.8)))
 
@@
 ```{r knndm-non-eval, eval=FALSE}
-points = sf::st_as_sf(task$coordinates(), crs = task$crs, coords = c("x", "y"))
+points = sf::st_as_sf(task$coordinates(), crs = task$crs,
+  coords = c("x", "y"))
 modeldomain = sf::st_as_sfc(sf::st_bbox(points))
@@
 Hierarchical clustering has been used together with the "ward.D2" link function and a target of five folds.', out.width="60%", fig.pos="ht", echo=FALSE}
-points = sf::st_as_sf(task$coordinates(), crs = task$crs, coords = c("x", "y"))
+points = sf::st_as_sf(task$coordinates(), crs = task$crs,
+  coords = c("x", "y"))
 modeldomain = sf::st_as_sfc(sf::st_bbox(points))
@@
   ) +
-  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 1.5))))
+  ggplot2::guides(colour = ggplot2::guide_legend(
+    override.aes = list(size = 1.5))))
 ```
@@
   ) *
-  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 0.8)))
+  ggplot2::guides(colour = ggplot2::guide_legend(
+    override.aes = list(size = 0.8)))
 ```

### 2024-03-15 paper/article.Rmd
use equations and proper transpose symbol

@@
 In CV for predictive model assessment, the following formal setting is considered.
-The interest is in predicting a numerical or categorical response $y$ of an object or instance using a feature vector $\mathbf{x} = (x^{(1)}, \ldots, x^{(p)})^t\in\mathbb{R}^p$ and a model $\hat{f}_\mathcal{L}$ that has been trained on a learning sample $\mathcal{L} = \{(y_i, \mathbf{x}_i),\ i = 1, ..., n\}$.
+The interest is in predicting a numerical or categorical response $y$ of an object or instance using a feature vector $\mathbf{x} = (x^\top{1}, \ldots, x^\top{p})^\top{t}\in\mathbb{R}^\top{p}$ and a model $\hat{f}_\mathcal{L}$ that has been trained on a learning sample $\mathcal{L} = \{(y_i, \mathbf{x}_i),\ i = 1, ..., n\}$.
 The goal is to estimate the expected value of the performance of $\hat{f}_\mathcal{L}$,
-\[
-\mathit{perf(\hat{f}_\mathcal{L})} := E(l(Y,\hat{f}_\mathcal{L}(X))),
-\]
+
+\begin{equation}
+\mathit{perf(\hat{f}_\mathcal{L})} := E(l(Y,\hat{f}_\mathcal{L}(X)))
+\label{eq:perf}
+\end{equation}
+
 where $l$ is a real-valued loss function, and the expected value is with respect to the probability distribution of $X$, the features of an instance $(Y,X)$ drawn randomly from the underlying population.
 This is referred to as the *actual* or *conditional* performance measure, as it is conditional on $\mathcal{L}$ [@hand1997].
-The loss function can take a variety of forms such as the misclassification error $I(Y\neq\hat{f}_\mathcal{L}(X))$ in classification, or the squared error $(Y-\hat{f}_\mathcal{L}(X))^2$ in regression, among many other possible measures.
+The loss function can take a variety of forms such as the misclassification error $I(Y\neq\hat{f}_\mathcal{L}(X))$ in classification, or the squared error $(Y-\hat{f}_\mathcal{L}(X))^\top{2}$ in regression, among many other possible measures.
 The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance measures for regression and classification (see, e.g., @hand1997 for classification, and @hyndman2006 for regression and forecasting tasks).
@@
 Since there is only a sample $\mathcal{T}$ of test data drawn from the population, one can only *estimate* the conditional performance of $\hat{f}_\mathcal{L}$:
-\[
+
+\begin{equation}
 \widehat{\mathit{perf}}_T(\hat{f}_\mathcal{L}) = \frac{1}{|\mathcal{T}|}\sum_{(Y,X)\in\mathcal{T}}l(Y,\hat{f}_\mathcal{L}(X)).
-\]
+\label{eq:condperf}
+\end{equation}
+
 This representation as a point estimator of $\mathit{perf(\hat{f}_\mathcal{L})}$ underlines the importance of using a random sample for model assessment to avoid estimation bias.
@@
 Specifically, $\mathcal{L}$ is split into $k$ partitions,
-\[
+
+\begin{equation}
 \mathcal{L} = \mathcal{L}_1 \cup \ldots \cup \mathcal{L}_k,\qquad \mathcal{L}_i\cap \mathcal{L}_j = \emptyset\quad \textrm{for all}\ i\neq j,
-\]
+\label{eq:part}
+\end{equation}
+
 and a model $\hat{f}_{(i)}$ is fitted on $\mathcal{L}_{(i)} := \mathcal{L}\setminus \mathcal{L}_i$, while $\mathcal{L}_i$ is withheld for testing.
@@
 The $k$-fold CV estimator can therefore be written as
-\[
-\widehat{\mathit{perf}}_{\mathcal{L}, CV}(f) := \frac{1}{k}\sum_{i=1}^k\widehat{\mathit{perf}}_{\mathcal{L}_i}(\hat{f}_{\mathcal{L}_{(i)}}),
-\]
+
+\begin{equation}
+\widehat{\mathit{perf}}_{\mathcal{L}, CV}(f) := \frac{1}{k}\sum_{i=1}^\top{k}\widehat{\mathit{perf}}_{\mathcal{L}_i}(\hat{f}_{\mathcal{L}_{(i)}}),
+\label{eq:kfold}
+\end{equation}
+
 where $f$ is a ML algorithm, i.e., a mapping that trains a model $\hat{f}_\mathcal{S}$ using any suitable training sample $\mathcal{S}$.
